### 2024-03-06 paper.tex
Meaningless change to test syncing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please note that whilst this template provides a
% preview of the typeset manuscript for submission, it
% will not necessarily be the final publication layout.
%
% letterpaper/a4paper: US/UK paper size toggle
% num-refs/alpha-refs: numeric/author-year citation and bibliography toggle

%\documentclass[letterpaper]{oup-contemporary}
\documentclass[a4paper,num-refs]{oup-contemporary}

%%% Journal toggle; only specific options recognised.
%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
\journal{gigascience}

\usepackage{graphicx}
\usepackage{siunitx}

% Additional packages used for code formatting
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}

% Additional packages for data model figure
\usepackage{forest}
\usepackage{tikz}
\usetikzlibrary{quotes,arrows.meta,3d}

% Local definitions to systematise tool naming
\newcommand{\sgapi}[1]{\texttt{#1}}
\newcommand{\toolname}[1]{\texttt{#1}}
\newcommand{\sgkit}{\texttt{sgkit}}

%%% Flushend: You can add this package to automatically balance the final page, but if things go awry (e.g. section contents appearing out-of-order or entire blocks or paragraphs are coloured), remove it!
% \usepackage{flushend}

% Macros etc for data model figure
\input{diagrams/data-model-preamble.tex}

% tentative
\title{Columnar VCF using Zarr}

%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
\author[1,\authfn{1},\authfn{2}]{First Author}
\author[2,\authfn{1},\authfn{2}]{Second Author}
\author[2]{Third Author}
\author[2,\authfn{1}]{Fourth Author}

\affil[1]{First Institution}
\affil[2]{Second Institution}

%%% Author Notes
\authnote{\authfn{1}abc@uni.edu; alphbeta@lab.edu}
\authnote{\authfn{2}Contributed equally.}

%%% Paper category
\papercat{Paper}

%%% "Short" author for running page header
\runningauthor{First et al.}

%%% Should only be set by an editor
\jvolume{00}
\jnumber{0}
\jyear{2024}

\begin{document}

\begin{frontmatter}
\maketitle

% The Abstract (250 words maximum) should be structured to
% include the following details:
% \textbf{Background}, the context and purpose of
% the study;
% \textbf{Results}, the main findings;
% \textbf{Conclusions}, brief
% summary and potential implications. Please minimize the use of abbreviations
% and do not cite references in the abstract.
% The Abstract (250 words maximum) should be structured to
% include the following details:
% \textbf{Background}, the context and purpose of % the study;
% \textbf{Results}, the main findings;
% \textbf{Conclusions}, brief summary and potential implications.

%% NOTE: this is much too long currently, but keeping for now so we
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
The largest dataset available today, the UK Biobank,
has a genotype matrix of approximately 1,000,000,000 rows and 500,000 columns,
along with metadata and quality control information of similar dimension.
Processing such matrices on a single computer is not feasible,
and computations must be distributed to complete in a reasonable timeframe.
Classical methods in genomics are inherently single-computer focused,
oriented around either streaming data through Unix pipelines or
storing the entire genotype matrix in memory.
The standard file format for genetic variation data (VCF: Variant Call Format)
is also fundamentally unsuited to distributed computation and is a severe
bottleneck at the Biobank scale.

\textbf{Results:}
key part of this approach is to use Zarr, a cloud-native
format for storing multi-dimensional numerical data, widely used in the
across the sciences to store petabyte-scale datasets, and with
multiple implementations.  We show that data stored in this format has
comparable compression levels to specialised methods to compress the
genotype matrix. We show that performing calculations on the genotype
matrix on a single computer is much more efficient at scale using sgkit than standard
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages.

\textbf{Conclusions:}
Such a standard, based on the language-agnostic Zarr format, would enable a
diverse ecosystem of tools to efficiently access subsets of the data without
copying, and substantially reduce research costs.
\end{abstract}

\begin{keywords}
Keyword1; keyword 2; keyword 3 (Three to ten keywords representing the main content of the article)
\end{keywords}
\end{frontmatter}

%%% Key points will be printed at top of second page
\begin{keypoints*}
\begin{itemize}
\item This is the first point
\item This is the second point
\item One last point.
\end{itemize}
\end{keypoints*}

\section{Background}

\section{Results}

\subsection{Overview of narrative for draft}

\paragraph{Storing genetic variation data}
High level discussion of
strategies used to store genotype matrix and accompanying metadata.
Emphasis on access by variant, not sample. Introduce Zarr, and
refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
and Fig~\ref{fig-data-storage}B to show that is competitive
with highly specialised approaches.
Note that the FC simulated dataset is probably a best-case
scenario for methods specialised to patterns in the genotype matrix.
Discuss storing additional data, and advantages of the
columnar approach (refs to genomics tools that do the same).

\paragraph{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.
We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations
of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks. We discuss parallelism, and why
bcftools and genozip can make limited use of available cores.
Also that single-machine parallelism is inevitably constrained
by IO (which is why savvy's thread utilisation drops off).
Figure: \ref{fig-whole-matrix-compute}. Also
supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.

\paragraph{Subsetting the genotype matrix}
Existing work overwhelmingly emphasises accessibility by variant.
We demonstrate that sgkit/zarr is good for getting subsets
in terms of both variants and samples via afdist on various subsets.
Discuss why storing AF values for crude population groupings is
intrinsically bad, and that it should be easy to do calculations
on aribitrary subsets. Also participants regularly drop-out of
large studies like UKB - do you update the AF values each time?

\subsection{Storing genetic variation data}
Storage is the foundation on which any large-scale data-driven computation
rests: how we encode and organise data on persistent storage (the format)
fundamentally affects and limits the efficiencies of downstream calculations.
Different formats are optimised for different purposes, and there are
often many competing requirements when designing a data storage format.
% TODO is there some literature we could cite on this? I'm making up the
% term "computational format" here, presumably there is lots  prior thought
% on these ideas?
For our  purposes here, it is important to distinguish between \emph{archival}
and \emph{computational} formats. An archival format is optimised for
maximum interoperability and future-proofing, and uses the simplest and
most robust technologies to ensure the long-term usability of the data.
Reasonable compression levels are important because
of storage and transfer costs, but the efficiency of data access is a
secondary concern.
A computational format, on the other hand, is optimised
for calculations, for \emph{using} the data, today. A computational format
is designed with particular data access patterns in mind, ultimately
to make a broad range of queries and downstream computations as efficient as
possible.

% Note: a quick explanation of VCF for CS people coming in fresh
VCF~\citep{danecek2011variant} is a text-based format in which each line
describes observations
and metadata for a single variant (a position on the genome at which
variation occurs in the sample). The first eight tab-delimited
columns are mandatory, and subsequent columns describe the per-sample
genotype calls and quality control data.

[WIP: currently working on weaving together the stuff below into
some sort of narrative]

As has been noted many times~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},

The shortcomings of VCF (and to a lesser degree, BCF) as a storage
format and the basis of efficient computation have long been
recognised
and many alternatives
have been proposed.

Purely archival formats try to get the maximum possible compression levels,
without much attention to query performance. Typically these use
highly specialised approaches to maximise compression performance
on genotype data.

These include:
TGC~\citep{deorowicz2013genome}
GTShark~\citep{deorowicz2019gtshark},
VCFShark~\citep{deorowicz2021vcfshark}.

SpVCF~\citep{lin2020sparse} is an extension of the VCF text
format that maximises compatibility, while gaining substantial
improvements in file sizes.

Many tools focus on balancing compression performance with
being able to run queries, typically outputting a VCF.
e.g.
BGT~\citep{li2016bgt},
GQT~\citep{layer2016efficient},
SAVVY~\citep{lefaive2021sparse},
XSI~\citep{wertenbroek2022xsi},
GTC~\citep{danek2018gtc},
GTRAC~\citep{tatwawadi2016gtrac}
genozip~\citep{lan2020genozip,lan2021genozip}
GBC~\citep{zhang2023gbc}.

SeqArray~\citep{zheng2017seqarray} and the GDS format~\citep{zheng2012high}
provide efficient storage and programmatic access in R.

Many recent tools are based on the Positional Burrows--Wheeler
Transform (PBWT)~\citep{durbin2014efficient}, which takes advantage
of the underlying haplotype structure of genomic data to provide
concise storage, and efficient haplotype matching.

The approach taken by \toolname{plink}~\citep{purcell2007plink,chang2015second} is to
store data in an uncompressed binary format. This has some major
advantages.

BGEN \citep{band2018bgen} is optimised for storing large-scale
genotype array data and stores probabilities, and is used
in particular by UK Biobank~\citep{bycroft2018genome}
[Should consider the "hard-call" nature of most of these formats]

Many methods working with large scale genetic
data now use the PBWT or variants of
it~\citep[e.g.][]{li2016bgt,lefaive2021sparse,wertenbroek2022xsi}. PBWTs have
disadvantages, however: data must be phased, and compression level
is quite sensitive to the presence of errors [citation? people have
worked around this]

Several tools provide querying interfaces in their CLIs
(reminiscent of SQL) to provide easy access to particular samples, or
variants~\cite{li2016bgt,layer2016efficient}.

More recently XSI~\citep{wertenbroek2022xsi} is an file format that
uses a combination of strategies to achieve excellent compression
performance, and can support some calculations directly on the
compressed format. It provides a command line interface and a
C API.

The SAV format is an extension of BCF that takes advantage
of the sparseness of large scale genomic datasets~\citep{lefaive2021sparse}.
They provide a command line interface and a C++ API.

\begin{figure}
    \begin{tabular}{c}
\resizebox{7cm}{!}{\input{diagrams/data-model.tex}} \\
\includegraphics[width=7cm]{figures/data-scaling}
\end{tabular}
\caption{Chunked compressed storage of the genotype matrix using Zarr.
(A) Schematic of the chunked array storage used in sgkit
(B) Compression performance compared with BCF, savvy and genozip
based on simulated data (see text for details).
\label{fig-data-storage}}
\end{figure}

High level discussion of
strategies used to store genotype matrix and accompanying metadata.
Emphasis on access by variant, not sample. Introduce Zarr, and
refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
and Fig~\ref{fig-data-storage}B to show that is competitive
with highly specialised approaches.
Note that the FC simulated dataset is probably a best-case
scenario for methods specialised to patterns in the genotype matrix.
Discuss storing additional data, and advantages of the
columnar approach (refs to genomics tools that do the same).

\subsection{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.
We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations
of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks. We discuss parallelism, and why
bcftools and genozip can make limited use of available cores.
Also that single-machine parallelism is inevitably constrained
by IO (which is why savvy's thread utilisation drops off).
Figure: \ref{fig-whole-matrix-compute}. Also
supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.

\begin{figure*}
\begin{tabular}{cc}
\includegraphics[width=6cm]{figures/whole-matrix-compute} &
\begin{tcolorbox}[width=7cm]
\begin{minted}[fontsize=\footnotesize]{python}
def sgkit_afdist(ds, num_bins=10):
    ds_vs = sgkit.variant_stats(ds).compute()
    bins = np.linspace(0, 1.0, num_bins + 1)
    bins[-1] += 0.01
    af = ds_vs.variant_allele_frequency.values[:, 1]
    pRA = 2 * af * (1 - af)
    pAA = af * af
    hets = ds_vs.variant_n_het.values
    homs = ds_vs.variant_n_hom_alt.values
    a = np.bincount(
        np.digitize(pRA, bins),
        weights=hets,
        minlength=num_bins + 1)
    b = np.bincount(
        np.digitize(pAA, bins),
        weights=homs,
        minlength=num_bins + 1)
    count = (a + b).astype(int)
    return pd.DataFrame({
        "start": bins[:-1],
        "stop": bins[1:],
        "prob_dist":
        count[1:]})

ds = sgkit.load_dataset(ds_path)
df = sgkit_afdist(ds)
\end{minted}
\end{tcolorbox}
\end{tabular}
\caption{Whole-matrix compute performance with increasing sample size.
(A) Total CPU time required to run \texttt{bcftools +afdist}
and equivalent operations in a single thread for various tools.
Elapsed time is also reported (dotted line).
(B) Relative speedup with 8 threads.
(C) The source code used for the sgkit version of afdist used in these
benchmarks. See Figure~\ref{fig-whole-matrix-compute-supplemental} for
more detailed comparisons with different numbers of threads, and comparing
the effects of HDD vs SSD storage.
\label{fig-whole-matrix-compute}}
\end{figure*}

\subsection{Subsetting the genotype matrix}
Existing work overwhelmingly emphasises accessibility by variant.
We demonstrate that sgkit/zarr is good for getting subsets
in terms of both variants and samples via afdist on various subsets.
Discuss why storing AF values for crude population groupings is
intrinsically bad, and that it should be easy to do calculations
on aribitrary subsets. Also participants regularly drop-out of
large studies like UKB - do you update the AF values each time?

\begin{figure*}
\begin{tabular}{cc}
\includegraphics[width=6cm]{figures/subset-matrix-compute} &
\begin{tcolorbox}[width=7cm]
\begin{minted}[fontsize=\footnotesize]{python}
# TODO not sure how best to illustrate this. Trying out various
# options

ds = sgkit.load_dataset(ds_path)
dss = ds.isel(
    variants=slice(...),
    samples=(slice(...)))
df = sgkit_afdist(dss)

# bcftools
bcftools view PATH -S ... -r...
| bcftools +fill-tags | bcftools +af-dist

# genozip
genocat PATH -r ... -s ...
| bcftools +fill-tags | bcftools +af-dist

# Complex query
df = pd.read_csv("sample_info.csv",
    sep="\t").set_index("sample_id")
dss = ds.isel(
    samples=df.decade[ds.sample_id.values] >= 1950)
sgkit_afdist(dss)

# Corresponding shell
% tail -n+2 sample_info.csv | awk '$2 >= 1950'
    | cut -f1 > samples.txt
bcftools view PATH -s samples.txt
    | bcftools +fill-tags | bcftools +af-dist
\end{minted}
\end{tcolorbox}
\end{tabular}
\caption{Compute performance on subsets of the matrix.
(A) Total CPU time required to run the afdist calculation for
a subset of 10 samples and 10000 variants from the middle of the matrix.
(B) Same for n/2 samples. Genozip did not run in this case for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute.
(C) A simplified version of the source code used for these benchmarks.
\label{fig-subset-matrix-compute}}
\end{figure*}

\subsection{Distributed computation}
Distributed computation has been the reality in the analysis
of genetic variation for many years,
where data is most often provided as per-chromosome VCFs.
This chunking provides a fairly straightforward way to parallelise
computations, using HPC schedulers to split work per-file.
When more fine-grained parallelism is required, users must
provide exact genome coordinates of the genome regions
as part of their submission scripts. Combining results
is a manual operation.
Workflow engines such as
Snakemake~\cite{koster2012snakemake,molder2021sustainable}
and [TODO mention WDL, Cromwell, etc?] make such analyses
much easier, but there is still a great deal of room for
mistakes.
Per-chromosome VCFs are not practical for biobank scale
projects, however, and variant call data tends to be split
into many (still large) files.
For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
hundreds of files per chromosome.
% This is horrible writing, but the point is important.
While these files provide a natural
unit for distributing work, the details of how they
are split differ and essentially there is
no interoperability among large-scale datasets because the VCFs are
so cumbersome that a separate orchestration layer is required to
access the data in parallel.

Despite this long history of de-facto distributed computation
and the clear need for better methods of accessing genetic
variation data in a distributed fashion, there is almost
no scientific literature on the topic (in stark contrast to the
study of methods to compress a VCF into a single file,
as discussed in section XXX).

[This is a really rough outline, also not sure how exactly
we sequence the ideas here. Just getting paragraph fragments
on paper. We do need to explain something of the hadoop ecosystem
to readers who are not familiar, though.]
Distributed analytics for large-scale data is a mature
techology, with many different approaches. Hadoop~\citep{white2012hadoop}
is the original method, and remains the basis of the
dominant ecosystem. HDFS. Spark [CITATION?] lets us
do more complex analytics. Parquet [CITATION?]

Although the Hadoop ecosystem has been hugely successful
in many domains, it has had a limited impact in genomics.
[List some things that have been tried for e.g variant calling
, like ADAM~\citep{nothaft2015rethinking}].
[Briefly mention two approaches using Hadoop clusters
~\citep{boufea2017managing,fan2020variant}, to not very
exiting results].

[Hadoop/Spark/Parquet are fundamentally unsuited to genetic
variation data [WHY]?]

Discuss Hail. It's awesome, has enabled current
generation of mega-scale projects. Drawbacks are the
architecture has emphasised performance etc over adaptability
and simplicity. Uses bits of Spark for X, but custom YZ .
File format is undocumented, etc.

Discuss why the sgkit architecture is suited to distributed
computation. Discuss Zarr's use in cloud stores, and existing
deployments. Discuss Dask.
We do not attempt a head-to-head performance comparison with
Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
with brief summary.
Forward ref to following sections on applications to demonstate
sgkit in a distributed context.

\section{Discussion}

Outline of a possible narrative by paragraph:
\begin{itemize}
\item When VCF was introduced as part of the work of the 1000 Genomes
project the software landscape was a real mess. It has entirely
succeeded in standardising the output of variant callers, and methods
have interoperated well for the past decade. As a method of interchange
following classical file-oriented paradigms it works well enough.
It is also a good archival format: when compressed with a well established
method like gzip and combined with the well-documented description of
the text, we can be sure that files will still be readable decades
from now.

\item VCF is holding back large-scale genomics, however, and in particular the
analysis of population scale genetic variation datasets like UKB and GeL.
Describe the current state of affairs. Variant datasets are chunked up
into reasonable sized bits (10s of gigabytes) so that they can be processed
in parallel. Doing anything with them requires interacting with workflow
managers. In practise, it is not possible to directly write programs to
work across multiple large-scale datasets: although the chunks of data
are given in VCF, orchestrating how the chunks are processed and
the results combined involves substantial work, involving the workflow
manager, and the details of how chromosomes are split into chunks.
Cite examples like the phasing pipelines for UKB from Brownings etc.

\item Distributed computing is the reality, and increasingly datasets
are being mandatorily stored in public clouds as part of
Trusted Research Environments. Cloud object stores are the reality
because it is much cheaper. POSIX file systems are emulated on top of
this. Benefits of breaking up data into manageable pieces are that
compute resources can be localised to the data. Compute resources are
most effectively obtained as short-lived, small VMs that operate
as part of a large cluster. Large vertically scaled ``pet'' servers
are much more expensive, and just not feasible for the current
scale of data. Lots of other architectural
benefits from the new reality of cloud-based storage.

\item Classical methods based around Unix pipelines and streaming
files between processes on a single machine are therefore a
significant mismatch to the reality of large-scale genetic variation
datasets. The majority of methods suggested to make working
with VCF data more efficient are based on these ideas: they all
create a single large file and regard distribution of work
in a cluster as extrinsic. Most output VCF as the result of queries,
which also assumes that the VCF data can be efficiently processed.
This is a reasonable assumption for thousands of samples, but not
for larger datasets.

\item The key issue here is that VCF is not a \emph{computational}
format. Contrast VCF with plink, which stores data on disk in a
form that can be computed on directly. Plink is limited in what
can be descibed though, does not compress, and is not designed
to distributed.

\item The large-scale analysis of genetic variation data does not
need a more efficient file format, which allows files to be downloaded
somewhat more quickly, and subsets extracted somewhat more efficiently
as VCF. It needs a standard storage platform that allows data to be
processed directly without conversion, subsetted efficiently across
both the variant and sample dimensions, with first-class support
for all the necessary additional information (GT values etc),
and that is suited to the realities of today's cloud-based
storage and processing facilities.

\item We suggest that the model used by Zarr is a good choice for this.
We have shown that chunks of the genotype matrix compressed with standard
methods are competitive with highly specialised methods for
genotype data. In reality most of the storage is for non-genotype
data anyway, where everyone is essentially doing the same thing.
Columnar storage allows getting just (e.g.) the variant positions
without reading everything. Another benefit is that by storing
fields separately in different cloud objects, they can have
different permissions. So, users can be given differential access
to read parts of the dataset, and this is managed directly
by the cloud storage. Give example.

\item A major strong point of Zarr is its simplicity. All it's doing
is compressing n-dimensional chunks of numerical data and storing
them using fixed address keys. This key-value approach means that storage
is very flexible and implementation is straightforward. There
are several implementations of the Zarr protocol, and while some of these
are rudimentary, they could be improved or replaced with relatively
little effort. Zarr has been successfully used to store petabytes
of data across different sciences --- genetic variation data is largely
arrays of numerical data, and not really that different ultimately.
Zarr is cloud native.
Zarr is an open specification, built on standard components (contrast
with Hail).

\item Zarr is not perfect. Discuss limitations around ragged arrays,
etc. Forthcoming improvements in zarr v3. The whole genotype matrix
paradigm is challenged by graph genomes as well. While this is
being worked out, there is still plenty of data to analyse as a
genotype matrix.

\item Vision. An ecosystem of tools that interoperate
based on an open, cloud-based format where distribution is naturally
managed by the stored chunks. Tools can be run on different datasets
with minimal tuning, not by writing a full layer of orchestration.
Tools read the data they need, not the entire dataset. Users read
directly from a single canonical dataset, and do not maintain
their own filtered copies. We suggest that the Zarr VCF approach
we have provided would be a good starting point for designing
such a compute platform.

\end{itemize}

\section{Methods}

\subsection{Benchmarking}
Full details of the benchmarking done in
Figure~\ref{fig-data-storage} and
Figure~\ref{fig-whole-matrix-compute}.
Quick notes:

\begin{itemize}
\item Dataset is based on subsets of French-Canadian sims. We simplify
for increasingly large subsets, keeping only variant sites. We then
convert to VCF and encode to BCF using
\texttt{tskit vcf | bcftools view -O b}.
\item We used a chunk size of XX for sgkit, after some experimentation.
This gave n files of around x MB each for the ``call\_genotypes`` array.
We used the defaults for savvy and genozip.
\item For the afdist CPU time we measure the sum of the total user and
system times required to execute the full command, as reported by GNU
time. Each tool was instructed to use one thread, where the options
were provided. For genozip, the time required to compute
\texttt{genocat file.genozip | bcftools +af-dist}. Because commands
in the pipeline execute concurrently on mutiple cores, the total CPU time is
greater than the elapsed time.
\item For the threading speedup plot, we report the time required to compute
afdist with 8 threads relative to 1 thread (a speedup of 8 is therefore
the maximum). For savvy, our C++ implementation of afdist calculates
on 8 chunks of genome, and aggegates once they are completed.
Bcftools has a ``--threads`` option, but it only applies to decoding.
\item Where possible in pipelines we use uncompressed BCF
 output \texttt{-Ou} to make processing more efficient (skip printf
and parsing costs, which are substantial).
\item We do not use BCF output in genozip because it doesn't support
it directly, only VCF (supports BCF by pipeing through bcftools).
\end{itemize}

\section{Availability of source code and requirements}

\section{Data availability}

\section{Declarations}

\subsection{List of abbreviations}
% If abbreviations are used in the text they should be defined in the text at
% first use, and a list of abbreviations should be provided in alphabetical
% order.

VCF: Variant Call Format;

\subsection{Competing Interests}

\subsection{Funding}

\subsection{Author's Contributions}

\section{Acknowledgements}

\bibliography{paper}

\clearpage
\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}

\section*{Supplementary Material}

\begin{figure*}[ht]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/whole-matrix-compute-supplemental}
	\end{center}
	\caption{\label{fig-whole-matrix-compute-supplemental}
    Limitations of vertical scaling on whole-matrix afdist. The elapsed
wall-clock time required to compute afdist on the whole genotype matrix using
different numbers of threads and storage backends (top row), and the relative
speedup obtained from threads as a fraction of optimum speedup (bottom row).
[TODO more detail and link to fig 1. Also add the times for bcftools and
genozip in col 1?]
}
\end{figure*}

\clearpage
\section*{Text graveyard}

Here's a bunch of text that contains bits that might be useful and readapted.

\subsection{Chunked storage and distributed computation}
How we store genetic data fundamentally affects how computation
can be structured, and ultimately the scalability of the calculations
we wish to perform. Classically, genetic variation data is stored via one
of two strategies. The first strategy, epitomised by
VCF~\citep{danecek2011variant} encodes all the data for a given
variant in a record, and then stores these records consecutively on file
(optionally compressed). This record-oriented approach maps well to the
powerful Unix pipelines abstraction and
has proven to be very successful.
The second classical strategy, epitomised
by the \texttt{plink} BED format~\citep{purcell2007plink}, is to
store genotype data in a structured binary format that can be
accessed directly. This memory-map approach tends to
be much faster to access, but leads to larger files and is more
limited in what can be represented. There are advantages and
disadvantages to both approaches, but one problem fundamentally
limits their utility for modern datasets: neither is well suited to
scaling-out across multiple machines.

[Probably need a paragraph here on what we've done classically
by splitting by variant. This was the dominant dimension, so that
worked fine. Now we need to also think about access-by-sample]

Given the scale of modern datasets, the ability to perform
calculations in parallel on multiple CPU cores
is now a basic prerequisite, and
running concurrently on multiple computers in a cluster
is increasingly becoming necessary. [For interactive analysis? People have
been doing concurrent computing with HPC clusters forever.]
Record-oriented variant storage such as VCF is not suitable
for  distributed computing because although it is relatively
straightforward to break workloads up by variant, we cannot efficiently
split by sample (see the section on Columnar Data below for details).
Storing data in a memory-mapped manner, where the genotypes
for each variant are written consecutive consecutively to a file,
suffers from the same problem. Although it is useful to conceptualise
the genotype matrix as two-dimensional, it is important to remember
that computer memory and file storage is ultimately one-dimensonal.
For example, plink's memory-mapped data layout linearises the 2D
genotype matrix by writing [TODO check] $n$ bytes representing
the genotypes for all samples at a variant position consecutively.
Because the genotypes for a given variant index can be found
in contiguous blocks at easily computable offsets from
this initial location in memory,
accessing the genotypes for a given variant is very efficient.
However, accessing the genotypes for a given sample at every
variant position is much less efficient because each of the $m$ bytes we
must read are $n$ bytes apart in the file.
In the distributed context, this often means
that each compute node  in a
cluster must effectively read the entire file in order to obtain the
data for a subset of samples. In practise, this means that calculations
are limited by I/0 bandwidth.

A simple and elegant solution to providing efficient access to such
large matrices across multiple dimensions is to store rectangular
\emph{chunks} rather than entire rows or columns. This is illustrated
in Fig~\ref{fig-data-storage}, where we show the basic approach
used by sgkit. By storing data in these regular chunks, we
can efficiently access data for both specific variants \emph{and}
samples.

Additionally, we can compress the individual chunks [more details].

This general strategy for storing multidimensional numerical data
has been in use for many years~\citep{folk2011overview}

This maps well to distributed computing because Zarr provides a
very flexible approach to storing the chunks. By using a native cloud store
for these chunks, compute nodes can directly access the chunks that
are needed, and not read irrelevant data.
[TODO flesh out and more stuff.]

\subsubsection{Array oriented computing}

NumPy~\citep{harris2020array} has been transformational, and provides
the basic framework for scientific computing in Python.

[ Things to cite: pandas~\citep{mckinney2010data},
BioNumpy~\citep{rand2022bionumpy}, OME-Zarr~\citep{moore2023ome}]

[NOTE: we may also want a section discussing the n-dimensional array model,
but I think this section is the right place to discuss that?]

\subsubsection{Columnar binary data}

Perhaps the most important factor limiting scalability in contemporary genomics
is the continued reliance on row-wise storage of data---most often
encoded as blockwise-compressed text.
[Explain why row-wise is bad without reference to specific genomics applications.]

The Variant Call Format (VCF) is the standard method for interchanging
genomics data~\citep{danecek2011variant}.
In VCF each row describes the observations for a set of samples
at a given position along the genome, and consists of fixed columns such as the
genome position as well as per-sample columns. [More info]
VCF was introduced as part of the 1000 Genomes project, and works reasonably
well at the scale of 1000s of samples. However, when we have hundreds of
thousands of samples it becomes extremely unwieldy
. For example, the
phased genotypes for 200K samples in UK Biobank produced by Browning [cite]
require X GB of space.
(Note that this file contains no extra INFO keys; in practice a great deal
more information is stored in the VCF leading to files that are orders of
magnitude larger, and requiring them to be split into chunks, further
increasing their unwieldiness.)
Running [a simple query that requires only looking at
one column] on this file like X required Y minutes.

[TODO be more concrete here; what is the thing we're summarising?]
This slow response time to a simple query is primarily caused by the row-wise
nature of VCF (and BCF), because in order to read a single piece of
information about a particular variant site we must read \emph{all}
the information
about that site. Then, if we wish to summarise that information over
all sites, then we must read the entire dataset.

A lesser but still significant cause of inefficiency is the use of
text to store the data. [More info]

These issues are of course well known, and there are numerous projects
that seek to alleviate them. [Discuss] They're all quite bespoke to genomics data
through.

In sgkit we use Zarr which although initially introduced to store
genomics data, has now been adopted widely in [X, Y and Z].
Zarr is [description] and has [advantages]

Also to discuss (not sure how these fit in to this specific sectio
on background technologies vs existing genomics infrastructure; can
sort out later):

\begin{itemize}
\item columnar (with small records) good for I/O elimination + sequential access.
Classical locality of reference stuff as well.
\item What does this mean with ND arrays? Slippery concept. Discuss.
\item Compression improvements from columnar.
\item Poor performance of text-based VCF has been noted many times,
and lots of prior work in this space e.g. \citep{kelleher2013processing}
\item Honourable mentions for bcftools~\citep{danecek2021twelve},
htslib~\citep{bonfield2021htslib}, CRAM~\citep{bonfield2014scramble,bonfield2022cram}
which uses a columnar approach
\end{itemize}

\subsection{Storing genetic variation data}

It is important to note that the example in Fig.~\ref{fig-data-storage}
is something of a best-base
scenario in terms genomic dataset storage, in that we are only
storing genotype data (per sample, per variant),
and, because it is error-free, is very
highly compressible. Noisier genotypes will not compress quite
so well, and storing less compressible data (such as genotype
quality values) results in a large increase in the overall
storage size. This is essentially unavoidable, but the Zarr-based
approach still has some significant advantages over VCF/BCF
[basically we can store separately and come up with distinct
compression strategies based on the data type. Dunno if we
want to get into this here. This is a complex topic with
a lot of existing work which we should review and reference,
and this is maybe not the place to do it.].

[SpVCF has some good text on high-entropy QC measures. We have to
reduce this entropy or we just can't store the data]

[Why are we reviewing all this? Well, we want to make the point
that people have been trying hard to address the problem of
how to deal with large-scale genomic data for quite a while,
and have come up with lots of specialised methods. There
are a few things to keep in mind. Our approach is to use
standard methods just on the "rectangles" of variant data
and it's already quite good. You could imagine writing specialised
compression filters for genotypes, if you wanted (maybe a
discussion item though). The key thing is the \emph{interface}.
They're all tying to the command line and to classical
sequential workflows. We have Python, and distributed compute. ]

% TEXT from HSG chapter section on compressing data

% The simplest approach to compressing such data is to organise it in a
% `variant-centric' fashion, such that all observed genotypes for a given variant
% are stored in a contiguous block. Because most variants are rare, this will
% result in long runs of identical values that compress well using standard
% methods. This is the approach taken by the Variant Call
% Format~\citep{danecek2011variant}, and its more efficient binary encoding, BCF.
% The compression levels achieved by this straightforward approach are good, and
% quite competitive with more sophisticated methods. The disadvantage
% is that many queries require full decompression of the data, which can be
% prohibitively time-consuming. The SpeedGene data format and software
% library~\citep{qiao2012handling} chooses from one of three encoding methods for
% each SNP determined by the allele frequency; \cite{sambo2014compression} extend
% this idea by identifying blocks of SNPs in LD, and adding two additional
% potential encodings utilising this information.

% An alternative to this variant-centric approach is to store genotypes in a
% `sample-centric' fashion. Here, all genotypes for a particular sample are
% stored consecutively. While this breaks the simple repetition structure of data
% stored in variant-centric form, other methods can be employed to find
% compressible structure. For example, TGC~\citep{deorowicz2013genome} compresses
% sample genotypes using a custom Lempel-Ziv style approach on runs of identity
% among samples. This explicit use of LD structure results in excellent
% compression performance (for example, 32MB for all SNPs in 1000 Genomes
% chromosome 1); unfortunately, querying a dataset requires full decompression,
% making the format unsuitable for online analysis. The GQT
% toolkit~\citep{layer2016efficient} also uses the sample-centric organisation of
% genotype data, but takes a different approach to compression. Variants are
% sorted by allele frequency (resulting in longer runs of identical values within
% samples) and then compressed using an efficient bit-vector
% representation~\citep{wu2002compressing}. The resulting file-sizes are similar to
% compressed BCF, but many queries can be performed directly on the compressed
% representation and are therefore many times faster.

% The positional Burrows-Wheeler transform~\citep{durbin2014efficient}, or PBWT,
% is another sample-centric method. Building on the success of Burrows-Wheeler
% transform applications in
% genomics~\citep{langmead2009ultrafast,li2009fast,li2009soap2}, PBWT provides
% very good compression performance and efficient algorithms for finding
% haplotype matches. The algorithm builds a representation of the data based
% on sorted haplotype prefixes in linear time, and the LD structure of the data
% ensures that the sorting orders between adjacent sites change very little
% on average. The method has been successfully applied
% to phasing~\citep{loh2016reference}, detection of IBD
% segments~\citep{naseri2017ultra}, improving the performance of the Li and
% Stephens model~\citep{lunter2016fast}, and a general query engine for genotype
% data~\citep{li2016bgt}. Recent extensions include privacy preserving
% search~\citep{shimizu2016efficient} and generalisation to the setting of graph
% genomes~\citep{novak2017graph}.

\end{document}

### 2024-03-08 src/collect_data.py
Basic version of afdist using zarr

@@ -9,6 +9,8 @@ import tempfile
 # Need to do this to avoid crashes with numba :cry:
 os.environ["SGKIT_DISABLE_NUMBA_CACHE"] = "1"
 
+import numba
+import zarr
 import psutil
 import humanize
 import numpy as np
@@ -176,6 +178,31 @@ def run_savvy_afdist(path, *, num_threads, num_sites, debug=False):
     return time_cli_command(cmd, debug)
 
 
+def zarr_afdist(path, num_bins=10):
+    root = zarr.open(path)
+    call_genotype = root["call_genotype"]
+    m = call_genotype.shape[0]
+    n = call_genotype.shape[1]
+
+    het = np.zeros(m, dtype=np.int32)
+    hom_alt = np.zeros(m, dtype=np.int32)
+    hom_ref = np.zeros(m, dtype=np.int32)
+    for j, genotypes in enumerate(call_genotype):
+        hom_ref[j], hom_alt[j], het[j] = count_genotypes(genotypes)
+    alt_count = 2 * hom_alt + het
+    af = alt_count / (n * 2)
+
+    bins = np.linspace(0, 1.0, num_bins + 1)
+    bins[-1] += 0.01
+    pRA = 2 * af * (1 - af)
+    pAA = af * af
+    a = np.bincount(np.digitize(pRA, bins), weights=het, minlength=num_bins + 1)
+    b = np.bincount(np.digitize(pAA, bins), weights=hom_alt, minlength=num_bins + 1)
+
+    count = (a + b).astype(int)
+    return pd.DataFrame({"start": bins[:-1], "stop": bins[1:], "prob_dist": count[1:]})
+
+
 def get_prob_dist(ds, num_bins=10):
     ds = sg.variant_stats(ds, merge=False).compute()
     bins = np.linspace(0, 1.0, num_bins + 1)
@@ -256,6 +283,66 @@ def run_sgkit_afdist_subset(
     return ProcessTimeResult(wall_time, sys_time, user_time)
 
 
+def _zarr_afdist_subset_worker(
+    ds_path, variant_slice, sample_slice, num_threads, debug, conn
+):
+    before = time.time()
+    assert variant_slice is None
+    assert variant_slice is None
+    assert num_threads == 1
+    df = zarr_afdist(ds_path)
+    wall_time = time.time() - before
+    cpu_times = psutil.Process().cpu_times()
+    if debug:
+        print(df)
+    conn.send(f"{wall_time} {cpu_times.user} {cpu_times.system}")
+
+
[…]

### 2024-03-08 src/collect_data.py
Add Zarr data

@@ -178,6 +178,20 @@ def run_savvy_afdist(path, *, num_threads, num_sites, debug=False):
     return time_cli_command(cmd, debug)
 
 
+@numba.njit("void(int64, int8[:], int32[:], int32[:], int32[:])")
+def count_genotypes(index, g, hom_ref, hom_alt, het):
+    n = g.shape[0] // 2
+    # NB Assuming no missing data!
+    for i in range(n):
+        j = 2 * i
+        if g[j] == 0 and g[j + 1] == 0:
+            hom_ref[index] += 1
+        elif g[j] > 0 and g[j + 1] > 0:
+            hom_alt[index] += 1
+        else:
+            het[index] += 1
+
+
 def zarr_afdist(path, num_bins=10):
     root = zarr.open(path)
     call_genotype = root["call_genotype"]
@@ -188,7 +202,7 @@ def zarr_afdist(path, num_bins=10):
     hom_alt = np.zeros(m, dtype=np.int32)
     hom_ref = np.zeros(m, dtype=np.int32)
     for j, genotypes in enumerate(call_genotype):
-        hom_ref[j], hom_alt[j], het[j] = count_genotypes(genotypes)
+        count_genotypes(j, genotypes.reshape(2 * n), hom_ref, hom_alt, het)
     alt_count = 2 * hom_alt + het
     af = alt_count / (n * 2)
 
@@ -548,23 +562,6 @@ def subset_processing_time(src, output, tool, slice_id, num_threads, debug):
                 df = pd.DataFrame(data).sort_values(["num_samples", "tool"])
                 df.to_csv(output, index=False)
                 print(df)
-
-
-@numba.njit
-def count_genotypes(g):
-    hom_ref = 0
-    hom_alt = 0
-    het = 0
-    # NB Assuming no missing data!
-    for j in range(len(g)):
-        gt = g[j]
-        if gt[0] == 0 and gt[1] == 0:
-            hom_ref += 1
-        elif gt[0] > 0 and gt[1] > 0:
-            hom_alt += 1
-        else:
-            het += 1
-    return hom_ref, hom_alt, het
 
 
 @click.command()

### 2024-03-08 src/plot.py

@@ -3,17 +3,21 @@ import matplotlib.pyplot as plt
 import humanize
 import pandas as pd
 import click
+import numpy as np
+import scipy.optimize as optimize
+
 
 sgkit_colour = "tab:blue"
 bcf_colour = "tab:orange"
 sav_colour = "tab:red"
 genozip_colour = "tab:purple"
+zarr_colour = "tab:blue"
 
 
 def plot_size(ax, df):
     colour_map = {
         "bcf": bcf_colour,
-        "sgkit": sgkit_colour,
+        "zarr": zarr_colour,
         "sav": sav_colour,
         "genozip": genozip_colour,
     }
@@ -55,7 +59,7 @@ def plot_size(ax, df):
 def plot_total_cpu(ax, df):
     colours = {
         "bcftools": bcf_colour,
-        "sgkit": sgkit_colour,
+        "zarr": zarr_colour,
         "savvy": sav_colour,
         "genozip": genozip_colour,
     }
@@ -78,7 +82,6 @@ def plot_total_cpu(ax, df):
         ax.loglog(
             dfs["num_samples"].values,
             dfs["wall_time"].values,
-            label=f"{tool}",
             linestyle=":",
             # marker=".",
             color=colours[tool],
@@ -87,13 +90,38 @@ def plot_total_cpu(ax, df):
 
         hours = total_cpu[-1] // 3600
 
-        ax.annotate(
-            f"{hours:.0f}h",
-            textcoords="offset points",
-            xytext=(15, 0),
-            xy=(row.num_samples, total_cpu[-1]),
-            xycoords="data",
-        )
+        if tool != "genozip":
+            ax.annotate(
+                f"{hours:.0f}h",
+                textcoords="offset points",
+                xytext=(15, 0),
+                xy=(row.num_samples, total_cpu[-1]),
+                xycoords="data",
+            )
+
+    # Extrapolate for genozip
+    def mulplicative_model(n, a, b):
+        # Fit a simple exponential function.
+        return a * np.power(n, b)
+
+    dfs = df[(df.threads == 1) & (df.tool == "genozip")]
+    fit_params, _ = optimize.curve_fit(
+        mulplicative_model, dfs.num_samples[2:], dfs.wall_time[2:])
[…]

### 2024-03-08 paper.tex
Update whole-matrix-compute figure

@@
 
-\begin{figure*}
-\begin{tabular}{cc}
-\includegraphics[width=6cm]{figures/whole-matrix-compute} &
-\begin{tcolorbox}[width=7cm]
-\begin{minted}[fontsize=\footnotesize]{python}
-def sgkit_afdist(ds, num_bins=10):
-    ds_vs = sgkit.variant_stats(ds).compute()
-    bins = np.linspace(0, 1.0, num_bins + 1)
-    bins[-1] += 0.01
-    af = ds_vs.variant_allele_frequency.values[:, 1]
-    pRA = 2 * af * (1 - af)
-    pAA = af * af
-    hets = ds_vs.variant_n_het.values
-    homs = ds_vs.variant_n_hom_alt.values
-    a = np.bincount(
-        np.digitize(pRA, bins),
-        weights=hets,
-        minlength=num_bins + 1)
-    b = np.bincount(
-        np.digitize(pAA, bins),
-        weights=homs,
-        minlength=num_bins + 1)
-    count = (a + b).astype(int)
-    return pd.DataFrame({
-        "start": bins[:-1],
-        "stop": bins[1:],
-        "prob_dist":
-        count[1:]})
-
-ds = sgkit.load_dataset(ds_path)
-df = sgkit_afdist(ds)
-\end{minted}
-\end{tcolorbox}
-\end{tabular}
+\begin{figure}
+\includegraphics[width=6cm]{figures/whole-matrix-compute}
 \caption{Whole-matrix compute performance with increasing sample size.
@@
 and equivalent operations in a single thread for various tools.
-Elapsed time is also reported (dotted line).
-(B) Relative speedup with 8 threads.
-(C) The source code used for the sgkit version of afdist used in these
-benchmarks. See Figure~\ref{fig-whole-matrix-compute-supplemental} for
-more detailed comparisons with different numbers of threads, and comparing
-the effects of HDD vs SSD storage.
+Elapsed time is also reported (dotted line). Run-time for genozip
+at $10^6$ samples was extrapolated by fitting an exponential.
 \label{fig-whole-matrix-compute}}
-\end{figure*}
+\end{figure}
 

### 2024-03-08 src/__init__.py
Add tested version of afdist

@@ -1,0 +1,2 @@
+# Not really a python package, just doing this so we can
+# do relative imports

### 2024-03-08 src/collect_data.py

@@ -178,44 +178,6 @@ def run_savvy_afdist(path, *, num_threads, num_sites, debug=False):
     return time_cli_command(cmd, debug)
 
 
-@numba.njit("void(int64, int8[:], int32[:], int32[:], int32[:])")
-def count_genotypes(index, g, hom_ref, hom_alt, het):
-    n = g.shape[0] // 2
-    # NB Assuming no missing data!
-    for i in range(n):
-        j = 2 * i
-        if g[j] == 0 and g[j + 1] == 0:
-            hom_ref[index] += 1
-        elif g[j] > 0 and g[j + 1] > 0:
-            hom_alt[index] += 1
-        else:
-            het[index] += 1
-
-
-def zarr_afdist(path, num_bins=10):
-    root = zarr.open(path)
-    call_genotype = root["call_genotype"]
-    m = call_genotype.shape[0]
-    n = call_genotype.shape[1]
-
-    het = np.zeros(m, dtype=np.int32)
-    hom_alt = np.zeros(m, dtype=np.int32)
-    hom_ref = np.zeros(m, dtype=np.int32)
-    for j, genotypes in enumerate(call_genotype):
-        count_genotypes(j, genotypes.reshape(2 * n), hom_ref, hom_alt, het)
-    alt_count = 2 * hom_alt + het
-    af = alt_count / (n * 2)
-
-    bins = np.linspace(0, 1.0, num_bins + 1)
-    bins[-1] += 0.01
-    pRA = 2 * af * (1 - af)
-    pAA = af * af
-    a = np.bincount(np.digitize(pRA, bins), weights=het, minlength=num_bins + 1)
-    b = np.bincount(np.digitize(pAA, bins), weights=hom_alt, minlength=num_bins + 1)
-
-    count = (a + b).astype(int)
-    return pd.DataFrame({"start": bins[:-1], "stop": bins[1:], "prob_dist": count[1:]})
-
 
 def get_prob_dist(ds, num_bins=10):
     ds = sg.variant_stats(ds, merge=False).compute()

### 2024-03-08 src/test_zarr_afdist.py

@@ -1,0 +1,125 @@
+import pytest
+import zarr
+import numpy as np
+import pandas as pd
+import numpy.testing as nt
+import msprime
+import sgkit as sg
+import bio2zarr.vcf
+import pysam
+
+
+from . import zarr_afdist
+
+
+def sgkit_afdist(ds, num_bins=10):
+    ds = sg.variant_stats(ds, merge=False).compute()
+    bins = np.linspace(0, 1.0, num_bins + 1)
+    bins[-1] += 0.0125
+    af = 1 - ds.variant_allele_count[:, 0].values / ds.variant_allele_total.values
+    pRA = 2 * af * (1 - af)
+    pAA = af * af
+    hets = ds.variant_n_het.values
+    homs = ds.variant_n_hom_alt.values
+    a = np.bincount(np.digitize(pRA, bins), weights=hets, minlength=num_bins + 1)
+    b = np.bincount(np.digitize(pAA, bins), weights=homs, minlength=num_bins + 1)
+
+    count = (a + b).astype(int)
+    return pd.DataFrame({"start": bins[:-1], "stop": bins[1:], "prob_dist": count[1:]})
+
+
+class TestSimulations:
+    def run_simulation(self, tmp_path, n, L, seed):
+        ts = msprime.sim_ancestry(
+            n, population_size=10_000, sequence_length=L, random_seed=seed
+        )
+        ts = msprime.sim_mutations(ts, rate=1e-8, random_seed=seed)
+        assert ts.num_sites > 5
+
+        vcf_file = tmp_path / "sim.vcf"
+        with open(vcf_file, "w") as f:
+            ts.write_vcf(f)
+        # This also compresses the input file
+        pysam.tabix_index(str(vcf_file), preset="vcf")
+        out = tmp_path / "example.vcf.zarr"
+        bio2zarr.vcf.convert([tmp_path / "sim.vcf.gz"], out)
+        return out
+
+    @pytest.mark.parametrize(
+        ["n", "L"],
+        [
+            (8, 10**5),
+            (8, 10**6),
+        ],
+    )
+    @pytest.mark.parametrize("seed", range(1, 10))
+    def test_afdist(self, tmp_path, n, L, seed):
+        zarr_path = self.run_simulation(tmp_path, n, L, seed)
+        afdist1 = zarr_afdist.zarr_afdist(zarr_path, 8)
+        ds = sg.load_dataset(zarr_path)
+        afdist2 = sgkit_afdist(ds, 8)
[…]

### 2024-03-08 src/collect_data.py
Tidy up

@@ -1,13 +1,11 @@
 import os
+import sys
 import pathlib
 import subprocess
 import time
 import dataclasses
 import multiprocessing
 import tempfile
-
-# Need to do this to avoid crashes with numba :cry:
-os.environ["SGKIT_DISABLE_NUMBA_CACHE"] = "1"
 
 import numba
 import zarr
@@ -18,7 +16,11 @@ import pandas as pd
 import tskit
 import click
 import sgkit as sg
-import dask.distributed
+
+
+# yuck - but simplest way to avoid changing directory structure
+sys.path.insert(0, "src")
+from zarr_afdist import zarr_afdist
 
 
 def get_file_size(file):
@@ -78,10 +80,6 @@ def genozip_version():
     cmd = "./software/genozip --version"
     out = subprocess.run(cmd, shell=True, check=True, capture_output=True)
     return out.stdout.decode()
-
-
-def sgkit_version():
-    return f"sgkit version: {sg.__version__}"
 
 
 def get_variant_slice_region(ds, variant_slice):
@@ -178,87 +176,6 @@ def run_savvy_afdist(path, *, num_threads, num_sites, debug=False):
     return time_cli_command(cmd, debug)
 
 
-
-def get_prob_dist(ds, num_bins=10):
-    ds = sg.variant_stats(ds, merge=False).compute()
-    bins = np.linspace(0, 1.0, num_bins + 1)
-    bins[-1] += 0.01
-    af = ds.variant_allele_frequency.values[:, 1]
-    pRA = 2 * af * (1 - af)
-    pAA = af * af
-    hets = ds.variant_n_het.values
-    homs = ds.variant_n_hom_alt.values
-    a = np.bincount(np.digitize(pRA, bins), weights=hets, minlength=num_bins + 1)
-    b = np.bincount(np.digitize(pAA, bins), weights=homs, minlength=num_bins + 1)
-
-    count = (a + b).astype(int)
-    return pd.DataFrame({"start": bins[:-1], "stop": bins[1:], "prob_dist": count[1:]})
-
-
-def _sgkit_afdist_subset_worker(
-    ds_path, variant_slice, sample_slice, num_threads, debug, conn
-):
-    before = time.time()
-    with dask.distributed.Client(
-        processes=False, threads_per_worker=num_threads
-    ) as client:
-        ds = sg.load_dataset(ds_path)
-        if sample_slice is not None:
-            assert variant_slice is not None
[…]

### 2024-03-08 src/test_zarr_afdist.py

@@ -12,6 +12,25 @@ import pysam
 from . import zarr_afdist
 
 
+# def _sgkit_afdist_subset_worker(
+#     ds_path, variant_slice, sample_slice, num_threads, debug, conn
+# ):
+#     before = time.time()
+#     with dask.distributed.Client(
+#         processes=False, threads_per_worker=num_threads
+#     ) as client:
+#         ds = sg.load_dataset(ds_path)
+#         if sample_slice is not None:
+#             assert variant_slice is not None
+#             ds = ds.isel(variants=variant_slice, samples=sample_slice)
+#         df = get_prob_dist(ds)
+#     wall_time = time.time() - before
+#     cpu_times = psutil.Process().cpu_times()
+#     if debug:
+#         print(df)
+#     conn.send(f"{wall_time} {cpu_times.user} {cpu_times.system}")
+
+
 def sgkit_afdist(ds, num_bins=10):
     ds = sg.variant_stats(ds, merge=False).compute()
     bins = np.linspace(0, 1.0, num_bins + 1)

### 2024-03-11 src/test_zarr_afdist.py
Implement block-wise genotype counting

@@ -48,7 +48,9 @@ def sgkit_afdist(ds, num_bins=10):
 
 
 class TestSimulations:
-    def run_simulation(self, tmp_path, n, L, seed):
+    def run_simulation(
+        self, tmp_path, n, L, seed, variant_chunk_size=None, sample_chunk_size=None
+    ):
         ts = msprime.sim_ancestry(
             n, population_size=10_000, sequence_length=L, random_seed=seed
         )
@@ -61,7 +63,12 @@ class TestSimulations:
         # This also compresses the input file
         pysam.tabix_index(str(vcf_file), preset="vcf")
         out = tmp_path / "example.vcf.zarr"
-        bio2zarr.vcf.convert([tmp_path / "sim.vcf.gz"], out)
+        bio2zarr.vcf.convert(
+            [tmp_path / "sim.vcf.gz"],
+            out,
+            chunk_length=variant_chunk_size,
+            chunk_width=sample_chunk_size,
+        )
         return out
 
     @pytest.mark.parametrize(
@@ -92,34 +99,22 @@ class TestSimulations:
     )
     @pytest.mark.parametrize("seed", range(1, 10))
     def test_classify_genotypes(self, tmp_path, n, L, seed):
-        zarr_path = self.run_simulation(tmp_path, n, L, seed)
-
-    @pytest.mark.parametrize(
-        ["n", "L"],
-        [
-            (10, 10**5),
-            (10, 10**6),
-            (100, 10**5),
-            (1000, 10**5),
-        ],
-    )
-    @pytest.mark.parametrize("seed", range(1, 10))
-    def test_classify_genotypes(self, tmp_path, n, L, seed):
-        zarr_path = self.run_simulation(tmp_path, n, L, seed)
-
-        counts = zarr_afdist.classify_genotypes(zarr.load(zarr_path)["call_genotype"])
+        zarr_path = self.run_simulation(
+            tmp_path, n, L, seed, variant_chunk_size=200, sample_chunk_size=103
+        )
+        root = zarr.open(zarr_path)
+        counts = zarr_afdist.classify_genotypes(root["call_genotype"])
         ds = sg.load_dataset(zarr_path)
         sg_counts = sg.variant_stats(ds, merge=False).compute()
         nt.assert_array_equal(sg_counts.variant_n_het, counts.het)
         nt.assert_array_equal(sg_counts.variant_n_hom_alt, counts.hom_alt)
[…]

### 2024-03-11 src/zarr_afdist.py

@@ -24,12 +24,46 @@ def count_genotypes(index, g, hom_ref, hom_alt, het, ref_count):
         ref_count[index] += (a == 0) + (b == 0)
 
 
+@numba.njit("void(int64, int8[:, :, :], int32[:], int32[:], int32[:], int32[:])")
+def count_genotypes_chunk(offset, G, hom_ref, hom_alt, het, ref_count):
+    # NB Assuming diploids and no missing data!
+    index = offset
+    for j in range(G.shape[0]):
+        for k in range(G.shape[1]):
+            a = G[j, k, 0]
+            b = G[j, k, 1]
+            if a == b:
+                if a == 0:
+                    hom_ref[index] += 1
+                else:
+                    hom_alt[index] += 1
+            else:
+                het[index] += 1
+            ref_count[index] += (a == 0) + (b == 0)
+        index += 1
+
+
 @dataclasses.dataclass
 class GenotypeCounts:
     hom_ref: list
     hom_alt: list
     het: list
     ref_count: list
+
+
+def classify_genotypes_variant_wise(call_genotype):
+    m = call_genotype.shape[0]
+    n = call_genotype.shape[1]
+
+    het = np.zeros(m, dtype=np.int32)
+    hom_alt = np.zeros(m, dtype=np.int32)
+    hom_ref = np.zeros(m, dtype=np.int32)
+    ref_count = np.zeros(m, dtype=np.int32)
+    # This way is quite a bit slower, leading to substantially higher sys time.
+    # Not clear why, since the IO should be done synchronously, but hey.
+    for j, genotypes in enumerate(call_genotype):
+        count_genotypes(j, genotypes.reshape(2 * n), hom_ref, hom_alt, het, ref_count)
+    return GenotypeCounts(hom_ref, hom_alt, het, ref_count)
 
 
 def classify_genotypes(call_genotype):
@@ -40,8 +74,12 @@ def classify_genotypes(call_genotype):
     hom_alt = np.zeros(m, dtype=np.int32)
     hom_ref = np.zeros(m, dtype=np.int32)
     ref_count = np.zeros(m, dtype=np.int32)
-    for j, genotypes in enumerate(call_genotype):
-        count_genotypes(j, genotypes.reshape(2 * n), hom_ref, hom_alt, het, ref_count)
+    j = 0
+    for v_chunk in range(call_genotype.cdata_shape[0]):
[…]

### 2024-03-11 src/test_zarr_afdist.py
Add simple mask-based subsetting

@@ -115,6 +115,35 @@ class TestSimulations:
         nt.assert_array_equal(sg_counts.variant_n_hom_alt, counts.hom_alt)
         nt.assert_array_equal(sg_counts.variant_n_hom_ref, counts.hom_ref)
 
+    @pytest.mark.parametrize(
+        ["n", "L"],
+        [
+            (10, 10**5),
+            (10, 10**6),
+            (100, 10**5),
+            (1000, 10**5),
+        ],
+    )
+    @pytest.mark.parametrize("seed", range(1, 10))
+    def test_classify_genotypes_subset(self, tmp_path, n, L, seed):
+        zarr_path = self.run_simulation(
+            tmp_path, n, L, seed, variant_chunk_size=200, sample_chunk_size=103
+        )
+        root = zarr.open(zarr_path)
+        G = root["call_genotype"]
+        m = G.shape[0]
+        n = G.shape[1]
+        variants_mask = np.ones(m, dtype=bool)
+        variants_mask[0 : m // 2] = 0
+        samples_mask = np.ones(n, dtype=bool)
+        samples_mask[n // 2 :] = 0
+        counts = zarr_afdist.classify_genotypes_subset(G, variants_mask, samples_mask)
+        ds = sg.load_dataset(zarr_path)
+        ds = ds.isel(variants=variants_mask, samples=samples_mask)
+        sg_counts = sg.variant_stats(ds, merge=False).compute()
+        nt.assert_array_equal(sg_counts.variant_n_het, counts.het)
+        nt.assert_array_equal(sg_counts.variant_n_hom_alt, counts.hom_alt)
+        nt.assert_array_equal(sg_counts.variant_n_hom_ref, counts.hom_ref)
 
     @pytest.mark.parametrize(
         ["n", "L"],
@@ -128,7 +157,7 @@ class TestSimulations:
     @pytest.mark.parametrize("seed", range(11, 15))
     def test_af(self, tmp_path, n, L, seed):
         zarr_path = self.run_simulation(tmp_path, n, L, seed)
-        counts = zarr_afdist.classify_genotypes(zarr.load(zarr_path)["call_genotype"])
+        counts = zarr_afdist.classify_genotypes(zarr.open(zarr_path)["call_genotype"])
         alt_count = 2 * n - counts.ref_count
         af = alt_count / (n * 2)
         ds = sg.load_dataset(zarr_path)

### 2024-03-11 src/zarr_afdist.py

@@ -43,6 +43,31 @@ def count_genotypes_chunk(offset, G, hom_ref, hom_alt, het, ref_count):
         index += 1
 
 
+@numba.njit(
+    "void(int64, int8[:, :, :], int8[:], int8[:], int32[:], int32[:], int32[:], int32[:])"
+)
+def count_genotypes_chunk_subset(
+    offset, G, variant_mask, sample_mask, hom_ref, hom_alt, het, ref_count
+):
+    # NB Assuming diploids and no missing data!
+    index = offset
+    for j in range(G.shape[0]):
+        if variant_mask[j]:
+            for k in range(G.shape[1]):
+                if sample_mask[k]:
+                    a = G[j, k, 0]
+                    b = G[j, k, 1]
+                    if a == b:
+                        if a == 0:
+                            hom_ref[index] += 1
+                        else:
+                            hom_alt[index] += 1
+                    else:
+                        het[index] += 1
+                    ref_count[index] += (a == 0) + (b == 0)
+            index += 1
+
+
 @dataclasses.dataclass
 class GenotypeCounts:
     hom_ref: list
@@ -68,7 +93,6 @@ def classify_genotypes_variant_wise(call_genotype):
 
 def classify_genotypes(call_genotype):
     m = call_genotype.shape[0]
-    n = call_genotype.shape[1]
 
     het = np.zeros(m, dtype=np.int32)
     hom_alt = np.zeros(m, dtype=np.int32)
@@ -83,11 +107,59 @@ def classify_genotypes(call_genotype):
     return GenotypeCounts(hom_ref, hom_alt, het, ref_count)
 
 
+def classify_genotypes_subset(call_genotype, variant_mask, sample_mask):
+    m = np.sum(variant_mask)
+
+    # Use zarr arrays to get mask chunks aligned with the main data
+    # for convenience.
+    z_variant_mask = zarr.array(
+        variant_mask, chunks=call_genotype.chunks[0], dtype=np.int8
+    )
+    z_sample_mask = zarr.array(
+        sample_mask, chunks=call_genotype.chunks[1], dtype=np.int8
+    )
+
+    het = np.zeros(m, dtype=np.int32)
+    hom_alt = np.zeros(m, dtype=np.int32)
+    hom_ref = np.zeros(m, dtype=np.int32)
+    ref_count = np.zeros(m, dtype=np.int32)
+    j = 0
+    # We should probably skip to the first non-zero chunk, but there probably
+    # isn't much difference unless we have a huge number of chunks, and we're
[…]

### 2024-03-12 paper.tex
Remove code block

@@
 \begin{figure*}
-\begin{tabular}{cc}
-\includegraphics[width=6cm]{figures/subset-matrix-compute} &
-\begin{tcolorbox}[width=7cm]
-\begin{minted}[fontsize=\footnotesize]{python}
-# TODO not sure how best to illustrate this. Trying out various
-# options
-
-ds = sgkit.load_dataset(ds_path)
-dss = ds.isel(
-    variants=slice(...),
-    samples=(slice(...)))
-df = sgkit_afdist(dss)
-
-# bcftools
-bcftools view PATH -S ... -r...
-| bcftools +fill-tags | bcftools +af-dist
-
-# genozip
-genocat PATH -r ... -s ...
-| bcftools +fill-tags | bcftools +af-dist
-
-# Complex query
-df = pd.read_csv("sample_info.csv",
-    sep="\t").set_index("sample_id")
-dss = ds.isel(
-    samples=df.decade[ds.sample_id.values] >= 1950)
-sgkit_afdist(dss)
-
-# Corresponding shell
-% tail -n+2 sample_info.csv | awk '$2 >= 1950'
-    | cut -f1 > samples.txt
-bcftools view PATH -s samples.txt
-    | bcftools +fill-tags | bcftools +af-dist
-\end{minted}
-\end{tcolorbox}
-\end{tabular}
+\includegraphics[width=6cm]{figures/subset-matrix-compute}
 \caption{Compute performance on subsets of the matrix.

### 2024-03-12 src/test_zarr_afdist.py
Update sample slicing benchmark code

@@ -89,6 +89,27 @@ class TestSimulations:
         nt.assert_array_equal(afdist1.prob_dist, afdist2.prob_dist)
 
     @pytest.mark.parametrize(
+        ["n", "L", "v_slice", "s_slice"],
+        [
+            (20, 10**5, slice(0, None), slice(0, None)),
+            (10, 10**5, slice(0, 2), slice(0, 100)),
+            (100, 10**5, slice(50, 100), slice(0, None)),
+            (100, 10**5, slice(0, None), slice(-100, None)),
+        ],
+    )
+    def test_afdist_subset(self, tmp_path, n, L, v_slice, s_slice):
+        zarr_path = self.run_simulation(tmp_path, n, L, 1234)
+        afdist1 = zarr_afdist.zarr_afdist(
+            zarr_path, variant_slice=v_slice, sample_slice=s_slice
+        )
+        ds = sg.load_dataset(zarr_path)
+        ds = ds.isel(variants=v_slice, samples=s_slice)
+        afdist2 = sgkit_afdist(ds)
+        nt.assert_array_equal(afdist1.start, afdist2.start)
+        nt.assert_array_equal(afdist1.stop, afdist2.stop)
+        nt.assert_array_equal(afdist1.prob_dist, afdist2.prob_dist)
+
+    @pytest.mark.parametrize(
         ["n", "L"],
         [
             (10, 10**5),

### 2024-03-12 src/zarr_afdist.py

@@ -149,18 +149,26 @@ def classify_genotypes_subset(call_genotype, variant_mask, sample_mask):
     return GenotypeCounts(hom_ref, hom_alt, het, ref_count)
 
 
-def zarr_afdist(path, num_bins=10):
+def zarr_afdist(path, num_bins=10, variant_slice=None, sample_slice=None):
     root = zarr.open(path)
     call_genotype = root["call_genotype"]
     m = call_genotype.shape[0]
     n = call_genotype.shape[1]
 
-    # Using the more general code is slightly slower, 35s vs 30s on one
-    # of the intermediate sized benchmarks.
-    # counts = classify_genotypes_subset(
-    #     call_genotype, np.ones(m, dtype=bool), np.ones(n, dtype=bool)
-    # )
-    counts = classify_genotypes(call_genotype)
+    if variant_slice is None and sample_slice is None:
+        # Using the more general code is slightly slower, 35s vs 30s on one
+        # of the intermediate sized benchmarks.
+        counts = classify_genotypes(call_genotype)
+    else:
+        variant_mask = np.zeros(m, dtype=bool)
+        variant_mask[variant_slice] = 1
+        sample_mask = np.zeros(n, dtype=bool)
+        sample_mask[sample_slice] = 1
+        counts = classify_genotypes_subset(
+            call_genotype, variant_mask, sample_mask
+        )
+        n = np.sum(sample_mask)
+
     alt_count = 2 * n - counts.ref_count
     af = alt_count / (n * 2)
     # print("af", af)

### 2024-03-12 src/plot.py
Data for Zarr subset matrix

@@ -278,7 +278,7 @@ def whole_matrix_compute_supplemental(time_data, output):
 def plot_subset_time(ax, df):
     colours = {
         "bcftools": bcf_colour,
-        "sgkit": sgkit_colour,
+        "zarr": zarr_colour,
         # "savvy": sav_colour,
         "genozip": genozip_colour,
     }

### 2024-03-12 paper.tex
Move some text to graveyard
Rough updates to show extrapolated genozip

@@
 
-\begin{figure*}
+\begin{figure}
 \includegraphics[width=6cm]{figures/subset-matrix-compute}
@@
 sample IDs, and the command line was therefore too long for the shell
-to execute.
-(C) A simplified version of the source code used for these benchmarks.
+to execute. [TODO Note that in this case the bcftools pipeline requires
+an additional step to compute AF etc.]
 \label{fig-subset-matrix-compute}}
-\end{figure*}
+\end{figure}
 
-\subsection{Distributed computation}
-Distributed computation has been the reality in the analysis
-of genetic variation for many years,
-where data is most often provided as per-chromosome VCFs.
-This chunking provides a fairly straightforward way to parallelise
-computations, using HPC schedulers to split work per-file.
-When more fine-grained parallelism is required, users must
-provide exact genome coordinates of the genome regions
-as part of their submission scripts. Combining results
-is a manual operation.
-Workflow engines such as
-Snakemake~\cite{koster2012snakemake,molder2021sustainable}
-and [TODO mention WDL, Cromwell, etc?] make such analyses
-much easier, but there is still a great deal of room for
-mistakes.
-Per-chromosome VCFs are not practical for biobank scale
-projects, however, and variant call data tends to be split
-into many (still large) files.
-For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
-are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
-hundreds of files per chromosome.
-% This is horrible writing, but the point is important.
-While these files provide a natural
-unit for distributing work, the details of how they
-are split differ and essentially there is
-no interoperability among large-scale datasets because the VCFs are
-so cumbersome that a separate orchestration layer is required to
-access the data in parallel.
+\subsection{Non genotype data}
 
-Despite this long history of de-facto distributed computation
-and the clear need for better methods of accessing genetic
-variation data in a distributed fashion, there is almost
-no scientific literature on the topic (in stark contrast to the
-study of methods to compress a VCF into a single file,
-as discussed in section XXX).
-
-[This is a really rough outline, also not sure how exactly
-we sequence the ideas here. Just getting paragraph fragments
-on paper. We do need to explain something of the hadoop ecosystem
-to readers who are not familiar, though.]
-Distributed analytics for large-scale data is a mature
-techology, with many different approaches. Hadoop~\citep{white2012hadoop}
-is the original method, and remains the basis of the
-dominant ecosystem. HDFS. Spark [CITATION?] lets us
-do more complex analytics. Parquet [CITATION?]
-
-Although the Hadoop ecosystem has been hugely successful
-in many domains, it has had a limited impact in genomics.
-[List some things that have been tried for e.g variant calling
-, like ADAM~\citep{nothaft2015rethinking}].
-[Briefly mention two approaches using Hadoop clusters
-~\citep{boufea2017managing,fan2020variant}, to not very
-exiting results].
-
-[Hadoop/Spark/Parquet are fundamentally unsuited to genetic
-variation data [WHY]?]
-
-Discuss Hail. It's awesome, has enabled current
-generation of mega-scale projects. Drawbacks are the
-architecture has emphasised performance etc over adaptability
-and simplicity. Uses bits of Spark for X, but custom YZ .
-File format is undocumented, etc.
-
-Discuss why the sgkit architecture is suited to distributed
-computation. Discuss Zarr's use in cloud stores, and existing
-deployments. Discuss Dask.
-We do not attempt a head-to-head performance comparison with
-Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
-with brief summary.
-Forward ref to following sections on applications to demonstate
-sgkit in a distributed context.
+\begin{table}
+\begin{tabular}{cc}
+Some & data on Genomics England\\
+\end{tabular}
+\caption{Summary of vcf-zarr performance on the Genomics England 100,000
+genomes VCF dataset.
+\label{tab-genomics-england-data}}
+\end{table}
 
@@
 
+\subsection{Distributed computation}
+Distributed computation has been the reality in the analysis
+of genetic variation for many years,
+where data is most often provided as per-chromosome VCFs.
+This chunking provides a fairly straightforward way to parallelise
+computations, using HPC schedulers to split work per-file.
+When more fine-grained parallelism is required, users must
+provide exact genome coordinates of the genome regions
+as part of their submission scripts. Combining results
+is a manual operation.
+Workflow engines such as
+Snakemake~\cite{koster2012snakemake,molder2021sustainable}
+and [TODO mention WDL, Cromwell, etc?] make such analyses
+much easier, but there is still a great deal of room for
+mistakes.
+Per-chromosome VCFs are not practical for biobank scale
+projects, however, and variant call data tends to be split
+into many (still large) files.
+For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
+are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
+hundreds of files per chromosome.
+% This is horrible writing, but the point is important.
+While these files provide a natural
+unit for distributing work, the details of how they
+are split differ and essentially there is
+no interoperability among large-scale datasets because the VCFs are
+so cumbersome that a separate orchestration layer is required to
+access the data in parallel.
+
+Despite this long history of de-facto distributed computation
+and the clear need for better methods of accessing genetic
+variation data in a distributed fashion, there is almost
+no scientific literature on the topic (in stark contrast to the
+study of methods to compress a VCF into a single file,
+as discussed in section XXX).
+
+[This is a really rough outline, also not sure how exactly
+we sequence the ideas here. Just getting paragraph fragments
+on paper. We do need to explain something of the hadoop ecosystem
+to readers who are not familiar, though.]
+Distributed analytics for large-scale data is a mature
+techology, with many different approaches. Hadoop~\citep{white2012hadoop}
+is the original method, and remains the basis of the
+dominant ecosystem. HDFS. Spark [CITATION?] lets us
+do more complex analytics. Parquet [CITATION?]
+
+Although the Hadoop ecosystem has been hugely successful
+in many domains, it has had a limited impact in genomics.
+[List some things that have been tried for e.g variant calling
+, like ADAM~\citep{nothaft2015rethinking}].
+[Briefly mention two approaches using Hadoop clusters
+~\citep{boufea2017managing,fan2020variant}, to not very
+exiting results].
+
+[Hadoop/Spark/Parquet are fundamentally unsuited to genetic
+variation data [WHY]?]
+
+Discuss Hail. It's awesome, has enabled current
+generation of mega-scale projects. Drawbacks are the
+architecture has emphasised performance etc over adaptability
+and simplicity. Uses bits of Spark for X, but custom YZ .
+File format is undocumented, etc.
+
+Discuss why the sgkit architecture is suited to distributed
+computation. Discuss Zarr's use in cloud stores, and existing
+deployments. Discuss Dask.
+We do not attempt a head-to-head performance comparison with
+Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
+with brief summary.
+Forward ref to following sections on applications to demonstate
+sgkit in a distributed context.
+
 \end{document}

### 2024-03-12 src/plot.py

@@ -106,7 +106,8 @@ def plot_total_cpu(ax, df):
 
     dfs = df[(df.threads == 1) & (df.tool == "genozip")]
     fit_params, _ = optimize.curve_fit(
-        mulplicative_model, dfs.num_samples[2:], dfs.wall_time[2:])
+        mulplicative_model, dfs.num_samples[2:], dfs.wall_time[2:]
+    )
     num_samples = df[(df.threads == 1) & (df.tool == "bcftools")].num_samples.values
     fit = mulplicative_model(num_samples, *fit_params)
     # print(fit)
@@ -121,7 +122,6 @@ def plot_total_cpu(ax, df):
         xycoords="data",
     )
     ax.legend()
-
 
 
 def plot_wall_time(ax, df, threads=1):
@@ -154,6 +154,7 @@ def plot_wall_time(ax, df, threads=1):
             xy=(row.num_samples, wall_time[-1]),
             xycoords="data",
         )
+
 
 def plot_thread_speedup(ax, df, threads):
     colours = {
@@ -247,11 +248,11 @@ def whole_matrix_compute_supplemental(time_data, output):
     ax5.sharey(ax4)
     ax6.sharey(ax4)
 
-    df_hdd = df[df.storage=="hdd"]
+    df_hdd = df[df.storage == "hdd"]
     plot_wall_time(ax1, df_hdd, 8)
     plot_thread_speedup(ax4, df_hdd, 8)
 
-    df_ssd = df[df.storage=="ssd"]
+    df_ssd = df[df.storage == "ssd"]
     plot_wall_time(ax2, df_ssd, 8)
     plot_thread_speedup(ax5, df_ssd, 8)
     plot_wall_time(ax3, df_ssd, 16)
@@ -275,7 +276,8 @@ def whole_matrix_compute_supplemental(time_data, output):
     plt.tight_layout()
     plt.savefig(output)
 
-def plot_subset_time(ax, df):
+
+def plot_subset_time(ax, df, extrapolate_genozip=False):
     colours = {
         "bcftools": bcf_colour,
         "zarr": zarr_colour,
@@ -302,7 +304,7 @@ def plot_subset_time(ax, df):
         ax.loglog(
             n,
             dfs["wall_time"].values,
-            label=f"{tool}",
+            # label=f"{tool}",
             linestyle=":",
             # marker=".",
             color=colours[tool],
@@ -319,6 +321,30 @@ def plot_subset_time(ax, df):
             xycoords="data",
         )
 
+    if extrapolate_genozip:
[…]

### 2024-03-12 paper.tex
Remove supplemental plot

@@
 
-\clearpage
-\renewcommand\thefigure{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand\thetable{S\arabic{table}}
-\setcounter{table}{0}
+% \clearpage
+% \renewcommand\thefigure{S\arabic{figure}}
+% \setcounter{figure}{0}
+% \renewcommand\thetable{S\arabic{table}}
+% \setcounter{table}{0}
 
-\section*{Supplementary Material}
+% \section*{Supplementary Material}
 
-\begin{figure*}[ht]
-	\begin{center}
-		\includegraphics[width=\textwidth]{figures/whole-matrix-compute-supplemental}
-	\end{center}
-	\caption{\label{fig-whole-matrix-compute-supplemental}
-    Limitations of vertical scaling on whole-matrix afdist. The elapsed
-wall-clock time required to compute afdist on the whole genotype matrix using
-different numbers of threads and storage backends (top row), and the relative
-speedup obtained from threads as a fraction of optimum speedup (bottom row).
-[TODO more detail and link to fig 1. Also add the times for bcftools and
-genozip in col 1?]
-}
-\end{figure*}
+% \begin{figure*}[ht]
+% 	\begin{center}
+% 		\includegraphics[width=\textwidth]{figures/whole-matrix-compute-supplemental}
+% 	\end{center}
+% 	\caption{\label{fig-whole-matrix-compute-supplemental}
+%     Limitations of vertical scaling on whole-matrix afdist. The elapsed
+% wall-clock time required to compute afdist on the whole genotype matrix using
+% different numbers of threads and storage backends (top row), and the relative
+% speedup obtained from threads as a fraction of optimum speedup (bottom row).
+% [TODO more detail and link to fig 1. Also add the times for bcftools and
+% genozip in col 1?]
+% }
+% \end{figure*}
 

### 2024-03-12 src/collect_data.py
Initial infrastructure for doing afdist with filtering

@@ -20,7 +20,7 @@ import sgkit as sg
 
 # yuck - but simplest way to avoid changing directory structure
 sys.path.insert(0, "src")
-from zarr_afdist import zarr_afdist
+from zarr_afdist import zarr_afdist, classify_genotypes_subset_filter
 
 
 def get_file_size(file):
@@ -442,6 +442,14 @@ def subset_processing_time(src, output, tool, slice_id, num_threads, debug):
 
 
 @click.command()
+@click.argument("zarr_path", type=click.Path())
+def genotype_filtering(zarr_path):
+    zarr_ds = zarr.open(zarr_path)
+    counts = classify_genotypes_subset_filter(zarr_ds)
+    print(counts)
+
+
+@click.command()
 def report_versions():
     for tool in all_tools:
         print(tool.name)
@@ -456,6 +464,7 @@ def cli():
 cli.add_command(file_size)
 cli.add_command(processing_time)
 cli.add_command(subset_processing_time)
+cli.add_command(genotype_filtering)
 cli.add_command(report_versions)
 
 

### 2024-03-12 src/zarr_afdist.py

@@ -44,7 +44,7 @@ def count_genotypes_chunk(offset, G, hom_ref, hom_alt, het, ref_count):
 
 
 @numba.njit(
-    "void(int64, int8[:, :, :], int8[:], int8[:], int32[:], int32[:], int32[:], int32[:])"
+    "void(int64, int8[:,:,:], b1[:], b1[:], int32[:], int32[:], int32[:], int32[:])"
 )
 def count_genotypes_chunk_subset(
     offset, G, variant_mask, sample_mask, hom_ref, hom_alt, het, ref_count
@@ -65,6 +65,40 @@ def count_genotypes_chunk_subset(
                     else:
                         het[index] += 1
                     ref_count[index] += (a == 0) + (b == 0)
+            index += 1
+
+
+@numba.njit(
+    "void(int64, int8[:,:,:], b1[:], b1[:], b1[:,:], int32[:], int32[:], int32[:], int32[:])"
+)
+def count_genotypes_chunk_subset_filter(
+    offset,
+    G,
+    variant_mask,
+    sample_mask,
+    genotype_mask,
+    hom_ref,
+    hom_alt,
+    het,
+    ref_count,
+):
+    # NB Assuming diploids and no missing data!
+    index = offset
+    for j in range(G.shape[0]):
+        if variant_mask[j]:
+            for k in range(G.shape[1]):
+                if sample_mask[k]:
+                    if genotype_mask[j, k]:
+                        a = G[j, k, 0]
+                        b = G[j, k, 1]
+                        if a == b:
+                            if a == 0:
+                                hom_ref[index] += 1
+                            else:
+                                hom_alt[index] += 1
+                        else:
+                            het[index] += 1
+                        ref_count[index] += (a == 0) + (b == 0)
             index += 1
 
 
@@ -112,12 +146,8 @@ def classify_genotypes_subset(call_genotype, variant_mask, sample_mask):
 
     # Use zarr arrays to get mask chunks aligned with the main data
     # for convenience.
-    z_variant_mask = zarr.array(
-        variant_mask, chunks=call_genotype.chunks[0], dtype=np.int8
-    )
-    z_sample_mask = zarr.array(
-        sample_mask, chunks=call_genotype.chunks[1], dtype=np.int8
-    )
+    z_variant_mask = zarr.array(variant_mask, chunks=call_genotype.chunks[0])
[…]

### 2024-03-13 src/collect_data.py
Some wip for doing genotype filtering

@@ -118,6 +118,25 @@ def run_bcftools_afdist_subset(
         )
         return time_cli_command(cmd, debug)
 
+def run_bcftools_afdist_filter(path, *, num_threads=1, debug=False):
+    cmd = (
+        "export BCFTOOLS_PLUGINS=software/bcftools-1.18/plugins; "
+        "/usr/bin/time -f'%S %U' "
+        # Need to run the pipeline in a subshell to make sure we're
+        # timing correctly
+        "sh -c '"
+        # bcftools view updates the AC and AN fields by default, but
+        # it doesn't update AF (which +af-dist uses). So, we use
+        # -I to prevent it updating, and then call fill-tags
+        f"software/bcftools view -I --include \"FORMAT/DP>10 & FORMAT/GQ>20\" "
+        # Output uncompressed BCF to make pipeline more efficient
+        f"-Ou --threads {num_threads} {path} | "
+        f"software/bcftools +fill-tags -Ou --threads {num_threads} | "
+        f"software/bcftools +af-dist --threads {num_threads}"
+        "'"
+    )
+    return time_cli_command(cmd, debug)
+
 
 # NOTE! These must be called on a file that has had fill-tags run on it.
 def run_bcftools_afdist(path, *, num_threads, num_sites, debug=False):
@@ -442,9 +461,12 @@ def subset_processing_time(src, output, tool, slice_id, num_threads, debug):
 
 
 @click.command()
-@click.argument("zarr_path", type=click.Path())
-def genotype_filtering(zarr_path):
-    zarr_ds = zarr.open(zarr_path)
+@click.argument("prefix", type=click.Path())
+def genotype_filtering_processing_time(prefix):
+    # run_bcftools_afdist_filter(prefix + ".vcf.gz", debug=True)
+
+    # run_zarr_afdist_filter(prefix + ".zarr", debug=True)
+    zarr_ds = zarr.open(prefix + ".zarr")
     counts = classify_genotypes_subset_filter(zarr_ds)
     print(counts)
 
@@ -464,7 +486,7 @@ def cli():
 cli.add_command(file_size)
 cli.add_command(processing_time)
 cli.add_command(subset_processing_time)
-cli.add_command(genotype_filtering)
+cli.add_command(genotype_filtering_processing_time)
 cli.add_command(report_versions)
 
 

### 2024-03-13 src/zarr_afdist.py

@@ -67,20 +67,11 @@ def count_genotypes_chunk_subset(
                     ref_count[index] += (a == 0) + (b == 0)
             index += 1
 
-
 @numba.njit(
     "void(int64, int8[:,:,:], b1[:], b1[:], b1[:,:], int32[:], int32[:], int32[:], int32[:])"
 )
 def count_genotypes_chunk_subset_filter(
-    offset,
-    G,
-    variant_mask,
-    sample_mask,
-    genotype_mask,
-    hom_ref,
-    hom_alt,
-    het,
-    ref_count,
+    offset, G, variant_mask, sample_mask, genotype_mask, hom_ref, hom_alt, het, ref_count
 ):
     # NB Assuming diploids and no missing data!
     index = offset
@@ -100,7 +91,6 @@ def count_genotypes_chunk_subset_filter(
                             het[index] += 1
                         ref_count[index] += (a == 0) + (b == 0)
             index += 1
-
 
 @dataclasses.dataclass
 class GenotypeCounts:
@@ -205,6 +195,8 @@ def classify_genotypes_subset_filter(zarr_ds, variant_mask=None, sample_mask=Non
     # We should probably skip to the first non-zero chunk, but there probably
     # isn't much difference unless we have a huge number of chunks, and we're
     # only selecting a tiny subset
+    # import tqdm
+    # for v_chunk in tqdm.tqdm(range(call_genotype.cdata_shape[0])):
     for v_chunk in range(call_genotype.cdata_shape[0]):
         variant_mask_chunk = z_variant_mask.blocks[v_chunk]
         count = np.sum(variant_mask_chunk)
@@ -214,7 +206,7 @@ def classify_genotypes_subset_filter(zarr_ds, variant_mask=None, sample_mask=Non
                 if np.sum(sample_mask_chunk) > 0:
                     DP = call_DP.blocks[v_chunk, s_chunk]
                     GQ = call_GQ.blocks[v_chunk, s_chunk]
-                    genotype_mask_chunk = (DP > 20) & (GQ > 10)
+                    genotype_mask_chunk = (DP > 20) & (GQ  > 10)
                     G = call_genotype.blocks[v_chunk, s_chunk]
                     count_genotypes_chunk_subset_filter(
                         j,
@@ -229,7 +221,6 @@ def classify_genotypes_subset_filter(zarr_ds, variant_mask=None, sample_mask=Non
                     )
             j += count
[…]

### 2024-03-13 src/collect_data.py
Add bcftools query example

@@ -118,6 +118,7 @@ def run_bcftools_afdist_subset(
         )
         return time_cli_command(cmd, debug)
 
+
 def run_bcftools_afdist_filter(path, *, num_threads=1, debug=False):
     cmd = (
         "export BCFTOOLS_PLUGINS=software/bcftools-1.18/plugins; "
@@ -128,7 +129,7 @@ def run_bcftools_afdist_filter(path, *, num_threads=1, debug=False):
         # bcftools view updates the AC and AN fields by default, but
         # it doesn't update AF (which +af-dist uses). So, we use
         # -I to prevent it updating, and then call fill-tags
-        f"software/bcftools view -I --include \"FORMAT/DP>10 & FORMAT/GQ>20\" "
+        f'software/bcftools view -I --include "FORMAT/DP>10 & FORMAT/GQ>20" '
         # Output uncompressed BCF to make pipeline more efficient
         f"-Ou --threads {num_threads} {path} | "
         f"software/bcftools +fill-tags -Ou --threads {num_threads} | "
@@ -472,6 +473,28 @@ def genotype_filtering_processing_time(prefix):
 
 
 @click.command()
+@click.argument("prefix", type=click.Path())
+def site_filtering_processing_time(prefix):
+    # bcftools command:
+
+    # bcftools query -i 'FILTER="PASS"' -f "%CHROM,%POS,%INFO/AN_EUR\n" tmp/1kg_chr20_all.vcf.gz > tmp1.csv
+    zarr_ds = zarr.open(prefix + ".zarr")
+    pass_filter = zarr_ds["variant_filter"][:, 0]
+    contig_id = zarr_ds.contig_id[:]
+    df = pd.DataFrame(
+        {
+            "CHROM": contig_id[zarr_ds["variant_contig"].vindex[pass_filter]],
+            "POS": zarr_ds["variant_position"].vindex[pass_filter],
+            "INFO/AN_EUR": zarr_ds["variant_AN_EUR"].vindex[pass_filter],
+        }
+    )
+    before = time.perf_counter()
+    df.to_csv("tmp2.csv", header=False, index=False)
+    duration = time.perf_counter() - before
+    print("csv writing took", duration, "for", len(df), " variants")
+
+
+@click.command()
 def report_versions():
     for tool in all_tools:
         print(tool.name)
@@ -487,6 +510,7 @@ cli.add_command(file_size)
 cli.add_command(processing_time)
 cli.add_command(subset_processing_time)
[…]

### 2024-03-25 scaling/run_conversion.sh
Update scripts for generating data

@@ -1,15 +1,0 @@
-#!/bin/bash 
-set -e
-
-for n in 10 100 1000 10000 100000 1000000; do
-# for n in 1000; do
-	echo "Subsetting for $n"
-
-	stem="data/chr21_n$n"
-	python3 ../src/convert.py subset-trees simulated_chrom_21.ts $n $stem.ts
-	make $stem.bcf
-	# make $stem.sgz
-	# make $stem.sav
-	# make $stem.genozip
-done
-

### 2024-03-25 scaling/subset_trees.py

@@ -1,0 +1,11 @@
+import tskit
+
+ts = tskit.load("simulated_chrom_21.ts")
+for k in range(1, 7):
+    num_individuals = 10**k
+    # Assuming diploids
+    ts = ts.simplify(ts.samples()[: 2 * num_individuals])
+    outfile = f"data/chr21_10_{k}.ts"
+    print("Writing", num_individuals, outfile)
+    ts.dump(outfile)
+

### 2024-03-25 scaling/subset_trees.py
Fix some bugs

@@ -1,10 +1,10 @@
 import tskit
 
-ts = tskit.load("simulated_chrom_21.ts")
+ts_full = tskit.load("simulated_chrom_21.ts")
 for k in range(1, 7):
     num_individuals = 10**k
     # Assuming diploids
-    ts = ts.simplify(ts.samples()[: 2 * num_individuals])
+    ts = ts_full.simplify(ts_full.samples()[: 2 * num_individuals])
     outfile = f"data/chr21_10_{k}.ts"
     print("Writing", num_individuals, outfile)
     ts.dump(outfile)

### 2024-03-25 src/convert.py

@@ -1,86 +1,0 @@
-import click
-import sgkit.io.vcf
-import tskit
-import dask.distributed
-import numpy as np
-
-
-@click.command()
-@click.argument("infile", type=click.File("rb"))
-@click.argument("num_individuals", type=click.INT)
-@click.argument("outfile", type=click.File("w"))
-def subset_trees(infile, num_individuals, outfile):
-    """
-    Subset the input trees file down to the specified number of
-    individuals
-    """
-    ts = tskit.load(infile)
-    # Assuming diploids
-    ts = ts.simplify(ts.samples()[: 2 * num_individuals])
-    ts.dump(outfile)
-
-
-# TODO replace this with a proper sgkit CLI
-@click.command()
-@click.argument("infile", type=click.Path())
-@click.argument("outfile", type=click.Path(exists=False))
-def vcf_to_sgkit_zarr(
-    infile,
-    outfile,
-):
-    """
-    Convert the input from vcf/bcf to sgkit
-    """
-    dask.config.set({'scheduler-profiler': None})
-    client = dask.distributed.Client(
-            n_workers=16, threads_per_worker=1,
-            )
-            # lifetime="100m", lifetime_stagger="5m",
-            # lifetime_restart=True)
-    print(client)
-    # Tried this, doesn't work:
-    # with dask.diagnostics.ProgressBar():
-    sgkit.io.vcf.vcf_to_zarr(infile,
-            outfile,
-            target_part_size="100MB",
-            chunk_width=10_000,
-            chunk_length=2000,
-            temp_chunk_length=1000,
-            read_chunk_length=500,
-            tempdir="tmp/")
-    print("done")
-
-
-@click.command()
-@click.argument("trees_file", type=click.Path(exists=True))
-@click.argument("sg_dataset", type=click.Path(exists=True))
-def verify_sgkit_conversion(trees_file, sg_dataset):
-    ds = sgkit.load_dataset(sg_dataset)
-    print(ds)
-    ts = tskit.load(trees_file)
-    assert ds.samples.shape[0] * 2 == ts.num_samples
-    np.testing.assert_array_equal(ts.sites_position, ds.variant_position)
-    ts_iter = ts.variants()
-    ds_iter = iter(ds.call_genotype)
-    with click.progressbar(range(ts.num_sites)) as bar:
-        for j in bar:
-            # This is probably horribly inefficient because it isn't
-            # doing things chunk-wise
[…]

### 2024-03-28 paper.tex
Initial methods bit about Zarr, Blosc and Zstd.

@@
 by IO (which is why savvy's thread utilisation drops off).
-Figure: \ref{fig-whole-matrix-compute}. Also
-supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.
+Figure: \ref{fig-whole-matrix-compute}. 
+% Also
+% supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.
 
@@
 by IO (which is why savvy's thread utilisation drops off).
-Figure: \ref{fig-whole-matrix-compute}. Also
-supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.
+% Figure: \ref{fig-whole-matrix-compute}. Also
+% supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.
 
@@
 \section{Methods}
+
+\subsection{Zarr and block-based compression}
+% Zarr is a simple layer on top of best-in-class, modern components
+In the interest of completeness it is useful to provide a high-level overview
+of Zarr and the techologies that it depends upon. Zarr is a specialised format
+for storing large-scale $n$-dimensional numerical data (arrays). Arrays
+are split into chunks, which are compressed and stored separately. Chunks are 
+addressed by their indexes along the dimensions of the array, and the 
+compressed data associated with this key. Chunks are
+are often stored in individual files, but a wide array of different
+stores are supported including cloud buckets and NoSQL databases. 
+Metadata describing the array and its properties is then stored 
+in JSON format along with the chunks. Zarr is best seen as a 
+\emph{specification} that describes the details of how chunks are 
+addressed and stored: there are multiple implementations across
+different languages.
+
+Zarr is flexible in allowing different compression codecs and 
+pre-compression filters to be specified on a per-array basis.
+Two key technologies used by Zarr are the Blosc
+% This seems like the best citation for Blosc? It is mentioned here
+meta-compressor~\cite{alted2010modern}
+and Zstandard compression algorithm~\citep{collet2021rfc}.
+Blosc is a high-performance compressor optimised for numerical
+data which uses ``blocking''~\citep{alted2010modern} to 
+optimise CPU-cache access patterns, as well as highly optimised
+bit and byte shuffle filters.  Remarkably, on highly 
+compressible datasets, Blosc decompression can be faster 
+than \texttt{memcpy}.
+Blosc is written in C, with APIs for C, Python, Julia, Rust
+and others.
+
+Blosc is a ``meta-compressor'' because it provides different 
+access to several different compression codecs. The 
+Zstandard compressor~\citep{collet2021rfc} is of particular 
+interest here as it achieves very high compression ratios
+with fast decompression speeds [see fig]. 
+% This may seem a bit off topic, but want to reassure reader that 
+% this isn't some niche technology
+Zstandard is widely
+used and is being incorporated, e.g., as a standard compression
+mechanism for HTTP traffic and to decompress Linux kernel images
+during boot.
 

### 2024-03-28 paper.tex
Stuff about vcf2zarr

@@
 during boot.
+% LZ4 is also notable - any citations for this?
+
+\subsection{The vcfzarr specification}
+The vcfzarr specification is a direct mapping from the VCF data model 
+to a columnar binary format, expressed using Zarr. [A HIGH LEVEL 
+DESCRIPTION OF THE SPEC, FOCUSING ON RATIONALE BEHIND DECISIONS]
+
+The vcfzarr specification can represent anything described by BCF
+(which is somewhat more restrictive than BCF) except for two corner
+cases related to the encoding of missing data. Firstly, vcfzarr does
+not distinguish between a field that is not present and one that 
+is present but contains missing data. [FOR EXAMPLE...]
+Because of the use of -1 to represent missing data and -2 to 
+represent dimension padding for integers, these values 
+cannot be stored. In practise this doesn't seem to be much of 
+an issue (we have not found a real VCF that contains negative 
+integers). If -1 and -2 need to be stored, a float field
+can be used without issues.
+
+\subsection{vcf2zarr}
+Converting VCF to Zarr format at Biobank scale is challenging. This main issue is 
+around memory usage: although we can view each record in the VCF one-by-one,
+we must buffer a full chunk (10,000 variants is reasonable choice) 
+in the variants dimension for each of the arrays. 
+For VCFs with many FORMAT fields and large numbers of samples this can
+easily require tens of gigabytes of RAM per worker, which makes 
+parallelism troublesome. Reading the VCF multiple times for different fields
+is possible, but would be prohibitively slow for multi-terabyte VCFs.
+
+The \texttt{vcf2zarr} utility solves this problem by first converting 
+the VCF data (which can be split across many files) into an Intermediate
+Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
+of VCFs and stores the data in per-field chunks (by default, 64MiB).
+In this way, the memory required is bounded by the size of single record 
+BCF record in htslib (plus the size of a decoded bgzf record
+[FIXME what is the correct terminology here?]), plus 64MiB for each 
+field.  The Intermediate Columnar Format stores the data for each 
+field spread across multiple partitions of the data (so that 
+multiple parts of the VCF can be read in parallel), and then 
+fixed size compressed chunks within those partitions. It also stores information
+about the number of records in each partition and chunk of a given
+field, so that the record at a particular index can be retrieved efficiently 
+later. Summaries such as the minimum and maximum value 
+of each field are also maintained, to aid choice of data types later.
+A set of VCF files can be converted to intermediate columnar 
+format in parallel on a single machine using a given number 
+of worker processes using the \texttt{explode} command,
+or can be distributed across a cluster using the 
+\texttt{dexplode-init},
+\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.
+
+Once the VCF data has converted to the intermediate columnar format,
+it can then be converted to Zarr using the \texttt{vcf2zarr encode}
+command. Alhough there is a one-to-one mapping between the VCF
+fields and the Zarr arrays, there is considerable flexibility 
+in how the data is stored, for example in integer width
+and compression and filter settings.
+By default vcf2zarr chooses integer type sizes based 
+on the range of values seen in the intermediate column format
+and reasonable compressor defaults, but these can be
+altered by generating a schema in JSON format and editing it.
+Encoding a given column (for example, \texttt{call\_AD})
+involves creating a buffer to hold a full variant-chunk of the 
+array in question, and then sequentially filling this buffer with 
+values read from the intermediate columnar format. Once the buffer 
+is full, the chunks in the samples dimension are compressed
+and written to file. [Something about parallelism and distributing]
 

### 2024-03-28 paper.tex
Cleanup

@@
 \section{Results}
-
-\subsection{Overview of narrative for draft}
-
-\paragraph{Storing genetic variation data}
-High level discussion of
-strategies used to store genotype matrix and accompanying metadata.
-Emphasis on access by variant, not sample. Introduce Zarr, and
-refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
-and Fig~\ref{fig-data-storage}B to show that is competitive
-with highly specialised approaches.
-Note that the FC simulated dataset is probably a best-case
-scenario for methods specialised to patterns in the genotype matrix.
-Discuss storing additional data, and advantages of the
-columnar approach (refs to genomics tools that do the same).
-
-\paragraph{Computing on the genotype matrix}
-Compressing data isn't ultimately the point - we want to use it.
-We discuss the problem of running computations on the entire
-genotype matrix, and how different formats limit what can be done.
-We illustrate the discussion by the afdist operation from
-the bcftools plugins package. This is perhaps a contrived example, but
-we chose it because it requires examining the entire genotype matrix
-and there is an efficient implementation available in bcftools.
-It is also a good example because the standard approach of
-divvying up the genome into chunks via HPC schedulers or workflow
-managers is quite tricky.
-We discuss the software development aspect, and how we developed
-the afdist-savvy program. We discuss the inherent limitations
-of the decode-to-vcf strategy, and how this inevitably
-leads to bottlenecks. We discuss parallelism, and why
-bcftools and genozip can make limited use of available cores.
-Also that single-machine parallelism is inevitably constrained
-by IO (which is why savvy's thread utilisation drops off).
-Figure: \ref{fig-whole-matrix-compute}. 
-% Also
-% supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.
-
-\paragraph{Subsetting the genotype matrix}
-Existing work overwhelmingly emphasises accessibility by variant.
-We demonstrate that sgkit/zarr is good for getting subsets
-in terms of both variants and samples via afdist on various subsets.
-Discuss why storing AF values for crude population groupings is
-intrinsically bad, and that it should be easy to do calculations
-on aribitrary subsets. Also participants regularly drop-out of
-large studies like UKB - do you update the AF values each time?
 
@@
 
-\subsection{Benchmarking}
+\subsection{Benchmarks}
 Full details of the benchmarking done in
@@
 greater than the elapsed time.
-\item For the threading speedup plot, we report the time required to compute
-afdist with 8 threads relative to 1 thread (a speedup of 8 is therefore
-the maximum). For savvy, our C++ implementation of afdist calculates
-on 8 chunks of genome, and aggegates once they are completed.
-Bcftools has a ``--threads`` option, but it only applies to decoding.
 \item Where possible in pipelines we use uncompressed BCF
  output \texttt{-Ou} to make processing more efficient (skip printf
-and parsing costs, which are substantial).
+and parsing costs, which are substantial). Following advice
+in~\citep{danecek2021twelve}.
 \item We do not use BCF output in genozip because it doesn't support
@@
 % \section*{Supplementary Material}
-
-% \begin{figure*}[ht]
-% 	\begin{center}
-% 		\includegraphics[width=\textwidth]{figures/whole-matrix-compute-supplemental}
-% 	\end{center}
-% 	\caption{\label{fig-whole-matrix-compute-supplemental}
-%     Limitations of vertical scaling on whole-matrix afdist. The elapsed
-% wall-clock time required to compute afdist on the whole genotype matrix using
-% different numbers of threads and storage backends (top row), and the relative
-% speedup obtained from threads as a fraction of optimum speedup (bottom row).
-% [TODO more detail and link to fig 1. Also add the times for bcftools and
-% genozip in col 1?]
-% }
-% \end{figure*}
 

### 2024-03-28 paper.tex

@@
 
-\subsection{Storing genetic variation data}
-Storage is the foundation on which any large-scale data-driven computation
-rests: how we encode and organise data on persistent storage (the format)
-fundamentally affects and limits the efficiencies of downstream calculations.
-Different formats are optimised for different purposes, and there are
-often many competing requirements when designing a data storage format.
-% TODO is there some literature we could cite on this? I'm making up the
-% term "computational format" here, presumably there is lots  prior thought
-% on these ideas?
-For our  purposes here, it is important to distinguish between \emph{archival}
-and \emph{computational} formats. An archival format is optimised for
-maximum interoperability and future-proofing, and uses the simplest and
-most robust technologies to ensure the long-term usability of the data.
-Reasonable compression levels are important because
-of storage and transfer costs, but the efficiency of data access is a
-secondary concern.
-A computational format, on the other hand, is optimised
-for calculations, for \emph{using} the data, today. A computational format
-is designed with particular data access patterns in mind, ultimately
-to make a broad range of queries and downstream computations as efficient as
-possible.
-
-% Note: a quick explanation of VCF for CS people coming in fresh
-VCF~\citep{danecek2011variant} is a text-based format in which each line
-describes observations
-and metadata for a single variant (a position on the genome at which
-variation occurs in the sample). The first eight tab-delimited
-columns are mandatory, and subsequent columns describe the per-sample
-genotype calls and quality control data.
-
-[WIP: currently working on weaving together the stuff below into
-some sort of narrative]
-
-As has been noted many times~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
-
-The shortcomings of VCF (and to a lesser degree, BCF) as a storage
-format and the basis of efficient computation have long been
-recognised
-and many alternatives
-have been proposed.
-
-Purely archival formats try to get the maximum possible compression levels,
-without much attention to query performance. Typically these use
-highly specialised approaches to maximise compression performance
-on genotype data.
-
-These include:
-TGC~\citep{deorowicz2013genome}
-GTShark~\citep{deorowicz2019gtshark},
-VCFShark~\citep{deorowicz2021vcfshark}.
-
-SpVCF~\citep{lin2020sparse} is an extension of the VCF text
-format that maximises compatibility, while gaining substantial
-improvements in file sizes.
-
-Many tools focus on balancing compression performance with
-being able to run queries, typically outputting a VCF.
-e.g.
-BGT~\citep{li2016bgt},
-GQT~\citep{layer2016efficient},
-SAVVY~\citep{lefaive2021sparse},
-XSI~\citep{wertenbroek2022xsi},
-GTC~\citep{danek2018gtc},
-GTRAC~\citep{tatwawadi2016gtrac}
-genozip~\citep{lan2020genozip,lan2021genozip}
-GBC~\citep{zhang2023gbc}.
-
-SeqArray~\citep{zheng2017seqarray} and the GDS format~\citep{zheng2012high}
-provide efficient storage and programmatic access in R.
-
-Many recent tools are based on the Positional Burrows--Wheeler
-Transform (PBWT)~\citep{durbin2014efficient}, which takes advantage
-of the underlying haplotype structure of genomic data to provide
-concise storage, and efficient haplotype matching.
-
-The approach taken by \toolname{plink}~\citep{purcell2007plink,chang2015second} is to
-store data in an uncompressed binary format. This has some major
-advantages.
-
-BGEN \citep{band2018bgen} is optimised for storing large-scale
-genotype array data and stores probabilities, and is used
-in particular by UK Biobank~\citep{bycroft2018genome}
-[Should consider the "hard-call" nature of most of these formats]
-
-Many methods working with large scale genetic
-data now use the PBWT or variants of
-it~\citep[e.g.][]{li2016bgt,lefaive2021sparse,wertenbroek2022xsi}. PBWTs have
-disadvantages, however: data must be phased, and compression level
-is quite sensitive to the presence of errors [citation? people have
-worked around this]
-
-Several tools provide querying interfaces in their CLIs
-(reminiscent of SQL) to provide easy access to particular samples, or
-variants~\cite{li2016bgt,layer2016efficient}.
-
-More recently XSI~\citep{wertenbroek2022xsi} is an file format that
-uses a combination of strategies to achieve excellent compression
-performance, and can support some calculations directly on the
-compressed format. It provides a command line interface and a
-C API.
-
-The SAV format is an extension of BCF that takes advantage
-of the sparseness of large scale genomic datasets~\citep{lefaive2021sparse}.
-They provide a command line interface and a C++ API.
+\subsection{Storing the genotype matrix}
+Although VCF is the standard format for exchanging genetic variation
+data, its limitations both in terms of compression ratios
+and query/compute performance are well known, and many methods 
+have been suggested to improve on these properties.[REVIEW METHODS]
+High level discussion of
+strategies used to store genotype matrix and accompanying metadata.
 
@@
 
-High level discussion of
-strategies used to store genotype matrix and accompanying metadata.
 Emphasis on access by variant, not sample. Introduce Zarr, and
@@
 of the decode-to-vcf strategy, and how this inevitably
-leads to bottlenecks. We discuss parallelism, and why
-bcftools and genozip can make limited use of available cores.
-Also that single-machine parallelism is inevitably constrained
-by IO (which is why savvy's thread utilisation drops off).
-% Figure: \ref{fig-whole-matrix-compute}. Also
-% supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.
+leads to bottlenecks.
 
@@
 in terms of both variants and samples via afdist on various subsets.
-Discuss why storing AF values for crude population groupings is
-intrinsically bad, and that it should be easy to do calculations
-on aribitrary subsets. Also participants regularly drop-out of
-large studies like UKB - do you update the AF values each time?
+Refer to fig~\ref{fig-subset-matrix-compute}.
 
@@
 
-\subsection{Non genotype data}
+\subsection{Case study: Genomics England 100,000 genomes}
+In this section we demonstrate the utility of vcfzarr on a large human dataset.
+[DETAILS ABOUT GeL DATA]
+We converted the data to the intermediate columnar format 
+which took X hours with a cluster of Y workers using the 
+distributed explode commands. When then converted to Zarr using 
+SOMETHING which took X hours using [RESOURCES]. In total, this gave 
+X TiB of data, spread across Y thousand files, a compression of 
+X over the original VCFs.
+The detailed breakdown of per-column compression ratios is 
+shown in Table~\ref{tab-genomics-england-data}.
 
@@
 \end{table}
+
+Table~\ref{tab-genomics-england-data} shows that the dataset size is 
+dominated by a few columns. This is well known and has been remarked 
+on several times [e.g. spVCF paper, etc]. [In this paragraph we discuss
+how some of these fields can be considerably compressed with very
+little loss of information. For example, the maximum Allele Depth (AD)
+value seen in this dataset is >3000. However, when we compute the 
+histogram of AD values we can see that something like 0.1\% of 
+AD values are > 127, so we can reasonably store these as 8 bit 
+ints. This reduces storage to X. Finish up by saying that more 
+could be done, but it's beyond the scope of this paper. Maybe note 
+that this kind of per-field analysis is easy now, e.g. computing 
+the hist took X minutes using xarray and Dask on a cluster with
+Y workers..]
+
+To illustrate the difference in query performance we ran some 
+\texttt{bcftools query} commands based on the Genomics England 
+user guides. For example, finding all variants passing 
+quality filters with an allele frequency X takes just 
+x seconds on a single core (y seconds of which is writing the 
+CSV file). The same query took X hours on one chunk, giving a 
+an approximate total time of XX hours over all Y chunks 
+in chromosomes 2.
 
@@
 and written to file. [Something about parallelism and distributing]
+
+\section{Choosing default compressor settings}
+To inform the choice of compression settings across different fields 
+in VCF data, we analysed their effect on compression ratio on 
+data from the 1000 Genomes project.
+
+% [FILL IN WITH Shadi's analysis from here:
+% https://github.com/sgkit-dev/bio2zarr/discussions/74
+% Discussed in:
+% https://github.com/sgkit-dev/vcf-zarr-publication/issues/26
 
@@
 
+Storage is the foundation on which any large-scale data-driven computation
+rests: how we encode and organise data on persistent storage (the format)
+fundamentally affects and limits the efficiencies of downstream calculations.
+Different formats are optimised for different purposes, and there are
+often many competing requirements when designing a data storage format.
+% TODO is there some literature we could cite on this? I'm making up the
+% term "computational format" here, presumably there is lots  prior thought
+% on these ideas?
+For our  purposes here, it is important to distinguish between \emph{archival}
+and \emph{computational} formats. An archival format is optimised for
+maximum interoperability and future-proofing, and uses the simplest and
+most robust technologies to ensure the long-term usability of the data.
+Reasonable compression levels are important because
+of storage and transfer costs, but the efficiency of data access is a
+secondary concern.
+A computational format, on the other hand, is optimised
+for calculations, for \emph{using} the data, today. A computational format
+is designed with particular data access patterns in mind, ultimately
+to make a broad range of queries and downstream computations as efficient as
+possible.
+
+% Note: a quick explanation of VCF for CS people coming in fresh
+VCF~\citep{danecek2011variant} is a text-based format in which each line
+describes observations
+and metadata for a single variant (a position on the genome at which
+variation occurs in the sample). The first eight tab-delimited
+columns are mandatory, and subsequent columns describe the per-sample
+genotype calls and quality control data.
+
+[WIP: currently working on weaving together the stuff below into
+some sort of narrative]
+
+As has been noted many times~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
+
+The shortcomings of VCF (and to a lesser degree, BCF) as a storage
+format and the basis of efficient computation have long been
+recognised
+and many alternatives
+have been proposed.
+
+Purely archival formats try to get the maximum possible compression levels,
+without much attention to query performance. Typically these use
+highly specialised approaches to maximise compression performance
+on genotype data.
+
+These include:
+TGC~\citep{deorowicz2013genome}
+GTShark~\citep{deorowicz2019gtshark},
+VCFShark~\citep{deorowicz2021vcfshark}.
+
+SpVCF~\citep{lin2020sparse} is an extension of the VCF text
+format that maximises compatibility, while gaining substantial
+improvements in file sizes.
+
+Many tools focus on balancing compression performance with
+being able to run queries, typically outputting a VCF.
+e.g.
+BGT~\citep{li2016bgt},
+GQT~\citep{layer2016efficient},
+SAVVY~\citep{lefaive2021sparse},
+XSI~\citep{wertenbroek2022xsi},
+GTC~\citep{danek2018gtc},
+GTRAC~\citep{tatwawadi2016gtrac}
+genozip~\citep{lan2020genozip,lan2021genozip}
+GBC~\citep{zhang2023gbc}.
+
+SeqArray~\citep{zheng2017seqarray} and the GDS format~\citep{zheng2012high}
+provide efficient storage and programmatic access in R.
+
+Many recent tools are based on the Positional Burrows--Wheeler
+Transform (PBWT)~\citep{durbin2014efficient}, which takes advantage
+of the underlying haplotype structure of genomic data to provide
+concise storage, and efficient haplotype matching.
+
+The approach taken by \toolname{plink}~\citep{purcell2007plink,chang2015second} is to
+store data in an uncompressed binary format. This has some major
+advantages.
+
+BGEN \citep{band2018bgen} is optimised for storing large-scale
+genotype array data and stores probabilities, and is used
+in particular by UK Biobank~\citep{bycroft2018genome}
+[Should consider the "hard-call" nature of most of these formats]
+
+Many methods working with large scale genetic
+data now use the PBWT or variants of
+it~\citep[e.g.][]{li2016bgt,lefaive2021sparse,wertenbroek2022xsi}. PBWTs have
+disadvantages, however: data must be phased, and compression level
+is quite sensitive to the presence of errors [citation? people have
+worked around this]
+
+Several tools provide querying interfaces in their CLIs
+(reminiscent of SQL) to provide easy access to particular samples, or
+variants~\cite{li2016bgt,layer2016efficient}.
+
+More recently XSI~\citep{wertenbroek2022xsi} is an file format that
+uses a combination of strategies to achieve excellent compression
+performance, and can support some calculations directly on the
+compressed format. It provides a command line interface and a
+C API.
+
+The SAV format is an extension of BCF that takes advantage
+of the sparseness of large scale genomic datasets~\citep{lefaive2021sparse}.
+They provide a command line interface and a C++ API.
 % TEXT from HSG chapter section on compressing data

### 2024-03-29 paper.tex
Update abstract

@@
 % tentative
-\title{Columnar VCF using Zarr}
+\title{Columnar VCF for Biobank scale datasets using Zarr}
 
@@
 \textbf{Background:}
-The largest dataset available today, the UK Biobank,
-has a genotype matrix of approximately 1,000,000,000 rows and 500,000 columns,
-along with metadata and quality control information of similar dimension.
-Processing such matrices on a single computer is not feasible,
-and computations must be distributed to complete in a reasonable timeframe.
-Classical methods in genomics are inherently single-computer focused,
-oriented around either streaming data through Unix pipelines or
-storing the entire genotype matrix in memory.
-The standard file format for genetic variation data (VCF: Variant Call Format)
-is also fundamentally unsuited to distributed computation and is a severe
-bottleneck at the Biobank scale.
+Variant Call Format (VCF) is the standard file format for interchanging
+genetic variation data and associated quality control metrics.
+It provides a well-defined data model and is central to a large ecosystem
+of tools interoperating through this format.
+The standard row-based encoding of the VCF data model (either as text
+or packed binary) emphasises efficient retrieval of all data for a given
+variant, but accessing subsets of the data within rows is inefficient.
+Biobank-scale datasets currently available 
+consist of hundreds of thousands of whole genomes with tens of terabytes
+of compressed VCF per chromosome.
+Row-based data storage is fundamentally unsuited to such datasets
+and these large VCFs represent a major computational bottleneck.
 
 \textbf{Results:}
-key part of this approach is to use Zarr, a cloud-native
-format for storing multi-dimensional numerical data, widely used in the
-across the sciences to store petabyte-scale datasets, and with
-multiple implementations.  We show that data stored in this format has
-comparable compression levels to specialised methods to compress the
-genotype matrix. We show that performing calculations on the genotype
-matrix on a single computer is much more efficient at scale using sgkit than standard
+We present the vcfzarr specification, a columnar encoding of the 
+VCF data model using Zarr which makes retrieving subsets of the 
+data much more efficient. Zarr is a cloud-native format for storing 
+multi-dimensional numerical data, widely used in scientific computing.
+We show how this format is competitive with specialised methods for 
+storing genotype data in terms of compression ratios, and also performs
+well on compressing the high-entropy quality-control fields that 
+tend to dominate. We show that performing calculations on the genotype
+matrix is much more efficient than standard
 approaches based on VCF, and is comparable with state-of-the-art methods
-using compiled languages.
+using compiled languages. We demonstrate 
+the vcfzarr format (and the vcf2zarr conversion utility) 
+on the Genomics England dataset of XX,XXX,XXX samples, showing
+an overall reduction in storage of XX, and a reduction in processing 
+time of common queries of up to XX fold.
 
 \textbf{Conclusions:}
-Such a standard, based on the language-agnostic Zarr format, would enable a
-diverse ecosystem of tools to efficiently access subsets of the data without
-copying, and substantially reduce research costs.
+Large row-encoded VCF files are a major bottleneck for current research, and 
+processing these files incurs a substantial economic cost.
+The vcfzarr specification, building on widely-used, open-source technologies
+has the potential to greatly reduce these costs,
+and may enable a diverse ecosystem of next-generation tools for analysing 
+genetic variation data.
 \end{abstract}
@@
 \section{Background}
+[Four paragraphs. (1) What is VCF, and the good things it has done. 
+(2) Explosion in dataset size. Summary of where we're at in terms of biobanks
+etc. (3) We have a problem. Existing solutions like TileDB etc
+are either proprietary, or tightly coupled to one specific software stack
+(Hail, e.g.).
+(4) We present the vcfzarr spec and the vcf2zarr conversion utility.]
 

### 2024-04-02 software/savvy-afdist/main.cpp
Simplify sav-afdist and add --decode-only option

@@ -5,9 +5,11 @@
 #include <string>
 #include <fstream>
 
-void process_slice(const std::string& filename, uint64_t start, uint64_t end, std::vector<double>& thread_af, std::vector<int>& thread_hets, std::vector<int>& thread_homs) {
+void classify_genotypes(const std::string& filename,
+        std::vector<double>& af, std::vector<int>& hets,
+        std::vector<int>& homs) {
+
     savvy::reader f(filename);
-    f.reset_bounds({start, end});
     savvy::variant var;
     std::vector<int> geno;
 
@@ -31,11 +33,29 @@ void process_slice(const std::string& filename, uint64_t start, uint64_t end, st
             }
         }
 
-        double allele_freq = total_alleles > 0 ? static_cast<double>(alt_alleles) / total_alleles : 0.0;
-        thread_af.push_back(allele_freq);
-        thread_hets.push_back(het_count);
-        thread_homs.push_back(hom_count);
+        double allele_freq = total_alleles > 0 ?
+            static_cast<double>(alt_alleles) / total_alleles : 0.0;
+        af.push_back(allele_freq);
+        hets.push_back(het_count);
+        homs.push_back(hom_count);
     }
+}
+
+void decode(const std::string& filename)
+{
+    savvy::reader f(filename);
+    savvy::variant var;
+    std::vector<int> geno;
+    int count = 0;
+
+    while (f >> var) {
+        var.get_format("GT", geno);
+        if (geno[0] >= 0) {
+            count++;
+        }
+    }
+
+    std::cout << "Decoded variants: " << count << std::endl;
 }
 
 // Use a lambda expression for finding the bin index
@@ -56,29 +76,20 @@ bool fileExists(const std::string& filename) {
 int main(int argc, char* argv[]) {
     int num_threads = std::thread::hardware_concurrency(); // Default to the number of cores
     std::string filename;
+    bool decode_only = false;
     int total_records = -1; // Initialize total_records to an invalid state
 
     for (int i = 1; i < argc; i++) {
         std::string arg = argv[i];
-        if (arg == "--threads") {
-            if (i + 1 < argc) {
-                num_threads = std::stoi(argv[++i]);
-            }
-        } else if (arg == "--sav-file") {
-            if (i + 1 < argc) {
-                filename = argv[++i];
-            }
-        } else if (arg == "--num-variants") {
[…]

### 2024-04-02 software/savvy-afdist/main.cpp
Add decoding option and some refactoring

@@ -20,6 +20,10 @@ void classify_genotypes(const std::string& filename,
         int het_count = 0;
         int hom_count = 0;
 
+        /* Note: this assumes binary alleles, and is therefore slightly
+         * different to the Zarr version. This shouldn't make any difference
+         * to the gross timings of the operations, though.
+         */
         for (size_t i = 0; i < geno.size(); i += 2) {
             int allele1 = geno[i];
             int allele2 = geno[i + 1];

### 2024-04-02 src/collect_data.py

@@ -20,7 +20,7 @@ import sgkit as sg
 
 # yuck - but simplest way to avoid changing directory structure
 sys.path.insert(0, "src")
-from zarr_afdist import zarr_afdist, classify_genotypes_subset_filter
+from zarr_afdist import zarr_afdist, classify_genotypes_subset_filter, zarr_decode
 
 
 def get_file_size(file):
@@ -190,8 +190,15 @@ def run_genozip_afdist_subset(
 def run_savvy_afdist(path, *, num_threads, num_sites, debug=False):
     cmd = (
         "/usr/bin/time -f'%S %U' "
-        f"software/savvy-afdist/sav-afdist --threads {num_threads} "
-        f"--sav-file {path} --num-variants {num_sites}"
+        f"software/savvy-afdist/sav-afdist {path}"
+    )
+    return time_cli_command(cmd, debug)
+
+
+def run_savvy_decode(path, *, debug=False):
+    cmd = (
+        "/usr/bin/time -f'%S %U' "
+        f"software/savvy-afdist/sav-afdist {path} --decode-only"
     )
     return time_cli_command(cmd, debug)
 
@@ -213,6 +220,16 @@ def zarr_afdist_worker(ds_path, num_threads, debug, conn):
     return _zarr_afdist_subset_worker(ds_path, None, None, num_threads, debug, conn)
 
 
+def zarr_decode_worker(ds_path, debug, conn):
+    before = time.time()
+    bytes_decoded = zarr_decode(ds_path)
+    wall_time = time.time() - before
+    cpu_times = psutil.Process().cpu_times()
+    if debug:
+        print(bytes_decoded)
+    conn.send(f"{wall_time} {cpu_times.user} {cpu_times.system}")
+
+
 def zarr_afdist_subset_worker(
     ds_path, variant_slice, sample_slice, num_threads, debug, conn
 ):
@@ -226,6 +243,19 @@ def run_zarr_afdist(ds_path, *, num_threads, num_sites, debug=False):
     p = multiprocessing.Process(
         target=zarr_afdist_worker, args=(ds_path, num_threads, debug, conn2)
     )
+    p.start()
+    value = conn1.recv()
+    wall_time, user_time, sys_time = map(float, value.split())
+    p.join()
+    if p.exitcode != 0:
+        raise ValueError()
+    p.close()
+    return ProcessTimeResult(wall_time, sys_time, user_time)
+
+
[…]

### 2024-04-02 src/zarr_afdist.py

@@ -4,6 +4,9 @@ import numpy as np
 import pandas as pd
 import numba
 import zarr
+import numcodecs
+
+numcodecs.nthreads = 1
 
 
 @numba.njit("void(int64, int8[:], int32[:], int32[:], int32[:], int32[:])")
@@ -67,11 +70,20 @@ def count_genotypes_chunk_subset(
                     ref_count[index] += (a == 0) + (b == 0)
             index += 1
 
+
 @numba.njit(
     "void(int64, int8[:,:,:], b1[:], b1[:], b1[:,:], int32[:], int32[:], int32[:], int32[:])"
 )
 def count_genotypes_chunk_subset_filter(
-    offset, G, variant_mask, sample_mask, genotype_mask, hom_ref, hom_alt, het, ref_count
+    offset,
+    G,
+    variant_mask,
+    sample_mask,
+    genotype_mask,
+    hom_ref,
+    hom_alt,
+    het,
+    ref_count,
 ):
     # NB Assuming diploids and no missing data!
     index = offset
@@ -91,6 +103,7 @@ def count_genotypes_chunk_subset_filter(
                             het[index] += 1
                         ref_count[index] += (a == 0) + (b == 0)
             index += 1
+
 
 @dataclasses.dataclass
 class GenotypeCounts:
@@ -206,7 +219,7 @@ def classify_genotypes_subset_filter(zarr_ds, variant_mask=None, sample_mask=Non
                 if np.sum(sample_mask_chunk) > 0:
                     DP = call_DP.blocks[v_chunk, s_chunk]
                     GQ = call_GQ.blocks[v_chunk, s_chunk]
-                    genotype_mask_chunk = (DP > 20) & (GQ  > 10)
+                    genotype_mask_chunk = (DP > 20) & (GQ > 10)
                     G = call_genotype.blocks[v_chunk, s_chunk]
                     count_genotypes_chunk_subset_filter(
                         j,
@@ -221,6 +234,7 @@ def classify_genotypes_subset_filter(zarr_ds, variant_mask=None, sample_mask=Non
                     )
             j += count
     return GenotypeCounts(hom_ref, hom_alt, het, ref_count)
+
 
 def zarr_afdist(path, num_bins=10, variant_slice=None, sample_slice=None):
     root = zarr.open(path)
@@ -237,9 +251,7 @@ def zarr_afdist(path, num_bins=10, variant_slice=None, sample_slice=None):
         variant_mask[variant_slice] = 1
[…]

### 2024-04-02 src/collect_data.py
Remove some complexity from timing code

@@ -95,7 +95,7 @@ def write_sample_names(ds, sample_slice, f):
 
 
 def run_bcftools_afdist_subset(
-    path, ds, variant_slice, sample_slice, *, num_threads, debug=False
+    path, ds, variant_slice, sample_slice, *, num_threads=1, debug=False
 ):
     region = get_variant_slice_region(ds, variant_slice)
     with tempfile.NamedTemporaryFile("w") as f:
@@ -140,7 +140,7 @@ def run_bcftools_afdist_filter(path, *, num_threads=1, debug=False):
 
 
 # NOTE! These must be called on a file that has had fill-tags run on it.
-def run_bcftools_afdist(path, *, num_threads, num_sites, debug=False):
+def run_bcftools_afdist(path, *, num_threads=1, debug=False):
     cmd = (
         "BCFTOOLS_PLUGINS=software/bcftools-1.18/plugins "
         "/usr/bin/time -f'%S %U' "
@@ -149,7 +149,7 @@ def run_bcftools_afdist(path, *, num_threads, num_sites, debug=False):
     return time_cli_command(cmd, debug)
 
 
-def run_genozip_afdist(path, *, num_threads, num_sites, debug=False):
+def run_genozip_afdist(path, *, num_threads=1, debug=False):
     cmd = (
         "export BCFTOOLS_PLUGINS=software/bcftools-1.18/plugins; "
         "/usr/bin/time -f'%S %U' "
@@ -164,7 +164,7 @@ def run_genozip_afdist(path, *, num_threads, num_sites, debug=False):
 
 
 def run_genozip_afdist_subset(
-    path, ds, variant_slice, sample_slice, *, num_threads, debug=False
+    path, ds, variant_slice, sample_slice, *, num_threads=1, debug=False
 ):
     region = get_variant_slice_region(ds, variant_slice)
     # There's no "file" option for specifying samples with genozip,
@@ -187,7 +187,7 @@ def run_genozip_afdist_subset(
     return time_cli_command(cmd, debug)
 
 
-def run_savvy_afdist(path, *, num_threads, num_sites, debug=False):
+def run_savvy_afdist(path, *, debug=False):
     cmd = (
         "/usr/bin/time -f'%S %U' "
         f"software/savvy-afdist/sav-afdist {path}"
@@ -204,10 +204,9 @@ def run_savvy_decode(path, *, debug=False):
 
 
 def _zarr_afdist_subset_worker(
[…]

### 2024-04-02 src/collect_data.py
Add plot and data for decoding
Interestingly, a lot of the Zarr time seems to be spent in blosc
shuffling, so could be possible to speed up  at the cost of
some drop in storage

@@ -318,7 +318,7 @@ all_tools = [
 
 @click.command()
 @click.argument("src", type=click.Path(), nargs=-1)
-@click.option("-o", "--output", type=click.File("w"), default="-")
+@click.option("-o", "--output", type=click.Path(), default=None)
 @click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools])
 @click.option("-s", "--storage", default="hdd")
 @click.option("--debug", is_flag=True)
@@ -361,13 +361,14 @@ def processing_time(src, output, tool, storage, debug):
                 }
             )
             df = pd.DataFrame(data).sort_values(["num_samples", "tool"])
-            df.to_csv(output, index=False)
+            if output is not None:
+                df.to_csv(output, index=False)
             print(df)
 
 
 @click.command()
 @click.argument("src", type=click.Path(), nargs=-1)
-@click.option("-o", "--output", type=click.File("w"), default="-")
+@click.option("-o", "--output", type=click.Path(), default=None)
 @click.option("--debug", is_flag=True)
 def file_size(src, output, debug):
     paths = [pathlib.Path(p) for p in sorted(src)]
@@ -413,7 +414,7 @@ def file_size(src, output, debug):
 
 @click.command()
 @click.argument("src", type=click.Path(), nargs=-1)
-@click.option("-o", "--output", type=click.File("w"), default="-")
+@click.option("-o", "--output", type=click.Path(), default=None)
 @click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools[:2]])
 @click.option("-s", "--storage", default="hdd")
 @click.option("--debug", is_flag=True)
@@ -478,7 +479,7 @@ def midslice(n, k):
 
 @click.command()
 @click.argument("src", type=click.Path(), nargs=-1)
-@click.option("-o", "--output", type=click.File("w"), default="-")
+@click.option("-o", "--output", type=click.Path(), default=None)
 @click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools])
 @click.option("-s", "--slice-id", multiple=True, default=["n10", "n/2"])
 @click.option("--debug", is_flag=True)

### 2024-04-02 src/plot.py

@@ -63,9 +63,10 @@ def plot_total_cpu(ax, df):
         "savvy": sav_colour,
         "genozip": genozip_colour,
     }
+    have_genozip = False
 
-    for tool in colours.keys():
-        dfs = df[(df.threads == 1) & (df.tool == tool)]
+    for tool in df.tool.unique():
+        dfs = df[(df.tool == tool)]
         total_cpu = dfs["user_time"].values + dfs["sys_time"].values
         ax.loglog(
             dfs["num_samples"].values,
@@ -98,89 +99,33 @@ def plot_total_cpu(ax, df):
                 xy=(row.num_samples, total_cpu[-1]),
                 xycoords="data",
             )
+        else:
+            have_genozip = True
 
     # Extrapolate for genozip
     def mulplicative_model(n, a, b):
         # Fit a simple exponential function.
         return a * np.power(n, b)
 
-    dfs = df[(df.threads == 1) & (df.tool == "genozip")]
-    fit_params, _ = optimize.curve_fit(
-        mulplicative_model, dfs.num_samples[2:], dfs.wall_time[2:]
-    )
-    num_samples = df[(df.threads == 1) & (df.tool == "bcftools")].num_samples.values
-    fit = mulplicative_model(num_samples, *fit_params)
-    # print(fit)
-
-    ax.loglog(num_samples[4:], fit[4:], linestyle=":", color="lightgray")
-    hours = fit[-1] // 3600
-    ax.annotate(
-        f"{hours:.0f}h*",
-        textcoords="offset points",
-        xytext=(15, 0),
-        xy=(num_samples[-1], fit[-1]),
-        xycoords="data",
-    )
-    ax.legend()
-
-
-def plot_wall_time(ax, df, threads=1):
-    colours = {
-        "bcftools": bcf_colour,
-        "genozip": genozip_colour,
-        "sgkit": sgkit_colour,
-        "savvy": sav_colour,
-    }
-
-    for tool in df.tool.unique():
-        dfs = df[(df.threads == threads) & (df.tool == tool)]
-        wall_time = dfs["wall_time"].values
-        ax.loglog(
-            dfs["num_samples"].values,
-            wall_time,
-            label=f"{tool}",
-            # linestyle=ls,
-            marker=".",
-            color=colours[tool],
+    if have_genozip:
[…]

### 2024-04-04 paper.tex

@@
 In this section we demonstrate the utility of vcfzarr on a large human dataset.
-[DETAILS ABOUT GeL DATA]
+Genomics England’s multi-sample VCF dataset (aggV2) is an 
+aggregate of 78,195 gVCFs from rare disease and cancer participants 
+recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
+aggV2 comprises approximately 722 million annotated single-nucleotide 
+variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
+totalling 165.3 Tib of data after \texttt{bgzip} compression. 
+The dataset is used for a variety of research purposes, ranging from 
+genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
+imputation~\cite{shi2023genomics} to 
+simple queries involving single gene 
+regions~\cite{leggatt2023genotype}~\cite{lam2023repeat}.
 We converted the data to the intermediate columnar format 

### 2024-04-08 paper.tex

@@
 
-\subsection{Storing the genotype matrix}
+\subsection{Storing genetic variation data}
 Although VCF is the standard format for exchanging genetic variation
-data, its limitations both in terms of compression ratios
-and query/compute performance are well known, and many methods 
-have been suggested to improve on these properties.[REVIEW METHODS]
-High level discussion of
-strategies used to store genotype matrix and accompanying metadata.
+data, its limitations both in terms of compression 
+and query/compute performance are well 
+known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
+and many methods 
+have been suggested to improve on these properties.
+Most approaches balance compression with
+performance on particular types of queries, 
+typically using a command line interface
+and outputting VCF text~\citep{
+layer2016efficient, %GQT
+li2016bgt, % BGT
+tatwawadi2016gtrac, % GTRAC
+danek2018gtc, % GTC
+lin2020sparse, % SpVCF
+lan2020genozip,lan2021genozip, %genozip
+lefaive2021sparse, % SAVVY
+wertenbroek2022xsi,% XSI
+zhang2023gbc}. %GBC
+Several specialised algorithms for compressing 
+the genotype matrix (i.e., just the genotype calls without additional
+VCF information) have been proposed
+\citep{deorowicz2013genome, %TGC
+deorowicz2019gtshark, %GTShark
+deorowicz2021vcfshark}, % VCFShark
+most notably the highly-influential Positional
+Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
+
+VCF is a text-based format in which each line
+describes observations
+and metadata for a single variant (usually a position on the genome
+where variation occurs in the samples)~\citep{danecek2011variant}. 
+Blocks of VCF text are compressed using \texttt{bgzip},
+which allows subsets of the file to be retrieved using secondary
+indexes stored alongside the original file~\citep{li2011tabix}.
+VCF is therefore a \emph{row-wise} format, i.e., where all of the 
+data for a record is stored continguously. BCF is similarly 
+row-wise, as are the majority of proposed alternative storage formats.
+Row-wise storage makes retrieving all information
+for a given record straightforward and efficient, and 
+allows data to be easily streamed through Unix pipelines (see Discussion).
+
+Row-wise storage works well when records are relatively small, 
+or when we typically want to analyse each record in its entirity.
+When we want to analyse only a subset of a record 
+row-wise storage can inefficient because we will usually need to
+retrieve more information than required from storage. In the case 
+of VCF (and BCF) where records are not of a fixed size and 
+are almost always compressed in blocks, accessing any information
+for all rows means retrieving and decompressing \emph{all} 
+information from all rows.
+The alternative to row-wise storage is \emph{columnar} storage:
+instead of grouping toether all the fields for a record,
+we group together all the records for a given field.
+To our knowledge, the only existing method using columnar 
+storage for genetic variation data is 
+SeqArray~\citep{zheng2017seqarray} built on 
+the GDS container format~\citep{zheng2012high}.
 
@@
 \caption{Chunked compressed storage of the genotype matrix using Zarr.
-(A) Schematic of the chunked array storage used in sgkit
+(A) Schematic of data stored under the \texttt{vcfzarr} specification.
 (B) Compression performance compared with BCF, savvy and genozip
@@
 
-Emphasis on access by variant, not sample. Introduce Zarr, and
-refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
-and Fig~\ref{fig-data-storage}B to show that is competitive
-with highly specialised approaches.
-Note that the FC simulated dataset is probably a best-case
-scenario for methods specialised to patterns in the genotype matrix.
-Discuss storing additional data, and advantages of the
-columnar approach (refs to genomics tools that do the same).
+We propose the \texttt{vcfzarr} specification which maps the 
+VCF data model into a columnar layout using the
+widely-used Zarr format (Fig~\ref{fig-data-storage}A). 
+Briefly, Zarr is a format 
+% FIXME two different uses of $n$ here in a paragraph
+for storing $n$-dimensional arrays of a fixed datatype, stored as 
+compressed chunks. In the \texttt{vcfzarr} specification, 
+each field in a VCF is mapped to an array, which is compressed 
+and stored individually allowing for efficient retrieval. 
+In particular, call-level data is stored as $m \times n$ arrays
+(for $m$ sites and $n$ samples), allowing for efficient 
+retrieval of subsets of those fields along both the 
+variants and samples axis.
+Note that this approach to storing higher dimensional 
+sets Zarr apart from more general columnar formats 
+such as Parquet [TODO more details and citations?].
+See the Methods for more detail on Zarr and the \texttt{vcfzarr}
+specification.
+
+Remarkably, the simple approach of compressing
+two dimensional chunks with the Zstandard 
+compressor~\citep{collet2021rfc} and the bit-shuffle
+filter from Blosc~\cite{alted2010modern} produces 
+compression levels competitive with specialised methods
+on the genotype matrix. In Fig~\ref{fig-data-storage}B
+we show compression performance on up to a million
+samples subset a large and highly realistic 
+simulation of French-Canadians~\cite{anderson2023on}
+(see Methods for details).
+Note that while simulations cannot capture 
+all the subtleties of real data, the allele frequency
+and population structure patterns in this dataset 
+have been shown to closely follow 
+observations~\cite{anderson2023on} and so provides 
+a reasonable data point when comparing such methods.
+% Could note here that this might represent a best 
+% case scenario for specialised methods, but maybe 
+% not worth bothering with.
 
@@
 genotype matrix, and how different formats limit what can be done.
+
+[Review computational accessibility of the available formats.
+We have those that only produce VCF in response to limited 
+query languages. Those with APIs.]
+
 We illustrate the discussion by the afdist operation from
@@
 the afdist-savvy program. We discuss the inherent limitations
+
 of the decode-to-vcf strategy, and how this inevitably
@@
 % this isn't some niche technology
-Zstandard is widely
-used and is being incorporated, e.g., as a standard compression
-mechanism for HTTP traffic and to decompress Linux kernel images
-during boot.
+Zstandard is also used in several recent VCF compression 
+methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
 % LZ4 is also notable - any citations for this?
@@
 
-As has been noted many times~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
-
-The shortcomings of VCF (and to a lesser degree, BCF) as a storage
-format and the basis of efficient computation have long been
-recognised
-and many alternatives
-have been proposed.
-
-Purely archival formats try to get the maximum possible compression levels,
-without much attention to query performance. Typically these use
-highly specialised approaches to maximise compression performance
-on genotype data.
-
-These include:
-TGC~\citep{deorowicz2013genome}
-GTShark~\citep{deorowicz2019gtshark},
-VCFShark~\citep{deorowicz2021vcfshark}.
-
-SpVCF~\citep{lin2020sparse} is an extension of the VCF text
-format that maximises compatibility, while gaining substantial
-improvements in file sizes.
-
-Many tools focus on balancing compression performance with
-being able to run queries, typically outputting a VCF.
-e.g.
-BGT~\citep{li2016bgt},
-GQT~\citep{layer2016efficient},
-SAVVY~\citep{lefaive2021sparse},
-XSI~\citep{wertenbroek2022xsi},
-GTC~\citep{danek2018gtc},
-GTRAC~\citep{tatwawadi2016gtrac}
-genozip~\citep{lan2020genozip,lan2021genozip}
-GBC~\citep{zhang2023gbc}.
-
 SeqArray~\citep{zheng2017seqarray} and the GDS format~\citep{zheng2012high}
 provide efficient storage and programmatic access in R.
-
-Many recent tools are based on the Positional Burrows--Wheeler
-Transform (PBWT)~\citep{durbin2014efficient}, which takes advantage
-of the underlying haplotype structure of genomic data to provide
-concise storage, and efficient haplotype matching.
 

### 2024-04-09 src/collect_data.py
Collect clean file-size data
- Include VCF
- Switch to stock "du" to capture all usage

@@ -23,18 +23,12 @@ sys.path.insert(0, "src")
 from zarr_afdist import zarr_afdist, classify_genotypes_subset_filter, zarr_decode
 
 
-def get_file_size(file):
-    return file.stat().st_size
-
-
-def get_dir_size(dir):
-    return sum(get_file_size(f) for f in dir.glob("**/*") if f.is_file())
-
-
-def du(file):
-    if file.is_file():
-        return get_file_size(file)
-    return get_dir_size(file)
+def du(path):
+    """
+    Return the total disk usage of path in bytes.
+    """
+    out = subprocess.run(f"du -sb {path}", shell=True, check=True, capture_output=True)
+    return int(out.stdout.decode().split()[0])
 
 
 @dataclasses.dataclass
@@ -188,10 +182,7 @@ def run_genozip_afdist_subset(
 
 
 def run_savvy_afdist(path, *, debug=False):
-    cmd = (
-        "/usr/bin/time -f'%S %U' "
-        f"software/savvy-afdist/sav-afdist {path}"
-    )
+    cmd = "/usr/bin/time -f'%S %U' " f"software/savvy-afdist/sav-afdist {path}"
     return time_cli_command(cmd, debug)
 
 
@@ -203,9 +194,7 @@ def run_savvy_decode(path, *, debug=False):
     return time_cli_command(cmd, debug)
 
 
-def _zarr_afdist_subset_worker(
-    ds_path, variant_slice, sample_slice, debug, conn
-):
+def _zarr_afdist_subset_worker(ds_path, variant_slice, sample_slice, debug, conn):
     before = time.time()
     df = zarr_afdist(ds_path, variant_slice=variant_slice, sample_slice=sample_slice)
     wall_time = time.time() - before
@@ -229,19 +218,13 @@ def zarr_decode_worker(ds_path, debug, conn):
     conn.send(f"{wall_time} {cpu_times.user} {cpu_times.system}")
 
 
-def zarr_afdist_subset_worker(
-    ds_path, variant_slice, sample_slice, debug, conn
-):
-    return _zarr_afdist_subset_worker(
-        ds_path, variant_slice, sample_slice, debug, conn
-    )
+def zarr_afdist_subset_worker(ds_path, variant_slice, sample_slice, debug, conn):
+    return _zarr_afdist_subset_worker(ds_path, variant_slice, sample_slice, debug, conn)
 
 
[…]

### 2024-04-09 paper.tex

@@
 \begin{figure}
-    \begin{tabular}{c}
-\resizebox{7cm}{!}{\input{diagrams/data-model.tex}} \\
-\includegraphics[width=7cm]{figures/data-scaling}
-\end{tabular}
-\caption{Chunked compressed storage of the genotype matrix using Zarr.
-(A) Schematic of data stored under the \texttt{vcfzarr} specification.
-(B) Compression performance compared with BCF, savvy and genozip
-based on simulated data (see text for details).
-\label{fig-data-storage}}
+\resizebox{225pt}{!}{\input{diagrams/data-model.tex}}
+\caption{Columnar storage of VCF data in chunked, compressed form using Zarr.
+[TODO more detail, including brief overview of how data maps from 
+VCF to vcfzarr in the examples]
+\label{fig-data-model}}
 \end{figure}
@@
 VCF data model into a columnar layout using the
-widely-used Zarr format (Fig~\ref{fig-data-storage}A). 
+widely-used Zarr format (Fig~\ref{fig-data-model}). 
 Briefly, Zarr is a format 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[]{figures/data-scaling}
+\end{center}
+\caption{Compression performance on simulated genotypes.
+Comparison of total stored bytes for VCF data produced 
+by subsets of a large simulation of French-Canadians.
+Sizes for $10^6$ samples are shown on the right. Sizes 
+for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
+\label{fig-data-storage}}
+\end{figure}
+
 Remarkably, the simple approach of compressing
@@
 compression levels competitive with specialised methods
-on the genotype matrix. In Fig~\ref{fig-data-storage}B
+on the genotype matrix. In Fig~\ref{fig-data-storage}
 we show compression performance on up to a million

### 2024-04-09 src/collect_data.py
Rename command

@@ -303,7 +303,7 @@ all_tools = [
 @click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools])
 @click.option("-s", "--storage", default="hdd")
 @click.option("--debug", is_flag=True)
-def processing_time(src, output, tool, storage, debug):
+def whole_matrix_compute(src, output, tool, storage, debug):
     if len(src) == 0:
         raise ValueError("Need at least one input file!")
     tool_map = {t.name: t for t in all_tools}
@@ -573,7 +573,7 @@ def cli():
 
 
 cli.add_command(file_size)
-cli.add_command(processing_time)
+cli.add_command(whole_matrix_compute)
 cli.add_command(decoding_time)
 cli.add_command(subset_processing_time)
 cli.add_command(genotype_filtering_processing_time)

### 2024-04-09 paper.tex
Some narrative

@@
 
-\subsection{Computing on the genotype matrix}
-Compressing data isn't ultimately the point - we want to use it.
-We discuss the problem of running computations on the entire
-genotype matrix, and how different formats limit what can be done.
-
-[Review computational accessibility of the available formats.
-We have those that only produce VCF in response to limited 
-query languages. Those with APIs.]
-
-We illustrate the discussion by the afdist operation from
-the bcftools plugins package. This is perhaps a contrived example, but
-we chose it because it requires examining the entire genotype matrix
-and there is an efficient implementation available in bcftools.
-It is also a good example because the standard approach of
-divvying up the genome into chunks via HPC schedulers or workflow
-managers is quite tricky.
-We discuss the software development aspect, and how we developed
-the afdist-savvy program. We discuss the inherent limitations
-
-of the decode-to-vcf strategy, and how this inevitably
-leads to bottlenecks.
+\subsection{Calculating with the genotype matrix}
+Storing genetic variation data compactly is important, but it is also
+important that we can analyse the data efficiently. Bioinformatics 
+workflows tend to emphasise text files and command line utilities 
+that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}. 
+Thus, many tools that compress VCF data provide a command line 
+utility with a query language to restrict the records
+examined, perform some pre-specified calculations and finally 
+output some text, typically VCF or tab/comma separated values
+[TODO check these citations]~\citep{
+layer2016efficient, %GQT
+li2016bgt, % BGT
+tatwawadi2016gtrac, % GTRAC
+danek2018gtc, % GTC
+lin2020sparse, % SpVCF
+lan2020genozip,lan2021genozip, %genozip
+lefaive2021sparse, % SAVVY
+wertenbroek2022xsi,% XSI
+zhang2023gbc}. %GBC
+These pre-defined calculations are by necessity limited in scope, however,
+and the volumes of text involved in Biobank scale datasets
+make the classical approach of custom
+analyses via Unix utilities in pipelines prohibitively slow. Thus, 
+methods have begun to provide Application Programming Interfaces
+(APIs), providing efficient access to genotype and other VCF data
+[TODO check these citations]~\cite[e.g.][]{lefaive2021sparse,
+wertenbroek2022xsi,zhang2023gbc}.
+
+To demonstrate the accessibility of genotype data and efficiency with 
+which calculations can be performed under the different formats,
+we use the \texttt{bcftools +af-dist} plugin as an example of such a whole-matrix
+operation. 
+% The details of the \texttt{af-dist} operation are not important:
+% as an example of a whole-matrix operation. 
+We chose this particular operation for several
+reasons. First, it is a simple but non-trival calculation that requires 
+examining every element in the genotype matrix.
+Secondly, it produces a small volume of output (a table of
+deviations from Hardy-Weinberg expectations in ten allele frequency
+bins) and therefore the time spent outputting results is negligible.
+Finally, it has an efficient implementation written using the 
+\texttt{htslib} C API~\citep{bonfield2021htslib}, distributed as 
+\texttt{bcftools} plugin.
 
@@
 \caption{Whole-matrix compute performance with increasing sample size.
-(A) Total CPU time required to run \texttt{bcftools +afdist}
+Total CPU time required to run \texttt{bcftools +af-dist}
 and equivalent operations in a single thread for various tools.
@@
 \end{figure}
+
+Figure~\ref{fig-whole-matrix-compute} shows the results of 
+running \texttt{bcftools +af-dist} and equivalent operations 
+on the data of Figure~\ref{fig-data-storage}.
+
+% We illustrate the discussion by the afdist operation from
+% the bcftools plugins package. This is perhaps a contrived example, but
+% we chose it because it requires examining the entire genotype matrix
+% and there is an efficient implementation available in bcftools.
+% It is also a good example because the standard approach of
+% divvying up the genome into chunks via HPC schedulers or workflow
+% managers is quite tricky.
+% We discuss the software development aspect, and how we developed
+% the afdist-savvy program. We discuss the inherent limitations
 

### 2024-04-09 src/plot.py
Update whole matrix compute fig

@@ -18,13 +18,16 @@ sav_colour = "tab:red"
 genozip_colour = "tab:purple"
 zarr_colour = "tab:blue"
 
+
 def one_panel_fig(**kwargs):
     # The columnwidth of the format is ~250pt, which is
     # 3 15/32 inch, = 3.46
     width = 3.46
-    fig, ax, = plt.subplots(1, 1, figsize=(width, 2 * width / 3), **kwargs)
+    (
+        fig,
+        ax,
+    ) = plt.subplots(1, 1, figsize=(width, 2 * width / 3), **kwargs)
     return fig, ax
-
 
 
 def plot_size(ax, df, label_y_offset=None):
@@ -47,7 +50,7 @@ def plot_size(ax, df, label_y_offset=None):
             dfs["size"].values,
             ".-",
             color=colour,
-            label= "vcf.gz" if tool == "vcf" else tool
+            label="vcf.gz" if tool == "vcf" else tool,
         )
         row = dfs.iloc[-1]
         size = row["size"] / GB
@@ -72,22 +75,24 @@ def plot_size(ax, df, label_y_offset=None):
     ax2.set_xticklabels([humanize.metric(m) for m in num_sites])
 
 
-def plot_total_cpu(ax, df):
+def plot_total_cpu(ax, df, toolname=None):
     colours = {
         "bcftools": bcf_colour,
+        "genozip": genozip_colour,
         "zarr": zarr_colour,
         "savvy": sav_colour,
-        "genozip": genozip_colour,
     }
     have_genozip = False
+    toolname = {} if toolname is None else toolname
 
-    for tool in df.tool.unique():
+    # for tool in df.tool.unique():
+    for tool in colours.keys():
         dfs = df[(df.tool == tool)]
         total_cpu = dfs["user_time"].values + dfs["sys_time"].values
         ax.loglog(
             dfs["num_samples"].values,
             total_cpu,
-            label=f"{tool}",
+            label=f"{toolname.get(tool, tool)}",
             # linestyle=ls,
             marker=".",
             color=colours[tool],
@@ -154,7 +159,7 @@ def data_scaling(size_data, output):
     df1 = pd.read_csv(size_data, index_col=None).sort_values("num_samples")
 
     fig, ax1 = one_panel_fig()
[…]

### 2024-04-09 paper.tex
Update whole matrix compute fig
Some text

@@
 \begin{figure}
-\includegraphics[width=6cm]{figures/whole-matrix-compute}
+\includegraphics{figures/whole-matrix-compute}
 \caption{Whole-matrix compute performance with increasing sample size.
@@
 running \texttt{bcftools +af-dist} and equivalent operations 
-on the data of Figure~\ref{fig-data-storage}.
+on the data of Figure~\ref{fig-data-storage}. We can see 
+that for $10^6$ samples, running \texttt{bcftools +af-dist}
+requires 20 hours (note that this is run on a BCF file, for 
+efficiency), which can be regarded as a reasonable baseline.
+Because genozip does not provide an API, we used the classical
+approach of piping the VCF text that it outputs into 
+\texttt{bcftools +af-dist}. The volumes of data involved here
+make this an expensive approach.
+[TODO discuss Zarr-python and Savvy C++ implementations]
 

### 2024-04-10 diagrams/data-model.tex

\begin{tikzpicture}

 %% sample_id
  \node at (2.0,5.9,-.5) {\ttfamily \scriptsize sample\textunderscore id};
  % bounds
  \draw[densely dashed] (0,5.5,-.5) -- (3.9,5.5,-.5);
  \draw[densely dashed] (0,5.7,-.5) -- (3.9,5.7,-.5);
 \tikzcuboidset{front face/.style={fill=yellow!20},right face/.style={fill=green!10},top face/.style={fill=yellow!20}}
  \pic at (0,5.5,-.5) {cuboid=.9--.2--0};
  \pic at (1,5.5,-.5) {cuboid=.9--.2--0};
  \node at (2.5,5.6,-.5) {\ttfamily \dots};
  \pic at (3,5.5,-.5) {cuboid=.9--.2--0};

 %% variant_contig
  \node[rotate=90] at (-1.5,2.5,-.3) {\ttfamily \scriptsize variant\textunderscore contig};
  % bounds
  \draw[densely dashed] (-1.4,0,-.5) -- (-1.4,4.9,-.5);
  \draw[densely dashed] (-1.2,0,-.5) -- (-1.2,4.9,-.5);
 \tikzcuboidset{front face/.style={fill=green!10},right face/.style={fill=green!10},top face/.style={fill=yellow!20}}
  \pic at (-1.4,0,-.5) {cuboid=.2--.9--0};
  \node at (-1.3,1.5,-.5) {\ttfamily \vdots};
  \pic at (-1.4,2,-.5) {cuboid=.2--.9--0};
  \pic at (-1.4,3,-.5) {cuboid=.2--.9--0};
  \pic at (-1.4,4,-.5) {cuboid=.2--.9--0};

 %% variant_position
  \node[rotate=90] at (-.8,2.5,-.3) {\ttfamily \scriptsize variant\textunderscore position};
  % bounds
  \draw[densely dashed] (-.7,0,-.5) -- (-.7,4.9,-.5);
  \draw[densely dashed] (-.5,0,-.5) -- (-.5,4.9,-.5);
 \tikzcuboidset{front face/.style={fill=green!10},right face/.style={fill=green!10},top face/.style={fill=yellow!20}}
  \pic at (-.7,0,-.5) {cuboid=.2--.9--0};
  \node at (-.6,1.5,-.5) {\ttfamily \vdots};
  \pic at (-.7,2,-.5) {cuboid=.2--.9--0};
  \pic at (-.7,3,-.5) {cuboid=.2--.9--0};
  \pic at (-.7,4,-.5) {cuboid=.2--.9--0};

 %% call_genotype
  \node at (2.0,5.3) {\ttfamily \scriptsize call\textunderscore genotype};
  % back box
  \draw[densely dashed] (0,0,-0.5) -- (3.9,0,-0.5);
  \draw[densely dashed] (0,0,-0.5) -- (0,4.9,-0.5);
  \draw[densely dashed] (0,4.9,-0.5) -- (3.9,4.9,-0.5);
  \draw[densely dashed] (3.9,0,-0.5) -- (3.9,4.9,-0.5);
 \tikzcuboidset{front face/.style={fill=blue!10},right face/.style={fill=green!10},top face/.style={fill=yellow!20}}
  \pic at (0,0) {cuboid=.9--.9--.5};
  \node at (0.45,0.5) {\ttfamily \tiny 100/0/0};
  \node at (0.5,1.6) {\ttfamily \vdots};
  \pic at (0,2) {cuboid=.9--.9--.5};
  \node at (0.45,2.5) {\ttfamily \tiny 2/0/0};
  \pic at (0,3) {cuboid=.9--.9--.5};
  \node at (0.45,3.5) {\ttfamily \tiny 1/0/0};
  \pic at (0,4) {cuboid=.9--.9--.5};
  \node at (0.45,4.5) {\ttfamily \tiny 0/0/0};
  \pic at (1,0) {cuboid=.9--.9--.5};
  \node at (1.45,0.5) {\ttfamily \tiny 100/1/0};
  \node at (1.5,1.6) {\ttfamily \vdots};
  \pic at (1,2) {cuboid=.9--.9--.5};
  \node at (1.45,2.5) {\ttfamily \tiny 2/1/0};
  \pic at (1,3) {cuboid=.9--.9--.5};
  \node at (1.45,3.5) {\ttfamily \tiny 1/1/0};
  \pic at (1,4) {cuboid=.9--.9--.5};
  \node at (1.45,4.5) {\ttfamily \tiny 0/1/0};
  \node at (2.5,0.5) {\ttfamily \dots};
  \node at (2.5,1.55) {\ttfamily $\ddots$};
  \node at (2.5,2.5) {\ttfamily \dots};
  \node at (2.5,3.5) {\ttfamily \dots};
  \node at (2.5,4.5) {\ttfamily \dots};
  \pic at (3,0) {cuboid=.9--.9--.5};
  \node at (3.45,0.5) {\ttfamily \tiny 100/10/0};
  \node at (3.5,1.6) {\ttfamily \vdots};
  \pic at (3,2) {cuboid=.9--.9--.5};
  \node at (3.45,2.5) {\ttfamily \tiny 2/10/0};
  \pic[ultra thick, red] at (3,3) {cuboid=.9--.9--.5};
  \node at (3.45,3.5) {\ttfamily \tiny 1/10/0};
  \pic at (3,4) {cuboid=.9--.9--.5};
  \node at (3.45,4.5) {\ttfamily \tiny 0/10/0};
  % front box
  \draw[densely dashed] (0,0) -- (3.9,0);
  \draw[densely dashed] (0,0) -- (0,4.9);
  \draw[densely dashed] (0,4.9) -- (3.9,4.9);
  \draw[densely dashed] (3.9,0) -- (3.9,4.9);

 %% dimensions
  \draw[stealth-stealth] (0.1,-0.2) -- (3.8,-0.2) node[midway,below] {\scriptsize samples};
  \draw[stealth-stealth] (4,-0.2, 0) -- (4,-0.2,-.5) node[midway,below right] {\scriptsize ploidy};
  \draw[stealth-stealth] (4.1,.1,-.5) -- (4.1,4.8,-.5) node[midway,below,rotate=90] {\scriptsize variants};

%% compresion
 \node (numcodecs) [rectangle, draw, densely dotted, thick, rounded corners, minimum width=2.4cm, minimum height=2.6cm, anchor= south west]
    at (4.8,2.2) {};
 \node[anchor=north,below] at (numcodecs.north) {\scriptsize Filter \& compress};
 %\node[anchor=south,below] at (numcodecs.south) {numcodecs};
 \node (blosk) [rectangle, ultra thick, red, draw, rounded corners, minimum width=1.8cm, minimum height=1.6cm, anchor=center,align=center,text=black]
    at (numcodecs) {\scriptsize Blosc\\\scriptsize ZSTD\\\scriptsize Autoshuffle};
 \draw[ultra thick, red,-stealth] (3.9,3.5,-.25) -- (5,3.5,-.25);

%% storage
 \node (location) [rectangle, draw, densely dotted, thick, rounded corners, minimum width=2.4cm, minimum height=4.5cm, anchor= south west]
    at (7.4,1.35) {};
 \node[anchor=south,below] at (location.south) {Storage};
 \node[anchor=center] at (location) {
        \begin{forest}
          for tree={
            font=\tiny ,
            grow'=0,
            child anchor=west,
            parent anchor=south,
            anchor=west,
            calign=first,
            edge path={
              \noexpand\path [draw, \forestoption{edge}]
              (!u.south west) +(7.5pt,0) |- node[fill,inner sep=1.25pt] {} (.child anchor)\forestoption{edge label};
            },
            before typesetting nodes={
              if n=1
                {insert before={[,phantom]}}
                {}
            },
            fit=band,
            before computing xy={l=15pt},
          }
        [{\scriptsize Cloud, file or blob}
          [\ttfamily call\textunderscore genotype
            [\ttfamily 0/0/0]
            [\ttfamily ...]
            [\ttfamily 1/10/0, draw={red,ultra thick, rounded corners}]
            [\ttfamily ...]
            [\ttfamily 100/10/0]
          ]
          [\ttfamily sample\textunderscore id]
          [\ttfamily ...]
        ]
        \end{forest}
 };
 \draw[ultra thick, red, -stealth,rounded corners=5pt] (6.8,3.5,-.25) -- (8,3.5,-.25);

\end{tikzpicture}

### 2024-04-10 paper.tex
Some tweaks

@@
 deorowicz2021vcfshark}, % VCFShark
-most notably the highly-influential Positional
+most notably the Positional
 Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
@@
 or when we typically want to analyse each record in its entirity.
-When we want to analyse only a subset of a record 
-row-wise storage can inefficient because we will usually need to
+When we want to analyse only a subset of a record,
+row-wise storage can be inefficient because we will usually need to
 retrieve more information than required from storage. In the case 
@@
 The alternative to row-wise storage is \emph{columnar} storage:
-instead of grouping toether all the fields for a record,
+instead of grouping together all the fields for a record,
 we group together all the records for a given field.
+Columnar storage has become standard [very quick summary of columnar
+storage with some refs].
 To our knowledge, the only existing method using columnar 
 storage for genetic variation data is 
-SeqArray~\citep{zheng2017seqarray} built on 
-the GDS container format~\citep{zheng2012high}.
+SeqArray~\citep{zheng2017seqarray,zheng2012high}.
 
@@
 variants and samples axis.
-Note that this approach to storing higher dimensional 
+Note that this approach to storing higher dimensional data
 sets Zarr apart from more general columnar formats 

### 2024-04-10 paper.tex
Pass through whole-matrix compute section

@@
 (APIs), providing efficient access to genotype and other VCF data
-[TODO check these citations]~\cite[e.g.][]{lefaive2021sparse,
-wertenbroek2022xsi,zhang2023gbc}.
+[TODO check these citations]~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
+wertenbroek2022xsi,zhang2023gbc}. By providing programmatic access,
+the data can be retrieved from storage, decoded and then analysed
+in the same memory space, avoiding the need for copying
+the decoded data and inter-process communication. [TODO some refs
+on zero-copy semantics and things like Arrow?]
 
@@
 we use the \texttt{bcftools +af-dist} plugin as an example of such a whole-matrix
-operation. 
+computation. 
 % The details of the \texttt{af-dist} operation are not important:
@@
 \texttt{htslib} C API~\citep{bonfield2021htslib}, distributed as 
-\texttt{bcftools} plugin.
+a \texttt{bcftools} plugin.
 
@@
 running \texttt{bcftools +af-dist} and equivalent operations 
-on the data of Figure~\ref{fig-data-storage}. We can see 
-that for $10^6$ samples, running \texttt{bcftools +af-dist}
-requires 20 hours (note that this is run on a BCF file, for 
-efficiency), which can be regarded as a reasonable baseline.
-Because genozip does not provide an API, we used the classical
-approach of piping the VCF text that it outputs into 
-\texttt{bcftools +af-dist}. The volumes of data involved here
-make this an expensive approach.
-[TODO discuss Zarr-python and Savvy C++ implementations]
-
-% We illustrate the discussion by the afdist operation from
-% the bcftools plugins package. This is perhaps a contrived example, but
-% we chose it because it requires examining the entire genotype matrix
-% and there is an efficient implementation available in bcftools.
-% It is also a good example because the standard approach of
-% divvying up the genome into chunks via HPC schedulers or workflow
-% managers is quite tricky.
-% We discuss the software development aspect, and how we developed
-% the afdist-savvy program. We discuss the inherent limitations
+on the data of Figure~\ref{fig-data-storage}. There is a large
+difference in the time required (note the log-log scale). 
+The slowest approach uses Genozip. Because genozip does not
+provide an API and only outputs VCF text, the best approach available 
+is to pipe its output into \texttt{bcftools +af-dist}. Thus,
+not only must we decode the data from the genozip format, we must 
+then parse the large volumes of VCF text and then finally do the 
+actual calculation. Running \texttt{bcftools +af-dist} directly
+on the gzipped VCF text incurs essentially the same cost, but
+the decompression cost is lower and without the interprocess 
+communication overhead. Using a BCF file is somewhat faster,
+because the packed binary format can be read into the internal 
+data structures used by the \texttt{htslib} API more efficiently.
+
+The data shown in Figure~\ref{fig-whole-matrix-compute} for Zarr and Savvy
+is based on custom programs written using their respective APIs
+to implement the \texttt{af-dist} operation. The Zarr program uses
+the Zarr-Python package to iterate over the decoded chunks of the 
+genotype matrix and classifies genotypes within a chunk using a 14 line Python
+function, accelerated using the Numba JIT compliler~\cite{lam2015numba}.
+The allele frequencies and genotype counts are then analysed to produce 
+the final counts within the allele frequency bins with 9 lines of 
+Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
+short and simple Python program is substantially faster than the 
+equivalent compiled C using \texttt{htslib} APIs on BCF (6 hours
+vs 20 hours for 1 million samples). The fastest method is the 
+C++ program written using the Savvy API. This would largely seem
+to be due to Savvy's excellent genotype decoding performance
+[TODO add fig for decode perf].
 

### 2024-04-10 src/zarr_afdist.py

@@ -256,8 +256,6 @@ def zarr_afdist(path, num_bins=10, variant_slice=None, sample_slice=None):
 
     alt_count = 2 * n - counts.ref_count
     af = alt_count / (n * 2)
-    # print("af", af)
-
     bins = np.linspace(0, 1.0, num_bins + 1)
     bins[-1] += 0.0125
     pRA = 2 * af * (1 - af)
@@ -266,8 +264,8 @@ def zarr_afdist(path, num_bins=10, variant_slice=None, sample_slice=None):
     b = np.bincount(
         np.digitize(pAA, bins), weights=counts.hom_alt, minlength=num_bins + 1
     )
-
     count = (a + b).astype(int)
+
     return pd.DataFrame({"start": bins[:-1], "stop": bins[1:], "prob_dist": count[1:]})
 
 

### 2024-04-10 paper.tex
Update subset-matrix compute figures

@@
 at $10^6$ samples was extrapolated by fitting an exponential.
+See Methods for full details.
 \label{fig-whole-matrix-compute}}
@@
 \begin{figure}
-\includegraphics[width=6cm]{figures/subset-matrix-compute}
+\includegraphics{figures/subset-matrix-compute}
 \caption{Compute performance on subsets of the matrix.
-(A) Total CPU time required to run the afdist calculation for
-a subset of 10 samples and 10000 variants from the middle of the matrix.
-(B) Same for n/2 samples. Genozip did not run in this case for
-$n > 10^4$ samples because it does not support a file to specify
-sample IDs, and the command line was therefore too long for the shell
-to execute. [TODO Note that in this case the bcftools pipeline requires
+Total CPU time required to run the af-dist calculation for
+a subset of 10 samples and 10000 variants from the middle of the matrix
+for the data in Figure~\ref{fig-data-storage}.
+[TODO Note that in this case the bcftools pipeline requires
 an additional step to compute AF etc.]
@@
 
-% \clearpage
-% \renewcommand\thefigure{S\arabic{figure}}
-% \setcounter{figure}{0}
-% \renewcommand\thetable{S\arabic{table}}
-% \setcounter{table}{0}
-
-% \section*{Supplementary Material}
+\clearpage
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
+
+\section*{Supplementary Material}
+
+\begin{figure}
+\includegraphics{figures/subset-matrix-compute-supplemental}
+\caption{Compute performance on a large subset of the genotype matrix.
+Total CPU time required to run the af-dist calculation for
+a subset of half of the samples and 10000 variants from the middle of the matrix
+for the data in Figure~\ref{fig-data-storage}.
+Genozip did not run for
+$n > 10^4$ samples because it does not support a file to specify
+sample IDs, and the command line was therefore too long for the shell
+to execute. 
+\label{fig-subset-matrix-compute-supplemental}}
+\end{figure}
 

### 2024-04-10 src/plot.py

@@ -23,11 +23,18 @@ def one_panel_fig(**kwargs):
     # The columnwidth of the format is ~250pt, which is
     # 3 15/32 inch, = 3.46
     width = 3.46
-    (
-        fig,
-        ax,
-    ) = plt.subplots(1, 1, figsize=(width, 2 * width / 3), **kwargs)
+    fig, ax = plt.subplots(1, 1, figsize=(width, 2 * width / 3), **kwargs)
     return fig, ax
+
+
+# def two_panel_fig(**kwargs):
+#     # The columnwidth of the genetics format is ~250pt, which is
+#     # 3 15/32 inch, = 3.46
+#     width = 3.46
+#     fig, (ax1, ax2) = plt.subplots(2, 1, figsize=(width, 4 * width / 3), **kwargs)
+#     ax1.set_title("(A)")
+#     ax2.set_title("(B)")
+#     return fig, (ax1, ax2)
 
 
 def plot_size(ax, df, label_y_offset=None):
@@ -197,7 +204,7 @@ def whole_matrix_compute(time_data, output):
     ax1.set_xlabel("Sample size (diploid)")
     ax1.set_ylabel("Time (seconds)")
 
-    ax1.set_title(f"af-dist CPU time")
+    # ax1.set_title(f"af-dist CPU time")
     ax1.legend()
 
     plt.tight_layout()
@@ -298,6 +305,21 @@ def plot_subset_time(ax, df, extrapolate_genozip=False):
         )
 
 
+def run_subset_matrix_plot(data, output, subset, extrapolate_genozip=False):
+    df = pd.read_csv(data, index_col=False).sort_values("num_samples")
+    fig, ax1 = one_panel_fig()
+    plot_subset_time(
+        ax1, df[df.slice == subset], extrapolate_genozip=extrapolate_genozip
+    )
+
+    ax1.set_xlabel("Sample size (diploid)")
+    ax1.set_ylabel("Time (seconds)")
+    ax1.legend()
+
+    plt.tight_layout()
+    plt.savefig(output)
+
+
 @click.command()
 @click.argument("data", type=click.File("r"))
 @click.argument("output", type=click.Path())
@@ -305,23 +327,17 @@ def subset_matrix_compute(data, output):
     """
     Plot the figure showing compute performance on subsets of matrix afdist.
     """
-    df = pd.read_csv(data, index_col=False).sort_values("num_samples")
+    run_subset_matrix_plot(data, output, "n10")
 
[…]

### 2024-04-10 src/plot.py
Update figures with makefile rule

@@ -239,9 +239,15 @@ def whole_matrix_decode(time_data, output):
 def plot_subset_time(ax, df, extrapolate_genozip=False):
     colours = {
         "bcftools": bcf_colour,
-        "zarr": zarr_colour,
-        # "savvy": sav_colour,
         "genozip": genozip_colour,
+        # "savvy": sav_colour,
+        "zarr": zarr_colour,
+    }
+
+    label_map = {
+        "bcftools": "bcftools pipeline",
+        "genozip": "genozip + bcftools pipeline",
+        "zarr": "zarr-python API",
     }
 
     for tool in colours.keys():
@@ -251,7 +257,7 @@ def plot_subset_time(ax, df, extrapolate_genozip=False):
         ax.loglog(
             n,
             total_cpu,
-            label=f"{tool}",
+            label=label_map[tool],
             # linestyle=ls,
             marker=".",
             color=colours[tool],

### 2024-04-10 src/collect_data.py
Update decode command

@@ -428,7 +428,7 @@ def file_size(src, output, debug):
 @click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools[:2]])
 @click.option("-s", "--storage", default="hdd")
 @click.option("--debug", is_flag=True)
-def decoding_time(src, output, tool, storage, debug):
+def whole_matrix_decode(src, output, tool, storage, debug):
     if len(src) == 0:
         raise ValueError("Need at least one input file!")
     tool_map = {t.name: t for t in all_tools[:2]}
@@ -602,7 +602,7 @@ def cli():
 
 cli.add_command(file_size)
 cli.add_command(whole_matrix_compute)
-cli.add_command(decoding_time)
+cli.add_command(whole_matrix_decode)
 cli.add_command(subset_processing_time)
 cli.add_command(genotype_filtering_processing_time)
 cli.add_command(site_filtering_processing_time)

### 2024-04-10 src/plot.py
Add initial whole-matrix decode supplemental figure

@@ -95,6 +95,8 @@ def plot_total_cpu(ax, df, toolname=None):
     # for tool in df.tool.unique():
     for tool in colours.keys():
         dfs = df[(df.tool == tool)]
+        if dfs.empty:
+            continue
         total_cpu = dfs["user_time"].values + dfs["sys_time"].values
         ax.loglog(
             dfs["num_samples"].values,
@@ -221,15 +223,16 @@ def whole_matrix_decode(time_data, output):
     df = pd.read_csv(time_data, index_col=False).sort_values("num_samples")
     df = df[df.storage == "hdd"]
 
-    # TODO set the width properly based on document
-    fig, ax1 = plt.subplots(1, 1, figsize=(4, 3))
+    fig, ax1 = one_panel_fig()
 
     plot_total_cpu(ax1, df)
+    df["genotypes_per_second"] = df["total_genotypes"] / df["wall_time"]
+    for tool in ["zarr", "savvy"]:
+        max_rate = df[df.tool == tool]["genotypes_per_second"].max()
+        print(tool, humanize.naturalsize(max_rate, binary=True))
 
     ax1.set_xlabel("Sample size (diploid)")
     ax1.set_ylabel("Time (seconds)")
-
-    ax1.set_title(f"Genotype decode time")
     ax1.legend()
 
     plt.tight_layout()

### 2024-04-10 paper.tex

@@
 to be due to Savvy's excellent genotype decoding performance
-[TODO add fig for decode perf].
+(up to 5.7GiB/s vs 1.2GiB/s for Zarr on this dataset;
+Figure~\ref{fig-whole-matrix-decode}).
 
@@
 \section*{Supplementary Material}
+
+\begin{figure}
+\includegraphics{figures/whole-matrix-decode}
+\caption{Genotype decoding performance.
+Total CPU time required to decode genotypes into memory using the zarr-python
+and Savvy C++ APIs for the data in Figure~\ref{fig-data-storage}.
+This corresponds to maximum rate of 1.2GiB/s for Zarr and 5.7GiB/s
+for Savvy. [TODO add data for 1 million samples]
+\label{fig-whole-matrix-decode}}
+\end{figure}
 

### 2024-04-10 paper.tex
Draft of text for subset matrix section

@@
 where variation occurs in the samples)~\citep{danecek2011variant}. 
-Blocks of VCF text are compressed using \texttt{bgzip},
-which allows subsets of the file to be retrieved using secondary
-indexes stored alongside the original file~\citep{li2011tabix}.
 VCF is therefore a \emph{row-wise} format, i.e., where all of the 
@@
 % not worth bothering with.
+
+[Short note here on why we choose these tools to
+benchmark against, and why there's no point in 
+a major benchmarkathon. Some recent papers 
+have good benchmarks which the interested 
+reader can cross-reference]
 
@@
 (up to 5.7GiB/s vs 1.2GiB/s for Zarr on this dataset;
-Figure~\ref{fig-whole-matrix-decode}).
+Figure~\ref{fig-whole-matrix-decode}). Different choices 
+of compressor and filters (zstandard with bit shuffle, in this case) 
+would likely lead to improved decoding performance in Zarr,
+at the expense of somewhat larger files.
 
 \subsection{Subsetting the genotype matrix}
-Existing work overwhelmingly emphasises accessibility by variant.
-We demonstrate that sgkit/zarr is good for getting subsets
-in terms of both variants and samples via afdist on various subsets.
-Refer to fig~\ref{fig-subset-matrix-compute}.
+As datasets grow ever larger, the ability to efficiently access subsets 
+of the data becomes more and more important. VCF/BCF achieve efficient 
+access to the data for genomic ranges 
+by compressing blocks of adjacent records using \texttt{bgzip},
+and storing secondary indexes alongside the original 
+files with a conventional suffix~\citep{li2011tabix}. 
+Thus, for a given range query we 
+decompress only the necessary blocks and can quickly access
+the required records. 
+Methods like Savvy and XSI largely follow the same principles [TODO check and 
+be more concrete]. The row-wise nature of these methods, however, mean 
+that we cannot efficiently subset \emph{by sample}. In the extreme
+case, if we want to access only the genotypes for a single sample
+we must still read in and decompress the entire dataset.
 
@@
 Total CPU time required to run the af-dist calculation for
-a subset of 10 samples and 10000 variants from the middle of the matrix
+a subset of 10000 variants $\times$ 10 samples from the middle of the matrix
 for the data in Figure~\ref{fig-data-storage}.
-[TODO Note that in this case the bcftools pipeline requires
-an additional step to compute AF etc.]
+The \texttt{bcftools} and \texttt{bcftools} pipelines involve
+multiple commands required to correctly calculate the AF INFO tag
+required by \texttt{bcftools +af-dist}. See the Methods for full details
+on the steps performed.
 \label{fig-subset-matrix-compute}}
 \end{figure}
+
+We illustrate the cost of row-wise encoding in
+Figure~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
+on a small fixed-size subset of the genotype matrices of
+Figure~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
+means that this small subset of the overall matrix can be efficiently
+extracted, and therefore the execution time depends very weakly on 
+the overall dataset size, with the overall computation requiring around
+1 second for 1 million samples. CPU time, for this simple calculation
+on a fixed size sub-matrix scales, with the dataset size.
+Figure~\ref{fig-subset-matrix-compute-supplemental} shows performance
+for the same operation when selecting half of the samples in the 
+dataset.
 
@@
 \end{itemize}
+[Random fragment to include somewhere]
+Row-wise storage of the genotype matrix makes accessing 
+sample haplotypes (i.e. the columns),
+required for many
+applications~\cite[e.g.][]{durbin2014efficient,kelleher2019inferring} 
+[TODO add phasing, imputation]
+inefficient. The simple approach of storing the genotype matrix
+in rectangular chunks allows efficient access both by row \emph{and}
+column. Similarly, these rectangular chunks capture patterns
+both of similarity between samples and linkage disequilibrium~\cite{mcvean2019linkage} 
+between sites, such that conventional compression schemes 
+perform very well~\ref{fig-data-storage}.
 

### 2024-04-11 paper.tex
Update figs for new data

@@
 to be due to Savvy's excellent genotype decoding performance
-(up to 5.7GiB/s vs 1.2GiB/s for Zarr on this dataset;
+(up to 6.5GiB/s vs 1.2GiB/s for Zarr on this dataset;
 Figure~\ref{fig-whole-matrix-decode}). Different choices 

### 2024-04-11 src/plot.py

@@ -65,7 +65,7 @@ def plot_size(ax, df, label_y_offset=None):
     add_number_of_variants(df, ax)
 
 
-def plot_total_cpu(ax, df, toolname=None):
+def plot_total_cpu(ax, df, toolname=None, time_units="h"):
     colours = {
         "bcftools": bcf_colour,
         "genozip": genozip_colour,
@@ -74,6 +74,7 @@ def plot_total_cpu(ax, df, toolname=None):
     }
     have_genozip = False
     toolname = {} if toolname is None else toolname
+    divisors = {"h": 3600, "m": 60}
 
     # for tool in df.tool.unique():
     for tool in colours.keys():
@@ -102,11 +103,11 @@ def plot_total_cpu(ax, df, toolname=None):
         )
         row = dfs.iloc[-1]
 
-        hours = total_cpu[-1] // 3600
+        hours = total_cpu[-1] // divisors[time_units]
 
         if tool != "genozip":
             ax.annotate(
-                f"{hours:.0f}h",
+                f"{hours:.0f}{time_units}",
                 textcoords="offset points",
                 xytext=(15, 0),
                 xy=(row.num_samples, total_cpu[-1]),
@@ -223,7 +224,7 @@ def whole_matrix_decode(time_data, output):
 
     fig, ax1 = one_panel_fig()
 
-    plot_total_cpu(ax1, df)
+    plot_total_cpu(ax1, df, time_units="m")
     df["genotypes_per_second"] = df["total_genotypes"] / df["wall_time"]
     for tool in ["zarr", "savvy"]:
         max_rate = df[df.tool == tool]["genotypes_per_second"].max()

### 2024-04-11 src/collect_data.py
Rename command

@@ -495,7 +495,7 @@ def midslice(n, k):
 @click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools])
 @click.option("-s", "--slice-id", multiple=True, default=["n10", "n/2"])
 @click.option("--debug", is_flag=True)
-def subset_processing_time(src, output, tool, slice_id, debug):
+def subset_matrix_compute(src, output, tool, slice_id, debug):
     if len(src) == 0:
         raise ValueError("Need at least one input file!")
     tool_map = {t.name: t for t in all_tools}
@@ -603,7 +603,7 @@ def cli():
 cli.add_command(file_size)
 cli.add_command(whole_matrix_compute)
 cli.add_command(whole_matrix_decode)
-cli.add_command(subset_processing_time)
+cli.add_command(subset_matrix_compute)
 cli.add_command(genotype_filtering_processing_time)
 cli.add_command(site_filtering_processing_time)
 cli.add_command(report_versions)

### 2024-04-11 src/plot.py
Add Savvy figs
Closes #20

@@ -60,7 +60,6 @@ def plot_size(ax, df, label_y_offset=None):
             xycoords="data",
         )
 
-
     ax.legend()
     add_number_of_variants(df, ax)
 
@@ -144,7 +143,6 @@ def plot_total_cpu(ax, df, toolname=None, time_units="h"):
 
 
 def add_number_of_variants(df, ax):
-
     dfs = df[df["tool"] == "zarr"]
     num_samples = dfs["num_samples"].values
     num_sites = dfs["num_sites"].values
@@ -242,7 +240,7 @@ def plot_subset_time(ax, df, extrapolate_genozip=False):
     colours = {
         "bcftools": bcf_colour,
         "genozip": genozip_colour,
-        # "savvy": sav_colour,
+        "savvy": sav_colour,
         "zarr": zarr_colour,
     }
 
@@ -250,10 +248,11 @@ def plot_subset_time(ax, df, extrapolate_genozip=False):
         "bcftools": "bcftools pipeline",
         "genozip": "genozip + bcftools pipeline",
         "zarr": "zarr-python API",
+        "savvy": "savvy C++ API",
     }
 
     for tool in colours.keys():
-        dfs = df[(df.threads == 1) & (df.tool == tool)]
+        dfs = df[(df.tool == tool)]
         total_cpu = dfs["user_time"].values + dfs["sys_time"].values
         n = dfs["num_samples"].values
         ax.loglog(
@@ -294,11 +293,11 @@ def plot_subset_time(ax, df, extrapolate_genozip=False):
             # Fit a simple exponential function.
             return a * np.power(n, b)
 
-        dfs = df[(df.threads == 1) & (df.tool == "genozip")]
+        dfs = df[(df.tool == "genozip")]
         fit_params, _ = optimize.curve_fit(
             mulplicative_model, dfs.num_samples[2:], dfs.wall_time[2:]
         )
-        num_samples = df[(df.threads == 1) & (df.tool == "bcftools")].num_samples.values
+        num_samples = df[(df.tool == "bcftools")].num_samples.values
         fit = mulplicative_model(num_samples, *fit_params)
         # print(fit)
 

### 2024-04-11 paper.tex

@@
 for the data in Figure~\ref{fig-data-storage}.
-The \texttt{bcftools} and \texttt{bcftools} pipelines involve
-multiple commands required to correctly calculate the AF INFO tag
+The \texttt{genozip} and \texttt{bcftools} pipelines involve
+multiple commands required to correctly calculate the AF INFO field
 required by \texttt{bcftools +af-dist}. See the Methods for full details
@@
 the overall dataset size, with the overall computation requiring around
-1 second for 1 million samples. CPU time, for this simple calculation
-on a fixed size sub-matrix scales, with the dataset size.
+1 second for 1 million samples. Because of their 
+row-wise encoding for all the other methods CPU time
+scales with the number of samples.
 Figure~\ref{fig-subset-matrix-compute-supplemental} shows performance

### 2024-04-11 src/collect_data.py
Add support for collecting afdist time on VCF

@@ -324,17 +324,28 @@ all_tools = [
     ),
 ]
 
+all_tools_vcf = all_tools + [
+    Tool(
+        "bcftools+vcf",
+        ".tags.vcf.gz",
+        run_bcftools_afdist,
+        run_bcftools_afdist_subset,
+        bcftools_version,
+    ),
+]
+
 
 @click.command()
 @click.argument("src", type=click.Path(), nargs=-1)
 @click.option("-o", "--output", type=click.Path(), default=None)
-@click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools])
+@click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools_vcf])
 @click.option("-s", "--storage", default="hdd")
 @click.option("--debug", is_flag=True)
 def whole_matrix_compute(src, output, tool, storage, debug):
     if len(src) == 0:
         raise ValueError("Need at least one input file!")
-    tool_map = {t.name: t for t in all_tools}
+
+    tool_map = {t.name: t for t in all_tools_vcf}
     tools = [tool_map[tool_name] for tool_name in tool]
 
     data = []

### 2024-04-11 src/plot.py
Add VCF lines and refactor plotting

@@ -12,8 +12,8 @@ plt.rcParams.update({"legend.fontsize": 6})
 plt.rcParams.update({"lines.markersize": 4})
 
 
-sgkit_colour = "tab:blue"
 bcf_colour = "tab:orange"
+vcf_colour = "tab:pink"
 sav_colour = "tab:red"
 genozip_colour = "tab:purple"
 zarr_colour = "tab:blue"
@@ -29,7 +29,7 @@ def one_panel_fig(**kwargs):
 
 def plot_size(ax, df, label_y_offset=None):
     colour_map = {
-        "vcf": "tab:pink",
+        "vcf": vcf_colour,
         "bcf": bcf_colour,
         "zarr": zarr_colour,
         "sav": sav_colour,
@@ -64,8 +64,9 @@ def plot_size(ax, df, label_y_offset=None):
     add_number_of_variants(df, ax)
 
 
-def plot_total_cpu(ax, df, toolname=None, time_units="h"):
+def plot_total_cpu(ax, df, toolname=None, time_units="h", extrapolate=None):
     colours = {
+        "bcftools+vcf": vcf_colour,
         "bcftools": bcf_colour,
         "genozip": genozip_colour,
         "zarr": zarr_colour,
@@ -73,7 +74,8 @@ def plot_total_cpu(ax, df, toolname=None, time_units="h"):
     }
     have_genozip = False
     toolname = {} if toolname is None else toolname
-    divisors = {"h": 3600, "m": 60}
+    divisors = {"s": 1, "h": 3600, "m": 60}
+    extrapolate = [] if extrapolate is None else extrapolate
 
     # for tool in df.tool.unique():
     for tool in colours.keys():
@@ -102,37 +104,33 @@ def plot_total_cpu(ax, df, toolname=None, time_units="h"):
         )
         row = dfs.iloc[-1]
 
-        hours = total_cpu[-1] // divisors[time_units]
-
-        if tool != "genozip":
+        if tool not in extrapolate:
+            time = total_cpu[-1] / divisors[time_units]
             ax.annotate(
-                f"{hours:.0f}{time_units}",
+                f"{time:.0f}{time_units}",
                 textcoords="offset points",
                 xytext=(15, 0),
                 xy=(row.num_samples, total_cpu[-1]),
                 xycoords="data",
             )
-        else:
-            have_genozip = True
 
-    # Extrapolate for genozip
-    def mulplicative_model(n, a, b):
+    def multiplicative_model(n, a, b):
         # Fit a simple exponential function.
[…]

### 2024-04-11 software/savvy-afdist/main.cpp
Add pos-only argument to savvy-afdist

@@ -53,6 +53,9 @@ void decode(const std::string& filename) {
 
     while (f >> var) {
         var.get_format("GT", geno);
+        /* This is just to make sure the compiler doesn't somehow ignore the
+         * get_format bit·
+         */
         if (geno[0] >= 0) {
             count++;
         }
@@ -60,6 +63,16 @@ void decode(const std::string& filename) {
 
     std::cout << "Decoded variants: " << count << std::endl;
 }
+
+void output_pos(const std::string& filename) {
+    savvy::reader f(filename);
+    savvy::variant var;
+
+    while (f >> var) {
+        std::cout << var.position() << std::endl;
+    }
+}
+
 
 // Use a lambda expression for finding the bin index
 auto findBinIndex = [](const std::vector<double>& bins, double value) -> int {
@@ -82,6 +95,7 @@ int main(int argc, char* argv[]) {
     std::string filename;
     std::string samples_file;
     bool decode_only = false;
+    bool pos_only = false;
     int start = -1;
     int end = -1;
 
@@ -89,6 +103,8 @@ int main(int argc, char* argv[]) {
         std::string arg = argv[i];
         if (arg == "--decode-only") {
             decode_only = true;
+        } else if (arg == "--pos-only") {
+            pos_only = true;
         } else if (arg == "--start") {
             i++;
             start = std::stoi(argv[i]);
@@ -122,6 +138,8 @@ int main(int argc, char* argv[]) {
 
     if (decode_only) {
         decode(filename);
+    } else if (pos_only) {
+        output_pos(filename);
     } else {
         std::vector<double> af;
         std::vector<int> hets;

### 2024-04-11 src/collect_data.py
Add timings for extracting the POS column

@@ -148,6 +148,18 @@ def run_bcftools_afdist(path, *, num_threads=1, debug=False):
     return time_cli_command(cmd, debug)
 
 
+def run_bcftools_pos_extract(path, *, debug=False):
+    with tempfile.NamedTemporaryFile("w") as f:
+        cmd = (
+            "/usr/bin/time -f'%S %U' "
+            f'./software/bcftools query --format "%POS\\n" {path} > {f.name}'
+        )
+        result = time_cli_command(cmd, debug)
+        if debug:
+            summarise_pos_file(f.name)
+        return result
+
+
 def run_genozip_afdist(path, *, num_threads=1, debug=False):
     cmd = (
         "export BCFTOOLS_PLUGINS=software/bcftools-1.18/plugins; "
@@ -199,6 +211,28 @@ def run_savvy_decode(path, *, debug=False):
     return time_cli_command(cmd, debug)
 
 
+def summarise_pos_file(path):
+    with open(path) as f:
+        num_lines = 0
+        for line in f:
+            num_lines += 1
+            if num_lines < 5:
+                print(line.strip())
+        print("Total rows:", num_lines)
+
+
+def run_savvy_pos_extract(path, *, debug=False):
+    with tempfile.NamedTemporaryFile("w") as f:
+        cmd = (
+            "/usr/bin/time -f'%S %U' "
+            f"software/savvy-afdist/sav-afdist {path} --pos-only > {f.name}"
+        )
+        result = time_cli_command(cmd, debug)
+        if debug:
+            summarise_pos_file(f.name)
+        return result
+
+
 def run_savvy_afdist_subset(
     path, ds, variant_slice, sample_slice, *, num_threads=1, debug=False
 ):
@@ -239,8 +273,27 @@ def zarr_decode_worker(ds_path, debug, conn):
     conn.send(f"{wall_time} {cpu_times.user} {cpu_times.system}")
 
 
+def zarr_pos_extract_worker(ds_path, debug, conn):
+    with tempfile.NamedTemporaryFile("w") as f:
+        before = time.time()
+        root = zarr.open(ds_path)
+        pos = root["variant_position"][:]
+        # Write the array to a text file to make the comparison fair
+        np.savetxt(f, pos, fmt="%d")
+        f.flush()
+        if debug:
+            summarise_pos_file(f.name)
+
+    wall_time = time.time() - before
+    cpu_times = psutil.Process().cpu_times()
[…]

### 2024-04-11 src/collect_data.py
Add extract position benchmark

@@ -279,7 +279,10 @@ def zarr_pos_extract_worker(ds_path, debug, conn):
         root = zarr.open(ds_path)
         pos = root["variant_position"][:]
         # Write the array to a text file to make the comparison fair
-        np.savetxt(f, pos, fmt="%d")
+        # np.savetxt(f, pos, fmt="%d")
+        # make a pandas dataframe because np.savetxt is too slow
+        df = pd.DataFrame(pos)
+        df.to_csv(f, header=False, index=False)
         f.flush()
         if debug:
             summarise_pos_file(f.name)

### 2024-04-11 paper.tex
More plot infrastructure for column extract

@@
 
+\begin{figure}
+\includegraphics{figures/column-extract}
+\caption{Time to extract the genome position and write to a text file.
+Total CPU time required to extract the POS field for BCF,
+sav and Zar formats
+for the data in Figure~\ref{fig-data-storage}.
+For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
+For sav, we used the Savvy C++ API to extract just the position, 
+and output text using the \texttt{std::cout} stream. For Zarr, we read 
+\texttt{variant\_position} into a NumPy array, and then wrote to
+a text file using the Pandas
+\texttt{write\_csv} method. Writing
+text dominates in the case of Zarr: populating the NumPy array 
+in memory takes less than a second in all cases. Genozip does
+not offer an option to extract individual columns, and so would require
+the output to be piped into \texttt{bcftools query} or a
+Unix utility such as \texttt{cut}.
+\label{fig-column-extract}}
+\end{figure}
+
+\subsection{Extracting fields}
+We have focused on the genotype matrix up to this point, contrasting
+Zarr with existing row-wise methods. The genotype matrix, however,
+is only one aspect of VCF data.
+Real-world VCFs encapsulate much more than just the genotype
+matrix, and can contain large numbers of additional fields. Efficient
+extraction of the data from these fields is a key use-case
+and an area where columnar storage has clear benefits.
+Figure~\ref{fig-column-extract} shows the time required to extract 
+the genomic position of each variant in the simulated benchmark 
+dataset, which we can use as an indicative example of a per-variant 
+query. Although Savvy is many times faster than \texttt{bcftools query}
+here, the row-wise storage strategy that they share means that 
+the entire dataset must be read into memory and 
+decompressed to extract just one field from each record. Zarr's
+columnar storage excels at these tasks: we only read and decompress
+the information required.
+
 \subsection{Case study: Genomics England 100,000 genomes}
@@
 perform very well~\ref{fig-data-storage}.
+
+Reel off Python data science ecosystem: Pandas~\citep{mckinney2010data} etc
 

### 2024-04-11 paper.tex
Rough outline of per-field data

@@
 recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
-aggV2 comprises approximately 722 million annotated single-nucleotide 
+The dataset comprises approximately 722 million annotated single-nucleotide 
 variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
-totalling 165.3 Tib of data after \texttt{bgzip} compression. 
+totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
 The dataset is used for a variety of research purposes, ranging from 
@@
 simple queries involving single gene 
-regions~\cite{leggatt2023genotype}~\cite{lam2023repeat}.
+regions~\cite{leggatt2023genotype,lam2023repeat}.
+
 We converted the data to the intermediate columnar format 
@@
 \begin{table}
-\begin{tabular}{cc}
-Some & data on Genomics England\\
-\end{tabular}
 \caption{Summary of vcf-zarr performance on the Genomics England 100,000
@@
 \label{tab-genomics-england-data}}
+\begin{tabular}{llccc}
+\toprule
+Field & type & storage & compress & \%total \\
+\midrule
+call\_AD &  int16 & 179G & 26 & 24\\
+call\_GQ &  int16 & 171G & 13 & 22 \\
+call\_DP &  int16 & 141G & 20 & 18.8 \\
+call\_FT &  string & 58G & 160 & 7.3 \\
+call\_PL &  int16 & 51G & 140 & 6.8 \\
+call\_GQX &  int16 & 12G & 190 & 1.6 \\
+call\_genotype & int8 & 6.13G & 380 & 0.8 \\
+call\_genotype\_mask & bool & 3.66G  & 630 & 0.48\\
+call\_genotype\_phased & bool & 3.66G  & 630 & 0.48 \\
+variant\_quality & float32 & 22.6M & 2.7 & 0.0003 \\
+variant\_allele & string & 21.5M & 11 \\
+variant\_position & int32 & 12.7M & 4.7 \\
+variant\_ABratio & float32 & 9.6M & 6.3 \\
+variant\_AN & int32 & 9.6M & 6.3 \\
+variant\_filter & bool & 1.96M & 490 \\
+\bottomrule
+\end{tabular}
 \end{table}

### 2024-04-11 paper.tex
Rough draft of table with indicitive data

@@
 \label{tab-genomics-england-data}}
-\begin{tabular}{llccc}
+\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
 \toprule
-Field & type & storage & compress & \%total \\
+{Field} & {type} & {storage} & {compress} & {\%total} \\
 \midrule
-call\_AD &  int16 & 179G & 26 & 24\\
-call\_GQ &  int16 & 171G & 13 & 22 \\
-call\_DP &  int16 & 141G & 20 & 18.8 \\
-call\_FT &  string & 58G & 160 & 7.3 \\
-call\_PL &  int16 & 51G & 140 & 6.8 \\
-call\_GQX &  int16 & 12G & 190 & 1.6 \\
-call\_genotype & int8 & 6.13G & 380 & 0.8 \\
-call\_genotype\_mask & bool & 3.66G  & 630 & 0.48\\
-call\_genotype\_phased & bool & 3.66G  & 630 & 0.48 \\
-variant\_quality & float32 & 22.6M & 2.7 & 0.0003 \\
+call\_AD &  int16 & 179.0G & 26 & 24.1\\
+call\_GQ &  int16 & 171.0G & 13 & 22.2 \\
+call\_DP &  int16 & 141.0G & 20 & 18.8 \\
+call\_FT &  string & 58.0G & 160 & 7.3 \\
+call\_PL &  int16 & 51.0G & 140 & 6.8 \\
+call\_GQX &  int16 & 12.0G & 190 & 1.6 \\
+call\_genotype & int8 & 6.1G & 380 & 0.8 \\
+call\_genotype\_mask & bool & 3.6G  & 630 & 0.48\\
+call\_genotype\_phased & bool & 3.6G  & 1700 & 0.48 \\
+variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
 variant\_allele & string & 21.5M & 11 \\
@@
 variant\_AN & int32 & 9.6M & 6.3 \\
-variant\_filter & bool & 1.96M & 490 \\
+variant\_filter & bool & 1.9M & 490 \\
 \bottomrule

### 2024-04-11 paper.tex
outline of GeL narrative.
Closes #4

@@
 \subsection{Case study: Genomics England 100,000 genomes}
-In this section we demonstrate the utility of vcfzarr on a large human dataset.
+In this section we demonstrate the utility of \texttt{vcfzarr} on a large human dataset
+and the scalability of the \texttt{vcf2zarr} conversion utility.
 Genomics England’s multi-sample VCF dataset (aggV2) is an 
@@
 
-We converted the data to the intermediate columnar format 
-which took X hours with a cluster of Y workers using the 
-distributed explode commands. When then converted to Zarr using 
-SOMETHING which took X hours using [RESOURCES]. In total, this gave 
-X TiB of data, spread across Y thousand files, a compression of 
-X over the original VCFs.
-The detailed breakdown of per-column compression ratios is 
-shown in Table~\ref{tab-genomics-england-data}.
+As described in the Methods, conversion to Zarr using 
+\texttt{vcf2zarr} is a two-step 
+process to limit memory usage and ensure reliability. We 
+first convert the X VCF files (YTiB) for chromosome 2
+into the intermediate columnar format. This task was 
+distributed over a cluster of N workers, using a 
+total of X CPU hours and an elapsed wall-clock time 
+of X hours. When then converted the intermediate 
+column format to Zarr using 
+SOMETHING which took X hours using [RESOURCES]. 
+This produced a dataset with N arrays, consuming a 
+total of Y TiB of storage over X directories and 
+Y files. The top fields in terms 
+of storage are detailed in Table~\ref{tab-genomics-england-data}.
 
 \begin{table}
-\caption{Summary of vcf-zarr performance on the Genomics England 100,000
-genomes VCF dataset.
+\caption{Summary of the top N vcfzarr columns produced for 
+Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
+default settings. Each field is stored independently 
+as Zarr array with the given type (sufficient to represent all values in the
+data). We show the total storage consumed (reported via \texttt{du}) in 
+power-of-two units, and the compression ratio achieved on that array.
+We also show the percentage of the overall storage that each array consumes.
 \label{tab-genomics-england-data}}
@@
 
-Table~\ref{tab-genomics-england-data} shows that the dataset size is 
-dominated by a few columns. This is well known and has been remarked 
-on several times [e.g. spVCF paper, etc]. [In this paragraph we discuss
-how some of these fields can be considerably compressed with very
-little loss of information. For example, the maximum Allele Depth (AD)
-value seen in this dataset is >3000. However, when we compute the 
-histogram of AD values we can see that something like 0.1\% of 
-AD values are > 127, so we can reasonably store these as 8 bit 
-ints. This reduces storage to X. Finish up by saying that more 
-could be done, but it's beyond the scope of this paper. Maybe note 
-that this kind of per-field analysis is easy now, e.g. computing 
-the hist took X minutes using xarray and Dask on a cluster with
-Y workers..]
-
-To illustrate the difference in query performance we ran some 
-\texttt{bcftools query} commands based on the Genomics England 
-user guides. For example, finding all variants passing 
-quality filters with an allele frequency X takes just 
-x seconds on a single core (y seconds of which is writing the 
-CSV file). The same query took X hours on one chunk, giving a 
-an approximate total time of XX hours over all Y chunks 
-in chromosomes 2.
+Table~\ref{tab-genomics-england-data} shows that the dataset storage
+size dominated by a few columns, with call\_AD and call\_GQ consuming 
+almost half of total. These fields are much less compressible
+than genotype data (which uses $<1\%$ of the total space here)
+due to their inherent noisiness~\citep{lin2020sparse}. Substantial
+reductions in space can be achieved, however, if we examine the 
+distributions of these fields and truncate some outliers.
+Figure~SX shows the histogram of the XX call\_AD values, and we can
+see that < 0.0X\% of the values are $>127$. Truncating and 
+re-encoding the call\_AD array as an int8 results in a total storage of 
+XX. A substantial advantage of Zarr is that such per-field optimisations
+can be made after conversion, without affecting the rest of the dataset.
+As discussed in Methods, judicious use of additional filters
+and lossy quantizing can result substantial space savings.
+
+Calculating the histogram of call\_AD values at this 
+scale is not a trivial task.
+Taking advantage of the rich software ecosystem growing around Zarr
+(see Discussion) we used xarray~\citep{hoyer2017xarray}
+and Dask~\citep{rocklin2015dask} to transparently distribute the 
+computation over a cluster of X workers. The elapsed time 
+was X minutes, and requires X lines of Python in a 
+Jupyter notebook~\citep{kluyver2016jupyter}.
+
+One of the key uses of the aggV2 dataset is to enable simple queries 
+of specific regions. For example, in [A PAPER] they [DID SOMETHING].
+We replicated this by running the appropriate bcftools query
+on the VCF chunk (note that the user is responsible for finding 
+the correct chunk file here, limiting the utility of genomic 
+indexes). This took XXX hours.
+The same query implemented with the Zarr-python 
+API required XXX seconds to produce identical output.
 

### 2024-04-12 paper.tex
First version of GeL data analysis

@@
 process to limit memory usage and ensure reliability. We 
-first convert the X VCF files (YTiB) for chromosome 2
-into the intermediate columnar format. This task was 
-distributed over a cluster of N workers, using a 
-total of X CPU hours and an elapsed wall-clock time 
-of X hours. When then converted the intermediate 
-column format to Zarr using 
-SOMETHING which took X hours using [RESOURCES]. 
-This produced a dataset with N arrays, consuming a 
+first converted the 28 VCF files (3.51 TiB) for chromosome 20
+into the intermediate columnar format (ICF). This task was 
+split into 3899 partitions, and distributed using the Genomics England
+HPC cluster. It required about 50 days of CPU, with an elapsed
+wall-clock time of about 2 hours. The ICF representation uses a total
+of XTiB over XX files and YY directories. We then encoded this intermediate
+columnar format to Zarr using 8 cores and XX GB of RAM in about 12 hours.
+This produced a dataset with 43 arrays, consuming a 
 total of Y TiB of storage over X directories and 
@@
 \begin{table}
-\caption{Summary of the top N vcfzarr columns produced for 
+\caption{Summary for a selection of the largest vcfzarr columns produced for 
 Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
 default settings. Each field is stored independently 
-as Zarr array with the given type (sufficient to represent all values in the
+as a Zarr array with the given type (sufficient to represent all values in the
 data). We show the total storage consumed (reported via \texttt{du}) in 
@@
 \midrule
-call\_AD &  int16 & 179.0G & 26 & 24.1\\
-call\_GQ &  int16 & 171.0G & 13 & 22.2 \\
-call\_DP &  int16 & 141.0G & 20 & 18.8 \\
-call\_FT &  string & 58.0G & 160 & 7.3 \\
-call\_PL &  int16 & 51.0G & 140 & 6.8 \\
-call\_GQX &  int16 & 12.0G & 190 & 1.6 \\
+call\_AD &  int16 & 179.7G & 26 & 24.0\\
+call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
+call\_DP &  int16 & 141.8G & 16 & 18.9 \\
+call\_DPF& int16  & 115.1G & 20 & 15.3\\
+call\_FT &  string & 58.5G & 160 & 7.8 \\
+call\_PL &  int16 & 51.4G & 140 & 6.9 \\
+call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
 call\_genotype & int8 & 6.1G & 380 & 0.8 \\
-call\_genotype\_mask & bool & 3.6G  & 630 & 0.48\\
-call\_genotype\_phased & bool & 3.6G  & 1700 & 0.48 \\
+call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
+call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
+call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
 variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
@@
 
+% array([23.956     , 22.90533333, 18.90133333, 15.34533333,  7.8       ,
+%         6.85333333,  1.61066667,  0.81733333,  0.488     ])
+
+% /call_AD                     int16    179.67 GiB  4.53 TiB       26 125768  37.74 MiB     1.46 MiB            (15912293, 78195, 2)  (10000, 1000, 2)  
+% /call_GQ                     int16    171.79 GiB  2.26 TiB       13 125768  18.87 MiB     1.4 MiB             (15912293, 78195)     (10000, 1000) 
+% /call_DP                     int16    141.76 GiB  2.26 TiB       16 125768  18.87 MiB     1.15 MiB            (15912293, 78195)     (10000, 1000)
+% /call_DPF                    int16    115.09 GiB  2.26 TiB       20 125768  18.87 MiB     959.56 KiB          (15912293, 78195)     (10000, 1000)
+% /call_FT                     object   58.5 GiB    9.05 TiB      160 125768  75.48 MiB     487.71 KiB          (15912293, 78195)     (10000, 1000)
+% /call_PL                     int16    51.4 GiB    6.79 TiB      140 125768  56.61 MiB     428.51 KiB          (15912293, 78195, 3)  (10000, 1000, 3)
+% /call_GQX                    int16    12.08 GiB   2.26 TiB      190 125768  18.87 MiB     100.73 KiB          (15912293, 78195)     (10000, 1000) 
+% /call_genotype               int8     6.13 GiB    2.26 TiB      380 125768  18.87 MiB     51.11 KiB           (15912293, 78195, 2)  (10000, 1000, 2)  
+% /call_genotype_mask          bool     3.66 GiB    2.26 TiB      630 125768  18.87 MiB     30.53 KiB           (15912293, 78195, 2)  (10000, 1000, 2)  
+% /call_genotype_phased        bool     692.51 MiB  1.13 TiB     1700 125768  9.43 MiB      5.64 KiB            (15912293, 78195)     (10000, 1000) 
+% /call_PS                     int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000) 
+% /call_ADR                    int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000)
+% /call_ADF                    int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000)
+% /variant_OLD_MULTIALLELIC    object   31.63 MiB   121.4 MiB       3.8 1592  78.09 KiB     20.34 KiB           (15912293,)           (10000,) 
+% /variant_quality             float32  22.6 MiB    60.7 MiB        2.7 1592  39.04 KiB     14.53 KiB           (15912293,)           (10000,) 
+% /variant_allele              object   21.52 MiB   242.8 MiB      11 1592  156.17 KiB    13.84 KiB           (15912293, 2)         (10000, 2) 
+% /variant_position            int32    12.83 MiB   60.7 MiB        4.7 1592  39.04 KiB     8.25 KiB            (15912293,)           (10000,) 
+% /variant_ABratio             float32  9.62 MiB    60.7 MiB        6.3 1592  39.04 KiB     6.18 KiB            (15912293,)           (10000,) 
+% /variant_AN                  int32    9.6 MiB     60.7 MiB        6.3 1592  39.04 KiB     6.18 KiB            (15912293,)           (10000,) 
+% /variant_AC                  int32    9.22 MiB    60.7 MiB        6.6 1592  39.04 KiB     5.93 KiB            (15912293,)           (10000,) 
+% /variant_AC_Het              int32    8.98 MiB    60.7 MiB        6.8 1592  39.04 KiB     5.78 KiB            (15912293,)           (10000,)
+
+% Approx total = 179.67 + 171.79 + 141.76 + 115.09 + 58.5 + 51.4 + 12.08 + 6.13 + 3.66 
+% 740, round up to 750
+% [179.67 , 171.79 , 141.76 , 115.09 , 58.5 , 51.4 , 12.08 , 6.13 , 3.66 ]
+% >>> import numpy as np
+% >>> a = np.array([179.67 , 171.79 , 141.76 , 115.09 , 58.5 , 51.4 , 12.08 ,
+% 6.13 , 3.66 ]  )
+% >>> a
+% array([179.67, 171.79, 141.76, 115.09,  58.5 ,  51.4 ,  12.08,   6.13,
+%          3.66])
+% >>> a / 750 * 100
+% array([23.956     , 22.90533333, 18.90133333, 15.34533333,  7.8       ,
+%         6.85333333,  1.61066667,  0.81733333,  0.488     ])
+
 Table~\ref{tab-genomics-england-data} shows that the dataset storage
-size dominated by a few columns, with call\_AD and call\_GQ consuming 
-almost half of total. These fields are much less compressible
+size is dominated by a few columns with the top four
+(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
+80\% of the total. These fields are much less compressible
 than genotype data (which uses $<1\%$ of the total space here)
-due to their inherent noisiness~\citep{lin2020sparse}. Substantial
-reductions in space can be achieved, however, if we examine the 
-distributions of these fields and truncate some outliers.
-Figure~SX shows the histogram of the XX call\_AD values, and we can
-see that < 0.0X\% of the values are $>127$. Truncating and 
-re-encoding the call\_AD array as an int8 results in a total storage of 
-XX. A substantial advantage of Zarr is that such per-field optimisations
-can be made after conversion, without affecting the rest of the dataset.
-As discussed in Methods, judicious use of additional filters
-and lossy quantizing can result substantial space savings.
-
-Calculating the histogram of call\_AD values at this 
+because of their inherent noisiness~\citep{lin2020sparse}. By 
+default the \texttt{bio2zarr} conversion is lossless, and integers
+of the narrowest type required to represent all of the data are
+used. In this case, this has resulted in 16 bit integers being used 
+for these top four fields. Examining the distribution of the depth
+fields, (call\_AD, call\_DP, call\_DPF), we can 
+we can see that a tiny fraction of the values stored
+are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
+Fig SX for histograms).
+One might then reasonably truncate these fields to 8 bit integers.
+After converting to 8 bit, call\_AD is reduced by 
+x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
+and call\_DPF by x\% (XX GiB). 
+Similarly, the vast majority of call\_GQ values (xx\%) are less than 
+128 (corresponding to a conditional probability that the genotype call is 
+wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
+be truncated. Here, this results in a reduction of x\% (XX GiB).
+
+% and can reasonably be truncated (a 
+% much information would be lost if we used 8 bit
+% Substantial
+% reductions in space can be achieved, however, if we examine the 
+% distributions of these fields and truncate some outliers.
+% Figure~SX shows the histogram of the XX call\_AD values, and we can
+% see that < 0.0X\% of the values are $>127$. Truncating and 
+% re-encoding the call\_AD array as an int8 results in a total storage of 
+% XX. A substantial advantage of Zarr is that such per-field optimisations
+% can be made after conversion, without affecting the rest of the dataset.
+% As discussed in Methods, judicious use of additional filters
+% and lossy quantizing can result substantial space savings.
+
+Calculating the histogram of per call values at this 
 scale is not a trivial task.
@@
 and Dask~\citep{rocklin2015dask} to transparently distribute the 
-computation over a cluster of X workers. The elapsed time 
-was X minutes, and requires X lines of Python in a 
+computations over a cluster of X workers. 
+[ADD DETAILS]
+The elapsed time was X minutes, and requires X lines of Python in a 
 Jupyter notebook~\citep{kluyver2016jupyter}.

### 2024-04-12 paper.tex
Updates

@@
 This produced a dataset with 43 arrays, consuming a 
-total of Y TiB of storage over X directories and 
-Y files. The top fields in terms 
+total of Y GiB of storage over X directories and 
+Y files. This is about a 4.8X reduction over the original VCF. 
+The top fields in terms 
 of storage are detailed in Table~\ref{tab-genomics-england-data}.
@@
 be truncated. Here, this results in a reduction of x\% (XX GiB).
+Combined, truncating these top four fields to 8 bits reduces the 
+overall storage of the dataset to XXGiB, corresponding to a 
+X-fold compression of the VCF.
 

### 2024-04-12 paper.tex
Ditch the text graveyard

@@
 Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
+See~\citep{mcvean2019linkage} for a review of the techniques
+employed in genetic data compression.
 
@@
 storage with some refs].
-To our knowledge, the only existing method using columnar 
+Although columnar methods have been successfully used in alignment 
+% TODO check
+storage~\citep{bonfield2014scramble,bonfield2022cram},
+the only existing method using columnar 
 storage for genetic variation data is 
-SeqArray~\citep{zheng2017seqarray,zheng2012high}.
+SeqArray~\citep{zheng2017seqarray,zheng2012high} to our knowledge.
 
@@
 \section{Discussion}
-
-Outline of a possible narrative by paragraph:
-\begin{itemize}
-\item When VCF was introduced as part of the work of the 1000 Genomes
-project the software landscape was a real mess. It has entirely
-succeeded in standardising the output of variant callers, and methods
-have interoperated well for the past decade. As a method of interchange
-following classical file-oriented paradigms it works well enough.
-It is also a good archival format: when compressed with a well established
-method like gzip and combined with the well-documented description of
-the text, we can be sure that files will still be readable decades
-from now.
-
-\item VCF is holding back large-scale genomics, however, and in particular the
-analysis of population scale genetic variation datasets like UKB and GeL.
-Describe the current state of affairs. Variant datasets are chunked up
-into reasonable sized bits (10s of gigabytes) so that they can be processed
-in parallel. Doing anything with them requires interacting with workflow
-managers. In practise, it is not possible to directly write programs to
-work across multiple large-scale datasets: although the chunks of data
-are given in VCF, orchestrating how the chunks are processed and
-the results combined involves substantial work, involving the workflow
-manager, and the details of how chromosomes are split into chunks.
-Cite examples like the phasing pipelines for UKB from Brownings etc.
-
-\item Distributed computing is the reality, and increasingly datasets
-are being mandatorily stored in public clouds as part of
-Trusted Research Environments. Cloud object stores are the reality
-because it is much cheaper. POSIX file systems are emulated on top of
-this. Benefits of breaking up data into manageable pieces are that
-compute resources can be localised to the data. Compute resources are
-most effectively obtained as short-lived, small VMs that operate
-as part of a large cluster. Large vertically scaled ``pet'' servers
-are much more expensive, and just not feasible for the current
-scale of data. Lots of other architectural
-benefits from the new reality of cloud-based storage.
-
-\item Classical methods based around Unix pipelines and streaming
-files between processes on a single machine are therefore a
-significant mismatch to the reality of large-scale genetic variation
-datasets. The majority of methods suggested to make working
-with VCF data more efficient are based on these ideas: they all
-create a single large file and regard distribution of work
-in a cluster as extrinsic. Most output VCF as the result of queries,
-which also assumes that the VCF data can be efficiently processed.
-This is a reasonable assumption for thousands of samples, but not
-for larger datasets.
-
-\item The key issue here is that VCF is not a \emph{computational}
-format. Contrast VCF with plink, which stores data on disk in a
-form that can be computed on directly. Plink is limited in what
-can be descibed though, does not compress, and is not designed
-to distributed.
-
-\item The large-scale analysis of genetic variation data does not
-need a more efficient file format, which allows files to be downloaded
-somewhat more quickly, and subsets extracted somewhat more efficiently
-as VCF. It needs a standard storage platform that allows data to be
-processed directly without conversion, subsetted efficiently across
-both the variant and sample dimensions, with first-class support
-for all the necessary additional information (GT values etc),
-and that is suited to the realities of today's cloud-based
-storage and processing facilities.
-
-\item We suggest that the model used by Zarr is a good choice for this.
-We have shown that chunks of the genotype matrix compressed with standard
-methods are competitive with highly specialised methods for
-genotype data. In reality most of the storage is for non-genotype
-data anyway, where everyone is essentially doing the same thing.
-Columnar storage allows getting just (e.g.) the variant positions
-without reading everything. Another benefit is that by storing
-fields separately in different cloud objects, they can have
-different permissions. So, users can be given differential access
-to read parts of the dataset, and this is managed directly
-by the cloud storage. Give example.
-
-\item A major strong point of Zarr is its simplicity. All it's doing
-is compressing n-dimensional chunks of numerical data and storing
-them using fixed address keys. This key-value approach means that storage
-is very flexible and implementation is straightforward. There
-are several implementations of the Zarr protocol, and while some of these
-are rudimentary, they could be improved or replaced with relatively
-little effort. Zarr has been successfully used to store petabytes
-of data across different sciences --- genetic variation data is largely
-arrays of numerical data, and not really that different ultimately.
-Zarr is cloud native.
-Zarr is an open specification, built on standard components (contrast
-with Hail).
-
-\item Zarr is not perfect. Discuss limitations around ragged arrays,
-etc. Forthcoming improvements in zarr v3. The whole genotype matrix
-paradigm is challenged by graph genomes as well. While this is
-being worked out, there is still plenty of data to analyse as a
-genotype matrix.
-
-\item Vision. An ecosystem of tools that interoperate
-based on an open, cloud-based format where distribution is naturally
-managed by the stored chunks. Tools can be run on different datasets
-with minimal tuning, not by writing a full layer of orchestration.
-Tools read the data they need, not the entire dataset. Users read
-directly from a single canonical dataset, and do not maintain
-their own filtered copies. We suggest that the Zarr VCF approach
-we have provided would be a good starting point for designing
-such a compute platform.
-
-\end{itemize}
+The introduction of VCF as part of the work of the 1000 Genomes
+project~\cite{10002015global} has transformed a fragmented
+
+% A computational format, on the other hand, is optimised
+% for calculations, for \emph{using} the data, today. A computational format
+% is designed with particular data access patterns in mind, ultimately
+% to make a broad range of queries and downstream computations as efficient as
+% possible.
+
+% Outline of a possible narrative by paragraph:
+% \begin{itemize}
+% \item When VCF was introduced as part of the work of the 1000 Genomes
+% project the software landscape was a real mess. It has entirely
+% succeeded in standardising the output of variant callers, and methods
+% have interoperated well for the past decade. As a method of interchange
+% following classical file-oriented paradigms it works well enough.
+% It is also a good archival format: when compressed with a well established
+% method like gzip and combined with the well-documented description of
+% the text, we can be sure that files will still be readable decades
+% from now.
+
+% \item VCF is holding back large-scale genomics, however, and in particular the
+% analysis of population scale genetic variation datasets like UKB and GeL.
+% Describe the current state of affairs. Variant datasets are chunked up
+% into reasonable sized bits (10s of gigabytes) so that they can be processed
+% in parallel. Doing anything with them requires interacting with workflow
+% managers. In practise, it is not possible to directly write programs to
+% work across multiple large-scale datasets: although the chunks of data
+% are given in VCF, orchestrating how the chunks are processed and
+% the results combined involves substantial work, involving the workflow
+% manager, and the details of how chromosomes are split into chunks.
+% Cite examples like the phasing pipelines for UKB from Brownings etc.
+
+% \item Distributed computing is the reality, and increasingly datasets
+% are being mandatorily stored in public clouds as part of
+% Trusted Research Environments. Cloud object stores are the reality
+% because it is much cheaper. POSIX file systems are emulated on top of
+% this. Benefits of breaking up data into manageable pieces are that
+% compute resources can be localised to the data. Compute resources are
+% most effectively obtained as short-lived, small VMs that operate
+% as part of a large cluster. Large vertically scaled ``pet'' servers
+% are much more expensive, and just not feasible for the current
+% scale of data. Lots of other architectural
+% benefits from the new reality of cloud-based storage.
+
+% \item Classical methods based around Unix pipelines and streaming
+% files between processes on a single machine are therefore a
+% significant mismatch to the reality of large-scale genetic variation
+% datasets. The majority of methods suggested to make working
+% with VCF data more efficient are based on these ideas: they all
+% create a single large file and regard distribution of work
+% in a cluster as extrinsic. Most output VCF as the result of queries,
+% which also assumes that the VCF data can be efficiently processed.
+% This is a reasonable assumption for thousands of samples, but not
+% for larger datasets.
+
+% \item The key issue here is that VCF is not a \emph{computational}
+% format. Contrast VCF with plink, which stores data on disk in a
+% form that can be computed on directly. Plink is limited in what
+% can be descibed though, does not compress, and is not designed
+% to distributed.
+
+% \item The large-scale analysis of genetic variation data does not
+% need a more efficient file format, which allows files to be downloaded
+% somewhat more quickly, and subsets extracted somewhat more efficiently
+% as VCF. It needs a standard storage platform that allows data to be
+% processed directly without conversion, subsetted efficiently across
+% both the variant and sample dimensions, with first-class support
+% for all the necessary additional information (GT values etc),
+% and that is suited to the realities of today's cloud-based
+% storage and processing facilities.
+
+% \item We suggest that the model used by Zarr is a good choice for this.
+% We have shown that chunks of the genotype matrix compressed with standard
+% methods are competitive with highly specialised methods for
+% genotype data. In reality most of the storage is for non-genotype
+% data anyway, where everyone is essentially doing the same thing.
+% Columnar storage allows getting just (e.g.) the variant positions
+% without reading everything. Another benefit is that by storing
+% fields separately in different cloud objects, they can have
+% different permissions. So, users can be given differential access
+% to read parts of the dataset, and this is managed directly
+% by the cloud storage. Give example.
+
+% \item A major strong point of Zarr is its simplicity. All it's doing
+% is compressing n-dimensional chunks of numerical data and storing
+% them using fixed address keys. This key-value approach means that storage
+% is very flexible and implementation is straightforward. There
+% are several implementations of the Zarr protocol, and while some of these
+% are rudimentary, they could be improved or replaced with relatively
+% little effort. Zarr has been successfully used to store petabytes
+% of data across different sciences --- genetic variation data is largely
+% arrays of numerical data, and not really that different ultimately.
+% Zarr is cloud native.
+% Zarr is an open specification, built on standard components (contrast
+% with Hail).
+
+% \item Zarr is not perfect. Discuss limitations around ragged arrays,
+% etc. Forthcoming improvements in zarr v3. The whole genotype matrix
+% paradigm is challenged by graph genomes as well. While this is
+% being worked out, there is still plenty of data to analyse as a
+% genotype matrix.
+
+% \item Vision. An ecosystem of tools that interoperate
+% based on an open, cloud-based format where distribution is naturally
+% managed by the stored chunks. Tools can be run on different datasets
+% with minimal tuning, not by writing a full layer of orchestration.
+% Tools read the data they need, not the entire dataset. Users read
+% directly from a single canonical dataset, and do not maintain
+% their own filtered copies. We suggest that the Zarr VCF approach
+% we have provided would be a good starting point for designing
+% such a compute platform.
+
+% \end{itemize}
+
 [Random fragment to include somewhere]
@@
 Reel off Python data science ecosystem: Pandas~\citep{mckinney2010data} etc
-
-\section{Methods}
-
-\subsection{Zarr and block-based compression}
-% Zarr is a simple layer on top of best-in-class, modern components
-In the interest of completeness it is useful to provide a high-level overview
-of Zarr and the techologies that it depends upon. Zarr is a specialised format
-for storing large-scale $n$-dimensional numerical data (arrays). Arrays
-are split into chunks, which are compressed and stored separately. Chunks are 
-addressed by their indexes along the dimensions of the array, and the 
-compressed data associated with this key. Chunks are
-are often stored in individual files, but a wide array of different
-stores are supported including cloud buckets and NoSQL databases. 
-Metadata describing the array and its properties is then stored 
-in JSON format along with the chunks. Zarr is best seen as a 
-\emph{specification} that describes the details of how chunks are 
-addressed and stored: there are multiple implementations across
-different languages.
-
-Zarr is flexible in allowing different compression codecs and 
-pre-compression filters to be specified on a per-array basis.
-Two key technologies used by Zarr are the Blosc
-% This seems like the best citation for Blosc? It is mentioned here
-meta-compressor~\cite{alted2010modern}
-and Zstandard compression algorithm~\citep{collet2021rfc}.
-Blosc is a high-performance compressor optimised for numerical
-data which uses ``blocking''~\citep{alted2010modern} to 
-optimise CPU-cache access patterns, as well as highly optimised
-bit and byte shuffle filters.  Remarkably, on highly 
-compressible datasets, Blosc decompression can be faster 
-than \texttt{memcpy}.
-Blosc is written in C, with APIs for C, Python, Julia, Rust
-and others.
-
-Blosc is a ``meta-compressor'' because it provides different 
-access to several different compression codecs. The 
-Zstandard compressor~\citep{collet2021rfc} is of particular 
-interest here as it achieves very high compression ratios
-with fast decompression speeds [see fig]. 
-% This may seem a bit off topic, but want to reassure reader that 
-% this isn't some niche technology
-Zstandard is also used in several recent VCF compression 
-methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
-% LZ4 is also notable - any citations for this?
-
-\subsection{The vcfzarr specification}
-The vcfzarr specification is a direct mapping from the VCF data model 
-to a columnar binary format, expressed using Zarr. [A HIGH LEVEL 
-DESCRIPTION OF THE SPEC, FOCUSING ON RATIONALE BEHIND DECISIONS]
-
-The vcfzarr specification can represent anything described by BCF
-(which is somewhat more restrictive than BCF) except for two corner
-cases related to the encoding of missing data. Firstly, vcfzarr does
-not distinguish between a field that is not present and one that 
-is present but contains missing data. [FOR EXAMPLE...]
-Because of the use of -1 to represent missing data and -2 to 
-represent dimension padding for integers, these values 
-cannot be stored. In practise this doesn't seem to be much of 
-an issue (we have not found a real VCF that contains negative 
-integers). If -1 and -2 need to be stored, a float field
-can be used without issues.
-
-\subsection{vcf2zarr}
-Converting VCF to Zarr format at Biobank scale is challenging. This main issue is 
-around memory usage: although we can view each record in the VCF one-by-one,
-we must buffer a full chunk (10,000 variants is reasonable choice) 
-in the variants dimension for each of the arrays. 
-For VCFs with many FORMAT fields and large numbers of samples this can
-easily require tens of gigabytes of RAM per worker, which makes 
-parallelism troublesome. Reading the VCF multiple times for different fields
-is possible, but would be prohibitively slow for multi-terabyte VCFs.
-
-The \texttt{vcf2zarr} utility solves this problem by first converting 
-the VCF data (which can be split across many files) into an Intermediate
-Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
-of VCFs and stores the data in per-field chunks (by default, 64MiB).
-In this way, the memory required is bounded by the size of single record 
-BCF record in htslib (plus the size of a decoded bgzf record
-[FIXME what is the correct terminology here?]), plus 64MiB for each 
-field.  The Intermediate Columnar Format stores the data for each 
-field spread across multiple partitions of the data (so that 
-multiple parts of the VCF can be read in parallel), and then 
-fixed size compressed chunks within those partitions. It also stores information
-about the number of records in each partition and chunk of a given
-field, so that the record at a particular index can be retrieved efficiently 
-later. Summaries such as the minimum and maximum value 
-of each field are also maintained, to aid choice of data types later.
-A set of VCF files can be converted to intermediate columnar 
-format in parallel on a single machine using a given number 
-of worker processes using the \texttt{explode} command,
-or can be distributed across a cluster using the 
-\texttt{dexplode-init},
-\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.
-
-Once the VCF data has converted to the intermediate columnar format,
-it can then be converted to Zarr using the \texttt{vcf2zarr encode}
-command. Alhough there is a one-to-one mapping between the VCF
-fields and the Zarr arrays, there is considerable flexibility 
-in how the data is stored, for example in integer width
-and compression and filter settings.
-By default vcf2zarr chooses integer type sizes based 
-on the range of values seen in the intermediate column format
-and reasonable compressor defaults, but these can be
-altered by generating a schema in JSON format and editing it.
-Encoding a given column (for example, \texttt{call\_AD})
-involves creating a buffer to hold a full variant-chunk of the 
-array in question, and then sequentially filling this buffer with 
-values read from the intermediate columnar format. Once the buffer 
-is full, the chunks in the samples dimension are compressed
-and written to file. [Something about parallelism and distributing]
-
-\section{Choosing default compressor settings}
-To inform the choice of compression settings across different fields 
-in VCF data, we analysed their effect on compression ratio on 
-data from the 1000 Genomes project.
-
-% [FILL IN WITH Shadi's analysis from here:
-% https://github.com/sgkit-dev/bio2zarr/discussions/74
-% Discussed in:
-% https://github.com/sgkit-dev/vcf-zarr-publication/issues/26
-
-\subsection{Benchmarks}
-Full details of the benchmarking done in
-Figure~\ref{fig-data-storage} and
-Figure~\ref{fig-whole-matrix-compute}.
-Quick notes:
-
-\begin{itemize}
-\item Dataset is based on subsets of French-Canadian sims. We simplify
-for increasingly large subsets, keeping only variant sites. We then
-convert to VCF and encode to BCF using
-\texttt{tskit vcf | bcftools view -O b}.
-\item We used a chunk size of XX for sgkit, after some experimentation.
-This gave n files of around x MB each for the ``call\_genotypes`` array.
-We used the defaults for savvy and genozip.
-\item For the afdist CPU time we measure the sum of the total user and
-system times required to execute the full command, as reported by GNU
-time. Each tool was instructed to use one thread, where the options
-were provided. For genozip, the time required to compute
-\texttt{genocat file.genozip | bcftools +af-dist}. Because commands
-in the pipeline execute concurrently on mutiple cores, the total CPU time is
-greater than the elapsed time.
-\item Where possible in pipelines we use uncompressed BCF
- output \texttt{-Ou} to make processing more efficient (skip printf
-and parsing costs, which are substantial). Following advice
-in~\citep{danecek2021twelve}.
-\item We do not use BCF output in genozip because it doesn't support
-it directly, only VCF (supports BCF by pipeing through bcftools).
-\end{itemize}
-
-\section{Availability of source code and requirements}
-
-\section{Data availability}
-
-\section{Declarations}
-
-\subsection{List of abbreviations}
-% If abbreviations are used in the text they should be defined in the text at
-% first use, and a list of abbreviations should be provided in alphabetical
-% order.
-
-VCF: Variant Call Format;
-
-\subsection{Competing Interests}
-
-\subsection{Funding}
-
-\subsection{Author's Contributions}
-
-\section{Acknowledgements}
-
-\bibliography{paper}
-
-\clearpage
-\renewcommand\thefigure{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand\thetable{S\arabic{table}}
-\setcounter{table}{0}
-
-\section*{Supplementary Material}
-
-\begin{figure}
-\includegraphics{figures/whole-matrix-decode}
-\caption{Genotype decoding performance.
-Total CPU time required to decode genotypes into memory using the zarr-python
-and Savvy C++ APIs for the data in Figure~\ref{fig-data-storage}.
-This corresponds to maximum rate of 1.2GiB/s for Zarr and 5.7GiB/s
-for Savvy. [TODO add data for 1 million samples]
-\label{fig-whole-matrix-decode}}
-\end{figure}
-
-\begin{figure}
-\includegraphics{figures/subset-matrix-compute-supplemental}
-\caption{Compute performance on a large subset of the genotype matrix.
-Total CPU time required to run the af-dist calculation for
-a subset of half of the samples and 10000 variants from the middle of the matrix
-for the data in Figure~\ref{fig-data-storage}.
-Genozip did not run for
-$n > 10^4$ samples because it does not support a file to specify
-sample IDs, and the command line was therefore too long for the shell
-to execute. 
-\label{fig-subset-matrix-compute-supplemental}}
-\end{figure}
-
-\clearpage
-\section*{Text graveyard}
-
-Here's a bunch of text that contains bits that might be useful and readapted.
-
-\subsection{Chunked storage and distributed computation}
-How we store genetic data fundamentally affects how computation
-can be structured, and ultimately the scalability of the calculations
-we wish to perform. Classically, genetic variation data is stored via one
-of two strategies. The first strategy, epitomised by
-VCF~\citep{danecek2011variant} encodes all the data for a given
-variant in a record, and then stores these records consecutively on file
-(optionally compressed). This record-oriented approach maps well to the
-powerful Unix pipelines abstraction and
-has proven to be very successful.
-The second classical strategy, epitomised
-by the \texttt{plink} BED format~\citep{purcell2007plink}, is to
-store genotype data in a structured binary format that can be
-accessed directly. This memory-map approach tends to
-be much faster to access, but leads to larger files and is more
-limited in what can be represented. There are advantages and
-disadvantages to both approaches, but one problem fundamentally
-limits their utility for modern datasets: neither is well suited to
-scaling-out across multiple machines.
-
-[Probably need a paragraph here on what we've done classically
-by splitting by variant. This was the dominant dimension, so that
-worked fine. Now we need to also think about access-by-sample]
-
-Given the scale of modern datasets, the ability to perform
-calculations in parallel on multiple CPU cores
-is now a basic prerequisite, and
-running concurrently on multiple computers in a cluster
-is increasingly becoming necessary. [For interactive analysis? People have
-been doing concurrent computing with HPC clusters forever.]
-Record-oriented variant storage such as VCF is not suitable
-for  distributed computing because although it is relatively
-straightforward to break workloads up by variant, we cannot efficiently
-split by sample (see the section on Columnar Data below for details).
-Storing data in a memory-mapped manner, where the genotypes
-for each variant are written consecutive consecutively to a file,
-suffers from the same problem. Although it is useful to conceptualise
-the genotype matrix as two-dimensional, it is important to remember
-that computer memory and file storage is ultimately one-dimensonal.
-For example, plink's memory-mapped data layout linearises the 2D
-genotype matrix by writing [TODO check] $n$ bytes representing
-the genotypes for all samples at a variant position consecutively.
-Because the genotypes for a given variant index can be found
-in contiguous blocks at easily computable offsets from
-this initial location in memory,
-accessing the genotypes for a given variant is very efficient.
-However, accessing the genotypes for a given sample at every
-variant position is much less efficient because each of the $m$ bytes we
-must read are $n$ bytes apart in the file.
-In the distributed context, this often means
-that each compute node  in a
-cluster must effectively read the entire file in order to obtain the
-data for a subset of samples. In practise, this means that calculations
-are limited by I/0 bandwidth.
-
-A simple and elegant solution to providing efficient access to such
-large matrices across multiple dimensions is to store rectangular
-\emph{chunks} rather than entire rows or columns. This is illustrated
-in Fig~\ref{fig-data-storage}, where we show the basic approach
-used by sgkit. By storing data in these regular chunks, we
-can efficiently access data for both specific variants \emph{and}
-samples.
-
-Additionally, we can compress the individual chunks [more details].
-
-This general strategy for storing multidimensional numerical data
-has been in use for many years~\citep{folk2011overview}
-
-This maps well to distributed computing because Zarr provides a
-very flexible approach to storing the chunks. By using a native cloud store
-for these chunks, compute nodes can directly access the chunks that
-are needed, and not read irrelevant data.
-[TODO flesh out and more stuff.]
-
-\subsubsection{Array oriented computing}
-
-NumPy~\citep{harris2020array} has been transformational, and provides
-the basic framework for scientific computing in Python.
-
 [ Things to cite: pandas~\citep{mckinney2010data},
@@
 
-[NOTE: we may also want a section discussing the n-dimensional array model,
-but I think this section is the right place to discuss that?]
-
-\subsubsection{Columnar binary data}
-
-Perhaps the most important factor limiting scalability in contemporary genomics
-is the continued reliance on row-wise storage of data---most often
-encoded as blockwise-compressed text.
-[Explain why row-wise is bad without reference to specific genomics applications.]
-
-The Variant Call Format (VCF) is the standard method for interchanging
-genomics data~\citep{danecek2011variant}.
-In VCF each row describes the observations for a set of samples
-at a given position along the genome, and consists of fixed columns such as the
-genome position as well as per-sample columns. [More info]
-VCF was introduced as part of the 1000 Genomes project, and works reasonably
-well at the scale of 1000s of samples. However, when we have hundreds of
-thousands of samples it becomes extremely unwieldy
-. For example, the
-phased genotypes for 200K samples in UK Biobank produced by Browning [cite]
-require X GB of space.
-(Note that this file contains no extra INFO keys; in practice a great deal
-more information is stored in the VCF leading to files that are orders of
-magnitude larger, and requiring them to be split into chunks, further
-increasing their unwieldiness.)
-Running [a simple query that requires only looking at
-one column] on this file like X required Y minutes.
-
-[TODO be more concrete here; what is the thing we're summarising?]
-This slow response time to a simple query is primarily caused by the row-wise
-nature of VCF (and BCF), because in order to read a single piece of
-information about a particular variant site we must read \emph{all}
-the information
-about that site. Then, if we wish to summarise that information over
-all sites, then we must read the entire dataset.
-
-A lesser but still significant cause of inefficiency is the use of
-text to store the data. [More info]
-
-These issues are of course well known, and there are numerous projects
-that seek to alleviate them. [Discuss] They're all quite bespoke to genomics data
-through.
-
-In sgkit we use Zarr which although initially introduced to store
-genomics data, has now been adopted widely in [X, Y and Z].
-Zarr is [description] and has [advantages]
-
-Also to discuss (not sure how these fit in to this specific sectio
-on background technologies vs existing genomics infrastructure; can
-sort out later):
-
-\begin{itemize}
-\item columnar (with small records) good for I/O elimination + sequential access.
-Classical locality of reference stuff as well.
-\item What does this mean with ND arrays? Slippery concept. Discuss.
-\item Compression improvements from columnar.
-\item Poor performance of text-based VCF has been noted many times,
-and lots of prior work in this space e.g. \citep{kelleher2013processing}
-\item Honourable mentions for bcftools~\citep{danecek2021twelve},
-htslib~\citep{bonfield2021htslib}, CRAM~\citep{bonfield2014scramble,bonfield2022cram}
-which uses a columnar approach
-\end{itemize}
-
-\subsection{Storing genetic variation data}
-
-It is important to note that the example in Fig.~\ref{fig-data-storage}
-is something of a best-base
-scenario in terms genomic dataset storage, in that we are only
-storing genotype data (per sample, per variant),
-and, because it is error-free, is very
-highly compressible. Noisier genotypes will not compress quite
-so well, and storing less compressible data (such as genotype
-quality values) results in a large increase in the overall
-storage size. This is essentially unavoidable, but the Zarr-based
-approach still has some significant advantages over VCF/BCF
-[basically we can store separately and come up with distinct
-compression strategies based on the data type. Dunno if we
-want to get into this here. This is a complex topic with
-a lot of existing work which we should review and reference,
-and this is maybe not the place to do it.].
-
-[SpVCF has some good text on high-entropy QC measures. We have to
-reduce this entropy or we just can't store the data]
-
-[Why are we reviewing all this? Well, we want to make the point
-that people have been trying hard to address the problem of
-how to deal with large-scale genomic data for quite a while,
-and have come up with lots of specialised methods. There
-are a few things to keep in mind. Our approach is to use
-standard methods just on the "rectangles" of variant data
-and it's already quite good. You could imagine writing specialised
-compression filters for genotypes, if you wanted (maybe a
-discussion item though). The key thing is the \emph{interface}.
-They're all tying to the command line and to classical
-sequential workflows. We have Python, and distributed compute. ]
-
-Storage is the foundation on which any large-scale data-driven computation
-rests: how we encode and organise data on persistent storage (the format)
-fundamentally affects and limits the efficiencies of downstream calculations.
-Different formats are optimised for different purposes, and there are
-often many competing requirements when designing a data storage format.
-% TODO is there some literature we could cite on this? I'm making up the
-% term "computational format" here, presumably there is lots  prior thought
-% on these ideas?
-For our  purposes here, it is important to distinguish between \emph{archival}
-and \emph{computational} formats. An archival format is optimised for
-maximum interoperability and future-proofing, and uses the simplest and
-most robust technologies to ensure the long-term usability of the data.
-Reasonable compression levels are important because
-of storage and transfer costs, but the efficiency of data access is a
-secondary concern.
-A computational format, on the other hand, is optimised
-for calculations, for \emph{using} the data, today. A computational format
-is designed with particular data access patterns in mind, ultimately
-to make a broad range of queries and downstream computations as efficient as
-possible.
-
-% Note: a quick explanation of VCF for CS people coming in fresh
-VCF~\citep{danecek2011variant} is a text-based format in which each line
-describes observations
-and metadata for a single variant (a position on the genome at which
-variation occurs in the sample). The first eight tab-delimited
-columns are mandatory, and subsequent columns describe the per-sample
-genotype calls and quality control data.
-
-[WIP: currently working on weaving together the stuff below into
-some sort of narrative]
-
-SeqArray~\citep{zheng2017seqarray} and the GDS format~\citep{zheng2012high}
-provide efficient storage and programmatic access in R.
-
-The approach taken by \toolname{plink}~\citep{purcell2007plink,chang2015second} is to
-store data in an uncompressed binary format. This has some major
-advantages.
-
-BGEN \citep{band2018bgen} is optimised for storing large-scale
-genotype array data and stores probabilities, and is used
-in particular by UK Biobank~\citep{bycroft2018genome}
-[Should consider the "hard-call" nature of most of these formats]
-
-Many methods working with large scale genetic
-data now use the PBWT or variants of
-it~\citep[e.g.][]{li2016bgt,lefaive2021sparse,wertenbroek2022xsi}. PBWTs have
-disadvantages, however: data must be phased, and compression level
-is quite sensitive to the presence of errors [citation? people have
-worked around this]
-
-Several tools provide querying interfaces in their CLIs
-(reminiscent of SQL) to provide easy access to particular samples, or
-variants~\cite{li2016bgt,layer2016efficient}.
-
-More recently XSI~\citep{wertenbroek2022xsi} is an file format that
-uses a combination of strategies to achieve excellent compression
-performance, and can support some calculations directly on the
-compressed format. It provides a command line interface and a
-C API.
-
-The SAV format is an extension of BCF that takes advantage
-of the sparseness of large scale genomic datasets~\citep{lefaive2021sparse}.
-They provide a command line interface and a C++ API.
-% TEXT from HSG chapter section on compressing data
-
-% The simplest approach to compressing such data is to organise it in a
-% `variant-centric' fashion, such that all observed genotypes for a given variant
-% are stored in a contiguous block. Because most variants are rare, this will
-% result in long runs of identical values that compress well using standard
-% methods. This is the approach taken by the Variant Call
-% Format~\citep{danecek2011variant}, and its more efficient binary encoding, BCF.
-% The compression levels achieved by this straightforward approach are good, and
-% quite competitive with more sophisticated methods. The disadvantage
-% is that many queries require full decompression of the data, which can be
-% prohibitively time-consuming. The SpeedGene data format and software
-% library~\citep{qiao2012handling} chooses from one of three encoding methods for
-% each SNP determined by the allele frequency; \cite{sambo2014compression} extend
-% this idea by identifying blocks of SNPs in LD, and adding two additional
-% potential encodings utilising this information.
-
-% An alternative to this variant-centric approach is to store genotypes in a
-% `sample-centric' fashion. Here, all genotypes for a particular sample are
-% stored consecutively. While this breaks the simple repetition structure of data
-% stored in variant-centric form, other methods can be employed to find
-% compressible structure. For example, TGC~\citep{deorowicz2013genome} compresses
-% sample genotypes using a custom Lempel-Ziv style approach on runs of identity
-% among samples. This explicit use of LD structure results in excellent
-% compression performance (for example, 32MB for all SNPs in 1000 Genomes
-% chromosome 1); unfortunately, querying a dataset requires full decompression,
-% making the format unsuitable for online analysis. The GQT
-% toolkit~\citep{layer2016efficient} also uses the sample-centric organisation of
-% genotype data, but takes a different approach to compression. Variants are
-% sorted by allele frequency (resulting in longer runs of identical values within
-% samples) and then compressed using an efficient bit-vector
-% representation~\citep{wu2002compressing}. The resulting file-sizes are similar to
-% compressed BCF, but many queries can be performed directly on the compressed
-% representation and are therefore many times faster.
-
-% The positional Burrows-Wheeler transform~\citep{durbin2014efficient}, or PBWT,
-% is another sample-centric method. Building on the success of Burrows-Wheeler
-% transform applications in
-% genomics~\citep{langmead2009ultrafast,li2009fast,li2009soap2}, PBWT provides
-% very good compression performance and efficient algorithms for finding
-% haplotype matches. The algorithm builds a representation of the data based
-% on sorted haplotype prefixes in linear time, and the LD structure of the data
-% ensures that the sorting orders between adjacent sites change very little
-% on average. The method has been successfully applied
-% to phasing~\citep{loh2016reference}, detection of IBD
-% segments~\citep{naseri2017ultra}, improving the performance of the Li and
-% Stephens model~\citep{lunter2016fast}, and a general query engine for genotype
-% data~\citep{li2016bgt}. Recent extensions include privacy preserving
-% search~\citep{shimizu2016efficient} and generalisation to the setting of graph
-% genomes~\citep{novak2017graph}.
-
-\subsection{Distributed computation}
+Although the Hadoop ecosystem has been hugely successful
+in many domains, it has had a limited impact in genomics.
+[List some things that have been tried for e.g variant calling
+, like ADAM~\citep{nothaft2015rethinking}].
+[Briefly mention two approaches using Hadoop clusters
+~\citep{boufea2017managing,fan2020variant}, to not very
+exiting results].
+Distributed analytics for large-scale data is a mature
+techology, with many different approaches. Hadoop~\citep{white2012hadoop}
+is the original method, and remains the basis of the
+dominant ecosystem. HDFS. Spark [CITATION?] lets us
+do more complex analytics. Parquet [CITATION?]
 Distributed computation has been the reality in the analysis
@@
 
-Despite this long history of de-facto distributed computation
-and the clear need for better methods of accessing genetic
-variation data in a distributed fashion, there is almost
-no scientific literature on the topic (in stark contrast to the
-study of methods to compress a VCF into a single file,
-as discussed in section XXX).
-
-[This is a really rough outline, also not sure how exactly
-we sequence the ideas here. Just getting paragraph fragments
-on paper. We do need to explain something of the hadoop ecosystem
-to readers who are not familiar, though.]
-Distributed analytics for large-scale data is a mature
-techology, with many different approaches. Hadoop~\citep{white2012hadoop}
-is the original method, and remains the basis of the
-dominant ecosystem. HDFS. Spark [CITATION?] lets us
-do more complex analytics. Parquet [CITATION?]
-
-Although the Hadoop ecosystem has been hugely successful
-in many domains, it has had a limited impact in genomics.
-[List some things that have been tried for e.g variant calling
-, like ADAM~\citep{nothaft2015rethinking}].
-[Briefly mention two approaches using Hadoop clusters
-~\citep{boufea2017managing,fan2020variant}, to not very
-exiting results].
-
-[Hadoop/Spark/Parquet are fundamentally unsuited to genetic
-variation data [WHY]?]
-
-Discuss Hail. It's awesome, has enabled current
-generation of mega-scale projects. Drawbacks are the
-architecture has emphasised performance etc over adaptability
-and simplicity. Uses bits of Spark for X, but custom YZ .
-File format is undocumented, etc.
-
-Discuss why the sgkit architecture is suited to distributed
-computation. Discuss Zarr's use in cloud stores, and existing
-deployments. Discuss Dask.
-We do not attempt a head-to-head performance comparison with
-Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
-with brief summary.
-Forward ref to following sections on applications to demonstate
-sgkit in a distributed context.
+\section{Methods}
+
+\subsection{Zarr and block-based compression}
+% Zarr is a simple layer on top of best-in-class, modern components
+In the interest of completeness it is useful to provide a high-level overview
+of Zarr and the techologies that it depends upon. Zarr is a specialised format
+for storing large-scale $n$-dimensional numerical data (arrays). Arrays
+are split into chunks, which are compressed and stored separately. Chunks are 
+addressed by their indexes along the dimensions of the array, and the 
+compressed data associated with this key. Chunks are
+are often stored in individual files, but a wide array of different
+stores are supported including cloud buckets and NoSQL databases. 
+Metadata describing the array and its properties is then stored 
+in JSON format along with the chunks. Zarr is best seen as a 
+\emph{specification} that describes the details of how chunks are 
+addressed and stored: there are multiple implementations across
+different languages.
+
+Zarr is flexible in allowing different compression codecs and 
+pre-compression filters to be specified on a per-array basis.
+Two key technologies used by Zarr are the Blosc
+% This seems like the best citation for Blosc? It is mentioned here
+meta-compressor~\cite{alted2010modern}
+and Zstandard compression algorithm~\citep{collet2021rfc}.
+Blosc is a high-performance compressor optimised for numerical
+data which uses ``blocking''~\citep{alted2010modern} to 
+optimise CPU-cache access patterns, as well as highly optimised
+bit and byte shuffle filters.  Remarkably, on highly 
+compressible datasets, Blosc decompression can be faster 
+than \texttt{memcpy}.
+Blosc is written in C, with APIs for C, Python, Julia, Rust
+and others.
+
+Blosc is a ``meta-compressor'' because it provides different 
+access to several different compression codecs. The 
+Zstandard compressor~\citep{collet2021rfc} is of particular 
+interest here as it achieves very high compression ratios
+with fast decompression speeds [see fig]. 
+% This may seem a bit off topic, but want to reassure reader that 
+% this isn't some niche technology
+Zstandard is also used in several recent VCF compression 
+methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
+% LZ4 is also notable - any citations for this?
+
+% Zarr was originally developed as a simpler version of 
+% HDF5~\citep{folk2011overview}
+
+\subsection{The vcfzarr specification}
+The vcfzarr specification is a direct mapping from the VCF data model 
+to a columnar binary format, expressed using Zarr. [A HIGH LEVEL 
+DESCRIPTION OF THE SPEC, FOCUSING ON RATIONALE BEHIND DECISIONS]
+
+The vcfzarr specification can represent anything described by BCF
+(which is somewhat more restrictive than BCF) except for two corner
+cases related to the encoding of missing data. Firstly, vcfzarr does
+not distinguish between a field that is not present and one that 
+is present but contains missing data. [FOR EXAMPLE...]
+Because of the use of -1 to represent missing data and -2 to 
+represent dimension padding for integers, these values 
+cannot be stored. In practise this doesn't seem to be much of 
+an issue (we have not found a real VCF that contains negative 
+integers). If -1 and -2 need to be stored, a float field
+can be used without issues.
+
+The vcfzarr specification is general and can be mapped to 
+file formats such as \toolname{plink}~\citep{purcell2007plink,chang2015second}
+and BGEN~\citep{band2018bgen} with some minor extensions.
+
+\subsection{vcf2zarr}
+Converting VCF to Zarr format at Biobank scale is challenging. This main issue is 
+around memory usage: although we can view each record in the VCF one-by-one,
+we must buffer a full chunk (10,000 variants is reasonable choice) 
+in the variants dimension for each of the arrays. 
+For VCFs with many FORMAT fields and large numbers of samples this can
+easily require tens of gigabytes of RAM per worker, which makes 
+parallelism troublesome. Reading the VCF multiple times for different fields
+is possible, but would be prohibitively slow for multi-terabyte VCFs.
+
+The \texttt{vcf2zarr} utility solves this problem by first converting 
+the VCF data (which can be split across many files) into an Intermediate
+Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
+of VCFs and stores the data in per-field chunks (by default, 64MiB).
+In this way, the memory required is bounded by the size of single record 
+BCF record in htslib (plus the size of a decoded bgzf record
+[FIXME what is the correct terminology here?]), plus 64MiB for each 
+field.  The Intermediate Columnar Format stores the data for each 
+field spread across multiple partitions of the data (so that 
+multiple parts of the VCF can be read in parallel), and then 
+fixed size compressed chunks within those partitions. It also stores information
+about the number of records in each partition and chunk of a given
+field, so that the record at a particular index can be retrieved efficiently 
+later. Summaries such as the minimum and maximum value 
+of each field are also maintained, to aid choice of data types later.
+A set of VCF files can be converted to intermediate columnar 
+format in parallel on a single machine using a given number 
+of worker processes using the \texttt{explode} command,
+or can be distributed across a cluster using the 
+\texttt{dexplode-init},
+\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.
+
+Once the VCF data has converted to the intermediate columnar format,
+it can then be converted to Zarr using the \texttt{vcf2zarr encode}
+command. Alhough there is a one-to-one mapping between the VCF
+fields and the Zarr arrays, there is considerable flexibility 
+in how the data is stored, for example in integer width
+and compression and filter settings.
+By default vcf2zarr chooses integer type sizes based 
+on the range of values seen in the intermediate column format
+and reasonable compressor defaults, but these can be
+altered by generating a schema in JSON format and editing it.
+Encoding a given column (for example, \texttt{call\_AD})
+involves creating a buffer to hold a full variant-chunk of the 
+array in question, and then sequentially filling this buffer with 
+values read from the intermediate columnar format. Once the buffer 
+is full, the chunks in the samples dimension are compressed
+and written to file. [Something about parallelism and distributing]
+
+Fields such as PL are currently problematic when there are large 
+numbers of alleles. However, we plan on implementing the 
+local alleles strategy~\citep{poterba2024scalable,danecek2021twelve}
+which will likely become part of the VCF specification.
+
+\section{Choosing default compressor settings}
+To inform the choice of compression settings across different fields 
+in VCF data, we analysed their effect on compression ratio on 
+data from the 1000 Genomes project.
+
+% [FILL IN WITH Shadi's analysis from here:
+% https://github.com/sgkit-dev/bio2zarr/discussions/74
+% Discussed in:
+% https://github.com/sgkit-dev/vcf-zarr-publication/issues/26
+
+\subsection{Benchmarks}
+Full details of the benchmarking done in
+Figure~\ref{fig-data-storage} and
+Figure~\ref{fig-whole-matrix-compute}.
+Quick notes:
+
+\begin{itemize}
+\item Dataset is based on subsets of French-Canadian sims. We simplify
+for increasingly large subsets, keeping only variant sites. We then
+convert to VCF and encode to BCF using
+\texttt{tskit vcf | bcftools view -O b}.
+\item We used a chunk size of XX for sgkit, after some experimentation.
+This gave n files of around x MB each for the ``call\_genotypes`` array.
+We used the defaults for savvy and genozip.
+\item For the afdist CPU time we measure the sum of the total user and
+system times required to execute the full command, as reported by GNU
+time. Each tool was instructed to use one thread, where the options
+were provided. For genozip, the time required to compute
+\texttt{genocat file.genozip | bcftools +af-dist}. Because commands
+in the pipeline execute concurrently on mutiple cores, the total CPU time is
+greater than the elapsed time.
+\item Where possible in pipelines we use uncompressed BCF
+ output \texttt{-Ou} to make processing more efficient (skip printf
+and parsing costs, which are substantial). Following advice
+in~\citep{danecek2021twelve}.
+\item We do not use BCF output in genozip because it doesn't support
+it directly, only VCF (supports BCF by pipeing through bcftools).
+\end{itemize}
+
+\section{Availability of source code and requirements}
+
+\section{Data availability}
+
+\section{Declarations}
+
+\subsection{List of abbreviations}
+% If abbreviations are used in the text they should be defined in the text at
+% first use, and a list of abbreviations should be provided in alphabetical
+% order.
+
+VCF: Variant Call Format;
+
+\subsection{Competing Interests}
+
+\subsection{Funding}
+
+\subsection{Author's Contributions}
+
+\section{Acknowledgements}
+
+\bibliography{paper}
+
+\clearpage
+\renewcommand\thefigure{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{table}{0}
+
+\section*{Supplementary Material}
+
+\begin{figure}
+\includegraphics{figures/whole-matrix-decode}
+\caption{Genotype decoding performance.
+Total CPU time required to decode genotypes into memory using the zarr-python
+and Savvy C++ APIs for the data in Figure~\ref{fig-data-storage}.
+This corresponds to maximum rate of 1.2GiB/s for Zarr and 5.7GiB/s
+for Savvy. [TODO add data for 1 million samples]
+\label{fig-whole-matrix-decode}}
+\end{figure}
+
+\begin{figure}
+\includegraphics{figures/subset-matrix-compute-supplemental}
+\caption{Compute performance on a large subset of the genotype matrix.
+Total CPU time required to run the af-dist calculation for
+a subset of half of the samples and 10000 variants from the middle of the matrix
+for the data in Figure~\ref{fig-data-storage}.
+Genozip did not run for
+$n > 10^4$ samples because it does not support a file to specify
+sample IDs, and the command line was therefore too long for the shell
+to execute. 
+\label{fig-subset-matrix-compute-supplemental}}
+\end{figure}
 

### 2024-04-12 paper.tex
Quarter-baked discussion

@@
 project~\cite{10002015global} has transformed a fragmented 
-
-% A computational format, on the other hand, is optimised
-% for calculations, for \emph{using} the data, today. A computational format
-% is designed with particular data access patterns in mind, ultimately
-% to make a broad range of queries and downstream computations as efficient as
-% possible.
-
-% Outline of a possible narrative by paragraph:
-% \begin{itemize}
-% \item When VCF was introduced as part of the work of the 1000 Genomes
-% project the software landscape was a real mess. It has entirely
-% succeeded in standardising the output of variant callers, and methods
-% have interoperated well for the past decade. As a method of interchange
-% following classical file-oriented paradigms it works well enough.
-% It is also a good archival format: when compressed with a well established
-% method like gzip and combined with the well-documented description of
-% the text, we can be sure that files will still be readable decades
-% from now.
-
-% \item VCF is holding back large-scale genomics, however, and in particular the
-% analysis of population scale genetic variation datasets like UKB and GeL.
-% Describe the current state of affairs. Variant datasets are chunked up
-% into reasonable sized bits (10s of gigabytes) so that they can be processed
-% in parallel. Doing anything with them requires interacting with workflow
-% managers. In practise, it is not possible to directly write programs to
-% work across multiple large-scale datasets: although the chunks of data
-% are given in VCF, orchestrating how the chunks are processed and
-% the results combined involves substantial work, involving the workflow
-% manager, and the details of how chromosomes are split into chunks.
-% Cite examples like the phasing pipelines for UKB from Brownings etc.
-
-% \item Distributed computing is the reality, and increasingly datasets
-% are being mandatorily stored in public clouds as part of
-% Trusted Research Environments. Cloud object stores are the reality
-% because it is much cheaper. POSIX file systems are emulated on top of
-% this. Benefits of breaking up data into manageable pieces are that
-% compute resources can be localised to the data. Compute resources are
-% most effectively obtained as short-lived, small VMs that operate
-% as part of a large cluster. Large vertically scaled ``pet'' servers
-% are much more expensive, and just not feasible for the current
-% scale of data. Lots of other architectural
-% benefits from the new reality of cloud-based storage.
-
-% \item Classical methods based around Unix pipelines and streaming
-% files between processes on a single machine are therefore a
-% significant mismatch to the reality of large-scale genetic variation
-% datasets. The majority of methods suggested to make working
-% with VCF data more efficient are based on these ideas: they all
-% create a single large file and regard distribution of work
-% in a cluster as extrinsic. Most output VCF as the result of queries,
-% which also assumes that the VCF data can be efficiently processed.
-% This is a reasonable assumption for thousands of samples, but not
-% for larger datasets.
-
-% \item The key issue here is that VCF is not a \emph{computational}
-% format. Contrast VCF with plink, which stores data on disk in a
-% form that can be computed on directly. Plink is limited in what
-% can be descibed though, does not compress, and is not designed
-% to distributed.
-
-% \item The large-scale analysis of genetic variation data does not
-% need a more efficient file format, which allows files to be downloaded
-% somewhat more quickly, and subsets extracted somewhat more efficiently
-% as VCF. It needs a standard storage platform that allows data to be
-% processed directly without conversion, subsetted efficiently across
-% both the variant and sample dimensions, with first-class support
-% for all the necessary additional information (GT values etc),
-% and that is suited to the realities of today's cloud-based
-% storage and processing facilities.
-
-% \item We suggest that the model used by Zarr is a good choice for this.
-% We have shown that chunks of the genotype matrix compressed with standard
-% methods are competitive with highly specialised methods for
-% genotype data. In reality most of the storage is for non-genotype
-% data anyway, where everyone is essentially doing the same thing.
-% Columnar storage allows getting just (e.g.) the variant positions
-% without reading everything. Another benefit is that by storing
-% fields separately in different cloud objects, they can have
-% different permissions. So, users can be given differential access
-% to read parts of the dataset, and this is managed directly
-% by the cloud storage. Give example.
+landscape of mutually incompatible per-program file formats
+into an ecosystem of tools that largely
+interoperate~\cite{garrison2022spectrum}. In this 
+sense VCF has been a huge success, and the underlying data model
+is now deeply embedded in bioinformatics practice. As we have 
+illustrated here, the shortcomings of the VCF and BCF file
+formats (and more generally \emph{any} row-wise encoding of the 
+data model) in terms of computational accessibility are severe,
+and there can be little doubt that these file formats 
+are incurring a large economic cost and 
+holding back progress. 
+As we have illustrated in the Genomics England
+example here, these shortcomings are most acutely felt in 
+large-scale data human datasets but any setting in which we 
+wish to extract columns of the data from a row-wise 
+VCF format will be inefficient 
+when datasets are larger than a few hundred megabytes.
+
+As we have demonstrated, the columnar approach, based on the 
+popular Zarr standard and widely used compression libraries, 
+can support many different types of data access
+patterm efficiently, and has excellent computational accessibility.
+In this approach there are some significant departures from 
+standard practise, which warrant some discussion here.
+
+Perhaps the largest deviation in the approach that we advocate here 
+from standard practise is to decouple our concept of a dataset from
+a file. Each VCF is broken up into a set of Zarr arrays, each 
+corresponding to a large number of chunks stored in a directory
+hierarchy (Zarr also supports other forms of chunk storage 
+including single files and key-value stores however; see Methods).
+While there are some disadvantages to large numbers of small files,
+we argue that the benefits greatly outweigh them, particularly 
+in the large, centrally controlled dataset setting. 
+Firstly, this approach is inherently parallelisable, and chunks across the 
+dataset written and read in parallel without needing complex 
+synchronisation and prior layout planning. 
+Such parallel write access is very difficult 
+in the single-file setting, and one of the key weaknesses 
+of the HDF5 format~\citep{folk2011overview} [CITATION with 
+parallel troubles from HDF5?]
+Secondly, mapping chunks in arrays to individually addressed
+keys maps well to large-scale cloud object storage. Indeed,
+Zarr is specifically designed for such usage, and 
+several petabyte scale datasets have been successfully deployed
+and used [citations?].
+We have focused here on the classical single-server and 
+HPC cluster settings, but a major advantage of Zarr is 
+its cloud readiness.
+% Probably getting into the weeds too much.
+% Some large cloud-hosted VCF datasets are currently stored in 
+% filesystems that emulate POSIX semantics on top of such object 
+% stores, at a substantial cost in performance. Directly hosting 
+% these datasets in object stores would have substantial 
+% benefits, both in terms of storage costs and performance.
+Thirdly, breaking up fields into different directories (or 
+cloud buckets) enables fine-grained access control to the 
+dataset. For example, all users may be allowed access to
+variant-level data, but only a subset permitted to access
+call-level individual data. Such access controls can be 
+straightforwardly applied using file or cloud-hosting 
+methods on the relevant directores/buckets. Such concerns
+are entirely decoupled from the file format.
+
+% This is poorly explained
+Along with the move away from the single-file-storage paradigm,
+the approach we outline here also suggests a shift from 
+the classical Unix pipeline-based processing approach 
+(but see below for a discussion of how existing pipelines can
+be supported). Datasets flowing through pipelines inherently 
+implies creating muliple copies of the data, and when 
+multiple terabytes of either unrelated or unchanged data 
+get copied several times through these pipes large inefficiencies 
+are incurred (see Fig~\ref{fig-whole-matrix-compute}). 
+We may think of this strategy as ``copy-oriented'' workflows.
+A more efficient way to structure such multilayered computations
+is in a  ``mask-oriented'' way. For example, to implement some
+complex multilayered call filtering strategy we could imagine sequentially
+updating a given per-call mask. Masks are cheap (e.g., the 
+standard genotype missingness mask in~\ref{tab-genomics-england-data}
+uses less than 1\% of the overall storage).
+
+Zarr provides pragmatic solutions to some of the more pressing 
+problems facing the analysis of large-scale genetic variation
+data, but it is not a solution to all problems. [large number
+of alleles, dataset is basically static, ragged arrays,
+strings ... any others? Zarr v3 on the way.]
+
+Nonetheless, at least in the single-shared-copy model currently
+used by most large-scale variation datasets, Zarr provides
+an excellent basis for building a new generation of tools 
+that compute \emph{directly} on the data, efficiently
+accessing data by variant and by sample.
+The vcfzarr specification and scalable vcf2zarr conversion utility
+that we have provided here are the starting points of such 
+an ecosystem, but much more would be needed to make a viable 
+alternative for everyday bioinformatics workflows. However, a few key
+pieces of infrastructure would provide a starting point, and would 
+be sufficient to greatly improve researcher productivity on 
+large, centrally managed datasets such as Genomics England.
+Firstly, a \texttt{vcfzarrtools} command line utility which 
+implements a subset of \texttt{bcftools} functionality (such as 
+\texttt{view} and \texttt{query}) would immediately speed up 
+limited region and field specific queries by orders of magnitude,
+and would provide compatibility with existing workflows.
+Developing such a utility would be a non-trivial but routine
+engineering task.
+
+The second key piece of infrastructure required at this large 
+scale concerns distributed computation. Even with more efficient
+columnar data representations, computations that required examing 
+large volumes of per-call data must be distributed to complete 
+in a reasonable amount of time. [Dask, Xarray, sgkit?]
+Currently users are responsible for orchestrating such calculations
+on clusters, and automation would be a significant boon
+for reasearcher productivity. [Hail has demonstrated the value of this?]
+
+% We suggest that the future of VCF is to explicitly define the underlying
+% data model, and to decouple it from the current row-wise encodings 
+% of the VCF text and BCF binary file formats. This will enable data 
+% to flow more freely, and provide scope for much-needed innovation.
 
@@
 
-% \item Zarr is not perfect. Discuss limitations around ragged arrays,
-% etc. Forthcoming improvements in zarr v3. The whole genotype matrix
-% paradigm is challenged by graph genomes as well. While this is
-% being worked out, there is still plenty of data to analyse as a
-% genotype matrix.
-
 % \item Vision. An ecosystem of tools that interoperate
@@
 
-Although the Hadoop ecosystem has been hugely successful
-in many domains, it has had a limited impact in genomics.
-[List some things that have been tried for e.g variant calling
-, like ADAM~\citep{nothaft2015rethinking}].
-[Briefly mention two approaches using Hadoop clusters
-~\citep{boufea2017managing,fan2020variant}, to not very
-exiting results].
-Distributed analytics for large-scale data is a mature
-techology, with many different approaches. Hadoop~\citep{white2012hadoop}
-is the original method, and remains the basis of the
-dominant ecosystem. HDFS. Spark [CITATION?] lets us
-do more complex analytics. Parquet [CITATION?]
-Distributed computation has been the reality in the analysis
-of genetic variation for many years,
-where data is most often provided as per-chromosome VCFs.
-This chunking provides a fairly straightforward way to parallelise
-computations, using HPC schedulers to split work per-file.
-When more fine-grained parallelism is required, users must
-provide exact genome coordinates of the genome regions
-as part of their submission scripts. Combining results
-is a manual operation.
-Workflow engines such as
-Snakemake~\cite{koster2012snakemake,molder2021sustainable}
-and [TODO mention WDL, Cromwell, etc?] make such analyses
-much easier, but there is still a great deal of room for
-mistakes.
-Per-chromosome VCFs are not practical for biobank scale
-projects, however, and variant call data tends to be split
-into many (still large) files.
-For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
-are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
-hundreds of files per chromosome.
-% This is horrible writing, but the point is important.
-While these files provide a natural
-unit for distributing work, the details of how they
-are split differ and essentially there is
-no interoperability among large-scale datasets because the VCFs are
-so cumbersome that a separate orchestration layer is required to
-access the data in parallel.
+% Although the Hadoop ecosystem has been hugely successful
+% in many domains, it has had a limited impact in genomics.
+% [List some things that have been tried for e.g variant calling
+% , like ADAM~\citep{nothaft2015rethinking}].
+% [Briefly mention two approaches using Hadoop clusters
+% ~\citep{boufea2017managing,fan2020variant}, to not very
+% exiting results].
+% Distributed analytics for large-scale data is a mature
+% techology, with many different approaches. Hadoop~\citep{white2012hadoop}
+% is the original method, and remains the basis of the
+% dominant ecosystem. HDFS. Spark [CITATION?] lets us
+% do more complex analytics. Parquet [CITATION?]
+% Distributed computation has been the reality in the analysis
+% of genetic variation for many years,
+% where data is most often provided as per-chromosome VCFs.
+% This chunking provides a fairly straightforward way to parallelise
+% computations, using HPC schedulers to split work per-file.
+% When more fine-grained parallelism is required, users must
+% provide exact genome coordinates of the genome regions
+% as part of their submission scripts. Combining results
+% is a manual operation.
+% Workflow engines such as
+% Snakemake~\cite{koster2012snakemake,molder2021sustainable}
+% and [TODO mention WDL, Cromwell, etc?] make such analyses
+% much easier, but there is still a great deal of room for
+% mistakes.
+% Per-chromosome VCFs are not practical for biobank scale
+% projects, however, and variant call data tends to be split
+% into many (still large) files.
+% For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
+% are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
+% hundreds of files per chromosome.
+% % This is horrible writing, but the point is important.
+% While these files provide a natural
+% unit for distributing work, the details of how they
+% are split differ and essentially there is
+% no interoperability among large-scale datasets because the VCFs are
+% so cumbersome that a separate orchestration layer is required to
+% access the data in parallel.
 

### 2024-04-12 src/plot.py
Missing tight_layout

@@ -62,6 +62,7 @@ def plot_size(ax, df, label_y_offset=None):
 
     ax.legend()
     add_number_of_variants(df, ax)
+    plt.tight_layout()
 
 
 def plot_total_cpu(ax, df, toolname=None, time_units="h", extrapolate=None):

### 2024-04-12 paper.tex

@@
 accessing data by variant and by sample.
+[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
 The vcfzarr specification and scalable vcf2zarr conversion utility
@@
 
+Zarr is now widely used.
+
+Multi terabyte numerical weather prediction models~\citep{gowan2022using}
+Petabyte scale geoscience datasets in the Pangeo project
+
+OME-Zarr~\cite{rzepka2023toward}
+ITS\_LIVE ice velocity dataset~\citep{fahnestock2023mappin} using 
+the Zarr.js Javascript interface to visualise the ``massive''
+dataset [how big is it?]
+
 \subsection{The vcfzarr specification}

### 2024-04-12 src/collect_data.py
Some infrastructure for testing noshuffle

@@ -18,6 +18,7 @@ import click
 import sgkit as sg
 
 
+
 # yuck - but simplest way to avoid changing directory structure
 sys.path.insert(0, "src")
 from zarr_afdist import zarr_afdist, classify_genotypes_subset_filter, zarr_decode
@@ -728,6 +729,19 @@ def report_versions():
         print(tool.name)
         print(tool.version_func())
 
+@click.command()
+@click.argument("path", type=click.Path())
+def benchmark_zarr_decode(path):
+    before = time.time()
+    bytes_decoded = zarr_decode(path)
+    wall_time = time.time() - before
+
+    print(
+        "time = ", wall_time, " s ",
+        "rate = ",
+        humanize.naturalsize(bytes_decoded / wall_time, format="%.1f"),
+        "/ s",
+    )
 
 @click.group()
 def cli():
@@ -742,6 +756,7 @@ cli.add_command(column_extract)
 cli.add_command(genotype_filtering_processing_time)
 cli.add_command(site_filtering_processing_time)
 cli.add_command(report_versions)
+cli.add_command(benchmark_zarr_decode)
 
 
 if __name__ == "__main__":

### 2024-04-15 paper.tex
Some Zarr bio refs
Closes #7
Closes #23

@@
 
+Zarr has recently been gaining traction in biological applications
+also. EUGENe~\citep{klie2023predictive} is a toolkit for 
+for applying deep-learning methods to genomic sequence data,
+integrating with PyTorch [citation?].
+
+Spatial Omics:
+emObject~\citep{baker2023emobject}
+SpatialData~\citep{marconato2024spatialdata}
+
 \subsection{The vcfzarr specification}

### 2024-04-15 paper.tex
DivBrowse
Closes #3

@@
 
-OME-Zarr~\cite{rzepka2023toward}
 ITS\_LIVE ice velocity dataset~\citep{fahnestock2023mappin} using 
@@
 
-Spatial Omics:
-emObject~\citep{baker2023emobject}
-SpatialData~\citep{marconato2024spatialdata}
+Spatially resolved omics data~\citep{baker2023emobject,marconato2024spatialdata}
+building on the Zarr-based OME-NGFF format~\citep{moore2023ome,rzepka2023toward}
+
+DivBrowse~\citep{konig2023divbrowse}. Works with VCF data using 
+the scikit-allel \texttt{vcf\_to\_zarr} function, (produces 
+a precursor to the VCF-Zarr specification here). It's a 
+browser, using Zarr as the storage backend.
 

### 2024-04-16 src/collect_data.py
Add some infrastructure for reading noshuffle results

@@ -18,7 +18,6 @@ import click
 import sgkit as sg
 
 
-
 # yuck - but simplest way to avoid changing directory structure
 sys.path.insert(0, "src")
 from zarr_afdist import zarr_afdist, classify_genotypes_subset_filter, zarr_decode
@@ -480,7 +479,6 @@ def file_size(src, output, debug):
         click.echo(f"{ts_path} n={ts.num_samples // 2}, m={ts.num_sites}")
         bcf_path = ts_path.with_suffix(".bcf")
         vcf_path = ts_path.with_suffix(".vcf.gz")
-        # FIXME
         zarr_path = ts_path.with_suffix(".zarr")
         sav_path = ts_path.with_suffix(".sav")
         genozip_path = ts_path.with_suffix(".genozip")
@@ -496,6 +494,7 @@ def file_size(src, output, debug):
             "vcf": vcf_path,
             "bcf": bcf_path,
             "zarr": zarr_path,
+            "zarr_nshf": ts_path.with_suffix(".noshuffle.zarr"),
             "sav": sav_path,
             "genozip": genozip_path,
         }
@@ -515,16 +514,29 @@ def file_size(src, output, debug):
         print(df)
 
 
+decode_tools = all_tools[:2] + [
+    Tool(
+        "zarr_nshf",
+        ".noshuffle.zarr",
+        run_zarr_afdist,
+        run_zarr_afdist_subset,
+        None,
+        run_zarr_decode,
+        run_zarr_pos_extract,
+    )
+]
+
+
 @click.command()
 @click.argument("src", type=click.Path(), nargs=-1)
 @click.option("-o", "--output", type=click.Path(), default=None)
-@click.option("-t", "--tool", multiple=True, default=[t.name for t in all_tools[:2]])
+@click.option("-t", "--tool", multiple=True, default=[t.name for t in decode_tools])
 @click.option("-s", "--storage", default="hdd")
 @click.option("--debug", is_flag=True)
 def whole_matrix_decode(src, output, tool, storage, debug):
     if len(src) == 0:
         raise ValueError("Need at least one input file!")
-    tool_map = {t.name: t for t in all_tools[:2]}
+    tool_map = {t.name: t for t in decode_tools}
     tools = [tool_map[tool_name] for tool_name in tool]
 
     data = []
@@ -729,6 +741,7 @@ def report_versions():
         print(tool.name)
         print(tool.version_func())
 
+
 @click.command()
[…]

### 2024-04-17 paper.tex

@@
 
-\subsection{The vcfzarr specification}
-The vcfzarr specification is a direct mapping from the VCF data model 
-to a columnar binary format, expressed using Zarr. [A HIGH LEVEL 
-DESCRIPTION OF THE SPEC, FOCUSING ON RATIONALE BEHIND DECISIONS]
-
-The vcfzarr specification can represent anything described by BCF
+\subsection{The VCF Zarr specification}
+The VCF Zarr specification is a direct mapping from the VCF data model
+to a chunked binary array format using Zarr. VCF Zarr takes advantage
+of Zarr's hierarchical structure by representing a VCF file as a top-level
+Zarr group containing Zarr arrays. Each VCF field (fixed fields, INFO fields,
+and FORMAT fields) is represented as a separate array in the Zarr hierarchy.
+Some of the structures from the VCF header are also represented as arrays,
+including contigs, filters, and samples. Making every item in VCF Zarr an
+array means that the data can be accessed via a uniform API.
+
+% Do we need to say that VCF Zarr is an evolution of the Zarr format in scikit-allel?
+
+The specification defines the name, shape, dimension names, and data type
+for each array in the Zarr store. These ``logical'' properties are mandated,
+in contrast to ``physical'' Zarr array properties such as chunk sizes and
+compression, which can be freely chosen by the implementation. This
+separation makes it straightforward for tools and applications to consume
+VCF Zarr data since the data has a well-defined structure, while allowing
+implementations enough room to optimise chunk sizes and compression
+according to the application's needs.
+
+The specification defines a clear mapping of VCF field names (keys) to
+array names, VCF Number to array shape, and VCF Type to array data type.
+To take one example, consider the VCF AD genotype field defined by the
+following VCF header: \texttt{\#\#FORMAT=<ID=AD,Number=A,Type=Integer,Description="Allele Depths">}
+
+The FORMAT key \texttt{ID} maps to an array name of \texttt{call\_AD}
+(FORMAT fields have a \texttt{call\_} prefix, while INFO fields have a
+\texttt{variant\_} prefix; both are followed by the key name). Arrays
+corresponding to FORMAT fields are 3-dimensional with shapes that look
+like \texttt{(variants, samples, <Number>)} in general. In this case, the
+Number A entry indicates that the field has one value per alternate allele,
+which in VCF Zarr is represented as the \texttt{alt\_alleles} dimension name,
+so the shape of this array is \texttt{(variants, samples, alt\_alleles)}.
+The VCF Integer type can be represented as any Zarr integer type, and in
+fact the specification doesn't mandate a particular integer width, so it
+could be encoded as a 16 bit integer for example.
+% Refer to GeL case study section?
+
+VCF Zarr uses special sentinel values to encode missing values to indicate
+that a value is not present, and fill values to pad variable length fields.
+The latter is necessary since Zarr arrays have fixed size dimensions. The
+specification also permits extra boolean arrays to accompany a field array
+to redundantly encode the missing and fill status, which can be useful for
+efficiently masking out values that shouldn't be processed, at the cost of
+a small amount of extra storage.
+
+The VCF Zarr specification can represent anything described by BCF
 (which is somewhat more restrictive than BCF) except for two corner
-cases related to the encoding of missing data. Firstly, vcfzarr does
+cases related to the encoding of missing data. Firstly, VCF Zarr does
 not distinguish between a field that is not present and one that 
-is present but contains missing data. [FOR EXAMPLE...]
-Because of the use of -1 to represent missing data and -2 to 
-represent dimension padding for integers, these values 
-cannot be stored. In practise this doesn't seem to be much of 
+is present but contains missing data. For example, a variant with an
+INFO field \texttt{NS=.} is represented in the same way in VCF Zarr
+as an INFO field with no \texttt{NS} key. Secondly, because of the use
+of sentinel values to represent missing and fill values for integers
+(-1 and -2, respectively), a field containing these original values
+cannot be stored. In practice this doesn't seem to be much of 
 an issue (we have not found a real VCF that contains negative 
-integers). If -1 and -2 need to be stored, a float field
-can be used without issues.
-
-The vcfzarr specification is general and can be mapped to 
+integers). However, if -1 and -2 need to be stored, a float field
+can be used without issues (since the float sentinel is a special NaN
+value).
+
+The VCF Zarr specification is general and can be mapped to 
 file formats such as \toolname{plink}~\citep{purcell2007plink,chang2015second}

### 2024-04-17 src/plot.py
Update numbers for whole-matrix decode
NOTE: big whitespace cleanup as well (oops)

Add updated data on storage, plus no shuffle

@@ -17,6 +17,7 @@ vcf_colour = "tab:pink"
 sav_colour = "tab:red"
 genozip_colour = "tab:purple"
 zarr_colour = "tab:blue"
+zarr_nshf_colour = "tab:cyan"
 
 
 def one_panel_fig(**kwargs):
@@ -32,6 +33,7 @@ def plot_size(ax, df, label_y_offset=None):
         "vcf": vcf_colour,
         "bcf": bcf_colour,
         "zarr": zarr_colour,
+        # "zarr_nshf": zarr_nshf_colour,
         "sav": sav_colour,
         "genozip": genozip_colour,
     }
@@ -51,7 +53,6 @@ def plot_size(ax, df, label_y_offset=None):
         )
         row = dfs.iloc[-1]
         size = row["size"] / GB
-        print(f"{tool} : {size:.2f}")
         ax.annotate(
             f"{size:.0f}G",
             textcoords="offset points",
@@ -59,6 +60,10 @@ def plot_size(ax, df, label_y_offset=None):
             xy=(row.num_samples, row["size"]),
             xycoords="data",
         )
+
+    df_large = df[df.num_samples == 10**6]
+    df_large["size"] /= GB
+    print(df_large)
 
     ax.legend()
     add_number_of_variants(df, ax)
@@ -71,6 +76,7 @@ def plot_total_cpu(ax, df, toolname=None, time_units="h", extrapolate=None):
         "bcftools": bcf_colour,
         "genozip": genozip_colour,
         "zarr": zarr_colour,
+        "zarr_nshf": zarr_nshf_colour,
         "savvy": sav_colour,
     }
     have_genozip = False
@@ -211,13 +217,19 @@ def whole_matrix_decode(time_data, output):
     Plot the figure showing raw decode performance on whole-matrix afdist.
     """
     df = pd.read_csv(time_data, index_col=False).sort_values("num_samples")
+    print(df)
     df = df[df.storage == "hdd"]
 
     fig, ax1 = one_panel_fig()
 
-    plot_total_cpu(ax1, df, time_units="m")
-    df["genotypes_per_second"] = df["total_genotypes"] / df["wall_time"]
-    for tool in ["zarr", "savvy"]:
+    name_map = {
+        "zarr": "Zarr (Zstd + bit shuffle)",
+        "savvy": "Savvy",
+        "zarr_nshf": "Zarr (Zstd)",
+    }
+    plot_total_cpu(ax1, df, toolname=name_map, time_units="m")
[…]

### 2024-04-18 paper.tex
Add data to text

@@
 to be due to Savvy's excellent genotype decoding performance
-(up to 6.5GiB/s vs 1.2GiB/s for Zarr on this dataset;
-Figure~\ref{fig-whole-matrix-decode}). Different choices 
-of compressor and filters (zstandard with bit shuffle, in this case) 
-would likely lead to improved decoding performance in Zarr,
-at the expense of somewhat larger files.
+% zarr 1.2 GiB
+% savvy 6.6 GiB
+% zarr_nshf 3.9 GiB
+(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
+Figure~\ref{fig-whole-matrix-decode}).
+% sequence_length  num_samples  num_sites       tool       size
+% 40         48129895      1000000    7254858       zarr  22.071687
+% 35         48129895      1000000    7254858        bcf  51.749294
+% 36         48129895      1000000    7254858    genozip  10.691384
+% 37         48129895      1000000    7254858        sav  21.249436
+% 38         48129895      1000000    7254858        tsk   1.802636
+% 39         48129895      1000000    7254858        vcf  81.375831
+% 41         48129895      1000000    7254858  zarr_nshf  29.897540
+Turning off the BitShuffle filter for the Zarr dataset,
+however, leads to a substantial increase in decoding speed
+(3.9GiB/s) at the cost of a roughly 25\% increase in storage
+space (29.9GiB up from 22.1GiB for 1 million samples; data not
+shown). Given the relatively small contribution of genotypes to the
+overall storage of real datasets (see the Genomics England example)
+and the frequency that they are likely to be accessed, this
+would seem like a good tradeoff in most cases.
+This ability to easily tune compression performance
+and decoding speed on a field-by-field basis is a major strong
+point of Zarr. The \texttt{vcf2zarr} utility also provides
+functionality to aid with such storage schema tuning.
 

### 2024-04-18 src/plot.py
Update whole matrix compute fig

@@ -61,7 +61,7 @@ def plot_size(ax, df, label_y_offset=None):
             xycoords="data",
         )
 
-    df_large = df[df.num_samples == 10**6]
+    df_large = df[df.num_samples == 10**6].copy()
     df_large["size"] /= GB
     print(df_large)
 
@@ -120,6 +120,10 @@ def plot_total_cpu(ax, df, toolname=None, time_units="h", extrapolate=None):
                 xy=(row.num_samples, total_cpu[-1]),
                 xycoords="data",
             )
+
+    df_large = df[df.num_samples == 10**6].copy()
+    df_large["total_time"] = (df["user_time"] + df["sys_time"]) / 3600
+    print(df_large)
 
     def multiplicative_model(n, a, b):
         # Fit a simple exponential function.
@@ -234,6 +238,7 @@ def whole_matrix_decode(time_data, output):
         print(tool, humanize.naturalsize(max_rate, binary=True))
 
     plt.savefig(output)
+
 
 @click.command()
 @click.argument("time_data", type=click.File("r"))

### 2024-04-18 paper.tex

@@
 short and simple Python program is substantially faster than the 
-equivalent compiled C using \texttt{htslib} APIs on BCF (6 hours
-vs 20 hours for 1 million samples). The fastest method is the 
+equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
+vs 20.6 hours for 1 million samples). 
+% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
+%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
+%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
+%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
+The fastest method is the 
 C++ program written using the Savvy API. This would largely seem

### 2024-04-24 paper.tex
Standardise on VCF Zarr
Closes #66

@@
 \usepackage{siunitx}
-
-% Additional packages used for code formatting
-\usepackage{minted}
-\usepackage{tcolorbox}
-\usepackage{etoolbox}
 
@@
 \textbf{Results:}
-We present the vcfzarr specification, a columnar encoding of the 
+We present the VCF Zarr specification, a columnar encoding of the 
 VCF data model using Zarr which makes retrieving subsets of the 
@@
 using compiled languages. We demonstrate 
-the vcfzarr format (and the vcf2zarr conversion utility) 
+the VCF Zarr format (and the vcf2zarr conversion utility) 
 on the Genomics England dataset of XX,XXX,XXX samples, showing
@@
 processing these files incurs a substantial economic cost.
-The vcfzarr specification, building on widely-used, open-source technologies
+The VCF Zarr specification, building on widely-used, open-source technologies
 has the potential to greatly reduce these costs,
@@
 (Hail, e.g.).
-(4) We present the vcfzarr spec and the vcf2zarr conversion utility.]
+(4) We present the VCF Zarr spec and the vcf2zarr conversion utility.]
 
@@
 [TODO more detail, including brief overview of how data maps from 
-VCF to vcfzarr in the examples]
+VCF to VCF Zarr in the examples]
 \label{fig-data-model}}
@@
 
-We propose the \texttt{vcfzarr} specification which maps the 
+We propose the VCF Zarr specification which maps the 
 VCF data model into a columnar layout using the
@@
 for storing $n$-dimensional arrays of a fixed datatype, stored as 
-compressed chunks. In the \texttt{vcfzarr} specification, 
+compressed chunks. In the VCF Zarr specification, 
 each field in a VCF is mapped to an array, which is compressed 
@@
 such as Parquet [TODO more details and citations?].
-See the Methods for more detail on Zarr and the \texttt{vcfzarr}
+See the Methods for more detail on Zarr and the VCF Zarr
 specification.
@@
 \subsection{Case study: Genomics England 100,000 genomes}
-In this section we demonstrate the utility of \texttt{vcfzarr} on a large human dataset
+In this section we demonstrate the utility of VCF Zarr on a large human dataset
 and the scalability of the \texttt{vcf2zarr} conversion utility.
@@
 \begin{table}
-\caption{Summary for a selection of the largest vcfzarr columns produced for 
+\caption{Summary for a selection of the largest VCF Zarr columns produced for 
 Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
@@
 [Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
-The vcfzarr specification and scalable vcf2zarr conversion utility
+The VCF Zarr specification and scalable vcf2zarr conversion utility
 that we have provided here are the starting points of such 

### 2024-04-24 paper.tex
A pass at the intro

@@
 \section{Background}
-[Four paragraphs. (1) What is VCF, and the good things it has done. 
-(2) Explosion in dataset size. Summary of where we're at in terms of biobanks
-etc. (3) We have a problem. Existing solutions like TileDB etc
-are either proprietary, or tightly coupled to one specific software stack
-(Hail, e.g.).
-(4) We present the VCF Zarr spec and the vcf2zarr conversion utility.]
+Variant Call Format (VCF) is the standard format for interchanging genetic
+variation data~\citep{danecek2011variant}, storing information about 
+DNA sequence polymorphisms among a set of samples with associated 
+quality control metrics and metadata. 
+Originally defined specifically as a text file,
+it has been refined and standardised~\citep{rehm2021ga4gh} and the 
+underlying data-model is now deeply embedded in bioinformatics practice.
+Dataset sizes have grown explosively since the introduction of 
+VCF as part of 1000 Genomes project~\citep{10002015global},
+with Biobank-scale projects such as 
+Genomics England~\cite{turnbull2018100},
+UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
+and the All of Us research program~\citep{all2024genomic}
+collecting genome sequence data for hundreds of thousands of humans.
+While there are many advantages to encoding data in text files,
+VCF is simply not fit for purpose as part of computational workflows 
+at this scale. [A sentence somehow establishing that methods analysing 
+large-scale genomic data are held back by VCF, needing to first 
+convert to intermediate formats like tsinfer.]
+
+The main solution that has emerged to these problems of 
+computation on genetic variation data at scale is Hail~\cite{ganna2016ultra,hail2024}.
+Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
+Hail distributes computations transparently over a cluster, enabling
+interactive analysis of very large datasets. It has been successfully used 
+in many studies, and instrumental in generating datasets such as 
+gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
+While Hail is powerful software, it is not a suitable platform for 
+developing inference or performing complex analysis. It is complex,
+vertically integrated software, with an undocumented file format
+[say this better]. 
+Similarly, commerical solutions that have emerged to facilitate
+the analysis of large-scale genetic variation data are either
+based on proprietary~[Terra? BigQuery?]
+or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
+
+VCF has enabled a vibrant ecosystem of interoperating
+utilities~\cite{danecek2021twelve,garrison2022spectrum}, but these tools
+are fundamentally scale-limited by the format. 
+The most intrinsically limiting aspect of VCF's design is its 
+row-wise layout of data within a file, which means that (for example)
+data for a particular sample cannot be efficiently retrieved.
+The file-oriented paradigm is also unsuited to the realities 
+of modern datasets, which are too large to download and 
+often required to stay in-situ by data-access agreements.
+Large files are stored in cloud environments, where the 
+file systems that are required by file-oriented tools
+are expensively emulated on the basic building blocks
+of object storage. 
+These multiple layers of inefficiences around processing
+VCF data at scale in the cloud mean that it is 
+time-consuming and expensive, effectively making 
+% Dunno if we want to get into this, but it's true
+analysis of these datasets the preserve of a well-funded elite.
+
+Here, we decouple the VCF data model from its row-oriented
+file definition, and show how the corresponding data model can be 
+efficiently stored and analysed in a cloud-native manner.
+We do this by translating VCF data into Zarr format,
+a popular method of storing compressed N-dimensional arrays
+used to store petabyte-scale datasets across the sciences [citations].
+We present the VCF Zarr specification which formalises this 
+mapping, and and the \texttt{vcf2zarr} 
+utility which reliably converts large-scale VCFs into Zarr 
+format. We show that VCF Zarr is much more compact than 
+VCF (and BCF), and it competitive with state-of-the-art
+VCF compression tools. Moreover, we show that Zarr's 
+storage of data in an analysis-ready 
+format~\cite{abernathey2021cloud} greatly facilitates computation,
+with various benchmarks being substantially faster than
+\texttt{bcftools} based pipelines, and again competetive
+with state-of-the-art methods. Finally, we show the 
+utility of VCF Zarr on the Genomics England aggV2 dataset,
+demonstrating that common queries can be performed orders
+of magnitude more quickly based on simple Python scripts.
 
@@
 reader can cross-reference]
+
+Although Zarr is optimised for cloud settings, it also
+works very well in the classical (shared) file-system setting
+and we therefore focus on benchmarks in this setting
+to allow us to compare with existing tools.
+Those seeking benchmarks of Zarr in cloud deployments are 
+directed to [citations].
 

### 2024-04-26 paper.tex
Some work on the intro narrative

@@
 VCF is simply not fit for purpose as part of computational workflows 
-at this scale. [A sentence somehow establishing that methods analysing 
-large-scale genomic data are held back by VCF, needing to first 
-convert to intermediate formats like tsinfer.]
+at this scale. 
+Methods that require efficient access to data
+either require VCF data to be converted into 
+PLINK~\cite{purcell2007plink,chang2015second} 
+or BGEN~\citep{band2018bgen} 
+format~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
+or use 
+custom binary formats~\citep[e.g.][]{
+% Uses custom "bref3" format,
+% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
+browning2018one, 
+% .samples Zarr format
+kelleher2019inferring,
+% Has a "xcftools" package, but it still looks pretty experimental
+hofmeister2023accurate} 
+that support the required access patterns.
+While PLINK and BGEN support efficient access to genotype data, neither
+can accommodate the full flexibility of the VCF data model and conversion
+is lossy. PLINK's approach of storing the genotype matrix in uncompressed
+packed-binary format also leads to large files, particularly for recent
+whole-genome sequence datasets. For example, at two bits 
+per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
+WGS data~\citep{uk2023whole} is 116 TiB.
+% 1,037,556,156 SNPs x  490,640 samples
+% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
+% '115.7 TiB'
 
@@
 
-VCF: Variant Call Format;
+\begin{itemize}
+    \item VCF: Variant Call Format;
+    \item UKB: UK Biobank
+    \item WGS: Whole Genome Sequence
+\end{itemize}
 
 \subsection{Competing Interests}
-
 \subsection{Funding}

### 2024-04-26 paper.tex

@@
 
-%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
-\author[1,\authfn{1},\authfn{2}]{First Author}
-\author[2,\authfn{1},\authfn{2}]{Second Author}
-\author[2]{Third Author}
-\author[2,\authfn{1}]{Fourth Author}
-
-\affil[1]{First Institution}
-\affil[2]{Second Institution}
+%%% Use the \authfn to add symbols for additional footnotes, if any. 
+% 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
+% First author
+\author[1,\authfn{1}]{Eric Czech}
+\author[2,\authfn{1}]{Tim Millar}
+\author[3,\authfn{1}]{Tom White}
+
+% Middle
+\author[4]{Ben Jeffery}
+\author[5]{Alistair Miles}
+\author[6]{Sam Tallman}
+\author[1]{Rafal Wojdyla}
+\author[7]{Shadi Zabad}
+
+% Senior 
+\author[8,\authfn{2}]{Jeff Hammerbacher}
+\author[4,\authfn{2},\authfn{3}]{Jerome Kelleher}
+
+\affil[1]{Related Sciences}
+\affil[2]{Tim's Institute}
+\affil[3]{Tom's Institute}
+\affil[4]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
+University of Oxford, UK}
+\affil[5]{Wellcome Sanger Institute}
+\affil[6]{Genomics England}
+\affil[7]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[8]{Jeff's Institute}
 
 %%% Author Notes
-\authnote{\authfn{1}abc@uni.edu; alphbeta@lab.edu}
-\authnote{\authfn{2}Contributed equally.}
+\authnote{\authfn{1}Joint first author.}
+\authnote{\authfn{2}Joint senior author.}
+\authnote{\authfn{3}jerome.kelleher@bdi.ox.ac.uk}
 
@@
 %%% "Short" author for running page header
-\runningauthor{First et al.}
+\runningauthor{Czech et al.}
 
@@
 \begin{keywords}
-Keyword1; keyword 2; keyword 3 (Three to ten keywords representing the main content of the article)
+Variant Call Format; Zarr; Analysis ready data.
 \end{keywords}
@@
 \begin{itemize}
-\item This is the first point
-\item This is the second point
-\item One last point.
+\item VCF is widely supported, and the underlying data model entrenched 
+in Bioinformatics pipelines.
+\item The standard row-wise encoding as text (or binary) is inherently
+inefficient for large-scale data processing.
+\item The Zarr format provides an efficient solution, by encoding fields
+in the VCF separately in chunk-compressed binary format.
 \end{itemize}

### 2024-04-28 src/plot.py
Added basic skeleton/setup for compression experiments

@@ -1,5 +1,6 @@
 import matplotlib
 import matplotlib.pyplot as plt
+import seaborn as sns
 import humanize
 import pandas as pd
 import click
@@ -301,6 +302,78 @@ def subset_matrix_compute_supplemental(data, output):
     run_subset_matrix_plot(data, output, "n/2", extrapolate=["genozip"])
 
 
+@click.command()
+@click.argument("data", type=click.File("r"))
+@click.argument("output", type=click.Path())
+def plot_compression_ratio_grid(data, output):
+
+    df = pd.read_csv(data)
+    
+    # Filter to only arrays in the same dim order as the original:
+    df_plot = df.loc[df.dim_order.isin(['(0, 1)', '(0, 1, 2)'])]
+
+    # Filter out results with Quantize / PackBits:
+    df_plot = df_plot.loc[df_plot.cname == 'zstd']
+
+    g = sns.catplot(df_plot, kind='bar', 
+                    row='ArrayName', col='shuffle', 
+                    x='named_chunksize', y='CompressionRatio', 
+                    sharex=False, sharey='row')
+    
+    for i, ax in enumerate(g.fig.axes):   ## getting all axes of the fig object
+         ax.set_xticklabels(ax.get_xticklabels(), rotation = 90)
+    
+    plt.subplots_adjust(hspace = 1.)
+    plt.savefig(output)
+
+
+@click.command()
+@click.argument("data", type=click.File("r"))
+@click.argument("output", type=click.Path())
+def plot_compression_dim_shuffle(data, output):
+
+    df = pd.read_csv(data)
+    # Effect was mainly on `call_AD`, so we filter here?
+    df_plot = df.loc[df.ArrayName == 'call_AD']
+
+    g = sns.catplot(df_plot, kind='bar',
+                    col='shuffle',
+                    x='named_chunksize', y='CompressionRatio',
+                    sharex=False, sharey='row')
+
+    for i, ax in enumerate(g.fig.axes):   ## getting all axes of the fig object
+         ax.set_xticklabels(ax.get_xticklabels(), rotation = 90)
+
+    plt.subplots_adjust(hspace = 1.)
+    plt.savefig(output)
+
+
+@click.command()
+@click.argument("data", type=click.File("r"))
+@click.argument("output", type=click.Path())
+def plot_compression_packbits(data, output):
+
+    df = pd.read_csv(data)
[…]

### 2024-04-29 paper.tex
Fixup affils
Closes #91

@@
 % First author
-\author[1,\authfn{1}]{Eric Czech}
-\author[2,\authfn{1}]{Tim Millar}
-\author[3,\authfn{1}]{Tom White}
+\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
+\author[2,3\authfn{1}]{Timothy R. Millar}
+\author[4,\authfn{1}]{Tom White}
 
 % Middle
-\author[4]{Ben Jeffery}
-\author[5]{Alistair Miles}
-\author[6]{Sam Tallman}
+\author[5]{Ben Jeffery}
+\author[6]{Alistair Miles}
+\author[7]{Sam Tallman}
 \author[1]{Rafal Wojdyla}
-\author[7]{Shadi Zabad}
+\author[8]{Shadi Zabad}
 
 % Senior 
-\author[8,\authfn{2}]{Jeff Hammerbacher}
-\author[4,\authfn{2},\authfn{3}]{Jerome Kelleher}
+\author[1,\authfn{2}]{Jeff Hammerbacher}
+\author[5,\authfn{2},\authfn{3}]{Jerome Kelleher}
 
 \affil[1]{Related Sciences}
-\affil[2]{Tim's Institute}
-\affil[3]{Tom's Institute}
-\affil[4]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
+\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
+New Zealand}
+\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
+\affil[4]{Tom's Institute}
+\affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
 University of Oxford, UK}
-\affil[5]{Wellcome Sanger Institute}
-\affil[6]{Genomics England}
-\affil[7]{School of Computer Science, McGill University, Montreal, QC, Canada}
-\affil[8]{Jeff's Institute}
+\affil[6]{Wellcome Sanger Institute}
+\affil[7]{Genomics England}
+\affil[8]{School of Computer Science, McGill University, Montreal, QC, Canada}
 

### 2024-04-29 paper.tex
Add orcids for most uathors
Add GRG ref
Closes #90

@@
 \author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
-\author[2,3\authfn{1}]{Timothy R. Millar}
+\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
 \author[4,\authfn{1}]{Tom White}
@@
 % Middle
-\author[5]{Ben Jeffery}
-\author[6]{Alistair Miles}
-\author[7]{Sam Tallman}
-\author[1]{Rafal Wojdyla}
-\author[8]{Shadi Zabad}
+\author[5]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[6]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[7]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
+\author[8]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
 % Senior 
-\author[1,\authfn{2}]{Jeff Hammerbacher}
-\author[5,\authfn{2},\authfn{3}]{Jerome Kelleher}
+\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
+\author[5,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 deorowicz2019gtshark, %GTShark
-deorowicz2021vcfshark}, % VCFShark
+deorowicz2021vcfshark, % VCFShark
+dehaas2024genotype} %  GRG
 most notably the Positional

### 2024-04-30 paper.tex
Some more Zarr refs

@@
 
-% tentative
-\title{Columnar VCF for Biobank scale datasets using Zarr}
+% See https://github.com/sgkit-dev/vcf-zarr-publication/issues/87
+% for discussion
+\title{Analysis-ready VCF at Biobank scale using Zarr}
 
@@
 stores are supported including cloud buckets and NoSQL databases. 
+In principle, Zarr can store data into any key-value store.
 Metadata describing the array and its properties is then stored 
@@
 different languages.
+
+% Good points from  the new Horizons technical paper:
+
+% From a technical standpoint, a standout feature of Zarr is its simplicity.
+% The format is easy for developers to
+% understand because the internals are exposed transparently. Consequently, Zarr
+% is considered easy to “hack,”
+% i.e. to adapt to new use cases and scenarios, such as implementing a reader
+% from a new programming language
+% or creating a new backend storage layer. (This is a significant contrast from
+% NetCDF4 / HDF5, whose internal file
+% format is relatively opaque, and for which all access must go through the
+% official C library.)
+
+% The Zarr project is governed by a steering council of contributors from
+% different institutions. No single company or institution controls the
+% specification and its implementations.
+% Second, although Zarr is relatively new, it is currently being evaluated by the
+% Open Geospatial Consortium (
+
+The Open GeoSpatial consortium has formally recognised Zarr as a community
+standard~\cite{ogc_zarr2_standard}
+and has formed
+a new GeoZarr Standards Working Group to establish a Zarr encoding for
+geospatial data~\cite{ogc_geozarr_news}.
+
+Zarr is being considered a potential format for migrating the 
+Copernicus global land service to the cloud~\citep{abernathey2021opening}
 
@@
 
+Mature zarr implementations: Zarr-Python~\citep{zarrpython},
+TensorStore~\citep{tensorstore}
+
+and more experimental~\citep[e.g.][]{xtensor_zarr}
+
 Zarr is now widely used.
@@
 Petabyte scale geoscience datasets in the Pangeo project
+
+CMIP6 dataset in Google Cloud~\cite{cmip6_dataset}
 

### 2024-04-30 paper.tex

--- 2024-04-29 paper.tex
@@
 of Zarr and the techologies that it depends upon. Zarr is a specialised format
-for storing large-scale $n$-dimensional numerical data (arrays). Arrays
+for storing large-scale $n$-dimensional data (arrays). Arrays
 are split into chunks, which are compressed and stored separately. Chunks are 
@@
 are often stored in individual files, but a wide array of different
-stores are supported including cloud buckets and NoSQL databases. 
+stores are supported including cloud buckets and NoSQL databases;
+in principle, Zarr can store data in any key-value store.
 Metadata describing the array and its properties is then stored 
-in JSON format along with the chunks. Zarr is best seen as a 
-\emph{specification} that describes the details of how chunks are 
-addressed and stored: there are multiple implementations across
-different languages.
+in JSON format along with the chunks. The simplicity and transparency
+of this design has substantial advantages over other technologies
+such as HDF5~\citep{folk2011overview} which are relatively complex and opaque.
+This simplicity has led to numerous implementations of the Zarr specification
+being developed, ranging from the mature Zarr-Python~\citep{zarrpython}
+and TensorStore~\citep{tensorstore} implementations
+to more experimental extensions to packages like
+GDAL~\citep{gdal_zarr},
+NetCDF~\citep{netcfd_c},
+N5~\citep{n5zarr}
+and xtensor~\citep{xtensor_zarr}
+as well as  standalone libraries for JavaScript~\cite{zarrjs},
+Julia~\cite{zarrjl}, Rust~\citep{zarrs}
+and R~\cite{pizzarr}.
 
@@
 and others.
-
 Blosc is a ``meta-compressor'' because it provides different 
@@
 
-% Zarr was originally developed as a simpler version of 
-% HDF5~\citep{folk2011overview}
-
-Zarr is now widely used.
+This combination of Zarr's simple and cloud-friendly storage 
+of data chunks with state-of-the-art compression methods has 
+led to Zarr gaining significant traction in large-scale 
+scientific computing, in particular in the geosciences.
 
 Multi terabyte numerical weather prediction models~\citep{gowan2022using}
-Petabyte scale geoscience datasets in the Pangeo project
+
+CMIP6 dataset in Google Cloud~\cite{cmip6_dataset}
+
+The Open GeoSpatial consortium has formally recognised Zarr as a community
+standard~\cite{ogc_zarr2_standard}
+and has formed
+a new GeoZarr Standards Working Group to establish a Zarr encoding for
+geospatial data~\cite{ogc_geozarr_news}.
+
+Zarr is being considered a potential format for migrating the 
+Copernicus global land service to the cloud~\citep{abernathey2021opening}
 

### 2024-04-30 paper.tex

@@
 Zarr is flexible in allowing different compression codecs and 
-pre-compression filters to be specified on a per-array basis.
+pre-compression filters to be specified on a per-array basis
+(see section XXX).
 Two key technologies used by Zarr are the Blosc
@@
 
-This combination of Zarr's simple and cloud-friendly storage 
+Scientific datasets are increasingly overwhelming the classical
+model of downloading and analysing locally, and are migrating to 
+centralised cloud repositories \citep{abernathey2021cloud}.
+The combination of Zarr's simple and cloud-friendly storage 
 of data chunks with state-of-the-art compression methods has 
-led to Zarr gaining significant traction in large-scale 
-scientific computing, in particular in the geosciences.
-
-Multi terabyte numerical weather prediction models~\citep{gowan2022using}
-
-CMIP6 dataset in Google Cloud~\cite{cmip6_dataset}
-
+led to Zarr gaining significant traction in these settings.
+Multiple petabyte-scale datasets are now stored using 
+Zarr~\cite[e.g.][]{gowan2022using, % weather model data
+fahnestock2023mappin, % ITS_LIVE dataset
+cmip6_dataset}
+or under active consideration for migration~\citep{abernathey2021opening}.
 The Open GeoSpatial consortium has formally recognised Zarr as a community
@@
 geospatial data~\cite{ogc_geozarr_news}.
-
-Zarr is being considered a potential format for migrating the 
-Copernicus global land service to the cloud~\citep{abernathey2021opening}
-
-ITS\_LIVE ice velocity dataset~\citep{fahnestock2023mappin} using 
-the Zarr.js Javascript interface to visualise the ``massive''
-dataset [how big is it?]
 

### 2024-05-01 paper.tex
Some single-cell refs and narrative
Closes #67

@@
 Those seeking benchmarks of Zarr in cloud deployments are 
-directed to [citations].
+directed to~\citep{durbin2020task,moore2021ome,gowan2022using} [ANY OTHERS?]
 
@@
 
-Reel off Python data science ecosystem: Pandas~\citep{mckinney2010data} etc
-[ Things to cite: pandas~\citep{mckinney2010data},
-BioNumpy~\citep{rand2022bionumpy}, OME-Zarr~\citep{moore2023ome}]
+The Python data science ecosystem is increasingly popular, 
+with packages such as NumPy~\citep{harris2020array},
+SciPy~\citep{virtanen2020scipy},
+Pandas~\citep{mckinney2010data},
+Jupyter~\citep{kluyver2016jupyter}
+enabling much of contemporary science.
+
+BioNumpy~\citep{rand2022bionumpy},
 
@@
 model of downloading and analysing locally, and are migrating to 
-centralised cloud repositories \citep{abernathey2021cloud}.
+centralised cloud repositories \citep{abernathey2021cloud,moore2021ome}.
 The combination of Zarr's simple and cloud-friendly storage 
@@
 cmip6_dataset}
-or under active consideration for migration~\citep{abernathey2021opening}.
+or under active consideration for migration~\citep{durbin2020task,abernathey2021opening}.
 The Open GeoSpatial consortium has formally recognised Zarr as a community
@@
 
-Zarr has recently been gaining traction in biological applications
-also. EUGENe~\citep{klie2023predictive} is a toolkit for 
+Zarr has recently been gaining popularity in biological applications.
+The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
+of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
+OME-Zarr already has a rich suite of supporting
+tools~\cite{moore2023ome,rzepka2023toward}. 
+Zarr has also seen recent uptake 
+in single-scell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
+and multimodal spatial omics
+data~\citep{marconato2024spatialdata,baker2023emobject}.
+
+EUGENe~\citep{klie2023predictive} is a toolkit for 
 for applying deep-learning methods to genomic sequence data,
 integrating with PyTorch [citation?].
-
-Spatially resolved omics data~\citep{baker2023emobject,marconato2024spatialdata}
-building on the Zarr-based OME-NGFF format~\citep{moore2023ome,rzepka2023toward}
 
@@
 browser, using Zarr as the storage backend.
+
+Zarr was originally developed as a means efficiently accessing VCF data,
+ and has been in use within the MalariaGEN project for XX years. [
+ALISTAIR - details and citations here!]
 

### 2024-05-01 paper.tex

@@
 than \texttt{memcpy}.
+% Mentioning this detail to reassure readers who might be thinking
+% "how would this be implemented and what kind of horrible dependencies
+% would I need?"
 Blosc is written in C, with APIs for C, Python, Julia, Rust
 and others.
-Blosc is a ``meta-compressor'' because it provides different 
+Blosc is a ``meta-compressor'' because it provides 
 access to several different compression codecs. The 
@@
 data~\citep{marconato2024spatialdata,baker2023emobject}.
-
-EUGENe~\citep{klie2023predictive} is a toolkit for 
-for applying deep-learning methods to genomic sequence data,
-integrating with PyTorch [citation?].
-
-DivBrowse~\citep{konig2023divbrowse}. Works with VCF data using 
-the scikit-allel \texttt{vcf\_to\_zarr} function, (produces 
-a precursor to the VCF-Zarr specification here). It's a 
-browser, using Zarr as the storage backend.
-
-Zarr was originally developed as a means efficiently accessing VCF data,
- and has been in use within the MalariaGEN project for XX years. [
-ALISTAIR - details and citations here!]
+Recent additions using Zarr include the application of deep learning
+models to genomic sequence data~\citep{klie2023predictive} 
+and a browser for genetic variation data~\citep{konig2023divbrowse}.
+Zarr has in fact been in use in genomics for many years, and was originally 
+developed to [ALISTAIR - details and citations here!]
 
@@
 following VCF header: \texttt{\#\#FORMAT=<ID=AD,Number=A,Type=Integer,Description="Allele Depths">}
-
 The FORMAT key \texttt{ID} maps to an array name of \texttt{call\_AD}

### 2024-05-02 paper.tex
Another pass at intro narrative

@@
 It provides a well-defined data model and is central to a large ecosystem
-of tools interoperating through this format.
-The standard row-based encoding of the VCF data model (either as text
+of interoperating tools.
+The standard row-wise encoding of the VCF data model (either as text
 or packed binary) emphasises efficient retrieval of all data for a given
@@
 Biobank-scale datasets currently available 
-consist of hundreds of thousands of whole genomes with tens of terabytes
-of compressed VCF per chromosome.
+consist of hundreds of thousands of whole genomes 
+and hundreds of terabytes of compressed VCF.
 Row-based data storage is fundamentally unsuited to such datasets
@@
 collecting genome sequence data for hundreds of thousands of humans.
-While there are many advantages to encoding data in text files,
-VCF is simply not fit for purpose as part of computational workflows 
-at this scale. 
-Methods that require efficient access to data
-either require VCF data to be converted into 
+
+VCF's simple text-based design and widespread support makes it an 
+excellent archival format, but it is a poor basis for computational
+workflows. Methods that require efficient access to genotype data
+either require convertion to the
 PLINK~\cite{purcell2007plink,chang2015second} 
 or BGEN~\citep{band2018bgen} 
-format~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
+formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
 or use 
-custom binary formats~\citep[e.g.][]{
+bespoke binary formats~\citep[e.g.][]{
 % Uses custom "bref3" format,
@@
 
-The main solution that has emerged to these problems of 
-computation on genetic variation data at scale is Hail~\cite{ganna2016ultra,hail2024}.
+Downstream processing of these datasets can be split into a 
+few broad categories. The first level of analysis that is 
+done is quality control (QC). Variant calling QC is an 
+involved and multi-faceted 
+task [CITATIONS?], requiring analysis of many different metrics
+and properties of the data and substantial computation.
+Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
+for quality control of large-scale variation datasets.
 Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
-Hail distributes computations transparently over a cluster, enabling
+Hail distributes calculations and data transparently over a cluster, enabling
 interactive analysis of very large datasets. It has been successfully used 
-in many studies, and instrumental in generating datasets such as 
+in many studies, and instrumental in datasets such as 
 gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
+After QC, a common task is to perform a genome wide association study
+(GWAS)~\cite{uffelmann2021genome}. The majority of tools for 
+performing GWAS and associated tasks require
+data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
+and so data must be ``hard-called'' according to some QC criteria
+and exported to additional (large) files.
+Many applications examine data on a variant-by-variant
+basis, but others,
+e.g.\ reference-panel based phasing and imputation 
+methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
+and Ancestral Recombination Graph 
+(ARG) inference~\cite{brandt2024promise,lewanski2024era,wong2024general}
+require access on a sample-by-sample basis.
+Finally, variation datasets are often queried in exploratory
+analyses, to find regions or variants of interest for a particular 
+study (see section XXX).
+
 While Hail is powerful software, it is not a suitable platform for 

### 2024-05-03 paper.tex
Pass through intro

@@
 % for discussion
-\title{Analysis-ready VCF at Biobank scale using Zarr}
+\title{Analysis-ready VCF at Biobank-scale using Zarr}
 
@@
 data much more efficient. Zarr is a cloud-native format for storing 
-multi-dimensional numerical data, widely used in scientific computing.
-We show how this format is competitive with specialised methods for 
-storing genotype data in terms of compression ratios, and also performs
-well on compressing the high-entropy quality-control fields that 
-tend to dominate. We show that performing calculations on the genotype
-matrix is much more efficient than standard
-approaches based on VCF, and is comparable with state-of-the-art methods
-using compiled languages. We demonstrate 
-the VCF Zarr format (and the vcf2zarr conversion utility) 
+multi-dimensional data, widely used in scientific computing.
+We show how this format is far more efficient than
+standard VCF based approaches,
+and competitive with specialised methods for 
+storing genotype data in terms of compression ratios
+and calculation performance.
+We demonstrate the VCF Zarr format (and the vcf2zarr conversion utility) 
 on the Genomics England dataset of XX,XXX,XXX samples, showing
@@
 Large row-encoded VCF files are a major bottleneck for current research, and 
-processing these files incurs a substantial economic cost.
+storing and processing these files incurs a substantial economic cost.
 The VCF Zarr specification, building on widely-used, open-source technologies
@@
 and may enable a diverse ecosystem of next-generation tools for analysing 
-genetic variation data.
+genetic variation data directly from cloud-based object stores.
 \end{abstract}
@@
 VCF as part of 1000 Genomes project~\citep{10002015global},
-with Biobank-scale projects such as 
+with Biobank-scale initiatives such as 
 Genomics England~\cite{turnbull2018100},
@@
 excellent archival format, but it is a poor basis for computational
-workflows. Methods that require efficient access to genotype data
-either require convertion to the
-PLINK~\cite{purcell2007plink,chang2015second} 
-or BGEN~\citep{band2018bgen} 
-formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
-or use 
-bespoke binary formats~\citep[e.g.][]{
-% Uses custom "bref3" format,
-% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
-browning2018one, 
-% .samples Zarr format
-kelleher2019inferring,
-% Has a "xcftools" package, but it still looks pretty experimental
-hofmeister2023accurate} 
-that support the required access patterns.
-While PLINK and BGEN support efficient access to genotype data, neither
-can accommodate the full flexibility of the VCF data model and conversion
-is lossy. PLINK's approach of storing the genotype matrix in uncompressed
-packed-binary format also leads to large files, particularly for recent
-whole-genome sequence datasets. For example, at two bits 
-per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
-WGS data~\citep{uk2023whole} is 116 TiB.
-% 1,037,556,156 SNPs x  490,640 samples
-% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
-% '115.7 TiB'
-
-Downstream processing of these datasets can be split into a 
+workflows. 
+% Methods that require efficient access to genotype data
+% either require convertion to the
+% PLINK~\cite{purcell2007plink,chang2015second} 
+% or BGEN~\citep{band2018bgen} 
+% formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
+% or use 
+% bespoke binary formats~\citep[e.g.][]{
+% % Uses custom "bref3" format,
+% % https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
+% browning2018one, 
+% % .samples Zarr format
+% kelleher2019inferring,
+% % Has a "xcftools" package, but it still looks pretty experimental
+% hofmeister2023accurate} 
+% that support the required access patterns.
+% While PLINK and BGEN support efficient access to genotype data, neither
+% can accommodate the full flexibility of the VCF data model and conversion
+% is lossy.
+Processing of Biobank-scale datasets can be split into a 
 few broad categories. The first level of analysis that is 
@@
 task [CITATIONS?], requiring analysis of many different metrics
-and properties of the data and substantial computation.
+and incurring substantial computation.
 Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
@@
 Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
-Hail distributes calculations and data transparently over a cluster, enabling
+Hail distributes data and calculations over a cluster, enabling
 interactive analysis of very large datasets. It has been successfully used 
@@
 and so data must be ``hard-called'' according to some QC criteria
-and exported to additional (large) files.
+and exported to additional files.
+PLINK's approach of storing the genotype matrix in uncompressed
+packed-binary format provides efficient access to genotype 
+data, but file sizes are substantially larger than 
+the equivalent compressed VCF. 
+For example, at two bits 
+per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
+WGS data~\citep{uk2023whole} is 116 TiB.
+% 1,037,556,156 SNPs x  490,640 samples
+% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
+% '115.7 TiB'
 Many applications examine data on a variant-by-variant
@@
 methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
-and Ancestral Recombination Graph 
-(ARG) inference~\cite{brandt2024promise,lewanski2024era,wong2024general}
+% and Ancestral Recombination Graph 
+% (ARG) inference~\cite{brandt2024promise,lewanski2024era,wong2024general}
 require access on a sample-by-sample basis.
 Finally, variation datasets are often queried in exploratory
-analyses, to find regions or variants of interest for a particular 
+analyses, to find regions or samples of interest for a particular 
 study (see section XXX).
 
-While Hail is powerful software, it is not a suitable platform for 
-developing inference or performing complex analysis. It is complex,
-vertically integrated software, with an undocumented file format
-[say this better]. 
-Similarly, commerical solutions that have emerged to facilitate
-the analysis of large-scale genetic variation data are either
-based on proprietary~[Terra? BigQuery?]
-or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
-
-VCF has enabled a vibrant ecosystem of interoperating
-utilities~\cite{danecek2021twelve,garrison2022spectrum}, but these tools
-are fundamentally scale-limited by the format. 
+VCF (and BCF) cannot support any of these workflows efficiently
+at the Biobank scale, and workarounds of creating subsets 
+of data (i.e., discarding QC metrics in VCF) or copying to 
+PLINK format are becoming less effective.
 The most intrinsically limiting aspect of VCF's design is its 
@@
 Large files are stored in cloud environments, where the 
-file systems that are required by file-oriented tools
+file systems that are required by classical file-oriented tools
 are expensively emulated on the basic building blocks
@@
 
+The analysis of Biobank-scale variation 
+datasets requires a primary data representation that can 
+efficiently support different types of query without 
+conversion or intermediate copies, and operate 
+directly from cloud object stores. Such a representation
+can be termed ``analysis-ready''
+and ``cloud-native''~\citep{abernathey2021cloud}.
+For such a representation to be FAIR~\citep{wilkinson2016fair},
+it must also \emph{accessible}: ``open, free, and universally
+implementable''.
+There is no currently no efficient, FAIR representation of genetic variation
+data suitable for cloud deployments.
+While Hail is powerful software its file formats are not documented or 
+designed for reuse.
+Similarly, commerical solutions that have emerged to facilitate
+the analysis of large-scale genetic variation data are either
+based on proprietary~[Terra? BigQuery?]
+or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
+The next generation of VCF analysis methods requires
+an open, free and transparent data representation 
+with multiple independent implementations.
+
 Here, we decouple the VCF data model from its row-oriented
-file definition, and show how the corresponding data model can be 
-efficiently stored and analysed in a cloud-native manner.
+file definition, and show how the data can be 
+efficiently stored and analysed in a cloud-native, FAIR manner.
 We do this by translating VCF data into Zarr format,
-a popular method of storing compressed N-dimensional arrays
-used to store petabyte-scale datasets across the sciences [citations].
+a method of storing large multidimensional data as a regular
+grid of compressed chunks. 
+Zarr's simplicity and its direct mapping of chunks to 
+keys in an object store have lead to 
+it gaining substantial traction
+across the sciences, and it is now used in multiple petabyte-scale
+datasets in cloud deployments (see Methods for details).
 We present the VCF Zarr specification which formalises this 
-mapping, and and the \texttt{vcf2zarr} 
+mapping, and the \texttt{vcf2zarr} 
 utility which reliably converts large-scale VCFs into Zarr 
 format. We show that VCF Zarr is much more compact than 
-VCF (and BCF), and it competitive with state-of-the-art
-VCF compression tools. Moreover, we show that Zarr's 
-storage of data in an analysis-ready 
-format~\cite{abernathey2021cloud} greatly facilitates computation,
+VCF (and BCF), and is competitive with state-of-the-art
+VCF compression tools. 
+Moreover, we show that Zarr's storage of data in an analysis-ready 
+format greatly facilitates computation,
 with various benchmarks being substantially faster than
 \texttt{bcftools} based pipelines, and again competetive
-with state-of-the-art methods. Finally, we show the 
+with state-of-the-art file-oriented methods. Finally, we show the 
 utility of VCF Zarr on the Genomics England aggV2 dataset,
-demonstrating that common queries can be performed orders
-of magnitude more quickly based on simple Python scripts.
+demonstrating that common \texttt{bcftools} queries can be performed orders
+of magnitude more quickly using simple Python scripts.
 

### 2024-05-03 paper.tex

@@
 employed in genetic data compression.
+[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
+in here.]
 
@@
 describes observations
-and metadata for a single variant (usually a position on the genome
-where variation occurs in the samples)~\citep{danecek2011variant}. 
+and metadata for a single variant~\citep{danecek2011variant}. 
 VCF is therefore a \emph{row-wise} format, i.e., where all of the 
-data for a record is stored continguously. BCF is similarly 
+data for a record is stored continguously. BCF (a packed binary 
+representation) is similarly 
 row-wise, as are the majority of proposed alternative storage formats.
@@
 are almost always compressed in blocks, accessing any information
-for all rows means retrieving and decompressing \emph{all} 
-information from all rows.
+for a set of rows means retrieving and decompressing \emph{all} 
+information from these rows.
 The alternative to row-wise storage is \emph{columnar} storage:
@@
 we group together all the records for a given field.
-Columnar storage has become standard [very quick summary of columnar
-storage with some refs].
-Although columnar methods have been successfully used in alignment 
+Columnar storage formats such as Parquet~\citep{parquet2024},
+makes retrieving particular columns much 
+more efficient and can lead to substantially better compression.
+While columnar techniques have been successfully applied 
+alignment
 % TODO check
-storage~\citep{bonfield2014scramble,bonfield2022cram},
-the only existing method using columnar 
-storage for genetic variation data is 
-SeqArray~\citep{zheng2017seqarray,zheng2012high} to our knowledge.
+% Jeff: OK with the ADAM reference in here? You did some parquet based
+% alignment storage, right?
+storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
+use of technologies such as Parquet for 
+storing and analysing variation data have had limited
+success~\citep{boufea2017managing,fan2020variant}.
+Directly mapping VCF data to a columnar layout, in which there is a 
+a column for the genotypes (and other per-sample QC metrics) 
+for each sample, scales poorly. Firstly, it leads to a large number 
+of columns (millions, at Biobank scale), which techologies such 
+as Parquet are not designed for. Secondly, compression performance 
+within columns is poor because there is much less similarity among
+the genotypes for a given sample than there is among the genotypes
+for a given variant. [TODO check this - reference?]
 
@@
 \end{figure}
+
+[TODO intro passage here on array-based or ``tensor'' storage. Separate
+out the different interleaved matrices with the VCF, and store them
+in regular chunked grids.]
 

### 2024-05-07 paper.tex
Nother intro pass

@@
 Variant Call Format (VCF) is the standard format for interchanging genetic
-variation data~\citep{danecek2011variant}, storing information about 
+variation data~\citep{danecek2011variant}, encoding information about 
 DNA sequence polymorphisms among a set of samples with associated 
@@
 workflows. 
-% Methods that require efficient access to genotype data
-% either require convertion to the
-% PLINK~\cite{purcell2007plink,chang2015second} 
-% or BGEN~\citep{band2018bgen} 
-% formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
-% or use 
-% bespoke binary formats~\citep[e.g.][]{
-% % Uses custom "bref3" format,
-% % https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
-% browning2018one, 
-% % .samples Zarr format
-% kelleher2019inferring,
-% % Has a "xcftools" package, but it still looks pretty experimental
-% hofmeister2023accurate} 
-% that support the required access patterns.
-% While PLINK and BGEN support efficient access to genotype data, neither
-% can accommodate the full flexibility of the VCF data model and conversion
-% is lossy.
+Methods that require efficient access to genotype data
+either require convertion to the
+PLINK~\cite{purcell2007plink,chang2015second} 
+or BGEN~\citep{band2018bgen} 
+formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
+or use 
+bespoke binary formats~\citep[e.g.][]{
+% Uses custom "bref3" format,
+% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
+browning2018one, 
+% .samples Zarr format
+kelleher2019inferring,
+% Has a "xcftools" package, but it still looks pretty experimental
+hofmeister2023accurate} 
+that support the required access patterns.
+While PLINK and BGEN support efficient access to genotype data, neither
+can accommodate the full flexibility of the VCF data model and conversion
+is lossy.
+PLINK's approach of storing the genotype matrix in uncompressed
+packed-binary format provides efficient access to genotype 
+data, but file sizes are substantially larger than 
+the equivalent compressed VCF. 
+For example, at two bits 
+per genotype, the full genotype matrix for the GraphTyper SNP dataset 
+in the UKB WGS data~\citep{uk2023whole} is 116 TiB.
+% 1,037,556,156 SNPs x  490,640 samples
+% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
+% '115.7 TiB'
+
 Processing of Biobank-scale datasets can be split into a 
-few broad categories. The first level of analysis that is 
-done is quality control (QC). Variant calling QC is an 
+few broad categories. The most basic analysis 
+is quality control (QC). Variant calling QC is an 
 involved and multi-faceted 
@@
 gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
-After QC, a common task is to perform a genome wide association study
+A common task to perform after QC is 
+a genome wide association study
 (GWAS)~\cite{uffelmann2021genome}. The majority of tools for 
-performing GWAS and associated tasks require
+performing GWAS and related analyses require
 data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
@@
 and exported to additional files.
-PLINK's approach of storing the genotype matrix in uncompressed
-packed-binary format provides efficient access to genotype 
-data, but file sizes are substantially larger than 
-the equivalent compressed VCF. 
-For example, at two bits 
-per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
-WGS data~\citep{uk2023whole} is 116 TiB.
-% 1,037,556,156 SNPs x  490,640 samples
-% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
-% '115.7 TiB'
-Many applications examine data on a variant-by-variant
-basis, but others,
-e.g.\ reference-panel based phasing and imputation 
-methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
-% and Ancestral Recombination Graph 
-% (ARG) inference~\cite{brandt2024promise,lewanski2024era,wong2024general}
-require access on a sample-by-sample basis.
+Genotype calls are often further refined via statistical methods,
+% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
+for example by phasing~[CITES] and imputation[CITES], usually 
+creating an additional copy of the dataset.
 Finally, variation datasets are often queried in exploratory
 analyses, to find regions or samples of interest for a particular 
-study (see section XXX).
-
-VCF (and BCF) cannot support any of these workflows efficiently
-at the Biobank scale, and workarounds of creating subsets 
+study (see section XXX for an example).
+
+VCF cannot support any of these workflows efficiently
+at the Biobank scale, and workarounds of copying subsets 
 of data (i.e., discarding QC metrics in VCF) or copying to 
-PLINK format are becoming less effective.
-The most intrinsically limiting aspect of VCF's design is its 
-row-wise layout of data within a file, which means that (for example)
-data for a particular sample cannot be efficiently retrieved.
+PLINK format have reached their limits.
+The most intrinsically limiting aspect of VCF's design 
+is its row-wise layout of data, which means that (for example)
+information for a particular sample or field cannot be obtained without
+retrieving the entire dataset.
 The file-oriented paradigm is also unsuited to the realities 
@@
 VCF data at scale in the cloud mean that it is 
-time-consuming and expensive, effectively making 
-% Dunno if we want to get into this, but it's true
+time-consuming and expensive, 
+% Can easily delete this last bit:
+effectively making 
 analysis of these datasets the preserve of a well-funded elite.
 
-The analysis of Biobank-scale variation 
-datasets requires a primary data representation that can 
-efficiently support different types of query without 
+To fully utilise these vast variation datasets,
+we require a primary data representation that can 
+support different types of query without 
 conversion or intermediate copies, and operate 
@@
 and ``cloud-native''~\citep{abernathey2021cloud}.
-For such a representation to be FAIR~\citep{wilkinson2016fair},
-it must also \emph{accessible}: ``open, free, and universally
-implementable''.
+For the representation to be FAIR~\citep{wilkinson2016fair},
+it must also \emph{accessible} using protocols that are 
+``open, free, and universally implementable''.
 There is no currently no efficient, FAIR representation of genetic variation
@@
 
-Here, we decouple the VCF data model from its row-oriented
+In this article, we decouple the VCF data model from its row-oriented
 file definition, and show how the data can be 
@@
 grid of compressed chunks. 
-Zarr's simplicity and its direct mapping of chunks to 
+Zarr's elegant simplicity and direct mapping of chunks to 
 keys in an object store have lead to 
@@
 format. We show that VCF Zarr is much more compact than 
-VCF (and BCF), and is competitive with state-of-the-art
+VCF and is competitive with state-of-the-art
 VCF compression tools. 

### 2024-05-07 paper.tex
More updates

@@
 
-VCF is a text-based format in which each line
+VCF is row-wise format in which 
 describes observations
-and metadata for a single variant~\citep{danecek2011variant}. 
-VCF is therefore a \emph{row-wise} format, i.e., where all of the 
-data for a record is stored continguously. BCF (a packed binary 
-representation) is similarly 
+and metadata for a single variant~\citep{danecek2011variant} are
+encoded as a line of text.
+BCF (a packed binary representation) is similarly 
 row-wise, as are the majority of proposed alternative storage formats.
@@
 allows data to be easily streamed through Unix pipelines (see Discussion).
-
-Row-wise storage works well when records are relatively small, 
+The row-wise approach works well when records are relatively small, 
 or when we typically want to analyse each record in its entirity.
@@
 information from these rows.
-The alternative to row-wise storage is \emph{columnar} storage:
+
+The usual alternative to row-wise storage is \emph{columnar} storage:
 instead of grouping together all the fields for a record,
@@
 Columnar storage formats such as Parquet~\citep{parquet2024},
-makes retrieving particular columns much 
+make retrieving particular columns much 
 more efficient and can lead to substantially better compression.
 While columnar techniques have been successfully applied 
-alignment
+in alignment
 % TODO check
@@
 storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
-use of technologies such as Parquet for 
+the use of technologies such as Parquet for 
 storing and analysing variation data have had limited
 success~\citep{boufea2017managing,fan2020variant}.
-Directly mapping VCF data to a columnar layout, in which there is a 
+Columnar techologies tend to be 
+optimised for ``tall and narrow'' data there is a large number of records
+for a relatively small number of columns. 
+Mapping VCF directly to a columnar layout, in which there is a 
 a column for the genotypes (and other per-sample QC metrics) 
-for each sample, scales poorly. Firstly, it leads to a large number 
-of columns (millions, at Biobank scale), which techologies such 
-as Parquet are not designed for. Secondly, compression performance 
-within columns is poor because there is much less similarity among
+leads to millions of columns, and substantial technical challenges.
+Additionally, compression performance within columns is poor 
+because there is much less similarity among
 the genotypes for a given sample than there is among the genotypes
@@
 \end{figure}
+
+VCF is at its core an encoding of the genotype matrix $G$, where each entry
+encodes the observed genotypes for a given sample at a give variant site,
+interleaved with per-variant information
+and other matrices (e.g., GQ or AD).
+The data is largely numerical and of a fixed dimension, and is therefore a 
+natural mapping to an array-oriented approach [PICK UP HERE]
 

### 2024-05-07 paper.tex
Tweaks

@@
 
-VCF is at its core an encoding of the genotype matrix $G$, where each entry
-encodes the observed genotypes for a given sample at a give variant site,
+VCF is at its core an encoding of the genotype matrix where each entry
+describes the observed genotypes for a given sample at a give variant site,
 interleaved with per-variant information
-and other matrices (e.g., GQ or AD).
-The data is largely numerical and of a fixed dimension, and is therefore a 
-natural mapping to an array-oriented approach [PICK UP HERE]
-
-[TODO intro passage here on array-based or ``tensor'' storage. Separate
-out the different interleaved matrices with the VCF, and store them
-in regular chunked grids.]
-
+and other matrices (e.g., the GQ or AD fields).
+The data is largely numerical and each field is of a fixed dimension, 
+and is therefore a natural mapping to array-oriented 
+or ``tensor'' storage.
 We propose the VCF Zarr specification which maps the 
-VCF data model into a columnar layout using the
+VCF data model into an array-oriented layout using the
 widely-used Zarr format (Fig~\ref{fig-data-model}). 
@@
 variants and samples axis.
-Note that this approach to storing higher dimensional data
-sets Zarr apart from more general columnar formats 
-such as Parquet [TODO more details and citations?].
 See the Methods for more detail on Zarr and the VCF Zarr

### 2024-05-07 src/plot.py
Adding complete pipeline with figures + usability additions.

@@ -324,6 +324,7 @@ def plot_compression_ratio_grid(data, output):
          ax.set_xticklabels(ax.get_xticklabels(), rotation = 90)
     
     plt.subplots_adjust(hspace = 1.)
+    plt.tight_layout()
     plt.savefig(output)
 
 
@@ -333,18 +334,21 @@ def plot_compression_ratio_grid(data, output):
 def plot_compression_dim_shuffle(data, output):
 
     df = pd.read_csv(data)
+
     # Effect was mainly on `call_AD`, so we filter here?
     df_plot = df.loc[df.ArrayName == 'call_AD']
 
     g = sns.catplot(df_plot, kind='bar',
                     col='shuffle',
                     x='named_chunksize', y='CompressionRatio',
-                    sharex=False, sharey='row')
+                    sharex=True, sharey=True)
 
     for i, ax in enumerate(g.fig.axes):   ## getting all axes of the fig object
          ax.set_xticklabels(ax.get_xticklabels(), rotation = 90)
 
-    plt.subplots_adjust(hspace = 1.)
+    plt.suptitle("Compression Ratios for call_AD across dimension shuffles")
+
+    plt.tight_layout()
     plt.savefig(output)
 
 
@@ -354,7 +358,8 @@ def plot_compression_dim_shuffle(data, output):
 def plot_compression_packbits(data, output):
 
     df = pd.read_csv(data)
-    df_plot = df.loc[df.ArrayName == 'call_genotype_mask']
+    df_plot = df.loc[(df.ArrayName == 'call_genotype_mask') & 
+                     df.dim_order.isin(['(0, 1)', '(0, 1, 2)'])]
 
     if len(df_plot) < 1:
         raise ValueError("Information about call_genotype_mask isn't present.")
@@ -362,12 +367,13 @@ def plot_compression_packbits(data, output):
     g = sns.catplot(df_plot, kind='bar',
                     col='shuffle', row='cname',
                     x='named_chunksize', y='CompressionRatio',
-                    sharex=False, sharey='row')
+                    sharex=False, sharey=True)
 
     for i, ax in enumerate(g.fig.axes):   ## getting all axes of the fig object
          ax.set_xticklabels(ax.get_xticklabels(), rotation = 90)
 
     plt.subplots_adjust(hspace = 1.)
+    plt.tight_layout()
     plt.savefig(output)
 
 

### 2024-05-08 paper.tex
another intro pass

@@
 collecting genome sequence data for hundreds of thousands of humans.
-
 VCF's simple text-based design and widespread support makes it an 
-excellent archival format, but it is a poor basis for computational
-workflows. 
+excellent archival format, but it is an inefficient basis for analysis.
 Methods that require efficient access to genotype data
-either require convertion to the
+either require conversion to the
 PLINK~\cite{purcell2007plink,chang2015second} 
@@
 per genotype, the full genotype matrix for the GraphTyper SNP dataset 
-in the UKB WGS data~\citep{uk2023whole} is 116 TiB.
+in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
 % 1,037,556,156 SNPs x  490,640 samples
@@
 few broad categories. The most basic analysis 
-is quality control (QC). Variant calling QC is an 
+is quality control (QC). Variant QC is an 
 involved and multi-faceted 
-task [CITATIONS?], requiring analysis of many different metrics
-and incurring substantial computation.
-Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
-for quality control of large-scale variation datasets.
-Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
-Hail distributes data and calculations over a cluster, enabling
-interactive analysis of very large datasets. It has been successfully used 
-in many studies, and instrumental in datasets such as 
-gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
+task [CITATIONS?], often requiring interactive, exploratory analysis
+and incurring substantial computation over multiple QC fields.
+Genotype calls are often refined via statistical methods,
+% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
+for example by phasing~[CITES] and imputation[CITES], 
+creating additional copies of the dataset.
 A common task to perform after QC is 
-a genome wide association study
-(GWAS)~\cite{uffelmann2021genome}. The majority of tools for 
+a genome wide association study (GWAS)~\cite{uffelmann2021genome}. 
+The majority of tools for 
 performing GWAS and related analyses require
@@
 and so data must be ``hard-called'' according to some QC criteria
-and exported to additional files.
-Genotype calls are often further refined via statistical methods,
-% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
-for example by phasing~[CITES] and imputation[CITES], usually 
-creating an additional copy of the dataset.
+and exported to additional copies.
 Finally, variation datasets are often queried in exploratory
@@
 VCF data at scale in the cloud mean that it is 
-time-consuming and expensive, 
-% Can easily delete this last bit:
-effectively making 
-analysis of these datasets the preserve of a well-funded elite.
-
-To fully utilise these vast variation datasets,
-we require a primary data representation that can 
-support different types of query without 
-conversion or intermediate copies, and operate 
-directly from cloud object stores. Such a representation
-can be termed ``analysis-ready''
+time-consuming and expensive, and these vast datasets are 
+not utilised to their full potential.
+
+To achieve this full potential we 
+need a new generation of tools that operate directly
+on a primary data representation that supports 
+efficient access across a range of applications,
+with native support for cloud object storage.
+Such a representation can be termed ``analysis-ready''
 and ``cloud-native''~\citep{abernathey2021cloud}.
 For the representation to be FAIR~\citep{wilkinson2016fair},
-it must also \emph{accessible} using protocols that are 
+it must also be \emph{accessible}, using protocols that are 
 ``open, free, and universally implementable''.
@@
 data suitable for cloud deployments.
-While Hail is powerful software its file formats are not documented or 
-designed for reuse.
+Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
+for quality control of large-scale variation dataset, 
+and has been instrumental in the delivery of datasets such as 
+gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
+While Hail is built on open components
+from the Hadoop distributed computing ecosytem~\citep{white2012hadoop},
+it is complex, vertically integrated software,
+and the custom storage formats are not intended for reuse.
 Similarly, commerical solutions that have emerged to facilitate
@@
 We do this by translating VCF data into Zarr format,
-a method of storing large multidimensional data as a regular
+a method of storing large-scale multidimensional data as a regular
 grid of compressed chunks. 
-Zarr's elegant simplicity and direct mapping of chunks to 
-keys in an object store have lead to 
+Zarr's elegant simplicity and first-class support for 
+cloud object stores have led to 
 it gaining substantial traction
@@
 VCF and is competitive with state-of-the-art
-VCF compression tools. 
+file-based VCF compression tools. 
 Moreover, we show that Zarr's storage of data in an analysis-ready 

### 2024-05-08 paper.tex
Update caption for fig1

@@
 VCF is row-wise format in which 
-describes observations
-and metadata for a single variant~\citep{danecek2011variant} are
-encoded as a line of text.
-BCF (a packed binary representation) is similarly 
+observations and metadata for a single variant are
+encoded as a line of text~\citep{danecek2011variant}.
+BCF (the standard binary representation of VCF) is similarly 
 row-wise, as are the majority of proposed alternative storage formats.
 Row-wise storage makes retrieving all information
-for a given record straightforward and efficient, and 
-allows data to be easily streamed through Unix pipelines (see Discussion).
+for a given record straightforward and efficient. 
 The row-wise approach works well when records are relatively small, 
@@
 we group together all the records for a given field.
-Columnar storage formats such as Parquet~\citep{parquet2024},
+Columnar storage formats such as Parquet~\citep{parquet2024}
 make retrieving particular columns much 
@@
 Columnar techologies tend to be 
-optimised for ``tall and narrow'' data there is a large number of records
+optimised for ``tall and narrow'' data where there is a large number of records
 for a relatively small number of columns. 
 Mapping VCF directly to a columnar layout, in which there is a 
-a column for the genotypes (and other per-sample QC metrics) 
+column for the genotypes (and other per-call QC metrics) 
 leads to millions of columns, and substantial technical challenges.
@@
 \resizebox{225pt}{!}{\input{diagrams/data-model.tex}}
-\caption{Columnar storage of VCF data in chunked, compressed form using Zarr.
-[TODO more detail, including brief overview of how data maps from 
-VCF to VCF Zarr in the examples]
-\label{fig-data-model}}
+\caption{Chunked compressed storage of VCF data using Zarr. 
+The \texttt{call\_genotype} array is a three-dimensional (variants, samples,
+ploidy) array of integers, split into a uniform grid of 
+chunks determined by the variant and sample chunk sizes (10,000
+and 1,000 by default in \texttt{vcf2zarr}). Each chunk is associated 
+with a key defining its location in this grid, which can be stored 
+in any key-value store such as a standard file-system or cloud object
+store. Chunks are compressed independently using standard 
+codecs and pre-compression filters, which can be specified on a per-array
+basis. Also shown are the one-dimensional \texttt{variant\_contig} (CHROM)
+and \texttt{variant\_position} arrays (POS). Other fields are stored 
+in a similar fashion. \label{fig-data-model}}
 \end{figure}
 
-VCF is at its core an encoding of the genotype matrix where each entry
-describes the observed genotypes for a given sample at a give variant site,
+VCF is at its core an encoding of the genotype matrix, where each entry
+describes the observed genotypes for a given sample at a given variant site,
 interleaved with per-variant information
-and other matrices (e.g., the GQ or AD fields).
+and other call-level matrices (e.g., the GQ or AD fields).
 The data is largely numerical and each field is of a fixed dimension, 
@@
 We propose the VCF Zarr specification which maps the 
-VCF data model into an array-oriented layout using the
+VCF data model into an array-oriented layout based on the
 widely-used Zarr format (Fig~\ref{fig-data-model}). 
-Briefly, Zarr is a format 
-% FIXME two different uses of $n$ here in a paragraph
-for storing $n$-dimensional arrays of a fixed datatype, stored as 
-compressed chunks. In the VCF Zarr specification, 
-each field in a VCF is mapped to an array, which is compressed 
-and stored individually allowing for efficient retrieval. 
-In particular, call-level data is stored as $m \times n$ arrays
-(for $m$ sites and $n$ samples), allowing for efficient 
-retrieval of subsets of those fields along both the 
-variants and samples axis.
+Briefly, Zarr stores multidimensional arrays of a given datatype
+as compressed fixed-size chunks. In the VCF Zarr specification, 
+each field in a VCF is mapped to an array, 
+which is stored individually allowing for efficient retrieval and 
+high levels of compression.
+% In particular, call-level data is stored as $m \times n$ arrays
+% (for $m$ sites and $n$ samples), allowing for efficient 
+% retrieval of subsets of those fields along both the 
+% variants and samples axis.
 See the Methods for more detail on Zarr and the VCF Zarr

### 2024-05-08 diagrams/data-model.tex
Change Autoshuffle to Bitshuffle

@@
  \node (blosk) [rectangle, ultra thick, red, draw, rounded corners, minimum width=1.8cm, minimum height=1.6cm, anchor=center,align=center,text=black]
-    at (numcodecs) {\scriptsize Blosc\\\scriptsize ZSTD\\\scriptsize Autoshuffle};
+    at (numcodecs) {\scriptsize Blosc\\\scriptsize ZSTD\\\scriptsize BITSHUFFLE};
  \draw[ultra thick, red,-stealth] (3.9,3.5,-.25) -- (5,3.5,-.25);

### 2024-05-08 paper.tex
Pass through storage section

@@
 performance on particular types of queries, 
-typically using a command line interface
+typically using a command line interface (CLI)
 and outputting VCF text~\citep{
@@
 for a given variant. [TODO check this - reference?]
+[Note this is the approach used in OpenCGA, as shown in their storage 
+benchmark numbers]
 
@@
 and other call-level matrices (e.g., the GQ or AD fields).
-The data is largely numerical and each field is of a fixed dimension, 
+The data is largely numerical and of fixed dimension, 
 and is therefore a natural mapping to array-oriented 
@@
 We propose the VCF Zarr specification which maps the 
-VCF data model into an array-oriented layout based on the
-widely-used Zarr format (Fig~\ref{fig-data-model}). 
-Briefly, Zarr stores multidimensional arrays of a given datatype
-as compressed fixed-size chunks. In the VCF Zarr specification, 
+VCF data model into an array-oriented layout using Zarr 
+(Fig~\ref{fig-data-model}). 
+In the VCF Zarr specification, 
 each field in a VCF is mapped to an array, 
@@
 
-Remarkably, the simple approach of compressing
-two dimensional chunks with the Zstandard 
-compressor~\citep{collet2021rfc} and the bit-shuffle
-filter from Blosc~\cite{alted2010modern} produces 
-compression levels competitive with specialised methods
-on the genotype matrix. In Fig~\ref{fig-data-storage}
-we show compression performance on up to a million
-samples subset a large and highly realistic 
+One of the key benefits of Zarr is its cloud-native design, 
+but it also works well on standard file systems, where 
+arrays and chunks are stored hierarchically in directories
+and files (storage as a single Zip archive is also supported).
+To enable comparison with the existing file-based ecosystem
+of tools, we focus on this scenario in a series of illustrative 
+benchmarks in the following sections. 
+(See \citep{durbin2020task,moore2021ome,gowan2022using} for Zarr
+benchmarks in cloud settings.)
+We compare primarily with
+VCF/BCF based workflows using \texttt{bcftools} because this
+is the standard practice, used in the vast majority of cases.
+We also compare with two representative recent specialised utilities;
+see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of 
+these and other tools.
+Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused 
+on compression performance, which uses a custom file format 
+and a CLI to extract VCF as text with various filtering options.
+Savvy~\cite{lefaive2021sparse} is an extension of BCF which 
+takes advantage of sparsity in the genotype matrix as well
+as using PBWT-based approaches for improved compression.
+Savvy provides a CLI as well as a C++ API.
+Our benchmarks are based on genotype data 
+from subsets of a large and highly realistic 
 simulation of French-Canadians~\cite{anderson2023on}
-(see Methods for details).
+(see Methods for details on the dataset and benchmarking methodology).
 Note that while simulations cannot capture 
@@
 observations~\cite{anderson2023on} and so provides 
-a reasonable data point when comparing such methods.
-% Could note here that this might represent a best 
-% case scenario for specialised methods, but maybe 
-% not worth bothering with.
-
-[Short note here on why we choose these tools to
-benchmark against, and why there's no point in 
-a major benchmarkathon. Some recent papers 
-have good benchmarks which the interested 
-reader can cross-reference]
-
-Although Zarr is optimised for cloud settings, it also
-works very well in the classical (shared) file-system setting
-and we therefore focus on benchmarks in this setting
-to allow us to compare with existing tools.
-Those seeking benchmarks of Zarr in cloud deployments are 
-directed to~\citep{durbin2020task,moore2021ome,gowan2022using} [ANY OTHERS?]
+a reasonably and easily reproducible data point 
+when comparing such methods.
+The simulations only contain genotypes without any additional
+high-entropy QC fields, which is unrealistic (see Section XXX
+for benchmarks on a large human dataset that includes 
+many such fields). 
+Note, however, that such minimal, genotype-only data 
+is something of a best-case scenario for specialised genotype
+compression methods using row-based storage.
+
+Fig~\ref{fig-data-storage} shows compression performance 
+on up to a million samples for chromosome 21.
+Unsurprisingly, gzipped compressed VCF requires the most 
+storage, with BCF providing substantially better compression
+performance (note the log-log scale). Genozip has 
+superb compression, having far smaller file sizes that the 
+other methods (although somewhat losing its advantages at 
+larger sample sizes). Zarr and Savvy have 
+almost identical compression performance in this example.
+It is remarkable that the simple approach of compressing
+two dimensional chunks of the genotype matrix 
+using the Zstandard compressor~\citep{collet2021rfc} and the 
+bit-shuffle filter from Blosc~\cite{alted2010modern} produces 
+compression levels competitive with the highly specialised methods
+used by Savvy.
 

### 2024-05-08 src/plot.py
Fix missing axis labels

@@ -67,6 +67,8 @@ def plot_size(ax, df, label_y_offset=None):
 
     ax.legend()
     add_number_of_variants(df, ax)
+    ax.set_xlabel("Number of samples")
+    ax.set_ylabel("Time (seconds)")
     plt.tight_layout()
 
 

### 2024-05-08 diagrams/data-model.tex
Tweaks

@@
  \node (blosk) [rectangle, ultra thick, red, draw, rounded corners, minimum width=1.8cm, minimum height=1.6cm, anchor=center,align=center,text=black]
-    at (numcodecs) {\scriptsize Blosc\\\scriptsize ZSTD\\\scriptsize BITSHUFFLE};
+    at (numcodecs) {\scriptsize Blosc\\\scriptsize Zstandard\\\scriptsize BITSHUFFLE};
  \draw[ultra thick, red,-stealth] (3.9,3.5,-.25) -- (5,3.5,-.25);

### 2024-05-08 paper.tex
Tweaks

@@
 on up to a million samples for chromosome 21.
-Unsurprisingly, gzipped compressed VCF requires the most 
+Unsurprisingly, gzip compressed VCF requires the most 
 storage, with BCF providing substantially better compression

### 2024-05-08 paper.tex
Pass through other benchmarks

@@
 observations~\cite{anderson2023on} and so provides 
-a reasonably and easily reproducible data point 
+a reasonable and easily reproducible data point 
 when comparing such methods.
@@
 the data can be retrieved from storage, decoded and then analysed
-in the same memory space, avoiding the need for copying
-the decoded data and inter-process communication. [TODO some refs
-on zero-copy semantics and things like Arrow?]
+in the same memory space without additional copies and 
+interprocess communication through pipes.
 
@@
 which calculations can be performed under the different formats,
-we use the \texttt{bcftools +af-dist} plugin as an example of such a whole-matrix
-computation. 
+we use the \texttt{bcftools +af-dist} plugin as an example.
 % The details of the \texttt{af-dist} operation are not important:
 % as an example of a whole-matrix operation. 
-We chose this particular operation for several
-reasons. First, it is a simple but non-trival calculation that requires 
-examining every element in the genotype matrix.
+We chose this particular operation for several reasons. 
+First, it is a straightforward calculation that 
+requires examining every element in the genotype matrix,
+and can be reproduced in different programming languages
+without too much effort.
 Secondly, it produces a small volume of output (a table of
@@
 
-Figure~\ref{fig-whole-matrix-compute} shows the results of 
-running \texttt{bcftools +af-dist} and equivalent operations 
+Fig~\ref{fig-whole-matrix-compute} shows timing results 
+for running \texttt{bcftools +af-dist} and equivalent operations 
 on the data of Figure~\ref{fig-data-storage}. There is a large
@@
 not only must we decode the data from the genozip format, we must 
-then parse the large volumes of VCF text and then finally do the 
-actual calculation. Running \texttt{bcftools +af-dist} directly
-on the gzipped VCF text incurs essentially the same cost, but
-the decompression cost is lower and without the interprocess 
-communication overhead. Using a BCF file is somewhat faster,
-because the packed binary format can be read into the internal 
-data structures used by the \texttt{htslib} API more efficiently.
-
-The data shown in Figure~\ref{fig-whole-matrix-compute} for Zarr and Savvy
+then parse the large volumes of VCF text produced and then finally do the 
+actual calculation. 
+Running \texttt{bcftools +af-dist} directly on the gzipped VCF
+is substantially faster, indicating that Genozip's excellent
+compression performance comes at a substantial decompression cost.
+Using a BCF file is again significantly faster,
+because the packed binary format avoids the overhead of parsing 
+VCF text into \texttt{htslib}'s internal data structures. 
+We only use BCF for subsequent \texttt{bcftools} benchmarks.
+
+The data shown in Fig~\ref{fig-whole-matrix-compute} for Zarr and Savvy
 is based on custom programs written using their respective APIs
@@
 (up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
-Figure~\ref{fig-whole-matrix-decode}).
+Fig~\ref{fig-whole-matrix-decode}).
 % sequence_length  num_samples  num_sites       tool       size
@@
 case, if we want to access only the genotypes for a single sample
-we must still read in and decompress the entire dataset.
+we must still retrieve and decompress the entire dataset.
 
@@
 a subset of 10000 variants $\times$ 10 samples from the middle of the matrix
-for the data in Figure~\ref{fig-data-storage}.
+for the data in Fig~\ref{fig-data-storage}.
 The \texttt{genozip} and \texttt{bcftools} pipelines involve
@@
 We illustrate the cost of row-wise encoding in
-Figure~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
+Fig~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
 on a small fixed-size subset of the genotype matrices of
-Figure~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
+Fig~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
 means that this small subset of the overall matrix can be efficiently
@@
 scales with the number of samples.
-Figure~\ref{fig-subset-matrix-compute-supplemental} shows performance
+Fig~\ref{fig-subset-matrix-compute-supplemental} shows performance
 for the same operation when selecting half of the samples in the 
@@
 Total CPU time required to extract the POS field for BCF,
-sav and Zar formats
+sav and Zarr formats
 for the data in Figure~\ref{fig-data-storage}.
@@
 We have focused on the genotype matrix up to this point, contrasting
-Zarr with existing row-wise methods. The genotype matrix, however,
-is only one aspect of VCF data.
+Zarr with existing row-wise methods.
 Real-world VCFs encapsulate much more than just the genotype
-matrix, and can contain large numbers of additional fields. Efficient
-extraction of the data from these fields is a key use-case
-and an area where columnar storage has clear benefits.
-Figure~\ref{fig-column-extract} shows the time required to extract 
+matrix, and can contain large numbers of additional fields.
+Fig~\ref{fig-column-extract} shows the time required to extract 
 the genomic position of each variant in the simulated benchmark 
@@
 the entire dataset must be read into memory and 
-decompressed to extract just one field from each record. Zarr's
-columnar storage excels at these tasks: we only read and decompress
-the information required.
+decompressed to extract just one field from each record. Zarr
+excels at these tasks: we only read and decompress the information required.
 

### 2024-05-08 diagrams/data-model.tex

--- 2024-05-08 diagrams/data-model.tex
@@
  \node (blosk) [rectangle, ultra thick, red, draw, rounded corners, minimum width=1.8cm, minimum height=1.6cm, anchor=center,align=center,text=black]
-    at (numcodecs) {\scriptsize Blosc\\\scriptsize ZSTD\\\scriptsize BITSHUFFLE};
+    at (numcodecs) {\scriptsize Blosc\\\scriptsize Zstandard\\\scriptsize BitShuffle};
  \draw[ultra thick, red,-stealth] (3.9,3.5,-.25) -- (5,3.5,-.25);

### 2024-05-08 software/savvy-afdist/main.cpp
Add support for extracting to memomry

@@ -1,4 +1,6 @@
-/* To any C++ programmers reading this file: apologies! */
+/* To any C++ programmers reading this file: apologies! It's obviously the work
+ * of non C++ programmers, but hopefully the nastiness won't make it
+ * slower. If so, please do let us know and we'll fix. */
 
 #include <fstream>
 #include <iostream>
@@ -64,15 +66,20 @@ void decode(const std::string& filename) {
     std::cout << "Decoded variants: " << count << std::endl;
 }
 
-void output_pos(const std::string& filename) {
+void output_pos(const std::string& filename, bool mem_only) {
     savvy::reader f(filename);
     savvy::variant var;
+    std::vector<int> pos;
 
     while (f >> var) {
-        std::cout << var.position() << std::endl;
+        pos.push_back(var.position());
+    }
+    if (! mem_only) {
+        for(unsigned int j = 0; j < pos.size(); j++) {
+            std::cout << pos[j] << std::endl;
+        }
     }
 }
-
 
 // Use a lambda expression for finding the bin index
 auto findBinIndex = [](const std::vector<double>& bins, double value) -> int {
@@ -96,6 +103,7 @@ int main(int argc, char* argv[]) {
     std::string samples_file;
     bool decode_only = false;
     bool pos_only = false;
+    bool mem_only = false;
     int start = -1;
     int end = -1;
 
@@ -105,6 +113,8 @@ int main(int argc, char* argv[]) {
             decode_only = true;
         } else if (arg == "--pos-only") {
             pos_only = true;
+        } else if (arg == "--mem-only") {
+            mem_only = true;
         } else if (arg == "--start") {
             i++;
             start = std::stoi(argv[i]);
@@ -139,7 +149,7 @@ int main(int argc, char* argv[]) {
     if (decode_only) {
         decode(filename);
     } else if (pos_only) {
-        output_pos(filename);
+        output_pos(filename, mem_only);
     } else {
         std::vector<double> af;
         std::vector<int> hets;

### 2024-05-08 paper.tex
some notes on hardware

@@
 
+All simulation-based benchmarks were performed on a 
+dual CPU (Intel Xeon E5-2680 v2)
+server with 256GiB of RAM running Debian GNU/Linux 11.
+To ensure that the true effects of having data distributed over a large 
+number of files were reported, benchmarks for Zarr and Savvy were 
+performed on a cold disk cache by running 
+\texttt{echo 3 | sudo tee /proc/sys/vm/drop\_caches} before each run.
+The IO subsytem used is based on a RAID 5 of 12 SATA hard drives.
+
 \begin{itemize}

### 2024-05-08 paper.tex

@@
 \texttt{variant\_position} into a NumPy array, and then wrote to
-a text file using the Pandas
-\texttt{write\_csv} method. Writing
-text dominates in the case of Zarr: populating the NumPy array 
-in memory takes less than a second in all cases. Genozip does
-not offer an option to extract individual columns, and so would require
-the output to be piped into \texttt{bcftools query} or a
-Unix utility such as \texttt{cut}.
+a text file using the Pandas \texttt{write\_csv} method. 
+Zarr CPU time is dominated by writing the text output; we also show
+the time required to populate a NumPy array with the data in Zarr,
+which is less than a second. Wall-clock time (dotted line) is dominated 
+in this case by file I/O. Time to output text for Savvy is not significant
+for $> 1000$ samples (not shown).
 \label{fig-column-extract}}

### 2024-05-09 paper.tex
Tom's changes

@@
 % for discussion
-\title{Analysis-ready VCF at Biobank-scale using Zarr}
+\title{Analysis-ready VCF at Biobank scale using Zarr}
 
@@
 variant, but accessing subsets of the data within rows is inefficient.
-Biobank-scale datasets currently available 
+Biobank scale datasets currently available 
 consist of hundreds of thousands of whole genomes 
@@
 VCF as part of 1000 Genomes project~\citep{10002015global},
-with Biobank-scale initiatives such as 
+with Biobank scale initiatives such as 
 Genomics England~\cite{turnbull2018100},
@@
 
-Processing of Biobank-scale datasets can be split into a 
+Processing of Biobank scale datasets can be split into a 
 few broad categories. The most basic analysis 
@@
 To enable comparison with the existing file-based ecosystem
-of tools, we focus on this scenario in a series of illustrative 
+of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
 benchmarks in the following sections. 
@@
 be more concrete]. The row-wise nature of these methods, however, mean 
-that we cannot efficiently subset \emph{by sample}. In the extreme
+that we cannot efficiently subset \emph{by sample}
+(e.g., to calculate statistics within a particular cohort). In the extreme
 case, if we want to access only the genotypes for a single sample
@@
 In this approach there are some significant departures from 
-standard practise, which warrant some discussion here.
+standard practice, which warrant some discussion here.
 
 Perhaps the largest deviation in the approach that we advocate here 
-from standard practise is to decouple our concept of a dataset from
+from standard practice is to decouple our concept of a dataset from
 a file. Each VCF is broken up into a set of Zarr arrays, each 
@@
 Zarr has also seen recent uptake 
-in single-scell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
+in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
 and multimodal spatial omics
@@
 The VCF Zarr specification can represent anything described by BCF
-(which is somewhat more restrictive than BCF) except for two corner
+(which is somewhat more restrictive than VCF) except for two corner
 cases related to the encoding of missing data. Firstly, VCF Zarr does
@@
 \texttt{echo 3 | sudo tee /proc/sys/vm/drop\_caches} before each run.
-The IO subsytem used is based on a RAID 5 of 12 SATA hard drives.
+The IO subsystem used is based on a RAID 5 of 12 SATA hard drives.
 

### 2024-05-09 src/plot.py
Run black on new code

@@ -73,7 +73,9 @@ def plot_size(ax, df, label_y_offset=None):
     plt.tight_layout()
 
 
-def plot_total_cpu(ax, df, toolname=None, colours=None, time_units="h", extrapolate=None):
+def plot_total_cpu(
+    ax, df, toolname=None, colours=None, time_units="h", extrapolate=None
+):
     if colours is None:
         colours = {
             "bcftools+vcf": vcf_colour,
@@ -262,8 +264,13 @@ def column_extract(time_data, output):
     }
     fig, ax1 = one_panel_fig()
     plot_total_cpu(ax1, df, toolname=toolname, time_units="s", extrapolate=["bcftools"])
-    plot_total_cpu(ax1, df_mem, colours={"zarr": "black"}, toolname={"zarr": "Zarr (memory)"},
-            time_units="s")
+    plot_total_cpu(
+        ax1,
+        df_mem,
+        colours={"zarr": "black"},
+        toolname={"zarr": "Zarr (memory)"},
+        time_units="s",
+    )
     plt.savefig(output)
 
 
@@ -312,24 +319,29 @@ def subset_matrix_compute_supplemental(data, output):
 @click.argument("data", type=click.File("r"))
 @click.argument("output", type=click.Path())
 def plot_compression_ratio_grid(data, output):
-
     df = pd.read_csv(data)
-    
+
     # Filter to only arrays in the same dim order as the original:
-    df_plot = df.loc[df.dim_order.isin(['(0, 1)', '(0, 1, 2)'])]
+    df_plot = df.loc[df.dim_order.isin(["(0, 1)", "(0, 1, 2)"])]
 
     # Filter out results with Quantize / PackBits:
-    df_plot = df_plot.loc[df_plot.cname == 'zstd']
+    df_plot = df_plot.loc[df_plot.cname == "zstd"]
 
-    g = sns.catplot(df_plot, kind='bar', 
-                    row='ArrayName', col='shuffle', 
-                    x='named_chunksize', y='CompressionRatio', 
-                    sharex=False, sharey='row')
-    
-    for i, ax in enumerate(g.fig.axes):   ## getting all axes of the fig object
-         ax.set_xticklabels(ax.get_xticklabels(), rotation = 90)
-    
-    plt.subplots_adjust(hspace = 1.)
+    g = sns.catplot(
+        df_plot,
+        kind="bar",
+        row="ArrayName",
+        col="shuffle",
+        x="named_chunksize",
+        y="CompressionRatio",
[…]

### 2024-05-09 src/plot.py
Add compression shuffle plot

@@ -349,6 +349,50 @@ def plot_compression_ratio_grid(data, output):
 @click.command()
 @click.argument("data", type=click.File("r"))
 @click.argument("output", type=click.Path())
+def compression_shuffle(data, output):
+    """
+    Plot figure showing the effect of shuffle settings on compression ratio.
+    """
+    df = pd.read_csv(data)
+
+    # Note this is ordered by best-to-worst compression for viz
+    arrays = [
+        "call_GQ",
+        "call_DP",
+        "call_AD",
+        "call_AB",
+        "call_genotype",
+    ]
+    df_sub = df[
+        df.ArrayName.isin(arrays)
+        & (df.sample_chunksize == 1000)
+        & (df.variant_chunksize == 10000)
+        & (df.cname == "zstd")
+        & (df.dim_order.isin(["(0, 1)", "(0, 1, 2)"]))
+    ].copy()
+    names = np.array(["No Shuffle", "Byte Shuffle", "Bit Shuffle"])
+    df_sub["Shuffle"] = names[df_sub["shuffle"]]
+
+    fig, ax = one_panel_fig()
+    sns.barplot(
+        df_sub,
+        orient="h",
+        order=arrays,
+        y="ArrayName",
+        x="CompressionRatio",
+        hue="Shuffle",
+        ax=ax,
+    )
+    ax.set_ylabel("")
+    ax.get_legend().set_title("")
+
+    plt.tight_layout()
+    plt.savefig(output)
+
+
+@click.command()
+@click.argument("data", type=click.File("r"))
+@click.argument("output", type=click.Path())
 def plot_compression_dim_shuffle(data, output):
     df = pd.read_csv(data)
 
@@ -417,6 +461,7 @@ cli.add_command(whole_matrix_decode)
 cli.add_command(column_extract)
 cli.add_command(subset_matrix_compute)
 cli.add_command(subset_matrix_compute_supplemental)
+cli.add_command(compression_shuffle)
 cli.add_command(plot_compression_ratio_grid)
 cli.add_command(plot_compression_dim_shuffle)
 cli.add_command(plot_compression_packbits)

### 2024-05-09 paper.tex
Add compression shuffle plot

@@
 addressed by their indexes along the dimensions of the array, and the 
-compressed data associated with this key. Chunks are
-are often stored in individual files, but a wide array of different
-stores are supported including cloud buckets and NoSQL databases;
+compressed data associated with this key. Chunks can
+be stored in individual files (as we do here), but a wide array of different
+storage backends are supported including cloud object stores 
+and NoSQL databases;
 in principle, Zarr can store data in any key-value store.
@@
 (see section XXX).
-Two key technologies used by Zarr are the Blosc
+Two key technologies often used in conjunction with Zarr are the Blosc
 % This seems like the best citation for Blosc? It is mentioned here
@@
 access to several different compression codecs. The 
-Zstandard compressor~\citep{collet2021rfc} is of particular 
+Zstandard codec is of particular 
 interest here as it achieves very high compression ratios
-with fast decompression speeds [see fig]. 
+with good decompression speeds. See Fig~\ref{fig-whole-matrix-decode}
+for a comparison of Zstandard with LZ4 on decoding performance.
 % This may seem a bit off topic, but want to reassure reader that 
@@
 Some of the structures from the VCF header are also represented as arrays,
-including contigs, filters, and samples. Making every item in VCF Zarr an
-array means that the data can be accessed via a uniform API.
+including contigs, filters, and samples.
 
@@
 To take one example, consider the VCF AD genotype field defined by the
-following VCF header: \texttt{\#\#FORMAT=<ID=AD,Number=A,Type=Integer,Description="Allele Depths">}
+following VCF header: \texttt{
+\#\#FORMAT=<ID=AD,Number=A,Type=Integer,Description="Allele Depths">}.
 The FORMAT key \texttt{ID} maps to an array name of \texttt{call\_AD}
@@
 so the shape of this array is \texttt{(variants, samples, alt\_alleles)}.
-The VCF Integer type can be represented as any Zarr integer type, and in
-fact the specification doesn't mandate a particular integer width, so it
-could be encoded as a 16 bit integer for example.
-% Refer to GeL case study section?
-
-VCF Zarr uses special sentinel values to encode missing values to indicate
-that a value is not present, and fill values to pad variable length fields.
-The latter is necessary since Zarr arrays have fixed size dimensions. The
-specification also permits extra boolean arrays to accompany a field array
-to redundantly encode the missing and fill status, which can be useful for
-efficiently masking out values that shouldn't be processed, at the cost of
-a small amount of extra storage.
+The VCF Integer type can be represented as any Zarr integer type, 
+and the specification doesn't mandate particular integer widths. 
+The \texttt{vcf2zarr} (see the next section) conversion utility 
+chooses the narrowest integer width that can represent the data in
+each field.
+
+An important aspect of VCF Zarr is that field dimensions are 
+global and fixed, and defined as the maximum across all rows.
+Continuing the example above, the third dimension of the array 
+is the maximum number of alternate alleles across \emph{all} 
+variants. For variants at which there are less than the maximum
+number of alternative alleles, the third dimension of the 
+\texttt{call\_AD} array
+is padded with a sentinel value (-2 for integers and a specific
+non-signalling NaN for floats). While this is not a problem 
+in practise for datasets in which all four bases are observed, 
+it is a substantial issue for
+fields that have a quadratic dependency on the number of alleles
+(Number=G) such as PL. Such fields are already known to cause 
+significant problems, and the ``local alleles'' proposal 
+provides an elegant solution~\cite{poterba2024scalable}.
+As this approach is on a likely path to 
+standardisation~\cite{danecek2021twelve}, we would aim to 
+include support in later versions of VCF Zarr.
 
@@
 integers). However, if -1 and -2 need to be stored, a float field
-can be used without issues (since the float sentinel is a special NaN
-value).
+can be used without issues.
 
@@
 
+\begin{figure}
+\includegraphics{figures/compression-shuffle}
+\caption{Effects of shuffle setting on compression ratios on 1000 Genomes
+data. TODO MORE DETAIL. 
+\label{fig-compression-shuffle}}
+\end{figure}
+
 \end{document}

### 2024-05-10 paper.tex
Pass through the vcf2zarr section

@@
 \subsection{vcf2zarr}
-Converting VCF to Zarr format at Biobank scale is challenging. This main issue is 
-around memory usage: although we can view each record in the VCF one-by-one,
-we must buffer a full chunk (10,000 variants is reasonable choice) 
-in the variants dimension for each of the arrays. 
+Converting VCF to Zarr at Biobank scale is challenging. 
+One problem is to determine the dimension of fields,
+(i.e., finding the maximum number of alternate alleles and the 
+maximum size of \texttt{Number=.} fields) which requires a full
+pass through the data. Another challenge is to keep
+memory usage within reasonable limits: although 
+we can view each record in the VCF one-by-one,
+we must buffer a full chunk (10,000 variants is the default in
+\texttt{vcf2zarr}) 
+in the variants dimension for each of the fields to convert to Zarr. 
 For VCFs with many FORMAT fields and large numbers of samples this can
-easily require tens of gigabytes of RAM per worker, which makes 
-parallelism troublesome. Reading the VCF multiple times for different fields
+require tens of gigabytes of RAM per worker, making
+parallelism difficult. Reading the VCF multiple times for different fields
 is possible, but would be prohibitively slow for multi-terabyte VCFs.
@@
 Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
-of VCFs and stores the data in per-field chunks (by default, 64MiB).
-In this way, the memory required is bounded by the size of single record 
-BCF record in htslib (plus the size of a decoded bgzf record
-[FIXME what is the correct terminology here?]), plus 64MiB for each 
-field.  The Intermediate Columnar Format stores the data for each 
-field spread across multiple partitions of the data (so that 
-multiple parts of the VCF can be read in parallel), and then 
-fixed size compressed chunks within those partitions. It also stores information
-about the number of records in each partition and chunk of a given
-field, so that the record at a particular index can be retrieved efficiently 
-later. Summaries such as the minimum and maximum value 
+of VCFs, and reads through them using
+\texttt{cyvcf2}~\cite{pedersen2017cyvcf2},
+storing each field independently in (approximately) fixed-size 
+compressed chunks.
+Large files can be partitioned based on information extracted from the 
+CSI or Tabix indexes, and so different parts of a file can be 
+converted to ICF in parallel.
+Once all partitions have completed, information about 
+the number of records in each partition and chunk of a given
+field is stored so that the record at a particular index 
+can be efficiently retrieved. 
+Summaries such as maximum dimension and the minimum and maximum value 
 of each field are also maintained, to aid choice of data types later.
 A set of VCF files can be converted to intermediate columnar 
-format in parallel on a single machine using a given number 
-of worker processes using the \texttt{explode} command,
+format in parallel on a single machine 
+using the \texttt{explode} command,
 or can be distributed across a cluster using the 
@@
 
-Once the VCF data has converted to the intermediate columnar format,
+Once the VCF data has been converted to the intermediate columnar format,
 it can then be converted to Zarr using the \texttt{vcf2zarr encode}
-command. Alhough there is a one-to-one mapping between the VCF
-fields and the Zarr arrays, there is considerable flexibility 
-in how the data is stored, for example in integer width
-and compression and filter settings.
-By default vcf2zarr chooses integer type sizes based 
-on the range of values seen in the intermediate column format
+command. We take advantage of Zarr's flexibility
+in \texttt{vcf2zarr} by choosing integer widths based 
+on the range of values seen in the ICF
 and reasonable compressor defaults, but these can be
-altered by generating a schema in JSON format and editing it.
+modified by generating and editing a JSON-formatted storage schema.
 Encoding a given column (for example, \texttt{call\_AD})
@@
 array in question, and then sequentially filling this buffer with 
-values read from the intermediate columnar format. Once the buffer 
-is full, the chunks in the samples dimension are compressed
-and written to file. [Something about parallelism and distributing]
-
-Fields such as PL are currently problematic when there are large 
-numbers of alleles. However, we plan on implementing the 
-local alleles strategy~\citep{poterba2024scalable,danecek2021twelve}
-which will likely become part of the VCF specification.
+values read from ICF and flushing to file. 
+Similar to the \texttt{explode} command,
+encoding to Zarr can be done in parallel on a single 
+machine using the \texttt{encode} command,
+or can be distributed across a cluster using the 
+\texttt{dencode-init},
+\texttt{dencode-partition} and \texttt{dencode-finalise} commands.
+The distributed commands are fault-tolerant, reporting any failed
+parititions so that they can be retried.
 

### 2024-05-10 paper.tex
Some updates on benchmarking methodology

@@
 for running \texttt{bcftools +af-dist} and equivalent operations 
-on the data of Figure~\ref{fig-data-storage}. There is a large
+on the data of Fig~\ref{fig-data-storage}. There is a large
 difference in the time required (note the log-log scale). 
@@
 in \texttt{vcf2zarr} by choosing integer widths based
-on the range of values seen in the ICF
+
 and reasonable compressor defaults, but these can be
@@
 array in question, and then sequentially filling this buffer with 
-values read from ICF and flushing to file. 
-Similar to the \texttt{explode} command,
+values read from ICF and flushing to file. Similar to the \texttt{explode} command,
 encoding to Zarr can be done in parallel on a single 
@@
 \subsection{Benchmarks}
-Full details of the benchmarking done in
-Figure~\ref{fig-data-storage} and
-Figure~\ref{fig-whole-matrix-compute}.
-Quick notes:
-
+In this section we describe the methodology used for the simulation-based
+benchmarks of Figs~\ref{fig-data-storage},\ref{fig-whole-matrix-compute},
+\ref{fig-subset-matrix-compute} and \ref{fig-column-extract}.
+The benchmarks use data simulated by conditioning on a large
+pedigree of French-Canadians using 
+\texttt{msprime}~\citep{baumdicker2021efficient},
+which have been shown to follow patterns observed in real 
+data from the same population to a remarkable 
+degree~\cite{anderson2023on}.
+We begin by downloading the simulated ancestral recombination 
+graph~\cite{brandt2024promise,lewanski2024era,wong2024general}
+for chromosome 21 from Zenodo~\cite{anderson2023simulated} 
+in compressed \texttt{tszip} format. This 552M file 
+contains the simulated ancestry and mutations for 1.4 million
+present-day samples. We then subset the full simulation 
+down to ${10^1, 10^2, \dots, 10^6}$ samples using 
+ARG simplification~\cite{kelleher2018efficient,wong2024general},
+storing the subsets in \texttt{tskit} format~\cite{tskit2024}.
+Note that this procedure captures the growth 
+in the number of variants (shown in the top x-axis labels)
+as we increase sample sizes as a 
+natural consequence of population-genetic processes.
+We then export the variation data from each subset to VCF
+using \texttt{tskit vcf subset.ts | bgzip > subset.vcf.gz}
+as the starting point for other tools.
+
+We used bcftools version 1.18, Savvy 2.1.0, Genozip 5.0.26
+and vcf2zarr 0.0.9. All tools used default settings, 
+unless otherwise stated.
 All simulation-based benchmarks were performed on a 
@@
 \texttt{echo 3 | sudo tee /proc/sys/vm/drop\_caches} before each run.
-The IO subsystem used is based on a RAID 5 of 12 SATA hard drives.
-
-\begin{itemize}
-\item Dataset is based on subsets of French-Canadian sims. We simplify
-for increasingly large subsets, keeping only variant sites. We then
-convert to VCF and encode to BCF using
-\texttt{tskit vcf | bcftools view -O b}.
-\item We used a chunk size of XX for sgkit, after some experimentation.
-This gave n files of around x MB each for the ``call\_genotypes`` array.
-We used the defaults for savvy and genozip.
-\item For the afdist CPU time we measure the sum of the total user and
-system times required to execute the full command, as reported by GNU
-time. Each tool was instructed to use one thread, where the options
-were provided. For genozip, the time required to compute
-\texttt{genocat file.genozip | bcftools +af-dist}. Because commands
-in the pipeline execute concurrently on mutiple cores, the total CPU time is
-greater than the elapsed time.
-\item Where possible in pipelines we use uncompressed BCF
- output \texttt{-Ou} to make processing more efficient (skip printf
-and parsing costs, which are substantial). Following advice
-in~\citep{danecek2021twelve}.
-\item We do not use BCF output in genozip because it doesn't support
-it directly, only VCF (supports BCF by pipeing through bcftools).
-\end{itemize}
+The I/O subsystem used is based on a RAID 5 of 12 SATA hard drives.
+For the CPU time benchmarks we measure the sum of the total user and
+system times required to execute the full command (as reported by GNU
+\texttt{time}) as well as elapsed wall-clock time. 
+Each tool was instructed to use one thread, where the options
+were provided. 
+Where possible in pipelines we use uncompressed BCF
+output \texttt{-Ou} to make processing more efficient~\citep{danecek2021twelve}.
+We do not use BCF output in genozip because it is not supported
+directly.
 

### 2024-05-10 paper.tex

@@
 Elapsed time is also reported (dotted line). Run-time for genozip
-at $10^6$ samples was extrapolated by fitting an exponential.
+and bcftools on VCF
+at $10^6$ samples were extrapolated by fitting an exponential.
 See Methods for full details.
@@
 natural consequence of population-genetic processes.
+% jk@holly$ python3 src/collect_data.py site-allele-report
+% scaling/data/chr21_10_6.zarr/
+% 0 alleles:  100.00%
+% 1 alleles:  100.00%
+% 2 alleles:  7.92%
+% 3 alleles:  0.21%
+As a result of simulated mutational processes,
+most sites have one alternate allele, 
+with 7.9\% having two and 0.2\% having three
+alternate alleles in the $10^6$ samples dataset.
 We then export the variation data from each subset to VCF
@@
 
-We used bcftools version 1.18, Savvy 2.1.0, Genozip 5.0.26
-and vcf2zarr 0.0.9. All tools used default settings, 
+We used bcftools version 1.18, Savvy 2.1.0, Genozip 5.0.26,
+vcf2zarr 0.0.9, and Zarr-Python 2.17.2. 
+All tools used default settings, 
 unless otherwise stated.
@@
 system times required to execute the full command (as reported by GNU
-\texttt{time}) as well as elapsed wall-clock time. 
+\texttt{time}) as well as elapsed wall-clock time. Total CPU
+time is shown as a solid line, with wall-clock time as a dashed line 
+of the same colour. In the case of pipelines, where some processing 
+is conducted concurrently wall-clock time can be less than total
+CPU (e.g.\ genozip in Fig~\ref{fig-whole-matrix-compute}). 
+When I/0 costs are significant, wall-clock time can be greater 
+than total CPU (e.g.\ Zarr and Savvy in Fig~\ref{fig-subset-matrix-compute}).
 Each tool was instructed to use one thread, where the options
@@
 Where possible in pipelines we use uncompressed BCF
-output \texttt{-Ou} to make processing more efficient~\citep{danecek2021twelve}.
+output (\texttt{-Ou}) to make processing 
+more efficient~\citep{danecek2021twelve}.
 We do not use BCF output in genozip because it is not supported
 directly.
+
+Because \texttt{bcftools +af-dist} requires the AF INFO field
+and this is not kept in sync by \texttt{bcftools view}  
+(although the AC and AN fields are), the subset calculation 
+for Fig~\ref{fig-subset-matrix-compute} requires an additional step. 
+The resulting pipeline is
+\texttt{bcftools view -r REGION -S SAMPLESFILE -IOu BCFFILE | 
+bcftools +fill-tags -Ou | bcftools +af-dist}. Genozip similarly
+requires a \texttt{+fill-tags} step in the pipeline.
 

### 2024-05-10 paper.tex
A paragraph on Python data science
Closes #6
Closes #88

@@
 collecting genome sequence data for hundreds of thousands of humans.
-VCF's simple text-based design and widespread support makes it an 
+VCF's simple text-based design and widespread
+support~\cite{garrison2022spectrum} makes it an 
 excellent archival format, but it is an inefficient basis for analysis.
@@
 for example by phasing~[CITES] and imputation[CITES], 
-creating additional copies of the dataset.
+creating additional dataset copies.
 A common task to perform after QC is 
@@
 analyses, to find regions or samples of interest for a particular 
-study (see section XXX for an example).
+study~\cite[e.g.][]{chen2024novo}.
 
@@
 often required to stay in-situ by data-access agreements.
-Large files are stored in cloud environments, where the 
+Large files are currently stored in cloud environments, where the 
 file systems that are required by classical file-oriented tools
@@
 ``open, free, and universally implementable''.
-There is no currently no efficient, FAIR representation of genetic variation
+There is currently no efficient, FAIR representation of genetic variation
 data suitable for cloud deployments.
 Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
-for quality control of large-scale variation dataset, 
+for quality control of large-scale variation datasets, 
 and has been instrumental in the delivery of datasets such as 
@@
 from the Hadoop distributed computing ecosytem~\citep{white2012hadoop},
-it is complex, vertically integrated software,
-and the custom storage formats are not intended for reuse.
+it is complex, vertically integrated software
+with custom storage formats that are not intended for reuse.
 Similarly, commerical solutions that have emerged to facilitate
@@
 file definition, and show how the data can be 
-efficiently stored and analysed in a cloud-native, FAIR manner.
+compactly stored and efficiently analysed in a cloud-native, FAIR manner.
 We do this by translating VCF data into Zarr format,
@@
 datasets in cloud deployments (see Methods for details).
-We present the VCF Zarr specification which formalises this 
+We present the VCF Zarr specification that formalises this 
 mapping, and the \texttt{vcf2zarr} 
-utility which reliably converts large-scale VCFs into Zarr 
-format. We show that VCF Zarr is much more compact than 
+utility to reliably convert large-scale VCFs to Zarr.
+We show that VCF Zarr is much more compact than 
 VCF and is competitive with state-of-the-art
@@
 \section{Discussion}
-The introduction of VCF as part of the work of the 1000 Genomes
-project~\cite{10002015global} has transformed a fragmented 
-landscape of mutually incompatible per-program file formats
-into an ecosystem of tools that largely
-interoperate~\cite{garrison2022spectrum}. In this 
-sense VCF has been a huge success, and the underlying data model
-is now deeply embedded in bioinformatics practice. As we have 
-illustrated here, the shortcomings of the VCF and BCF file
-formats (and more generally \emph{any} row-wise encoding of the 
-data model) in terms of computational accessibility are severe,
-and there can be little doubt that these file formats 
-are incurring a large economic cost and 
-holding back progress. 
-As we have illustrated in the Genomics England
-example here, these shortcomings are most acutely felt in 
-large-scale data human datasets but any setting in which we 
-wish to extract columns of the data from a row-wise 
-VCF format will be inefficient 
-when datasets are larger than a few hundred megabytes.
-
-As we have demonstrated, the columnar approach, based on the 
-popular Zarr standard and widely used compression libraries, 
-can support many different types of data access
-patterm efficiently, and has excellent computational accessibility.
-In this approach there are some significant departures from 
-standard practice, which warrant some discussion here.
-
-Perhaps the largest deviation in the approach that we advocate here 
-from standard practice is to decouple our concept of a dataset from
-a file. Each VCF is broken up into a set of Zarr arrays, each 
-corresponding to a large number of chunks stored in a directory
-hierarchy (Zarr also supports other forms of chunk storage 
-including single files and key-value stores however; see Methods).
-While there are some disadvantages to large numbers of small files,
-we argue that the benefits greatly outweigh them, particularly 
-in the large, centrally controlled dataset setting. 
-Firstly, this approach is inherently parallelisable, and chunks across the 
-dataset written and read in parallel without needing complex 
-synchronisation and prior layout planning. 
-Such parallel write access is very difficult 
-in the single-file setting, and one of the key weaknesses 
-of the HDF5 format~\citep{folk2011overview} [CITATION with 
-parallel troubles from HDF5?]
-Secondly, mapping chunks in arrays to individually addressed
-keys maps well to large-scale cloud object storage. Indeed,
-Zarr is specifically designed for such usage, and 
-several petabyte scale datasets have been successfully deployed
-and used [citations?].
-We have focused here on the classical single-server and 
-HPC cluster settings, but a major advantage of Zarr is 
-its cloud readiness.
+% The introduction of VCF as part of the work of the 1000 Genomes
+% project~\cite{10002015global} has transformed a fragmented 
+% landscape of mutually incompatible per-program file formats
+% into an ecosystem of tools that largely
+% interoperate~\cite{garrison2022spectrum}. In this 
+% sense VCF has been a huge success, and the underlying data model
+% is now deeply embedded in bioinformatics practice. As we have 
+% illustrated here, the shortcomings of the VCF and BCF file
+% formats (and more generally \emph{any} row-wise encoding of the 
+% data model) in terms of computational accessibility are severe,
+% and there can be little doubt that these file formats 
+% are incurring a large economic cost and 
+% holding back progress. 
+% As we have illustrated in the Genomics England
+% example here, these shortcomings are most acutely felt in 
+% large-scale data human datasets but any setting in which we 
+% wish to extract columns of the data from a row-wise 
+% VCF format will be inefficient 
+% when datasets are larger than a few hundred megabytes.
+
+% As we have demonstrated, the columnar approach, based on the 
+% popular Zarr standard and widely used compression libraries, 
+% can support many different types of data access
+% patterm efficiently, and has excellent computational accessibility.
+% In this approach there are some significant departures from 
+% standard practice, which warrant some discussion here.
+
+% Perhaps the largest deviation in the approach that we advocate here 
+% from standard practice is to decouple our concept of a dataset from
+% a file. Each VCF is broken up into a set of Zarr arrays, each 
+% corresponding to a large number of chunks stored in a directory
+% hierarchy (Zarr also supports other forms of chunk storage 
+% including single files and key-value stores however; see Methods).
+% While there are some disadvantages to large numbers of small files,
+% we argue that the benefits greatly outweigh them, particularly 
+% in the large, centrally controlled dataset setting. 
+% Firstly, this approach is inherently parallelisable, and chunks across the 
+% dataset written and read in parallel without needing complex 
+% synchronisation and prior layout planning. 
+% Such parallel write access is very difficult 
+% in the single-file setting, and one of the key weaknesses 
+% of the HDF5 format~\citep{folk2011overview} [CITATION with 
+% parallel troubles from HDF5?]
+% Secondly, mapping chunks in arrays to individually addressed
+% keys maps well to large-scale cloud object storage. Indeed,
+% Zarr is specifically designed for such usage, and 
+% several petabyte scale datasets have been successfully deployed
+% and used [citations?].
+% We have focused here on the classical single-server and 
+% HPC cluster settings, but a major advantage of Zarr is 
+% its cloud readiness.
+
 % Probably getting into the weeds too much.
@@
 % benefits, both in terms of storage costs and performance.
-Thirdly, breaking up fields into different directories (or 
-cloud buckets) enables fine-grained access control to the 
-dataset. For example, all users may be allowed access to
-variant-level data, but only a subset permitted to access
-call-level individual data. Such access controls can be 
-straightforwardly applied using file or cloud-hosting 
-methods on the relevant directores/buckets. Such concerns
-are entirely decoupled from the file format.
+% Thirdly, breaking up fields into different directories (or 
+% cloud buckets) enables fine-grained access control to the 
+% dataset. For example, all users may be allowed access to
+% variant-level data, but only a subset permitted to access
+% call-level individual data. Such access controls can be 
+% straightforwardly applied using file or cloud-hosting 
+% methods on the relevant directores/buckets. Such concerns
+% are entirely decoupled from the file format.
 
@@
 updating a given per-call mask. Masks are cheap (e.g., the 
-standard genotype missingness mask in~\ref{tab-genomics-england-data}
+standard genotype missingness mask in Table~\ref{tab-genomics-england-data}
 uses less than 1\% of the overall storage).
@@
 [Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
+
 The VCF Zarr specification and scalable vcf2zarr conversion utility
@@
 
-[Random fragment to include somewhere]
-Row-wise storage of the genotype matrix makes accessing 
-sample haplotypes (i.e. the columns),
-required for many
-applications~\cite[e.g.][]{durbin2014efficient,kelleher2019inferring} 
-[TODO add phasing, imputation]
-inefficient. The simple approach of storing the genotype matrix
-in rectangular chunks allows efficient access both by row \emph{and}
-column. Similarly, these rectangular chunks capture patterns
-both of similarity between samples and linkage disequilibrium~\cite{mcvean2019linkage} 
-between sites, such that conventional compression schemes 
-perform very well~\ref{fig-data-storage}.
-
-The Python data science ecosystem is increasingly popular, 
-with packages such as NumPy~\citep{harris2020array},
-SciPy~\citep{virtanen2020scipy},
-Pandas~\citep{mckinney2010data},
-Jupyter~\citep{kluyver2016jupyter}
-enabling much of contemporary science.
-
-BioNumpy~\citep{rand2022bionumpy},
-
-% Although the Hadoop ecosystem has been hugely successful
-% in many domains, it has had a limited impact in genomics.
-% [List some things that have been tried for e.g variant calling
-% , like ADAM~\citep{nothaft2015rethinking}].
-% [Briefly mention two approaches using Hadoop clusters
-% ~\citep{boufea2017managing,fan2020variant}, to not very
-% exiting results].
-% Distributed analytics for large-scale data is a mature
-% techology, with many different approaches. Hadoop~\citep{white2012hadoop}
-% is the original method, and remains the basis of the
-% dominant ecosystem. HDFS. Spark [CITATION?] lets us
-% do more complex analytics. Parquet [CITATION?]
-% Distributed computation has been the reality in the analysis
-% of genetic variation for many years,
-% where data is most often provided as per-chromosome VCFs.
-% This chunking provides a fairly straightforward way to parallelise
-% computations, using HPC schedulers to split work per-file.
-% When more fine-grained parallelism is required, users must
-% provide exact genome coordinates of the genome regions
-% as part of their submission scripts. Combining results
-% is a manual operation.
+The Python data science ecosystem, 
+built on the foundations of efficient array-oriented 
+computing~\citep{harris2020array},
+has a rich suite of powerful tools~\cite{
+mckinney2010data,
+rocklin2015dask,
+kluyver2016jupyter,
+hoyer2017xarray,
+virtanen2020scipy} and gaining increasing
+popularity in recent biological applications~\citep[e.g.][]{
+abdennur2020cooler,
+rand2022bionumpy,
+open2c2024bioframe,
+hou2024admix}.
+Zarr has the potential to act as a unified, cloud-native 
+efficient storage layer, freeing the next generation of 
+bioinformatics tools
+from the traditional burdens of text parsing and format
+conversion.
+
+% Zarr has also seen recent uptake 
+% in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
+% and multimodal spatial omics
+% data~\citep{marconato2024spatialdata,baker2023emobject}.
+% Recent additions using Zarr include the application of deep learning
+% models to genomic sequence data~\citep{klie2023predictive} 
+% and a browser for genetic variation data~\citep{konig2023divbrowse}.
+
 % Workflow engines such as
@@
 
-\section{Choosing default compressor settings}
+\subsection{Choosing default compressor settings}
 To inform the choice of compression settings across different fields 
@@
 data from the 1000 Genomes project.
+[TODO, reffing Fig~\ref{fig-compression-shuffle} and other fig
+on chunk size.]
 

### 2024-05-10 paper.tex
Para on Zarr's limitations

@@
 than genotype data (which uses $<1\%$ of the total space here)
-because of their inherent noisiness~\citep{lin2020sparse}. By 
-default the \texttt{bio2zarr} conversion is lossless, and integers
-of the narrowest type required to represent all of the data are
-used. In this case, this has resulted in 16 bit integers being used 
-for these top four fields. Examining the distribution of the depth
-fields, (call\_AD, call\_DP, call\_DPF), we can 
-we can see that a tiny fraction of the values stored
-are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
-Fig SX for histograms).
-One might then reasonably truncate these fields to 8 bit integers.
-After converting to 8 bit, call\_AD is reduced by 
-x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
-and call\_DPF by x\% (XX GiB). 
-Similarly, the vast majority of call\_GQ values (xx\%) are less than 
-128 (corresponding to a conditional probability that the genotype call is 
-wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
-be truncated. Here, this results in a reduction of x\% (XX GiB).
-Combined, truncating these top four fields to 8 bits reduces the 
-overall storage of the dataset to XXGiB, corresponding to a 
-X-fold compression of the VCF.
+because of their inherent noisiness~\citep{lin2020sparse}.
+
+[TODO add in benchmarks discussed in issue. Extract POS, run af-dist, 
+do a filtering query. Note problems with orchestration when
+dealing with multiple VCF chunks. 
+Talk about mask-oriented computing there at the
+end, pointing out the cost of these multiple copies.]
+
+% By default the \texttt{bio2zarr} conversion is lossless, and integers
+% of the narrowest type required to represent all of the data are
+% used. In this case, this has resulted in 16 bit integers being used 
+% for these top four fields. Examining the distribution of the depth
+% fields, (call\_AD, call\_DP, call\_DPF), we can 
+% we can see that a tiny fraction of the values stored
+% are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
+% Fig SX for histograms).
+% One might then reasonably truncate these fields to 8 bit integers.
+% After converting to 8 bit, call\_AD is reduced by 
+% x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
+% and call\_DPF by x\% (XX GiB). 
+% Similarly, the vast majority of call\_GQ values (xx\%) are less than 
+% 128 (corresponding to a conditional probability that the genotype call is 
+% wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
+% be truncated. Here, this results in a reduction of x\% (XX GiB).
+% Combined, truncating these top four fields to 8 bits reduces the 
+% overall storage of the dataset to XXGiB, corresponding to a 
+% X-fold compression of the VCF.
 
@@
 
-Calculating the histogram of per call values at this 
-scale is not a trivial task.
-Taking advantage of the rich software ecosystem growing around Zarr
-(see Discussion) we used xarray~\citep{hoyer2017xarray}
-and Dask~\citep{rocklin2015dask} to transparently distribute the 
-computations over a cluster of X workers. 
-[ADD DETAILS]
-The elapsed time was X minutes, and requires X lines of Python in a 
-Jupyter notebook~\citep{kluyver2016jupyter}.
-
-One of the key uses of the aggV2 dataset is to enable simple queries 
-of specific regions. For example, in [A PAPER] they [DID SOMETHING].
-We replicated this by running the appropriate bcftools query
-on the VCF chunk (note that the user is responsible for finding 
-the correct chunk file here, limiting the utility of genomic 
-indexes). This took XXX hours.
-The same query implemented with the Zarr-python 
-API required XXX seconds to produce identical output.
+% Calculating the histogram of per call values at this 
+% scale is not a trivial task.
+% Taking advantage of the rich software ecosystem growing around Zarr
+% (see Discussion) we used xarray~\citep{hoyer2017xarray}
+% and Dask~\citep{rocklin2015dask} to transparently distribute the 
+% computations over a cluster of X workers. 
+% [ADD DETAILS]
+% The elapsed time was X minutes, and requires X lines of Python in a 
+% Jupyter notebook~\citep{kluyver2016jupyter}.
+
+% One of the key uses of the aggV2 dataset is to enable simple queries 
+% of specific regions. For example, in [A PAPER] they [DID SOMETHING].
+% We replicated this by running the appropriate bcftools query
+% on the VCF chunk (note that the user is responsible for finding 
+% the correct chunk file here, limiting the utility of genomic 
+% indexes). This took XXX hours.
+% The same query implemented with the Zarr-python 
+% API required XXX seconds to produce identical output.
 
 \section{Discussion}
-% The introduction of VCF as part of the work of the 1000 Genomes
-% project~\cite{10002015global} has transformed a fragmented 
-% landscape of mutually incompatible per-program file formats
-% into an ecosystem of tools that largely
-% interoperate~\cite{garrison2022spectrum}. In this 
-% sense VCF has been a huge success, and the underlying data model
-% is now deeply embedded in bioinformatics practice. As we have 
-% illustrated here, the shortcomings of the VCF and BCF file
-% formats (and more generally \emph{any} row-wise encoding of the 
-% data model) in terms of computational accessibility are severe,
-% and there can be little doubt that these file formats 
-% are incurring a large economic cost and 
-% holding back progress. 
-% As we have illustrated in the Genomics England
-% example here, these shortcomings are most acutely felt in 
-% large-scale data human datasets but any setting in which we 
-% wish to extract columns of the data from a row-wise 
-% VCF format will be inefficient 
-% when datasets are larger than a few hundred megabytes.
-
-% As we have demonstrated, the columnar approach, based on the 
-% popular Zarr standard and widely used compression libraries, 
-% can support many different types of data access
-% patterm efficiently, and has excellent computational accessibility.
-% In this approach there are some significant departures from 
-% standard practice, which warrant some discussion here.
+VCF is a central element of modern genomics, facilitating
+the exchange of data among interoperating tools. 
+Its current row-oriented form, however,
+is fundamentally inefficient,
+profoundly limiting the scalability of the current generation
+of bioinformatics tools. Large scale VCF data cannot 
+currently be 
+processed without incurring a substantial economic
+and likely environmental cost~\cite{grealey2022carbon}.
+We have shown here that this is not a necessary situation,
+and that greatly improved efficiency can be achieved by 
+using more appropriate storage representations tuned
+to the realities of modern computing. We have argued that 
+Zarr provides a powerful basis for cloud-based
+storage and analysis of large-scale genetic variation data.
+We propose the VCF Zarr specification which losslessly
+maps VCF data to Zarr, and provide an efficient and scalable
+tool to perform conversion.
+
+Zarr provides pragmatic solutions to some of the more pressing 
+problems facing the analysis of large-scale genetic variation
+data, but it is not a solution to all problems. Firstly, 
+any dataset containing a variant with a large number of alleles
+(perhaps due to indels) will cause problems because the 
+dimensions of various fields are determined by the \emph{maximum}
+dimension among all variants. Secondly, the design of 
+VCF Zarr emphasises efficiency of analysis for a fixed 
+dataset, and does not consider problems of ongoing updates
+(i.e, as new samples and their variants are added).
+Thirdly, Zarr is not suitable for handling unstructured
+string data, and for example likely not work well as a 
+basis for storing annotation data [CITES].
 
@@
 
-% Probably getting into the weeds too much.
-% Some large cloud-hosted VCF datasets are currently stored in 
-% filesystems that emulate POSIX semantics on top of such object 
-% stores, at a substantial cost in performance. Directly hosting 
-% these datasets in object stores would have substantial 
-% benefits, both in terms of storage costs and performance.
-% Thirdly, breaking up fields into different directories (or 
-% cloud buckets) enables fine-grained access control to the 
-% dataset. For example, all users may be allowed access to
-% variant-level data, but only a subset permitted to access
-% call-level individual data. Such access controls can be 
-% straightforwardly applied using file or cloud-hosting 
-% methods on the relevant directores/buckets. Such concerns
-% are entirely decoupled from the file format.
-
-% This is poorly explained
-Along with the move away from the single-file-storage paradigm,
-the approach we outline here also suggests a shift from 
-the classical Unix pipeline-based processing approach 
-(but see below for a discussion of how existing pipelines can
-be supported). Datasets flowing through pipelines inherently 
-implies creating muliple copies of the data, and when 
-multiple terabytes of either unrelated or unchanged data 
-get copied several times through these pipes large inefficiencies 
-are incurred (see Fig~\ref{fig-whole-matrix-compute}). 
-We may think of this strategy as ``copy-oriented'' workflows.
-A more efficient way to structure such multilayered computations
-is in a  ``mask-oriented'' way. For example, to implement some
-complex multilayered call filtering strategy we could imagine sequentially
-updating a given per-call mask. Masks are cheap (e.g., the 
-standard genotype missingness mask in Table~\ref{tab-genomics-england-data}
-uses less than 1\% of the overall storage).
-
-Zarr provides pragmatic solutions to some of the more pressing 
-problems facing the analysis of large-scale genetic variation
-data, but it is not a solution to all problems. [large number
-of alleles, dataset is basically static, ragged arrays,
-strings ... any others? Zarr v3 on the way.]
-
 Nonetheless, at least in the single-shared-copy model currently
@@
 conversion.
-
-% Zarr has also seen recent uptake 
-% in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
-% and multimodal spatial omics
-% data~\citep{marconato2024spatialdata,baker2023emobject}.
-% Recent additions using Zarr include the application of deep learning
-% models to genomic sequence data~\citep{klie2023predictive} 
-% and a browser for genetic variation data~\citep{konig2023divbrowse}.
 

### 2024-05-10 paper.tex
More discussion

@@
 \section{Discussion}
+% Zarr is great
 VCF is a central element of modern genomics, facilitating
@@
 
+% We're not pretending is a solution to all problems and
+% up front about weaknesses of the model.
 Zarr provides pragmatic solutions to some of the more pressing 
@@
 
-% Perhaps the largest deviation in the approach that we advocate here 
-% from standard practice is to decouple our concept of a dataset from
-% a file. Each VCF is broken up into a set of Zarr arrays, each 
-% corresponding to a large number of chunks stored in a directory
-% hierarchy (Zarr also supports other forms of chunk storage 
-% including single files and key-value stores however; see Methods).
-% While there are some disadvantages to large numbers of small files,
-% we argue that the benefits greatly outweigh them, particularly 
-% in the large, centrally controlled dataset setting. 
-% Firstly, this approach is inherently parallelisable, and chunks across the 
-% dataset written and read in parallel without needing complex 
-% synchronisation and prior layout planning. 
-% Such parallel write access is very difficult 
-% in the single-file setting, and one of the key weaknesses 
-% of the HDF5 format~\citep{folk2011overview} [CITATION with 
-% parallel troubles from HDF5?]
-% Secondly, mapping chunks in arrays to individually addressed
-% keys maps well to large-scale cloud object storage. Indeed,
-% Zarr is specifically designed for such usage, and 
-% several petabyte scale datasets have been successfully deployed
-% and used [citations?].
-% We have focused here on the classical single-server and 
-% HPC cluster settings, but a major advantage of Zarr is 
-% its cloud readiness.
-
-Nonetheless, at least in the single-shared-copy model currently
-used by most large-scale variation datasets, Zarr provides
-an excellent basis for building a new generation of tools 
-that compute \emph{directly} on the data, efficiently
-accessing data by variant and by sample.
-[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
+Nonetheless, there are numerous datasets that exist today
+what would benefit from cloud-native storage, 
+inherent parallelism and granular storage.
+[Mention benefits of fine-grained access control to 
+different parts of the dataset using native storage access
+control mechanism?]
 
@@
 in a reasonable amount of time. [Dask, Xarray, sgkit?]
-Currently users are responsible for orchestrating such calculations
-on clusters, and automation would be a significant boon
-for reasearcher productivity. [Hail has demonstrated the value of this?]
-
-% We suggest that the future of VCF is to explicitly define the underlying
-% data model, and to decouple it from the current row-wise encodings 
-% of the VCF text and BCF binary file formats. This will enable data 
-% to flow more freely, and provide scope for much-needed innovation.
-
-% \item A major strong point of Zarr is its simplicity. All it's doing
-% is compressing n-dimensional chunks of numerical data and storing
-% them using fixed address keys. This key-value approach means that storage
-% is very flexible and implementation is straightforward. There
-% are several implementations of the Zarr protocol, and while some of these
-% are rudimentary, they could be improved or replaced with relatively
-% little effort. Zarr has been successfully used to store petabytes
-% of data across different sciences --- genetic variation data is largely
-% arrays of numerical data, and not really that different ultimately.
-% Zarr is cloud native.
-% Zarr is an open specification, built on standard components (contrast
-% with Hail).
 
@@
 conversion.
-
-% Workflow engines such as
-% Snakemake~\cite{koster2012snakemake,molder2021sustainable}
-% and [TODO mention WDL, Cromwell, etc?] make such analyses
-% much easier, but there is still a great deal of room for
-% mistakes.
-% Per-chromosome VCFs are not practical for biobank scale
-% projects, however, and variant call data tends to be split
-% into many (still large) files.
-% For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
-% are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
-% hundreds of files per chromosome.
-% % This is horrible writing, but the point is important.
-% While these files provide a natural
-% unit for distributing work, the details of how they
-% are split differ and essentially there is
-% no interoperability among large-scale datasets because the VCFs are
-% so cumbersome that a separate orchestration layer is required to
-% access the data in parallel.
 

### 2024-05-21 src/plot.py
Add 2-bit encoding to data scaling fig
Closes #118

@@ -19,6 +19,7 @@ sav_colour = "tab:red"
 genozip_colour = "tab:purple"
 zarr_colour = "tab:blue"
 zarr_nshf_colour = "tab:cyan"
+two_bit_colour = "tab:pink"
 
 
 def one_panel_fig(**kwargs):
@@ -38,6 +39,7 @@ def two_panel_fig(**kwargs):
 
 def plot_size(ax, df, label_y_offset=None):
     colour_map = {
+        "2bit": two_bit_colour,
         "vcf": vcf_colour,
         "bcf": bcf_colour,
         "zarr": zarr_colour,
@@ -61,8 +63,12 @@ def plot_size(ax, df, label_y_offset=None):
         )
         row = dfs.iloc[-1]
         size = row["size"] / GB
+        label = f"{size:.0f}G"
+        if size > 100:
+            size /= 1024
+            label = f"{size:.1f}T"
         ax.annotate(
-            f"{size:.0f}G",
+            label,
             textcoords="offset points",
             xytext=(15, label_y_offset.get(tool, 0)),
             xy=(row.num_samples, row["size"]),
@@ -76,7 +82,7 @@ def plot_size(ax, df, label_y_offset=None):
     ax.legend()
     add_number_of_variants(df, ax)
     ax.set_xlabel("Number of samples")
-    ax.set_ylabel("Time (seconds)")
+    ax.set_ylabel("Storage size (bytes)")
     plt.tight_layout()
 
 
@@ -189,13 +195,23 @@ def data_scaling(size_data, output):
     """
     df1 = pd.read_csv(size_data, index_col=None).sort_values("num_samples")
 
-    fig, ax1 = one_panel_fig()
-    plot_size(ax1, df1, label_y_offset={"vcf": 4, "sav": -5.5, "genozip": -7})
-
     sav = df1[df1.tool == "sav"]
     zarr = df1[df1.tool == "zarr"]
     ratio = sav["size"].values / zarr["size"].values
     print("sav / zarr ratio:", ratio)
+
+    plink_ish = []
+    for _, row in zarr.iterrows():
+        d = dict(row)
+        d["tool"] = "2bit"
+        d["size"] = row.num_samples * row.num_sites / 4
+        plink_ish.append(d)
+
+    df1 = pd.concat([df1, pd.DataFrame(plink_ish)])
+
+    fig, ax1 = one_panel_fig()
[…]

### 2024-05-21 paper.tex

@@
 data, but file sizes are substantially larger than 
-the equivalent compressed VCF. 
+the equivalent compressed VCF (see Fig~\ref{fig-data-storage}).
 For example, at two bits 
-per genotype, the full genotype matrix for the GraphTyper SNP dataset 
+per diploid genotype, the full genotype matrix for the GraphTyper SNP dataset 
 in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
@@
 for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
+Also shown for reference is the size of genotype matrix 
+when encoded as two bits per diploid genotype (2bit), as used 
+in the PLINK binary format.
 \label{fig-data-storage}}
@@
 data from the 1000 Genomes project.
-[TODO, reffing Fig~\ref{fig-compression-shuffle} and other fig
-on chunk size.]
+[TODO, reffing Fig~\ref{fig-compression-shuffle} and 
+Fig~\ref{fig-compression-chunksize}.]
 
@@
 
+\begin{figure}
+\includegraphics{figures/compression-chunksize}
+\caption{Effects of chunk sizes on compression ratios on 1000 Genomes
+data. TODO MORE DETAIL. 
+\label{fig-compression-chunksize}}
+\end{figure}
+
 \end{document}

### 2024-05-21 paper.tex
Update table for new chr2 data

@@
 \midrule
-call\_AD &  int16 & 179.7G & 26 & 24.0\\
-call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
-call\_DP &  int16 & 141.8G & 16 & 18.9 \\
-call\_DPF& int16  & 115.1G & 20 & 15.3\\
-call\_FT &  string & 58.5G & 160 & 7.8 \\
-call\_PL &  int16 & 51.4G & 140 & 6.9 \\
-call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
-call\_genotype & int8 & 6.1G & 380 & 0.8 \\
-call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
-call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
-call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
-variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
-variant\_allele & string & 21.5M & 11 \\
-variant\_position & int32 & 12.7M & 4.7 \\
-variant\_ABratio & float32 & 9.6M & 6.3 \\
-variant\_AN & int32 & 9.6M & 6.3 \\
-variant\_filter & bool & 1.9M & 490 \\
+/call\_AD & int16 & 658.4G & 26.0 & 25.35\% \\
+/call\_GQ & int16 & 654.5G & 13.0 & 25.20\% \\
+/call\_DP & int16 & 570.0G & 15.0 & 21.95\% \\
+/call\_DPF & int16 & 447.1G & 20.0 & 17.22\% \\
+/call\_PL & int16 & 162.6G & 160.0 & 6.26\% \\
+/call\_GQX & int16 & 41.0G & 210.0 & 1.58\% \\
+/call\_FT & string & 25.0G & 1400.0 & 0.96\% \\
+/call\_genotype & int8 & 21.5G & 410.0 & 0.83\% \\
+/call\_genotype\_mask & bool & 12.8G & 680.0 & 0.49\% \\
+/call\_genotype\_phased & bool & 2.4G & 1900.0 & 0.09\% \\
+/call\_PS & int8 & 383.4M & 12000.0 & 0.01\% \\
+% /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
+% /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
+/variant\_position & int32 & 111.6M & 2.0 & \\
+/variant\_quality & float32 & 87.4M & 2.6 & \\
+/variant\_allele & string & 69.3M & 13.0 & \\
+% /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
+/variant\_AN & int32 & 47.3M & 4.8 & \\
+% /variant\_AC & int32 & 47M & 4.9 & 0.00\% \\
+% /variant\_AC\_Het & int32 & 46.62M & 4.9 & 0.00\% \\
+% /variant\_ABratio & float32 & 20.23M & 11.0 & 0.00\% \\
+% /variant\_MendelSite & float32 & 20.09M & 11.0 & 0.00\% \\
+% /variant\_medianGQ & float32 & 19.16M & 12.0 & 0.00\% \\
+% /variant\_completeGTRatio & float32 & 14.98M & 15.0 & 0.00\% \\
+% /variant\_phwe\_eur & float32 & 14.46M & 16.0 & 0.00\% \\
+% /variant\_phwe\_afr & float32 & 13.37M & 17.0 & 0.00\% \\
+% /variant\_medianDepthNonMiss & float32 & 12.17M & 19.0 & 0.00\% \\
+% /variant\_medianDepthAll & float32 & 11.99M & 19.0 & 0.00\% \\
+% /variant\_phwe\_sas & float32 & 11.59M & 20.0 & 0.00\% \\
+% /variant\_AC\_Hom & int32 & 11.12M & 21.0 & 0.00\% \\
+% /variant\_missingness & float32 & 9.3M & 25.0 & 0.00\% \\
+% /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
+/variant\_filter & bool & 6.4M & 570.0 & \\
+% /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
+% /variant\_composite\_filter & bool & 433.38 KiB & 130.0 & 0.00\% \\
+% /variant\_OLD\_CLUMPED & object & 384.18 KiB & 1200.0 & 0.00\% \\
+% /variant\_id & object & 298.76 KiB & 1600.0 & 0.00\% \\
+% /variant\_phwe\_amr & float32 & 269.52 KiB & 870.0 & 0.00\% \\
+% /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
+% /sample\_id & object & 268.12 KiB & 2.3 & 0.00\% \\
+% /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
+% /contig\_length & int64 & 3.79 KiB & 5.3 & 0.00\% \\
+% /contig\_id & object & 2.3 KiB & 8.8 & 0.00\% \\
+% /filter\_id & object & 808 bytes & 0.6 & 0.00\% \\
+%%%% OLD for chr20
+% call\_AD &  int16 & 179.7G & 26 & 24.0\\
+% call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
+% call\_DP &  int16 & 141.8G & 16 & 18.9 \\
+% call\_DPF& int16  & 115.1G & 20 & 15.3\\
+% call\_FT &  string & 58.5G & 160 & 7.8 \\
+% call\_PL &  int16 & 51.4G & 140 & 6.9 \\
+% call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
+% call\_genotype & int8 & 6.1G & 380 & 0.8 \\
+% call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
+% call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
+% call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
+% variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
+% variant\_allele & string & 21.5M & 11 \\
+% variant\_position & int32 & 12.7M & 4.7 \\
+% variant\_ABratio & float32 & 9.6M & 6.3 \\
+% variant\_AN & int32 & 9.6M & 6.3 \\
+% variant\_filter & bool & 1.9M & 490 \\
 \bottomrule
@@
 \end{table}
-
-% array([23.956     , 22.90533333, 18.90133333, 15.34533333,  7.8       ,
-%         6.85333333,  1.61066667,  0.81733333,  0.488     ])
-
-% /call_AD                     int16    179.67 GiB  4.53 TiB       26 125768  37.74 MiB     1.46 MiB            (15912293, 78195, 2)  (10000, 1000, 2)  
-% /call_GQ                     int16    171.79 GiB  2.26 TiB       13 125768  18.87 MiB     1.4 MiB             (15912293, 78195)     (10000, 1000) 
-% /call_DP                     int16    141.76 GiB  2.26 TiB       16 125768  18.87 MiB     1.15 MiB            (15912293, 78195)     (10000, 1000)
-% /call_DPF                    int16    115.09 GiB  2.26 TiB       20 125768  18.87 MiB     959.56 KiB          (15912293, 78195)     (10000, 1000)
-% /call_FT                     object   58.5 GiB    9.05 TiB      160 125768  75.48 MiB     487.71 KiB          (15912293, 78195)     (10000, 1000)
-% /call_PL                     int16    51.4 GiB    6.79 TiB      140 125768  56.61 MiB     428.51 KiB          (15912293, 78195, 3)  (10000, 1000, 3)
-% /call_GQX                    int16    12.08 GiB   2.26 TiB      190 125768  18.87 MiB     100.73 KiB          (15912293, 78195)     (10000, 1000) 
-% /call_genotype               int8     6.13 GiB    2.26 TiB      380 125768  18.87 MiB     51.11 KiB           (15912293, 78195, 2)  (10000, 1000, 2)  
-% /call_genotype_mask          bool     3.66 GiB    2.26 TiB      630 125768  18.87 MiB     30.53 KiB           (15912293, 78195, 2)  (10000, 1000, 2)  
-% /call_genotype_phased        bool     692.51 MiB  1.13 TiB     1700 125768  9.43 MiB      5.64 KiB            (15912293, 78195)     (10000, 1000) 
-% /call_PS                     int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000) 
-% /call_ADR                    int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000)
-% /call_ADF                    int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000)
-% /variant_OLD_MULTIALLELIC    object   31.63 MiB   121.4 MiB       3.8 1592  78.09 KiB     20.34 KiB           (15912293,)           (10000,) 
-% /variant_quality             float32  22.6 MiB    60.7 MiB        2.7 1592  39.04 KiB     14.53 KiB           (15912293,)           (10000,) 
-% /variant_allele              object   21.52 MiB   242.8 MiB      11 1592  156.17 KiB    13.84 KiB           (15912293, 2)         (10000, 2) 
-% /variant_position            int32    12.83 MiB   60.7 MiB        4.7 1592  39.04 KiB     8.25 KiB            (15912293,)           (10000,) 
-% /variant_ABratio             float32  9.62 MiB    60.7 MiB        6.3 1592  39.04 KiB     6.18 KiB            (15912293,)           (10000,) 
-% /variant_AN                  int32    9.6 MiB     60.7 MiB        6.3 1592  39.04 KiB     6.18 KiB            (15912293,)           (10000,) 
-% /variant_AC                  int32    9.22 MiB    60.7 MiB        6.6 1592  39.04 KiB     5.93 KiB            (15912293,)           (10000,) 
-% /variant_AC_Het              int32    8.98 MiB    60.7 MiB        6.8 1592  39.04 KiB     5.78 KiB            (15912293,)           (10000,)
-
-% Approx total = 179.67 + 171.79 + 141.76 + 115.09 + 58.5 + 51.4 + 12.08 + 6.13 + 3.66 
-% 740, round up to 750
-% [179.67 , 171.79 , 141.76 , 115.09 , 58.5 , 51.4 , 12.08 , 6.13 , 3.66 ]
-% >>> import numpy as np
-% >>> a = np.array([179.67 , 171.79 , 141.76 , 115.09 , 58.5 , 51.4 , 12.08 ,
-% 6.13 , 3.66 ]  )
-% >>> a
-% array([179.67, 171.79, 141.76, 115.09,  58.5 ,  51.4 ,  12.08,   6.13,
-%          3.66])
-% >>> a / 750 * 100
-% array([23.956     , 22.90533333, 18.90133333, 15.34533333,  7.8       ,
-%         6.85333333,  1.61066667,  0.81733333,  0.488     ])
 

### 2024-05-21 paper.tex
Pass through GEL section

@@
 process to limit memory usage and ensure reliability. We 
-first converted the 28 VCF files (3.51 TiB) for chromosome 20
+first converted the 106 VCF files (12.81 TiB) for chromosome 2
 into the intermediate columnar format (ICF). This task was 
-split into 3899 partitions, and distributed using the Genomics England
-HPC cluster. It required about 50 days of CPU, with an elapsed
-wall-clock time of about 2 hours. The ICF representation uses a total
-of XTiB over XX files and YY directories. We then encoded this intermediate
-columnar format to Zarr using 8 cores and XX GB of RAM in about 12 hours.
-This produced a dataset with 43 arrays, consuming a 
-total of Y GiB of storage over X directories and 
-Y files. This is about a 4.8X reduction over the original VCF. 
+split into 14,605 partitions, and distributed using the Genomics England
+HPC cluster. The average run-time per partition was 20.7 min.
+The ICF representation used a total
+of 9.94 TiB over 3,960,177 data storage files. 
+We then converted the ICF to Zarr [DETAILS].
+This produced a dataset with 44 arrays, consuming a 
+total of 2.54 TiB of storage over 6,312,488
+chunk files. This is a roughly 5X reduction in total storage space over the original VCF. 
 The top fields in terms 
@@
 \caption{Summary for a selection of the largest VCF Zarr columns produced for 
-Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
+Genomics England aggV2 VCFs on chromosome 2 using \texttt{vcf2zarr}
 default settings. Each field is stored independently 
@@
 power-of-two units, and the compression ratio achieved on that array.
-We also show the percentage of the overall storage that each array consumes.
+We also show the percentage of the overall storage that each array consumes
+(omitting values < 0.01\%).
 \label{tab-genomics-england-data}}
@@
 % /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
-% /sample\_id & object & 268.12 KiB & 2.3 & 0.00\% \\
+/sample\_id & str & 268.12K & 2.3 & \\
 % /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
@@
 size is dominated by a few columns with the top four
-(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
-80\% of the total. These fields are much less compressible
+(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for 
+90\% of the total. These fields are much less compressible
 than genotype data (which uses $<1\%$ of the total space here)
 because of their inherent noisiness~\citep{lin2020sparse}. 
-
-[TODO add in benchmarks discussed in issue. Extract POS, run af-dist, 
-do a filtering query. Note problems with orchestration when
-dealing with multiple VCF chunks. 
-Talk about mask-oriented computing there at the
-end, pointing out the cost of these multiple copies.]
-
-% By default the \texttt{bio2zarr} conversion is lossless, and integers
-% of the narrowest type required to represent all of the data are
-% used. In this case, this has resulted in 16 bit integers being used 
-% for these top four fields. Examining the distribution of the depth
-% fields, (call\_AD, call\_DP, call\_DPF), we can 
-% we can see that a tiny fraction of the values stored
-% are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
-% Fig SX for histograms).
-% One might then reasonably truncate these fields to 8 bit integers.
-% After converting to 8 bit, call\_AD is reduced by 
-% x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
-% and call\_DPF by x\% (XX GiB). 
-% Similarly, the vast majority of call\_GQ values (xx\%) are less than 
-% 128 (corresponding to a conditional probability that the genotype call is 
-% wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
-% be truncated. Here, this results in a reduction of x\% (XX GiB).
-% Combined, truncating these top four fields to 8 bits reduces the 
-% overall storage of the dataset to XXGiB, corresponding to a 
-% X-fold compression of the VCF.
-
-% and can reasonably be truncated (a 
-% much information would be lost if we used 8 bit
-% Substantial
-% reductions in space can be achieved, however, if we examine the 
-% distributions of these fields and truncate some outliers.
-% Figure~SX shows the histogram of the XX call\_AD values, and we can
-% see that < 0.0X\% of the values are $>127$. Truncating and 
-% re-encoding the call\_AD array as an int8 results in a total storage of 
-% XX. A substantial advantage of Zarr is that such per-field optimisations
-% can be made after conversion, without affecting the rest of the dataset.
-% As discussed in Methods, judicious use of additional filters
-% and lossy quantizing can result substantial space savings.
-
-% Calculating the histogram of per call values at this 
-% scale is not a trivial task.
-% Taking advantage of the rich software ecosystem growing around Zarr
-% (see Discussion) we used xarray~\citep{hoyer2017xarray}
-% and Dask~\citep{rocklin2015dask} to transparently distribute the 
-% computations over a cluster of X workers. 
-% [ADD DETAILS]
-% The elapsed time was X minutes, and requires X lines of Python in a 
-% Jupyter notebook~\citep{kluyver2016jupyter}.
-
-% One of the key uses of the aggV2 dataset is to enable simple queries 
-% of specific regions. For example, in [A PAPER] they [DID SOMETHING].
-% We replicated this by running the appropriate bcftools query
-% on the VCF chunk (note that the user is responsible for finding 
-% the correct chunk file here, limiting the utility of genomic 
-% indexes). This took XXX hours.
-% The same query implemented with the Zarr-python 
-% API required XXX seconds to produce identical output.
+Note that these top four fields are stored as 16 bit integers
+because they contain rare outliers that cannot be stored as 
+8 bits. While the fields could likely be truncated to have 
+a maximum of 127 with minimal loss of information, 
+the compression gains from doing so are relatively minor,
+and we therefore opt for fully lossless compression here 
+for simplicity.
+
+To demonstrate the computational accessibility of Zarr on this 
+large human dataset, we performed some illustrative benchmarks.
+Firstly, we ran the af-dist calculation (Figs~\ref{fig-whole-matrix-compute}
+and~\ref{fig-subset-matrix-compute}) on one 132GiB VCF file covering
+positions 58,219,159--60,650,943 (562,640 variants) using \texttt{bcftools
++af-dist} as before. The elapsed time for this operation was 716 min
+user+sys time almost exactly equal). We repeated this operation for the 
+same genome coordinates in Zarr (using Python code described in previous sections) 
+% 716 minutes / 4.25 = 168X
+with a total CPU time of 2min 19s and elapsed time of 4min 15s (168X faster).
+Second, we extracted the genome position for each variant in this single VCF 
+chunk using \texttt{bcftools query} and Python Zarr code as described in 
+Fig~\ref{fig-column-extract}. The \texttt{bcftools} command required 
+% def parse_time(time_str):
+%     minsplit = time_str.split("m")
+%     return datetime.timedelta(minutes=int(minsplit[0]),
+%         seconds=float(minsplit[1][:-1]))
+% def parse_time_output(s):
+%     values = {}
+%     for line in s.splitlines():
+%         split = line.split()
+%         values[split[0]] = parse_time(split[1])
+%     return values["real"].seconds / 60, (values["user"] + values["sys"]).seconds / 60
+% real  85m51.013s
+% user  54m39.986s
+% sys 0m45.277s
+55.42 min total CPU (85.85 min elapsed).
+% CPU times: user 846 ms, sys: 1.93 s, total: 2.78 s
+% Wall time: 1min 44s
+% 85.85 / 1.73 = 49.62 = 50X
+The Zarr code required 2.78 sec total CPU and 1.73 min elapsed (50X faster).
+Note that the Zarr code here is heavily I/O bound [TODO say something 
+about the GEL cluster?]. When we subsequently time the operation of 
+just populating an array of the variant positions in specified region,
+this is reduced to 1.13 sec, with an elapsed time of 20.9 sec.
+[Note: this is crazy that we still have a 19 second I/O wait. The
+5989 files for the POS chunks must be in cache already, so the delay
+has to be the file system checking cache coherency.]
+
+Finally, to illustrate performance on a common filtering task, we 
+created a copy of the VCF chunk which contains only variants 
+that pass some common filtering criteria using
+\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
+following standard practices [citations?].
+This took a total of 689 min CPU time, with the same elapsed time.
+In comparison, computing and storing a variant mask (i.e., a boolean value 
+for each variant denoting whether it should be considered or not
+for analysis) based on the same criteria using Zarr 
+required a total of 1.96 min CPU time with an elapsed time 
+of  11 min (63X faster).
+%689 / 11 = 62.6 = 63X
+There is an important distinction here between creating a copy
+of the data (an implicit part of VCF based workflows) and creating an
+an additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
+illustrates, call-level masks are cheap (the standard 
+missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
+storage) and variant or sample level masks requiring negligible storage.
+If downstream software can 
+use configurable masks (at variant, sample and call level) 
+rather than expecting full copies of the data, major storage savings
+and improvements in processing efficiency can be made.
+Such a transition from ``mask-oriented'' vs ``copy-oriented'' 
+analysis is one of the transformational changes enabled by Zarr.
 

### 2024-05-22 src/compression_benchmarks.py
Run Black on benchmarks code

@@ -8,10 +8,9 @@ from tqdm import tqdm
 import argparse
 
 
-
-def generate_chunksize_variations(default_variant_chunksize=10_000,
-                                  default_sample_chunksize=1_000):
-    
+def generate_chunksize_variations(
+    default_variant_chunksize=10_000, default_sample_chunksize=1_000
+):
     variant_chunksizes = [1000, 2000, 5000, 10000, 20000, 50000, None]
     sample_chunksizes = [100, 500, 1000, 1500, 2000, None]
 
@@ -25,9 +24,7 @@ def generate_chunksize_variations(default_variant_chunksize=10_000,
     return list(final_chunksizes)
 
 
-
-def test_compression_ratio_vs_chunksize(z_arr,
-                                        dry_run=False):
+def test_compression_ratio_vs_chunksize(z_arr, dry_run=False):
     """
     Test the impact of chunksize on compression ratio
 
@@ -35,17 +32,15 @@ def test_compression_ratio_vs_chunksize(z_arr,
     :param dry_run: If True, generate the variations table without copying any data.
     """
 
-    chunksize_var = generate_chunksize_variations()    
+    chunksize_var = generate_chunksize_variations()
     comp_results_table = []
 
     if not dry_run:
         # Extract the data
         arr = z_arr[:]
 
-    for cs_var in tqdm(chunksize_var,
-                       total=len(chunksize_var),
-                       desc='Chunksizes'):
-
+    for cs_var in tqdm(chunksize_var, total=len(chunksize_var), desc="Chunksizes"):
+        print(cs_var)
         cs_var = (cs_var[0] or z_arr.shape[0], cs_var[1] or z_arr.shape[1])
 
         variant_cs, sample_cs = cs_var
@@ -54,41 +49,45 @@ def test_compression_ratio_vs_chunksize(z_arr,
             cs_var += (z_arr.shape[2],)
 
         if not dry_run:
-
             if z_arr.filters is not None:
                 object_codec = z_arr.filters[0]
             else:
                 object_codec = None
 
-            z2 = zarr.empty_like(arr,
-                                 chunks=cs_var,
-                                 compressor=z_arr.compressor,
-                                 dtype=z_arr.dtype,
-                                 object_codec=object_codec)
+            z2 = zarr.empty_like(
+                arr,
+                chunks=cs_var,
+                compressor=z_arr.compressor,
+                dtype=z_arr.dtype,
+                object_codec=object_codec,
+            )
 
             z2[:] = arr
 
             n_chunks = z2.nchunks
[…]

### 2024-05-22 src/compression_benchmarks.py
Update spacing on chunk sizes

@@ -8,20 +8,20 @@ from tqdm import tqdm
 import argparse
 
 
-def generate_chunksize_variations(
-    default_variant_chunksize=10_000, default_sample_chunksize=1_000
-):
-    variant_chunksizes = [1000, 2000, 5000, 10000, 20000, 50000, None]
-    sample_chunksizes = [100, 500, 1000, 1500, 2000, None]
+def generate_chunksize_variations(z_arr):
+    default_variant_chunksize = z_arr.chunks[0]
+    default_sample_chunksize = z_arr.chunks[1]
+    variant_chunksizes = np.logspace(2, np.log10(z_arr.shape[0]), num=6).astype(int)
+    sample_chunksizes = np.linspace(100, z_arr.shape[1], num=6).astype(int)
 
-    final_chunksizes = set()
+    final_chunksizes = []
     for vcs in variant_chunksizes:
-        final_chunksizes.add((vcs, default_sample_chunksize))
+        final_chunksizes.append((vcs, default_sample_chunksize))
 
     for scs in sample_chunksizes:
-        final_chunksizes.add((default_variant_chunksize, scs))
+        final_chunksizes.append((default_variant_chunksize, scs))
 
-    return list(final_chunksizes)
+    return final_chunksizes
 
 
 def test_compression_ratio_vs_chunksize(z_arr, dry_run=False):
@@ -32,7 +32,7 @@ def test_compression_ratio_vs_chunksize(z_arr, dry_run=False):
     :param dry_run: If True, generate the variations table without copying any data.
     """
 
-    chunksize_var = generate_chunksize_variations()
+    chunksize_var = generate_chunksize_variations(z_arr)
     comp_results_table = []
 
     if not dry_run:
@@ -40,7 +40,6 @@ def test_compression_ratio_vs_chunksize(z_arr, dry_run=False):
         arr = z_arr[:]
 
     for cs_var in tqdm(chunksize_var, total=len(chunksize_var), desc="Chunksizes"):
-        print(cs_var)
         cs_var = (cs_var[0] or z_arr.shape[0], cs_var[1] or z_arr.shape[1])
 
         variant_cs, sample_cs = cs_var

### 2024-05-22 src/plot.py
Run Black on plot script

@@ -29,6 +29,7 @@ def one_panel_fig(**kwargs):
     fig, ax = plt.subplots(1, 1, figsize=(width, 2 * width / 3), **kwargs)
     return fig, ax
 
+
 def two_panel_fig(**kwargs):
     # The columnwidth of the format is ~250pt, which is
     # 3 15/32 inch, = 3.46
@@ -395,12 +396,16 @@ def compression_chunksize(data, output):
         arr_vdf = variant_df.loc[variant_df.ArrayName == arr].sort_values(
             "variant_chunksize"
         )
-        axes[0].plot(arr_sdf.sample_chunksize, arr_sdf.CompressionRatio, label=arr,
-                marker="o")
-        axes[1].semilogx(arr_vdf.variant_chunksize, arr_vdf.CompressionRatio, label=arr,
-                marker="o")
+        axes[0].plot(
+            arr_sdf.sample_chunksize, arr_sdf.CompressionRatio, label=arr, marker="o"
+        )
+        axes[1].semilogx(
+            arr_vdf.variant_chunksize, arr_vdf.CompressionRatio, label=arr, marker="o"
+        )
 
     plt.legend()
+    axes[0].set_title("(A)")
+    axes[1].set_title("(B)")
     axes[0].set_xlabel("Sample chunk size")
     axes[1].set_xlabel("Variant chunk size")
     axes[0].set_ylabel("Compression ratio")

### 2024-05-22 paper.tex
Some text

@@
 % /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
-/sample\_id & str & 268.12K & 2.3 & \\
+/sample\_id & str & 268.1K & 2.3 & \\
 % /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
@@
 it can then be converted to Zarr using the \texttt{vcf2zarr encode}
-command. We take advantage of Zarr's flexibility
-in \texttt{vcf2zarr} by choosing integer widths based
-
-and reasonable compressor defaults, but these can be
-modified by generating and editing a JSON-formatted storage schema.
+command. By default we choose integer widths based on the maximum
+and minimum values observed during conversion to ICF
+along with reasonable compressor defaults (see next section), 
+but these can be
+modified by generating a JSON-formatted storage schema, 
+which can be edited and supplied as an argument to \texttt{encode}.
 Encoding a given column (for example, \texttt{call\_AD})
@@
 array in question, and then sequentially filling this buffer with 
-values read from ICF and flushing to file. Similar to the \texttt{explode} command,
+values read from ICF and flushing to file.
+ Similar to the \texttt{explode} command,
 encoding to Zarr can be done in parallel on a single 
@@
 The distributed commands are fault-tolerant, reporting any failed
-parititions so that they can be retried.
+partitions so that they can be retried.
 
@@
 in VCF data, we analysed their effect on compression ratio on 
-data from the 1000 Genomes project.
+data from the 1000 Genomes project. We began by downloading 
+the high coverage WGS data for 2504 samples~\citep{byrska2022high}
+for chromosome 22 and converted to Zarr using 
+\texttt{vcf2zarr} with default settings. We then systematically 
+examined the effects of varying chunk sizes and compressor settings
+on the compression ratio for call-level fields. We excluded call\_PL
+from this analysis as it requires conversion to a ``local alleles''
+encoding~\citep{poterba2024scalable} to be efficient, 
+which is planned for implementation in a future 
+version of \texttt{vcf2zarr}.
+
+Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
+fairly weak influence on compression ratio. While larger 
+sample chunk sizes lead to a slight increase in compression
+ratio on the more compressable fields, the overall effect
+on total storage used will be slight. Variant chunk size 
+appears to have no effect on compression ratio. Thus, we 
+do not expect the choice of chunk size to have a significant
+impact on overall storage usage, and so choice should be
+determined by other considerations such as expected data
+access patterns.
+
 [TODO, reffing Fig~\ref{fig-compression-shuffle} and 
 Fig~\ref{fig-compression-chunksize}.]
-
-% [FILL IN WITH Shadi's analysis from here:
-% https://github.com/sgkit-dev/bio2zarr/discussions/74
-% Discussed in:
-% https://github.com/sgkit-dev/vcf-zarr-publication/issues/26
 
@@
 
+% Acking according to the request here:
+% https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/20190405_1000G_2504_high_cov_data_reuse_README.md
+The 1000 Genomes data used to analyse the effects of storage settings
+was generated at the New York Genome Center with funds provided by
+NHGRI Grant 3UM1HG008901-03S1.
+
 \bibliography{paper}
@@
 \begin{figure}
+\includegraphics{figures/compression-chunksize}
+\caption{Effects of chunk sizes on compression ratio on call-level 
+fields in 1000 Genomes data.
+(A) Varying sample chunk size, holding variant chunk size fixed at 10,000.
+(B) Varying variant chunk size, holding sample chunk size fixed at 1,000.
+In all cases the zstd compressor with clevel=7 was used. Bit shuffle
+was used for call\_genotype, and no shuffle used for the other fields.
+\label{fig-compression-chunksize}}
+\end{figure}
+
+\begin{figure}
 \includegraphics{figures/compression-shuffle}
-\caption{Effects of shuffle setting on compression ratios on 1000 Genomes
-data. TODO MORE DETAIL. 
+\caption{Effects of Blosc shuffle settings on compression ratio on call-level 
+fields in 1000 Genomes data. 
+In all cases the zstd compressor with clevel=7 was used, with a variant
+chunk size of 10,000 and sample chunk size of 1,000.
 \label{fig-compression-shuffle}}
@@
 
-\begin{figure}
-\includegraphics{figures/compression-chunksize}
-\caption{Effects of chunk sizes on compression ratios on 1000 Genomes
-data. TODO MORE DETAIL. 
-\label{fig-compression-chunksize}}
-\end{figure}
-
 \end{document}

### 2024-05-22 paper.tex
Write up compression benchmarks

@@
 Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
-fairly weak influence on compression ratio. While larger 
-sample chunk sizes lead to a slight increase in compression
-ratio on the more compressable fields, the overall effect
-on total storage used will be slight. Variant chunk size 
-appears to have no effect on compression ratio. Thus, we 
-do not expect the choice of chunk size to have a significant
+weak influence on compression ratio for most fields. Interestingly,
+the choice of chunk size has a somewhat unpredictable effect
+on compression for the genotype matrix, suggesting that how 
+samples are grouped together has a strong effect on compression.
+At around 50X, compression is excellent in all cases however, 
+and given that genotypes contribute relatively little to the
+overall storage cost in many datasets 
+(Table~\ref{tab-genomics-england-data}), the differences are slight.
+Otherwise, increasing sample chunk size slightly increases 
+compression on call\_AB, and has no effect on less compressible
+fields. Variant chunk size appears to have almost no effect on compression 
+ratio. Thus, we do not expect the choice of chunk size to have a significant
 impact on overall storage usage, and so choice should be
@@
 
-[TODO, reffing Fig~\ref{fig-compression-shuffle} and 
-Fig~\ref{fig-compression-chunksize}.]
+The ``shuffle'' parameter in the Blosc meta-compressor~\cite{alted2010modern}
+can result in substantially better compression, albeit at the 
+cost of somewhat slower decoding (see Fig~\ref{fig-whole-matrix-decode}).
+Fig~\ref{fig-compression-shuffle} shows the effect of 
+the bit shuffle (grouping together bits at the same position
+across bytes before compression), and byte shuffle (grouping together 
+bytes at the sample position across words before compression)
+operations on compression ratio. Bit shuffle provides a significant
+improvement in compression for the call\_genotype field  because 
+the vast majority of genotype calls will be 0 or 1, and therefore 
+bits 1 to 7 will be 0. Thus, grouping these bits together will 
+lead to significantly better compression. This strategy also 
+works extremely well when compressing boolean fields stored as 
+8 bit integers, where the top 7 bits are always 0. In practise,
+boolean fields stored in this way have very similar compression
+to using a bit-packing pre-compression filter (data not shown).
+Although byte shuffle leads to somewhat better compression for call\_AD
+and call\_DP, it gives substantially worse compression on call\_AB
+than no shuffling. 
+The default in \texttt{vcf2zarr}
+is therefore to use bit shuffle for \texttt{call\_genotype}
+and all boolean fields, and to not use 
+byte shuffling on any field. These defaults can be easily
+overruled, however, by outputting and modifying a JSON formatted 
+storage schema before encoding to Zarr.
 
@@
 was used for call\_genotype, and no shuffle used for the other fields.
+Values are chosen to be evenly spaced on a linear scale 
+between 100 and 2504 (the number of samples) in (A) and 
+evenly spaced between 100 and 96514 on a log scale in (B). 
 \label{fig-compression-chunksize}}

### 2024-05-22 paper.tex
Add fig for compression codec

@@
 
+Fig~\ref{fig-compression-compressor} shows the effect of varying compression
+codecs in Blosc. The combination of outstanding compression performance
+and good decoding speed makes zstd a good default choice.
+
 \subsection{Benchmarks}
@@
 (B) Varying variant chunk size, holding sample chunk size fixed at 1,000.
-In all cases the zstd compressor with clevel=7 was used. Bit shuffle
+In all cases the zstd compressor with compression level=7 was used. Bit shuffle
 was used for call\_genotype, and no shuffle used for the other fields.
@@
 fields in 1000 Genomes data. 
-In all cases the zstd compressor with clevel=7 was used, with a variant
+In all cases the zstd compressor with compression level=7 was used, with a variant
 chunk size of 10,000 and sample chunk size of 1,000.
@@
 
+\begin{figure}
+\includegraphics{figures/compression-compressor}
+\caption{Effects of Blosc compression codec on compression ratio on call-level 
+fields in 1000 Genomes data. 
+In all cases compression level=7 was used, with a variant
+chunk size of 10,000 and sample chunk size of 1,000.
+Bit shuffle was used for call\_genotype, and no shuffle used for the other fields.
+\label{fig-compression-compressor}}
+\end{figure}
+
 \end{document}

### 2024-05-22 src/plot.py

@@ -378,6 +378,42 @@ def compression_shuffle(data, output):
 @click.command()
 @click.argument("data", type=click.File("r"))
 @click.argument("output", type=click.Path())
+def compression_compressor(data, output):
+    """
+    Plot figure showing the effect of compressor codec on compression ratio.
+    """
+    df = pd.read_csv(data)
+
+    # Note this is ordered by best-to-worst compression for viz
+
+    arrays = [
+        "call_GQ",
+        "call_DP",
+        "call_AD",
+        "call_AB",
+        "call_genotype",
+    ]
+
+    fig, ax = one_panel_fig()
+    sns.barplot(
+        df,
+        orient="h",
+        order=arrays,
+        y="ArrayName",
+        x="CompressionRatio",
+        hue="Compressor",
+        ax=ax,
+    )
+    ax.set_ylabel("")
+    ax.get_legend().set_title("")
+
+    plt.tight_layout()
+    plt.savefig(output)
+
+
+@click.command()
+@click.argument("data", type=click.File("r"))
+@click.argument("output", type=click.Path())
 def compression_chunksize(data, output):
     """
     Plot figure showing the effect of chunksize settings on compression ratio.
@@ -427,6 +463,7 @@ cli.add_command(subset_matrix_compute)
 cli.add_command(subset_matrix_compute_supplemental)
 cli.add_command(compression_shuffle)
 cli.add_command(compression_chunksize)
+cli.add_command(compression_compressor)
 
 
 if __name__ == "__main__":

### 2024-05-23 paper.tex
Tweaks

@@
 
-VCF is row-wise format in which 
+VCF is a row-wise format in which 
 observations and metadata for a single variant are
@@
 Row-wise storage makes retrieving all information
-for a given record straightforward and efficient. 
-The row-wise approach works well when records are relatively small, 
+for a given record straightforward and efficient,
+and works well when records are relatively small, 
 or when we typically want to analyse each record in its entirity.
@@
 column for the genotypes (and other per-call QC metrics) 
+for each sample
 leads to millions of columns, and substantial technical challenges.
-Additionally, compression performance within columns is poor 
-because there is much less similarity among
+Additionally, compression performance within columns tends to
+be poor because there is much less similarity among
 the genotypes for a given sample than there is among the genotypes
-for a given variant. [TODO check this - reference?]
-[Note this is the approach used in OpenCGA, as shown in their storage 
-benchmark numbers]
+for a given variant, requiring
+specialised haplotype compression 
+techniques~\citep{layer2016efficient,mcvean2019linkage}.
 
@@
 In the VCF Zarr specification, 
-each field in a VCF is mapped to an array, 
-which is stored individually allowing for efficient retrieval and 
+each field in a VCF is mapped to a separately-stored array, 
+allowing for efficient retrieval and 
 high levels of compression.
@@
 The simulations only contain genotypes without any additional
-high-entropy QC fields, which is unrealistic (see Section XXX
+high-entropy QC fields, which is unrealistic 
+(see the Genomics England case-study
 for benchmarks on a large human dataset that includes 
@@
 Fig~\ref{fig-data-storage} shows compression performance 
-on up to a million samples for chromosome 21.
-Unsurprisingly, gzip compressed VCF requires the most 
-storage, with BCF providing substantially better compression
-performance (note the log-log scale). Genozip has 
+on up to a million samples for chromosome 21, with 
+the size of the genotype-matrix encoded as 1-bit per haploid
+call included for reference.
+% 2bit  1689.153258 G
+% vcf   81.375831 G
+% = 20.75 X, 4.8%
+Gzip compressed VCF performs remarkably well, compressing 
+the data to around 5\% of the 
+minimal binary encoding of a biallelic genotype matrix
+for 1 million samples. 
+BCF provides a significant improvement in compression
+performance over VCF (note the log-log scale). Genozip has 
 superb compression, having far smaller file sizes that the 
@@
 using the Zstandard compressor~\citep{collet2021rfc} and the 
-bit-shuffle filter from Blosc~\cite{alted2010modern} produces 
+bit-shuffle filter from Blosc~\cite{alted2010modern} 
+(see Methods for details) produces 
 compression levels competitive with the highly specialised methods
@@
 in VCF data, we analysed their effect on compression ratio on 
-data from the 1000 Genomes project. We began by downloading 
-the high coverage WGS data for 2504 samples~\citep{byrska2022high}
-for chromosome 22 and converted to Zarr using 
+recent high-coverage WGS data from the 1000 Genomes 
+project~\citep{byrska2022high}. We began by downloading 
+the first 100,000 lines of the VCF 
+for chromosome 22 (giving a 1.1GiB compressed VCF)
+and converted to Zarr using 
 \texttt{vcf2zarr} with default settings. We then systematically 
@@
 
-The ``shuffle'' parameter in the Blosc meta-compressor~\cite{alted2010modern}
+The \texttt{shuffle} parameter in the Blosc meta-compressor~\cite{alted2010modern}
 can result in substantially better compression, albeit at the 
@@
 Fig~\ref{fig-compression-shuffle} shows the effect of 
-the bit shuffle (grouping together bits at the same position
+bit shuffle (grouping together bits at the same position
 across bytes before compression), and byte shuffle (grouping together 
 bytes at the sample position across words before compression)
-operations on compression ratio. Bit shuffle provides a significant
+on compression ratio. Bit shuffle provides a significant
 improvement in compression for the call\_genotype field  because 
@@
 lead to significantly better compression. This strategy also 
-works extremely well when compressing boolean fields stored as 
+works well when compressing boolean fields stored as 
 8 bit integers, where the top 7 bits are always 0. In practise,
@@
 
-% Acking according to the request here:
-% https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/20190405_1000G_2504_high_cov_data_reuse_README.md
-The 1000 Genomes data used to analyse the effects of storage settings
-was generated at the New York Genome Center with funds provided by
-NHGRI Grant 3UM1HG008901-03S1.
-
 \bibliography{paper}

### 2024-05-23 paper.tex
Pass through results benchmarks

@@
 which calculations can be performed under the different formats,
-we use the \texttt{bcftools +af-dist} plugin as an example.
+we use the \texttt{bcftools +af-dist} plugin
+(which computes a table of
+deviations from Hardy-Weinberg expectations in 
+allele frequency bins) as an example.
 % The details of the \texttt{af-dist} operation are not important:
@@
 without too much effort.
-Secondly, it produces a small volume of output (a table of
-deviations from Hardy-Weinberg expectations in ten allele frequency
-bins) and therefore the time spent outputting results is negligible.
+Secondly, it produces a small volume of output  and therefore the 
+time spent outputting results is negligible.
 Finally, it has an efficient implementation written using the 
-\texttt{htslib} C API~\citep{bonfield2021htslib}, distributed as 
-a \texttt{bcftools} plugin.
+\texttt{htslib} C API~\citep{bonfield2021htslib},
+and therefore running this command on a VCF or BCF file provides 
+a reasonable approximation of the limit of what can be achieved in terms 
+of whole-matrix computation on these formats.
 
@@
 difference in the time required (note the log-log scale). 
-The slowest approach uses Genozip. Because genozip does not
+The slowest approach uses Genozip. Because Genozip does not
 provide an API and only outputs VCF text, the best approach available 
-is to pipe its output into \texttt{bcftools +af-dist}. Thus,
-not only must we decode the data from the genozip format, we must 
-then parse the large volumes of VCF text produced and then finally do the 
-actual calculation. 
+is to pipe its output into \texttt{bcftools +af-dist}. 
+This involves first decoding the data from Genozip format,
+then generating large volumes  of VCF text (terabytes, in the 
+largest examples here), which we must 
+subsequently parse before finally doing the actual calculation. 
 Running \texttt{bcftools +af-dist} directly on the gzipped VCF

### 2024-05-23 paper.tex

@@
 Total CPU time required to run the af-dist calculation for
-a subset of 10000 variants $\times$ 10 samples from the middle of the matrix
+a subset of 10,000 variants $\times$ 10 samples from the middle of the matrix
 for the data in Fig~\ref{fig-data-storage}.
@@
 
-We illustrate the cost of row-wise encoding in
+We illustrate this cost of row-wise encoding in
 Fig~\ref{fig-subset-matrix-compute}, where we run the af-dist calculation
@@
 Fig~\ref{fig-data-storage}. The two-dimensional chunking of Zarr
-means that this small subset of the overall matrix can be efficiently
+means that this sub-matrix can be efficiently
 extracted, and therefore the execution time depends very weakly on 
-the overall dataset size, with the overall computation requiring around
+the overall dataset size, with the computation requiring around
 1 second for 1 million samples. Because of their 
-row-wise encoding for all the other methods CPU time
-scales with the number of samples.
+row-wise encoding, CPU time scales with the number of samples
+for all the other methods.
 Fig~\ref{fig-subset-matrix-compute-supplemental} shows performance
@@
 For the BCF file we used \texttt{bcftools query -f"\%POS$\backslash$n"}.
-For sav, we used the Savvy C++ API to extract just the position, 
+For sav, we used the Savvy C++ API to extract position for each variant
 and output text using the \texttt{std::cout} stream. For Zarr, we read 
-\texttt{variant\_position} into a NumPy array, and then wrote to
+the variant\_position array into a NumPy array, and then wrote to
 a text file using the Pandas \texttt{write\_csv} method. 
@@
 The dataset comprises approximately 722 million annotated single-nucleotide 
-variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
+variants and small indels split into 1,371 roughly equal chunks and 
 totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
 The dataset is used for a variety of research purposes, ranging from 
-genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
+GWAS~\cite{kousathanas2022whole} and 
 imputation~\cite{shi2023genomics} to 
@@
 As described in the Methods, conversion to Zarr using 
-\texttt{vcf2zarr} is a two-step 
-process to limit memory usage and ensure reliability. We 
+\texttt{vcf2zarr} is a two-step process. We 
 first converted the 106 VCF files (12.81 TiB) for chromosome 2
@@
 \midrule
-/call\_AD & int16 & 658.4G & 26.0 & 25.35\% \\
-/call\_GQ & int16 & 654.5G & 13.0 & 25.20\% \\
-/call\_DP & int16 & 570.0G & 15.0 & 21.95\% \\
-/call\_DPF & int16 & 447.1G & 20.0 & 17.22\% \\
-/call\_PL & int16 & 162.6G & 160.0 & 6.26\% \\
-/call\_GQX & int16 & 41.0G & 210.0 & 1.58\% \\
-/call\_FT & string & 25.0G & 1400.0 & 0.96\% \\
-/call\_genotype & int8 & 21.5G & 410.0 & 0.83\% \\
-/call\_genotype\_mask & bool & 12.8G & 680.0 & 0.49\% \\
-/call\_genotype\_phased & bool & 2.4G & 1900.0 & 0.09\% \\
-/call\_PS & int8 & 383.4M & 12000.0 & 0.01\% \\
+/call\_AD & int16 & 658.4G & 26 & 25.35\% \\
+/call\_GQ & int16 & 654.5G & 13 & 25.20\% \\
+/call\_DP & int16 & 570.0G & 15 & 21.95\% \\
+/call\_DPF & int16 & 447.1G & 20 & 17.22\% \\
+/call\_PL & int16 & 162.6G & 160 & 6.26\% \\
+/call\_GQX & int16 & 41.0G & 210 & 1.58\% \\
+/call\_FT & string & 25.0G & 1400& 0.96\% \\
+/call\_genotype & int8 & 21.5G & 410& 0.83\% \\
+/call\_genotype\_mask & bool & 12.8G & 680& 0.49\% \\
+/call\_genotype\_phased & bool & 2.4G & 1900& 0.09\% \\
+/call\_PS & int8 & 383.4M & 12000& 0.01\% \\
 % /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
 % /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
-/variant\_position & int32 & 111.6M & 2.0 & \\
+/variant\_position & int32 & 111.6M & 2 & \\
 /variant\_quality & float32 & 87.4M & 2.6 & \\
-/variant\_allele & string & 69.3M & 13.0 & \\
+/variant\_allele & string & 69.3M & 13 & \\
 % /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
@@
 % /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
-/variant\_filter & bool & 6.4M & 570.0 & \\
+/variant\_filter & bool & 6.4M & 570 & \\
 % /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
@@
 and we therefore opt for fully lossless compression here 
-for simplicity.
+for simplicity. The call\_PS field here has an extremely high
+compression ratio because it consists entirely of missing data
+(i.e., it was listed in the header but never used in the VCF).
 
@@
 of the data (an implicit part of VCF based workflows) and creating an
-an additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
+additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
 illustrates, call-level masks are cheap (the standard 
@@
 and improvements in processing efficiency can be made.
-Such a transition from ``mask-oriented'' vs ``copy-oriented'' 
-analysis is one of the transformational changes enabled by Zarr.
+The transition from the manifold inefficiences of 
+present-day ``copy-oriented'' computing,
+to the ``mask-oriented'' analysis of large immutable, single-source
+datasets is a potentially transformational change enabled by Zarr.
 

### 2024-05-23 paper.tex
Add in bit about annotations
Closes #71
Closes #72
Closes #73

@@
 in here.]
+Methods have also been developed which store variation data 
+along with annotations in databases to facilitate
+efficient queries~\cite[e.g.][]{
+paila2013gemini,%GEMINI
+lopez2017hgva} %OpenCGA
+which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
+or require storage larger 
+than uncompressed VCF~\citep{al2023critical}.
 
@@
 illustrates, call-level masks are cheap (the standard 
-missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
-storage) and variant or sample level masks requiring negligible storage.
+genotype missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
+storage) and variant or sample level masks require negligible storage.
 If downstream software can 
@@
 VCF is a central element of modern genomics, facilitating
-the exchange of data among interoperating tools. 
+the exchange of data in a large ecosystem of interoperating tools. 
 Its current row-oriented form, however,
@@
 problems facing the analysis of large-scale genetic variation
-data, but it is not a solution to all problems. Firstly, 
+data, but it is not a panacea. Firstly, 
 any dataset containing a variant with a large number of alleles
 (perhaps due to indels) will cause problems because the 
-dimensions of various fields are determined by the \emph{maximum}
-dimension among all variants. Secondly, the design of 
+dimensions of fields are determined by their \emph{maximum}
+dimension among all variants. In particular this is problematic
+for fields like PL in which the dimension depends 
+quadratically on the number of alleles (although practical
+solutions have been suggested that can be 
+adapted~\cite{poterba2024scalable}).
+Secondly, the design of 
 VCF Zarr emphasises efficiency of analysis for a fixed 
-dataset, and does not consider problems of ongoing updates
-(i.e, as new samples and their variants are added).
-Thirdly, Zarr is not suitable for handling unstructured
-string data, and for example likely not work well as a 
-basis for storing annotation data [CITES].
+dataset, and does not consider how samples (and the 
+corresponding novel variants) should be added.
+Thirdly, Zarr is not suitable for handling 
+unstructured data often included in VCF INFO 
+fields, and would likely be a poor choice
+for representing variant 
+annotations, for example.
 

### 2024-05-23 paper.tex
Eric's changes

--- 2024-05-23 paper.tex
@@
 Row-based data storage is fundamentally unsuited to such datasets
-and these large VCFs represent a major computational bottleneck.
+and a more scalable, grid-based solution is needed.
 
@@
 of object storage. 
-These multiple layers of inefficiences around processing
+These multiple layers of inefficiencies around processing
 VCF data at scale in the cloud mean that it is 
@@
 While Hail is built on open components
-from the Hadoop distributed computing ecosytem~\citep{white2012hadoop},
+from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
 it is complex, vertically integrated software
 with custom storage formats that are not intended for reuse.
-Similarly, commerical solutions that have emerged to facilitate
+Similarly, commercial solutions that have emerged to facilitate
 the analysis of large-scale genetic variation data are either
@@
 with various benchmarks being substantially faster than
-\texttt{bcftools} based pipelines, and again competetive
+\texttt{bcftools} based pipelines, and again competitive
 with state-of-the-art file-oriented methods. Finally, we show the 
@@
 and works well when records are relatively small, 
-or when we typically want to analyse each record in its entirity.
+or when we typically want to analyse each record in its entirety.
 When we want to analyse only a subset of a record,
@@
 success~\citep{boufea2017managing,fan2020variant}.
-Columnar techologies tend to be 
+Columnar technologies tend to be 
 optimised for ``tall and narrow'' data where there is a large number of records
@@
 in the same memory space without additional copies and 
-interprocess communication through pipes.
+inter-process communication through pipes.
 
@@
 and improvements in processing efficiency can be made.
-The transition from the manifold inefficiences of 
+The transition from the manifold inefficiencies of 
 present-day ``copy-oriented'' computing,
@@
 scale concerns distributed computation. Even with more efficient
-columnar data representations, computations that required examing 
+columnar data representations, computations that required examining 
 large volumes of per-call data must be distributed to complete 
@@
 In the interest of completeness it is useful to provide a high-level overview
-of Zarr and the techologies that it depends upon. Zarr is a specialised format
+of Zarr and the technologies that it depends upon. Zarr is a specialised format
 for storing large-scale $n$-dimensional data (arrays). Arrays

### 2024-05-23 paper.tex
wip intro edits

@@
 \item VCF is widely supported, and the underlying data model entrenched 
-in Bioinformatics pipelines.
+in bioinformatics pipelines.
 \item The standard row-wise encoding as text (or binary) is inherently
@@
 collecting genome sequence data for hundreds of thousands of humans.
+Larger genome variation datasets are also being generated for other organisms
+and a variety of purposes including agriculture~\citep{wang2023rice}, conservation~\citep{shaffer2022landscape}
+and infectious disease surveillance~\citep{TODO}.
 VCF's simple text-based design and widespread

### 2024-05-23 paper.tex
add ref

--- 2024-05-23 paper.tex
@@
 collecting genome sequence data for hundreds of thousands of humans.
+Larger genome variation datasets are also being generated for other organisms
+and a variety of purposes including agriculture~\citep{wang2023rice}, conservation~\citep{shaffer2022landscape}
+and infectious disease surveillance~\citep{hamid2023pf7}.
 VCF's simple text-based design and widespread

### 2024-05-23 paper.tex
edits

@@
 genotype matrix and classifies genotypes within a chunk using a 14 line Python
-function, accelerated using the Numba JIT compliler~\cite{lam2015numba}.
+function, accelerated using the Numba JIT compiler~\cite{lam2015numba}.
 The allele frequencies and genotype counts are then analysed to produce 
@@
 Zarr has in fact been in use in genomics for many years, and was originally 
-developed to [ALISTAIR - details and citations here!]
+developed to store data from the \textit{Anopheles gambiae} 1000 Genomes Project~\citep{anopheles2017genetic}.
+A successor project, the MalariaGEN Vector Observatory,
+currently uses Zarr to store data from whole-genome sequencing of 23,000 \textit{Anopheles} mosquitoes from 31 African countries.\footnote{\url{https://www.malariagen.net/malaria-vector-genome-observatory/}}
+These data are hosted in Google Cloud Storage and can be analysed interactively
+using free cloud computing services like Google Colab,
+democratising the use of data by scientists in malaria-endemic countries
+where access to local computing infrastructure and sufficient network bandwidth to download large datasets may be limited.
 

### 2024-05-24 paper.tex
Updates from Overleaf

--- 2024-05-23 paper.tex
@@
 \item VCF is widely supported, and the underlying data model entrenched 
-in bioinformatics pipelines.
+in Bioinformatics pipelines.
 \item The standard row-wise encoding as text (or binary) is inherently
@@
 collecting genome sequence data for hundreds of thousands of humans.
-Larger genome variation datasets are also being generated for other organisms
-and a variety of purposes including agriculture~\citep{wang2023rice}, conservation~\citep{shaffer2022landscape}
-and infectious disease surveillance~\citep{TODO}.
 VCF's simple text-based design and widespread
@@
 hoyer2017xarray,
-virtanen2020scipy} and gaining increasing
+virtanen2020scipy} and is gaining increasing
 popularity in recent biological applications~\citep[e.g.][]{

### 2024-05-24 paper.tex
Tidy up some GeL details
Closes #107

@@
 of 9.94 TiB over 3,960,177 data storage files. 
-We then converted the ICF to Zarr [DETAILS].
+We then converted the ICF to Zarr, partitioned into
+5989 independent jobs, with an 18.6 min average run time.
 This produced a dataset with 44 arrays, consuming a 
 total of 2.54 TiB of storage over 6,312,488
-chunk files. This is a roughly 5X reduction in total storage space over the original VCF. 
+chunk files. This is a roughly 5X reduction in total storage 
+space over the original VCF. 
 The top fields in terms 
 of storage are detailed in Table~\ref{tab-genomics-england-data}.
+We do not compare with other tools such as Genozip and 
+Savvy here because  they have fundamental limitations 
+(as shown in earlier simulation-based benchmarks),
+and conversion of these large VCFs is a major undertaking.
 

### 2024-05-24 paper.tex
Add note on Genozip
Add QC citations from Eric

@@
 involved and multi-faceted 
-task [CITATIONS?], often requiring interactive, exploratory analysis
+task~\cite{marees2018tutorial,panoutsopoulou2018quality}, 
+often requiring interactive, exploratory analysis
 and incurring substantial computation over multiple QC fields.
@@
 
+Genozip was used under the terms of the free Genozip Academic license.
+Genozip was only used on simulated data, in compliance with
+the ``No Commercial Data'' criterion.
+
 \bibliography{paper}

### 2024-05-24 paper.tex
Cites for phasing and imputation

@@
 collecting genome sequence data for hundreds of thousands of humans.
-Larger genome variation datasets are also being generated for other organisms
-and a variety of purposes including agriculture~\citep{wang2023rice}, conservation~\citep{shaffer2022landscape}
+Large genome variation datasets are also being generated for other organisms
+and a variety of purposes including 
+agriculture~\citep{ros2020accuracy,wang2023rice}, 
+conservation~\citep{shaffer2022landscape}
 and infectious disease surveillance~\citep{hamid2023pf7}.
@@
 formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
-or use 
-bespoke binary formats~\citep[e.g.][]{
+or use bespoke binary formats 
+that support the required access patterns~\citep[e.g.][]{
 % Uses custom "bref3" format,
@@
 % Has a "xcftools" package, but it still looks pretty experimental
-hofmeister2023accurate} 
-that support the required access patterns.
-While PLINK and BGEN support efficient access to genotype data, neither
+hofmeister2023accurate}.
+While PLINK and BGEN are more efficient to access than VCF, neither
 can accommodate the full flexibility of the VCF data model and conversion
@@
 involved and multi-faceted 
-task~\cite{marees2018tutorial,panoutsopoulou2018quality}, 
+task~\cite{marees2018tutorial,panoutsopoulou2018quality,chen2024genomic}, 
 often requiring interactive, exploratory analysis
@@
 Genotype calls are often refined via statistical methods,
-% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
-for example by phasing~[CITES] and imputation[CITES], 
+for example by phasing~\citep{
+browning2021fast,browning2023statistical,hofmeister2023accurate,williams2024phasing},
+and imputation~\citep{browning2018one,rubinacci2020genotype,barton2021whole,
+rubinacci2023imputation}
 creating additional dataset copies.
@@
 \texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
-following standard practices [citations?].
+following standard practices~\citep[e.g.][]{chen2024genomic}.
 This took a total of 689 min CPU time, with the same elapsed time.

### 2024-05-24 paper.tex
Add GWAS cites

@@
 performing GWAS and related analyses require
-data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
+data to be in PLINK or BGEN formats~\cite[e.g][]{chang2015second,
+loh2015efficient,
+abraham2017flashpca2,
+mbatchou2021computationally},
 and so data must be ``hard-called'' according to some QC criteria

### 2024-05-24 paper.tex
Cites for DP/GQ

@@
 \texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
-following standard practices~\citep[e.g.][]{chen2024genomic}.
+following standard practices~\citep[e.g.][]{bergstrom2020insights,
+kousathanas2022whole,
+chen2024genomic}.
 This took a total of 689 min CPU time, with the same elapsed time.

### 2024-05-24 paper.tex
Work in plink2, some additional compression refs

@@
 hofmeister2023accurate}.
-While PLINK and BGEN are more efficient to access than VCF, neither
+While PLINK and BGEN formats are more efficient to access than VCF, neither
 can accommodate the full flexibility of the VCF data model and conversion
@@
 VCF information) have been proposed
-\citep{deorowicz2013genome, %TGC
+\citep{qiao2012handling, % SpeedGene
+deorowicz2013genome, %TGC
+sambo2014compression, % snpack
 deorowicz2019gtshark, %GTShark
@@
 in here.]
+The widely-used PLINK binary format stores genotypes in a 
+packed binary representation, supporting only biallelic
+variants without phase information.
+The PLINK 2 PGEN format~\citep{pgen2024} is more general
+and compact than PLINK, compressing data using specialised
+genotype compression methods~\cite{sambo2014compression}.
 Methods have also been developed which store variation data 
@@
 which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
-or require storage larger 
+or have storage requirements larger
 than uncompressed VCF~\citep{al2023critical}.
@@
 BCF (the standard binary representation of VCF) is similarly 
-row-wise, as are the majority of proposed alternative storage formats.
+row-wise, as are the majority of proposed alternative storage 
+formats.
 Row-wise storage makes retrieving all information

### 2024-05-24 paper.tex

@@
 The PLINK 2 PGEN format~\citep{pgen2024} is more general
-and compact than PLINK, compressing data using specialised
-genotype compression methods~\cite{sambo2014compression}.
+and compact than PLINK, compressing variant data using specialised
+algorithms~\cite{sambo2014compression}.
 Methods have also been developed which store variation data 

### 2024-05-24 paper.tex
Add genomics products refs

@@
 it is complex, vertically integrated software 
+(i.e., consisting of tightly coupled components)
 with custom storage formats that are not intended for reuse.
@@
 the analysis of large-scale genetic variation data are either
-based on proprietary~[Terra? BigQuery?]
-or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
+based on proprietary~\cite{basespace2024,graf2024,googlelifesciences2024,
+awshealthomics2024,microsoftgenomics2024}
+or single-vendor technologies~\cite[e.g.][]{tiledb2024,genomicsdb2024}.
 The next generation of VCF analysis methods requires

### 2024-05-24 paper.tex
Tidy up some stuff on intro

@@
 employed in genetic data compression.
-[FIXME need to work SeqArray~\citep{zheng2017seqarray,zheng2012high}
-in here.]
 The widely-used PLINK binary format stores genotypes in a 
@@
 than uncompressed VCF~\citep{al2023critical}. 
+The SeqArray package~\citep{zheng2017seqarray} builds on the 
+Genomic Data Storage container format~\cite{zheng2012high}
+to store VCF genotype data in a packed and compressed format,
+and is used in several downstream R packages~\cite[e.g.][]{
+gogarten2019genetic,fernandes2020simplephenotypes}.
 
@@
 encoded as a line of text~\citep{danecek2011variant}.
-BCF (the standard binary representation of VCF) is similarly 
-row-wise, as are the majority of proposed alternative storage 
+BCF~\citep{li2011statistical}, the standard binary representation of VCF,
+is similarly row-wise, as 
+are the majority of proposed alternative storage 
 formats.

### 2024-05-24 paper.tex
Switch to reference for URL

@@
 Zarr has in fact been in use in genomics for many years, and was originally 
-developed to store data from the \textit{Anopheles gambiae} 1000 Genomes Project~\citep{anopheles2017genetic}.
+developed to store data from the \textit{Anopheles gambiae} 
+1000 Genomes Project~\citep{anopheles2017genetic}.
 A successor project, the MalariaGEN Vector Observatory,
-currently uses Zarr to store data from whole-genome sequencing of 23,000 \textit{Anopheles} mosquitoes from 31 African countries.\footnote{\url{https://www.malariagen.net/malaria-vector-genome-observatory/}}
-These data are hosted in Google Cloud Storage and can be analysed interactively
+currently uses Zarr to store data from whole-genome sequencing 
+of 23,000 \textit{Anopheles} mosquitoes from 31 African 
+countries~\cite{mvgo2024}.
+The data is hosted in Google Cloud Storage and can be analysed interactively
 using free cloud computing services like Google Colab,
 democratising the use of data by scientists in malaria-endemic countries
-where access to local computing infrastructure and sufficient network bandwidth to download large datasets may be limited.
+where access to local computing infrastructure and sufficient network 
+bandwidth to download large datasets may be limited.
 

### 2024-05-24 src/compression_benchmarks.py
Adding fine-grained chunksize benchmarks

@@ -174,6 +174,57 @@ def test_compression_ratio_vs_shuffle(z_arr, dry_run=False):
     )
 
 
+def test_compression_ratio_vs_chunksize_finegrained(z_arr,
+                                                    min_chunksize=100,
+                                                    max_chunksize=256,
+                                                    dim='sample',
+                                                    dry_run=False):
+
+    dim_idx = ['variant', 'sample'].index(dim)
+
+    comp_results_table = []
+    if not dry_run:
+        # Extract the data
+        arr = z_arr[:]
+
+    for i in tqdm(range(min_chunksize, max_chunksize + 1),
+                  total=max_chunksize - min_chunksize):
+
+        chunksize = list(z_arr.chunks)
+        chunksize[dim_idx] = i
+        chunksize = tuple(chunksize)
+
+        if not dry_run:
+
+            z2 = zarr.empty_like(arr,
+                                 chunks=chunksize,
+                                 compressor=z_arr.compressor,
+                                 dtype=z_arr.dtype)
+            z2[:] = arr
+            nchunks = z2.nchunks
+            compress_ratio = float(dict(z2.info_items())["Storage ratio"])
+        else:
+            nchunks = None
+            compress_ratio = None
+
+        smallest_chunksize = z_arr.shape[dim_idx] % chunksize[dim_idx]
+
+        comp_results_table.append({
+             'chunksize': chunksize,
+            f'{dim}_chunksize': i,
+             'nchunks': nchunks,
+             'smallest_chunk_size': smallest_chunksize,
+             'CompressionRatio': compress_ratio
+        })
+
+    return pd.DataFrame(comp_results_table).sort_values(
+        "CompressionRatio", ascending=False
+    )
+
+
+
+
+
 def test_vcf2zarr_compression_variations(
     z_group,
     keys=("call_GQ", "call_DP", "call_AD", "call_AB", "call_genotype"),
@@ -213,6 +264,11 @@ def test_vcf2zarr_compression_variations(
             res = test_compression_ratio_vs_chunksize(z_group[k], dry_run=dry_run)
         elif test_config == "compressor":
             res = test_compression_ratio_vs_compressor(z_group[k], dry_run=dry_run)
+        elif test_config == "chunksize_finegrained":
+            # Test only call_genotype for now...
+            if k != 'call_genotype':
+                continue
[…]

### 2024-05-30 paper.tex

@@
 \affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
-\affil[4]{Tom's Institute}
+\affil[4]{Tom White Consulting Ltd.}
 \affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
@@
 Recent additions using Zarr include the application of deep learning
-models to genomic sequence data~\citep{klie2023predictive} 
+models to genomic sequence data~\citep{klie2023predictive}, storage and manipulation of large-scale Linkage-Disequilibrium (LD) matrices \cite{ZABAD2023741},  
 and a browser for genetic variation data~\citep{konig2023divbrowse}.
@@
 non-signalling NaN for floats). While this is not a problem 
-in practise for datasets in which all four bases are observed, 
+in practice for datasets in which all four bases are observed, 
 it is a substantial issue for
@@
 works well when compressing boolean fields stored as 
-8 bit integers, where the top 7 bits are always 0. In practise,
+8 bit integers, where the top 7 bits are always 0. In practice,
 boolean fields stored in this way have very similar compression

### 2024-05-30 paper.tex
Add Shadi's figure into document

@@
 Recent additions using Zarr include the application of deep learning
-models to genomic sequence data~\citep{klie2023predictive}, storage and manipulation of large-scale Linkage-Disequilibrium (LD) matrices \cite{ZABAD2023741},  
+models to genomic sequence data~\citep{klie2023predictive}, storage and 
+manipulation of large-scale linkage disequilibrium matrices \cite{zabad2023fast},  
 and a browser for genetic variation data~\citep{konig2023divbrowse}.
@@
 \begin{figure}
+\includegraphics{figures/compression-chunksize-finegrained.pdf}
+\caption{Effects of sample chunk size on compression ratio on call\_genotype
+field in 1000 Genomes data.
+The same analysis as in Fig~\ref{fig-compression-chunksize}, except we only
+consider call\_genotype and we consider all sample chunk sizes from 
+100 to 256. Distinct trend-lines emerge for odd, even and multiple-of-four
+chunk sizes (shown by markers). The size of the final chunk also has a minor
+effect (shown by colour).
+\label{fig-compression-chunksize-finegrained}}
+\end{figure}
+
+\begin{figure}
 \includegraphics{figures/compression-shuffle}

### 2024-05-30 src/plot.py

@@ -38,9 +38,7 @@ def two_panel_fig(**kwargs):
     return fig, ax
 
 
-def plot_size(ax, df, label_y_offset=None,
-              order=None):
-
+def plot_size(ax, df, label_y_offset=None, order=None):
     colour_map = {
         "2bit": two_bit_colour,
         "vcf": vcf_colour,
@@ -90,10 +88,8 @@ def plot_size(ax, df, label_y_offset=None,
 
 
 def plot_total_cpu(
-    ax, df, toolname=None, colours=None, time_units="h", extrapolate=None,
-    order=None
+    ax, df, toolname=None, colours=None, time_units="h", extrapolate=None, order=None
 ):
-
     if colours is None:
         colours = {
             "bcftools+vcf": vcf_colour,
@@ -245,8 +241,13 @@ def whole_matrix_compute(time_data, output):
         "zarr": "zarr-python API",
         "savvy": "savvy C++ API",
     }
-    plot_total_cpu(ax1, df, name_map, extrapolate=["genozip", "bcftools+vcf"],
-                   order=['genozip', 'bcftools+vcf', 'bcftools', 'zarr', 'savvy'])
+    plot_total_cpu(
+        ax1,
+        df,
+        name_map,
+        extrapolate=["genozip", "bcftools+vcf"],
+        order=["genozip", "bcftools+vcf", "bcftools", "zarr", "savvy"],
+    )
 
     plt.savefig(output)
 
@@ -269,8 +270,9 @@ def whole_matrix_decode(time_data, output):
         "savvy": "Savvy",
         "zarr_nshf": "Zarr (Zstd)",
     }
-    plot_total_cpu(ax1, df, toolname=name_map, time_units="m",
-                   order=['zarr', 'zarr_nshf', 'savvy'])
+    plot_total_cpu(
+        ax1, df, toolname=name_map, time_units="m", order=["zarr", "zarr_nshf", "savvy"]
+    )
     df["genotypes_per_second"] = df["total_genotypes"] / df["user_time"]
     for tool in name_map.keys():
         max_rate = df[df.tool == tool]["genotypes_per_second"].max()
@@ -296,8 +298,14 @@ def column_extract(time_data, output):
         "zarr": "Zarr + pandas to_csv",
     }
     fig, ax1 = one_panel_fig()
-    plot_total_cpu(ax1, df, toolname=toolname, time_units="s", extrapolate=["bcftools"],
[…]

### 2024-05-30 paper.tex

@@
 
-Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
-weak influence on compression ratio for most fields. Interestingly,
-the choice of chunk size has a somewhat unpredictable effect
-on compression for the genotype matrix, suggesting that how 
-samples are grouped together has a strong effect on compression.
-At around 50X, compression is excellent in all cases however, 
-and given that genotypes contribute relatively little to the
-overall storage cost in many datasets 
-(Table~\ref{tab-genomics-england-data}), the differences are slight.
-Otherwise, increasing sample chunk size slightly increases 
-compression on call\_AB, and has no effect on less compressible
-fields. Variant chunk size appears to have almost no effect on compression 
-ratio. Thus, we do not expect the choice of chunk size to have a significant
-impact on overall storage usage, and so choice should be
-determined by other considerations such as expected data
-access patterns.
+Fig~\ref{fig-compression-compressor} shows the effect of varying compression
+codecs in Blosc. The combination of outstanding compression performance
+and competitive decoding speed (Fig~\ref{fig-whole-matrix-decode}) 
+makes zstd a good default choice.
 
@@
 
-Fig~\ref{fig-compression-compressor} shows the effect of varying compression
-codecs in Blosc. The combination of outstanding compression performance
-and good decoding speed makes zstd a good default choice.
+Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
+weak influence on compression ratio for most fields. 
+Increasing sample chunk size slightly increases 
+compression on call\_AB, and has no effect on less compressible
+fields. Variant chunk size appears to have almost no effect 
+on compression ratio. 
+Interestingly, the choice chunk size along the sample dimension for the
+genotype matrix does a significant effect. 
+With six evenly spaced points between 100 and 2504,
+Fig~\ref{fig-compression-chunksize}A shows a somewhat unpredictable 
+relationship between sample chunk size and compression ratio.
+The more fine-grained analysis of
+Fig~\ref{fig-compression-chunksize-finegrained} shows that three
+distinct trend lines emerge depending on the chunk size 
+divisibility, with the modulus 
+(i.e., the remainder in the last chunk) also having a minor effect. 
+At greater than 40X,
+compression ratio is 
+high in all cases, and given that genotypes contribute relatively
+little to the total storage of real datasets
+(Table~\ref{tab-genomics-england-data}) the effect will
+likely be fairly minor in practice.
+Thus, we do not expect the choice of chunk size to have a significant
+impact on overall storage usage, and so choice may be
+determined by other considerations such as expected data
+access patterns.
 
@@
 \end{figure}
+\begin{figure}
+
+\includegraphics{figures/compression-compressor}
+\caption{Effects of Blosc compression codec on compression ratio on call-level 
+fields in 1000 Genomes data. 
+In all cases compression level=7 was used, with a variant
+chunk size of 10,000 and sample chunk size of 1,000.
+Bit shuffle was used for call\_genotype, and no shuffle used for the other fields.
+\label{fig-compression-compressor}}
+\end{figure}
+
+\begin{figure}
+\includegraphics{figures/compression-shuffle}
+\caption{Effects of Blosc shuffle settings on compression ratio on call-level 
+fields in 1000 Genomes data. 
+In all cases the zstd compressor with compression level=7 was used, with a variant
+chunk size of 10,000 and sample chunk size of 1,000.
+\label{fig-compression-shuffle}}
+\end{figure}
 
@@
 \includegraphics{figures/compression-chunksize-finegrained.pdf}
-\caption{Effects of sample chunk size on compression ratio on call\_genotype
+\caption{Effects of sample chunk size on compression ratio on the call\_genotype
 field in 1000 Genomes data.
 The same analysis as in Fig~\ref{fig-compression-chunksize}, except we only
-consider call\_genotype and we consider all sample chunk sizes from 
+consider call\_genotype and we examine all sample chunk sizes from 
 100 to 256. Distinct trend-lines emerge for odd, even and multiple-of-four
@@
 
-\begin{figure}
-\includegraphics{figures/compression-shuffle}
-\caption{Effects of Blosc shuffle settings on compression ratio on call-level 
-fields in 1000 Genomes data. 
-In all cases the zstd compressor with compression level=7 was used, with a variant
-chunk size of 10,000 and sample chunk size of 1,000.
-\label{fig-compression-shuffle}}
-\end{figure}
-
-\begin{figure}
-\includegraphics{figures/compression-compressor}
-\caption{Effects of Blosc compression codec on compression ratio on call-level 
-fields in 1000 Genomes data. 
-In all cases compression level=7 was used, with a variant
-chunk size of 10,000 and sample chunk size of 1,000.
-Bit shuffle was used for call\_genotype, and no shuffle used for the other fields.
-\label{fig-compression-compressor}}
-\end{figure}
-
 \end{document}

### 2024-05-30 src/plot.py
Tweak decode plot

@@ -88,7 +88,14 @@ def plot_size(ax, df, label_y_offset=None, order=None):
 
 
 def plot_total_cpu(
-    ax, df, toolname=None, colours=None, time_units="h", extrapolate=None, order=None
+    ax,
+    df,
+    toolname=None,
+    colours=None,
+    time_units="h",
+    extrapolate=None,
+    order=None,
+    label_y_offset=None,
 ):
     if colours is None:
         colours = {
@@ -104,6 +111,7 @@ def plot_total_cpu(
     divisors = {"s": 1, "h": 3600, "m": 60}
     extrapolate = [] if extrapolate is None else extrapolate
 
+    label_y_offset = {} if label_y_offset is None else label_y_offset
     if order is None:
         order = colours.keys()
 
@@ -139,7 +147,7 @@ def plot_total_cpu(
             ax.annotate(
                 f"{time:.0f}{time_units}" if time > 1 else f"{time:.1f}{time_units}",
                 textcoords="offset points",
-                xytext=(15, 0),
+                xytext=(15, label_y_offset.get(tool, 0)),
                 xy=(row.num_samples, total_cpu[-1]),
                 xycoords="data",
             )
@@ -166,7 +174,7 @@ def plot_total_cpu(
         ax.annotate(
             f"{time:.0f}{time_units}*",
             textcoords="offset points",
-            xytext=(15, 0),
+            xytext=(15, label_y_offset.get(tool, 0)),
             xy=(num_samples[-1], fit[-1]),
             xycoords="data",
         )
@@ -260,7 +268,6 @@ def whole_matrix_decode(time_data, output):
     Plot the figure showing raw decode performance on whole-matrix afdist.
     """
     df = pd.read_csv(time_data, index_col=False).sort_values("num_samples")
-    print(df)
     df = df[df.storage == "hdd"]
 
     fig, ax1 = one_panel_fig()
@@ -271,7 +278,12 @@ def whole_matrix_decode(time_data, output):
         "zarr_nshf": "Zarr (Zstd)",
     }
     plot_total_cpu(
-        ax1, df, toolname=name_map, time_units="m", order=["zarr", "zarr_nshf", "savvy"]
+        ax1,
+        df,
+        toolname=name_map,
+        time_units="m",
+        order=["zarr", "zarr_nshf", "savvy"],
+        label_y_offset={"savvy": -2},
     )
     df["genotypes_per_second"] = df["total_genotypes"] / df["user_time"]
[…]

### 2024-05-30 paper.tex
Fix caption
Various figure tidy-ups

@@
 \caption{Genotype decoding performance.
-Total CPU time required to decode genotypes into memory using the zarr-python
+Total CPU time required to decode genotypes into memory using the Zarr-Python
 and Savvy C++ APIs for the data in Figure~\ref{fig-data-storage}.
-This corresponds to maximum rate of 1.2GiB/s for Zarr and 5.7GiB/s
-for Savvy. [TODO add data for 1 million samples]
+This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
+3.9 GiB/s Zarr (Zstd), and 6.6 GiB/s for Savvy. 
 \label{fig-whole-matrix-decode}}

### 2024-05-30 src/plot.py

@@ -273,7 +273,7 @@ def whole_matrix_decode(time_data, output):
     fig, ax1 = one_panel_fig()
 
     name_map = {
-        "zarr": "Zarr (Zstd + bit shuffle)",
+        "zarr": "Zarr (Zstd + BitShuffle)",
         "savvy": "Savvy",
         "zarr_nshf": "Zarr (Zstd)",
     }
@@ -328,7 +328,7 @@ def column_extract(time_data, output):
     plt.savefig(output)
 
 
-def run_subset_matrix_plot(data, output, subset, extrapolate):
+def run_subset_matrix_plot(data, output, subset, extrapolate, **kwargs):
     df = pd.read_csv(data, index_col=False).sort_values("num_samples")
     fig, ax1 = one_panel_fig()
 
@@ -346,6 +346,7 @@ def run_subset_matrix_plot(data, output, subset, extrapolate):
         time_units="s",
         extrapolate=extrapolate,
         order=["genozip", "bcftools", "savvy", "zarr"],
+        **kwargs,
     )
     plt.savefig(output)
 
@@ -357,7 +358,8 @@ def subset_matrix_compute(data, output):
     """
     Plot the figure showing compute performance on subsets of matrix afdist.
     """
-    run_subset_matrix_plot(data, output, "n10", extrapolate=[])
+    run_subset_matrix_plot(data, output, "n10", extrapolate=[],
+            label_y_offset={"savvy":-1})
 
 
 @click.command()
@@ -367,7 +369,8 @@ def subset_matrix_compute_supplemental(data, output):
     """
     Plot the figure showing compute performance on subsets of matrix afdist.
     """
-    run_subset_matrix_plot(data, output, "n/2", extrapolate=["genozip"])
+    run_subset_matrix_plot(data, output, "n/2", extrapolate=["genozip"],
+            label_y_offset={"savvy": 2, "zarr": -2.5})
 
 
 @click.command()

### 2024-05-30 paper.tex
Clear up a minor comment from Shadi

@@
 As datasets grow ever larger, the ability to efficiently access subsets 
-of the data becomes more and more important. VCF/BCF achieve efficient 
+of the data becomes increasingly important. VCF/BCF achieve efficient 
 access to the data for genomic ranges 
@@
 Total CPU time required to run the af-dist calculation for
-a subset of 10,000 variants $\times$ 10 samples from the middle of the matrix
+a contiguous subset of 10,000 variants $\times$ 10 samples 
+from the middle of the matrix
 for the data in Fig~\ref{fig-data-storage}.

### 2024-05-30 paper.tex
Add some discussion of Cloud operation benefits

@@
 quadratically on the number of alleles (although practical
-solutions have been suggested that can be 
-adapted~\cite{poterba2024scalable}).
+solutions have been suggested that we plan to 
+implement~\cite{poterba2024scalable}).
 Secondly, the design of 
@@
 Thirdly, Zarr is not suitable for handling 
-unstructured data often included in VCF INFO 
+the unstructured data often included in VCF INFO 
 fields, and would likely be a poor choice
@@
 Nonetheless, there are numerous datasets that exist today
-what would benefit from cloud-native storage, 
-inherent parallelism and granular storage.
-[Mention benefits of fine-grained access control to 
-different parts of the dataset using native storage access
-control mechanism?]
-
-The VCF Zarr specification and scalable vcf2zarr conversion utility
-that we have provided here are the starting points of such 
-an ecosystem, but much more would be needed to make a viable 
-alternative for everyday bioinformatics workflows. However, a few key
-pieces of infrastructure would provide a starting point, and would 
-be sufficient to greatly improve researcher productivity on 
+that would likely reap significant benefits from being deployed 
+in a cloud-native fashion using Zarr. Object 
+stores typically allow for individual objects (chunks, in
+Zarr) to be associated with ``tags'', which can then be 
+used to associate storage class, user access control
+and encryption keys on a chunk-by-chunk basis. 
+Aside from the performance benefits
+we have focused on here provided by Zarr, the ability
+to (for example) use high-performance storage for commonly
+used data such as the variant position and 
+more cost-effective storage classes
+for infrequently used bulk QC data should provide 
+significant operational benefits.
+Granular access controls would similarly allow non-identifiable
+variant-level data to be shared relatively freely,
+with genotype and other data more tightly controlled
+as required.
+Even finer granularity is possible if samples are grouped according
+to access level within chunks (i.e, by padding partially filled 
+chunks and using an appropriate sample mask). 
+Delegating access control to the cloud provider in this way
+and providing client applications direct access to
+the data in an efficiently analysable form, 
+largely negates the 
+need for web API access~\cite{kelleher2019htsget} 
+and cryptographic container formats~\citep{senf2021crypt4gh}
+in this setting.
+
+The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
+provided here are a necessary starting point for such cloud-native
+biobank repositories, but significant investment and development
+would be needed 
+to provide a viable alternative to standard bioinformatics workflows. 
+A few key pieces of relatively straightforward infrastructure, 
+however, would provide a starting point  and be
+sufficient to greatly improve researcher productivity on 
 large, centrally managed datasets such as Genomics England.
-Firstly, a \texttt{vcfzarrtools} command line utility which 
+Firstly, a ``\texttt{vcztools}'' command line utility which 
 implements a subset of \texttt{bcftools} functionality (such as 
-\texttt{view} and \texttt{query}) would immediately speed up 
-limited region and field specific queries by orders of magnitude,
-and would provide compatibility with existing workflows.
+\texttt{view} and \texttt{query}) on a VCF Zarr dataset 
+would speed up some common queries by orders of magnitude,
+and provide compatibility with existing workflows.
 Developing such a utility would be a non-trivial but routine

### 2024-05-30 paper.tex
Pass through discussion

@@
 fields, and would likely be a poor choice
-for representing variant 
-annotations, for example.
+for representing variant annotations, for example.
 
@@
 used to associate storage class, user access control
-and encryption keys on a chunk-by-chunk basis. 
+and encryption keys. 
 Aside from the performance benefits
@@
 
-The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
+The draft VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
 provided here are a necessary starting point for such cloud-native
-biobank repositories, but significant investment and development
-would be needed 
+biobank repositories and open up many possibilities,
+but significant investment and development would be needed 
 to provide a viable alternative to standard bioinformatics workflows. 
-A few key pieces of relatively straightforward infrastructure, 
-however, would provide a starting point  and be
-sufficient to greatly improve researcher productivity on 
-large, centrally managed datasets such as Genomics England.
-Firstly, a ``\texttt{vcztools}'' command line utility which 
+Two initial directions for development, however, may provide 
+a sufficient starting point to both
+greatly improve researcher productivity on 
+large, centrally managed datasets such as Genomics England
+and motivate further development.
+The first direction is to provide compatibility with existing 
+workflows via a ``\texttt{vcztools}'' command line utility which 
 implements a subset of \texttt{bcftools} functionality (such as 
-\texttt{view} and \texttt{query}) on a VCF Zarr dataset 
-would speed up some common queries by orders of magnitude,
-and provide compatibility with existing workflows.
-Developing such a utility would be a non-trivial but routine
-engineering task.
-
-The second key piece of infrastructure required at this large 
-scale concerns distributed computation. Even with more efficient
-columnar data representations, computations that required examining 
-large volumes of per-call data must be distributed to complete 
-in a reasonable amount of time. [Dask, Xarray, sgkit?]
-
-% \item Vision. An ecosystem of tools that interoperate
-% based on an open, cloud-based format where distribution is naturally
-% managed by the stored chunks. Tools can be run on different datasets
-% with minimal tuning, not by writing a full layer of orchestration.
-% Tools read the data they need, not the entire dataset. Users read
-% directly from a single canonical dataset, and do not maintain
-% their own filtered copies. We suggest that the Zarr VCF approach
-% we have provided would be a good starting point for designing
-% such a compute platform.
-
-% \end{itemize}
-
-The Python data science ecosystem, 
-built on the foundations of efficient array-oriented 
-computing~\citep{harris2020array},
-has a rich suite of powerful tools~\cite{
+\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
+Such a tool would speed up some common queries by orders 
+of magnitude, and reduce the need for user orchestration of 
+operations among manually split VCF chunks (large VCF datasets
+are typically split into hundreds of files; see the Genomics
+England case study). Datasets could then be hosted in cloud
+object stores, while still presenting file-like semantics
+for existing workflows. This could provide an evolutionary path,
+allowing established analysis workflows to coexist with new Zarr-native
+approaches, working from the same primary data.
+
+The second direction is then to develop new methods 
+working directly with the Zarr data, and can leverage 
+a rich ecosystem of powerful and efficient tools in 
+multiple programming languages (see Methods).
+The Python data science ecosystem, in particular,
+has a rich suite of powerful tools~\cite[e.g.][]{
+harris2020array,
 mckinney2010data,
-rocklin2015dask,
 kluyver2016jupyter,
-hoyer2017xarray,
-virtanen2020scipy} and is gaining increasing
+virtanen2020scipy} 
+and multiple Zarr implementations (see Methods).
+Of particular interest for large-scale datasets are
+Dask~\cite{rocklin2015dask} and 
+Xarray~\citep{hoyer2017xarray} which provide powerful abstractions
+for distributed computing and can be
+straightforwardly used to perform interactive analysis
+of VCF Zarr datasets,
+with computations transparently distributed over a cluster.
+As shown in our benchmarks, lower-level computations
+in which numerical operations are applied on a chunk-by-chunk
+basis are also straightforward to implement and highly performant.
+Thus, a range of possibilities exist in which developers can 
+choose the appropriate level of abstraction and tool chain
+on a case-by-case basis.
+
+[NOTE: still trying to figure out how to string this together. 
+Want to finish on a positive, not too technical note.]
+
+and is gaining increasing
 popularity in recent biological applications~\citep[e.g.][]{

### 2024-05-30 paper.tex
Intro tweaks

@@
 Variant Call Format (VCF) is the standard format for interchanging genetic
-variation data~\citep{danecek2011variant}, encoding information about 
+variation data, encoding information about 
 DNA sequence polymorphisms among a set of samples with associated 
-quality control metrics and metadata. 
+quality control metrics and metadata~\citep{danecek2011variant}. 
 Originally defined specifically as a text file,
@@
 collecting genome sequence data for hundreds of thousands of humans.
-Large genome variation datasets are also being generated for other organisms
+Large genetic variation datasets are also being generated for other organisms
 and a variety of purposes including 
@@
 and incurring substantial computation over multiple QC fields.
-Genotype calls are often refined via statistical methods,
+Genotype calls are sometimes refined via statistical methods,
 for example by phasing~\citep{
@@
 creating additional dataset copies.
-A common task to perform after QC is 
+A common task to perform  is
 a genome wide association study (GWAS)~\cite{uffelmann2021genome}. 
@@
 
-VCF cannot support any of these workflows efficiently
-at the Biobank scale, and workarounds of copying subsets 
-of data (i.e., discarding QC metrics in VCF) or copying to 
-PLINK format have reached their limits.
+VCF cannot support any of these workflows efficiently at the Biobank scale.
 The most intrinsically limiting aspect of VCF's design 
@@
 for quality control of large-scale variation datasets, 
-and has been instrumental in the delivery of datasets such as 
+and has been instrumental in projects such as 
 gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.

### 2024-05-30 paper.tex
Add Glow, address some of Eric's points

@@
 gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
-While Hail is built on open components
+Glow~\cite{glow2024} has recently emerged as a similar approach
+to performing distributed analytics on genetic variation data.
+While Hail and Glow are built on open components
 from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
-it is complex, vertically integrated software 
-(i.e., consisting of tightly coupled components)
-with custom storage formats that are not intended for reuse.
+they are both complex projects consisting of 
+tightly coupled components that are not intended for external reuse.
 Similarly, commercial solutions that have emerged to facilitate
@@
 for a given record straightforward and efficient,
-and works well when records are relatively small, 
-or when we typically want to analyse each record in its entirety.
+and works well when records are either relatively small
+or we typically want to analyse each record in its entirety.
 When we want to analyse only a subset of a record,
@@
 storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
-the use of technologies such as Parquet for 
+the use of columnar technologies for
 storing and analysing variation data have had limited
 success~\citep{boufea2017managing,fan2020variant}.
-Columnar technologies tend to be 
-optimised for ``tall and narrow'' data where there is a large number of records
-for a relatively small number of columns. 
 Mapping VCF directly to a columnar layout, in which there is a 
 column for the genotypes (and other per-call QC metrics) 
-for each sample
-leads to millions of columns, and substantial technical challenges.
-Additionally, compression performance within columns tends to
-be poor because there is much less similarity among
-the genotypes for a given sample than there is among the genotypes
-for a given variant, requiring
-specialised haplotype compression 
-techniques~\citep{layer2016efficient,mcvean2019linkage}.
+for each sample leads to a large number of columns, which 
+can be cumbersome and cause scalability issues.
+Fundamentally, columnar methods are one-dimensional, storing a vector
+of values associated with a particular key, whereas 
+genetic variation data is usually modelled as a two-dimensional matrix
+in which we are interested in accessing both rows \emph{and} columns.
+Just as row-oriented storage makes accessing data for a given
+sample inefficient, columnar storage makes accessing all the data 
+for a given variant inefficient.
 

### 2024-05-30 paper.tex
Layout tweaks

@@
 have been shown to closely follow 
-observations~\cite{anderson2023on} and so provides 
+observations~\cite{anderson2023on} and so it provides 
 a reasonable and easily reproducible data point 
@@
 is something of a best-case scenario for specialised genotype
-compression methods using row-based storage.
+compression methods using row-wise storage.
 
@@
 superb compression, having far smaller file sizes that the 
-other methods (although somewhat losing its advantages at 
+other methods (although somewhat losing its advantage at 
 larger sample sizes). Zarr and Savvy have 
@@
 \subsection{Subsetting the genotype matrix}
+\begin{figure}[t]
+\includegraphics{figures/subset-matrix-compute}
+\caption{Compute performance on subsets of the matrix.
+Total CPU time required to run the af-dist calculation for
+a contiguous subset of 10,000 variants $\times$ 10 samples 
+from the middle of the matrix
+for the data in Fig~\ref{fig-data-storage}.
+The \texttt{genozip} and \texttt{bcftools} pipelines involve
+multiple commands required to correctly calculate the AF INFO field
+required by \texttt{bcftools +af-dist}. See the Methods for full details
+on the steps performed.
+\label{fig-subset-matrix-compute}}
+\end{figure}
+
 As datasets grow ever larger, the ability to efficiently access subsets 
@@
 
-\begin{figure}
-\includegraphics{figures/subset-matrix-compute}
-\caption{Compute performance on subsets of the matrix.
-Total CPU time required to run the af-dist calculation for
-a contiguous subset of 10,000 variants $\times$ 10 samples 
-from the middle of the matrix
-for the data in Fig~\ref{fig-data-storage}.
-The \texttt{genozip} and \texttt{bcftools} pipelines involve
-multiple commands required to correctly calculate the AF INFO field
-required by \texttt{bcftools +af-dist}. See the Methods for full details
-on the steps performed.
-\label{fig-subset-matrix-compute}}
-\end{figure}
-
 We illustrate this cost of row-wise encoding in
@@
 
+\subsection{Extracting fields}
 \begin{figure}
@@
 \end{figure}
-
-\subsection{Extracting fields}
 We have focused on the genotype matrix up to this point, contrasting

### 2024-05-30 paper.tex
Clear up some TODOs

@@
 examined, perform some pre-specified calculations and finally 
-output some text, typically VCF or tab/comma separated values
-[TODO check these citations]~\citep{
+output some text, typically VCF or tab/comma separated 
+values~\citep{
 layer2016efficient, %GQT
 li2016bgt, % BGT
-tatwawadi2016gtrac, % GTRAC
 danek2018gtc, % GTC
@@
 lan2020genozip,lan2021genozip, %genozip
-lefaive2021sparse, % SAVVY
-wertenbroek2022xsi,% XSI
 zhang2023gbc}. %GBC
@@
 methods have begun to provide Application Programming Interfaces
-(APIs), providing efficient access to genotype and other VCF data
-[TODO check these citations]~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
-wertenbroek2022xsi,zhang2023gbc}. By providing programmatic access,
+(APIs), providing efficient access to genotype and other VCF 
+data~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
+wertenbroek2022xsi}. By providing programmatic access,
 the data can be retrieved from storage, decoded and then analysed
@@
 the required records. 
-Methods like Savvy and XSI largely follow the same principles [TODO check and 
-be more concrete]. The row-wise nature of these methods, however, mean 
+The row-wise nature of VCF (and most proposed alternatives), however, means
 that we cannot efficiently subset \emph{by sample}

### 2024-05-31 paper.tex
Pass through discussion

@@
 is fundamentally inefficient,
-profoundly limiting the scalability of the current generation
+profoundly limiting the scalability of the present generation
 of bioinformatics tools. Large scale VCF data cannot 
@@
 processed without incurring a substantial economic
-and likely environmental cost~\cite{grealey2022carbon}.
+(and environmental~\cite{grealey2022carbon}) cost.
 We have shown here that this is not a necessary situation,
@@
 corresponding novel variants) should be added.
-Thirdly, Zarr is not suitable for handling 
-the unstructured data often included in VCF INFO 
-fields, and would likely be a poor choice
-for representing variant annotations, for example.
+Thirdly, Zarr works best for numerical data of a fixed 
+dimension, and therefore may not suitable for representing
+the unstructured data often included in VCF INFO fields.
 
@@
 as required.
-Even finer granularity is possible if samples are grouped according
-to access level within chunks (i.e, by padding partially filled 
-chunks and using an appropriate sample mask). 
-Delegating access control to the cloud provider in this way
-and providing client applications direct access to
-the data in an efficiently analysable form, 
-largely negates the 
-need for web API access~\cite{kelleher2019htsget} 
+Even finer granularity is possible if samples are grouped by
+access level within chunks (padding partially filled 
+chunks as needed and using an appropriate sample mask). 
+Providing client applications direct access to 
+the data over HTTP
+and delegating access control to the cloud provider
+makes custom web APIs~\cite{kelleher2019htsget} 
 and cryptographic container formats~\citep{senf2021crypt4gh}
-in this setting.
-
-The draft VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
+largely unnecessary.
+
+The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
 provided here are a necessary starting point for such cloud-native
@@
 to provide a viable alternative to standard bioinformatics workflows. 
-Two initial directions for development, however, may provide 
-a sufficient starting point to both
-greatly improve researcher productivity on 
+Two initial directions for development, however, may quickly
+yield sufficient results to both greatly improve researcher productivity on 
 large, centrally managed datasets such as Genomics England
-and motivate further development.
+and motivate further research and development.
 The first direction is to provide compatibility with existing 
@@
 
-The second direction is then to develop new methods 
-working directly with the Zarr data, and can leverage 
-a rich ecosystem of powerful and efficient tools in 
-multiple programming languages (see Methods).
+The second natural direction for development is to create
+these Zarr-native applications, which can take advantage
+of the efficient data representation
+across multiple programming languages (see Methods).
 The Python data science ecosystem, in particular,
 has a rich suite of powerful tools~\cite[e.g.][]{
+mckinney2010data,
+lam2015numba,
+kluyver2016jupyter,
 harris2020array,
-mckinney2010data,
-kluyver2016jupyter,
 virtanen2020scipy} 
-and multiple Zarr implementations (see Methods).
-Of particular interest for large-scale datasets are
+and is increasingly
+popular in recent biological applications~\citep[e.g.][]{
+abdennur2020cooler,
+rand2022bionumpy,
+open2c2024bioframe,
+hou2024admix}.
 Dask~\cite{rocklin2015dask} and 
-Xarray~\citep{hoyer2017xarray} which provide powerful abstractions
-for distributed computing and can be
-straightforwardly used to perform interactive analysis
-of VCF Zarr datasets,
+Xarray~\citep{hoyer2017xarray} provide powerful abstractions
+for large-scale data analysis and can be
+used to perform interactive analysis
+of large VCF Zarr datasets,
 with computations transparently distributed over a cluster.
 As shown in our benchmarks, lower-level computations
-in which numerical operations are applied on a chunk-by-chunk
+in which optimised kernels are applied on a chunk-by-chunk
 basis are also straightforward to implement and highly performant.
@@
 
-[NOTE: still trying to figure out how to string this together. 
-Want to finish on a positive, not too technical note.]
-
-and is gaining increasing
-popularity in recent biological applications~\citep[e.g.][]{
-abdennur2020cooler,
-rand2022bionumpy,
-open2c2024bioframe,
-hou2024admix}.
 Zarr has the potential to act as a unified, cloud-native 

### 2024-05-31 paper.tex
A crack at the discussion

@@
 
-Zarr has the potential to act as a unified, cloud-native 
-efficient storage layer, freeing the next generation of 
-bioinformatics tools
-from the traditional burdens of text parsing and format
-conversion.
+While Zarr is now widely used across the sciences (see Methods)
+it was originally 
+developed to store genetic variation data from the 
+\textit{Anopheles gambiae} 
+1000 Genomes Project~\citep{anopheles2017genetic}
+and is in active use in this 
+setting~\cite[e.g.][]{ahouidi2021open,trimarsanto2022molecular}.
+The VCF Zarr specification presented here builds on this real-world 
+experience but is still a draft proposal that would benefit 
+from wider input across a range of applications. With some 
+refinements and sufficient uptake it may be suitable 
+for standardisation~\cite{rehm2021ga4gh}.
+The benefits of Zarr are substantial, and, in certain settings,
+worth the cost of retooling away from classical file-oriented workflows.
+For example, the MalariaGEN Vector Observatory
+currently uses Zarr to store data from whole-genome sequencing 
+of 23,000 \textit{Anopheles} mosquitoes from 31 African 
+countries~\cite{mvgo2024}.
+The data is hosted in Google Cloud Storage and can be analysed interactively
+using free cloud computing services like Google Colab,
+enabling the use of data by scientists in malaria-endemic countries
+where access to local computing infrastructure and sufficient network 
+bandwidth to download large datasets may be limited.
+VCF Zarr could similarly reduce the costs of analysing large-scale
+human data, and effectively open access to biobanks
+for a much broader group of researchers than currently possible.
 
@@
 and a browser for genetic variation data~\citep{konig2023divbrowse}.
-Zarr has in fact been in use in genomics for many years, and was originally 
-developed to store data from the \textit{Anopheles gambiae} 
-1000 Genomes Project~\citep{anopheles2017genetic}.
-A successor project, the MalariaGEN Vector Observatory,
-currently uses Zarr to store data from whole-genome sequencing 
-of 23,000 \textit{Anopheles} mosquitoes from 31 African 
-countries~\cite{mvgo2024}.
-The data is hosted in Google Cloud Storage and can be analysed interactively
-using free cloud computing services like Google Colab,
-democratising the use of data by scientists in malaria-endemic countries
-where access to local computing infrastructure and sufficient network 
-bandwidth to download large datasets may be limited.
 
@@
 The VCF Zarr specification is a direct mapping from the VCF data model
-to a chunked binary array format using Zarr. VCF Zarr takes advantage
+to a chunked binary array format using Zarr,
+and is evolution of the Zarr format used in the \texttt{scikit-allel}
+package~\citep{miles2023scikit}.
+VCF Zarr takes advantage
 of Zarr's hierarchical structure by representing a VCF file as a top-level
@@
 including contigs, filters, and samples.
-
-% Do we need to say that VCF Zarr is an evolution of the Zarr format in scikit-allel?
 

### 2024-05-31 paper.tex
First pass at Acks

@@
 interest here as it achieves very high compression ratios
-with good decompression speeds. See Fig~\ref{fig-whole-matrix-decode}
-for a comparison of Zstandard with LZ4 on decoding performance.
+with good decompression speeds (Figs~\ref{fig-whole-matrix-decode},
+\ref{fig-compression-compressor}).
 % This may seem a bit off topic, but want to reassure reader that 
@@
 to a chunked binary array format using Zarr,
-and is evolution of the Zarr format used in the \texttt{scikit-allel}
+and is an evolution of the Zarr format used in the \texttt{scikit-allel}
 package~\citep{miles2023scikit}.
@@
 As this approach is on a likely path to 
-standardisation~\cite{danecek2021twelve}, we would aim to 
+standardisation~\cite{danecek2021twelve}, we plan to 
 include support in later versions of VCF Zarr.
@@
 and minimum values observed during conversion to ICF
-along with reasonable compressor defaults (see next section), 
-but these can be
+along with reasonable compressor defaults (see next section).
+Default choices can be 
 modified by generating a JSON-formatted storage schema,
 which can be edited and supplied as an argument to \texttt{encode}.
-Encoding a given column (for example, \texttt{call\_AD})
+Encoding a given field (for example, \texttt{call\_AD})
 involves creating a buffer to hold a full variant-chunk of the 
@@
 
-\section{Data availability}
-
-\section{Declarations}
+The VCF Zarr specification is available on GitHub at 
+\url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
+All source code for running benchmarks, analyses and creating 
+"
+plots in this article is available at
+\url{https://github.com/sgkit-dev/vcf-zarr-publication}.
+Vcf2zarr is freely available under the terms of the Apache 2.0 
+license as part of the \texttt{bio2zarr} 
+suite (\url{https://github.com/sgkit-dev/bio2zarr/}) 
+and can be installed from the Python Package Index
+(\url{https://pypi.org/project/bio2zarr/}).
 
@@
 \begin{itemize}
-    \item VCF: Variant Call Format;
+    \item ICF: Intermediate Columnar Format
+    \item GWAS: Genome Wide Association Study
+    \item PBWT: Positional Burrows-Wheeler Transform
+    \item QC: Quality Control
     \item UKB: UK Biobank
+    \item VCF: Variant Call Format
     \item WGS: Whole Genome Sequence
@@
 
-\subsection{Competing Interests}
+% \subsection{Competing Interests}
 \subsection{Funding}
-
-\subsection{Author's Contributions}
+JK acknowledges support from the Robertson Foundation, 
+NIH (research grants HG011395 and HG012473), 
+and Bill \& Melinda Gates Foundation (INV-001927; Sanger subaward G3594).
+AM acknowledges support from the Bill \& Melinda Gates Foundation 
+(INV-001927).
+
+% \subsection{Author's Contributions}
 
 \section{Acknowledgements}
+This research was made possible through access to data in the National Genomic
+Research Library, which is managed by Genomics England Limited (a wholly owned
+company of the Department of Health and Social Care). The National Genomic
+Research Library holds data provided by patients and collected by the NHS as
+part of their care and data collected as part of their participation in
+research. The National Genomic Research Library is funded by the National
+Institute for Health Research and NHS England. The Wellcome Trust, Cancer
+Research UK and the Medical Research Council have also funded research
+infrastructure.
+
+Computation used the Oxford Biomedical Research Computing (BMRC) facility, a
+joint development between the Wellcome Centre for Human Genetics and the Big
+Data Institute supported by Health Data Research UK and the NIHR Oxford
+Biomedical Research Centre. The views expressed are those of the author(s) and
+not necessarily those of the NHS, the NIHR or the Department of Health.
 

### 2024-05-31 paper.tex
Update GEL numbers

@@
 large human dataset, we performed some illustrative benchmarks.
-Firstly, we ran the af-dist calculation (Figs~\ref{fig-whole-matrix-compute}
-and~\ref{fig-subset-matrix-compute}) on one 132GiB VCF file covering
-positions 58,219,159--60,650,943 (562,640 variants) using \texttt{bcftools
-+af-dist} as before. The elapsed time for this operation was 716 min
-user+sys time almost exactly equal). We repeated this operation for the 
-same genome coordinates in Zarr (using Python code described in previous sections) 
-% 716 minutes / 4.25 = 168X
-with a total CPU time of 2min 19s and elapsed time of 4min 15s (168X faster).
-Second, we extracted the genome position for each variant in this single VCF 
+As these benchmarks take some time to run, we focus 
+on a single 132GiB compressed VCF file covering
+positions 58,219,159--60,650,943 (562,640 variants) 
+from the middle of the set of 106 files for chromosome 2.
+We report both the total CPU time and elapsed wall-clock time here
+as both are relevant.
+First, we extracted the genome position for each variant in this single VCF 
 chunk using \texttt{bcftools query} and Python Zarr code as described in 
@@
 % sys 0m45.277s
-55.42 min total CPU (85.85 min elapsed).
+55.42 min CPU and 85.85 min elapsed.
 % CPU times: user 846 ms, sys: 1.93 s, total: 2.78 s
 % Wall time: 1min 44s
-% 85.85 / 1.73 = 49.62 = 50X
-The Zarr code required 2.78 sec total CPU and 1.73 min elapsed (50X faster).
-Note that the Zarr code here is heavily I/O bound [TODO say something 
-about the GEL cluster?]. When we subsequently time the operation of 
-just populating an array of the variant positions in specified region,
-this is reduced to 1.13 sec, with an elapsed time of 20.9 sec.
-[Note: this is crazy that we still have a 19 second I/O wait. The
-5989 files for the POS chunks must be in cache already, so the delay
-has to be the file system checking cache coherency.]
+% 55.42 * 60 / 2.78  = 1196
+% 85.85 / 1.73 = 49.62 
+The Zarr code required 2.78 sec CPU and 1.73 min elapsed.
+This is a 1196X smaller CPU burden and a 50X speed-up in elapsed time.
+The major difference between CPU time and wall-time is noteworthy 
+here, and indicates some opportunities for improvement in VCF Zarr
+in high-latency environments such as the shared file system in the
+Genomics England HPC system. Currently VCF Zarr does not store any
+specialised index to map genomic coordinates to array positions
+along the variants dimension. Instead, to find the relevant slice
+of records corresponding to the range of positions in the target
+VCF file, we read in the entire variant\_position array and 
+binary search. This entails reading 5,989 chunk files
+(the chunk size is 100,000 variants) which incurs a substantial
+latency penalty on this system. Later versions of the specification
+may solve this problem by either storing a combined variant\_contig,
+variant\_position array as a single chunk, or use a specialised
+index that supports overlap queries.
+
+% When we subsequently time the operation of 
+% just populating an array of the variant positions in specified region,
+% this is reduced to 1.13 sec, with an elapsed time of 20.9 sec.
+% [Note: this is crazy that we still have a 19 second I/O wait. The
+% 5989 files for the POS chunks must be in cache already, so the delay
+% has to be the file system checking cache coherency.]
+
+We then ran the af-dist calculation (Figs~\ref{fig-whole-matrix-compute}
+and~\ref{fig-subset-matrix-compute}) on the VCF file
+\texttt{bcftools +af-dist} as before. 
+The elapsed time for this operation was 716.28 min CPU, 716.3 min elapsed.
+Repeating this operation for the same coordinates in Zarr 
+(using Python code described in previous sections) 
+% 716.28 / 2.32 = 308.74
+% 716.3 / 4.25 = 168.54
+with a total CPU time of 2.32 min and elapsed time of 4.25 min.
+This is a 309X reduction in CPU burden and a 169X speed-up in elapsed time.
 
@@
 chen2024genomic}.
-This took a total of 689 min CPU time, with the same elapsed time.
+This used 689.46 min CPU time, with an elapsed time of 689.48 min.
 In comparison, computing and storing a variant mask (i.e., a boolean value 
@@
 for analysis) based on the same criteria using Zarr 
-required a total of 1.96 min CPU time with an elapsed time 
-of  11 min (63X faster).
-%689 / 11 = 62.6 = 63X
+consumed 1.96 min CPU time with an elapsed time of 11 min.
+This is a 358X reduction in CPU usage, and 63X reduction in elapsed time.
+% 689.46 / 1.96 = 351.76
+% 689.48 / 11 = 62.6 = 63X
 There is an important distinction here between creating a copy

### 2024-05-31 paper.tex
Update abstract

@@
 genetic variation data and associated quality control metrics.
-It provides a well-defined data model and is central to a large ecosystem
-of interoperating tools.
-The standard row-wise encoding of the VCF data model (either as text
+% It provides a well-defined data model and is central to a large ecosystem
+% of interoperating tools.
+The usual row-wise encoding of the VCF data model (either as text
 or packed binary) emphasises efficient retrieval of all data for a given
-variant, but accessing subsets of the data within rows is inefficient.
+variant, but accessing data on a field or sample basis is inefficient.
 Biobank scale datasets currently available 
@@
 and hundreds of terabytes of compressed VCF.
-Row-based data storage is fundamentally unsuited to such datasets
-and a more scalable, grid-based solution is needed.
+Row-wise data storage is fundamentally unsuitable
+and a more scalable approach is needed.
 
 \textbf{Results:}
-We present the VCF Zarr specification, a columnar encoding of the 
+We present the VCF Zarr specification, an encoding of the 
 VCF data model using Zarr which makes retrieving subsets of the 
@@
 We demonstrate the VCF Zarr format (and the vcf2zarr conversion utility) 
-on the Genomics England dataset of XX,XXX,XXX samples, showing
-an overall reduction in storage of XX, and a reduction in processing 
-time of common queries of up to XX fold.
+on a subset of the Genomics England aggV2 dataset comprising
+78,195 samples and 59,880,903 variants,
+with a 5X reduction in storage and greater than 300X reduction in CPU usage
+in some representative benchmarks.
 
@@
 Large row-encoded VCF files are a major bottleneck for current research, and 
-storing and processing these files incurs a substantial economic cost.
+storing and processing these files incurs a substantial cost.
 The VCF Zarr specification, building on widely-used, open-source technologies

### 2024-06-03 paper.tex
Shadi's updates
Update acks

@@
 on compression ratio. 
-Interestingly, the choice chunk size along the sample dimension for the
-genotype matrix does a significant effect. 
+Interestingly, the choice of chunk size along the sample dimension for the
+genotype matrix does have a significant effect. 
 With six evenly spaced points between 100 and 2504,
@@
 JK acknowledges support from the Robertson Foundation, 
-NIH (research grants HG011395 and HG012473), 
-and Bill \& Melinda Gates Foundation (INV-001927; Sanger subaward G3594).
-AM acknowledges support from the Bill \& Melinda Gates Foundation 
-(INV-001927).
+NIH (research grants HG011395 and HG012473). 
+JK and AM acknowledge the Bill \& Melinda Gates Foundation (INV-001927).
 

### 2024-06-07 paper.tex
Eric's update
Tweaks to GEL section
Closes #140

@@
 positions 58,219,159--60,650,943 (562,640 variants) 
-from the middle of the set of 106 files for chromosome 2.
+from the middle of the list of 106 files for chromosome 2.
 We report both the total CPU time and elapsed wall-clock time here
@@
 of records corresponding to the range of positions in the target
-VCF file, we read in the entire variant\_position array and 
+VCF file, we load the entire variant\_position array and 
 binary search. This entails reading 5,989 chunk files
@@
 latency penalty on this system. Later versions of the specification
-may solve this problem by either storing a combined variant\_contig,
-variant\_position array as a single chunk, or use a specialised
-index that supports overlap queries.
+may solve this problem by storing an array of size 
+(approximately) the number variant chunks 
+which maps ranges of genome coordinates to chunk indexes,
+or a more specialised structure that supports overlap queries.
 
@@
 and~\ref{fig-subset-matrix-compute}) on the VCF file
-\texttt{bcftools +af-dist} as before. 
+using \texttt{bcftools +af-dist} as before. 
 The elapsed time for this operation was 716.28 min CPU, 716.3 min elapsed.
@@
 % 716.3 / 4.25 = 168.54
-with a total CPU time of 2.32 min and elapsed time of 4.25 min.
+gave a total CPU time of 2.32 min and elapsed time of 4.25 min.
 This is a 309X reduction in CPU burden and a 169X speed-up in elapsed time.
+It is worth noting here that \texttt{bcftools +af-dist} cannot be 
+performed in parallel across multiple slices of a chromosome, and if 
+we did want to run it on all of chromosome 2 we would need to 
+concatenate the 106 VCF files. While af-dist itself is not a common operation,
+many tasks share this property
+of not being straightforwardly decomposable across multiple VCF files.
 
@@
 choose the appropriate level of abstraction and tool chain
-on a case-by-case basis.
+on a case-by-case basis. These abstractions will extend, hopefully, to
+include more convenient primitives for genomics workflows through
+projects like sgkit in the future.
 

### 2024-06-07 paper.tex
Add point about annotations
Closes #141

@@
 
-\subsection{Extracting fields}
+\subsection{Extracting, inserting and updating fields}
 \begin{figure}
@@
 excels at these tasks: we only read and decompress the information required.
+
+Many of the additional fields that we find in real-world VCFs are 
+variant-level annotations, extensively used in downstream applications.
+For example, a common workflow is to 
+add or update variant IDs in a VCF using a reference database
+such as dbSNP \cite{Sherry2001dbSNP}. The standard approach to this
+(using e.g.\ \texttt{bcftools annotate}) is to create a \emph{copy} of 
+the VCF which includes these new annotations. Thus, even though 
+we many only be altering a single field comprising a tiny fraction 
+of the data, we still read, decompress, update, compress and 
+write the entire dataset to a new file. With Zarr,
+we can update an existing field or add arbitrary additional
+fields without touching the rest of the data or creating redundant
+copies.
 

### 2024-06-07 paper.tex
Apply Jeff's comments to intro

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%,%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Please note that whilst this template provides a
@@
 gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
-Glow~\cite{glow2024} has recently emerged as a similar approach
-to performing distributed analytics on genetic variation data.
-While Hail and Glow are built on open components
+% Glow~\cite{glow2024} has recently emerged as a similar approach
+% to performing distributed analytics on genetic variation data.
+While Hail is built on open components
 from the Hadoop distributed computing ecosystem~\citep{white2012hadoop},
-they are both complex projects consisting of 
-tightly coupled components that are not intended for external reuse.
+the details of its MatrixTable format are not documented
+or intended for external reuse. 
+% cite? https://dev.hail.is/t/matrixtable-file-format-reference/173/6
 Similarly, commercial solutions that have emerged to facilitate

### 2024-06-07 paper.tex

@@
 gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
-% Glow~\cite{glow2024} has recently emerged as a similar approach
-% to performing distributed analytics on genetic variation data.
 While Hail is built on open components
@@
 and cryptographic container formats~\citep{senf2021crypt4gh}
-largely unnecessary.
+largely unnecessary in this setting.
 
@@
 hou2024admix}.
-Dask~\cite{rocklin2015dask} and 
-Xarray~\citep{hoyer2017xarray} provide powerful abstractions
-for large-scale data analysis and can be
-used to perform interactive analysis
-of large VCF Zarr datasets,
-with computations transparently distributed over a cluster.
-As shown in our benchmarks, lower-level computations
-in which optimised kernels are applied on a chunk-by-chunk
-basis are also straightforward to implement and highly performant.
-Thus, a range of possibilities exist in which developers can 
-choose the appropriate level of abstraction and tool chain
-on a case-by-case basis. These abstractions will extend, hopefully, to
-include more convenient primitives for genomics workflows through
-projects like sgkit in the future.
+Xarray~\cite{hoyer2017xarray} provides a unified interface for working 
+with multi-dimensional arrays
+in Python, and libraries like Dask~\cite{rocklin2015dask}  
+and Cubed~\cite{cubed2024} allow these operations to be
+scaled out transparently across processors and clusters. 
+This scaling is achieved by distributing calculations over 
+grid-based array representations like Zarr, where chunks
+provide the basic unit for parallel computation.
+The VCF Zarr specification introduced here was created 
+to facilitate work on a scalable genetics toolkit for 
+Python~\cite{sgkit2024} built on Xarray. 
+While the high-level facilities for distributed 
+computation provided by Xarray 
+are very powerful, they are not needed or indeed
+appropriate in all contexts. Our benchmarks here 
+illustrate that working at the lowest level, 
+by sequentially applying optimised kernels on a chunk-by-chunk
+basis is both straightforward to implement and highly performant.
+Thus, a range of possibilities exist in which developers 
+can build utilities using the VCF Zarr specification
+using the appropriate level of abstraction and tool chain
+on a case-by-case basis.
 

### 2024-06-07 paper.tex
Tweak formatting for preprint
Improve typesetting of the URLs

@@
 %%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
-\journal{gigascience}
+\journal{preprint}
 
@@
 \usetikzlibrary{quotes,arrows.meta,3d}
-
-% Local definitions to systematise tool naming
-\newcommand{\sgapi}[1]{\texttt{#1}}
-\newcommand{\toolname}[1]{\texttt{#1}}
-\newcommand{\sgkit}{\texttt{sgkit}}
 
@@
 The VCF Zarr specification is general and can be mapped to 
-file formats such as \toolname{plink}~\citep{purcell2007plink,chang2015second}
+file formats such as PLINK~\citep{purcell2007plink,chang2015second}
 and BGEN~\citep{band2018bgen} with some minor extensions.

### 2024-06-07 paper.tex
Tweak funding
Typesetting tweaks
Shadi's typo fix

@@
 the VCF which includes these new annotations. Thus, even though 
-we many only be altering a single field comprising a tiny fraction 
+we may only be altering a single field comprising a tiny fraction 
 of the data, we still read, decompress, update, compress and 
@@
 \subsection{Funding}
-JK acknowledges support from the Robertson Foundation, 
+JK acknowledges the Robertson Foundation and 
 NIH (research grants HG011395 and HG012473). 
@@
 
-\clearpage
+%\clearpage
 \renewcommand\thefigure{S\arabic{figure}}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics{figures/whole-matrix-decode}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics{figures/subset-matrix-compute-supplemental}
@@
 \end{figure}
-\begin{figure}
+\begin{figure}[h]
 
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics{figures/compression-shuffle}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics{figures/compression-chunksize}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics{figures/compression-chunksize-finegrained.pdf}

### 2024-06-10 scaling/vcf2zarr_prototype.py
Remove vcf2zarr prototype

@@ -1,341 +1,0 @@
-# Experimental sequential multithreaded version of VCF to sgkit zarr
-# conversion. This is a minimal standalone version, forming the basis
-# of the version in sgkit.
-#
-# This is entirely untested, and only keeping here as a useful reference.
-# DO NOT USE!
-import contextlib
-import concurrent.futures
-import dataclasses
-import subprocess
-import time
-import multiprocessing
-import os
-from typing import Any
-import time
-
-import click
-import numpy as np
-import cyvcf2
-import zarr
-import numcodecs
-import tqdm
-import threading
-import queue
-
-numcodecs.blosc.use_threads = False
-
-
-# based on https://gist.github.com/everilae/9697228
-# Needs refactoring to allow for graceful handing of
-# errors in the main thread, and in the decode thread.
-class ThreadedGenerator:
-    def __init__(self, iterator, queue_maxsize=0, daemon=False):
-        self._iterator = iterator
-        self._sentinel = object()
-        self._queue = queue.Queue(maxsize=queue_maxsize)
-        self._thread = threading.Thread(name="generator_thread", target=self._run)
-        self._thread.daemon = daemon
-
-    def _run(self):
-        # TODO check whether this correctly handles errors in the decode
-        # thread.
-        try:
-            for value in self._iterator:
-                self._queue.put(value)
-
-        finally:
-            self._queue.put(self._sentinel)
-
-    def __iter__(self):
-        self._thread.start()
-        for value in iter(self._queue.get, self._sentinel):
-            yield value
-
-        self._thread.join()
-
-
-def sync_flush_array(np_buffer, zarr_array, offset):
-    zarr_array[offset : offset + np_buffer.shape[0]] = np_buffer
-
-
-def async_flush_array(executor, np_buffer, zarr_array, offset):
-    """
-    Flush the specified chunk aligned buffer to the specified zarr array.
-    """
-    assert zarr_array.shape[1:] == np_buffer.shape[1:]
-
-    if len(np_buffer.shape) == 1:
-        futures = [executor.submit(sync_flush_array, np_buffer, zarr_array, offset)]
-    else:
-        futures = async_flush_2d_array(executor, np_buffer, zarr_array, offset)
-    return futures
-
-
-def async_flush_2d_array(executor, np_buffer, zarr_array, offset):
-    # Flush each of the chunks in the second dimension separately
-    s = slice(offset, offset + np_buffer.shape[0])
-
-    def flush_chunk(start, stop):
[…]

### 2024-06-10 paper.tex
Add line numbers

@@
 \usepackage{siunitx}
+\usepackage{lineno}
+
+\linenumbers
 
@@
 Zarr is flexible in allowing different compression codecs and 
-pre-compression filters to be specified on a per-array basis
-(see section XXX).
+pre-compression filters to be specified on a per-array basis.
 Two key technologies often used in conjunction with Zarr are the Blosc
@@
 All source code for running benchmarks, analyses and creating 
-"
 plots in this article is available at

### 2024-06-19 paper.tex
Add explicit description of dotted lines
Add citation for NetCDF and tweak Zarr discussion.

@@
 \usepackage{siunitx}
-\usepackage{lineno}
+\usepackage[switch]{lineno}
 
@@
 for the data in Fig~\ref{fig-data-storage}.
+Elapsed time is also reported (dotted line). 
 The \texttt{genozip} and \texttt{bcftools} pipelines involve
@@
 Metadata describing the array and its properties is then stored 
-in JSON format along with the chunks. The simplicity and transparency
-of this design has substantial advantages over other technologies
-such as HDF5~\citep{folk2011overview} which are relatively complex and opaque.
-This simplicity has led to numerous implementations of the Zarr specification
-being developed, ranging from the mature Zarr-Python~\citep{zarrpython}
+in JSON format along with the chunks. 
+The simplicity and transparency
+of this design has substantial advantages over technologies
+such as HDF~\citep{folk2011overview} and 
+NetCDF~\citep{rew1990netcdf} which are based on complex
+layouts of multidimensional data within a single file,
+and cannot be accessed in practice without the corresponding library.
+(See~\cite{abernathey2021cloud} for further discussion of the benefits
+of Zarr over these monolithic file-oriented formats.)
+In contrast, there are numerous implementations 
+of the Zarr specification, ranging 
+from the mature Zarr-Python~\citep{zarrpython}
 and TensorStore~\citep{tensorstore} implementations
@@
 JK and AM acknowledge the Bill \& Melinda Gates Foundation (INV-001927).
+TM acknowledges funding from The New Zealand Institute for Plant \& Food
+Research Ltd Kiwifruit Royalty Investment Programme.
 
@@
 and Savvy C++ APIs for the data in Figure~\ref{fig-data-storage}.
+Elapsed time is also reported (dotted line). 
 This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
@@
 for the data in Figure~\ref{fig-data-storage}.
+Elapsed time is also reported (dotted line). 
 Genozip did not run for

### 2024-06-19 paper.tex
Minor tweaks, and ack Gil's comments.

@@
 and may enable a diverse ecosystem of next-generation tools for analysing 
-genetic variation data directly from cloud-based object stores.
+genetic variation data directly from cloud-based object stores,
+while maintaining compatibility with existing file-oriented workflows.
 \end{abstract}
@@
 involved and multi-faceted 
-task~\cite{marees2018tutorial,panoutsopoulou2018quality,chen2024genomic}, 
+task~\cite{marees2018tutorial,panoutsopoulou2018quality,chen2024genomic,hemstrom2024next}, 
 often requiring interactive, exploratory analysis
@@
 \section{Acknowledgements}
+We would like to thank Gil McVean for helpful comments on the manuscript.
+
 This research was made possible through access to data in the National Genomic

### 2024-11-14 paper.tex
Add subsection on spruce case study (DRAFT)

@@
 
+\subsection{Case study: 1,063 spruce whole-genome samples}
+
+To demonstrate the versatility of \texttt{vcf2zarr}, in this section
+we include a case study of a dataset originating from a species whose
+genome properties differ substantially from human. The conifer Norway
+spruce, \emph{Picea abies}, is one of the largest and most
+ecologically important species on Earth and is distributed over large
+parts of the Northern hemisphere. One of its features is its large
+genome, consisting of 12 autosomes with a haploid genome assembly size
+of 17.7 Gbp. Here, we assess the performance of \texttt{vcf2zarr} as
+applied to 1,063 whole-genome resequenced spruce individuals from a
+recent study (in preparation). The dataset consists of approximately
+3,745 million single-nucleotide variants and small indels from the 12
+autosomes (15.9 Gbp) totalling 7.4TiB of VCF data after \texttt{bgzip}
+compression, distributed over 165 VCF files. For computational and
+storage reasons, the genotypes only include the `PL` field.
+
+%% Regarding data storage files, which to use?
+%% find spruce.icf -name "*" | wc -l -> 1394080
+%% # chunks in spruce.icf.inspect.txt -> 1342569
+%% du -hs spruce.icf -> 6.8TiB
+%% sum size in spruce.icf.inspect.txt -> 6.8TiB
+%% du -hs spruce.vcz -> 
+%% sum size in spruce.vcz.inspect.txt -> 6.6TiB
+%% # chunks in spruce.vcz.inspect.txt -> 11984581
+%% More chunk files despite smaller input data: could selection of
+%% chunk size matter here?
+Following the conversion process detailed in the previous section, we
+first converted the 165 VCF files (7.4TiB) to ICF, followed by
+conversion to Zarr. The tasks were run as single jobs on a compute
+node with 2,048 GB RAM and 128 cores. Conversion to ICF required
+13h45min runtime. The ICF representation used a total of 6.8 TiB over
+1,394,080 data storage files. Conversion to Zarr required 18h51min
+runtime, generating a dataset with 32 arrays, consuming a total of XXX
+TiB storage over 11,984,581 chunk files. The reduction in total storage (XX)
+is negligible, presumably due to the small sample size. The top fields
+in terms of storage are detailed in Table~\ref{tab-spruce-data}.
+
+\begin{table}
+\caption{Summary for a selection of the largest VCF Zarr columns produced for 
+1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
+default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
+\label{tab-spruce-data}}
+\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
+\toprule
+{Field} & {type} & {storage} & {compress} & {\%total} \\
+  \midrule
+  /call\_PL &	int16 &	6.1T &	12 & 92.75\%\\
+  /call\_genotype & int8 & 282.5G & 26 & 4.18\%\\
+  /call\_genotype\_mask & bool& 42.2G & 180 & 0.62\%\\
+  /variant\_DP4 & int32 & 26.2G & 2.1 & 0.39\%\\
+  /variant\_MQSBZ & float32 & 13.1G & 1.1 & 0.19\%\\
+  /variant\_RPBZ & float32 & 13.0G & 1.1 & 0.19\%\\
+  /variant\_MQBZ & float32 & 13.0G & 1.1 & 0.19\%\\
+  /variant\_SCBZ & float32 & 12.9G & 1.1 & 0.19\%\\
+  /variant\_VDB & float32 & 12.8G & 1.1 & 0.19\%\\
+  /variant\_quality & float32 &12.8G & 1.1 & 0.19\%\\
+  /variant\_BQBZ & float32 & 12.6G & 1.1 & 0.19\%\\
+  /variant\_SGB & float32 & 12.1G & 1.2 & 0.18\%\\
+  /variant\_position & int32 & 9.7G & 1.4 & 0.14\%\\
+  /variant\_AC & int16 & 7.4G & 2.8 & 0.11\%\\
+  /variant\_DP & int32 & 6.7G & 2.1 & 0.10\%\\
+  /variant\_allele & object & 5.4G & 21 & 0.08\%\\
+  /variant\_AN & int16 & 3.1G & 2.3 & 0.05\%\\
+  /call\_genotype\_phased & bool & 2.0G & 1800 & 0.03\%\\
+  /variant\_filter & bool & 1.5G & 2.4 & 0.02\%\\
+  /variant\_MQ & int8 &	489.8M & 7.3 &	0.01\%\\
+%   "/variant_IMF"	"float32"	"161.37 MiB"	89	"0.00%"
+% "/variant_IDV"	"int16"	"121.14 MiB"	59	"0.00%"
+% "/variant_INDEL"	"bool"	"72.6 MiB"	49	"0.00%"
+% "/variant_id"	"object"	"34.95 MiB"	820	"0.00%"
+% "/variant_id_mask"	"bool"	"33.16 MiB"	110	"0.00%"
+% "/variant_contig"	"int16"	"32.44 MiB"	220	"0.00%"
+% "/variant_FS"	"float32"	"32.09 MiB"	450	"0.00%"
+% "/variant_MQ0F"	"float32"	"32.09 MiB"	450	"0.00%"
+% "/contig_length"	"int64"	"6.77 KiB"	1.2	"0.00%"
+% "/contig_id"	"object"	"6.2 KiB"	1.3	"0.00%"
+% "/sample_id"	"object"	"5.46 KiB"	1.5	"0.00%"
+% "/filter_id"	"object"	"4.43 KiB"	0	"0.00%"
+  \bottomrule
+\end{tabular}
+\end{table}
+
+Table~\ref{tab-spruce-data} shows that the dataset storage size is
+dominated by a major component, call\_PL, which accounts for nearly
+93\% of the total. call\_PL is yet another field with low compression potential due to inherent noisiness.
+
+%% FIXME: reference regarding VCF POS?
+The spruce data set poses an interesting challenge to the VCF storage
+format due to the large chromosome sizes (>1Gbp). The VCF POS field is
+encoded in 32-bit format (corresponding to a maximum of approximately
+537Mbp), which mandates splitting up larger chromosomes into chunks,
+adding additional complexity to downstream analyses. Zarr does not
+impose this limitation, and we therefore demonstrate the versatility
+of Zarr by updating contig information and the coordinate system for
+the spruce dataset.
+
 \section{Discussion}
@@
 
-% \subsection{Competing Interests}
+\subsection{Competing Interests}
+JK and BJ are consultants for Genomics England Limited.
+The authors declare that they have no other competing interests.
+
 \subsection{Funding}

### 2024-11-14 paper.tex
Add spruce citation and update text

@@
 parts of the Northern hemisphere. One of its features is its large
-genome, consisting of 12 autosomes with a haploid genome assembly size
-of 17.7 Gbp. Here, we assess the performance of \texttt{vcf2zarr} as
-applied to 1,063 whole-genome resequenced spruce individuals from a
-recent study (in preparation). The dataset consists of approximately
-3,745 million single-nucleotide variants and small indels from the 12
-autosomes (15.9 Gbp) totalling 7.4TiB of VCF data after \texttt{bgzip}
-compression, distributed over 165 VCF files. For computational and
-storage reasons, the genotypes only include the `PL` field.
+genome, consisting of 12 autosomes with a genome size of 19.6
+Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
+performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
+resequenced spruce individuals from a recent study (in preparation),
+where resequencing data has been mapped to a chromosome-scale
+reference genome. The dataset consists of approximately 3,745 million
+single-nucleotide variants and small indels from the 12 autosomes
+totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
+distributed over 165 VCF files. Since the dataset originally was
+generated for downstream analyses that require genotype likelihoods,
+the genotypes include the PL field.
 
@@
 %% sum size in spruce.icf.inspect.txt -> 6.8TiB
-%% du -hs spruce.vcz -> 
+%% du -hs spruce.vcz -> 6.7TiB
 %% sum size in spruce.vcz.inspect.txt -> 6.6TiB
@@
 1,394,080 data storage files. Conversion to Zarr required 18h51min
-runtime, generating a dataset with 32 arrays, consuming a total of XXX
-TiB storage over 11,984,581 chunk files. The reduction in total storage (XX)
-is negligible, presumably due to the small sample size. The top fields
-in terms of storage are detailed in Table~\ref{tab-spruce-data}.
+runtime, generating a dataset with 32 arrays, consuming a total of 6.7
+TiB storage over 11,984,581 chunk files. The reduction in total
+storage (1.1X) is negligible, presumably due to the small sample size.
+The top fields in terms of storage are detailed in
+Table~\ref{tab-spruce-data}.
 

### 2024-11-14 paper.tex

@@
 Following the conversion process detailed in the previous section, we
-first converted the 165 VCF files (7.4TiB) to ICF, followed by
+first converted the 165 VCF files (7.33TiB) to ICF, followed by
 conversion to Zarr. The tasks were run as single jobs on a compute
 node with 2,048 GB RAM and 128 cores. Conversion to ICF required
-13h45min runtime. The ICF representation used a total of 6.8 TiB over
-1,394,080 data storage files. Conversion to Zarr required 18h51min
-runtime, generating a dataset with 32 arrays, consuming a total of 6.7
+13h45min runtime. The ICF representation used a total of 6.77 TiB over
+1,342,569 data storage files. Conversion to Zarr required 18h51min
+runtime, generating a dataset with 32 arrays, consuming a total of 6.6
 TiB storage over 11,984,581 chunk files. The reduction in total
-storage (1.1X) is negligible, presumably due to the small sample size.
-The top fields in terms of storage are detailed in
-Table~\ref{tab-spruce-data}.
+storage (1.1X) is negligible, mainly due to the inherent difficulty in
+compressing the PL field. The top fields in terms of storage are
+detailed in Table~\ref{tab-spruce-data}.
 
@@
   \midrule
-  /call\_PL &	int16 &	6.1T &	12 & 92.75\%\\
-  /call\_genotype & int8 & 282.5G & 26 & 4.18\%\\
-  /call\_genotype\_mask & bool& 42.2G & 180 & 0.62\%\\
-  /variant\_DP4 & int32 & 26.2G & 2.1 & 0.39\%\\
-  /variant\_MQSBZ & float32 & 13.1G & 1.1 & 0.19\%\\
-  /variant\_RPBZ & float32 & 13.0G & 1.1 & 0.19\%\\
-  /variant\_MQBZ & float32 & 13.0G & 1.1 & 0.19\%\\
-  /variant\_SCBZ & float32 & 12.9G & 1.1 & 0.19\%\\
-  /variant\_VDB & float32 & 12.8G & 1.1 & 0.19\%\\
-  /variant\_quality & float32 &12.8G & 1.1 & 0.19\%\\
-  /variant\_BQBZ & float32 & 12.6G & 1.1 & 0.19\%\\
-  /variant\_SGB & float32 & 12.1G & 1.2 & 0.18\%\\
-  /variant\_position & int32 & 9.7G & 1.4 & 0.14\%\\
-  /variant\_AC & int16 & 7.4G & 2.8 & 0.11\%\\
-  /variant\_DP & int32 & 6.7G & 2.1 & 0.10\%\\
-  /variant\_allele & object & 5.4G & 21 & 0.08\%\\
-  /variant\_AN & int16 & 3.1G & 2.3 & 0.05\%\\
-  /call\_genotype\_phased & bool & 2.0G & 1800 & 0.03\%\\
-  /variant\_filter & bool & 1.5G & 2.4 & 0.02\%\\
-  /variant\_MQ & int8 &	489.8M & 7.3 &	0.01\%\\
-%   "/variant_IMF"	"float32"	"161.37 MiB"	89	"0.00%"
-% "/variant_IDV"	"int16"	"121.14 MiB"	59	"0.00%"
-% "/variant_INDEL"	"bool"	"72.6 MiB"	49	"0.00%"
-% "/variant_id"	"object"	"34.95 MiB"	820	"0.00%"
-% "/variant_id_mask"	"bool"	"33.16 MiB"	110	"0.00%"
-% "/variant_contig"	"int16"	"32.44 MiB"	220	"0.00%"
-% "/variant_FS"	"float32"	"32.09 MiB"	450	"0.00%"
-% "/variant_MQ0F"	"float32"	"32.09 MiB"	450	"0.00%"
-% "/contig_length"	"int64"	"6.77 KiB"	1.2	"0.00%"
-% "/contig_id"	"object"	"6.2 KiB"	1.3	"0.00%"
-% "/sample_id"	"object"	"5.46 KiB"	1.5	"0.00%"
-% "/filter_id"	"object"	"4.43 KiB"	0	"0.00%"
+  /call\_PL & int16 & 6.12 TiB & 12.0 & 92.75\% \\
+  /call\_genotype & int8 & 282.45 GiB & 26.0 & 4.18\% \\
+  /call\_genotype\_mask & bool & 42.15 GiB & 180.0 & 0.62\% \\
+  /variant\_DP4 & int32 & 26.24 GiB & 2.1 & 0.39\% \\
+  /variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.19\% \\
+  /variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.19\% \\
+  /variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.19\% \\
+  /variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.19\% \\
+  /variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.19\% \\
+  /variant\_quality & float32 & 12.84 GiB & 1.1 & 0.19\% \\
+  /variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.19\% \\
+  /variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.18\% \\
+  /variant\_position & int32 & 9.65 GiB & 1.4 & 0.14\% \\
+  /variant\_AC & int16 & 7.42 GiB & 2.8 & 0.11\% \\
+  /variant\_DP & int32 & 6.71 GiB & 2.1 & 0.10\% \\
+  /variant\_allele & object & 5.37 GiB & 21.0 & 0.08\% \\
+  /variant\_AN & int16 & 3.05 GiB & 2.3 & 0.05\% \\
+  /call\_genotype\_phased & bool & 2.03 GiB & 1800.0 & 0.03\% \\
+  /variant\_filter & bool & 1.46 GiB & 2.4 & 0.02\% \\
+  /variant\_MQ & int8 & 489.83 MiB & 7.3 & 0.01\% \\
   \bottomrule

### 2024-11-15 paper.tex
Add Will Tyler to authorship list for preprint update

@@
 \author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
-\author[4,\authfn{1}]{Tom White}
+\author[4,\authfn{1}]{Will Tyler}
+\author[5,\authfn{1}]{Tom White}
 
 % Middle
-\author[5]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[6]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[7]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[7]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[8]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[9]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
@@
 \author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[5,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[6,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 \affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
-\affil[4]{Tom White Consulting Ltd.}
-\affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
+\affil[4]{Independent researcher}
+\affil[5]{Tom White Consulting Ltd.}
+\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
 University of Oxford, UK}
-\affil[6]{Wellcome Sanger Institute}
-\affil[7]{Genomics England}
-\affil[8]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[7]{Wellcome Sanger Institute}
+\affil[8]{Genomics England}
+\affil[9]{School of Computer Science, McGill University, Montreal, QC, Canada}
 
@@
 
-\subsection{Case study: 1,063 spruce whole-genome samples}
-
-To demonstrate the versatility of \texttt{vcf2zarr}, in this section
-we include a case study of a dataset originating from a species whose
-genome properties differ substantially from human. The conifer Norway
-spruce, \emph{Picea abies}, is one of the largest and most
-ecologically important species on Earth and is distributed over large
-parts of the Northern hemisphere. One of its features is its large
-genome, consisting of 12 autosomes with a genome size of 19.6
-Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
-performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
-resequenced spruce individuals from a recent study (in preparation),
-where resequencing data has been mapped to a chromosome-scale
-reference genome. The dataset consists of approximately 3,745 million
-single-nucleotide variants and small indels from the 12 autosomes
-totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
-distributed over 165 VCF files. Since the dataset originally was
-generated for downstream analyses that require genotype likelihoods,
-the genotypes include the PL field.
-
-%% Regarding data storage files, which to use?
-%% find spruce.icf -name "*" | wc -l -> 1394080
-%% # chunks in spruce.icf.inspect.txt -> 1342569
-%% du -hs spruce.icf -> 6.8TiB
-%% sum size in spruce.icf.inspect.txt -> 6.8TiB
-%% du -hs spruce.vcz -> 6.7TiB
-%% sum size in spruce.vcz.inspect.txt -> 6.6TiB
-%% # chunks in spruce.vcz.inspect.txt -> 11984581
-%% More chunk files despite smaller input data: could selection of
-%% chunk size matter here?
-Following the conversion process detailed in the previous section, we
-first converted the 165 VCF files (7.33TiB) to ICF, followed by
-conversion to Zarr. The tasks were run as single jobs on a compute
-node with 2,048 GB RAM and 128 cores. Conversion to ICF required
-13h45min runtime. The ICF representation used a total of 6.77 TiB over
-1,342,569 data storage files. Conversion to Zarr required 18h51min
-runtime, generating a dataset with 32 arrays, consuming a total of 6.6
-TiB storage over 11,984,581 chunk files. The reduction in total
-storage (1.1X) is negligible, mainly due to the inherent difficulty in
-compressing the PL field. The top fields in terms of storage are
-detailed in Table~\ref{tab-spruce-data}.
-
-\begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns produced for 
-1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
-default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
-\label{tab-spruce-data}}
-\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
-\toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
-  \midrule
-  /call\_PL & int16 & 6.12 TiB & 12.0 & 92.75\% \\
-  /call\_genotype & int8 & 282.45 GiB & 26.0 & 4.18\% \\
-  /call\_genotype\_mask & bool & 42.15 GiB & 180.0 & 0.62\% \\
-  /variant\_DP4 & int32 & 26.24 GiB & 2.1 & 0.39\% \\
-  /variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.19\% \\
-  /variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.19\% \\
-  /variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.19\% \\
-  /variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.19\% \\
-  /variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.19\% \\
-  /variant\_quality & float32 & 12.84 GiB & 1.1 & 0.19\% \\
-  /variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.19\% \\
-  /variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.18\% \\
-  /variant\_position & int32 & 9.65 GiB & 1.4 & 0.14\% \\
-  /variant\_AC & int16 & 7.42 GiB & 2.8 & 0.11\% \\
-  /variant\_DP & int32 & 6.71 GiB & 2.1 & 0.10\% \\
-  /variant\_allele & object & 5.37 GiB & 21.0 & 0.08\% \\
-  /variant\_AN & int16 & 3.05 GiB & 2.3 & 0.05\% \\
-  /call\_genotype\_phased & bool & 2.03 GiB & 1800.0 & 0.03\% \\
-  /variant\_filter & bool & 1.46 GiB & 2.4 & 0.02\% \\
-  /variant\_MQ & int8 & 489.83 MiB & 7.3 & 0.01\% \\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-Table~\ref{tab-spruce-data} shows that the dataset storage size is
-dominated by a major component, call\_PL, which accounts for nearly
-93\% of the total. call\_PL is yet another field with low compression potential due to inherent noisiness.
-
-%% FIXME: reference regarding VCF POS?
-The spruce data set poses an interesting challenge to the VCF storage
-format due to the large chromosome sizes (>1Gbp). The VCF POS field is
-encoded in 32-bit format (corresponding to a maximum of approximately
-537Mbp), which mandates splitting up larger chromosomes into chunks,
-adding additional complexity to downstream analyses. Zarr does not
-impose this limitation, and we therefore demonstrate the versatility
-of Zarr by updating contig information and the coordinate system for
-the spruce dataset.
-
 \section{Discussion}
@@
 \section{Acknowledgements}
-We would like to thank Gil McVean for helpful comments on the manuscript.
-
+% We would like to thank Gil McVean for helpful comments on the manuscript.
 This research was made possible through access to data in the National Genomic

### 2024-11-18 paper.tex
Remove comments regarding data storage files
Details are now taken from notebooks

--- 2024-11-14 paper.tex
@@
 
-%% Regarding data storage files, which to use?
-%% find spruce.icf -name "*" | wc -l -> 1394080
-%% # chunks in spruce.icf.inspect.txt -> 1342569
-%% du -hs spruce.icf -> 6.8TiB
-%% sum size in spruce.icf.inspect.txt -> 6.8TiB
-%% du -hs spruce.vcz -> 6.7TiB
-%% sum size in spruce.vcz.inspect.txt -> 6.6TiB
-%% # chunks in spruce.vcz.inspect.txt -> 11984581
-%% More chunk files despite smaller input data: could selection of
-%% chunk size matter here?
 Following the conversion process detailed in the previous section, we

### 2024-11-21 paper.tex

@@
 \author[9]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[10]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 
@@
 \affil[9]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[10]{Department of Cell and Molecular Biology, National
+  Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
+  Uppsala University, Uppsala, Sweden}
 
@@
   \midrule
-  /call\_PL & int16 & 6.12 TiB & 12.0 & 92.75\% \\
-  /call\_genotype & int8 & 282.45 GiB & 26.0 & 4.18\% \\
-  /call\_genotype\_mask & bool & 42.15 GiB & 180.0 & 0.62\% \\
-  /variant\_DP4 & int32 & 26.24 GiB & 2.1 & 0.39\% \\
-  /variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.19\% \\
-  /variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.19\% \\
-  /variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.19\% \\
-  /variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.19\% \\
-  /variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.19\% \\
-  /variant\_quality & float32 & 12.84 GiB & 1.1 & 0.19\% \\
-  /variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.19\% \\
-  /variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.18\% \\
-  /variant\_position & int32 & 9.65 GiB & 1.4 & 0.14\% \\
-  /variant\_AC & int16 & 7.42 GiB & 2.8 & 0.11\% \\
-  /variant\_DP & int32 & 6.71 GiB & 2.1 & 0.10\% \\
-  /variant\_allele & object & 5.37 GiB & 21.0 & 0.08\% \\
-  /variant\_AN & int16 & 3.05 GiB & 2.3 & 0.05\% \\
-  /call\_genotype\_phased & bool & 2.03 GiB & 1800.0 & 0.03\% \\
-  /variant\_filter & bool & 1.46 GiB & 2.4 & 0.02\% \\
-  /variant\_MQ & int8 & 489.83 MiB & 7.3 & 0.01\% \\
+  /call\_PL & int16 & 6.1 TiB & 12.0 & 92.75\% \\
+  /call\_genotype & int8 & 282.5 GiB & 26.0 & 4.18\% \\
+  /call\_genotype\_mask & bool & 42.2 GiB & 180.0 & 0.62\% \\
+  /variant\_DP4 & int32 & 26.2 GiB & 2.1 & 0.39\% \\
+  /variant\_MQSBZ & float32 & 13.1 GiB & 1.1 & 0.19\% \\
+  /variant\_RPBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
+  /variant\_MQBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
+  /variant\_SCBZ & float32 & 12.9 GiB & 1.1 & 0.19\% \\
+  /variant\_VDB & float32 & 12.8 GiB & 1.1 & 0.19\% \\
+  /variant\_quality & float32 & 12.8 GiB & 1.1 & 0.19\% \\
+  /variant\_BQBZ & float32 & 12.6 GiB & 1.1 & 0.19\% \\
+  /variant\_SGB & float32 & 12.1 GiB & 1.2 & 0.18\% \\
+  /variant\_position & int32 & 9.7 GiB & 1.4 & 0.14\% \\
+  /variant\_AC & int16 & 7.4 GiB & 2.8 & 0.11\% \\
+  /variant\_DP & int32 & 6.7 GiB & 2.1 & 0.10\% \\
+  /variant\_allele & object & 5.4 GiB & 21.0 & 0.08\% \\
+  /variant\_AN & int16 & 3.1 GiB & 2.3 & 0.05\% \\
+  /call\_genotype\_phased & bool & 2.0 GiB & 1800.0 & 0.03\% \\
+  /variant\_filter & bool & 1.5 GiB & 2.4 & 0.02\% \\
+  /variant\_MQ & int8 & 489.8 MiB & 7.3 & 0.01\% \\
   \bottomrule
@@
 dominated by a major component, call\_PL, which accounts for nearly
-93\% of the total. call\_PL is yet another field with low compression potential due to inherent noisiness.
+93\% of the total. call\_PL is yet another field with low compression
+potential due to inherent noisiness.
 
@@
 Research Ltd Kiwifruit Royalty Investment Programme.
-
+PU was supported by the SciLifeLab \& Wallenberg Data Driven Life
+Science Program, Knut and Alice Wallenberg Foundation (grants: KAW
+2020.0239 and KAW 2017.0003), and by the National Bioinformatics
+Infrastructure Sweden (NBIS) at SciLifeLab
 % \subsection{Author's Contributions}
@@
 not necessarily those of the NHS, the NIHR or the Department of Health.
+
+Computation for the Spruce case study were enabled by resources
+provided by the National Academic Infrastructure for Supercomputing in
+Sweden (NAISS), partially funded by the Swedish Research Council
+through grant agreement no. 2022-06725.
 

### 2025-01-08 paper.tex
Add BAI/VCF POS paragraphs
- exemplify old indexing formats that are ill suited for large genomes
- motivate why the VCFs were partitioned
- mention VCF POS limitation and how this could affect other genomes
  with large chromosomes

@@
 
-%% FIXME: reference regarding VCF POS?
-The spruce data set poses an interesting challenge to the VCF storage
-format due to the large chromosome sizes (>1Gbp). The VCF POS field is
-encoded in 32-bit format (corresponding to a maximum of approximately
-537Mbp), which mandates splitting up larger chromosomes into chunks,
-adding additional complexity to downstream analyses. Zarr does not
-impose this limitation, and we therefore demonstrate the versatility
-of Zarr by updating contig information and the coordinate system for
-the spruce dataset.
+Large genomes from non-model organisms pose particular problems for
+analysis workflows that often are tailored for human conditions. For
+instance, the binary Sequence Alignment/Map index format BAI has a 512
+Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
+The spruce BAM files used for variant calling were BAI indexed, which
+motivated the partitioning into multiple VCF files in the first place,
+adding complexity to downstream processing.
+
+A similar limitation is imposed by the VCF POS field which is encoded
+in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
+spruce does not have chromosomes that exceed the limit, other species
+do, such as the recently sequence mistletoe, with chromosomes larger
+than 8 Gbp. Zarr does not impose this limitation, and we therefore
+demonstrate the versatility of Zarr by updating contig information and
+the coordinate system for the spruce dataset. This updating procedure
+could be applied to any other species.
 

### 2025-01-08 paper.tex
Add BAI/VCF POS paragraphs
- exemplify old indexing formats that are ill suited for large genomes
- motivate why the VCFs were partitioned
- mention VCF POS limitation and how this could affect other genomes
  with large chromosomes

@@
 variants without phase information.
-The PLINK 2 PGEN format~\citep{pgen2024} is more general
+The PLINK 2 PGEN format~\citep{rivas2024efficient} is more general
 and compact than PLINK, compressing variant data using specialised
@@
 of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
-benchmarks in the following sections. 
-(See \citep{durbin2020task,moore2021ome,gowan2022using} for Zarr
-benchmarks in cloud settings.)
+benchmarks in the following sections before presenting cloud-based benchmarks. 
 We compare primarily with
@@
 space (29.9GiB up from 22.1GiB for 1 million samples; data not
-shown). Given the relatively small contribution of genotypes to the
+shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
+decoding speed, but at the cost of much larger storage requirements 
+(219GiB; data not shown).
+Given the relatively small contribution of genotypes to the
 overall storage of real datasets (see the Genomics England example)
-and the frequency that they are likely to be accessed, this
-would seem like a good tradeoff in most cases.
+and the frequency that they are likely to be accessed, using 
+Zstandard without BitShuffle 
+would seem like a good tradeoff in many cases.
 This ability to easily tune compression performance
@@
 % has to be the file system checking cache coherency.]
+% [Note BJ: The compute is on EC2 and storage is on-prem, I believe there is
+% no ec2-side cache and the link is saturated, hence the latency.]
 
@@
 
-Finally, to illustrate performance on a common filtering task, we 
+Additionally, to illustrate performance on a common filtering task, we 
 created a copy of the VCF chunk which contains only variants 
@@
 datasets is a potentially transformational change enabled by Zarr.
+
+Finally, to explore the use of cloud computing services with VCF Zarr,
+the zarr files for chromosome 2 were uploaded to an AWS S3 standard bucket.
+Using a single-threaded process on an EC2 instance the benchmarks above were
+repeated. Extracting the position array required 15s CPU, 23.8s elapsed.
+The af-dist calculation required 2.70min CPU, 5.23min elapsed. Creating the filter mask
+required 3.05min CPU, 11.15min elapsed. These times are comparable to the on-premises
+HPC system above.
+
+This single-threaded approach is not optimal for cloud computing, as to optimise
+both runtime and costs parallel processing is required. Using 64 parallel processes
+on a single c5.24xlarge EC2 instance with 96 vCPUs (48 physical) the zarr chunks of
+genotype array could be retrived from S3 and decompressed at a rate of 50.8 GB/s.
+The network data rate was 127 MB/s, far below the capbility of the instance,
+indicating that this process was compute-constrained. When calculating af-dist after
+decompression the throughput was 24.6 GB/s. With this parallel approach the 
+af-dist calculation above was then completed in 4.12s elapsed time.
+This level of throughput would complete af-dist for all chromosomes for all samples
+% compute: $4.08 * 1.27
+% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
+in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
+(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
+benchmarks in cloud settings.)
+
+\subsection{Case study: All of Us exome-like data} 
+The All of Us dataset comprises 245,388 clinical-grade whole genome sequences (WGS), 
+identifying over 1 billion variants, including 3.9 million novel variants 
+with coding consequences~\citep{all2024genomic}. Participants are from the United States, 
+with an emphasis on historically underrepresented genetic ancestry groups. 
+The genomic data, generated through short-read sequencing on the Illumina NovaSeq 6000, 
+is of high quality, with an average coverage depth of 30x, and is linked to extensive 
+health-related information, providing a valuable resource for biomedical research. 
+WGS analysis is conducted on the Illumina DRAGEN platform, with alignment performed 
+using the GRCh38dh reference genome. GTC files are converted to VCF format using 
+Picard GtcToVcf, and every step of the WGS procedure is rigorously controlled by 
+predefined quality control measures.
+
+For this study, we applied the bio2zarr compression to the chromosome 20 exome-like 
+data from the All of Us dataset. The exome-like data is constructed by extracting from all WGS samples
+the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
+transcripts. This dataset, stored as a compressed VCF file of 7.44 GiB, 
+includes 715,256 variants, of which 746 are classified as pathogenic or likely pathogenic.
+
+\begin{table}
+\caption{Summary for a selection of the largest VCF Zarr columns produced 
+the from the 245,388 exome-like samples from chromosome 20
+using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
+10,000 samples.
+For column details see the caption of Table~\ref{tab-genomics-england-data}.
+\label{tab-all-of-us-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
+\toprule
+{Field} & {type} & {storage} & {compress} & {\%total} \\
+\midrule
+/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
+/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
+/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
+/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
+/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
+/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
+/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
+% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
+% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
+% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
+% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
+% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
+% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
+% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
+% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
+% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
+% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
+% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
+% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 and xtensor~\citep{xtensor_zarr}
-as well as  standalone libraries for JavaScript~\cite{zarrjs},
+as well as  standalone libraries for 
+C++~\citep{ruan2024image}, JavaScript~\cite{zarrjs},
 Julia~\cite{zarrjl}, Rust~\citep{zarrs}
@@
 tools~\cite{moore2023ome,rzepka2023toward}. 
-Zarr has also seen recent uptake 
-in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
+Zarr has also seen recent uptake in other large-scale processing for
+microscopy applications~\citep{ruan2024image}, as well as 
+single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
 and multimodal spatial omics
@@
 This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
-3.9 GiB/s Zarr (Zstd), and 6.6 GiB/s for Savvy. 
+3.9 GiB/s Zarr (Zstd), 4.1 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
 \label{fig-whole-matrix-decode}}

### 2025-01-09 src/plot.py
Add zarr versions plot

@@ -20,6 +20,9 @@ genozip_colour = "tab:purple"
 zarr_colour = "tab:blue"
 zarr_nshf_colour = "tab:cyan"
 two_bit_colour = "tab:pink"
+zarr_java_colour = "tab:olive"
+ts_py_colour = "tab:brown"
+ts_cpp_colour = "tab:gray"
 
 
 def one_panel_fig(**kwargs):
@@ -105,6 +108,9 @@ def plot_total_cpu(
             "zarr": zarr_colour,
             "zarr_nshf": zarr_nshf_colour,
             "savvy": sav_colour,
+            "zarr_java": zarr_java_colour,
+            "ts_py": ts_py_colour,
+            "ts_cpp": ts_cpp_colour,
         }
     have_genozip = False
     toolname = {} if toolname is None else toolname
@@ -259,6 +265,31 @@ def whole_matrix_compute(time_data, output):
 
     plt.savefig(output)
 
+@click.command()
+@click.argument("time_data", type=click.File("r"))
+@click.argument("output", type=click.Path())
+def whole_matrix_compute_zarr_versions(time_data, output):
+    """
+    Plot the figure showing compute performance on whole-matrix afdist for the zarr implementations.
+    """
+    df = pd.read_csv(time_data, index_col=False).sort_values("num_samples")
+    df = df[df.storage == "hdd"]
+
+    fig, ax1 = one_panel_fig()
+    name_map = {
+        "zarr": "zarr-python API",
+        "zarr_java": "zarr-java API",
+        "ts_py": "tensorstore Python API",
+        "ts_cpp": "tensorstore C++ API",
+    }
+    plot_total_cpu(
+        ax1,
+        df,
+        name_map,
+        order=["zarr", "zarr_java", "ts_py", "ts_cpp"],
+    )
+
+    plt.savefig(output)
 
 @click.command()
 @click.argument("time_data", type=click.File("r"))
@@ -543,6 +574,7 @@ def cli():
 
 cli.add_command(data_scaling)
 cli.add_command(whole_matrix_compute)
+cli.add_command(whole_matrix_compute_zarr_versions)
 cli.add_command(whole_matrix_decode)
 cli.add_command(column_extract)
 cli.add_command(subset_matrix_compute)

### 2025-01-10 src/plot.py
Add GeL throughput figure

@@ -86,6 +86,39 @@ def plot_size(ax, df, label_y_offset=None, order=None):
     ax.set_ylabel("Storage size (bytes)")
     plt.tight_layout()
 
+def plot_s3_throughput(ax, df):
+    ax.loglog(
+        df["processes"].values,
+        df["throughput_decompress"].values,
+        marker=".",
+        label="Decompression only"
+    )
+    ax.annotate(
+        f"{df['throughput_decompress'][6]:.0f} MB/s",
+        textcoords="offset points",
+        xytext=(-8, -10),
+        xy=(df['processes'][6], df['throughput_decompress'][6]),
+        xycoords="data",
+    )
+    ax.loglog(
+        df["processes"].values,
+        df["throughput_afdist"].values,
+        marker=".",
+        label="AF Dist"
+    )
+    ax.annotate(
+        f"{df['throughput_afdist'][6]:.0f} MB/s",
+        textcoords="offset points",
+        xytext=(-8, -10),
+        xy=(df['processes'][6], df['throughput_afdist'][6]),
+        xycoords="data",
+    )
+
+    ax.legend()
+    ax.set_xlabel("Number of processes")
+    ax.set_ylabel("Genotype data throughput (MB/s)")
+    plt.tight_layout()
+
 
 def plot_total_cpu(
     ax,
@@ -227,6 +260,23 @@ def data_scaling(size_data, output):
     # I tried putting an inset axis showing the ratio, but it was too small.
     # ax_inset = ax1.inset_axes([0.70, 0.1, 0.25, 0.25])
     # ax_inset.semilogx(sav["num_samples"], ratio)
+
+    plt.savefig(output)
+
+@click.command()
+@click.argument("data", type=click.File("r"))
+@click.argument("output", type=click.Path())
+def s3_throughput(data, output):
+    """
+    Plot the figure showing compute performance on whole-matrix afdist.
+    """
+    df = pd.read_csv(data, index_col=False).sort_values("processes")
+
+    fig, ax1 = one_panel_fig()
+    plot_s3_throughput(
+        ax1,
+        df,
+    )
 
     plt.savefig(output)
 
@@ -542,6 +592,7 @@ def cli():
 
 
 cli.add_command(data_scaling)
+cli.add_command(s3_throughput)
 cli.add_command(whole_matrix_compute)
 cli.add_command(whole_matrix_decode)
 cli.add_command(column_extract)

### 2025-01-12 saige_prototype/cleanup.sh
Add notebook to demonstrate SAIGE prototype

@@ -1,0 +1,9 @@
+#!/bin/bash
+
+export PATH="/opt/miniconda3/bin:$PATH"
+
+rm -rf SAIGE
+rm SAIGE_1.3.6.tar.gz
+rm chr21_10_4.*
+rm chr21_10_5.*
+conda env remove -n RSAIGE -y

### 2025-01-12 saige_prototype/setup.sh

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+if [ ! -d SAIGE ]; then
+    git clone https://github.com/Will-Tyler/SAIGE.git
+fi
+
+export PATH="/opt/miniconda3/bin:$PATH"
+
+if ! conda list -n RSAIGE; then
+    CONDA_SUBDIR=osx-64 conda env create -f env.yml
+fi
+
+conda activate RSAIGE
+
+R CMD build SAIGE
+R CMD INSTALL SAIGE

### 2025-01-12 src/collect_data.py
Add decoding data for LZ4

@@ -315,7 +315,7 @@ def zarr_afdist_worker(ds_path, debug, conn, azure=False):
 
         service_client = BlobServiceClient.from_connection_string(connect_str)
         container_client = service_client.get_container_client("data")
-        
+
         store = zarr.storage.ABSStore(client=container_client)
 
     assert azure is False
@@ -621,6 +621,15 @@ decode_tools = all_tools[:2] + [
     Tool(
         "zarr_nshf",
         ".noshuffle.zarr",
+        run_zarr_afdist,
+        run_zarr_afdist_subset,
+        None,
+        run_zarr_decode,
+        run_zarr_pos_extract,
+    ),
+    Tool(
+        "zarr_lz4_nshf",
+        ".lz4.noshuffle.zarr",
         run_zarr_afdist,
         run_zarr_afdist_subset,
         None,

### 2025-01-13 paper.tex
Add text on S3

@@
 datasets is a potentially transformational change enabled by Zarr.
+
+Finally, to explore the use of cloud computing services with VCF Zarr,
+the zarr files for chromosome 2 were uploaded to an AWS S3 standard bucket.
+Using a single-threaded process on an EC2 instance the benchmarks above were
+repeated. Extracting the position array required 15s CPU, 23.8s elapsed.
+The af-dist calculation required 2.70min CPU, 5.23min elapsed. Creating the filter mask
+required 3.05min CPU, 11.15min elapsed. These times are comparable to the on-premises
+HPC system above.
+
+This single-threaded approach is not optimal for cloud computing, as to optimise
+both runtime and costs parallel processing is required. Using 64 parallel processes
+on a single c5.24xlarge EC2 instance with 96 vCPUs (48 physical) the zarr chunks of
+genotype array could be retrived from S3 and decompressed at a rate of 50.8 GB/s.
+The network data rate was 127 MB/s, far below the capbility of the instance,
+indicating that this process was compute-constrained. When calculating af-dist after
+decompression the throughput was 24.6 GB/s. With this parallel approach the 
+af-dist calculation above was then completed in 4.12s elapsed time.
+This level of throughput would complete af-dist for all chromosomes for all samples
+% compute: $4.08 * 1.27
+% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
+in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
+(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
+benchmarks in cloud settings.)
 

### 2025-01-15 paper.tex

@@
 
+\subsection{Case study: All of Us exome-like data} 
+The All of Us dataset comprises 245,388 clinical-grade whole genome sequences,
+identifying over 1 billion variants, including 3.9 million novel variants with
+coding consequences~\citep{all2024genomic}. 
+Participants are from the United States, with an emphasis
+on historically underrepresented genetic ancestry groups. The genomic data is
+of high quality, with an average coverage depth of 30x, and is linked to
+extensive health-related information, providing a valuable resource for
+biomedical research. For this study, we applied the bio2zarr compression method
+to the exome-like data for chromosome 20 from All of Us. This dataset, stored as a
+compressed VCF file of 7.44 GiB, includes 681,000 variants, of which 746 are
+classified as pathogenic or likely pathogenic.
+
+\begin{table}
+\caption{Summary for a selection of the largest VCF Zarr columns produced 
+the from the 245,388 exome-like samples from chromosome 20
+using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
+10,000 samples.
+For column details see the caption of Table~\ref{tab-genomics-england-data}.
+\label{tab-all-of-us-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
+\toprule
+{Field} & {type} & {storage} & {compress} & {\%total} \\
+\midrule
+/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
+/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
+/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
+/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
+/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
+/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
+/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
+% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
+% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
+% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
+% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
+% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
+% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
+% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
+% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
+% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
+% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
+% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
+% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \subsection{Case study: 1,063 spruce whole-genome samples}

### 2025-01-15 paper.tex
Correct and add details to AoU text

@@
 \subsection{Case study: All of Us exome-like data} 
-The All of Us dataset comprises 245,388 clinical-grade whole genome sequences,
-identifying over 1 billion variants, including 3.9 million novel variants with
-coding consequences~\citep{all2024genomic}. 
-Participants are from the United States, with an emphasis
-on historically underrepresented genetic ancestry groups. The genomic data is
-of high quality, with an average coverage depth of 30x, and is linked to
-extensive health-related information, providing a valuable resource for
-biomedical research. For this study, we applied the bio2zarr compression method
-to the exome-like data for chromosome 20 from All of Us. This dataset, stored as a
-compressed VCF file of 7.44 GiB, includes 681,000 variants, of which 746 are
-classified as pathogenic or likely pathogenic.
+The All of Us dataset comprises 245,388 clinical-grade whole genome sequences (WGS), 
+identifying over 1 billion variants, including 3.9 million novel variants 
+with coding consequences~\citep{all2024genomic}. Participants are from the United States, 
+with an emphasis on historically underrepresented genetic ancestry groups. 
+The genomic data, generated through short-read sequencing on the Illumina NovaSeq 6000, 
+is of high quality, with an average coverage depth of 30x, and is linked to extensive 
+health-related information, providing a valuable resource for biomedical research. 
+WGS analysis is conducted on the Illumina DRAGEN platform, with alignment performed 
+using the GRCh38dh reference genome. GTC files are converted to VCF format using 
+Picard GtcToVcf, and every step of the WGS procedure is rigorously controlled by 
+predefined quality control measures.
+
+For this study, we applied the bio2zarr compression to the chromosome 20 exome-like 
+data from the All of Us dataset. The exome-like data is constructed by extracting from the WGS 
+the SNP and Indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
+transcripts for all samples. This dataset, stored as a compressed VCF file of 7.44 GiB, 
+includes 715,256 variants, of which 746 are classified as pathogenic or likely pathogenic.
 

### 2025-01-15 paper.tex
Minor change in AoU text

@@
 data from the All of Us dataset. The exome-like data is constructed by extracting from the WGS 
-the SNP and Indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
+the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
 transcripts for all samples. This dataset, stored as a compressed VCF file of 7.44 GiB, 

### 2025-01-15 paper.tex
Minor change in AoU text

@@
 For this study, we applied the bio2zarr compression to the chromosome 20 exome-like 
-data from the All of Us dataset. The exome-like data is constructed by extracting from the WGS 
+data from the All of Us dataset. The exome-like data is constructed by extracting from all WGS samples
 the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
-transcripts for all samples. This dataset, stored as a compressed VCF file of 7.44 GiB, 
+transcripts. This dataset, stored as a compressed VCF file of 7.44 GiB, 
 includes 715,256 variants, of which 746 are classified as pathogenic or likely pathogenic.

### 2025-01-16 src/collect_data.py
Add zip data and a rule to create 10kx1k data for Zarr

@@ -597,7 +597,7 @@ def file_size(src, output, debug):
             "vcf": vcf_path,
             "bcf": bcf_path,
             "zarr": zarr_path,
-            "zarr_nshf": ts_path.with_suffix(".noshuffle.zarr"),
+            "zarr.zip": ts_path.with_suffix(".zarr.zip"),
             "sav": sav_path,
             "genozip": genozip_path,
         }

### 2025-01-17 paper.tex
add OFH subsection

@@
 variants without phase information.
-The PLINK 2 PGEN format~\citep{rivas2024efficient} is more general
+The PLINK 2 PGEN format~\citep{pgen2024} is more general
 and compact than PLINK, compressing variant data using specialised
@@
 of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
-benchmarks in the following sections before presenting cloud-based benchmarks. 
+benchmarks in the following sections. 
+(See \citep{durbin2020task,moore2021ome,gowan2022using} for Zarr
+benchmarks in cloud settings.)
 We compare primarily with
@@
 space (29.9GiB up from 22.1GiB for 1 million samples; data not
-shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
-decoding speed, but at the cost of much larger storage requirements 
-(219GiB; data not shown).
-Given the relatively small contribution of genotypes to the
+shown). Given the relatively small contribution of genotypes to the
 overall storage of real datasets (see the Genomics England example)
-and the frequency that they are likely to be accessed, using 
-Zstandard without BitShuffle 
-would seem like a good tradeoff in many cases.
+and the frequency that they are likely to be accessed, this
+would seem like a good tradeoff in most cases.
 This ability to easily tune compression performance
@@
 % has to be the file system checking cache coherency.]
-% [Note BJ: The compute is on EC2 and storage is on-prem, I believe there is
-% no ec2-side cache and the link is saturated, hence the latency.]
 
@@
 
-Additionally, to illustrate performance on a common filtering task, we 
+Finally, to illustrate performance on a common filtering task, we 
 created a copy of the VCF chunk which contains only variants 
@@
 
-Finally, to explore the use of cloud computing services with VCF Zarr,
-the zarr files for chromosome 2 were uploaded to an AWS S3 standard bucket.
-Using a single-threaded process on an EC2 instance the benchmarks above were
-repeated. Extracting the position array required 15s CPU, 23.8s elapsed.
-The af-dist calculation required 2.70min CPU, 5.23min elapsed. Creating the filter mask
-required 3.05min CPU, 11.15min elapsed. These times are comparable to the on-premises
-HPC system above.
-
-This single-threaded approach is not optimal for cloud computing, as to optimise
-both runtime and costs parallel processing is required. Using 64 parallel processes
-on a single c5.24xlarge EC2 instance with 96 vCPUs (48 physical) the zarr chunks of
-genotype array could be retrived from S3 and decompressed at a rate of 50.8 GB/s.
-The network data rate was 127 MB/s, far below the capbility of the instance,
-indicating that this process was compute-constrained. When calculating af-dist after
-decompression the throughput was 24.6 GB/s. With this parallel approach the 
-af-dist calculation above was then completed in 4.12s elapsed time.
-This level of throughput would complete af-dist for all chromosomes for all samples
-% compute: $4.08 * 1.27
-% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
-in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
-(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
-benchmarks in cloud settings.)
-
-\subsection{Case study: All of Us exome-like data} 
-The All of Us dataset comprises 245,388 clinical-grade whole genome sequences (WGS), 
-identifying over 1 billion variants, including 3.9 million novel variants 
-with coding consequences~\citep{all2024genomic}. Participants are from the United States, 
-with an emphasis on historically underrepresented genetic ancestry groups. 
-The genomic data, generated through short-read sequencing on the Illumina NovaSeq 6000, 
-is of high quality, with an average coverage depth of 30x, and is linked to extensive 
-health-related information, providing a valuable resource for biomedical research. 
-WGS analysis is conducted on the Illumina DRAGEN platform, with alignment performed 
-using the GRCh38dh reference genome. GTC files are converted to VCF format using 
-Picard GtcToVcf, and every step of the WGS procedure is rigorously controlled by 
-predefined quality control measures.
-
-For this study, we applied the bio2zarr compression to the chromosome 20 exome-like 
-data from the All of Us dataset. The exome-like data is constructed by extracting from all WGS samples
-the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
-transcripts. This dataset, stored as a compressed VCF file of 7.44 GiB, 
-includes 715,256 variants, of which 746 are classified as pathogenic or likely pathogenic.
-
-\begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns produced 
-the from the 245,388 exome-like samples from chromosome 20
-using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
-10,000 samples.
-For column details see the caption of Table~\ref{tab-genomics-england-data}.
-\label{tab-all-of-us-data}}
-\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
-\toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
-\midrule
-/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
-/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
-/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
-/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
-/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
-/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
-/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
-% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
-% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
-% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
-% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
-% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
-% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
-% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
-% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
-% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
-% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
-% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
-% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
-\bottomrule
-\end{tabular}
-\end{table}
+\subsection{Case study: Our Future Health genotype release}
+
+Our Future Health is the UK’s largest ever health research programme. 
+It is designed to help people live healthier lives for longer through the 
+discovery and testing of more effective approaches to prevention, earlier 
+detection and treatment of diseases.
+Our Future Health is inviting millions of people, from all backgrounds and 
+from right across the UK to take part. Volunteers are providing information 
+about their health and lifestyles to create an incredibly detailed picture 
+that represents the whole of the UK.
+
+% this section is probably better placed in acknowledgements?
+This study makes use of de-identified data held by Our Future Health. 
+We would like to acknowledge all the research participants who have donated 
+their data to the Our Future Health research programme 
+\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
+
+Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
+was made available in December 2024. The genetic component includes genotype data 
+from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
+chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
+three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
+10,221 variants.
 
@@
 and xtensor~\citep{xtensor_zarr}
-as well as  standalone libraries for 
-C++~\citep{ruan2024image}, JavaScript~\cite{zarrjs},
+as well as  standalone libraries for JavaScript~\cite{zarrjs},
 Julia~\cite{zarrjl}, Rust~\citep{zarrs}
@@
 tools~\cite{moore2023ome,rzepka2023toward}. 
-Zarr has also seen recent uptake in other large-scale processing for
-microscopy applications~\citep{ruan2024image}, as well as 
-single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
+Zarr has also seen recent uptake 
+in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
 and multimodal spatial omics
@@
 This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
-3.9 GiB/s Zarr (Zstd), 4.1 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
+3.9 GiB/s Zarr (Zstd), and 6.6 GiB/s for Savvy. 
 \label{fig-whole-matrix-decode}}

### 2025-01-17 paper.tex
add citation
Add citation and update statement on coordinate transformation

@@
 
-\subsection{Case study: Our Future Health genotype release}
-
-Our Future Health is the UK’s largest ever health research programme. 
-It is designed to help people live healthier lives for longer through the 
-discovery and testing of more effective approaches to prevention, earlier 
-detection and treatment of diseases.
-Our Future Health is inviting millions of people, from all backgrounds and 
-from right across the UK to take part. Volunteers are providing information 
-about their health and lifestyles to create an incredibly detailed picture 
-that represents the whole of the UK.
-
-% this section is probably better placed in acknowledgements?
-This study makes use of de-identified data held by Our Future Health. 
-We would like to acknowledge all the research participants who have donated 
-their data to the Our Future Health research programme 
-\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
-
-Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
-was made available in December 2024. The genetic component includes genotype data 
-from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
-chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
-three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
-10,221 variants.
-
 \subsection{Case study: 1,063 spruce whole-genome samples}
@@
 
-%% FIXME: reference regarding VCF POS?
-The spruce data set poses an interesting challenge to the VCF storage
-format due to the large chromosome sizes (>1Gbp). The VCF POS field is
-encoded in 32-bit format (corresponding to a maximum of approximately
-537Mbp), which mandates splitting up larger chromosomes into chunks,
-adding additional complexity to downstream analyses. Zarr does not
-impose this limitation, and we therefore demonstrate the versatility
-of Zarr by updating contig information and the coordinate system for
-the spruce dataset.
+Large genomes from non-model organisms pose particular problems for
+analysis workflows that often are tailored for human conditions. For
+instance, the binary Sequence Alignment/Map index format BAI has a 512
+Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
+The spruce BAM files used for variant calling were BAI indexed, which
+motivated the partitioning into multiple VCF files in the first place,
+adding complexity to downstream processing.
+
+A similar limitation is imposed by the VCF POS field which is encoded
+in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
+spruce does not have chromosomes that exceed the limit, other species
+do, such as the recently sequence mistletoe, with chromosomes larger
+than 8 Gbp~\cite{dtol_mistletoe}. Zarr does not impose this
+limitation, and the coordinate system could easily be converted to
+true genome coordinates for storage in Zarr. This updating procedure
+could be applied to any species of choice.
 

### 2025-01-17 paper.tex

@@
 
+\subsection{Case study: Our Future Health genotype release}
+Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
+It is designed to help people live healthier lives for longer through the 
+discovery and testing of more effective approaches to prevention, earlier 
+detection and treatment of diseases.
+Our Future Health is inviting millions of people, from all backgrounds and 
+from right across the UK to take part. Volunteers are providing information 
+about their health and lifestyles to create an incredibly detailed picture 
+that represents the whole of the UK.
+
+% this section is probably better placed in acknowledgements?
+This study makes use of de-identified data held by Our Future Health. 
+We would like to acknowledge all the research participants who have donated 
+their data to the Our Future Health research programme 
+\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
+
+Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
+was made available in December 2024. The genetic component includes genotype data 
+from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
+chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
+three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
+10,221 variants.
+
 \subsection{Case study: All of Us exome-like data} 
@@
 \end{table}
+>>>>>>> main
 
@@
 
-Large genomes from non-model organisms pose particular problems for
-analysis workflows that often are tailored for human conditions. For
-instance, the binary Sequence Alignment/Map index format BAI has a 512
-Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
-The spruce BAM files used for variant calling were BAI indexed, which
-motivated the partitioning into multiple VCF files in the first place,
-adding complexity to downstream processing.
-
-A similar limitation is imposed by the VCF POS field which is encoded
-in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
-spruce does not have chromosomes that exceed the limit, other species
-do, such as the recently sequence mistletoe, with chromosomes larger
-than 8 Gbp~\cite{dtol_mistletoe}. Zarr does not impose this
-limitation, and the coordinate system could easily be converted to
-true genome coordinates for storage in Zarr. This updating procedure
-could be applied to any species of choice.
+%% FIXME: reference regarding VCF POS?
+The spruce data set poses an interesting challenge to the VCF storage
+format due to the large chromosome sizes (>1Gbp). The VCF POS field is
+encoded in 32-bit format (corresponding to a maximum of approximately
+537Mbp), which mandates splitting up larger chromosomes into chunks,
+adding additional complexity to downstream analyses. Zarr does not
+impose this limitation, and we therefore demonstrate the versatility
+of Zarr by updating contig information and the coordinate system for
+the spruce dataset.
 

### 2025-01-19 paper.tex
Redo chr20 AoU conversion and start adding results in text

--- 2025-01-17 paper.tex
@@
 
-\subsection{Case study: Our Future Health genotype release}
-Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
-It is designed to help people live healthier lives for longer through the 
-discovery and testing of more effective approaches to prevention, earlier 
-detection and treatment of diseases.
-Our Future Health is inviting millions of people, from all backgrounds and 
-from right across the UK to take part. Volunteers are providing information 
-about their health and lifestyles to create an incredibly detailed picture 
-that represents the whole of the UK.
-
-% this section is probably better placed in acknowledgements?
-This study makes use of de-identified data held by Our Future Health. 
-We would like to acknowledge all the research participants who have donated 
-their data to the Our Future Health research programme 
-\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
-
-Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
-was made available in December 2024. The genetic component includes genotype data 
-from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
-chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
-three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
-10,221 variants.
-
 \subsection{Case study: All of Us exome-like data} 
@@
 the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
-transcripts. This dataset, stored as a compressed VCF file of 7.44 GiB, 
-includes 715,256 variants, of which 746 are classified as pathogenic or likely pathogenic.
+transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB, 
+includes 715,256 variants, of which more than 700 are classified as pathogenic or likely pathogenic.
+
+The compression was done on a single compute node with 64 CPUs and 416 GiB of RAM.
+We first converted the VCF to and ICF, which required 6h17min effective runtime and
+343 hours CPU time (parallel efficiency of 86\%), for a cost of \$23.88. The ICF 
+representation used a total of 7.5 GiB over 91,232 data storage files. We then 
+converted the ICF to Zarr (effective runtime: 50min32s, CPU time: 49h49min, 
+parallel efficiency: 92\%, cost: \$3.19), consuming a total of 13 GiB storage 
+over 36,9656 chunk files.
 
@@
 \end{table}
->>>>>>> main
 

### 2025-01-19 paper.tex
Rewrite AoU results section

@@
 
-The compression was done on a single compute node with 64 CPUs and 416 GiB of RAM.
-We first converted the VCF to and ICF, which required 6h17min effective runtime and
-343 hours CPU time (parallel efficiency of 86\%), for a cost of \$23.88. The ICF 
-representation used a total of 7.5 GiB over 91,232 data storage files. We then 
-converted the ICF to Zarr (effective runtime: 50min32s, CPU time: 49h49min, 
-parallel efficiency: 92\%, cost: \$3.19), consuming a total of 13 GiB storage 
-over 36,9656 chunk files.
+The compression was performed on a single compute node with 64 CPUs and 416 GiB of RAM.  
+We first converted the VCF to an ICF, which took 6 hours and 17 minutes of  
+effective runtime and utilized 343 hours of CPU time, achieving a parallel efficiency of 86\%.  
+This step incurred a cost of \$23.88. The resulting ICF representation required 7.5 GiB of storage  
+distributed across 91,232 files. Next, the ICF was converted to Zarr, which took 50 minutes and 32 
+seconds of effective runtime and 49 hours and 49 minutes of CPU time, with a parallel efficiency of 92\%, 
+at a cost of \$3.19. The final Zarr representation occupied 13 GiB of storage across 369,656 chunk files.
 

### 2025-01-19 paper.tex

--- 2025-01-17 paper.tex
@@
 the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
-transcripts. This dataset, stored as a compressed VCF file of 7.44 GiB, 
-includes 715,256 variants, of which 746 are classified as pathogenic or likely pathogenic.
+transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB, 
+includes 715,256 variants, of which more than 700 are classified as pathogenic or likely pathogenic.
+
+The compression was performed on a single compute node with 64 CPUs and 416 GiB of RAM.  
+We first converted the VCF to an ICF, which took 6 hours and 17 minutes of  
+effective runtime and utilized 343 hours of CPU time, achieving a parallel efficiency of 86\%.  
+This step incurred a cost of \$23.88. The resulting ICF representation required 7.5 GiB of storage  
+distributed across 91,232 files. Next, the ICF was converted to Zarr, which took 50 minutes and 32 
+seconds of effective runtime and 49 hours and 49 minutes of CPU time, with a parallel efficiency of 92\%, 
+at a cost of \$3.19. The final Zarr representation occupied 13 GiB of storage across 369,656 chunk files.
 

### 2025-01-19 src/collect_data.py
Update data scaling benchmark data

@@ -598,6 +598,9 @@ def file_size(src, output, debug):
             "bcf": bcf_path,
             "zarr": zarr_path,
             "zarr.zip": ts_path.with_suffix(".zarr.zip"),
+            "10kx1k.zarr": ts_path.with_suffix(".10kx1k.zarr"),
+            "noshuffle.zarr": ts_path.with_suffix(".noshuffle.zarr"),
+            "lz4.noshuffle.zarr": ts_path.with_suffix(".lz4.noshuffle.zarr"),
             "sav": sav_path,
             "genozip": genozip_path,
         }

### 2025-01-20 paper.tex
Update text for 1kx10k Zarr versions

@@
 by subsets of a large simulation of French-Canadians.
-Sizes for $10^6$ samples are shown on the right. Sizes 
-for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
+Sizes for $10^6$ samples are shown on the right.
 Also shown for reference is the size of genotype matrix 
 when encoded as two bits per diploid genotype (2bit), as used 
-in the PLINK binary format.
+in the PLINK binary format. Default compression settings were 
+used in all cases.
 \label{fig-data-storage}}
@@
 larger sample sizes). Zarr and Savvy have 
-almost identical compression performance in this example.
+very similar compression performance in this example (23.75GiB
+and 21.25GiB for $10^6$ samples, respectively). When using
+a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
+storage is reduced to 22.07GiB (not shown).
 It is remarkable that the simple approach of compressing
@@
 This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
-3.9 GiB/s Zarr (Zstd), 4.1 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
+2.7 GiB/s Zarr (Zstd), 2.9 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
 \label{fig-whole-matrix-decode}}

### 2025-01-20 paper.tex
Narrative updates for additional benchmarks.

@@
 % savvy 6.6 GiB
-% zarr_nshf 3.9 GiB
+% zarr_nshf 2.7 GiB
+% zarr_lz4_nshf 2.9 GiB
 (up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
 Fig~\ref{fig-whole-matrix-decode}).
+% sav / zarr ratio: [0.68777206 0.75870945 0.69865406 0.64644888 0.68131752
+% 0.89460906]
 %     sequence_length  num_samples  num_sites                tool         size
-% 40         48129895      1000000    7254858       zarr  22.071687
-% 35         48129895      1000000    7254858        bcf  51.749294
-% 36         48129895      1000000    7254858    genozip  10.691384
-% 37         48129895      1000000    7254858        sav  21.249436
-% 38         48129895      1000000    7254858        tsk   1.802636
-% 39         48129895      1000000    7254858        vcf  81.375831
-% 41         48129895      1000000    7254858  zarr_nshf  29.897540
+% 58         48129895      1000000    7254858                zarr    23.752762
+% 50         48129895      1000000    7254858         10kx1k.zarr    22.068925
+% 51         48129895      1000000    7254858                 bcf    51.749294
+% 52         48129895      1000000    7254858             genozip    10.691384
+% 53         48129895      1000000    7254858  lz4.noshuffle.zarr   214.868005
+% 54         48129895      1000000    7254858      noshuffle.zarr    31.208149
+% 55         48129895      1000000    7254858                 sav    21.249436
+% 56         48129895      1000000    7254858                 tsk     1.802636
+% 57         48129895      1000000    7254858                 vcf    81.375831
+% 59         48129895      1000000    7254858            zarr.zip    24.240338
+% 5          48129895      1000000    7254858                2bit  1689.153258
+[TODO revise this text!!]
 Turning off the BitShuffle filter for the Zarr dataset,
 however, leads to a substantial increase in decoding speed
-(3.9GiB/s) at the cost of a roughly 25\% increase in storage
+(2.9GiB/s) at the cost of a roughly 25\% increase in storage
 space (29.9GiB up from 22.1GiB for 1 million samples; data not
@@
 functionality to aid with such storage schema tuning.
+
+A substantial advantage of Zarr is the availabilty of several 
+independent implementations across different programming languages
+(see Methods). Figure~\ref{fig-whole-matrix-compute-zarr-versions}
+compares the CPU time required to perform the computations
+of Figure~\ref{fig-whole-matrix-compute} 
+for af-dist implementations using the Zarr-Python~\cite{zarrpython},
+JZarr~\cite{jzarr},
+and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
+The version using the Tensorstore C++ API is fastest, requiring 
+about 20\% less time than Zarr-Python (which is used in other
+benchmarks).
+
+% num_samples  num_sites       tool  user_time  sys_time     wall_time
+% storage  total_time
+% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917
+% hdd    4.652136
+% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810
+% hdd    3.701753
+% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377
+% hdd   12.229383
+% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182
+% hdd    5.186064
+% Cpp / Zarr python [0.79197226]
 
@@
 as well as  standalone libraries for 
-C++~\citep{ruan2024image}, JavaScript~\cite{zarrjs},
+C++~\citep{ruan2024image}, 
+Java~\cite{jzarr},
+JavaScript~\cite{zarrjs},
 Julia~\cite{zarrjl}, Rust~\citep{zarrs}
@@
 \begin{figure}[h]
+\includegraphics{figures/whole-matrix-compute-zarr-versions}
+\caption{Whole-matrix computation performance using different Zarr 
+implementations. 
+Total CPU time required to run the af-dist calculation
+the data in Figure~\ref{fig-data-storage} using different Zarr implementations.
+Elapsed time is also reported (dotted line). 
+These benchmarks were run on [TODO insert hardware details].
+Note the difference between the
+time reported here and in Fig~\ref{fig-whole-matrix-compute} for Zarr-Python
+is due to different hardware platforms.
+\label{fig-whole-matrix-compute-zarr-versions}}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics{figures/subset-matrix-compute-supplemental}

### 2025-01-20 src/plot.py

@@ -90,31 +90,32 @@ def plot_size(ax, df, label_y_offset=None, order=None):
     ax.set_ylabel("Storage size (bytes)")
     plt.tight_layout()
 
+
 def plot_s3_throughput(ax, df):
     ax.loglog(
         df["processes"].values,
         df["throughput_decompress"].values,
         marker=".",
-        label="Decompression only"
+        label="Decompression only",
     )
     ax.annotate(
         f"{df['throughput_decompress'][6]:.0f} MB/s",
         textcoords="offset points",
         xytext=(-8, -10),
-        xy=(df['processes'][6], df['throughput_decompress'][6]),
+        xy=(df["processes"][6], df["throughput_decompress"][6]),
         xycoords="data",
     )
     ax.loglog(
         df["processes"].values,
         df["throughput_afdist"].values,
         marker=".",
-        label="AF Dist"
+        label="AF Dist",
     )
     ax.annotate(
         f"{df['throughput_afdist'][6]:.0f} MB/s",
         textcoords="offset points",
         xytext=(-8, -10),
-        xy=(df['processes'][6], df['throughput_afdist'][6]),
+        xy=(df["processes"][6], df["throughput_afdist"][6]),
         xycoords="data",
     )
 
@@ -133,6 +134,7 @@ def plot_total_cpu(
     extrapolate=None,
     order=None,
     label_y_offset=None,
+    side_digits=None,
 ):
     if colours is None:
         colours = {
@@ -185,8 +187,14 @@ def plot_total_cpu(
 
         if tool not in extrapolate:
             time = total_cpu[-1] / divisors[time_units]
+            if side_digits is None:
+                digits = 0 if time > 1 else 1
+            else:
+                digits = side_digits
+
+            side_label = f"{time:.{digits}f}{time_units}"
             ax.annotate(
-                f"{time:.0f}{time_units}" if time > 1 else f"{time:.1f}{time_units}",
+                side_label,
                 textcoords="offset points",
                 xytext=(15, label_y_offset.get(tool, 0)),
                 xy=(row.num_samples, total_cpu[-1]),
@@ -271,6 +279,7 @@ def data_scaling(size_data, output):
 
     plt.savefig(output)
 
+
 @click.command()
 @click.argument("data", type=click.File("r"))
 @click.argument("output", type=click.Path())
[…]

### 2025-01-20 src/collect_data.py
Add Zarr zip decoding data

@@ -638,6 +638,15 @@ decode_tools = all_tools[:2] + [
         None,
         run_zarr_decode,
         run_zarr_pos_extract,
+    ),
+    Tool(
+        "zarr.zip",
+        ".zarr.zip",
+        run_zarr_afdist,
+        run_zarr_afdist_subset,
+        None,
+        run_zarr_decode,
+        run_zarr_pos_extract,
     )
 ]
 

### 2025-01-21 paper.tex
Update storage size discussion for new benchmarks

@@
 a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
-storage is reduced to 22.07GiB (not shown).
+storage is reduced to 22.07GiB for $10^6$ samples (data not shown;
+see Figs~\ref{fig-compression-chunksize} and \ref{fig-compression-chunksize-finegrained}
+for more analysis on the effect of chunk size on compression).
 It is remarkable that the simple approach of compressing
@@
 % 5          48129895      1000000    7254858                2bit  1689.153258
-[TODO revise this text!!]
 Turning off the BitShuffle filter for the Zarr dataset,
 however, leads to a substantial increase in decoding speed
-(2.9GiB/s) at the cost of a roughly 25\% increase in storage
-space (29.9GiB up from 22.1GiB for 1 million samples; data not
+(2.7GiB/s) at the cost of a roughly 25\% increase in storage
+space (31.2GiB up from 23.75GiB for 1 million samples; data not
 shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
-decoding speed, but at the cost of much larger storage requirements 
-(219GiB; data not shown).
+decoding speed (2.9GiB/s), but at the cost of much larger storage requirements 
+(214GiB; data not shown).
 Given the relatively small contribution of genotypes to the
@@
 
-A substantial advantage of Zarr is the availabilty of several 
+Another advantage of Zarr is the availabilty of several 
 independent implementations across different programming languages

### 2025-01-21 src/plot.py

@@ -271,7 +271,7 @@ def data_scaling(size_data, output):
     df1 = pd.concat([df1, pd.DataFrame(plink_ish)])
 
     fig, ax1 = one_panel_fig()
-    plot_size(ax1, df1, label_y_offset={"vcf": 4, "sav": -5.5, "genozip": -7})
+    plot_size(ax1, df1, label_y_offset={"vcf": 4, "bcf":1, "sav": -5.5, "genozip": -7})
 
     # I tried putting an inset axis showing the ratio, but it was too small.
     # ax_inset = ax1.inset_axes([0.70, 0.1, 0.25, 0.25])

### 2025-01-21 paper.tex
Update fig and text for whole-matrix compute data

@@
 short and simple Python program is substantially faster than the 
-equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
+equivalent compiled C using \texttt{htslib} APIs on BCF (7.3 hours
 vs 20.6 hours for 1 million samples). 
 % num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
-%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
-%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
-%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
+% 21      1000000    7254858     savvy    8451.25     16.59   8471.293500     hdd 2.352178
+% 10      1000000    7254858  bcftools   73939.32     46.09  74023.083678     hdd 20.551503
+% 27      1000000    7254858      zarr   26227.92    187.22  31476.507443     hdd 7.337539
 The fastest method is the 

### 2025-01-21 paper.tex
Update text for subset compute

@@
 the overall dataset size, with the computation requiring around
-1 second for 1 million samples. Because of their 
+2 seconds for 1 million samples. Because of their 
 row-wise encoding, CPU time scales with the number of samples

### 2025-01-21 paper.tex
Update text for column extract figure

@@
 the time required to populate a NumPy array with the data in Zarr,
-which is less than a second. Wall-clock time (dotted line) is dominated 
-in this case by file I/O. Time to output text for Savvy is not significant
+which is 2 seconds. Wall-clock time (dotted line) is dominated 
+in this case by file I/O. Note that our benchmarking methodology of 
+dropping the file system caches before each run (see Methods) is
+likely not reflective of real world usage where a small and 
+frequently used field such as POS would remain in cache.
+Time to output text for Savvy is not significant
 for $> 1000$ samples (not shown).

### 2025-01-21 paper.tex
Tweak

@@
 and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
-The version using the Tensorstore C++ API is fastest, requiring 
-about 20\% less time than Zarr-Python (which is used in other
-benchmarks).
+The program using the Tensorstore C++ API is fastest, requiring 
+about 20\% less CPU time than the version using Zarr-Python and Numba
+used in our other benchmarks.
 

### 2025-01-21 src/plot.py
Add discussion of zip files.
Closes #214

@@ -20,6 +20,7 @@ genozip_colour = "tab:purple"
 zarr_colour = "tab:blue"
 zarr_nshf_colour = "tab:cyan"
 zarr_lz4_nshf_colour = "tab:purple"
+zarr_zip_colour = "tab:green"
 two_bit_colour = "tab:pink"
 zarr_java_colour = "tab:olive"
 ts_py_colour = "tab:brown"
@@ -144,6 +145,7 @@ def plot_total_cpu(
             "zarr": zarr_colour,
             "zarr_nshf": zarr_nshf_colour,
             "zarr_lz4_nshf": zarr_lz4_nshf_colour,
+            "zarr.zip": zarr_zip_colour,
             "savvy": sav_colour,
             "zarr_java": zarr_java_colour,
             "ts_py": ts_py_colour,
@@ -271,7 +273,7 @@ def data_scaling(size_data, output):
     df1 = pd.concat([df1, pd.DataFrame(plink_ish)])
 
     fig, ax1 = one_panel_fig()
-    plot_size(ax1, df1, label_y_offset={"vcf": 4, "bcf":1, "sav": -5.5, "genozip": -7})
+    plot_size(ax1, df1, label_y_offset={"vcf": 4, "bcf": 1, "sav": -5.5, "genozip": -7})
 
     # I tried putting an inset axis showing the ratio, but it was too small.
     # ax_inset = ax1.inset_axes([0.70, 0.1, 0.25, 0.25])
@@ -390,6 +392,13 @@ def whole_matrix_decode(time_data, output):
         print(tool, humanize.naturalsize(max_rate, binary=True))
 
     plt.savefig(output)
+
+    df["cpu_time"] = df["user_time"] + df["sys_time"]
+    df = df[((df.tool == "zarr") | (df.tool == "zarr.zip")) & (df.num_samples == 10**6)]
+    dfs = df[["tool", "user_time", "sys_time", "cpu_time", "wall_time"]].copy()
+    dfs["wall_time"] /= 60
+    dfs["cpu_time"] /= 60
+    print(dfs)
 
 
 @click.command()

### 2025-01-21 paper.tex

@@
 
+The discrepancy between elapsed time and total CPU time for Zarr 
+benchmarks in
+Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
+and~\ref{fig-whole-matrix-decode} are mostly due to the overhead
+of accessing a large number of small chunk files and the reduced
+effectiveness of file-system read-ahead caching.
+While having many independent chunks maps 
+naturally to high-performance parallel access patterns
+in the cloud store setting~\citep{durner2023exploiting} 
+(see also the section on accessing Genomics England data stored on S3), 
+such file access patterns can be problematic in the 
+classical file-system setting.
+The Zarr Zip store (which is simply a Zip archive of the standard directory 
+hierarchy) provides a useful alternative here and can 
+significantly reduce overall processing time. To illustrate this
+we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
+which we measure the time required to decode the entire genotype
+matrix (thus serving as a lower-bound for any subsequent computations).
+In the case of $10^6$ samples, the Zip file requires slightly more 
+% Full data table above ^
+% 58         48129895      1000000    7254858                zarr    23.752762
+% 59         48129895      1000000    7254858            zarr.zip    24.240338
+storage space than the directory hierarchy (23.75GiB and 24.24G
+for file system and Zip file, respectively)
+and while total 
+CPU time is similar (193m vs 187m), the elapsed wall clock
+time for the Zip file version is significantly less 
+(258m vs 188m).
+        % tool  user_time  sys_time    cpu_time   wall_time
+% 11      zarr   11383.82    191.01  192.913833  258.598778
+% 29  zarr.zip   11198.37     43.13  187.358333  188.659252
+
 Another advantage of Zarr is the availabilty of several 
 independent implementations across different programming languages
-(see Methods). Figure~\ref{fig-whole-matrix-compute-zarr-versions}
+(see Methods). Fig~\ref{fig-whole-matrix-compute-zarr-versions}
 compares the CPU time required to perform the computations
-of Figure~\ref{fig-whole-matrix-compute} 
+of Fig~\ref{fig-whole-matrix-compute} 
 for af-dist implementations using the Zarr-Python~\cite{zarrpython},
@@
 % storage  total_time
-% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917
-% hdd    4.652136
-% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810
-% hdd    3.701753
-% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377
-% hdd   12.229383
-% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182
-% hdd    5.186064
+% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917 hdd    4.652136
+% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810 hdd    3.701753
+% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377 hdd   12.229383
+% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182 hdd    5.186064
 % Cpp / Zarr python [0.79197226]

### 2025-01-21 paper.tex
Update S3 benchmark text

@@
 arrays and chunks are stored hierarchically in directories
-and files (storage as a single Zip archive is also supported).
+and files (storage as a single Zip archive is also supported,
+and has certain advantages as discussed later).
 To enable comparison with the existing file-based ecosystem
@@
 Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
-and~\ref{fig-whole-matrix-decode} are mostly due to the overhead
-of accessing a large number of small chunk files and the reduced
-effectiveness of file-system read-ahead caching.
+and~\ref{fig-whole-matrix-decode} is mostly due to the overhead
+of accessing a large number of small chunk files and 
+consequent reduced effectiveness of file-system read-ahead caching.
 While having many independent chunks maps 
@@
 total of 2.54 TiB of storage over 6,312,488
-chunk files. This is a roughly 5X reduction in total storage 
+chunk files (using a chunk size of $10^4$ variants $\times 10^3$ 
+samples). This is a roughly 5X reduction in total storage 
 space over the original VCF. 
@@
 Finally, to explore the use of cloud computing services with VCF Zarr,
-the zarr files for chromosome 2 were uploaded to an AWS S3 standard bucket.
-Using a single-threaded process on an EC2 instance the benchmarks above were
-repeated. Extracting the position array required 15s CPU, 23.8s elapsed.
-The af-dist calculation required 2.70min CPU, 5.23min elapsed. Creating the filter mask
-required 3.05min CPU, 11.15min elapsed. These times are comparable to the on-premises
-HPC system above.
-
-This single-threaded approach is not optimal for cloud computing, as to optimise
-both runtime and costs parallel processing is required. Using 64 parallel processes
-on a single c5.24xlarge EC2 instance with 96 vCPUs (48 physical) the zarr chunks of
-genotype array could be retrived from S3 and decompressed at a rate of 50.8 GB/s.
-The network data rate was 127 MB/s, far below the capbility of the instance,
-indicating that this process was compute-constrained. When calculating af-dist after
+we uploaded the Zarr directory hierarchy for chromosome 2 to an AWS S3 
+standard bucket. We repeated the benchmarks above the same subset
+562,640 variants, and the same 
+single-threaded sequential Zarr code.
+Extracting the position array required 15s CPU, 23.8s elapsed.
+The af-dist calculation required 2.70min CPU, 5.23min elapsed. 
+Creating the filter mask
+required 3.05min CPU, 11.15min elapsed.
+
+While these benchmarks are comparable with results obtained from the 
+shared file-system in the Genomics England research environment,
+the sequential single-threaded access patterns used do not map
+well to modern cloud computing environments.
+As each request to a cloud object store has a high latency, it 
+is important to have many requests running concurrently to achieve 
+good performance~\citep{durner2023exploiting}.
+To explore the limits of performance we developed a simple prototype
+Zarr implementation, designed to maximise concurrency using asynchrous
+IO and multiple processes (see Methods for details).
+Fig~\ref{fig-s3-throughput} shows the rate at which genotypes 
+can be decoded to RAM as well as rate at which we can compute
+af-dist (using Numba to classify genotypes, as before)
+as we increase the number of parallel processes on
+an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
+The maximum decompression rate of 47.4GiB/s 
+occurs at 64 processes, with a corresponding 
+network data rate of 
+% FIXME MiB/s
+127 MB/s, far below the capbility of the instance,
+indicating that this process was compute-constrained.
+
+When calculating af-dist after
 decompression the throughput was 24.6 GB/s. With this parallel approach the 
@@
 
-The compression was performed on a single compute node with 64 CPUs and 416 GiB of RAM.  
-We first converted the VCF to an ICF, which took 6 hours and 17 minutes of  
-effective runtime and utilized 343 hours of CPU time, achieving a parallel efficiency of 86\%.  
-This step incurred a cost of \$23.88. The resulting ICF representation required 7.5 GiB of storage  
-distributed across 91,232 files. Next, the ICF was converted to Zarr, which took 50 minutes and 32 
-seconds of effective runtime and 49 hours and 49 minutes of CPU time, with a parallel efficiency of 92\%, 
-at a cost of \$3.19. The final Zarr representation occupied 13 GiB of storage across 369,656 chunk files.
+The compression was performed on the All of Us Researcher Workbench platform 
+(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
+of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
+took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
+time, achieving a parallel efficiency of 86\%. This step incurred a cost of 
+\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
+across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with 
+a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a 
+cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across 
+232,809 chunk files.
 
@@
 \caption{Summary for a selection of the largest VCF Zarr columns produced 
-the from the 245,388 exome-like samples from chromosome 20
-using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
-10,000 samples.
-For column details see the caption of Table~\ref{tab-genomics-england-data}.
-\label{tab-all-of-us-data}}
-\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
+    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
+    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
+    \label{tab-top-fields-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
 \toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
+{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
 \midrule
-/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
-/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
-/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
-/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
-/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
-/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
-/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
-% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
-% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
-% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
-% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
-% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
-% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
-% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
-% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
-% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
-% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
-% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
-% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
+/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
+/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
+/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
+/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
+/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
+/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
+/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
 \bottomrule
@@
 \end{figure}
+
 \begin{figure}[h]
-
+\includegraphics{figures/s3-throughput}
+\caption{Scalability of genotype data processing on AWS.
+\label{fig-s3-throughput}}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics{figures/compression-compressor}

### 2025-01-21 src/plot.py

@@ -93,36 +93,61 @@ def plot_size(ax, df, label_y_offset=None, order=None):
 
 
 def plot_s3_throughput(ax, df):
+
+    GB = 2**30
+
+    argmax = 6
     ax.loglog(
         df["processes"].values,
         df["throughput_decompress"].values,
         marker=".",
         label="Decompression only",
+        base=2,
     )
+    ax.plot(
+        df["processes"][argmax],
+        df["throughput_decompress"][argmax],
+        marker="+",
+        color="black",
+    )
+    maxval = df["throughput_decompress"][argmax]
     ax.annotate(
-        f"{df['throughput_decompress'][6]:.0f} MB/s",
+        f"{maxval / GB:.1f} GiB/s",
         textcoords="offset points",
         xytext=(-8, -10),
-        xy=(df["processes"][6], df["throughput_decompress"][6]),
+        xy=(df["processes"][argmax], df["throughput_decompress"][argmax]),
         xycoords="data",
     )
     ax.loglog(
         df["processes"].values,
         df["throughput_afdist"].values,
         marker=".",
-        label="AF Dist",
+        label="Compute af-dist",
+        base=2,
     )
+    argmax = 7
+    ax.plot(
+        df["processes"][argmax],
+        df["throughput_afdist"][argmax],
+        marker="+",
+        color="black",
+    )
+    maxval = df["throughput_afdist"][argmax]
     ax.annotate(
-        f"{df['throughput_afdist'][6]:.0f} MB/s",
+        f"{maxval / GB:.1f} GiB/s",
         textcoords="offset points",
         xytext=(-8, -10),
-        xy=(df["processes"][6], df["throughput_afdist"][6]),
+        xy=(df["processes"][argmax], df["throughput_afdist"][argmax]),
         xycoords="data",
     )
 
+    ax.set_xticks(df["processes"].values)
+    ax.get_xaxis().set_major_formatter(matplotlib.ticker.ScalarFormatter())
+    # ax.set_xticklabels(labels)
+
     ax.legend()
     ax.set_xlabel("Number of processes")
-    ax.set_ylabel("Genotype data throughput (MB/s)")
+    ax.set_ylabel("Genotype data throughput (bytes/s)")
     plt.tight_layout()
 
 
@@ -290,13 +315,14 @@ def s3_throughput(data, output):
     Plot the figure showing compute performance on whole-matrix afdist.
     """
[…]

### 2025-01-22 paper.tex

--- 2025-01-21 paper.tex
@@
 
-The compression was performed on a single compute node with 64 CPUs and 416 GiB of RAM.  
-We first converted the VCF to an ICF, which took 6 hours and 17 minutes of  
-effective runtime and utilized 343 hours of CPU time, achieving a parallel efficiency of 86\%.  
-This step incurred a cost of \$23.88. The resulting ICF representation required 7.5 GiB of storage  
-distributed across 91,232 files. Next, the ICF was converted to Zarr, which took 50 minutes and 32 
-seconds of effective runtime and 49 hours and 49 minutes of CPU time, with a parallel efficiency of 92\%, 
-at a cost of \$3.19. The final Zarr representation occupied 13 GiB of storage across 369,656 chunk files.
+The compression was performed on the All of Us Researcher Workbench platform 
+(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
+of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
+took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
+time, achieving a parallel efficiency of 86\%. This step incurred a cost of 
+\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
+across 91,232 files. Next, the ICF was converted to Zarr, 
+after removing the field \texttt{call\_FT}, which is 
+suspected to be problematic. The conversion took 53 minutes and 40 seconds, with 
+a CPU time of 52 hours and 34 minutes and a parallel efficiency of 92\%, at a 
+cost of \$3.40. This second Zarr representation occupied 8 GiB of storage across 
+214,190 chunk files.
 
@@
 \caption{Summary for a selection of the largest VCF Zarr columns produced 
-the from the 245,388 exome-like samples from chromosome 20
-using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
-10,000 samples.
-For column details see the caption of Table~\ref{tab-genomics-england-data}.
-\label{tab-all-of-us-data}}
-\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
+    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
+    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
+    \label{tab-top-fields-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
 \toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
+{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
 \midrule
-/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
-/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
-/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
-/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
-/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
-/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
-/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
-% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
-% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
-% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
-% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
-% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
-% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
-% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
-% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
-% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
-% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
-% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
-% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
+/call_LAD & int16 & 2.26 GiB & 290.0 & 29.63% \\
+/call_GQ & int8 & 1.89 GiB & 87.0 & 24.78% \\
+/call_LA & int8 & 1.24 GiB & 260.0 & 16.26% \\
+/call_genotype & int8 & 914.22 MiB & 370.0 & 11.70% \\
+/call_RGQ & int16 & 729.57 MiB & 460.0 & 9.34% \\
+/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.76% \\
 \bottomrule

### 2025-01-22 paper.tex

@@
 \$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
-across 91,232 files. Next, the ICF was converted to Zarr, 
-after removing the field \texttt{call\_FT}, which is 
-suspected to be problematic. The conversion took 53 minutes and 40 seconds, with 
-a CPU time of 52 hours and 34 minutes and a parallel efficiency of 92\%, at a 
-cost of \$3.40. This second Zarr representation occupied 8 GiB of storage across 
-214,190 chunk files.
+across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with 
+a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a 
+cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across 
+232,809 chunk files.
 
@@
 \midrule
-/call_LAD & int16 & 2.26 GiB & 290.0 & 29.63% \\
-/call_GQ & int8 & 1.89 GiB & 87.0 & 24.78% \\
-/call_LA & int8 & 1.24 GiB & 260.0 & 16.26% \\
-/call_genotype & int8 & 914.22 MiB & 370.0 & 11.70% \\
-/call_RGQ & int16 & 729.57 MiB & 460.0 & 9.34% \\
-/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.76% \\
+/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
+/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
+/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
+/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
+/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
+/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
+/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
 \bottomrule

### 2025-01-22 paper.tex
Add figure for network throughput

--- 2025-01-21 paper.tex
@@
 we uploaded the Zarr directory hierarchy for chromosome 2 to an AWS S3 
-standard bucket. We repeated the benchmarks above the same subset
-562,640 variants, and the same 
+standard bucket. We repeated the benchmarks above on the same subset
+of 562,640 variants, using the same 
 single-threaded sequential Zarr code.
@@
 
-While these benchmarks are comparable with results obtained from the 
+While these benchmarks are comparable with results obtained on the 
 shared file-system in the Genomics England research environment,
@@
 occurs at 64 processes, with a corresponding 
-network data rate of 
-% FIXME MiB/s
-127 MB/s, far below the capbility of the instance,
-indicating that this process was compute-constrained.
+network data rate of 127 MB/s [FIXME confirm and convert to MiB/s].
+Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
+network access by measuring the rate at which we can download the chunks
+into RAM (without decompressing) using the same parallel and asynchronous
+architecture. Here, the maximum throughput of
+
+To assess the limitations imposed by network access, we repeated this
+experiment but measured only the rate at which chunks can be downloaded
 
@@
 \begin{figure}[h]
+\includegraphics{figures/s3-network-throughput}
+\caption{Chunk data download rate on AWS.
+\label{fig-s3-network-throughput}}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics{figures/compression-compressor}

### 2025-01-22 src/plot.py

@@ -92,6 +92,39 @@ def plot_size(ax, df, label_y_offset=None, order=None):
     plt.tight_layout()
 
 
+def plot_s3_network_throughput(ax, df):
+
+    GB = 2**30
+
+    x = df["processes"].values
+    for col in ["c5.9xlarge", "c5n.9xlarge"]:
+        y = df[col].values
+        ax.loglog(
+            x ,
+            y, 
+            marker=".",
+            label=col,
+            base=2,
+        )
+        argmax = np.argmax(y)
+        ax.plot(x[argmax], y[argmax], marker="+", color="black")
+        ax.annotate(
+            f"{y[argmax] / GB:.1f} GiB/s",
+            textcoords="offset points",
+            xytext=(-8, -10),
+            xy=(x[argmax], y[argmax]),
+            xycoords="data",
+        )
+
+    ax.set_xticks(x)
+    ax.get_xaxis().set_major_formatter(matplotlib.ticker.ScalarFormatter())
+
+    ax.legend()
+    ax.set_xlabel("Number of processes")
+    ax.set_ylabel("Data download rate (bytes/s)")
+    plt.tight_layout()
+
+
 def plot_s3_throughput(ax, df):
 
     GB = 2**30
@@ -323,6 +356,50 @@ def s3_throughput(data, output):
 
     fig, ax1 = one_panel_fig()
     plot_s3_throughput(ax1, df)
+    plt.savefig(output)
+
+
+@click.command()
+@click.argument("data", type=click.File("r"))
+@click.argument("output", type=click.Path())
+def s3_network_throughput(data, output):
+    """
+    Plot the figure showing compute performance on whole-matrix afdist.
+    """
+
+    # Data derived from S3-1-DownloadChunks-MP.ipynb notebook. Numbers there]
+    # are in power-of-10 MB
+
+    # x = [1, 2, 4, 8, 16, 32, 64]
+    # data1 = np.array([
+    #     159.0341786762204,
+    #     349.88301727773677,
+    #     660.1621287063576,
+    #     1009.4825168651081,
+    #     1253.8112536680967,
+    #     1280.6974940263988,
+    #     1274.494689412922,
+    # ])
+    # data2 = np.array([
+    #     162.80624306237462,
+    #     345.7732721919105,
+    #     658.9358771498762,
+    #     1062.419542658184,
+    #     1972.057722720383,
[…]

### 2025-01-22 paper.tex
Pass through the S3 section

@@
 into RAM (without decompressing) using the same parallel and asynchronous
-architecture. Here, the maximum throughput of
-
-To assess the limitations imposed by network access, we repeated this
-experiment but measured only the rate at which chunks can be downloaded
-
-When calculating af-dist after
-decompression the throughput was 24.6 GB/s. With this parallel approach the 
-af-dist calculation above was then completed in 4.12s elapsed time.
+architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
+and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
+was achieved with 32 processes. Genotype data can therefore be fetched 
+from S3 around an order of magnitude more quickly than it can be decompressed 
+into RAM. This has some interesting consequences
+when we consider the goal of seeking the highest possible compression levels 
+for genotype data and the usual cloud storage cost structuring models (see
+Discussion).
+
+Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
+calculation is performed at a maximum rate of 25.5GiB/s using 128 
+processes. At this rate, the af-dist calculation for the target region 
+was completed in 4.12s [CHECKME] (compare to 5.23min for the naive single-threaded
+method above).
 This level of throughput would complete af-dist for all chromosomes for all samples
 % compute: $4.08 * 1.27
+% What are the numbers here??
 % io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
-in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
+in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
 (See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr

### 2025-01-22 src/plot.py

@@ -169,14 +169,13 @@ def plot_s3_throughput(ax, df):
     ax.annotate(
         f"{maxval / GB:.1f} GiB/s",
         textcoords="offset points",
-        xytext=(-8, -10),
+        xytext=(-30, -14),
         xy=(df["processes"][argmax], df["throughput_afdist"][argmax]),
         xycoords="data",
     )
 
     ax.set_xticks(df["processes"].values)
     ax.get_xaxis().set_major_formatter(matplotlib.ticker.ScalarFormatter())
-    # ax.set_xticklabels(labels)
 
     ax.legend()
     ax.set_xlabel("Number of processes")

### 2025-01-22 paper.tex
Add fig S2 hardware details

--- 2025-01-22 paper.tex
@@
 unless otherwise stated.
-All simulation-based benchmarks were performed on a 
+All simulation-based benchmarks (excluding Fig~\ref{fig-whole-matrix-compute-zarr-versions}) were performed on a 
 dual CPU (Intel Xeon E5-2680 v2)
@@
 Elapsed time is also reported (dotted line). 
-These benchmarks were run on [TODO insert hardware details].
+These benchmarks were run on an 8-core CPU (Intel i7-9700) with 32GiB RAM running
+Linux Mint 21.3 with data on an NVMe SSD.
 Note the difference between the

### 2025-01-22 paper.tex
Tidy up OFH section

@@
 
-\subsection{Case study: Genomics England 100,000 genomes}
+\subsection{Case study: Genomics England WGS data}
 In this section we demonstrate the utility of VCF Zarr on a large human dataset
@@
 
-\subsection{Case study: Our Future Health genotype release}
+\subsection{Case study: Our Future Health genotype data}
 Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
@@
 
-% this section is probably better placed in acknowledgements?
-This study makes use of de-identified data held by Our Future Health. 
-We would like to acknowledge all the research participants who have donated 
-their data to the Our Future Health research programme 
-\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
-
-Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
+Release 9~\citep{ofhdatarelease9}
 was made available in December 2024. The genetic component includes genotype data 
@@
 \midrule
-/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
-/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
-/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
-/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
-/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
-/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
-/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
+% /call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
+% /call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
+% /call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
+% /call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
+% /call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
+% /call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
+% /call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
 \bottomrule
@@
 conversion to Zarr. The tasks were run as single jobs on a compute
-node with 2,048 GB RAM and 128 cores. Conversion to ICF required
-13h45min runtime. The ICF representation used a total of 6.77 TiB over
-1,342,569 data storage files. Conversion to Zarr required 18h51min
-runtime, generating a dataset with 32 arrays, consuming a total of 6.6
-TiB storage over 11,984,581 chunk files. The reduction in total
-storage (1.1X) is negligible, mainly due to the inherent difficulty in
-compressing the PL field. The top fields in terms of storage are
-detailed in Table~\ref{tab-spruce-data}.
+node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
+required 13h45min runtime. The ICF representation used a total of 6.77
+TiB over 1,340,804 data storage files. Conversion to Zarr required
+18h36min runtime, generating a dataset with 32 arrays, consuming a
+total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
+total storage (1.1X) is negligible, mainly due to the inherent
+difficulty in compressing the PL field. The output dataset storage
+size is dominated by a major component, call\_PL, which accounts for
+nearly 93\% of the total. call\_PL is yet another field with low
+compression potential due to inherent noisiness.
+
+Therefore, to reduce storage requirements we generated a Zarr schema
+for encoding that uses local allele fields with \texttt{vcf2zarr
+mkschema --local-alleles}, and used the generated schema as input to
+\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
+generating a dataset with 35 arrays, consuming a total of 2.3TiB
+storage over 11,235,556 chunk files, with a reduction in storage of
+3.2X. Therefore, by using local allele fields we achieve approximately
+3X improvement of compression. The top fields in terms of storage are
+detailed inTable~\ref{tab-spruce-data}.
 
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns produced for 
-1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
-default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
+\caption{Summary for a selection of the largest VCF Zarr columns
+produced for 1,063 VCFs on 12 spruce chromosomes using
+\texttt{vcf2zarr} with samples chunk size set to 1,063, variants
+chunk size to 10,000, and using local allele fields to reduce output
+size. For column details see the caption of
+Table~\ref{tab-genomics-england-data}.
 \label{tab-spruce-data}}
-\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
+
+\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
 \toprule
@@
 \midrule
-  /call\_PL & int16 & 6.1 TiB & 12.0 & 92.75\% \\
-  /call\_genotype & int8 & 282.5 GiB & 26.0 & 4.18\% \\
-  /call\_genotype\_mask & bool & 42.2 GiB & 180.0 & 0.62\% \\
-  /variant\_DP4 & int32 & 26.2 GiB & 2.1 & 0.39\% \\
-  /variant\_MQSBZ & float32 & 13.1 GiB & 1.1 & 0.19\% \\
-  /variant\_RPBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
-  /variant\_MQBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
-  /variant\_SCBZ & float32 & 12.9 GiB & 1.1 & 0.19\% \\
-  /variant\_VDB & float32 & 12.8 GiB & 1.1 & 0.19\% \\
-  /variant\_quality & float32 & 12.8 GiB & 1.1 & 0.19\% \\
-  /variant\_BQBZ & float32 & 12.6 GiB & 1.1 & 0.19\% \\
-  /variant\_SGB & float32 & 12.1 GiB & 1.2 & 0.18\% \\
-  /variant\_position & int32 & 9.7 GiB & 1.4 & 0.14\% \\
-  /variant\_AC & int16 & 7.4 GiB & 2.8 & 0.11\% \\
-  /variant\_DP & int32 & 6.7 GiB & 2.1 & 0.10\% \\
-  /variant\_allele & object & 5.4 GiB & 21.0 & 0.08\% \\
-  /variant\_AN & int16 & 3.1 GiB & 2.3 & 0.05\% \\
-  /call\_genotype\_phased & bool & 2.0 GiB & 1800.0 & 0.03\% \\
-  /variant\_filter & bool & 1.5 GiB & 2.4 & 0.02\% \\
-  /variant\_MQ & int8 & 489.8 MiB & 7.3 & 0.01\% \\
+/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
+/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
+/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
+/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
+/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
+/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
+/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
+/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
+/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
+/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
+/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
+/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
+/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
+/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
+/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
+/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
+/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
+/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
+/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
+/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
+/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
+/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
+/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
+% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
+% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
+% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
+% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
+% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
+% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
+% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
+% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
+% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
+% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
+% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
+% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
 \bottomrule
@@
 \end{table}
-
-Table~\ref{tab-spruce-data} shows that the dataset storage size is
-dominated by a major component, call\_PL, which accounts for nearly
-93\% of the total. call\_PL is yet another field with low compression
-potential due to inherent noisiness.
 
@@
 
+This study makes use of de-identified data held by Our Future Health. 
+We would like to acknowledge all the research participants who have donated 
+their data to the Our Future Health research programme~\cite{ofhpubpolicy}.
+
 Computation used the Oxford Biomedical Research Computing (BMRC) facility, a

### 2025-01-22 paper.tex
Add OFH to intro
Closes #153

@@
 UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
+Our Future Health~\citep{ofhmain_2025},
 and the All of Us research program~\citep{all2024genomic}

### 2025-01-22 paper.tex
Update spruce example to use local alleles

--- 2025-01-22 paper.tex
@@
 by subsets of a large simulation of French-Canadians.
-Sizes for $10^6$ samples are shown on the right.
+Sizes for $10^6$ samples are shown on the right. Sizes 
+for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
 Also shown for reference is the size of genotype matrix 
 when encoded as two bits per diploid genotype (2bit), as used 
-in the PLINK binary format. Default compression settings were 
-used in all cases.
+in the PLINK binary format.
 \label{fig-data-storage}}
@@
 arrays and chunks are stored hierarchically in directories
-and files (storage as a single Zip archive is also supported,
-and has certain advantages as discussed later).
+and files (storage as a single Zip archive is also supported).
 To enable comparison with the existing file-based ecosystem
@@
 larger sample sizes). Zarr and Savvy have 
-very similar compression performance in this example (23.75GiB
-and 21.25GiB for $10^6$ samples, respectively). When using
-a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
-storage is reduced to 22.07GiB for $10^6$ samples (data not shown;
-see Figs~\ref{fig-compression-chunksize} and \ref{fig-compression-chunksize-finegrained}
-for more analysis on the effect of chunk size on compression).
+almost identical compression performance in this example.
 It is remarkable that the simple approach of compressing
@@
 short and simple Python program is substantially faster than the 
-equivalent compiled C using \texttt{htslib} APIs on BCF (7.3 hours
+equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
 vs 20.6 hours for 1 million samples). 
 % num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
-% 21      1000000    7254858     savvy    8451.25     16.59   8471.293500     hdd 2.352178
-% 10      1000000    7254858  bcftools   73939.32     46.09  74023.083678     hdd 20.551503
-% 27      1000000    7254858      zarr   26227.92    187.22  31476.507443     hdd 7.337539
+%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
+%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
+%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
 The fastest method is the 
@@
 % savvy 6.6 GiB
-% zarr_nshf 2.7 GiB
-% zarr_lz4_nshf 2.9 GiB
+% zarr_nshf 3.9 GiB
 (up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
 Fig~\ref{fig-whole-matrix-decode}).
-% sav / zarr ratio: [0.68777206 0.75870945 0.69865406 0.64644888 0.68131752
-% 0.89460906]
 % sequence_length  num_samples  num_sites       tool       size
-% 58         48129895      1000000    7254858                zarr    23.752762
-% 50         48129895      1000000    7254858         10kx1k.zarr    22.068925
-% 51         48129895      1000000    7254858                 bcf    51.749294
-% 52         48129895      1000000    7254858             genozip    10.691384
-% 53         48129895      1000000    7254858  lz4.noshuffle.zarr   214.868005
-% 54         48129895      1000000    7254858      noshuffle.zarr    31.208149
-% 55         48129895      1000000    7254858                 sav    21.249436
-% 56         48129895      1000000    7254858                 tsk     1.802636
-% 57         48129895      1000000    7254858                 vcf    81.375831
-% 59         48129895      1000000    7254858            zarr.zip    24.240338
-% 5          48129895      1000000    7254858                2bit  1689.153258
+% 40         48129895      1000000    7254858       zarr  22.071687
+% 35         48129895      1000000    7254858        bcf  51.749294
+% 36         48129895      1000000    7254858    genozip  10.691384
+% 37         48129895      1000000    7254858        sav  21.249436
+% 38         48129895      1000000    7254858        tsk   1.802636
+% 39         48129895      1000000    7254858        vcf  81.375831
+% 41         48129895      1000000    7254858  zarr_nshf  29.897540
 Turning off the BitShuffle filter for the Zarr dataset,
 however, leads to a substantial increase in decoding speed
-(2.7GiB/s) at the cost of a roughly 25\% increase in storage
-space (31.2GiB up from 23.75GiB for 1 million samples; data not
+(3.9GiB/s) at the cost of a roughly 25\% increase in storage
+space (29.9GiB up from 22.1GiB for 1 million samples; data not
 shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
-decoding speed (2.9GiB/s), but at the cost of much larger storage requirements 
-(214GiB; data not shown).
+decoding speed, but at the cost of much larger storage requirements 
+(219GiB; data not shown).
 Given the relatively small contribution of genotypes to the
@@
 functionality to aid with such storage schema tuning.
-
-The discrepancy between elapsed time and total CPU time for Zarr 
-benchmarks in
-Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
-and~\ref{fig-whole-matrix-decode} is mostly due to the overhead
-of accessing a large number of small chunk files and 
-consequent reduced effectiveness of file-system read-ahead caching.
-While having many independent chunks maps 
-naturally to high-performance parallel access patterns
-in the cloud store setting~\citep{durner2023exploiting} 
-(see also the section on accessing Genomics England data stored on S3), 
-such file access patterns can be problematic in the 
-classical file-system setting.
-The Zarr Zip store (which is simply a Zip archive of the standard directory 
-hierarchy) provides a useful alternative here and can 
-significantly reduce overall processing time. To illustrate this
-we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
-which we measure the time required to decode the entire genotype
-matrix (thus serving as a lower-bound for any subsequent computations).
-In the case of $10^6$ samples, the Zip file requires slightly more 
-% Full data table above ^
-% 58         48129895      1000000    7254858                zarr    23.752762
-% 59         48129895      1000000    7254858            zarr.zip    24.240338
-storage space than the directory hierarchy (23.75GiB and 24.24G
-for file system and Zip file, respectively)
-and while total 
-CPU time is similar (193m vs 187m), the elapsed wall clock
-time for the Zip file version is significantly less 
-(258m vs 188m).
-        % tool  user_time  sys_time    cpu_time   wall_time
-% 11      zarr   11383.82    191.01  192.913833  258.598778
-% 29  zarr.zip   11198.37     43.13  187.358333  188.659252
-
-Another advantage of Zarr is the availabilty of several 
-independent implementations across different programming languages
-(see Methods). Fig~\ref{fig-whole-matrix-compute-zarr-versions}
-compares the CPU time required to perform the computations
-of Fig~\ref{fig-whole-matrix-compute} 
-for af-dist implementations using the Zarr-Python~\cite{zarrpython},
-JZarr~\cite{jzarr},
-and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
-The program using the Tensorstore C++ API is fastest, requiring 
-about 20\% less CPU time than the version using Zarr-Python and Numba
-used in our other benchmarks.
-
-% num_samples  num_sites       tool  user_time  sys_time     wall_time
-% storage  total_time
-% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917 hdd    4.652136
-% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810 hdd    3.701753
-% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377 hdd   12.229383
-% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182 hdd    5.186064
-% Cpp / Zarr python [0.79197226]
 
@@
 the overall dataset size, with the computation requiring around
-2 seconds for 1 million samples. Because of their 
+1 second for 1 million samples. Because of their 
 row-wise encoding, CPU time scales with the number of samples
@@
 the time required to populate a NumPy array with the data in Zarr,
-which is 2 seconds. Wall-clock time (dotted line) is dominated 
-in this case by file I/O. Note that our benchmarking methodology of 
-dropping the file system caches before each run (see Methods) is
-likely not reflective of real world usage where a small and 
-frequently used field such as POS would remain in cache.
-Time to output text for Savvy is not significant
+which is less than a second. Wall-clock time (dotted line) is dominated 
+in this case by file I/O. Time to output text for Savvy is not significant
 for $> 1000$ samples (not shown).
@@
 total of 2.54 TiB of storage over 6,312,488
-chunk files (using a chunk size of $10^4$ variants $\times 10^3$ 
-samples). This is a roughly 5X reduction in total storage 
+chunk files. This is a roughly 5X reduction in total storage 
 space over the original VCF. 
@@
 Finally, to explore the use of cloud computing services with VCF Zarr,
-we uploaded the Zarr directory hierarchy for chromosome 2 to an AWS S3 
-standard bucket. We repeated the benchmarks above on the same subset
-of 562,640 variants, using the same 
-single-threaded sequential Zarr code.
-Extracting the position array required 15s CPU, 23.8s elapsed.
-The af-dist calculation required 2.70min CPU, 5.23min elapsed. 
-Creating the filter mask
-required 3.05min CPU, 11.15min elapsed.
-
-While these benchmarks are comparable with results obtained on the 
-shared file-system in the Genomics England research environment,
-the sequential single-threaded access patterns used do not map
-well to modern cloud computing environments.
-As each request to a cloud object store has a high latency, it 
-is important to have many requests running concurrently to achieve 
-good performance~\citep{durner2023exploiting}.
-To explore the limits of performance we developed a simple prototype
-Zarr implementation, designed to maximise concurrency using asynchrous
-IO and multiple processes (see Methods for details).
-Fig~\ref{fig-s3-throughput} shows the rate at which genotypes 
-can be decoded to RAM as well as rate at which we can compute
-af-dist (using Numba to classify genotypes, as before)
-as we increase the number of parallel processes on
-an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
-The maximum decompression rate of 47.4GiB/s 
-occurs at 64 processes, with a corresponding 
-network data rate of 127 MB/s [FIXME confirm and convert to MiB/s].
-Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
-network access by measuring the rate at which we can download the chunks
-into RAM (without decompressing) using the same parallel and asynchronous
-architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
-and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
-was achieved with 32 processes. Genotype data can therefore be fetched 
-from S3 around an order of magnitude more quickly than it can be decompressed 
-into RAM. This has some interesting consequences
-when we consider the goal of seeking the highest possible compression levels 
-for genotype data and the usual cloud storage cost structuring models (see
-Discussion).
-
-Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
-calculation is performed at a maximum rate of 25.5GiB/s using 128 
-processes. At this rate, the af-dist calculation for the target region 
-was completed in 4.12s [CHECKME] (compare to 5.23min for the naive single-threaded
-method above).
+the zarr files for chromosome 2 were uploaded to an AWS S3 standard bucket.
+Using a single-threaded process on an EC2 instance the benchmarks above were
+repeated. Extracting the position array required 15s CPU, 23.8s elapsed.
+The af-dist calculation required 2.70min CPU, 5.23min elapsed. Creating the filter mask
+required 3.05min CPU, 11.15min elapsed. These times are comparable to the on-premises
+HPC system above.
+
+This single-threaded approach is not optimal for cloud computing, as to optimise
+both runtime and costs parallel processing is required. Using 64 parallel processes
+on a single c5.24xlarge EC2 instance with 96 vCPUs (48 physical) the zarr chunks of
+genotype array could be retrived from S3 and decompressed at a rate of 50.8 GB/s.
+The network data rate was 127 MB/s, far below the capbility of the instance,
+indicating that this process was compute-constrained. When calculating af-dist after
+decompression the throughput was 24.6 GB/s. With this parallel approach the 
+af-dist calculation above was then completed in 4.12s elapsed time.
 This level of throughput would complete af-dist for all chromosomes for all samples
 % compute: $4.08 * 1.27
-% What are the numbers here??
 % io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
-in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
+in 1.27 hours for a compute cost of \$5.18 and an io cost of \$2.28.
 (See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
@@
 
-The compression was performed on the All of Us Researcher Workbench platform 
-(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
-of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
-took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
-time, achieving a parallel efficiency of 86\%. This step incurred a cost of 
-\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed 
-across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with 
-a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a 
-cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across 
-232,809 chunk files.
+The compression was performed on a single compute node with 64 CPUs and 416 GiB of RAM.  
+We first converted the VCF to an ICF, which took 6 hours and 17 minutes of  
+effective runtime and utilized 343 hours of CPU time, achieving a parallel efficiency of 86\%.  
+This step incurred a cost of \$23.88. The resulting ICF representation required 7.5 GiB of storage  
+distributed across 91,232 files. Next, the ICF was converted to Zarr, which took 50 minutes and 32 
+seconds of effective runtime and 49 hours and 49 minutes of CPU time, with a parallel efficiency of 92\%, 
+at a cost of \$3.19. The final Zarr representation occupied 13 GiB of storage across 369,656 chunk files.
 
@@
 \caption{Summary for a selection of the largest VCF Zarr columns produced 
-    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
-    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
-    \label{tab-top-fields-data}}
-\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
+the from the 245,388 exome-like samples from chromosome 20
+using \texttt{vcf2zarr} and chunk sizes of 1,000 variants and 
+10,000 samples.
+For column details see the caption of Table~\ref{tab-genomics-england-data}.
+\label{tab-all-of-us-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
 \toprule
-{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
+{Field} & {type} & {storage} & {compress} & {\%total} \\
 \midrule
-/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
-/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
-/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
-/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
-/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
-/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
-/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
+/call\_GQ & int8 & 1.89G & 87.0 & 45.94\% \\
+/call\_genotype & int8 & 914.22M & 370.0 & 21.70\% \\
+/call\_RGQ & int16 & 729.57M & 460.0 & 17.32\% \\
+/call\_genotype\_mask & bool & 606.02M & 550.0 & 14.39\% \\
+/call\_genotype\_phased & bool & 17.17M & 9700.0 & 0.41\% \\
+/variant\_allele & str & 4.74M & 110.0 & 0.11\% \\
+/variant\_filter & bool & 2.87M & 0.9 & 0.07\% \\
+% /variant\_AN & int32 & 908.75 K & 3.1 & 0.02\% \\
+% /variant\_position & int32 & 810.77 K & 3.4 & 0.02\% \\
+% /sample\_id & object & 357.15 K & 5.4 & 0.01\% \\
+% /variant\_length & int16 & 152.4 K & 9.2 & 0.00\% \\
+% /variant\_id & object & 56.18 K & 99.0 & 0.00\% \\
+% /variant\_quality & float32 & 51.92 K & 54.0 & 0.00\% \\
+% /variant\_id\_mask & bool & 51.89 K & 13.0 & 0.00\% \\
+% /variant\_contig & int16 & 50.6 K & 28.0 & 0.00\% \\
+% /region\_index & int32 & 13.93 K & 1.2 & 0.00\% \\
+% /contig\_length & int64 & 9.24 K & 2.8 & 0.00\% \\
+% /contig\_id & object & 8.33 K & 3.2 & 0.00\% \\
+% /filter\_id & object & 4.48 K & 0.0 & 0.00\% \\
 \bottomrule
@@
 conversion to Zarr. The tasks were run as single jobs on a compute
-node with 2,048 GB RAM and 128 cores. Conversion to ICF required
-13h45min runtime. The ICF representation used a total of 6.77 TiB over
-1,342,569 data storage files. Conversion to Zarr required 18h51min
-runtime, generating a dataset with 32 arrays, consuming a total of 6.6
-TiB storage over 11,984,581 chunk files. The reduction in total
-storage (1.1X) is negligible, mainly due to the inherent difficulty in
-compressing the PL field. The top fields in terms of storage are
-detailed in Table~\ref{tab-spruce-data}.
+node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
+required 13h45min runtime. The ICF representation used a total of 6.77
+TiB over 1,340,804 data storage files. Conversion to Zarr required
+18h36min runtime, generating a dataset with 32 arrays, consuming a
+total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
+total storage (1.1X) is negligible, mainly due to the inherent
+difficulty in compressing the PL field. The output dataset storage
+size is dominated by a major component, call\_PL, which accounts for
+nearly 93\% of the total. call\_PL is yet another field with low
+compression potential due to inherent noisiness.
+
+Therefore, to reduce storage
+requirements we generated a Zarr schema for encoding that uses local
+allele fields with \texttt{vcf2zarr mkschema --local-alleles}, and
+used the generated schema as input to \texttt{vcf2zarr encode}.
+Zarr conversion required 14h09min runtime, generating a dataset with 35 arrays, consuming a total of 2.3TiB storage over 11,235,556 chunk files, with a reduction in storage of 3.2X.
+Therefore, by using local allele fields we achieve approximately 3X improvement of compression.
+The top fields in terms of storage are detailed in Table~\ref{tab-spruce-data}.
 
@@
 \caption{Summary for a selection of the largest VCF Zarr columns produced for 
-1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
-default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
+1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr} with samples chunk size set to 1,063 , variants chunk size to 10,000, and using local allele fields to reduce output size.
+For column details see the caption of Table~\ref{tab-genomics-england-data}.
 \label{tab-spruce-data}}
+
 \begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
 \toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
+Field & type & storage & compress & percentage \\
 \midrule
-  /call\_PL & int16 & 6.1 TiB & 12.0 & 92.75\% \\
-  /call\_genotype & int8 & 282.5 GiB & 26.0 & 4.18\% \\
-  /call\_genotype\_mask & bool & 42.2 GiB & 180.0 & 0.62\% \\
-  /variant\_DP4 & int32 & 26.2 GiB & 2.1 & 0.39\% \\
-  /variant\_MQSBZ & float32 & 13.1 GiB & 1.1 & 0.19\% \\
-  /variant\_RPBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
-  /variant\_MQBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
-  /variant\_SCBZ & float32 & 12.9 GiB & 1.1 & 0.19\% \\
-  /variant\_VDB & float32 & 12.8 GiB & 1.1 & 0.19\% \\
-  /variant\_quality & float32 & 12.8 GiB & 1.1 & 0.19\% \\
-  /variant\_BQBZ & float32 & 12.6 GiB & 1.1 & 0.19\% \\
-  /variant\_SGB & float32 & 12.1 GiB & 1.2 & 0.18\% \\
-  /variant\_position & int32 & 9.7 GiB & 1.4 & 0.14\% \\
-  /variant\_AC & int16 & 7.4 GiB & 2.8 & 0.11\% \\
-  /variant\_DP & int32 & 6.7 GiB & 2.1 & 0.10\% \\
-  /variant\_allele & object & 5.4 GiB & 21.0 & 0.08\% \\
-  /variant\_AN & int16 & 3.1 GiB & 2.3 & 0.05\% \\
-  /call\_genotype\_phased & bool & 2.0 GiB & 1800.0 & 0.03\% \\
-  /variant\_filter & bool & 1.5 GiB & 2.4 & 0.02\% \\
-  /variant\_MQ & int8 & 489.8 MiB & 7.3 & 0.01\% \\
+/call_LPL & int16 & 1.26 TiB & 17.0 & 54.76% \\
+/call_LA & int8 & 415.39 GiB & 18.0 & 17.63% \\
+/call_genotype & int8 & 381.28 GiB & 19.0 & 16.18% \\
+/call_genotype_mask & bool & 103.84 GiB & 71.0 & 4.41% \\
+/variant_DP4 & int32 & 26.24 GiB & 2.1 & 1.11% \\
+/variant_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56% \\
+/variant_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55% \\
+/variant_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55% \\
+/variant_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55% \\
+/variant_VDB & float32 & 12.84 GiB & 1.1 & 0.55% \\
+/variant_quality & float32 & 12.84 GiB & 1.1 & 0.55% \\
+/variant_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54% \\
+/variant_SGB & float32 & 12.12 GiB & 1.2 & 0.51% \\
+/variant_position & int32 & 9.65 GiB & 1.4 & 0.41% \\
+/variant_AC & int16 & 7.42 GiB & 2.8 & 0.31% \\
+/variant_DP & int32 & 6.71 GiB & 2.1 & 0.28% \\
+/variant_allele & object & 5.37 GiB & 21.0 & 0.23% \\
+/variant_AN & int16 & 3.05 GiB & 2.3 & 0.13% \\
+/call_genotype_phased & bool & 1.75 GiB & 2100.0 & 0.07% \\
+/variant_filter & bool & 1.46 GiB & 2.4 & 0.06% \\
+/variant_MQ & int8 & 489.41 MiB & 7.3 & 0.02% \\
+/variant_IMF & float32 & 160.95 MiB & 89.0 & 0.01% \\
+/variant_IDV & int16 & 120.72 MiB & 59.0 & 0.01% \\
 \bottomrule
@@
 as well as  standalone libraries for 
-C++~\citep{ruan2024image}, 
-Java~\cite{jzarr},
-JavaScript~\cite{zarrjs},
+C++~\citep{ruan2024image}, JavaScript~\cite{zarrjs},
 Julia~\cite{zarrjl}, Rust~\citep{zarrs}
@@
 unless otherwise stated.
-All simulation-based benchmarks (excluding Fig~\ref{fig-whole-matrix-compute-zarr-versions}) were performed on a 
+All simulation-based benchmarks were performed on a 
 dual CPU (Intel Xeon E5-2680 v2)
@@
 This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
-2.7 GiB/s Zarr (Zstd), 2.9 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
+3.9 GiB/s Zarr (Zstd), 4.1 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy. 
 \label{fig-whole-matrix-decode}}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics{figures/whole-matrix-compute-zarr-versions}
-\caption{Whole-matrix computation performance using different Zarr 
-implementations. 
-Total CPU time required to run the af-dist calculation
-the data in Figure~\ref{fig-data-storage} using different Zarr implementations.
-Elapsed time is also reported (dotted line). 
-These benchmarks were run on an 8-core CPU (Intel i7-9700) with 32GiB RAM running
-Linux Mint 21.3 with data on an NVMe SSD.
-Note the difference between the
-time reported here and in Fig~\ref{fig-whole-matrix-compute} for Zarr-Python
-is due to different hardware platforms.
-\label{fig-whole-matrix-compute-zarr-versions}}
 \end{figure}
@@
 \end{figure}
-
 \begin{figure}[h]
-\includegraphics{figures/s3-throughput}
-\caption{Scalability of genotype data processing on AWS.
-\label{fig-s3-throughput}}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics{figures/s3-network-throughput}
-\caption{Chunk data download rate on AWS.
-\label{fig-s3-network-throughput}}
-\end{figure}
-
-\begin{figure}[h]
+
 \includegraphics{figures/compression-compressor}

### 2025-01-22 paper.tex
Add missing data and update table format

@@
 
-Therefore, to reduce storage
-requirements we generated a Zarr schema for encoding that uses local
-allele fields with \texttt{vcf2zarr mkschema --local-alleles}, and
-used the generated schema as input to \texttt{vcf2zarr encode}.
-Zarr conversion required 14h09min runtime, generating a dataset with 35 arrays, consuming a total of 2.3TiB storage over 11,235,556 chunk files, with a reduction in storage of 3.2X.
-Therefore, by using local allele fields we achieve approximately 3X improvement of compression.
-The top fields in terms of storage are detailed in Table~\ref{tab-spruce-data}.
+Therefore, to reduce storage requirements we generated a Zarr schema
+for encoding that uses local allele fields with \texttt{vcf2zarr
+mkschema --local-alleles}, and used the generated schema as input to
+\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
+generating a dataset with 35 arrays, consuming a total of 2.3TiB
+storage over 11,235,556 chunk files, with a reduction in storage of
+3.2X. Therefore, by using local allele fields we achieve approximately
+3X improvement of compression. The top fields in terms of storage are
+detailed inTable~\ref{tab-spruce-data}.
 
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns produced for 
-1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr} with samples chunk size set to 1,063 , variants chunk size to 10,000, and using local allele fields to reduce output size.
-For column details see the caption of Table~\ref{tab-genomics-england-data}.
+\caption{Summary for a selection of the largest VCF Zarr columns
+produced for 1,063 VCFs on 12 spruce chromosomes using
+\texttt{vcf2zarr} with samples chunk size set to 1,063, variants
+chunk size to 10,000, and using local allele fields to reduce output
+size. For column details see the caption of
+Table~\ref{tab-genomics-england-data}.
 \label{tab-spruce-data}}
 
-\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
+\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
 \toprule
-Field & type & storage & compress & percentage \\
+{Field} & {type} & {storage} & {compress} & {\%total} \\
 \midrule
-/call_LPL & int16 & 1.26 TiB & 17.0 & 54.76% \\
-/call_LA & int8 & 415.39 GiB & 18.0 & 17.63% \\
-/call_genotype & int8 & 381.28 GiB & 19.0 & 16.18% \\
-/call_genotype_mask & bool & 103.84 GiB & 71.0 & 4.41% \\
-/variant_DP4 & int32 & 26.24 GiB & 2.1 & 1.11% \\
-/variant_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56% \\
-/variant_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55% \\
-/variant_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55% \\
-/variant_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55% \\
-/variant_VDB & float32 & 12.84 GiB & 1.1 & 0.55% \\
-/variant_quality & float32 & 12.84 GiB & 1.1 & 0.55% \\
-/variant_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54% \\
-/variant_SGB & float32 & 12.12 GiB & 1.2 & 0.51% \\
-/variant_position & int32 & 9.65 GiB & 1.4 & 0.41% \\
-/variant_AC & int16 & 7.42 GiB & 2.8 & 0.31% \\
-/variant_DP & int32 & 6.71 GiB & 2.1 & 0.28% \\
-/variant_allele & object & 5.37 GiB & 21.0 & 0.23% \\
-/variant_AN & int16 & 3.05 GiB & 2.3 & 0.13% \\
-/call_genotype_phased & bool & 1.75 GiB & 2100.0 & 0.07% \\
-/variant_filter & bool & 1.46 GiB & 2.4 & 0.06% \\
-/variant_MQ & int8 & 489.41 MiB & 7.3 & 0.02% \\
-/variant_IMF & float32 & 160.95 MiB & 89.0 & 0.01% \\
-/variant_IDV & int16 & 120.72 MiB & 59.0 & 0.01% \\
+/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
+/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
+/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
+/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
+/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
+/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
+/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
+/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
+/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
+/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
+/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
+/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
+/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
+/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
+/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
+/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
+/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
+/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
+/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
+/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
+/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
+/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
+/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
+% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
+% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
+% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
+% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
+% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
+% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
+% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
+% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
+% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
+% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
+% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
+% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
 \bottomrule

### 2025-01-22 paper.tex

--- 2025-01-22 paper.tex
@@
 UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
-Our Future Health~\citep{ofhmain_2025},
 and the All of Us research program~\citep{all2024genomic}
@@
 
-\subsection{Case study: Genomics England WGS data}
+\subsection{Case study: Genomics England 100,000 genomes}
 In this section we demonstrate the utility of VCF Zarr on a large human dataset
@@
 
-\subsection{Case study: Our Future Health genotype data}
+\subsection{Case study: Our Future Health genotype release}
 Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
@@
 
-Release 9~\citep{ofhdatarelease9}
+% this section is probably better placed in acknowledgements?
+This study makes use of de-identified data held by Our Future Health. 
+We would like to acknowledge all the research participants who have donated 
+their data to the Our Future Health research programme 
+\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
+
+Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
 was made available in December 2024. The genetic component includes genotype data 
@@
 \midrule
-% /call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
-% /call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
-% /call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
-% /call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
-% /call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
-% /call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
-% /call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
+/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
+/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
+/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
+/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
+/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
+/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
+/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
 \bottomrule
@@
 
-This study makes use of de-identified data held by Our Future Health. 
-We would like to acknowledge all the research participants who have donated 
-their data to the Our Future Health research programme~\cite{ofhpubpolicy}.
-
 Computation used the Oxford Biomedical Research Computing (BMRC) facility, a

### 2025-01-22 paper.tex
First version of OFH narrative

@@
 
-\subsection{Case study: Our Future Health genotype release}
+\subsection{Case study: Our Future Health genotype data}
 Our Future Health~\cite{ofhmain_2025}  is the UK’s largest ever health research programme. 
@@
 
-% this section is probably better placed in acknowledgements?
-This study makes use of de-identified data held by Our Future Health. 
-We would like to acknowledge all the research participants who have donated 
-their data to the Our Future Health research programme 
-\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
-
-Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
+\begin{table}
+\caption{Summary of the largest VCF Zarr columns for the Our Future Health
+    genotype dataset, with chunk sizes of 1,000 variants and 
+    10,000 samples and applying the BitRound(5) filter to the floating point 
+    fields.
+    For column details, see Table~\ref{tab-genomics-england-data}.
+    \label{tab-ofh-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
+\toprule
+{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
+\midrule
+/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
+/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
+/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
+/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
+/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
+/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
+/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
+/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
+/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
+/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
+/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
+/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
+/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
+/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
+/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
+/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
+/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
+\bottomrule
+\bottomrule
+\end{tabular}
+\end{table}
+
+Release 9~\citep{ofhdatarelease9}
 was made available in December 2024. The genetic component includes genotype data 
@@
 three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
-10,221 variants.
+% 20G     ofh.chr22-b0.vcf.gz
+% 19G     ofh.chr22-b1.vcf.gz
+% 3.3G    ofh.chr22-b2.vcf.gz
+% ==42.3 GiB - assuming the G is power-of-two here
+10,221 variants and 42.3GiB of storage. We first converted the VCF 
+data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
+The ICF representation required a total of 41GiB over XX files.
+We then converted to VCF Zarr format using the default schema, which resulted
+in a 41GiB VCZ. On inspection, this was dominated by three floating point 
+fields LRR, BAF and GS stored 
+to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
+places for GS).
+On inspecting the histograms for these fields
+(Fig~\ref{fig-ofh-field-distributions}) it seemed that such high precision
+was likely unneccessary, and we investigated applying the Quantize
+and Bitround filters from \texttt{numcodecs}~\citep{numcodecs}
+to reduce the precision in stored floating point data, 
+and therefore improve compression.
+Using the Quantize filter with 5 digits of precision on these three fields reduced the 
+overall storage space to 31G with a mean absolute error (MAE, i.e., the average
+difference between the truncated and original values) of 
+% 3   call_LRR    quantize    1.91e-06
+% 4   call_BAF    quantize    1.18e-06
+% 5   call_GS quantize    1.88e-06
+$1.91\times10^{-6}$ (LRR), 
+$1.18\times10^{-6}$ (BAF)  and 
+$1.88\times10^{-6}$ (GS). 
+Using the Bitround filter with a parameter of 5 bits reduced the overall storage for the Zarr to
+XX with an MAE of
+% 0   call_LRR    bitround    6.93e-04
+% 1   call_BAF    bitround    7.86e-04
+% 2   call_GS bitround    3.38e-03
+$6.93\times10^{-4}$ (LRR), 
+$7.86\times10^{-4}$ (BAF)  and 
+$3.38\times10^{-3}$ (GS). 
+The histograms of the truncated fields in all cases were visually indistinguishable 
+from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
+Table~\ref{tab-ofh-data} shows a summary of the Zarr using the BitRound filter.
 
@@
 \begin{figure}[h]
+\begin{tabular}{ccc}
+\includegraphics[width=2.5cm]{ofh_example/plots/default_call_LRR.png}&
+\includegraphics[width=2.5cm]{ofh_example/plots/default_call_BAF.png}&
+\includegraphics[width=2.5cm]{ofh_example/plots/default_call_GS.png}
+\end{tabular}
+\caption{Distribution of floating point fields in the Our Future Health data.
+\label{fig-ofh-field-distributions}}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics{figures/compression-compressor}

### 2025-01-22 paper.tex

@@
 
-For this study, we applied the bio2zarr compression to the chromosome 20 exome-like 
-data from the All of Us dataset. The exome-like data is constructed by extracting from all WGS samples
+For this study, we converted the chromosome 20 exome-like 
+data from the All of Us dataset to VCF Zarr.
+The exome-like data is constructed by extracting from all WGS samples
 the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
@@
 
-The compression was performed on the All of Us Researcher Workbench platform 
+Conversion was performed on the All of Us Researcher Workbench platform 
 (based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
-of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
+of RAM, which cost \$3.80/hour. We first converted the VCF to  ICF, which 
 took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
@@
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns produced 
-    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
-    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
-    \label{tab-top-fields-data}}
+\caption{Summary for a selection of the largest VCF Zarr columns 
+    for the All of Us exome-like data. The chunk size is
+    1,000 variants $\times$ 10,000 samples. 
+    For column details, see Table~\ref{tab-genomics-england-data}.
+    \label{tab-auo-data}}
 \begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
@@
 \midrule
-% /call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
-% /call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
-% /call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
-% /call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
-% /call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
-% /call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
-% /call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
+/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
+/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
+/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
+/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
+/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
+/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
+/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
 \bottomrule
@@
 \end{table}
+
+[TODO discuss local alleles and mention Table~\ref{tab-auo-data}].
 
@@
 analysis workflows that often are tailored for human conditions. For
-instance, the binary Sequence Alignment/Map index format BAI has a 512
+instance, the binary Sequence Alignment/Map index format
+BAI~\cite{li2009sequence} has a 512
 Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
@@
 do, such as the recently sequence mistletoe, with chromosomes larger
-than 8 Gbp~\cite{dtol_mistletoe}. Zarr does not impose this
-limitation, and the coordinate system could easily be converted to
-true genome coordinates for storage in Zarr. This updating procedure
-could be applied to any species of choice.
+than 8 Gbp~\cite{dtol_mistletoe}. Zarr's flexible type 
+system means that coordinates can be stored as 64 or 32 bit integers
+as needed, and there is therefore no particular limit on genome size.
+% limitation, and the coordinate system could easily be converted to
+% true genome coordinates for storage in Zarr. This updating procedure
+% could be applied to any species of choice.
 

### 2025-01-22 paper.tex
Draft of SC2 case study section

@@
 \bottomrule
-\bottomrule
 \end{tabular}
@@
 
+\subsection{Case study: SARS-Cov-2 alignment data}
+The SARS-CoV-2 pandemic resulted in viral sequence data collection at 
+an unprecedented scale, with over 15 million whole genomes incorporated into 
+one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale 
+overwhelmed classical phylogenetic methods, leading to the development
+of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}. 
+The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled 
+set of consensus sequences for 4,471,579 samples, addressing a range of systematic 
+errors in earlier collections. The \texttt{sc2ts} inference tool uses VCF-Zarr as its 
+input format, and Zhan et al have made the Viridian dataset,
+aligned to the reference using MAFFT~\citep{katoh2013mafft},
+available for download~\citep{zhan2023towards}. This dataset has quite
+different properties to those examined in previous case studies, and 
+illustrates the flexibility of Zarr across a range of biological applications.
+
+One of the unusual properties of this dataset is that whole genome alignments 
+are stored, such that information on all 29903 bases is stored for every 
+sample. Given the number of samples and the fact that nearly all bases
+[TODO be concrete?] contain variation it is simply more convenient to 
+store the entire alignment for each sample than to manage the complexity
+of subsetting the sites. Similarly, the usual convention in VCF Zarr 
+is that the \texttt{variant\_alleles} field contains the REF and ALT 
+alleles as observed (such that the zeroth allele is always the reference).
+In this case it is more convenient to encode the alleles as the original
+bases and IUPAC uncertainty codes, such that the allele 0 is always A, 1 is C etc.
+Even in this case with very high sequence diversity, uncertainty codes and 
+storing data for the entire genome, compression rates are very high.
+
+Based on the batch 1 release of Viridian, the Zarr dataset contains
+3,960,704 sequences of length 29903 base pairs along with 31 metadata columns
+encoding information about those samples. This full dataset is stored in a
+single 345MiB Zip file, 241MiB of which is used by the alignment data. This
+gives an average compression ratio of 468X.
+
+Computational accessibility is excellent. For example, computing the number of
+samples with missing or ambiguous data at each site requires around 3 minutes,
+iterating over sample chunks sequentially (5 lines of Python).
+Retrieving the full alignment for a given sample requires 163ms and 
+getting the calls for all samples at a given site 209ms. [TODO confirm hardware
+details]. The storage of metadata and genetic data in a single file also
+makes many queries efficient and convenient. For example, computing the 
+average number of deletions in samples with a Scorpio designation of Alpha
+required x seconds and 5 lines of code using Dask.
+
 \section{Discussion}

### 2025-01-22 paper.tex
Rough outlines of missing sections

@@
 
+\subsection{Interoperability: vcztools}
+To provide compatibility with existing workflows we have 
+developed the ``\texttt{vcztools}'' command line utility which 
+implements a subset of \texttt{bcftools} functionality.
+[TODO DETAILS]
+% (such as 
+% \texttt{view} and \texttt{query}) on a VCF Zarr dataset.
+% Such a tool would speed up some common queries by orders 
+% of magnitude, and reduce the need for user orchestration of 
+% operations among manually split VCF chunks (large VCF datasets
+% are typically split into hundreds of files; see the Genomics
+% England case study). Datasets could then be hosted in cloud
+% object stores, while still presenting file-like semantics
+% for existing workflows. This could provide an evolutionary path,
+% allowing established analysis workflows to coexist with new Zarr-native
+% approaches, working from the same primary data.
+
+\subsection{Interoperability: SAIGE proof-of-concept}
+To demonstrate the potential for integrating VCF Zarr into existing utilities,
+we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
+zhou2020scalable}, a popular package for performing genome wide
+association tests. Using the Tensorstore C++ library~\cite{tensorstore}
+we added support for VCF Zarr via an additional backend class [some details 
+about the implementation].
+
+To test the performance of this implemenentation we compared the time required
+to perform an association test using the simulated dataset of 10,000 samples 
+from Fig~\ref{fig-data-storage}. We generated a phenotype using
+\texttt{tstrait}~\citep{tagami2024tstrait}.
+
+We compared the same analyis using the existing VCF, BCF and Savvy backends
+to our prototype Zarr backend. 
+[Times: BCF 3m, VCF6.5m, Savvy 39s, Zarr 55s]
+
+% BCF:
+% real 3m1.068s user 2m56.944s sys 0m2.941s
+
+% VCF:
+% real 6m30.335s user 6m25.193s sys 0m2.842s
+
+% Savvy:
+% real 0m39.660s user 0m37.498s sys 0m2.183s
+
+% Zarr:
+% real 0m55.957s user 1m10.375s sys 0m12.111s
+
 \section{Discussion}
@@
 
+[TODO REWRITE]
 The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
@@
 
+[TODO REWRITE]
 The second natural direction for development is to create
@@
 
+\subsection{Local alleles}
+[TODO Implementation of local alleles in vcf2zarr]
+
 \subsection{Choosing default compressor settings}
@@
 access patterns.
+
+\subsection{vcztools}
+[TODO Implementation of vcztools]
 
@@
 
+\subsection{AWS parallel Zarr prototype}
+[TODO Implementation of prototype Zarr implementaion]
+
 \section{Availability of source code and requirements}
@@
 (\url{https://pypi.org/project/bio2zarr/}).
+Vcztools (\url{https://github.com/sgkit-dev/vcztools/}) 
+is freely available under the terms of the Apache 2.0 
+and can be installed from the Python Package Index
+(\url{https://pypi.org/project/vcztools/}).
 

### 2025-01-22 paper.tex

@@
 
-% Middle
+% Middle, alphabetical order
 \author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
@@
 \author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[9]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[9]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
-\author[10]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+\author[10]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
@@
 \affil[8]{Genomics England}
-\affil[9]{School of Computer Science, McGill University, Montreal, QC, Canada}
-\affil[10]{Department of Cell and Molecular Biology, National
+\affil[9]{Department of Cell and Molecular Biology, National
   Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
   Uppsala University, Uppsala, Sweden}
+\affil[10]{School of Computer Science, McGill University, Montreal, QC, Canada}
 

### 2025-01-22 paper.tex
Add AOU authors

@@
 % Middle, alphabetical order
-\author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[7]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
-\author[9]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+\author[6,7]{Jeremy Guez} % ?
+\author[8]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[6,7,9]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
+\author[10]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[11]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[12]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[10]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[13]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
@@
 \author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[6,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[8,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 \affil[5]{Tom White Consulting Ltd.}
-\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
+\affil[6]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[7]{Analytic and Translational Genetics Unit, Massachusetts General
+Hospital, Boston, Massachusetts 02114, USA}
+\affil[8]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
 University of Oxford, UK}
-\affil[7]{Wellcome Sanger Institute}
-\affil[8]{Genomics England}
-\affil[9]{Department of Cell and Molecular Biology, National
+\affil[9]{Analytic and Translational Genetics Unit, Massachusetts General
+Hospital, Boston, Massachusetts 02114, USA}
+\affil[10]{Wellcome Sanger Institute}
+\affil[11]{Genomics England}
+\affil[12]{Department of Cell and Molecular Biology, National
   Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
   Uppsala University, Uppsala, Sweden}
-\affil[10]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[13]{School of Computer Science, McGill University, Montreal, QC, Canada}
 

### 2025-01-22 paper.tex
Add Ben Elsworth to author list

--- 2025-01-22 paper.tex
@@
 % Middle, alphabetical order
-\author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[7]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
-\author[9]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+\author[6]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
+\author[7,8]{Jeremy Guez} % ?
+\author[9]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[7,8,10]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
+\author[11]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[12]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[13]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[10]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[14]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
@@
 \author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[6,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[9,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 \affil[5]{Tom White Consulting Ltd.}
-\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
+\affil[6]{Our Future Health}
+\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
+Hospital, Boston, Massachusetts 02114, USA}
+\affil[9]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
 University of Oxford, UK}
-\affil[7]{Wellcome Sanger Institute}
-\affil[8]{Genomics England}
-\affil[9]{Department of Cell and Molecular Biology, National
+\affil[10]{Analytic and Translational Genetics Unit, Massachusetts General
+Hospital, Boston, Massachusetts 02114, USA}
+\affil[11]{Wellcome Sanger Institute}
+\affil[12]{Genomics England}
+\affil[13]{Department of Cell and Molecular Biology, National
   Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
   Uppsala University, Uppsala, Sweden}
-\affil[10]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[14]{School of Computer Science, McGill University, Montreal, QC, Canada}
 
@@
 \section{Acknowledgements}
-% We would like to thank Gil McVean for helpful comments on the manuscript.
+We would like to thank Gil McVean and Ben Neale for helpful discussions and 
+comments on the manuscript.
 This research was made possible through access to data in the National Genomic

### 2025-01-23 paper.tex
Update Jeremy's orcid and accents
Added Jonny Hancox to author list

@@
 \author[6]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
-\author[7,8]{Jeremy Guez} % ?
-\author[9]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[7,8,10]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
-\author[11]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[12]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
-\author[13]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+\author[7,8]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
+\author[9]{Jonny Hancox} % ?
+\author[10]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[7,8,11]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
+\author[12]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[13]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[14]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[14]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[15]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
@@
 \author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[9,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 Hospital, Boston, Massachusetts 02114, USA}
-\affil[9]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
+\affil[9]{Nvidia}
+\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
 University of Oxford, UK}
-\affil[10]{Analytic and Translational Genetics Unit, Massachusetts General
-Hospital, Boston, Massachusetts 02114, USA}
-\affil[11]{Wellcome Sanger Institute}
-\affil[12]{Genomics England}
-\affil[13]{Department of Cell and Molecular Biology, National
+\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[12]{Wellcome Sanger Institute}
+\affil[13]{Genomics England}
+\affil[14]{Department of Cell and Molecular Biology, National
   Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
   Uppsala University, Uppsala, Sweden}
-\affil[14]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}
 

### 2025-01-23 paper.tex
First v. rough draft of GPU stuff

@@
 
+\subsection{Extensibility: GPU acceleration}
+[JK: this is a super rough first narrative draft.]
+Machine learning, and in particular deep learning methods, have led to 
+major breakthroughs in many different fields in recent years [CITES].
+GPUs have been a key enabling factor in this [details and cites].
+The PyData ecosystem [cites] has also enabled this. Techologies like
+[Cupy, lots of others] substantially lower the bar for writing efficient
+GPU code.
+Deep learning methods depend crucially on efficient access to vast 
+datasets for training, and Zarr is emerging as a popular choice
+[cites from geosciences?]. The RAPIDs GPU Accelerated data science 
+library [which does] now supports Zarr through [kvikio, more?].
+
+To demonstrate the ease and efficiency with this data stored in Zarr 
+format can be processed on a GPU, we developed two proof-of-concept
+applications. In the first we re-implement the af-dist calculation
+used throughout as an example of a non-trivial whole-matrix operation
+on GPU. In the second, we demonstrate some emerging technology 
+in which Zarr data can be decompressed directly in GPU memory. 
+In both cases we demonstrate on chromosome 22 of 
+1000 Genomes WGS data~\citep{byrska2022high}.
+
+[Details of af-dist benchmark]
+
+[Details of decompression benchmark]
+
 \section{Discussion}

### 2025-01-23 software/vcf_parse/vcf_parse.c
Add basic VCF parser to measure rate of reading

@@ -1,0 +1,39 @@
+#include <assert.h>
+#include <err.h>
+#include <hts.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <vcf.h>
+
+static void read_vcf(char *path) {
+    htsFile *h = hts_open(path, "r");
+
+    if (h == NULL) {
+        errx(EXIT_FAILURE, "Error opening VCF");
+    }
+    bcf_hdr_t *hdr = bcf_hdr_read(h);
+    if (hdr == NULL) {
+        errx(EXIT_FAILURE, "Error reading header");
+    }
+    bcf1_t *b = bcf_init();
+    int ret = 0;
+    int records = 0;
+    while ((ret = bcf_read(h, hdr, b)) == 0) {
+        /* printf("%d\n", b->pos); */
+        records++;
+    }
+    printf("read = %d records\n", records);
+
+    bcf_destroy(b);
+    bcf_hdr_destroy(hdr);
+    hts_close(h);
+}
+
+int main(int argc, char **argv) {
+    if (argc != 2) {
+        errx(EXIT_FAILURE, "usage: vcf_parse <filename>\n");
+    }
+    printf("Reading VCF %s\n", argv[1]);
+    read_vcf(argv[1]);
+    return 0;
+}

### 2025-01-23 paper.tex

@@
 implements a subset of \texttt{bcftools} functionality.
-[TODO DETAILS]
-% (such as 
-% \texttt{view} and \texttt{query}) on a VCF Zarr dataset.
-% Such a tool would speed up some common queries by orders 
-% of magnitude, and reduce the need for user orchestration of 
-% operations among manually split VCF chunks (large VCF datasets
-% are typically split into hundreds of files; see the Genomics
-% England case study). Datasets could then be hosted in cloud
-% object stores, while still presenting file-like semantics
-% for existing workflows. This could provide an evolutionary path,
-% allowing established analysis workflows to coexist with new Zarr-native
-% approaches, working from the same primary data.
+Vcztools is intended to be a drop-in replacement for \texttt{bcftools},
+and compatibility being a primary objective.
+Vcztools is primarily intended to provide an evolutionary path for 
+VCF Zarr adoption,
+allowing established analysis workflows to coexist with new Zarr-native
+approaches, working from the same primary data.
+Vcztools currently supports Zarr stored in the local file system
+and Zip archives, and support for major cloud stores will be added in the near
+future. [JK: note this is a pretty simple feature, maybe we should just try 
+to add it?]
+% Mention Pyparsing?
+It is written in a mixture of Python and C, using
+PyRanges~\citep{stovner2020pyranges} to perform efficient range queries.
+
+The initial version of vcztools implements a subset of the \texttt{query}
+and \texttt{view} commands. We provide an example based on the simulated 
+dataset of $10^5$ samples of Fig~\ref{fig-data-storage} to illustrate 
+the functionality and 
+performance of vcztools. First, we extract the position column 
+using the \texttt{query} command. Using bcftools
+(\texttt{bcftools query -f '\%POS\textbackslash n' chr21\_10\_5.bcf})
+it took 8m57s to extract the position values for the 2,365,367
+variants from the BCF file. 
+Using vcztools
+(\texttt{vcztools query -f '\%POS\textbackslash n' chr21\_10\_5.zarr})
+required 51s (note that this operation could be performed much
+more quickly and may be worth optimising for, see 
+Fig~\ref{fig-column-extract}). This illustrates the general principle
+that vcztools only retrieves information necessary to fulfil a particular 
+query, which can have a significant impact with large datasets.
+
+To form a practical replacement for bcftools in production pipelines
+it is important that vcztools can produce VCF text at a reasonable
+rate. To do this we measure the time taken to generate VCF text for the 
+last 10,000 variants in the dataset (using the previously extracted 
+position column to determine the coordinates, \texttt{vcztools view
+chr21\_10\_5.zarr -r1:47958263-}). This took 30s, at rate of 124.3 MiB
+of VCF text produced per second. The bcftools command on the corresponding
+BCF file took 25s (152.64 MiB/s) and on the VCF file it took 63s (60.57 MiB/s).
+Thus, vcztools view can produce VCF at an intermediate rate between 
+bcftools reading from BCF and VCF files.
+
+Vcztools is primarily intended to used as a replacement for bcftools 
+in existing workflows based on Unix pipes, where VCF text is sequentially
+filtered and processed by different processes in the classical
+Unix manner. In this context the most important property is that vcztools
+can produce VCF text as quickly as other processes in the pipeline can consume
+it. To measure the rate at which VCF can be parsed, we wrote a simple
+program that reads a VCF file and parses it line-by-line using \texttt{htslib}
+APIs~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
+above (3.73 GiB) this took 31s, giving a parse rate of 123.1 MiB/s
+(almost identical to the write-rate of vcztools above). Given this limiting
+factor on overall throughput,
+it is unlikely that \texttt{vcztools view} would be a bottleneck 
+in practical processing pipelines.
 

### 2025-01-23 paper.tex
Tidy up GeL notebooks and text fixes

--- 2025-01-23 paper.tex
@@
 
-% Middle, alphabetical order
-\author[6]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
-\author[7,8]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
-\author[9]{Jonny Hancox} % ?
-\author[10]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[7,8,11]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
-\author[12]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[13]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
-\author[14]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+% Middle
+\author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[7]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[15]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[9]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[10]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 
@@
 \author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[6,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 \affil[5]{Tom White Consulting Ltd.}
-\affil[6]{Our Future Health}
-\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
-Hospital, Boston, Massachusetts 02114, USA}
-\affil[9]{Nvidia}
-\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
+\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
 University of Oxford, UK}
-\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[12]{Wellcome Sanger Institute}
-\affil[13]{Genomics England}
-\affil[14]{Department of Cell and Molecular Biology, National
+\affil[7]{Wellcome Sanger Institute}
+\affil[8]{Genomics England}
+\affil[9]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[10]{Department of Cell and Molecular Biology, National
   Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
   Uppsala University, Uppsala, Sweden}
-\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}
 
@@
 occurs at 64 processes, with a corresponding 
-network data rate of 127 MB/s [FIXME confirm and convert to MiB/s].
+network data rate of 121 MiB/s
 Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
@@
 processes. At this rate, the af-dist calculation for the target region 
-was completed in 4.12s [CHECKME] (compare to 5.23min for the naive single-threaded
+was completed in 4.15s (compare to 5.23min for the naive single-threaded
 method above).
 This level of throughput would complete af-dist for all chromosomes for all samples
-% compute: $4.08 * 1.27
-% What are the numbers here??
-% io: (((722000000/10000)*(79000/1000))/1000)*$0.0004
+% compute: cost of c5.9xlarge per hour * 1.27
+% ==                             $4.08 * 1.27
+% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
+% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
 in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
@@
 
-\begin{table}
-\caption{Summary of the largest VCF Zarr columns for the Our Future Health
-    genotype dataset, with chunk sizes of 1,000 variants and 
-    10,000 samples and applying the BitRound(5) filter to the floating point 
-    fields.
-    For column details, see Table~\ref{tab-genomics-england-data}.
-    \label{tab-ofh-data}}
-\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
-\toprule
-{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
-\midrule
-/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
-/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
-/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
-/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
-/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
-/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
-/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
-/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
-/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
-/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
-/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
-/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
-/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
-/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
-/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
-/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
-/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Release 9~\citep{ofhdatarelease9}
+% this section is probably better placed in acknowledgements?
+This study makes use of de-identified data held by Our Future Health. 
+We would like to acknowledge all the research participants who have donated 
+their data to the Our Future Health research programme 
+\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}
+
+Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
 was made available in December 2024. The genetic component includes genotype data 
@@
 three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
-% 20G     ofh.chr22-b0.vcf.gz
-% 19G     ofh.chr22-b1.vcf.gz
-% 3.3G    ofh.chr22-b2.vcf.gz
-% ==42.3 GiB - assuming the G is power-of-two here
-10,221 variants and 42.3GiB of storage. We first converted the VCF 
-data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
-The ICF representation required a total of 41GiB over XX files.
-We then converted to VCF Zarr format using the default schema, which resulted
-in a 41GiB VCZ. On inspection, this was dominated by three floating point 
-fields LRR, BAF and GS stored 
-to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
-places for GS).
-On inspecting the histograms for these fields
-(Fig~\ref{fig-ofh-field-distributions}) it seemed that such high precision
-was likely unneccessary, and we investigated applying the Quantize
-and Bitround filters from \texttt{numcodecs}~\citep{numcodecs}
-to reduce the precision in stored floating point data, 
-and therefore improve compression.
-Using the Quantize filter with 5 digits of precision on these three fields reduced the 
-overall storage space to 31G with a mean absolute error (MAE, i.e., the average
-difference between the truncated and original values) of 
-% 3   call_LRR    quantize    1.91e-06
-% 4   call_BAF    quantize    1.18e-06
-% 5   call_GS quantize    1.88e-06
-$1.91\times10^{-6}$ (LRR), 
-$1.18\times10^{-6}$ (BAF)  and 
-$1.88\times10^{-6}$ (GS). 
-Using the Bitround filter with a parameter of 5 bits reduced the overall storage for the Zarr to
-XX with an MAE of
-% 0   call_LRR    bitround    6.93e-04
-% 1   call_BAF    bitround    7.86e-04
-% 2   call_GS bitround    3.38e-03
-$6.93\times10^{-4}$ (LRR), 
-$7.86\times10^{-4}$ (BAF)  and 
-$3.38\times10^{-3}$ (GS). 
-The histograms of the truncated fields in all cases were visually indistinguishable 
-from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
-Table~\ref{tab-ofh-data} shows a summary of the Zarr using the BitRound filter.
+10,221 variants.
 
@@
 
-For this study, we converted the chromosome 20 exome-like 
-data from the All of Us dataset to VCF Zarr.
-The exome-like data is constructed by extracting from all WGS samples
+For this study, we applied the bio2zarr compression to the chromosome 20 exome-like 
+data from the All of Us dataset. The exome-like data is constructed by extracting from all WGS samples
 the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic 
@@
 
-Conversion was performed on the All of Us Researcher Workbench platform 
+The compression was performed on the All of Us Researcher Workbench platform 
 (based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB 
-of RAM, which cost \$3.80/hour. We first converted the VCF to  ICF, which 
+of RAM, which cost \$3.80/hour. We first converted the VCF to an ICF, which 
 took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU 
@@
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns 
-    for the All of Us exome-like data. The chunk size is
-    1,000 variants $\times$ 10,000 samples. 
-    For column details, see Table~\ref{tab-genomics-england-data}.
-    \label{tab-auo-data}}
+\caption{Summary for a selection of the largest VCF Zarr columns produced 
+    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
+    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
+    \label{tab-top-fields-data}}
 \begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
@@
 \midrule
-/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
-/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
-/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
-/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
-/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
-/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
-/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
+/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
+/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
+/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
+/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
+/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
+/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
+/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
 \bottomrule
@@
 \end{table}
-
-[TODO discuss local alleles and mention Table~\ref{tab-auo-data}].
 
@@
 conversion to Zarr. The tasks were run as single jobs on a compute
-node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
-required 13h45min runtime. The ICF representation used a total of 6.77
-TiB over 1,340,804 data storage files. Conversion to Zarr required
-18h36min runtime, generating a dataset with 32 arrays, consuming a
-total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
-total storage (1.1X) is negligible, mainly due to the inherent
-difficulty in compressing the PL field. The output dataset storage
-size is dominated by a major component, call\_PL, which accounts for
-nearly 93\% of the total. call\_PL is yet another field with low
-compression potential due to inherent noisiness.
-
-Therefore, to reduce storage requirements we generated a Zarr schema
-for encoding that uses local allele fields with \texttt{vcf2zarr
-mkschema --local-alleles}, and used the generated schema as input to
-\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
-generating a dataset with 35 arrays, consuming a total of 2.3TiB
-storage over 11,235,556 chunk files, with a reduction in storage of
-3.2X. Therefore, by using local allele fields we achieve approximately
-3X improvement of compression. The top fields in terms of storage are
-detailed inTable~\ref{tab-spruce-data}.
+node with 2,048 GB RAM and 128 cores. Conversion to ICF required
+13h45min runtime. The ICF representation used a total of 6.77 TiB over
+1,342,569 data storage files. Conversion to Zarr required 18h51min
+runtime, generating a dataset with 32 arrays, consuming a total of 6.6
+TiB storage over 11,984,581 chunk files. The reduction in total
+storage (1.1X) is negligible, mainly due to the inherent difficulty in
+compressing the PL field. The top fields in terms of storage are
+detailed in Table~\ref{tab-spruce-data}.
 
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns
-produced for 1,063 VCFs on 12 spruce chromosomes using
-\texttt{vcf2zarr} with samples chunk size set to 1,063, variants
-chunk size to 10,000, and using local allele fields to reduce output
-size. For column details see the caption of
-Table~\ref{tab-genomics-england-data}.
+\caption{Summary for a selection of the largest VCF Zarr columns produced for 
+1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
+default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
 \label{tab-spruce-data}}
-
-\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
+\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
 \toprule
@@
   \midrule
-/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
-/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
-/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
-/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
-/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
-/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
-/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
-/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
-/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
-/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
-/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
-/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
-/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
-/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
-/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
-/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
-/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
-/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
-/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
-/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
-/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
-% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
-% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
-% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
-% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
-% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
-% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
-% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
-% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
-% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
-% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
+  /call\_PL & int16 & 6.1 TiB & 12.0 & 92.75\% \\
+  /call\_genotype & int8 & 282.5 GiB & 26.0 & 4.18\% \\
+  /call\_genotype\_mask & bool & 42.2 GiB & 180.0 & 0.62\% \\
+  /variant\_DP4 & int32 & 26.2 GiB & 2.1 & 0.39\% \\
+  /variant\_MQSBZ & float32 & 13.1 GiB & 1.1 & 0.19\% \\
+  /variant\_RPBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
+  /variant\_MQBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
+  /variant\_SCBZ & float32 & 12.9 GiB & 1.1 & 0.19\% \\
+  /variant\_VDB & float32 & 12.8 GiB & 1.1 & 0.19\% \\
+  /variant\_quality & float32 & 12.8 GiB & 1.1 & 0.19\% \\
+  /variant\_BQBZ & float32 & 12.6 GiB & 1.1 & 0.19\% \\
+  /variant\_SGB & float32 & 12.1 GiB & 1.2 & 0.18\% \\
+  /variant\_position & int32 & 9.7 GiB & 1.4 & 0.14\% \\
+  /variant\_AC & int16 & 7.4 GiB & 2.8 & 0.11\% \\
+  /variant\_DP & int32 & 6.7 GiB & 2.1 & 0.10\% \\
+  /variant\_allele & object & 5.4 GiB & 21.0 & 0.08\% \\
+  /variant\_AN & int16 & 3.1 GiB & 2.3 & 0.05\% \\
+  /call\_genotype\_phased & bool & 2.0 GiB & 1800.0 & 0.03\% \\
+  /variant\_filter & bool & 1.5 GiB & 2.4 & 0.02\% \\
+  /variant\_MQ & int8 & 489.8 MiB & 7.3 & 0.01\% \\
   \bottomrule
@@
 
+Table~\ref{tab-spruce-data} shows that the dataset storage size is
+dominated by a major component, call\_PL, which accounts for nearly
+93\% of the total. call\_PL is yet another field with low compression
+potential due to inherent noisiness.
+
 Large genomes from non-model organisms pose particular problems for
 analysis workflows that often are tailored for human conditions. For
-instance, the binary Sequence Alignment/Map index format
-BAI~\cite{li2009sequence} has a 512
+instance, the binary Sequence Alignment/Map index format BAI has a 512
 Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
@@
 do, such as the recently sequence mistletoe, with chromosomes larger
-than 8 Gbp~\cite{dtol_mistletoe}. Zarr's flexible type 
-system means that coordinates can be stored as 64 or 32 bit integers
-as needed, and there is therefore no particular limit on genome size.
-% limitation, and the coordinate system could easily be converted to
-% true genome coordinates for storage in Zarr. This updating procedure
-% could be applied to any species of choice.
-
-\subsection{Case study: SARS-Cov-2 alignment data}
-The SARS-CoV-2 pandemic resulted in viral sequence data collection at 
-an unprecedented scale, with over 15 million whole genomes incorporated into 
-one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale 
-overwhelmed classical phylogenetic methods, leading to the development
-of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}. 
-The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled 
-set of consensus sequences for 4,471,579 samples, addressing a range of systematic 
-errors in earlier collections. The \texttt{sc2ts} inference tool uses VCF-Zarr as its 
-input format, and Zhan et al have made the Viridian dataset,
-aligned to the reference using MAFFT~\citep{katoh2013mafft},
-available for download~\citep{zhan2023towards}. This dataset has quite
-different properties to those examined in previous case studies, and 
-illustrates the flexibility of Zarr across a range of biological applications.
-
-One of the unusual properties of this dataset is that whole genome alignments 
-are stored, such that information on all 29903 bases is stored for every 
-sample. Given the number of samples and the fact that nearly all bases
-[TODO be concrete?] contain variation it is simply more convenient to 
-store the entire alignment for each sample than to manage the complexity
-of subsetting the sites. Similarly, the usual convention in VCF Zarr 
-is that the \texttt{variant\_alleles} field contains the REF and ALT 
-alleles as observed (such that the zeroth allele is always the reference).
-In this case it is more convenient to encode the alleles as the original
-bases and IUPAC uncertainty codes, such that the allele 0 is always A, 1 is C etc.
-Even in this case with very high sequence diversity, uncertainty codes and 
-storing data for the entire genome, compression rates are very high.
-
-Based on the batch 1 release of Viridian, the Zarr dataset contains
-3,960,704 sequences of length 29903 base pairs along with 31 metadata columns
-encoding information about those samples. This full dataset is stored in a
-single 345MiB Zip file, 241MiB of which is used by the alignment data. This
-gives an average compression ratio of 468X.
-
-Computational accessibility is excellent. For example, computing the number of
-samples with missing or ambiguous data at each site requires around 3 minutes,
-iterating over sample chunks sequentially (5 lines of Python).
-Retrieving the full alignment for a given sample requires 163ms and 
-getting the calls for all samples at a given site 209ms. [TODO confirm hardware
-details]. The storage of metadata and genetic data in a single file also
-makes many queries efficient and convenient. For example, computing the 
-average number of deletions in samples with a Scorpio designation of Alpha
-required x seconds and 5 lines of code using Dask.
-
-\subsection{Interoperability: vcztools}
-To provide compatibility with existing workflows we have 
-developed the ``\texttt{vcztools}'' command line utility which 
-implements a subset of \texttt{bcftools} functionality.
-[TODO DETAILS]
-% (such as 
-% \texttt{view} and \texttt{query}) on a VCF Zarr dataset.
-% Such a tool would speed up some common queries by orders 
-% of magnitude, and reduce the need for user orchestration of 
-% operations among manually split VCF chunks (large VCF datasets
-% are typically split into hundreds of files; see the Genomics
-% England case study). Datasets could then be hosted in cloud
-% object stores, while still presenting file-like semantics
-% for existing workflows. This could provide an evolutionary path,
-% allowing established analysis workflows to coexist with new Zarr-native
-% approaches, working from the same primary data.
-
-\subsection{Interoperability: SAIGE proof-of-concept}
-To demonstrate the potential for integrating VCF Zarr into existing utilities,
-we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
-zhou2020scalable}, a popular package for performing genome wide
-association tests. Using the Tensorstore C++ library~\cite{tensorstore}
-we added support for VCF Zarr via an additional backend class [some details 
-about the implementation].
-
-To test the performance of this implemenentation we compared the time required
-to perform an association test using the simulated dataset of 10,000 samples 
-from Fig~\ref{fig-data-storage}. We generated a phenotype using
-\texttt{tstrait}~\citep{tagami2024tstrait}.
-
-We compared the same analyis using the existing VCF, BCF and Savvy backends
-to our prototype Zarr backend. 
-[Times: BCF 3m, VCF6.5m, Savvy 39s, Zarr 55s]
-
-% BCF:
-% real 3m1.068s user 2m56.944s sys 0m2.941s
-
-% VCF:
-% real 6m30.335s user 6m25.193s sys 0m2.842s
-
-% Savvy:
-% real 0m39.660s user 0m37.498s sys 0m2.183s
-
-% Zarr:
-% real 0m55.957s user 1m10.375s sys 0m12.111s
+than 8 Gbp~\cite{dtol_mistletoe}. Zarr does not impose this
+limitation, and the coordinate system could easily be converted to
+true genome coordinates for storage in Zarr. This updating procedure
+could be applied to any species of choice.
 
@@
 
-[TODO REWRITE]
 The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
@@
 
-[TODO REWRITE]
 The second natural direction for development is to create
@@
 
-\subsection{Local alleles}
-[TODO Implementation of local alleles in vcf2zarr]
-
 \subsection{Choosing default compressor settings}
@@
 access patterns.
-
-\subsection{vcztools}
-[TODO Implementation of vcztools]
 
@@
 
-\subsection{AWS parallel Zarr prototype}
-[TODO Implementation of prototype Zarr implementaion]
-
 \section{Availability of source code and requirements}
@@
 (\url{https://pypi.org/project/bio2zarr/}).
-Vcztools (\url{https://github.com/sgkit-dev/vcztools/}) 
-is freely available under the terms of the Apache 2.0 
-and can be installed from the Python Package Index
-(\url{https://pypi.org/project/vcztools/}).
 
@@
 \section{Acknowledgements}
-We would like to thank Gil McVean and Ben Neale for helpful discussions and 
-comments on the manuscript.
+% We would like to thank Gil McVean for helpful comments on the manuscript.
 This research was made possible through access to data in the National Genomic
@@
 infrastructure.
-
-This study makes use of de-identified data held by Our Future Health. 
-We would like to acknowledge all the research participants who have donated 
-their data to the Our Future Health research programme~\cite{ofhpubpolicy}.
 
@@
 \begin{figure}[h]
-\begin{tabular}{ccc}
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_LRR.png}&
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_BAF.png}&
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_GS.png}
-\end{tabular}
-\caption{Distribution of floating point fields in the Our Future Health data.
-\label{fig-ofh-field-distributions}}
-\end{figure}
-
-\begin{figure}[h]
 \includegraphics{figures/compression-compressor}

### 2025-01-23 paper.tex
First para for GPU section

@@
 In both cases we demonstrate on chromosome 22 of 
-1000 Genomes WGS data~\citep{byrska2022high}.
-
-[Details of af-dist benchmark]
-
-[Details of decompression benchmark]
+1000 Genomes WGS data~\citep{byrska2022high}. This consists
+of 26,504 variants and 2504 samples [and more details about the zarr].
+Benchmarks in this section were performed on a
+machine with an Intel Xeon Silver 4216 processor,
+Nvidia A6000 RTX GPU and NVMe SSD.
+
+We first illustrate the ease with which computations can be 
+be peformed on the GPU, with little or no preprocessing of the 
+data. To do this we reimplemented the af-dist calculation
+described above using CuPy~\citep{cupy}, a GPU accelerated library
+of core operations for scientific computing.
+All of the operations required to compute  af-dist are 
+performed on the GPU, mostly with minimal changes required 
+(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
+We perform the core operation of genotype classification using the
+CuPy JIT compiler to generate a custom CUDA kernel, which 
+is also very similar to the Numba compiled function used in other
+benchmarks. In the simplest version, we iterate over the chunks
+in the \texttt{call\_genotype} array, decompress each to a local
+NumPy array and then copying the data to GPU memory by creating 
+the corresponding CuPy array. For this version, the total time
+in the genotype classification kernel was 0.045s, with an 
+overall data transfer time of 0.19s (we discount decompression
+time here; see next paragraph). For comparison, 
+our standard Numba compiled classification function required 
+a total of 1.44s running on a single CPU core.
+Another approach in which we load the entire 
+\texttt{call\_genotype} array into GPU memory at once (possible
+here as the $96514 \times 2504 \times 2$ array requires 
+460MiB) and use
+more sophisticated kernel in which we map GPU resources to
+subsets of the array in two dimensions, reduces transfer 
+time to 0.136s (and processing time to $2.43\times10^{-4}$s.
+
+Chunk decompression comprises a significant fraction of processing 
+time [FIXME crossrefs?]
 

### 2025-01-23 paper.tex
Update second GPU section

@@
 1000 Genomes WGS data~\citep{byrska2022high}. This consists
-of 26,504 variants and 2504 samples [and more details about the zarr].
+of 96,514 variants and 2504 samples [and more details about the zarr].
 Benchmarks in this section were performed on a
@@
 Chunk decompression comprises a significant fraction of processing 
-time [FIXME crossrefs?]
+time [FIXME crossrefs?] and host-to-GPU data transfer is 
+often a bottleneck. It is therefore useful in some applications 
+to directly decompress Zarr chunks into GPU memory. 
+KvikIO is a Python package that provides access to optimised
+GPU IO routines, and providing support for Zarr along with
+access to several GPU optimised compression codecs.
+To explore the performance of these codecs we recompressed
+the \texttt{call\_AD} (allele depth) array
+(96514, 2504, 7) [int16 more details]. In call cases it took around 1s 
+to compress and store this 3.15GiB array. [Details about compression codecs?]
+We then measured the time to decompress these arrays and 
+copy back to host memory [Why??]. This required about 0.6s with the 
+fastest codecs. For reference, using the Zstandard codec on a single
+CPU thread required about 7.6s.
 

### 2025-01-23 paper.tex
Update some author details

@@
 \author[7,8]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
-\author[9]{Jonny Hancox} % ?
+\author[9]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
 \author[10]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
@@
 \affil[5]{Tom White Consulting Ltd.}
-\affil[6]{Our Future Health}
+\affil[6]{Our Future Health, Manchester, UK.}
 \affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
@@
 Hospital, Boston, Massachusetts 02114, USA}
-\affil[9]{Nvidia}
+\affil[9]{NVIDIA Ltd, Reading, UK}
 \affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
@@
 to directly decompress Zarr chunks into GPU memory. 
-KvikIO is a Python package that provides access to optimised
+KvikIO~\cite{kvikio} is a Python package that provides access to optimised
 GPU IO routines, and providing support for Zarr along with

### 2025-01-23 paper.tex
Add lungfish as example of large genome

@@
 spruce does not have chromosomes that exceed the limit, other species
-do, such as the recently sequence mistletoe, with chromosomes larger
-than 8 Gbp~\cite{dtol_mistletoe}. Zarr's flexible type 
+do, such as the 87.2 Gbp South American lungfish, whose 19 chromosomes
+are all but one larger than the 3.1 Gbp human
+genome~\cite{schartl_GenomesAllLungfish_2024}. Zarr's flexible type
 system means that coordinates can be stored as 64 or 32 bit integers

### 2025-01-23 paper.tex

@@
 1000 Genomes WGS data~\citep{byrska2022high}. This consists
-of 96,514 variants and 2504 samples [and more details about the zarr].
+of 96,514 variants and 2504 samples, converted to Zarr format
+using a chunk size of 10,000 variants $\times$ 1,000 variants.
 Benchmarks in this section were performed on a
@@
 benchmarks. In the simplest version, we iterate over the chunks
-in the \texttt{call\_genotype} array, decompress each to a local
+in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
+a total of 7.82MiB), decompress each to a local
 NumPy array and then copying the data to GPU memory by creating 
 the corresponding CuPy array. For this version, the total time
-in the genotype classification kernel was 0.045s, with an 
-overall data transfer time of 0.19s (we discount decompression
+in the genotype classification kernel was 25ms with an 
+overall data transfer time of 179ms (we discount decompression
 time here; see next paragraph). For comparison, 
 our standard Numba compiled classification function required 
-a total of 1.44s running on a single CPU core.
+a total of 1.46s running on a single CPU core.
 Another approach in which we load the entire 
@@
 here as the $96514 \times 2504 \times 2$ array requires 
-460MiB) and use
-more sophisticated kernel in which we map GPU resources to
-subsets of the array in two dimensions, reduces transfer 
-time to 0.136s (and processing time to $2.43\times10^{-4}$s.
+460MiB) and uses
+a more sophisticated kernel in which we map GPU resources to
+subsets of the array in two dimensions, 
+reduces transfer 
+time to 138ms and processing time to 6ms.
 
@@
 the \texttt{call\_AD} (allele depth) array
-(96514, 2504, 7) [int16 more details]. In call cases it took around 1s 
+(96514, 2504, 7) [int16 more details]. In all cases it took around 1s 
 to compress and store this 3.15GiB array. [Details about compression codecs?]
 We then measured the time to decompress these arrays and 
-copy back to host memory [Why??]. This required about 0.6s with the 
-fastest codecs. For reference, using the Zstandard codec on a single
-CPU thread required about 7.6s.
+copy back to host memory. This required about 0.6s with the 
+fastest codecs. For reference, using the Zstandard codec and the
+standard CPU Zarr code (which does some limited multithreading for decompression
+by default) required about 7.6s. [Some discussion of how to interpret
+these benchmarks. We're just trying to illustrate the powerful toolchain
+ that is immediately to hand when data is stored in Zarr format.]
 

### 2025-01-25 paper.tex

@@
 To provide compatibility with existing workflows we have
-developed the ``\texttt{vcztools}'' command line utility which 
+developed the \texttt{vcztools} command line utility which
 implements a subset of \texttt{bcftools} functionality.
 Vcztools is intended to be a drop-in replacement for \texttt{bcftools},
-and compatibility being a primary objective.
-Vcztools is primarily intended to provide an evolutionary path for 
-VCF Zarr adoption,
-allowing established analysis workflows to coexist with new Zarr-native
-approaches, working from the same primary data.
+providing an evolutionary path for VCF Zarr adoption,
+with compatibility being a core objective.
 Vcztools currently supports Zarr stored in the local file system
@@
 to add it?]
-% Mention Pyparsing?
-It is written in a mixture of Python and C, using
-PyRanges~\citep{stovner2020pyranges} to perform efficient range queries.
-
-The initial version of vcztools implements a subset of the \texttt{query}
+Thus, users can convert their VCF data to Zarr using \texttt{bio2zarr},
+store it in a local file system, Zip file or cloud store as desired
+and then access this data source within existing processing pipelines,
+essentially replacing \texttt{bcftools view <vcf file>} with
+\texttt{vcztools view <Zarr URL>}.
+This allows established analysis workflows to coexist with new Zarr-native
+approaches, working from the same primary data.
+
+The initial version of \texttt{vcztools} implements a subset of the \texttt{query}
 and \texttt{view} commands. We provide an example based on the simulated
@@
 the functionality and
-performance of vcztools. First, we extract the position column 
-using the \texttt{query} command. Using bcftools
+performance of \texttt{vcztools}. First, we extract the position column
+using the \texttt{query} command. Using \texttt{bcftools}
 (\texttt{bcftools query -f '\%POS\textbackslash n' chr21\_10\_5.bcf})
@@
 Using vcztools
-(\texttt{vcztools query -f '\%POS\textbackslash n' chr21\_10\_5.zarr})
+(\texttt{vcztools query -f '\%POS\textbackslash n' chr21\_10\_5.vcz})
 required 51s (note that this operation could be performed much
-more quickly and may be worth optimising for, see 
+more quickly and may be worth optimising for within \texttt{vcztools}, see
 Fig~\ref{fig-column-extract}). This illustrates the general principle
-that vcztools only retrieves information necessary to fulfil a particular 
+that \texttt{vcztools} only retrieves information necessary to fulfil a particular
 query, which can have a significant impact with large datasets.
@@
 To form a practical replacement for bcftools in production pipelines
-it is important that vcztools can produce VCF text at a reasonable
+it is important that \texttt{vcztools} can produce VCF text at a reasonable
 rate. To do this we measure the time taken to generate VCF text for the
 last 10,000 variants in the dataset (using the previously extracted
-position column to determine the coordinates, \texttt{vcztools view
-chr21\_10\_5.zarr -r1:47958263-}). This took 30s, at rate of 124.3 MiB
-of VCF text produced per second. The bcftools command on the corresponding
+position column to determine the coordinates), \texttt{vcztools view
+chr21\_10\_5.vcz -r1:47958263-}). This took 30s, at a rate of 124.3 MiB
+of VCF text per second. The bcftools command on the corresponding
 BCF file took 25s (152.64 MiB/s) and on the VCF file it took 63s (60.57 MiB/s).
-Thus, vcztools view can produce VCF at an intermediate rate between 
-bcftools reading from BCF and VCF files.
-
+
+% TODO texttt
 Vcztools is primarily intended to used as a replacement for bcftools
-in existing workflows based on Unix pipes, where VCF text is sequentially
+in existing workflows based on pipes, where VCF text is sequentially
 filtered and processed by different processes in the classical
@@
 can produce VCF text as quickly as other processes in the pipeline can consume
-it. To measure the rate at which VCF can be parsed, we wrote a simple
-program that reads a VCF file and parses it line-by-line using \texttt{htslib}
-APIs~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
+it. To get an estimate of the rate at which VCF can be parsed, we wrote a simple
+program that reads a VCF file and parses it line-by-line using the \texttt{htslib}
+API~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
 above (3.73 GiB) this took 31s, giving a parse rate of 123.1 MiB/s
@@
 \subsection{vcztools}
-[TODO Implementation of vcztools]
+Vcztools is available for download on the Python package index, and
+supports all major platforms [TODO do we support Windows??],
+and is written in a combination of Python and C.
+We use
+the Zarr-Python package~\citep{zarrpython} for all Zarr operations.
+The mini-language in the \texttt{query} command is parsed using
+PyParsing [TODO cite].
+As a general principle, data is only retrieved from the Zarr store
+when it is required to fulfil a particular query. For example,
+when a subset of samples are specified, we only retrieve chunks
+in arrays such as \texttt{call\_genotype} for those
+samples. Chunks in the variants dimension are retrieved as needed,
+and we use a multithreaded strategy to fetch and decode variant chunks
+while another thread encodes VCF. Generating VCF text at speed
+is a challenging problem, and we use a C extension module to convert
+a given set of Zarr chunks represented as numpy arrays to VCF.
+Based on the \texttt{region\_index} array computed by \texttt{vcf2zarr},
+we use PyRanges~\citep{stovner2020pyranges} to perform efficient interval
+search queries.
 

### 2025-01-25 paper.tex
Minor edits

@@
 and the All of Us research program~\citep{all2024genomic}
-collecting genome sequence data for hundreds of thousands of humans.
+collecting genome sequence data for millions of humans.
 Large genetic variation datasets are also being generated for other organisms
@@
 We present the VCF Zarr specification that formalises this
-mapping, and the \texttt{vcf2zarr}
-utility to reliably convert large-scale VCFs to Zarr.
+mapping, the \texttt{vcf2zarr}
+utility to reliably convert large-scale VCFs to Zarr,
+and the \texttt{vcztools} program to form a compatibility layer by
+emulating the ubiquitous \texttt{bcftools}.
 We show that VCF Zarr is much more compact than
@@
 Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
-See~\citep{mcvean2019linkage} for a review of the techniques
-employed in genetic data compression.
+(See~\citep{mcvean2019linkage} for a review of the techniques
+employed in genetic data compression.)
 The widely-used PLINK binary format stores genotypes in a
@@
 storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
-the use of columnar technologies for
+columnar technologies for
 storing and analysing variation data have had limited
@@
 high-entropy QC fields, which is unrealistic
-(see the Genomics England case-study
-for benchmarks on a large human dataset that includes
-many such fields).
+(see the later case-studies
+for benchmarks on datasets that contain a wide variety
+of such fields).
 Note, however, that such minimal, genotype-only data
@@
 of accessing a large number of small chunk files and
-consequent reduced effectiveness of file-system read-ahead caching.
+consequent reduced effectiveness of operating system read-ahead caching.
 While having many independent chunks maps
@@
 
-Another advantage of Zarr is the availabilty of several
+Zarr is a standard with multiple 
 independent implementations across different programming languages
@@
 and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
-The program using the Tensorstore C++ API is fastest, requiring
+The program using the TensorStore C++ API is fastest, requiring
 about 20\% less CPU time than the version using Zarr-Python and Numba

### 2025-01-25 paper.tex
Move S3 GEL stuff to own section

@@
 
-Finally, to explore the use of cloud computing services with VCF Zarr,
-we uploaded the Zarr directory hierarchy for chromosome 2 to an AWS S3
-standard bucket. We repeated the benchmarks above on the same subset
-of 562,640 variants, using the same
-single-threaded sequential Zarr code.
-Extracting the position array required 15s CPU, 23.8s elapsed.
-The af-dist calculation required 2.70min CPU, 5.23min elapsed.
-Creating the filter mask
-required 3.05min CPU, 11.15min elapsed.
-
-While these benchmarks are comparable with results obtained on the
-shared file-system in the Genomics England research environment,
-the sequential single-threaded access patterns used do not map
-well to modern cloud computing environments.
-As each request to a cloud object store has a high latency, it
-is important to have many requests running concurrently to achieve
-good performance~\citep{durner2023exploiting}.
-To explore the limits of performance we developed a simple prototype
-Zarr implementation, designed to maximise concurrency using asynchrous
-IO and multiple processes (see Methods for details).
-Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
-can be decoded to RAM as well as rate at which we can compute
-af-dist (using Numba to classify genotypes, as before)
-as we increase the number of parallel processes on
-an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
-The maximum decompression rate of 47.4GiB/s
-occurs at 64 processes, with a corresponding
-network data rate of 121 MiB/s
-Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
-network access by measuring the rate at which we can download the chunks
-into RAM (without decompressing) using the same parallel and asynchronous
-architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
-and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
-was achieved with 32 processes. Genotype data can therefore be fetched
-from S3 around an order of magnitude more quickly than it can be decompressed
-into RAM. This has some interesting consequences
-when we consider the goal of seeking the highest possible compression levels
-for genotype data and the usual cloud storage cost structuring models (see
-Discussion).
-
-Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
-calculation is performed at a maximum rate of 25.5GiB/s using 128
-processes. At this rate, the af-dist calculation for the target region
-was completed in 4.15s (compare to 5.23min for the naive single-threaded
-method above).
-This level of throughput would complete af-dist for all chromosomes for all samples
-% compute: cost of c5.9xlarge per hour * 1.27
-% ==                             $4.08 * 1.27
-% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
-% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
-in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
-(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
-benchmarks in cloud settings.)
-
 \subsection{Case study: Our Future Health genotype data}
@@
 
+\subsection{Extensibility: Cloud computing}
+To explore the use of cloud computing services with VCF Zarr,
+we uploaded the Zarr directory hierarchy for the Genomics England
+data chromosome 2 described in Table~\ref{tab-genomics-england-data}.
+an AWS S3 standard bucket. 
+[TODO clarify what we're doing and remind reader]
+We repeated the benchmarks above on the same subset
+of 562,640 variants, using the same
+single-threaded sequential Zarr code.
+Extracting the position array required 15s CPU, 23.8s elapsed.
+The af-dist calculation required 2.70min CPU, 5.23min elapsed.
+Creating the filter mask
+required 3.05min CPU, 11.15min elapsed.
+
+While these benchmarks are comparable with results obtained on the
+shared file-system in the Genomics England research environment,
+the sequential single-threaded access patterns used do not map
+well to modern cloud computing environments.
+As each request to a cloud object store has a high latency, it
+is important to have many requests running concurrently to achieve
+good performance~\citep{durner2023exploiting}.
+To explore the limits of performance we developed a simple prototype
+Zarr implementation, designed to maximise concurrency using asynchrous
+IO and multiple processes (see Methods for details).
+Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
+can be decoded to RAM as well as rate at which we can compute
+af-dist (using Numba to classify genotypes, as before)
+as we increase the number of parallel processes on
+an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
+The maximum decompression rate of 47.4GiB/s
+occurs at 64 processes, with a corresponding
+network data rate of 121 MiB/s
+Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
+network access by measuring the rate at which we can download the chunks
+into RAM (without decompressing) using the same parallel and asynchronous
+architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
+and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
+was achieved with 32 processes. Genotype data can therefore be fetched
+from S3 around an order of magnitude more quickly than it can be decompressed
+into RAM. [TODO QUESTION: can such rates be sustained, and be expected to 
+happen in parallel? Do Durner et al consider this?]
+
+% Maybe don't bother with this, the same trade-off always exists.
+% but I guess it's an extra factor to take into account when weighing
+% up practical engineering decisions.
+% This has some interesting consequences
+% when we consider the goal of seeking the highest possible compression levels
+% for genotype data and the usual cloud storage cost structuring models (see
+% Discussion).
+
+Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
+calculation is performed at a maximum rate of 25.5GiB/s using 128
+processes. At this rate, the af-dist calculation for the target region
+was completed in 4.15s (compare to 5.23min for the naive single-threaded
+method above).
+This level of throughput would complete af-dist for all chromosomes for all samples
+% compute: cost of c5.9xlarge per hour * 1.27
+% ==                             $4.08 * 1.27
+% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
+% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
+in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
+(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
+benchmarks in cloud settings.)
+
 \subsection{Extensibility: GPU acceleration}

### 2025-01-25 paper.tex
Update OFH and GEL data

@@
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns produced for
-Genomics England aggV2 VCFs on chromosome 2 using \texttt{vcf2zarr}
-default settings. Each field is stored independently
+\caption{Summary for VCF Zarr conversion of 
+Genomics England WGS data for chromosome 2 (78,195 samples, 59,880,903 variants),
+consisting of 44 fields and 2.54TiB of storage ($\sim$5X compression over
+source gzipped VCF).
+Each field is stored independently
 as a Zarr array with the given type (sufficient to represent all values in the
@@
 \begin{table}
-\caption{Summary of the largest VCF Zarr columns for the Our Future Health
-    genotype dataset, with chunk sizes of 1,000 variants and
-    10,000 samples and applying the BitRound(5) filter to the floating point
-    fields.
+% \caption{Summary for VCF Zarr conversion of 
+% Genomics England WGS data for chromosome 2 (78,195 samples, 59,880,903 variants),
+% consisting of 44 fields and 2.54TiB of storage ($\sim$5X compression over gzipped VCF).
+\caption{Summary for VCF Zarr conversion of 
+Our Future Health genotype data for chromosome 22 (651,050 samples, 10,221
+variants) consisting of 19 fields and 17GiB of storage ($\sim2.5$X compression
+over shouce gzipped VCF).
+% 42.3  / 17 = 2.4882
+Here we use a chunk size of 1,000 variants $\times$ 10,000 samples and 
+apply the BitRound(5) filter to the large floating point fields (see text for 
+details).
 For column details, see Table~\ref{tab-genomics-england-data}.

### 2025-01-25 paper.tex
Updates to Zarr table for AoU

@@
 \begin{table}
-% \caption{Summary for VCF Zarr conversion of 
-% Genomics England WGS data for chromosome 2 (78,195 samples, 59,880,903 variants),
-% consisting of 44 fields and 2.54TiB of storage ($\sim$5X compression over gzipped VCF).
 \caption{Summary for VCF Zarr conversion of 
@@
 variants) consisting of 19 fields and 17GiB of storage ($\sim2.5$X compression
-over shouce gzipped VCF).
+over source gzipped VCF).
 % 42.3  / 17 = 2.4882
@@
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns
-    for the All of Us exome-like data. The chunk size is
-    1,000 variants $\times$ 10,000 samples.
+\caption{Summary for VCF Zarr conversion of 
+All of Us exome-like data genotype data for chromosome 22 (245,394 samples, 
+715,256 variants) 
+consisting of 25 fields and 8.5GiB of storage ($\sim1.1$X larger
+% 8.5 / 7.44 = 1.142
+than source gzipped VCF).
+Here we use the local alleles fields \texttt{call\_LA} and \texttt{call\_LAD};
+see text for details.
 For column details, see Table~\ref{tab-genomics-england-data}.

### 2025-01-25 paper.tex

@@
 % 8.5 / 7.44 = 1.142
-than source gzipped VCF).
+than source gzipped VCF; see text for discussion).
 Here we use the local alleles fields \texttt{call\_LA} and \texttt{call\_LAD}; 
@@
 
+[Discuss why we're bigger than VCF here, people will want to understand.
+
 \subsection{Case study: 1,063 spruce whole-genome samples}
@@
 where resequencing data has been mapped to a chromosome-scale
-reference genome. The dataset consists of approximately 3,745 million
+reference genome. The dataset consists of approximately 3.8 million
 single-nucleotide variants and small indels from the 12 autosomes
@@
 \begin{table}
-\caption{Summary for a selection of the largest VCF Zarr columns
-produced for 1,063 VCFs on 12 spruce chromosomes using
-\texttt{vcf2zarr} with samples chunk size set to 1,063, variants
-chunk size to 10,000, and using local allele fields to reduce output
-size. For column details see the caption of
+\caption{Summary for VCF Zarr conversion of the Norway spruce
+WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
+consisting of 34 fields and 2.3TiB of storage ($\sim 3.2X$ smaller 
+than source gzipped VCFs). The chunk size is 10,000 variants times 
+1,063 samples (i..e, a single sample chunk). This uses local 
+alleles to reduce the 
+space required by the PL field; see the text for discussion.
+For column details see the caption of
 Table~\ref{tab-genomics-england-data}.

### 2025-01-25 paper.tex
Tweaks

@@
 \subsection{Case study: All of Us exome-like data}
-The All of Us dataset comprises 245,388 clinical-grade whole genome sequences (WGS),
+The All of Us dataset comprises 245K clinical-grade whole genome sequences,
 identifying over 1 billion variants, including 3.9 million novel variants
@@
 WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
-consisting of 34 fields and 2.3TiB of storage ($\sim 3.2X$ smaller 
-than source gzipped VCFs). The chunk size is 10,000 variants times 
+consisting of 34 fields and 2.3TiB of storage ($\sim 3.2$X smaller 
+than source gzipped VCFs). The chunk size is 10,000 variants $\times$
 1,063 samples (i..e, a single sample chunk). This uses local 

### 2025-01-25 paper.tex

@@
 \subsection{Extensibility: GPU acceleration}
-[JK: this is a super rough first narrative draft.]
-Machine learning, and in particular deep learning methods, have led to
-major breakthroughs in many different fields in recent years [CITES].
-GPUs have been a key enabling factor in this [details and cites].
-The PyData ecosystem [cites] has also enabled this. Techologies like
-[Cupy, lots of others] substantially lower the bar for writing efficient
-GPU code.
-Deep learning methods depend crucially on efficient access to vast
-datasets for training, and Zarr is emerging as a popular choice
-[cites from geosciences?]. The RAPIDs GPU Accelerated data science
-library [which does] now supports Zarr through [kvikio, more?].
+GPU acceleration has revolutionized scientific computing by leveraging the
+massively parallel architecture of GPUs to achieve significant speedups in
+computationally intensive tasks such as simulations, data analysis, and machine
+learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
+mittal2014survey,owens2008gpu}
+% (Nickolls & Dally, 2010;
+% Mittal & Vetter, 2015; 
+% Owens et al., 2008). 
+This has profound implications for
+fields ranging from molecular dynamics~\citep{stone2007accelerating}
+% (Stone et al., 2010) 
+to deep learning~\citep{lecun2015deep}
+% (LeCun et al., 2015). 
+While early GPU programming required expertise in
+low-level languages like CUDA or OpenCL, the advent of high-level Python
+libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
+%(Okuta et al., 2017), 
+and
+PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
+% (Klöckner, 2012) 
+has democratized access to GPU acceleration.
+These tools offer user-friendly interfaces and just-in-time compilation,
+allowing scientists to harness the power of GPUs within the familiar Python
+ecosystem, thereby significantly accelerating scientific discovery and
+broadening the adoption of accelerated computing across various research
+domains.
+
+% [JK: this is a super rough first narrative draft.]
+% Machine learning, and in particular deep learning methods, have led to
+% major breakthroughs in many different fields in recent years [CITES].
+% GPUs have been a key enabling factor in this [details and cites].
+% The PyData ecosystem [cites] has also enabled this. Techologies like
+% [Cupy, lots of others] substantially lower the bar for writing efficient
+% GPU code.
+% Deep learning methods depend crucially on efficient access to vast
+% datasets for training, and Zarr is emerging as a popular choice
+% [cites from geosciences?]. The RAPIDs GPU Accelerated data science
+% library [which does] now supports Zarr through [kvikio, more?].
 
@@
 data. To do this we reimplemented the af-dist calculation
-described above using CuPy~\citep{cupy}, a GPU accelerated library
+described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
 of core operations for scientific computing.

### 2025-01-26 paper.tex
Tweaks to S3 benchmarks

@@
 we uploaded the Zarr directory hierarchy for the Genomics England
-data chromosome 2 described in Table~\ref{tab-genomics-england-data}.
-an AWS S3 standard bucket. 
-[TODO clarify what we're doing and remind reader]
+data chromosome 2 described in Table~\ref{tab-genomics-england-data}
+to an AWS S3 standard bucket. 
 We repeated the benchmarks above on the same subset
-of 562,640 variants, using the same
+of 562,640 variants and 78,195 samples using the same
 single-threaded sequential Zarr code.
@@
 The af-dist calculation required 2.70min CPU, 5.23min elapsed.
-Creating the filter mask
-required 3.05min CPU, 11.15min elapsed.
+Creating the filter mask required 3.05min CPU, 11.15min elapsed.
 
@@
 and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
-was achieved with 32 processes. Genotype data can therefore be fetched
-from S3 around an order of magnitude more quickly than it can be decompressed
-into RAM. [TODO QUESTION: can such rates be sustained, and be expected to 
-happen in parallel? Do Durner et al consider this?]
+was achieved with 32 processes. 
+% Richard Durbin asked whether we were just using up all of a scarce 
+% resource on this point.
+While these measurments are simplistic and may not be an 
+accurate predictor of sustained bandwidth, it is still clear that 
+genotype data can be fetched
+from S3 around an order of magnitude more quickly 
+than it can be decompressed into RAM. 
+We have also not measured
+the achievable bandwidth while scaling out across many instances,
+but given the ubiquity of cloud-based services and the scale 
+at which they operate, it seems reasonable to assume that the claimed
+terabits per second of aggregate bandwidth from
+S3~\cite{s3-performance-optimising} are achievable.
 
@@
 in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
-(See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
-benchmarks in cloud settings.)
+See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
+benchmarks in cloud settings.
 

### 2025-01-26 paper.tex
Pass at the abstract

@@
 genetic variation data and associated quality control metrics.
-% It provides a well-defined data model and is central to a large ecosystem
-% of interoperating tools.
 The usual row-wise encoding of the VCF data model (either as text
@@
 \textbf{Results:}
+Zarr is a format for storing
+multi-dimensional data that is widely used across the sciences,
+and is ideally suited to massively parallel processing.
 We present the VCF Zarr specification, an encoding of the
-VCF data model using Zarr which makes retrieving subsets of the
-data much more efficient. Zarr is a cloud-native format for storing
-multi-dimensional data, widely used in scientific computing.
+VCF data model using Zarr, along with fundamental software infrastructure
+for efficient and reliable conversion at scale.
 We show how this format is far more efficient than
@@
 storing genotype data in terms of compression ratios
-and calculation performance.
-We demonstrate the VCF Zarr format (and the vcf2zarr conversion utility)
-on a subset of the Genomics England aggV2 dataset comprising
-78,195 samples and 59,880,903 variants,
-with a 5X reduction in storage and greater than 300X reduction in CPU usage
-in some representative benchmarks.
+and single-threaded calculation performance.
+We present case studies on subsets of three large 
+human datasets (Genomics England: $n$=78,195;
+Our Future Health: $n$=651,050; 
+All of Us: $n$=245,394) along with whole genome
+datasets for Norway Spruce ($n$=1,063) and 
+SARS-Cov-2 ($n$=3,960,704) % TODO update if we use tranche 2 data also
+demonstrating the scalability and flexibility of VCF Zarr.
+We demonstrate the potential for VCF Zarr to enable a new generation
+of high-performance and cost-effective applications
+via illustrative highly parallel benchmarks on AWS and GPUs.
 
@@
 For column details, see Table~\ref{tab-genomics-england-data}.
-\label{tab-auo-data}}
+\label{tab-aou-data}}
 \begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
@@
 
-[TODO discuss local alleles and mention Table~\ref{tab-auo-data}].
+[TODO discuss local alleles and mention Table~\ref{tab-aou-data}].
 

### 2025-01-26 paper.tex
Draft of intro

@@
 datasets for Norway Spruce ($n$=1,063) and 
-SARS-Cov-2 ($n$=3,960,704) % TODO update if we use tranche 2 data also
+SARS-Cov-2 ($n$=3,960,704) % TODO update for tranche 2 data 
 demonstrating the scalability and flexibility of VCF Zarr.
@@
 Zarr's elegant simplicity and first-class support for
-cloud object stores have led to
+cloud object stores and parallel computing have led to
 it gaining substantial traction
@@
 \texttt{bcftools} based pipelines, and again competitive
-with state-of-the-art file-oriented methods. Finally, we show the
-utility of VCF Zarr on the Genomics England aggV2 dataset,
-demonstrating that common \texttt{bcftools} queries can be performed orders
-of magnitude more quickly using simple Python scripts.
+with state-of-the-art file-oriented methods. 
+We demonstrate the scalability and flexibility of VCF Zarr
+via case-studies on five boundary-pushing datasets.
+We examine conversion and compression performance on three different
+data modalities for large human datasets: whole genome sequence
+data from Genomics England (5X smaller than gzipped VCF);
+genotype data from Our Future Health (2.5X smaller than gzipped VCF);
+and exome-like data from All of Us (1.1X larger than gzipped VCF). 
+We then examine performance 
+for species with large genomes using Norway Spruce data (3.2X smaller
+than gzipped VCF),
+and in a large collection of short whole genome alignments
+using SARS-Cov-2 data 
+(100X % FIXME! this is roughly correct, but need to get the numbers straight
+smaller than gzipped FASTA). 
+Interoperability with the vast ecosystem of tools
+based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
+is crucial. 
+We first present the \texttt{vcztools} program,
+which implements a subset of the functionality in \texttt{bcftools}
+allowing data stored in Zarr at some URL to be 
+treated as if it were a VCF or BCF file,
+and demonstrate that it can be used as a practical replacement 
+in processing pipelines.
+We then present an illustrative example of how existing
+methods can adopt VCF Zarr by modifying the popular 
+SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
+(written in C++), and show significant speedups over the existing 
+VCF (6.5X) and BCF (3X) backends in a simple benchmark.
 

### 2025-01-26 paper.tex
Pass through the abstract

@@
 We present the VCF Zarr specification that formalises this
-mapping, the \texttt{vcf2zarr}
-utility to reliably convert large-scale VCFs to Zarr,
-and the \texttt{vcztools} program to form a compatibility layer by
-emulating the ubiquitous \texttt{bcftools}.
+mapping, along with the \texttt{vcf2zarr} and \texttt{vcztools} 
+utilities to reliably convert VCF to and from Zarr at scale.
 We show that VCF Zarr is much more compact than
@@
 is crucial. 
+% TODO sentence too long
 We first present the \texttt{vcztools} program,
 which implements a subset of the functionality in \texttt{bcftools}
-allowing data stored in Zarr at some URL to be 
+and allows data stored in Zarr at some URL to be 
 treated as if it were a VCF or BCF file,
@@
 VCF (6.5X) and BCF (3X) backends in a simple benchmark.
+Finally, we illustrate the potential 
+for a new generation of Zarr-based applications via some illustative
+benchmarks. First, we demonstrate the very high levels of throughput 
+performance available on public clouds when data in an object 
+store is accessed in an asynchronous and highly parallel manner.
+Finally, we demonstrate the ease and efficiency with which data
+in Zarr format can be processed on a GPU, opening many exiting
+avenues for future research.
 

### 2025-01-26 paper.tex

@@
 datasets for Norway Spruce ($n$=1,063) and 
-SARS-Cov-2 ($n$=3,960,704) % TODO update for tranche 2 data 
+SARS-Cov-2 ($n$=3,960,704), % TODO update for tranche 2 data 
 demonstrating the scalability and flexibility of VCF Zarr.
@@
 
-\subsection{Extensibility: Cloud computing}
+\subsection{Future applications: Cloud computing}
 To explore the use of cloud computing services with VCF Zarr,
@@
 
-\subsection{Extensibility: GPU acceleration}
+\subsection{Future applications: GPU acceleration}
 GPU acceleration has revolutionized scientific computing by leveraging the
@@
 We propose the VCF Zarr specification which losslessly
-maps VCF data to Zarr, and provide an efficient and scalable
-tool to perform conversion.
+maps VCF data to Zarr, and provide the basic software
+infrastructure required for bidirectional conversion.
 
@@
 for fields like PL in which the dimension depends
-quadratically on the number of alleles (although practical
-solutions have been suggested that we plan to
-implement~\cite{poterba2024scalable}).
+quadratically on the number of alleles.
+While the local alleles representation~\cite{poterba2024scalable}
+is a substantial improvement (see All of Us and Norway Spruce
+case studies) it may be lossy (see Methods). 
 Secondly, the design of
@@
 
-[TODO REWRITE]
-The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
-provided here are a necessary starting point for such cloud-native
-biobank repositories and open up many possibilities,
-but significant investment and development would be needed
-to provide a viable alternative to standard bioinformatics workflows.
-Two initial directions for development, however, may quickly
-yield sufficient results to both greatly improve researcher productivity on
-large, centrally managed datasets such as Genomics England
-and motivate further research and development.
-The first direction is to provide compatibility with existing
-workflows via a ``\texttt{vcztools}'' command line utility which
-implements a subset of \texttt{bcftools} functionality (such as
-\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
-Such a tool would speed up some common queries by orders
-of magnitude, and reduce the need for user orchestration of
-operations among manually split VCF chunks (large VCF datasets
-are typically split into hundreds of files; see the Genomics
-England case study). Datasets could then be hosted in cloud
-object stores, while still presenting file-like semantics
-for existing workflows. This could provide an evolutionary path,
-allowing established analysis workflows to coexist with new Zarr-native
-approaches, working from the same primary data.
-
-[TODO REWRITE]
-The second natural direction for development is to create
-these Zarr-native applications, which can take advantage
-of the efficient data representation
-across multiple programming languages (see Methods).
-The Python data science ecosystem, in particular,
-has a rich suite of powerful tools~\cite[e.g.][]{
+% Streaming is inherently incompatible with efficient utilisation
+% of cloud object stores.
+Modern computing is increasingly parallel, from multicore CPUs, GPUs 
+and the massively parallel scale provided by public clouds.
+Making optimal use of these parallel resources requires new programming
+models, and in particular adapting analytics workloads to the properties of 
+disaggregated cloud object stores is essential.
+Classical workflows in genomics emphasise streaming, 
+in which large volumes of data are read sequentially from file,
+streamed through the processor, and the results written sequentially to 
+storage or to the input of another process in a pipeline. 
+Using this model on data held on a cloud object store requires
+us to emulate file-system semantics,
+and places hard requirements on receiving data in a specific order.
+As Durner et al~\cite{durner2023exploiting} show (and we confirm 
+in Fig~\ref{fig-s3-network-throughput}), sustained data throughput speeds similar to
+a locally attached SSD can be obtained from cost-effective and 
+massively scalable cloud object stores like Amazon S3,
+but on the condition that we do \emph{not} place constraints on
+the order in which data is received (due to the high variance in latency).
+In this context, Zarr provides a simple and elegant model for computation
+in which the \emph{chunk} is the basic unit of parallelism. For many
+applications, each chunk can be processed independently and in no particular order.
+
+% In the streaming model, large volumes of data are read sequentially from storage,
+% streamed through the processor, with results written sequentially to 
+% storage or read by other processes in a pipeline. 
+% Row-oriented storage of data (see the Introduction) is also well-suited to 
+% streaming computation
+% % The focus on large sequential 
+% % reads is derived ultimately from the realities of disk drive storage, and leads
+% % to efficient utilisation through multiple layers of operating system and 
+% % on-device caching, with heuristic parameters tuned through decades of experience.
+% % While there is some potential for parallelism in chaining multiple processes
+% % in a Unix pipeline, it tends to be limited by the slowest step and is not
+% % an effective way to use the tens of CPU cores available on a modern machines.
+% Splitting work up along the genome has provided ample opportunities to 
+% to parallelise in the past, and is part of the motivation 
+% for the current practise of providing VCF data in sub-chromosome partitions 
+% (see the Genomics England, Our Future Health and Norway Spruce case-studies). 
+% However, with increasingly large sample sizes, this approach to parallelism
+
+There is therefore an opportunity to develop a new 
+generation of Zarr-native applications, taking advantage of the efficient data
+representation and natural model of chunk-based parallelism.
+Although there are several independent Zarr implementations
+across programming languages (see Methods),
+the Python data science ecosystem is a particularly powerful
+platform,
+with a rich suite of tools~\cite[e.g.][]{
 mckinney2010data,
@@
 illustrate that working at the lowest level,
-by sequentially applying optimised kernels on a chunk-by-chunk
+by applying optimised kernels on a chunk-by-chunk
 basis is both straightforward to implement and highly performant.
@@
 PyParsing [TODO cite].
+[Add a note here on why it was easier to to write a Python application
+from scratch than to update bcftools.]
+
 As a general principle, data is only retrieved from the Zarr store

### 2025-01-26 reviewed-paper.tex
Add reviewed version for easy diffing

--- 2025-01-26 paper.tex
@@
 \author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
-\author[4,\authfn{1}]{Will Tyler}
-\author[5,\authfn{1}]{Tom White}
-
-% Middle, alphabetical order
-\author[6]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
-\author[7,8]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
-\author[9]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
-\author[10]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[7,8,11]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
-\author[12]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[13]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
-\author[14]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+\author[4,\authfn{1}]{Tom White}
+
+% Middle
+\author[5]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[6]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[7]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[15]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[8]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
@@
 \author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[5,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 \affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
-\affil[4]{Independent researcher}
-\affil[5]{Tom White Consulting Ltd.}
-\affil[6]{Our Future Health, Manchester, UK.}
-\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
-Hospital, Boston, Massachusetts 02114, USA}
-\affil[9]{NVIDIA Ltd, Reading, UK}
-\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
+\affil[4]{Tom White Consulting Ltd.}
+\affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
 University of Oxford, UK}
-\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[12]{Wellcome Sanger Institute}
-\affil[13]{Genomics England}
-\affil[14]{Department of Cell and Molecular Biology, National
-  Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
-  Uppsala University, Uppsala, Sweden}
-\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[6]{Wellcome Sanger Institute}
+\affil[7]{Genomics England}
+\affil[8]{School of Computer Science, McGill University, Montreal, QC, Canada}
 
@@
 genetic variation data and associated quality control metrics.
+% It provides a well-defined data model and is central to a large ecosystem
+% of interoperating tools.
 The usual row-wise encoding of the VCF data model (either as text
@@
 \textbf{Results:}
-Zarr is a format for storing
-multi-dimensional data that is widely used across the sciences,
-and is ideally suited to massively parallel processing.
 We present the VCF Zarr specification, an encoding of the 
-VCF data model using Zarr, along with fundamental software infrastructure
-for efficient and reliable conversion at scale.
+VCF data model using Zarr which makes retrieving subsets of the 
+data much more efficient. Zarr is a cloud-native format for storing 
+multi-dimensional data, widely used in scientific computing.
 We show how this format is far more efficient than
@@
 storing genotype data in terms of compression ratios
-and single-threaded calculation performance.
-We present case studies on subsets of three large 
-human datasets (Genomics England: $n$=78,195;
-Our Future Health: $n$=651,050; 
-All of Us: $n$=245,394) along with whole genome
-datasets for Norway Spruce ($n$=1,063) and 
-SARS-Cov-2 ($n$=3,960,704) % TODO update for tranche 2 data 
-demonstrating the scalability and flexibility of VCF Zarr.
-We demonstrate the potential for VCF Zarr to enable a new generation
-of high-performance and cost-effective applications
-via illustrative highly parallel benchmarks on AWS and GPUs.
+and calculation performance.
+We demonstrate the VCF Zarr format (and the vcf2zarr conversion utility) 
+on a subset of the Genomics England aggV2 dataset comprising
+78,195 samples and 59,880,903 variants,
+with a 5X reduction in storage and greater than 300X reduction in CPU usage
+in some representative benchmarks.
 
@@
 UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
-Our Future Health~\citep{ofhmain_2025},
 and the All of Us research program~\citep{all2024genomic}
-collecting genome sequence data for millions of humans.
+collecting genome sequence data for hundreds of thousands of humans.
 Large genetic variation datasets are also being generated for other organisms
@@
 Zarr's elegant simplicity and first-class support for 
-cloud object stores and parallel computing have led to
+cloud object stores have led to 
 it gaining substantial traction
@@
 We present the VCF Zarr specification that formalises this 
-mapping, the \texttt{vcf2zarr}
-utility to reliably convert large-scale VCFs to Zarr,
-and the \texttt{vcztools} program to form a compatibility layer by
-emulating the ubiquitous \texttt{bcftools}.
+mapping, and the \texttt{vcf2zarr} 
+utility to reliably convert large-scale VCFs to Zarr.
 We show that VCF Zarr is much more compact than 
@@
 \texttt{bcftools} based pipelines, and again competitive
-with state-of-the-art file-oriented methods. 
-We demonstrate the scalability and flexibility of VCF Zarr
-via case-studies on five boundary-pushing datasets.
-We examine conversion and compression performance on three different
-data modalities for large human datasets: whole genome sequence
-data from Genomics England (5X smaller than gzipped VCF);
-genotype data from Our Future Health (2.5X smaller than gzipped VCF);
-and exome-like data from All of Us (1.1X larger than gzipped VCF). 
-We then examine performance 
-for species with large genomes using Norway Spruce data (3.2X smaller
-than gzipped VCF),
-and in a large collection of short whole genome alignments
-using SARS-Cov-2 data 
-(100X % FIXME! this is roughly correct, but need to get the numbers straight
-smaller than gzipped FASTA). 
-Interoperability with the vast ecosystem of tools
-based on VCF~\citep{danecek2021twelve,garrison2022spectrum}
-is crucial. 
-We first present the \texttt{vcztools} program,
-which implements a subset of the functionality in \texttt{bcftools}
-allowing data stored in Zarr at some URL to be 
-treated as if it were a VCF or BCF file,
-and demonstrate that it can be used as a practical replacement 
-in processing pipelines.
-We then present an illustrative example of how existing
-methods can adopt VCF Zarr by modifying the popular 
-SAIGE~\cite{zhou2018efficiently, zhou2020scalable} GWAS software
-(written in C++), and show significant speedups over the existing 
-VCF (6.5X) and BCF (3X) backends in a simple benchmark.
+with state-of-the-art file-oriented methods. Finally, we show the 
+utility of VCF Zarr on the Genomics England aggV2 dataset,
+demonstrating that common \texttt{bcftools} queries can be performed orders
+of magnitude more quickly using simple Python scripts.
 
@@
 Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
-(See~\citep{mcvean2019linkage} for a review of the techniques
-employed in genetic data compression.)
+See~\citep{mcvean2019linkage} for a review of the techniques
+employed in genetic data compression.
 The widely-used PLINK binary format stores genotypes in a 
@@
 storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
-columnar technologies for
+the use of columnar technologies for
 storing and analysing variation data have had limited
@@
 by subsets of a large simulation of French-Canadians.
-Sizes for $10^6$ samples are shown on the right.
+Sizes for $10^6$ samples are shown on the right. Sizes 
+for Savvy (21.25GiB) and Zarr (22.06GiB) are very similar.
 Also shown for reference is the size of genotype matrix 
 when encoded as two bits per diploid genotype (2bit), as used 
-in the PLINK binary format. Default compression settings were
-used in all cases.
+in the PLINK binary format.
 \label{fig-data-storage}}
@@
 arrays and chunks are stored hierarchically in directories
-and files (storage as a single Zip archive is also supported,
-and has certain advantages as discussed later).
+and files (storage as a single Zip archive is also supported).
 To enable comparison with the existing file-based ecosystem
 of tools, we focus on Zarr's file system chunk storage in a series of illustrative 
-benchmarks in the following sections before presenting cloud-based benchmarks.
+benchmarks in the following sections. 
+(See \citep{durbin2020task,moore2021ome,gowan2022using} for Zarr
+benchmarks in cloud settings.)
 We compare primarily with
@@
 high-entropy QC fields, which is unrealistic 
-(see the later case-studies
-for benchmarks on datasets that contain a wide variety
-of such fields).
+(see the Genomics England case-study
+for benchmarks on a large human dataset that includes 
+many such fields). 
 Note, however, that such minimal, genotype-only data 
@@
 % savvy 6.6 GiB
-% zarr_nshf 2.7 GiB
-% zarr_lz4_nshf 2.9 GiB
+% zarr_nshf 3.9 GiB
 (up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
 Fig~\ref{fig-whole-matrix-decode}).
-% sav / zarr ratio: [0.68777206 0.75870945 0.69865406 0.64644888 0.68131752
-% 0.89460906]
 % sequence_length  num_samples  num_sites       tool       size
-% 58         48129895      1000000    7254858                zarr    23.752762
-% 50         48129895      1000000    7254858         10kx1k.zarr    22.068925
-% 51         48129895      1000000    7254858                 bcf    51.749294
-% 52         48129895      1000000    7254858             genozip    10.691384
-% 53         48129895      1000000    7254858  lz4.noshuffle.zarr   214.868005
-% 54         48129895      1000000    7254858      noshuffle.zarr    31.208149
-% 55         48129895      1000000    7254858                 sav    21.249436
-% 56         48129895      1000000    7254858                 tsk     1.802636
-% 57         48129895      1000000    7254858                 vcf    81.375831
-% 59         48129895      1000000    7254858            zarr.zip    24.240338
-% 5          48129895      1000000    7254858                2bit  1689.153258
+% 40         48129895      1000000    7254858       zarr  22.071687
+% 35         48129895      1000000    7254858        bcf  51.749294
+% 36         48129895      1000000    7254858    genozip  10.691384
+% 37         48129895      1000000    7254858        sav  21.249436
+% 38         48129895      1000000    7254858        tsk   1.802636
+% 39         48129895      1000000    7254858        vcf  81.375831
+% 41         48129895      1000000    7254858  zarr_nshf  29.897540
 Turning off the BitShuffle filter for the Zarr dataset,
 however, leads to a substantial increase in decoding speed
-(2.7GiB/s) at the cost of a roughly 25\% increase in storage
-space (31.2GiB up from 23.75GiB for 1 million samples; data not
-shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
-decoding speed (2.9GiB/s), but at the cost of much larger storage requirements
-(214GiB; data not shown).
-Given the relatively small contribution of genotypes to the
+(3.9GiB/s) at the cost of a roughly 25\% increase in storage
+space (29.9GiB up from 22.1GiB for 1 million samples; data not
+shown). Given the relatively small contribution of genotypes to the
 overall storage of real datasets (see the Genomics England example)
-and the frequency that they are likely to be accessed, using
-Zstandard without BitShuffle
-would seem like a good tradeoff in many cases.
+and the frequency that they are likely to be accessed, this
+would seem like a good tradeoff in most cases.
 This ability to easily tune compression performance
@@
 functionality to aid with such storage schema tuning.
-
-The discrepancy between elapsed time and total CPU time for Zarr
-benchmarks in
-Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
-and~\ref{fig-whole-matrix-decode} is mostly due to the overhead
-of accessing a large number of small chunk files and
-consequent reduced effectiveness of operating system read-ahead caching.
-While having many independent chunks maps
-naturally to high-performance parallel access patterns
-in the cloud store setting~\citep{durner2023exploiting}
-(see also the section on accessing Genomics England data stored on S3),
-such file access patterns can be problematic in the
-classical file-system setting.
-The Zarr Zip store (which is simply a Zip archive of the standard directory
-hierarchy) provides a useful alternative here and can
-significantly reduce overall processing time. To illustrate this
-we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
-which we measure the time required to decode the entire genotype
-matrix (thus serving as a lower-bound for any subsequent computations).
-In the case of $10^6$ samples, the Zip file requires slightly more
-% Full data table above ^
-% 58         48129895      1000000    7254858                zarr    23.752762
-% 59         48129895      1000000    7254858            zarr.zip    24.240338
-storage space than the directory hierarchy (23.75GiB and 24.24G
-for file system and Zip file, respectively)
-and while total
-CPU time is similar (193m vs 187m), the elapsed wall clock
-time for the Zip file version is significantly less
-(258m vs 188m).
-        % tool  user_time  sys_time    cpu_time   wall_time
-% 11      zarr   11383.82    191.01  192.913833  258.598778
-% 29  zarr.zip   11198.37     43.13  187.358333  188.659252
-
-Zarr is a standard with multiple 
-independent implementations across different programming languages
-(see Methods). Fig~\ref{fig-whole-matrix-compute-zarr-versions}
-compares the CPU time required to perform the computations
-of Fig~\ref{fig-whole-matrix-compute}
-for af-dist implementations using the Zarr-Python~\cite{zarrpython},
-JZarr~\cite{jzarr},
-and TensorStore~\citep{tensorstore} (C++ and Python) libraries.
-The program using the TensorStore C++ API is fastest, requiring
-about 20\% less CPU time than the version using Zarr-Python and Numba
-used in our other benchmarks.
-
-% num_samples  num_sites       tool  user_time  sys_time     wall_time
-% storage  total_time
-% 22      1000000    7254858       zarr   16709.55     38.14  16910.057917 hdd    4.652136
-% 20      1000000    7254858     ts_cpp   13233.50     92.81  13573.456810 hdd    3.701753
-% 21      1000000    7254858  zarr_java   43780.50    245.28  44134.462377 hdd   12.229383
-% 23      1000000    7254858      ts_py   18027.43    642.40  19163.174182 hdd    5.186064
-% Cpp / Zarr python [0.79197226]
 
@@
 the overall dataset size, with the computation requiring around
-2 seconds for 1 million samples. Because of their
+1 second for 1 million samples. Because of their 
 row-wise encoding, CPU time scales with the number of samples
@@
 the time required to populate a NumPy array with the data in Zarr,
-which is 2 seconds. Wall-clock time (dotted line) is dominated
-in this case by file I/O. Note that our benchmarking methodology of
-dropping the file system caches before each run (see Methods) is
-likely not reflective of real world usage where a small and
-frequently used field such as POS would remain in cache.
-Time to output text for Savvy is not significant
+which is less than a second. Wall-clock time (dotted line) is dominated 
+in this case by file I/O. Time to output text for Savvy is not significant
 for $> 1000$ samples (not shown).
@@
 total of 2.54 TiB of storage over 6,312,488
-chunk files (using a chunk size of $10^4$ variants $\times 10^3$
-samples). This is a roughly 5X reduction in total storage
+chunk files. This is a roughly 5X reduction in total storage 
 space over the original VCF. 
@@
 \begin{table}
-\caption{Summary for VCF Zarr conversion of 
-Genomics England WGS data for chromosome 2 (78,195 samples, 59,880,903 variants),
-consisting of 44 fields and 2.54TiB of storage ($\sim$5X compression over
-source gzipped VCF).
-Each field is stored independently
+\caption{Summary for a selection of the largest VCF Zarr columns produced for 
+Genomics England aggV2 VCFs on chromosome 2 using \texttt{vcf2zarr}
+default settings. Each field is stored independently 
 as a Zarr array with the given type (sufficient to represent all values in the
@@
 
-Additionally, to illustrate performance on a common filtering task, we
+Finally, to illustrate performance on a common filtering task, we 
 created a copy of the VCF chunk which contains only variants 
@@
 
-\subsection{Case study: Our Future Health genotype data}
-Our Future Health~\cite{ofhmain_2025}  is the UK’s largest ever health research programme.
-It is designed to help people live healthier lives for longer through the
-discovery and testing of more effective approaches to prevention, earlier
-detection and treatment of diseases.
-Our Future Health is inviting millions of people, from all backgrounds and
-from right across the UK to take part. Volunteers are providing information
-about their health and lifestyles to create an incredibly detailed picture
-that represents the whole of the UK.
-
-\begin{table}
-\caption{Summary for VCF Zarr conversion of 
-Our Future Health genotype data for chromosome 22 (651,050 samples, 10,221
-variants) consisting of 19 fields and 17GiB of storage ($\sim2.5$X compression
-over source gzipped VCF).
-% 42.3  / 17 = 2.4882
-Here we use a chunk size of 1,000 variants $\times$ 10,000 samples and 
-apply the BitRound(5) filter to the large floating point fields (see text for 
-details).
-For column details, see Table~\ref{tab-genomics-england-data}.
-    \label{tab-ofh-data}}
-\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
-\toprule
-{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
-\midrule
-/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
-/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
-/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
-/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
-/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
-/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
-/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
-/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
-/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
-/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
-/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
-/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
-/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
-/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
-/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
-/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
-/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Release 9~\citep{ofhdatarelease9}
-was made available in December 2024. The genetic component includes genotype data
-from 651,050 individuals on 707,522 variants. Here we use the genotype data from
-chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
-three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling
-% 20G     ofh.chr22-b0.vcf.gz
-% 19G     ofh.chr22-b1.vcf.gz
-% 3.3G    ofh.chr22-b2.vcf.gz
-% ==42.3 GiB - assuming the G is power-of-two here
-10,221 variants and 42.3GiB of storage. We first converted the VCF
-data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
-The ICF representation required a total of 41GiB over XX files.
-We then converted to VCF Zarr format using the default schema, which resulted
-in a 41GiB VCZ. On inspection, this was dominated by three floating point
-fields LRR, BAF and GS stored
-to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
-places for GS).
-On inspecting the histograms for these fields
-(Fig~\ref{fig-ofh-field-distributions}) it seemed that such high precision
-was likely unneccessary, and we investigated applying the Quantize
-and Bitround filters from \texttt{numcodecs}~\citep{numcodecs}
-to reduce the precision in stored floating point data,
-and therefore improve compression.
-Using the Quantize filter with 5 digits of precision on these three fields reduced the
-overall storage space to 31G with a mean absolute error (MAE, i.e., the average
-difference between the truncated and original values) of
-% 3   call_LRR    quantize    1.91e-06
-% 4   call_BAF    quantize    1.18e-06
-% 5   call_GS quantize    1.88e-06
-$1.91\times10^{-6}$ (LRR),
-$1.18\times10^{-6}$ (BAF)  and
-$1.88\times10^{-6}$ (GS).
-Using the Bitround filter with a parameter of 5 bits reduced the overall storage for the Zarr to
-XX with an MAE of
-% 0   call_LRR    bitround    6.93e-04
-% 1   call_BAF    bitround    7.86e-04
-% 2   call_GS bitround    3.38e-03
-$6.93\times10^{-4}$ (LRR),
-$7.86\times10^{-4}$ (BAF)  and
-$3.38\times10^{-3}$ (GS).
-The histograms of the truncated fields in all cases were visually indistinguishable
-from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
-Table~\ref{tab-ofh-data} shows a summary of the Zarr using the BitRound filter.
-
-\subsection{Case study: All of Us exome-like data}
-The All of Us dataset comprises 245K clinical-grade whole genome sequences,
-identifying over 1 billion variants, including 3.9 million novel variants
-with coding consequences~\citep{all2024genomic}. Participants are from the United States,
-with an emphasis on historically underrepresented genetic ancestry groups.
-The genomic data, generated through short-read sequencing on the Illumina NovaSeq 6000,
-is of high quality, with an average coverage depth of 30x, and is linked to extensive
-health-related information, providing a valuable resource for biomedical research.
-WGS analysis is conducted on the Illumina DRAGEN platform, with alignment performed
-using the GRCh38dh reference genome. GTC files are converted to VCF format using
-Picard GtcToVcf, and every step of the WGS procedure is rigorously controlled by
-predefined quality control measures.
-
-For this study, we converted the chromosome 20 exome-like
-data from the All of Us dataset to VCF Zarr.
-The exome-like data is constructed by extracting from all WGS samples
-the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic
-transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB,
-includes 715,256 variants, of which more than 700 are classified as pathogenic or likely pathogenic.
-
-Conversion was performed on the All of Us Researcher Workbench platform
-(based on Google Cloud), using a single compute node with 64 CPUs and 416 GiB
-of RAM, which cost \$3.80/hour. We first converted the VCF to  ICF, which
-took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU
-time, achieving a parallel efficiency of 86\%. This step incurred a cost of
-\$23.88. The resulting ICF representation required 7.5 GiB of storage distributed
-across 91,232 files. Next, the ICF was converted to Zarr, which took 1 hour and 8 minutes, with
-a CPU time of 68 hours and 8 minutes and a parallel efficiency of 94\%, at a
-cost of \$4.31. This second Zarr representation occupied 8.5 GiB of storage across
-232,809 chunk files.
-
-\begin{table}
-\caption{Summary for VCF Zarr conversion of 
-All of Us exome-like data genotype data for chromosome 22 (245,394 samples, 
-715,256 variants) 
-consisting of 25 fields and 8.5GiB of storage ($\sim1.1$X larger
-% 8.5 / 7.44 = 1.142
-than source gzipped VCF; see text for discussion).
-Here we use the local alleles fields \texttt{call\_LA} and \texttt{call\_LAD}; 
-see text for details.
-For column details, see Table~\ref{tab-genomics-england-data}.
-\label{tab-aou-data}}
-\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
-\toprule
-{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
-\midrule
-/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
-/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
-/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
-/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
-/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
-/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
-/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-[TODO discuss local alleles and mention Table~\ref{tab-aou-data}].
-
-[Discuss why we're bigger than VCF here, people will want to understand.
-
-\subsection{Case study: 1,063 spruce whole-genome samples}
-
-To demonstrate the versatility of \texttt{vcf2zarr}, in this section
-we include a case study of a dataset originating from a species whose
-genome properties differ substantially from human. The conifer Norway
-spruce, \emph{Picea abies}, is one of the largest and most
-ecologically important species on Earth and is distributed over large
-parts of the Northern hemisphere. One of its features is its large
-genome, consisting of 12 autosomes with a genome size of 19.6
-Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
-performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
-resequenced spruce individuals from a recent study (in preparation),
-where resequencing data has been mapped to a chromosome-scale
-reference genome. The dataset consists of approximately 3.8 million
-single-nucleotide variants and small indels from the 12 autosomes
-totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
-distributed over 165 VCF files. Since the dataset originally was
-generated for downstream analyses that require genotype likelihoods,
-the genotypes include the PL field.
-
-Following the conversion process detailed in the previous section, we
-first converted the 165 VCF files (7.33TiB) to ICF, followed by
-conversion to Zarr. The tasks were run as single jobs on a compute
-node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
-required 13h45min runtime. The ICF representation used a total of 6.77
-TiB over 1,340,804 data storage files. Conversion to Zarr required
-18h36min runtime, generating a dataset with 32 arrays, consuming a
-total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
-total storage (1.1X) is negligible, mainly due to the inherent
-difficulty in compressing the PL field. The output dataset storage
-size is dominated by a major component, call\_PL, which accounts for
-nearly 93\% of the total. call\_PL is yet another field with low
-compression potential due to inherent noisiness.
-
-Therefore, to reduce storage requirements we generated a Zarr schema
-for encoding that uses local allele fields with \texttt{vcf2zarr
-mkschema --local-alleles}, and used the generated schema as input to
-\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
-generating a dataset with 35 arrays, consuming a total of 2.3TiB
-storage over 11,235,556 chunk files, with a reduction in storage of
-3.2X. Therefore, by using local allele fields we achieve approximately
-3X improvement of compression. The top fields in terms of storage are
-detailed inTable~\ref{tab-spruce-data}.
-
-\begin{table}
-\caption{Summary for VCF Zarr conversion of the Norway spruce
-WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
-consisting of 34 fields and 2.3TiB of storage ($\sim 3.2$X smaller 
-than source gzipped VCFs). The chunk size is 10,000 variants $\times$
-1,063 samples (i..e, a single sample chunk). This uses local 
-alleles to reduce the 
-space required by the PL field; see the text for discussion.
-For column details see the caption of
-Table~\ref{tab-genomics-england-data}.
-\label{tab-spruce-data}}
-
-\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
-\toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
-\midrule
-/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
-/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
-/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
-/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
-/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
-/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
-/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
-/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
-/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
-/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
-/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
-/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
-/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
-/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
-/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
-/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
-/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
-/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
-/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
-/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
-/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
-% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
-% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
-% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
-% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
-% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
-% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
-% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
-% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
-% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
-% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Large genomes from non-model organisms pose particular problems for
-analysis workflows that often are tailored for human conditions. For
-instance, the binary Sequence Alignment/Map index format
-BAI~\cite{li2009sequence} has a 512
-Mbp limit, but spruce chromosomes range in size from 1.0 to 1.7 Gbp.
-The spruce BAM files used for variant calling were BAI indexed, which
-motivated the partitioning into multiple VCF files in the first place,
-adding complexity to downstream processing.
-
-A similar limitation is imposed by the VCF POS field which is encoded
-in 32-bit format and can hold chromomosomes up to 2.14 Gbp. Although
-spruce does not have chromosomes that exceed the limit, other species
-do, such as the 87.2 Gbp South American lungfish, whose 19 chromosomes
-are all but one larger than the 3.1 Gbp human
-genome~\cite{schartl_GenomesAllLungfish_2024}. Zarr's flexible type
-system means that coordinates can be stored as 64 or 32 bit integers
-as needed, and there is therefore no particular limit on genome size.
-% limitation, and the coordinate system could easily be converted to
-% true genome coordinates for storage in Zarr. This updating procedure
-% could be applied to any species of choice.
-
-\subsection{Case study: SARS-Cov-2 alignment data}
-The SARS-CoV-2 pandemic resulted in viral sequence data collection at
-an unprecedented scale, with over 15 million whole genomes incorporated into
-one resource as of 2024~\citep{hinrichs2024ongoing}. Data at such scale
-overwhelmed classical phylogenetic methods, leading to the development
-of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}.
-The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled
-set of consensus sequences for 4,471,579 samples, addressing a range of systematic
-errors in earlier collections. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
-input format, and Zhan et al have made the Viridian dataset,
-aligned to the reference using MAFFT~\citep{katoh2013mafft},
-available for download~\citep{zhan2023towards}. This dataset has quite
-different properties to those examined in previous case studies, and
-illustrates the flexibility of Zarr across a range of biological applications.
-
-One of the unusual properties of this dataset is that whole genome alignments
-are stored, such that information on all 29903 bases is stored for every
-sample. Given the number of samples and the fact that nearly all bases
-[TODO be concrete?] contain variation it is simply more convenient to
-store the entire alignment for each sample than to manage the complexity
-of subsetting the sites. Similarly, the usual convention in VCF Zarr
-is that the \texttt{variant\_alleles} field contains the REF and ALT
-alleles as observed (such that the zeroth allele is always the reference).
-In this case it is more convenient to encode the alleles as the original
-bases and IUPAC uncertainty codes, such that the allele 0 is always A, 1 is C etc.
-Even in this case with very high sequence diversity, uncertainty codes and
-storing data for the entire genome, compression rates are very high.
-
-Based on the batch 1 release of Viridian, the Zarr dataset contains
-3,960,704 sequences of length 29903 base pairs along with 31 metadata columns
-encoding information about those samples. This full dataset is stored in a
-single 345MiB Zip file, 241MiB of which is used by the alignment data. This
-gives an average compression ratio of 468X.
-
-Computational accessibility is excellent. For example, computing the number of
-samples with missing or ambiguous data at each site requires around 3 minutes,
-iterating over sample chunks sequentially (5 lines of Python).
-Retrieving the full alignment for a given sample requires 163ms and
-getting the calls for all samples at a given site 209ms. [TODO confirm hardware
-details]. The storage of metadata and genetic data in a single file also
-makes many queries efficient and convenient. For example, computing the
-average number of deletions in samples with a Scorpio designation of Alpha
-required x seconds and 5 lines of code using Dask.
-
-\subsection{Interoperability: vcztools}
-To provide compatibility with existing workflows we have
-developed the \texttt{vcztools} command line utility which
-implements a subset of \texttt{bcftools} functionality.
-Vcztools is intended to be a drop-in replacement for \texttt{bcftools},
-providing an evolutionary path for VCF Zarr adoption,
-with compatibility being a core objective.
-Vcztools currently supports Zarr stored in the local file system
-and Zip archives, and support for major cloud stores will be added in the near
-future. [JK: note this is a pretty simple feature, maybe we should just try
-to add it?]
-Thus, users can convert their VCF data to Zarr using \texttt{bio2zarr},
-store it in a local file system, Zip file or cloud store as desired
-and then access this data source within existing processing pipelines,
-essentially replacing \texttt{bcftools view <vcf file>} with
-\texttt{vcztools view <Zarr URL>}.
-This allows established analysis workflows to coexist with new Zarr-native
-approaches, working from the same primary data.
-
-The initial version of \texttt{vcztools} implements a subset of the \texttt{query}
-and \texttt{view} commands. We provide an example based on the simulated
-dataset of $10^5$ samples of Fig~\ref{fig-data-storage} to illustrate
-the functionality and
-performance of \texttt{vcztools}. First, we extract the position column
-using the \texttt{query} command. Using \texttt{bcftools}
-(\texttt{bcftools query -f '\%POS\textbackslash n' chr21\_10\_5.bcf})
-it took 8m57s to extract the position values for the 2,365,367
-variants from the BCF file.
-Using vcztools
-(\texttt{vcztools query -f '\%POS\textbackslash n' chr21\_10\_5.vcz})
-required 51s (note that this operation could be performed much
-more quickly and may be worth optimising for within \texttt{vcztools}, see
-Fig~\ref{fig-column-extract}). This illustrates the general principle
-that \texttt{vcztools} only retrieves information necessary to fulfil a particular
-query, which can have a significant impact with large datasets.
-
-To form a practical replacement for bcftools in production pipelines
-it is important that \texttt{vcztools} can produce VCF text at a reasonable
-rate. To do this we measure the time taken to generate VCF text for the
-last 10,000 variants in the dataset (using the previously extracted
-position column to determine the coordinates), \texttt{vcztools view
-chr21\_10\_5.vcz -r1:47958263-}). This took 30s, at a rate of 124.3 MiB
-of VCF text per second. The bcftools command on the corresponding
-BCF file took 25s (152.64 MiB/s) and on the VCF file it took 63s (60.57 MiB/s).
-
-% TODO texttt
-Vcztools is primarily intended to used as a replacement for bcftools
-in existing workflows based on pipes, where VCF text is sequentially
-filtered and processed by different processes in the classical
-Unix manner. In this context the most important property is that vcztools
-can produce VCF text as quickly as other processes in the pipeline can consume
-it. To get an estimate of the rate at which VCF can be parsed, we wrote a simple
-program that reads a VCF file and parses it line-by-line using the \texttt{htslib}
-API~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
-above (3.73 GiB) this took 31s, giving a parse rate of 123.1 MiB/s
-(almost identical to the write-rate of vcztools above). Given this limiting
-factor on overall throughput,
-it is unlikely that \texttt{vcztools view} would be a bottleneck
-in practical processing pipelines.
-
-\subsection{Interoperability: SAIGE proof-of-concept}
-To demonstrate the potential for integrating VCF Zarr into existing utilities,
-we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
-zhou2020scalable}, a popular package for performing genome wide
-association tests. Using the Tensorstore C++ library~\cite{tensorstore}
-we added support for VCF Zarr via an additional backend class [some details
-about the implementation].
-
-To test the performance of this implemenentation we compared the time required
-to perform an association test using the simulated dataset of 10,000 samples
-from Fig~\ref{fig-data-storage}. We generated a phenotype using
-\texttt{tstrait}~\citep{tagami2024tstrait}.
-
-We compared the same analyis using the existing VCF, BCF and Savvy backends
-to our prototype Zarr backend.
-[Times: BCF 3m, VCF6.5m, Savvy 39s, Zarr 55s]
-
-% BCF:
-% real 3m1.068s user 2m56.944s sys 0m2.941s
-
-% VCF:
-% real 6m30.335s user 6m25.193s sys 0m2.842s
-
-% Savvy:
-% real 0m39.660s user 0m37.498s sys 0m2.183s
-
-% Zarr:
-% real 0m55.957s user 1m10.375s sys 0m12.111s
-
-\subsection{Extensibility: Cloud computing}
-To explore the use of cloud computing services with VCF Zarr,
-we uploaded the Zarr directory hierarchy for the Genomics England
-data chromosome 2 described in Table~\ref{tab-genomics-england-data}
-to an AWS S3 standard bucket. 
-We repeated the benchmarks above on the same subset
-of 562,640 variants and 78,195 samples using the same
-single-threaded sequential Zarr code.
-Extracting the position array required 15s CPU, 23.8s elapsed.
-The af-dist calculation required 2.70min CPU, 5.23min elapsed.
-Creating the filter mask required 3.05min CPU, 11.15min elapsed.
-
-While these benchmarks are comparable with results obtained on the
-shared file-system in the Genomics England research environment,
-the sequential single-threaded access patterns used do not map
-well to modern cloud computing environments.
-As each request to a cloud object store has a high latency, it
-is important to have many requests running concurrently to achieve
-good performance~\citep{durner2023exploiting}.
-To explore the limits of performance we developed a simple prototype
-Zarr implementation, designed to maximise concurrency using asynchrous
-IO and multiple processes (see Methods for details).
-Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
-can be decoded to RAM as well as rate at which we can compute
-af-dist (using Numba to classify genotypes, as before)
-as we increase the number of parallel processes on
-an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
-The maximum decompression rate of 47.4GiB/s
-occurs at 64 processes, with a corresponding
-network data rate of 121 MiB/s
-Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
-network access by measuring the rate at which we can download the chunks
-into RAM (without decompressing) using the same parallel and asynchronous
-architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
-and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
-was achieved with 32 processes. 
-% Richard Durbin asked whether we were just using up all of a scarce 
-% resource on this point.
-While these measurments are simplistic and may not be an 
-accurate predictor of sustained bandwidth, it is still clear that 
-genotype data can be fetched
-from S3 around an order of magnitude more quickly 
-than it can be decompressed into RAM. 
-We have also not measured
-the achievable bandwidth while scaling out across many instances,
-but given the ubiquity of cloud-based services and the scale 
-at which they operate, it seems reasonable to assume that the claimed
-terabits per second of aggregate bandwidth from
-S3~\cite{s3-performance-optimising} are achievable.
-
-% Maybe don't bother with this, the same trade-off always exists.
-% but I guess it's an extra factor to take into account when weighing
-% up practical engineering decisions.
-% This has some interesting consequences
-% when we consider the goal of seeking the highest possible compression levels
-% for genotype data and the usual cloud storage cost structuring models (see
-% Discussion).
-
-Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
-calculation is performed at a maximum rate of 25.5GiB/s using 128
-processes. At this rate, the af-dist calculation for the target region
-was completed in 4.15s (compare to 5.23min for the naive single-threaded
-method above).
-This level of throughput would complete af-dist for all chromosomes for all samples
-% compute: cost of c5.9xlarge per hour * 1.27
-% ==                             $4.08 * 1.27
-% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
-% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
-in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
-See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
-benchmarks in cloud settings.
-
-\subsection{Extensibility: GPU acceleration}
-GPU acceleration has revolutionized scientific computing by leveraging the
-massively parallel architecture of GPUs to achieve significant speedups in
-computationally intensive tasks such as simulations, data analysis, and machine
-learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
-mittal2014survey,owens2008gpu}
-% (Nickolls & Dally, 2010;
-% Mittal & Vetter, 2015; 
-% Owens et al., 2008). 
-This has profound implications for
-fields ranging from molecular dynamics~\citep{stone2007accelerating}
-% (Stone et al., 2010) 
-to deep learning~\citep{lecun2015deep}
-% (LeCun et al., 2015). 
-While early GPU programming required expertise in
-low-level languages like CUDA or OpenCL, the advent of high-level Python
-libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
-%(Okuta et al., 2017), 
-and
-PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
-% (Klöckner, 2012) 
-has democratized access to GPU acceleration.
-These tools offer user-friendly interfaces and just-in-time compilation,
-allowing scientists to harness the power of GPUs within the familiar Python
-ecosystem, thereby significantly accelerating scientific discovery and
-broadening the adoption of accelerated computing across various research
-domains.
-
-% [JK: this is a super rough first narrative draft.]
-% Machine learning, and in particular deep learning methods, have led to
-% major breakthroughs in many different fields in recent years [CITES].
-% GPUs have been a key enabling factor in this [details and cites].
-% The PyData ecosystem [cites] has also enabled this. Techologies like
-% [Cupy, lots of others] substantially lower the bar for writing efficient
-% GPU code.
-% Deep learning methods depend crucially on efficient access to vast
-% datasets for training, and Zarr is emerging as a popular choice
-% [cites from geosciences?]. The RAPIDs GPU Accelerated data science
-% library [which does] now supports Zarr through [kvikio, more?].
-
-To demonstrate the ease and efficiency with this data stored in Zarr
-format can be processed on a GPU, we developed two proof-of-concept
-applications. In the first we re-implement the af-dist calculation
-used throughout as an example of a non-trivial whole-matrix operation
-on GPU. In the second, we demonstrate some emerging technology
-in which Zarr data can be decompressed directly in GPU memory.
-In both cases we demonstrate on chromosome 22 of
-1000 Genomes WGS data~\citep{byrska2022high}. This consists
-of 96,514 variants and 2504 samples, converted to Zarr format
-using a chunk size of 10,000 variants $\times$ 1,000 variants.
-Benchmarks in this section were performed on a
-machine with an Intel Xeon Silver 4216 processor,
-Nvidia A6000 RTX GPU and NVMe SSD.
-
-We first illustrate the ease with which computations can be
-be peformed on the GPU, with little or no preprocessing of the
-data. To do this we reimplemented the af-dist calculation
-described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
-of core operations for scientific computing.
-All of the operations required to compute  af-dist are
-performed on the GPU, mostly with minimal changes required
-(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
-We perform the core operation of genotype classification using the
-CuPy JIT compiler to generate a custom CUDA kernel, which
-is also very similar to the Numba compiled function used in other
-benchmarks. In the simplest version, we iterate over the chunks
-in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
-a total of 7.82MiB), decompress each to a local
-NumPy array and then copying the data to GPU memory by creating
-the corresponding CuPy array. For this version, the total time
-in the genotype classification kernel was 25ms with an
-overall data transfer time of 179ms (we discount decompression
-time here; see next paragraph). For comparison,
-our standard Numba compiled classification function required
-a total of 1.46s running on a single CPU core.
-Another approach in which we load the entire
-\texttt{call\_genotype} array into GPU memory at once (possible
-here as the $96514 \times 2504 \times 2$ array requires
-460MiB) and uses
-a more sophisticated kernel in which we map GPU resources to
-subsets of the array in two dimensions,
-reduces transfer
-time to 138ms and processing time to 6ms.
-
-Chunk decompression comprises a significant fraction of processing
-time [FIXME crossrefs?] and host-to-GPU data transfer is
-often a bottleneck. It is therefore useful in some applications
-to directly decompress Zarr chunks into GPU memory.
-KvikIO~\cite{kvikio} is a Python package that provides access to optimised
-GPU IO routines, and providing support for Zarr along with
-access to several GPU optimised compression codecs.
-To explore the performance of these codecs we recompressed
-the \texttt{call\_AD} (allele depth) array
-(96514, 2504, 7) [int16 more details]. In all cases it took around 1s
-to compress and store this 3.15GiB array. [Details about compression codecs?]
-We then measured the time to decompress these arrays and
-copy back to host memory. This required about 0.6s with the
-fastest codecs. For reference, using the Zstandard codec and the
-standard CPU Zarr code (which does some limited multithreading for decompression
-by default) required about 7.6s. [Some discussion of how to interpret
-these benchmarks. We're just trying to illustrate the powerful toolchain
- that is immediately to hand when data is stored in Zarr format.]
-
 \section{Discussion}
@@
 and xtensor~\citep{xtensor_zarr}
-as well as  standalone libraries for
-C++~\citep{ruan2024image},
-Java~\cite{jzarr},
-JavaScript~\cite{zarrjs},
+as well as  standalone libraries for JavaScript~\cite{zarrjs},
 Julia~\cite{zarrjl}, Rust~\citep{zarrs}
@@
 tools~\cite{moore2023ome,rzepka2023toward}. 
-Zarr has also seen recent uptake in other large-scale processing for
-microscopy applications~\citep{ruan2024image}, as well as
-single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
+Zarr has also seen recent uptake 
+in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
 and multimodal spatial omics
@@
 partitions so that they can be retried.
-
-\subsection{Local alleles}
-[TODO Implementation of local alleles in vcf2zarr]
 
@@
 access patterns.
-
-\subsection{vcztools}
-Vcztools is available for download on the Python package index, and
-supports all major platforms [TODO do we support Windows??],
-and is written in a combination of Python and C.
-We use
-the Zarr-Python package~\citep{zarrpython} for all Zarr operations.
-The mini-language in the \texttt{query} command is parsed using
-PyParsing [TODO cite].
-As a general principle, data is only retrieved from the Zarr store
-when it is required to fulfil a particular query. For example,
-when a subset of samples are specified, we only retrieve chunks
-in arrays such as \texttt{call\_genotype} for those
-samples. Chunks in the variants dimension are retrieved as needed,
-and we use a multithreaded strategy to fetch and decode variant chunks
-while another thread encodes VCF. Generating VCF text at speed
-is a challenging problem, and we use a C extension module to convert
-a given set of Zarr chunks represented as numpy arrays to VCF.
-Based on the \texttt{region\_index} array computed by \texttt{vcf2zarr},
-we use PyRanges~\citep{stovner2020pyranges} to perform efficient interval
-search queries.
 
@@
 unless otherwise stated.
-All simulation-based benchmarks (excluding Fig~\ref{fig-whole-matrix-compute-zarr-versions}) were performed on a
+All simulation-based benchmarks were performed on a 
 dual CPU (Intel Xeon E5-2680 v2)
@@
 (\url{https://pypi.org/project/bio2zarr/}).
-Vcztools (\url{https://github.com/sgkit-dev/vcztools/})
-is freely available under the terms of the Apache 2.0
-and can be installed from the Python Package Index
-(\url{https://pypi.org/project/vcztools/}).
 
@@
 Research Ltd Kiwifruit Royalty Investment Programme.
-PU was supported by the SciLifeLab \& Wallenberg Data Driven Life
-Science Program, Knut and Alice Wallenberg Foundation (grants: KAW
-2020.0239 and KAW 2017.0003), and by the National Bioinformatics
-Infrastructure Sweden (NBIS) at SciLifeLab
+
 % \subsection{Author's Contributions}
@@
 \section{Acknowledgements}
-We would like to thank Gil McVean and Ben Neale for helpful discussions and
-comments on the manuscript.
+We would like to thank Gil McVean for helpful comments on the manuscript.
+
 This research was made possible through access to data in the National Genomic
@@
 
-This study makes use of de-identified data held by Our Future Health.
-We would like to acknowledge all the research participants who have donated
-their data to the Our Future Health research programme~\cite{ofhpubpolicy}.
-
 Computation used the Oxford Biomedical Research Computing (BMRC) facility, a
@@
 not necessarily those of the NHS, the NIHR or the Department of Health.
-
-Computation for the Spruce case study were enabled by resources
-provided by the National Academic Infrastructure for Supercomputing in
-Sweden (NAISS), partially funded by the Swedish Research Council
-through grant agreement no. 2022-06725.
 
@@
 This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
-2.7 GiB/s Zarr (Zstd), 2.9 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy.
+3.9 GiB/s Zarr (Zstd), and 6.6 GiB/s for Savvy. 
 \label{fig-whole-matrix-decode}}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics{figures/whole-matrix-compute-zarr-versions}
-\caption{Whole-matrix computation performance using different Zarr
-implementations.
-Total CPU time required to run the af-dist calculation
-the data in Figure~\ref{fig-data-storage} using different Zarr implementations.
-Elapsed time is also reported (dotted line).
-These benchmarks were run on an 8-core CPU (Intel i7-9700) with 32GiB RAM running
-Linux Mint 21.3 with data on an NVMe SSD.
-Note the difference between the
-time reported here and in Fig~\ref{fig-whole-matrix-compute} for Zarr-Python
-is due to different hardware platforms.
-\label{fig-whole-matrix-compute-zarr-versions}}
 \end{figure}
@@
 \end{figure}
-
 \begin{figure}[h]
-\includegraphics{figures/s3-throughput}
-\caption{Scalability of genotype data processing on AWS.
-\label{fig-s3-throughput}}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics{figures/s3-network-throughput}
-\caption{Chunk data download rate on AWS.
-\label{fig-s3-network-throughput}}
-\end{figure}
-
-\begin{figure}[h]
-\begin{tabular}{ccc}
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_LRR.png}&
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_BAF.png}&
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_GS.png}
-\end{tabular}
-\caption{Distribution of floating point fields in the Our Future Health data.
-\label{fig-ofh-field-distributions}}
-\end{figure}
-
-\begin{figure}[h]
+
 \includegraphics{figures/compression-compressor}

### 2025-01-27 paper.tex
add 2025 article

@@
 \subsection{Case study: Our Future Health genotype data}
-Our Future Health~\cite{ofhmain_2025}  is the UK’s largest ever health research programme.
+Our Future Health~\cite{ofhmain_2025, Cook2025}  is the UK’s largest ever health research programme.
 It is designed to help people live healthier lives for longer through the

### 2025-01-28 paper.tex
Fixup OFH refs

--- 2025-01-26 paper.tex
@@
 UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
-Our Future Health~\citep{ofhmain_2025},
+Our Future Health~\citep{cook2025our},
 and the All of Us research program~\citep{all2024genomic}
@@
 \subsection{Case study: Our Future Health genotype data}
-Our Future Health~\cite{ofhmain_2025}  is the UK’s largest ever health research programme.
+Our Future Health~\cite{cook2025our}  is the UK’s largest ever health research programme.
 It is designed to help people live healthier lives for longer through the

### 2025-01-28 paper.tex
Add call-to-action ref
Closes #171

@@
 and the All of Us research program~\citep{all2024genomic}
-collecting genome sequence data for millions of humans.
+collecting genome sequence data for millions of humans~\citep{stark2024call}.
 Large genetic variation datasets are also being generated for other organisms

### 2025-01-28 paper.tex
Update paper.tex

@@
 of high-performance and cost-effective applications
-via illustrative highly parallel benchmarks on AWS and GPUs.
+via illustrative highly parallel benchmarks on AWS and experimentation with GPUs.
 
@@
 using a chunk size of 10,000 variants $\times$ 1,000 variants.
-Benchmarks in this section were performed on a
+Experiments in this section were performed on a
 machine with an Intel Xeon Silver 4216 processor,
@@
 is also very similar to the Numba compiled function used in other
-benchmarks. In the simplest version, we iterate over the chunks
+experiments. In the simplest version, we iterate over the chunks
 in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
@@
 by default) required about 7.6s. [Some discussion of how to interpret
-these benchmarks. We're just trying to illustrate the powerful toolchain
+these experiments. We're just trying to illustrate the powerful toolchain
  that is immediately to hand when data is stored in Zarr format.]

### 2025-01-28 paper.tex
Add mention of zarr v3

@@
 % First author
-\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
-\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
-\author[4,\authfn{1}]{Will Tyler}
-\author[5,\authfn{1}]{Tom White}
+\author[1,2\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
+\author[3,4\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
+\author[5,\authfn{1}]{Will Tyler}
+\author[6,\authfn{1}]{Tom White}
 
 % Middle, alphabetical order
-\author[6]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
-\author[7,8]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
-\author[9]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
-\author[10]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[7,8,11]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
-\author[12]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[13]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
-\author[14]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+\author[7]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
+\author[8,9]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
+\author[10]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
+\author[11]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[8,9,12]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
+\author[13]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[14]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[15]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[15]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[16]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
 % Senior
-\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
-
-\affil[1]{Related Sciences}
-\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
+\author[1,2,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
+\author[11,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+
+\affil[1]{Open Athena AI Foundation}
+\affil[2]{Related Sciences}
+\affil[3]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
 New Zealand}
-\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
-\affil[4]{Independent researcher}
-\affil[5]{Tom White Consulting Ltd.}
-\affil[6]{Our Future Health, Manchester, UK.}
-\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
+\affil[4]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
+\affil[5]{Independent researcher}
+\affil[6]{Tom White Consulting Ltd.}
+\affil[7]{Our Future Health, Manchester, UK.}
+\affil[8]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[9]{Analytic and Translational Genetics Unit, Massachusetts General
 Hospital, Boston, Massachusetts 02114, USA}
-\affil[9]{NVIDIA Ltd, Reading, UK}
-\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
+\affil[10]{NVIDIA Ltd, Reading, UK}
+\affil[11]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
 University of Oxford, UK}
-\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[12]{Wellcome Sanger Institute}
-\affil[13]{Genomics England}
-\affil[14]{Department of Cell and Molecular Biology, National
+\affil[12]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[13]{Wellcome Sanger Institute}
+\affil[14]{Genomics England}
+\affil[15]{Department of Cell and Molecular Biology, National
   Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
   Uppsala University, Uppsala, Sweden}
-\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[16]{School of Computer Science, McGill University, Montreal, QC, Canada}
 
@@
 datasets for Norway Spruce ($n$=1,063) and 
-SARS-Cov-2 ($n$=3,960,704), % TODO update for tranche 2 data 
-demonstrating the scalability and flexibility of VCF Zarr.
+SARS-CoV-2 ($n$=4,384,310).
 We demonstrate the potential for VCF Zarr to enable a new generation
 of high-performance and cost-effective applications
-via illustrative highly parallel benchmarks on AWS and experimentation with GPUs.
+via illustrative examples using cloud computing and GPUs.
 
@@
 UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
-Our Future Health~\citep{cook2025our},
+Our Future Health~\citep{cook2025our,ofhmain_2025},
 and the All of Us research program~\citep{all2024genomic}
@@
 VCF's simple text-based design and widespread
-support~\cite{garrison2022spectrum} makes it an
+support~\cite{garrison2022spectrum} make it an
 excellent archival format, but it is an inefficient basis for analysis.
@@
 and in a large collection of short whole genome alignments
-using SARS-Cov-2 data 
-(100X % FIXME! this is roughly correct, but need to get the numbers straight
-smaller than gzipped FASTA). 
+using SARS-CoV-2 data 
+(81X smaller than gzipped FASTA). 
 Interoperability with the vast ecosystem of tools
@@
 (written in C++), and show significant speedups over the existing 
-VCF (6.5X) and BCF (3X) backends in a simple benchmark.
+VCF (7.9X) and BCF (3.4X) backends in a simple benchmark.
 Finally, we illustrate the potential 
@@
 Finally, we demonstrate the ease and efficiency with which data
-in Zarr format can be processed on a GPU, opening many exiting
+in Zarr format can be processed on a GPU, opening many exciting
 avenues for future research.
@@
 in alignment
-% TODO check
-% Jeff: OK with the ADAM reference in here? You did some parquet based
-% alignment storage, right?
 storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
@@
 The data is largely numerical and of fixed dimension,
-and is therefore a natural mapping to array-oriented
-or ``tensor'' storage.
+making it a natural fit for array-oriented or ``tensor'' storage.
 We propose the VCF Zarr specification which maps the
@@
 larger sample sizes). Zarr and Savvy have
-very similar compression performance in this example (23.75GiB
-and 21.25GiB for $10^6$ samples, respectively). When using
+very similar compression performance in this example (23.75 GiB
+and 21.25 GiB for $10^6$ samples, respectively). When using
 a chunk size of $10^4$ variants $\times$ $10^3$ samples, Zarr
-storage is reduced to 22.07GiB for $10^6$ samples (data not shown;
+storage is reduced to 22.07 GiB for $10^6$ samples (data not shown;
 see Figs~\ref{fig-compression-chunksize} and \ref{fig-compression-chunksize-finegrained}
@@
 % zarr_lz4_nshf 2.9 GiB
-(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
+(up to 6.6 GiB/s vs 1.2 GiB/s for Zarr on this dataset;
 Fig~\ref{fig-whole-matrix-decode}).
@@
 however, leads to a substantial increase in decoding speed
-(2.7GiB/s) at the cost of a roughly 25\% increase in storage
-space (31.2GiB up from 23.75GiB for 1 million samples; data not
+(2.7 GiB/s) at the cost of a roughly 25\% increase in storage
+space (31.2 GiB up from 23.75 GiB for 1 million samples; data not
 shown). The LZ4 codec (again without BitShuffle) leads to slightly higher
-decoding speed (2.9GiB/s), but at the cost of much larger storage requirements
-(214GiB; data not shown).
+decoding speed (2.9 GiB/s), but at the cost of much larger storage requirements
+(214 GiB; data not shown).
 Given the relatively small contribution of genotypes to the
@@
 hierarchy) provides a useful alternative here and can
-significantly reduce overall processing time. To illustrate this
+significantly reduce overall processing time. 
+(The nascent Zarr v3 specification provides another option 
+via  ``sharding'' which essentially 
+allows multiple chunks to be stored in a single file.)
+To illustrate this
 we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
@@
 % 59         48129895      1000000    7254858            zarr.zip    24.240338
-storage space than the directory hierarchy (23.75GiB and 24.24G
+storage space than the directory hierarchy (23.75 GiB and 24.24 GiB
 for file system and Zip file, respectively)
@@
 Genomics England WGS data for chromosome 2 (78,195 samples, 59,880,903 variants),
-consisting of 44 fields and 2.54TiB of storage ($\sim$5X compression over
+consisting of 44 fields and 2.54 TiB of storage ($\sim$5X compression over
 source gzipped VCF).
@@
 power-of-two units, and the compression ratio achieved on that array.
-We also show the percentage of the overall storage that each array consumes
-(omitting values < 0.01\%).
+We also show the percentage of the overall storage that each array consumes.
+Shown are the top 11 fields consuming at least 0.01\% of the overall storage.
 \label{tab-genomics-england-data}}
-\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
+\begin{tabular}{llS[table-format=3.2]S[table-format=5.1]S[table-format=3.2]}
 \toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
+{Field} & {Type} & {Storage} & {Compress} & {\%Total} \\
 \midrule
-/call\_AD & int16 & 658.4G & 26 & 25.35\% \\
-/call\_GQ & int16 & 654.5G & 13 & 25.20\% \\
-/call\_DP & int16 & 570.0G & 15 & 21.95\% \\
-/call\_DPF & int16 & 447.1G & 20 & 17.22\% \\
-/call\_PL & int16 & 162.6G & 160 & 6.26\% \\
-/call\_GQX & int16 & 41.0G & 210 & 1.58\% \\
-/call\_FT & string & 25.0G & 1400& 0.96\% \\
-/call\_genotype & int8 & 21.5G & 410& 0.83\% \\
-/call\_genotype\_mask & bool & 12.8G & 680& 0.49\% \\
-/call\_genotype\_phased & bool & 2.4G & 1900& 0.09\% \\
-/call\_PS & int8 & 383.4M & 12000& 0.01\% \\
-% /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
-% /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
-/variant\_position & int32 & 111.6M & 2 & \\
-/variant\_quality & float32 & 87.4M & 2.6 & \\
-/variant\_allele & string & 69.3M & 13 & \\
-% /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
-/variant\_AN & int32 & 47.3M & 4.8 & \\
-% /variant\_AC & int32 & 47M & 4.9 & 0.00\% \\
-% /variant\_AC\_Het & int32 & 46.62M & 4.9 & 0.00\% \\
-% /variant\_ABratio & float32 & 20.23M & 11.0 & 0.00\% \\
-% /variant\_MendelSite & float32 & 20.09M & 11.0 & 0.00\% \\
-% /variant\_medianGQ & float32 & 19.16M & 12.0 & 0.00\% \\
-% /variant\_completeGTRatio & float32 & 14.98M & 15.0 & 0.00\% \\
-% /variant\_phwe\_eur & float32 & 14.46M & 16.0 & 0.00\% \\
-% /variant\_phwe\_afr & float32 & 13.37M & 17.0 & 0.00\% \\
-% /variant\_medianDepthNonMiss & float32 & 12.17M & 19.0 & 0.00\% \\
-% /variant\_medianDepthAll & float32 & 11.99M & 19.0 & 0.00\% \\
-% /variant\_phwe\_sas & float32 & 11.59M & 20.0 & 0.00\% \\
-% /variant\_AC\_Hom & int32 & 11.12M & 21.0 & 0.00\% \\
-% /variant\_missingness & float32 & 9.3M & 25.0 & 0.00\% \\
-% /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
-/variant\_filter & bool & 6.4M & 570 & \\
-% /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
-% /variant\_composite\_filter & bool & 433.38 KiB & 130.0 & 0.00\% \\
-% /variant\_OLD\_CLUMPED & object & 384.18 KiB & 1200.0 & 0.00\% \\
-% /variant\_id & object & 298.76 KiB & 1600.0 & 0.00\% \\
-% /variant\_phwe\_amr & float32 & 269.52 KiB & 870.0 & 0.00\% \\
-% /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
-/sample\_id & str & 268.1K & 2.3 & \\
-% /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
-% /contig\_length & int64 & 3.79 KiB & 5.3 & 0.00\% \\
-% /contig\_id & object & 2.3 KiB & 8.8 & 0.00\% \\
-% /filter\_id & object & 808 bytes & 0.6 & 0.00\% \\
-%%%% OLD for chr20
-% call\_AD &  int16 & 179.7G & 26 & 24.0\\
-% call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
-% call\_DP &  int16 & 141.8G & 16 & 18.9 \\
-% call\_DPF& int16  & 115.1G & 20 & 15.3\\
-% call\_FT &  string & 58.5G & 160 & 7.8 \\
-% call\_PL &  int16 & 51.4G & 140 & 6.9 \\
-% call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
-% call\_genotype & int8 & 6.1G & 380 & 0.8 \\
-% call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
-% call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
-% call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
-% variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
-% variant\_allele & string & 21.5M & 11 \\
-% variant\_position & int32 & 12.7M & 4.7 \\
-% variant\_ABratio & float32 & 9.6M & 6.3 \\
-% variant\_AN & int32 & 9.6M & 6.3 \\
-% variant\_filter & bool & 1.9M & 490 \\
+call\_AD & int16 & 658.44 GiB & 26.0 & 25.35\% \\
+call\_GQ & int16 & 654.45 GiB & 13.0 & 25.20\% \\
+call\_DP & int16 & 570.03 GiB & 15.0 & 21.95\% \\
+call\_DPF & int16 & 447.09 GiB & 20.0 & 17.22\% \\
+call\_PL & int16 & 162.56 GiB & 160.0 & 6.26\% \\
+call\_GQX & int16 & 40.99 GiB & 210.0 & 1.58\% \\
+call\_FT & str & 25.04 GiB & 1400.0 & 0.96\% \\
+call\_genotype & int8 & 21.46 GiB & 410.0 & 0.83\% \\
+call\_genotype\_mask & bool & 12.78 GiB & 680.0 & 0.49\% \\
+call\_genotype\_phased & bool & 2.35 GiB & 1900.0 & 0.09\% \\
+call\_PS & int8 & 383.38 MiB & 12000.0 & 0.01\% \\
+call\_ADF & int8 & 383.38 MiB & 12000.0 & 0.01\% \\
+call\_ADR & int8 & 383.38 MiB & 12000.0 & 0.01\% \\
 \bottomrule
@@
 As these benchmarks take some time to run, we focus
-on a single 132GiB compressed VCF file covering
+on a single 132 GiB compressed VCF file covering
 positions 58,219,159--60,650,943 (562,640 variants)
@@
 The major difference between CPU time and wall-time is noteworthy
-here, and indicates some opportunities for improvement in VCF Zarr
-in high-latency environments such as the shared file system in the
-Genomics England HPC system. Currently VCF Zarr does not store any
-specialised index to map genomic coordinates to array positions
-along the variants dimension. Instead, to find the relevant slice
+here, and highlights the importance of using an index to find
+the location of variant chunks.
+This benchmark predates the \texttt{region\_index}
+array computed by \texttt{vcf2zarr} (see Methods) which 
+stores sufficient information to identify which variant chunks intersect
+with a given range query.
+Here, to find the relevant slice
 of records corresponding to the range of positions in the target
-VCF file, we load the entire variant\_position array and
+VCF file, we simply load the entire \texttt{variant\_position} array and
 binary search. This entails reading 5,989 chunk files
-(the chunk size is 100,000 variants) which incurs a substantial
-latency penalty on this system. Later versions of the specification
-may solve this problem by storing an array of size
-(approximately) the number variant chunks
-which maps ranges of genome coordinates to chunk indexes,
-or a more specialised structure that supports overlap queries.
+(the chunk size is 100,000 variants) and incurred a substantial
+latency penalty on the Genomics England HPC shared filesystem.
 
@@
 \subsection{Case study: Our Future Health genotype data}
-Our Future Health~\cite{cook2025our}  is the UK’s largest ever health research programme.
+Our Future Health~\cite{cook2025our,ofhmain_2025}  is the UK’s largest ever health research programme.
 It is designed to help people live healthier lives for longer through the
@@
 detection and treatment of diseases.
-Our Future Health is inviting millions of people, from all backgrounds and
-from right across the UK to take part. Volunteers are providing information
-about their health and lifestyles to create an incredibly detailed picture
-that represents the whole of the UK.
+Data Release 9~\citep{ofhdatarelease9} was made available in December 2024,
+including genotype array data for from 651,050 individuals
+and 707,522 variants.
+For this study, we converted the genotype data for chromosome 22 consisting 
+of 10,221 variants over three gzipped VCF files and a total of
+42.3 GiB of storage. 
+% 20G     ofh.chr22-b0.vcf.gz
+% 19G     ofh.chr22-b1.vcf.gz
+% 3.3G    ofh.chr22-b2.vcf.gz
+% ==42.3 GiB - assuming the G is power-of-two here
+% 10,221 variants and 42.3 GiB of storage.
 
@@
 Our Future Health genotype data for chromosome 22 (651,050 samples, 10,221
-variants) consisting of 19 fields and 17GiB of storage ($\sim2.5$X compression
+variants) consisting of 19 fields and 17 GiB of storage ($\sim2.5$X compression
 over source gzipped VCF).
@@
 details). 
-For column details, see Table~\ref{tab-genomics-england-data}.
+Shown are the top 6 fields consuming at least 0.01\% of the overall storage 
+(see Table~\ref{tab-genomics-england-data} for column details).
 \label{tab-ofh-data}}
@@
 \midrule
-/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
-/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
-/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
-/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
-/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
-/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
-/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
-/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
-/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
-/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
-/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
-/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
-/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
-/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
-/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
-/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
-/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
+call\_LRR & float32 & 8.35 GiB & 3.0 & 52.08\% \\
+call\_BAF & float32 & 6.02 GiB & 4.1 & 37.55\% \\
+call\_GS & float32 & 961.37 MiB & 26.0 & 5.86\% \\
+call\_genotype & int8 & 657.21 MiB & 19.0 & 4.00\% \\
+call\_genotype\_mask & bool & 79.48 MiB & 160.0 & 0.48\% \\
+sample\_id & str & 2.93 MiB & 1.7 & 0.02\% \\
 \bottomrule
@@
 
-Release 9~\citep{ofhdatarelease9}
-was made available in December 2024. The genetic component includes genotype data
-from 651,050 individuals on 707,522 variants. Here we use the genotype data from
-chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
-three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling
-% 20G     ofh.chr22-b0.vcf.gz
-% 19G     ofh.chr22-b1.vcf.gz
-% 3.3G    ofh.chr22-b2.vcf.gz
-% ==42.3 GiB - assuming the G is power-of-two here
-10,221 variants and 42.3GiB of storage. We first converted the VCF
+Conversion was performed on a Standard\_D32s\_v5 Azure VM instance, with 
+32 CPUs and 128 GB RAM. We first converted the VCF
 data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
-The ICF representation required a total of 41GiB over XX files.
+The ICF representation required a total of 41 GiB of storage.
 We then converted to VCF Zarr format using the default schema, which resulted
-in a 41GiB VCZ. On inspection, this was dominated by three floating point
-fields LRR, BAF and GS stored
+in a 41 GiB VCZ over 4514 chunk files. 
+On inspection, this was dominated by three floating point
+fields LRR (Log R Ratio), BAF (B allele frequency) and GS (GenCall Score)
+stored
 to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
@@
 (Fig~\ref{fig-ofh-field-distributions}) it seemed that such high precision
-was likely unneccessary, and we investigated applying the Quantize
-and Bitround filters from \texttt{numcodecs}~\citep{numcodecs}
+was likely unnecessary, and we investigated applying the Quantize
+and BitRound filters from \texttt{numcodecs}~\citep{numcodecs}
 to reduce the precision in stored floating point data,
@@
 Using the Quantize filter with 5 digits of precision on these three fields reduced the
-overall storage space to 31G with a mean absolute error (MAE, i.e., the average
+overall storage space to 31 GiB with a mean absolute error (MAE, i.e., the average
 difference between the truncated and original values) of
@@
 $1.88\times10^{-6}$ (GS).
-Using the Bitround filter with a parameter of 5 bits reduced the overall storage for the Zarr to
-XX with an MAE of
+Using the BitRound filter with a parameter of 5 bits reduced the overall storage for the Zarr to
+17 GiB with an MAE of
 % 0   call_LRR    bitround    6.93e-04
@@
 $3.38\times10^{-3}$ (GS).
-The histograms of the truncated fields in all cases were visually indistinguishable
-from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
-Table~\ref{tab-ofh-data} shows a summary of the Zarr using the BitRound filter.
+The histograms of the truncated fields in all cases were very similar to the
+to the original (Fig~\ref{fig-ofh-field-distributions}).
+Table~\ref{tab-ofh-data} shows a summary of the Zarr dataset using the 
+BitRound filter.
 
@@
 identifying over 1 billion variants, including 3.9 million novel variants
-with coding consequences~\citep{all2024genomic}. Participants are from the United States,
+with coding consequences~\citep{all2024genomic}. 
+Participants are from the United States,
 with an emphasis on historically underrepresented genetic ancestry groups.
@@
 is of high quality, with an average coverage depth of 30x, and is linked to extensive
-health-related information, providing a valuable resource for biomedical research.
-WGS analysis is conducted on the Illumina DRAGEN platform, with alignment performed
-using the GRCh38dh reference genome. GTC files are converted to VCF format using
-Picard GtcToVcf, and every step of the WGS procedure is rigorously controlled by
-predefined quality control measures.
-
+health-related information.
 For this study, we converted the chromosome 20 exome-like
@@
 The exome-like data is constructed by extracting from all WGS samples
-the SNP and indel variants that lie within the chromosome 20 exon regions of the Gencode v42 basic
+the SNP and indel variants that lie within the 
+chromosome 20 exon regions of the Gencode v42 basic
 transcripts. This dataset, stored as a single compressed VCF file of 7.44 GiB,
-includes 715,256 variants, of which more than 700 are classified as pathogenic or likely pathogenic.
+includes 715,256 variants, of which more than 700 are classified as 
+pathogenic or likely pathogenic.
 
@@
 \caption{Summary for VCF Zarr conversion of 
-All of Us exome-like data genotype data for chromosome 22 (245,394 samples, 
+All of Us exome-like data genotype data for chromosome 20 (245,394 samples, 
 715,256 variants) 
-consisting of 25 fields and 8.5GiB of storage ($\sim1.1$X larger
+consisting of 25 fields and 8.5 GiB of storage ($\sim1.1$X larger
 % 8.5 / 7.44 = 1.142
@@
 see text for details.
-For column details, see Table~\ref{tab-genomics-england-data}.
+Shown are the top 14 fields consuming at least 0.01\% of the 
+overall storage (see Table~\ref{tab-genomics-england-data} for column details).
+The \texttt{variant\_homozygote\_count} field has been renamed 
+\texttt{variant\_hom\_count} for display purposes.
 \label{tab-aou-data}}
@@
 \midrule
-/call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
-/call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
-/call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
-/call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
-/call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
-/call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
-/call\_FT & object & 498.61 MiB & 2700.0 & 6.00\% \\
+call\_LAD & int16 & 2.26 GiB & 290.0 & 27.85\% \\
+call\_GQ & int8 & 1.89 GiB & 87.0 & 23.29\% \\
+call\_LA & int8 & 1.24 GiB & 260.0 & 15.28\% \\
+call\_genotype & int8 & 914.22 MiB & 370.0 & 11.00\% \\
+call\_RGQ & int16 & 729.57 MiB & 460.0 & 8.78\% \\
+call\_genotype\_mask & bool & 606.02 MiB & 550.0 & 7.29\% \\
+call\_FT & str & 498.61 MiB & 2700.0 & 6.00\% \\
+call\_genotype\_phased & bool & 17.17 MiB & 9700.0 & 0.21\% \\
+variant\_AF & float32 & 5.95 MiB & 43.0 & 0.07\% \\
+% /variant\_homozygote\_count & int32 & 4.78 MiB & 54.0 & 0.06\% \\
+variant\_hom\_count & int32 & 4.78 MiB & 54.0 & 0.06\% \\
+variant\_allele & str & 4.74 MiB & 110.0 & 0.06\% \\
+variant\_AC & int32 & 4.24 MiB & 61.0 & 0.05\% \\
+variant\_filter & bool & 2.87 MiB & 0.9 & 0.03\% \\
+variant\_AN & int32 & 904.75 KiB & 3.1 & 0.01\% \\
 \bottomrule
@@
 
-[TODO discuss local alleles and mention Table~\ref{tab-aou-data}].
-
-[Discuss why we're bigger than VCF here, people will want to understand.
+Local alleles (see Methods) is essential in this example
+as the maximum number of alleles observed at a site is 95.
+Thus, as defined, the AD field field has an inner dimension of 95
+each $1000\times10000$ chunk 
+% >>> 1000 * 10_000 * 95 * 2 / 1024**3
+% 1.7695128917694092
+has an uncompressed size of 1.77 GiB (in this 16 bit integers are required
+to store the values). Given that the vast majority
+of this encoding consists of the dimension padding sentinel (-2, see 
+Methods), it is an inefficient means of representing and processing 
+allele depth data. Translating to the local alleles representation
+here reduces the inner dimension to 2, and the uncompressed 
+% >>> 1000 * 10_000 * 2 * 2 / 1024**2
+% 38.14697265625
+size of each chunk to 38.15 MiB. 
+Note that there is also no loss of information in this case 
+by encoding as local alleles, as the per-sample gVCFs used 
+in earlier stages in the All of Us pipeline were truncated
+to a maximum of two ALT alleles per site.
+
+[Discussion of the sparsity of the QC field data, and how this 
+explains the difference to the VCF compression.]
 
@@
 where resequencing data has been mapped to a chromosome-scale
-reference genome. The dataset consists of approximately 3.8 million
+reference genome. The dataset consists of approximately 3.75 billion
 single-nucleotide variants and small indels from the 12 autosomes
 totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
-distributed over 165 VCF files. Since the dataset originally was
-generated for downstream analyses that require genotype likelihoods,
-the genotypes include the PL field.
-
-Following the conversion process detailed in the previous section, we
-first converted the 165 VCF files (7.33TiB) to ICF, followed by
+distributed over 165 VCF files. 
+% Since the dataset originally was
+% generated for downstream analyses that require genotype likelihoods,
+% the genotypes include the PL field.
+
+We first converted the 165 VCF files (7.33 TiB) to ICF, followed by
 conversion to Zarr. The tasks were run as single jobs on a compute
@@
 required 13h45min runtime. The ICF representation used a total of 6.77
-TiB over 1,340,804 data storage files. Conversion to Zarr required
-18h36min runtime, generating a dataset with 32 arrays, consuming a
-total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
-total storage (1.1X) is negligible, mainly due to the inherent
-difficulty in compressing the PL field. The output dataset storage
-size is dominated by a major component, call\_PL, which accounts for
-nearly 93\% of the total. call\_PL is yet another field with low
-compression potential due to inherent noisiness.
-
-Therefore, to reduce storage requirements we generated a Zarr schema
-for encoding that uses local allele fields with \texttt{vcf2zarr
+TiB over 1,340,804 data storage files. Conversion to Zarr 
+using the default schema and a chunk size of 10,000 variants
+$\times$ 1,063 samples
+required 18h36min runtime, generating a dataset with 34 arrays, 
+and required a total of 6.6 TiB storage over 10,861,038 chunk files. 
+Over 90\% of this was used by the call\_PL field, mostly due to 
+the maximum number of alleles at a site being 4,
+and consequently the inner dimension of call\_PL 10 because 
+of its quadratic dependency on the number of alleles (see Methods).
+
+We therefore generated a vcf2zarr schema
+using local allele fields with \texttt{vcf2zarr
 mkschema --local-alleles}, and used the generated schema as input to
 \texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
-generating a dataset with 35 arrays, consuming a total of 2.3TiB
-storage over 11,235,556 chunk files, with a reduction in storage of
-3.2X. Therefore, by using local allele fields we achieve approximately
-3X improvement of compression. The top fields in terms of storage are
-detailed inTable~\ref{tab-spruce-data}.
+generating a dataset with 35 arrays, consuming a total of 2.3 TiB
+storage over 11,235,556 chunk files. 
+Using local allele fields in this example we therefore 
+achieve approximately 3X improvement of compression. 
+The top fields in terms of storage are
+detailed in Table~\ref{tab-spruce-data}.
 
 \begin{table}
-\caption{Summary for VCF Zarr conversion of the Norway spruce
-WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
-consisting of 34 fields and 2.3TiB of storage ($\sim 3.2$X smaller 
+\caption{Summary for VCF Zarr conversion of the Norway Spruce
+WGS data on all 12 autosomes (1,063 samples, 
+%3745170452
+3,745,170,452 variants)
+consisting of 34 fields and 2.3 TiB of storage ($\sim 3.2$X smaller 
 than source gzipped VCFs). The chunk size is 10,000 variants $\times$
-1,063 samples (i..e, a single sample chunk). This uses local 
+1,063 samples (i.e., a single sample chunk). This uses local 
 alleles to reduce the 
 space required by the PL field; see the text for discussion.
-For column details see the caption of
-Table~\ref{tab-genomics-england-data}.
+Shown are the top 21 fields consuming at least 0.01\% of the 
+overall storage (see Table~\ref{tab-genomics-england-data} for column details).
 \label{tab-spruce-data}}
-
 \begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
 \toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
+{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
 \midrule
-/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
-/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
-/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
-/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
-/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
-/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
-/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
-/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
-/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
-/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
-/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
-/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
-/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
-/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
-/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
-/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
-/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
-/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
-/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
-/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
-/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
-% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
-% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
-% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
-% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
-% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
-% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
-% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
-% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
-% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
-% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
+call\_LPL & int16 & 1.26 TiB & 17.0 & 54.76\% \\
+call\_LA & int8 & 415.39 GiB & 18.0 & 17.63\% \\
+call\_genotype & int8 & 381.28 GiB & 19.0 & 16.18\% \\
+call\_genotype\_mask & bool & 103.84 GiB & 71.0 & 4.41\% \\
+variant\_DP4 & int32 & 26.24 GiB & 2.1 & 1.11\% \\
+variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56\% \\
+variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55\% \\
+variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55\% \\
+variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55\% \\
+variant\_quality & float32 & 12.84 GiB & 1.1 & 0.55\% \\
+variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.55\% \\
+variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54\% \\
+variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.51\% \\
+variant\_position & int32 & 9.65 GiB & 1.4 & 0.41\% \\
+variant\_AC & int16 & 7.42 GiB & 2.8 & 0.31\% \\
+variant\_DP & int32 & 6.71 GiB & 2.1 & 0.28\% \\
+variant\_allele & str & 5.37 GiB & 21.0 & 0.23\% \\
+variant\_AN & int16 & 3.05 GiB & 2.3 & 0.13\% \\
+call\_genotype\_phased & bool & 1.75 GiB & 2100.0 & 0.07\% \\
+variant\_filter & bool & 1.46 GiB & 2.4 & 0.06\% \\
+variant\_MQ & int8 & 489.41 MiB & 7.3 & 0.02\% \\
 \bottomrule
@@
 adding complexity to downstream processing.
-
 A similar limitation is imposed by the VCF POS field which is encoded
@@
 
-\subsection{Case study: SARS-Cov-2 alignment data}
+\subsection{Case study: SARS-CoV-2 alignment data}
 The SARS-CoV-2 pandemic resulted in viral sequence data collection at
@@
 overwhelmed classical phylogenetic methods, leading to the development
-of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}.
+of a new class of tools~\citep{turakhia2021ultrafast,ye2022matoptimize,zhan2023towards}
+capable of handling millions of samples.
 The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled
-set of consensus sequences for 4,471,579 samples, addressing a range of systematic
-errors in earlier collections. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
+set of consensus sequences for 4,384,310 samples, addressing a range of systematic
+errors. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
 input format, and Zhan et al have made the Viridian dataset,
@@
 One of the unusual properties of this dataset is that whole genome alignments
-are stored, such that information on all 29903 bases is stored for every
-sample. Given the number of samples and the fact that nearly all bases
-[TODO be concrete?] contain variation it is simply more convenient to
+are stored, such that information on all 29,903 bases is stored for every
+sample. Given the number of samples and that nearly all bases have 
+variation (all sites have at least one sample with missing data,
+and 99.4\% have more than one allele)
+it is simply more convenient to
 store the entire alignment for each sample than to manage the complexity
@@
 is that the \texttt{variant\_alleles} field contains the REF and ALT
-alleles as observed (such that the zeroth allele is always the reference).
+alleles as observed (such that the zero'th allele is always the reference).
 In this case it is more convenient to encode the alleles as the original
-bases and IUPAC uncertainty codes, such that the allele 0 is always A, 1 is C etc.
-Even in this case with very high sequence diversity, uncertainty codes and
-storing data for the entire genome, compression rates are very high.
-
-Based on the batch 1 release of Viridian, the Zarr dataset contains
-3,960,704 sequences of length 29903 base pairs along with 31 metadata columns
-encoding information about those samples. This full dataset is stored in a
-single 345MiB Zip file, 241MiB of which is used by the alignment data. This
-gives an average compression ratio of 468X.
-
-Computational accessibility is excellent. For example, computing the number of
-samples with missing or ambiguous data at each site requires around 3 minutes,
-iterating over sample chunks sequentially (5 lines of Python).
-Retrieving the full alignment for a given sample requires 163ms and
-getting the calls for all samples at a given site 209ms. [TODO confirm hardware
-details]. The storage of metadata and genetic data in a single file also
-makes many queries efficient and convenient. For example, computing the
-average number of deletions in samples with a Scorpio designation of Alpha
-required x seconds and 5 lines of code using Dask.
+bases and IUPAC uncertainty codes~\citep{cornish1985nomenclature}, 
+such that allele 0 is always A, 1 is C etc.
+Other than these minor deviations in interpretation, the Zarr follows
+the VCF Zarr specification and can be processed using the same tools.
+
+\begin{table}
+\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
+whole genome alignments and metadata 
+(4,384,310 samples, 29903 variants)
+consisting of 37 fields and 391 MiB of storage, 
+of which 280MiB is used by aligments ($\sim 81$X smaller 
+than gzipped FASTA). The chunk size is 100 variants $\times$
+10,000 samples.
+Shown are the top 9 fields consuming at least 1\% of the 
+overall storage (see Table~\ref{tab-genomics-england-data} for column details).
+Field names (derived from source TSV file) are shortened for 
+display purposes by replacing ``Genbank'' with ``G''
+and ``Viridian'' with ``V''.
+\label{tab-sc2-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
+\toprule
+{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
+\midrule
+call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
+sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
+sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
+sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
+sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
+sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
+sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
+sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
+sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Decompressed, the Viridian dataset consists of 48
+FASTA files totaling 125 GiB of storage and a 1.4 GiB
+TSV file for sample metadata.
+When aligned to the reference and converted to Zarr, the
+entire dataset is stored in a single 391 MiB Zip file.
+When compressed with bgzip (so they can be indexed; see next 
+paragraph) the alignments require 22 GiB of storage.
+Table~\ref{tab-sc2-data} summarises the storage used by individual
+arrays, and shows that the alignment data requires 298 MiB,
+around 81X smaller than gzipped FASTA.
+The original dataset is distributed as xz compressed FASTA files, which
+achieves even higher compression levels at a total of 218 MiB.
+However, these files must be decompressed in order to access the data
+using standard FASTA processing techniques.
+
+Computational accessibility for the Zarr is excellent. For example,
+computing the total number of missing or ambiguous bases for 
+all 4.4 million samples required 55.5 seconds elapsed (6min 45s CPU)
+using 8 threads. This required 11 lines of code, using only the 
+Python standard library and NumPy. Similarly, computing allele
+counts at each site required 4min 53s (25min 35s CPU) using a slightly
+more complicated 16 line Python function.
+Existing methods for working with FASTA data provided efficient 
+access to individual alignments using the fai index format, which
+works with uncompressed or bgzip compressed files.
+For example, retrieving the alignment for the 1 millionth sample 
+from the corresponding chunk file requires only a few milliseconds
+using pyfaidx~\citep{shirley2015efficient} or 
+pysam~\citep{pysam}. Retrieving the same alignment using Zarr 
+required 365ms. 
+Accessing calls for all samples at at a given 
+site requires a similar time (here, accessing site 14,951
+required 276ms). 
+FASTA cannot support efficient retrieval
+of data for a particular site in a set of alignments.
+Storing data chunked in two dimensions means that we can avoid
+the classic choice between sample-major or variant-major storage,
+and access it efficiently in both dimensions.
 
@@
 Vcztools currently supports Zarr stored in the local file system
-and Zip archives, and support for major cloud stores will be added in the near
-future. [JK: note this is a pretty simple feature, maybe we should just try
-to add it?]
+and Zip archives, and support for major cloud stores will be added in the near future. 
 Thus, users can convert their VCF data to Zarr using \texttt{bio2zarr},
@@
 last 10,000 variants in the dataset (using the previously extracted
-position column to determine the coordinates), \texttt{vcztools view
-chr21\_10\_5.vcz -r1:47958263-}). This took 30s, at a rate of 124.3 MiB
+position column to determine the coordinates), 
+\texttt{vcztools view
+chr21\_10\_5.vcz -r1:47958263-}. This took 30s, at a rate of 124.3 MiB
 of VCF text per second. The bcftools command on the corresponding
@@
 in existing workflows based on pipes, where VCF text is sequentially
-filtered and processed by different processes in the classical
+filtered and processed in the classical
 Unix manner. In this context the most important property is that vcztools
@@
 it. To get an estimate of the rate at which VCF can be parsed, we wrote a simple
-program that reads a VCF file and parses it line-by-line using the \texttt{htslib}
-API~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
-above (3.73 GiB) this took 31s, giving a parse rate of 123.1 MiB/s
+C program that reads a VCF file and parses it line-by-line using 
+\texttt{htslib}~\citep{bonfield2021htslib}. 
+For the 10,000 variant subset discussed
+above (3.73 GiB) this took 31s, giving a parsing rate of 123.1 MiB/s
 (almost identical to the write-rate of vcztools above). Given this limiting
@@
 
+Vcztools is implemented in Python and C, and is available for download 
+from the Python package index. 
+Based on the \texttt{region\_index} array computed by \texttt{vcf2zarr},
+we use PyRanges~\citep{stovner2020pyranges} to perform efficient interval
+search queries. 
+% While support direct support for VCF Zarr in \texttt{bcftools} would be the 
+% ideal outcome, implementation in C within \texttt{htslib} would be significant 
+% undertaking.
+
 \subsection{Interoperability: SAIGE proof-of-concept}
-To demonstrate the potential for integrating VCF Zarr into existing utilities,
+To demonstrate the potential for integrating VCF Zarr into existing
+high-performance analysis tools,
 we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
 zhou2020scalable}, a popular package for performing genome wide
-association tests. Using the Tensorstore C++ library~\cite{tensorstore}
-we added support for VCF Zarr via an additional backend class [some details
-about the implementation].
-
-To test the performance of this implemenentation we compared the time required
-to perform an association test using the simulated dataset of 10,000 samples
-from Fig~\ref{fig-data-storage}. We generated a phenotype using
-\texttt{tstrait}~\citep{tagami2024tstrait}.
-
-We compared the same analyis using the existing VCF, BCF and Savvy backends
-to our prototype Zarr backend.
-[Times: BCF 3m, VCF6.5m, Savvy 39s, Zarr 55s]
-
-% BCF:
-% real 3m1.068s user 2m56.944s sys 0m2.941s
-
-% VCF:
-% real 6m30.335s user 6m25.193s sys 0m2.842s
-
-% Savvy:
-% real 0m39.660s user 0m37.498s sys 0m2.183s
-
-% Zarr:
-% real 0m55.957s user 1m10.375s sys 0m12.111s
+association tests. 
+Using the TensorStore C++ library~\cite{tensorstore}
+we added support for VCF Zarr via an additional backend 
+(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats),
+for which the core functionality is defined in a 264 line C++ file.
+TensorStore provides efficient read and write access to array
+data in several formats, used, for example, in
+checkpointing~\cite{orbax-tensorstore} 
+TensorFlow model training~\citep{tensorflow2015-whitepaper}.
+It supports several cloud platforms, and is based on an
+asyncronous, parallel processing model (see Discussion).
+
+To assess the performance of this implementation we compared the time required
+to perform an association test using the simulated dataset of 100,000 samples
+from Fig~\ref{fig-data-storage}. 
+We performed a single-variant association
+test based on a phenotype simulated using
+\texttt{tstrait}~\citep{tagami2024tstrait},
+and measured the overall time required when using the 
+equivalent VCF (162m), BCF (70m), SAV (7m) and Zarr (20m) files.
+Thus, in this case, substantially improved performance over the 
+most popular formats, along with native cloud-store support are 
+possible with relatively minor coding effort.
+We validated the implementation by comparing the association
+test results with the output of the other backends,
+confirming that they were numerically identical. 
+As TensorStore does not currently support string data for Zarr, 
+however,
+we were not able to fully replicate the output of fields 
+such as MarkerID, Allele1 etc.
+
+%BCF:
+%user   system  elapsed 
+%4195.218    4.671 4200.508
+%VCF:
+%user  system elapsed 
+%9718.428    4.735 9727.995 
+%Savvy:
+%user  system elapsed 
+%415.371   2.420 418.098 
+%Zarr:
+%user   system  elapsed 
+%1758.976  259.924 1223.767
+
+% BCF: 4200.508 / 1223.767 = 3.43
+% VCF 9727.995 / 1223.767 = 7.95
 
@@
 To explore the use of cloud computing services with VCF Zarr,
-we uploaded the Zarr directory hierarchy for the Genomics England
-data chromosome 2 described in Table~\ref{tab-genomics-england-data}
-to an AWS S3 standard bucket. 
-We repeated the benchmarks above on the same subset
+Genomics England provided access to Aggv2 data 
+in VCF Zarr format within their secure
+cloud environment. The VCF Zarr files for Chromosome 2
+described in Table~\ref{tab-genomics-england-data}
+were copied to a standard AWS S3 bucket.
+Using this data we repeated the benchmarks above on the same subset
 of 562,640 variants and 78,195 samples using the same
@@
 To explore the limits of performance we developed a simple prototype
-Zarr implementation, designed to maximise concurrency using asynchrous
+Zarr implementation, designed to maximise concurrency using asynchronous
 IO and multiple processes (see Methods for details).
@@
 an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
-The maximum decompression rate of 47.4GiB/s
-occurs at 64 processes, with a corresponding
-network data rate of 121 MiB/s
+The maximum decompression rate of 47.4 GiB/s
+occurs at 64 processes (each with 100 concurrent requests), 
+with a corresponding network data rate of 121 MiB/s
 Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
@@
 into RAM (without decompressing) using the same parallel and asynchronous
-architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
-and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
+architecture. Here, the maximum rate of 1.2 GiB/s for a c5.9xlarge instance
+and 2.4 GiB/s for a c5n.9xlarge (high performance network) instance
 was achieved with 32 processes. 
@@
 % resource on this point.
-While these measurments are simplistic and may not be an 
+While these measurements are simplistic and may not be an 
 accurate predictor of sustained bandwidth, it is still clear that 
@@
 Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
-calculation is performed at a maximum rate of 25.5GiB/s using 128
-processes. At this rate, the af-dist calculation for the target region
-was completed in 4.15s (compare to 5.23min for the naive single-threaded
-method above).
+calculation is performed at a maximum rate of 25.5 GiB/s using 128
+processes. At this rate, the af-dist calculation for the target 
+% 180.80 MB in 4503 files
+% Total data: 90060.00 MB in 4503 chnnks
+region---172 MiB retrieved in 4503 chunks, 83.87 GiB of genotype data
+decompressed and processed---
+% NOTE: this doesn't quite up to the 25.5G above:  83.87 / 4.15 = 20.2
+was completed in 4.15s.
 This level of throughput would complete af-dist for all chromosomes for all samples
@@
 in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
-a total of 7.82MiB), decompress each to a local
+a total of 7.82 MiB), decompress each to a local
 NumPy array and then copying the data to GPU memory by creating
@@
 here as the $96514 \times 2504 \times 2$ array requires
-460MiB) and uses
+460 MiB) and uses
 a more sophisticated kernel in which we map GPU resources to
@@
 (96514, 2504, 7) [int16 more details]. In all cases it took around 1s
-to compress and store this 3.15GiB array. [Details about compression codecs?]
+to compress and store this 3.15 GiB array. [Details about compression codecs?]
 We then measured the time to decompress these arrays and
@@
 quadratically on the number of alleles.
-While the local alleles representation~\cite{poterba2024scalable}
+While the local alleles representation~\cite{poterba2025scalable}
 is a substantial improvement (see All of Us and Norway Spruce
@@
 % to parallelise in the past, and is part of the motivation 
-% for the current practise of providing VCF data in sub-chromosome partitions 
+% for the current practice of providing VCF data in sub-chromosome partitions 
 % (see the Genomics England, Our Future Health and Norway Spruce case-studies). 
@@
 optimise CPU-cache access patterns, as well as highly optimised
-bit and byte shuffle filters.  Remarkably, on highly
+bit-wise and byte-wise shuffle filters.  Remarkably, on highly
 compressible datasets, Blosc decompression can be faster
@@
 is padded with a sentinel value (-2 for integers and a specific
-non-signalling NaN for floats). While this is not a problem
-in practice for datasets in which all four bases are observed,
-it is a substantial issue for
-fields that have a quadratic dependency on the number of alleles
-(Number=G) such as PL. Such fields are already known to cause
-significant problems, and the ``local alleles'' proposal
-provides an elegant solution~\cite{poterba2024scalable}.
-As this approach is on a likely path to
-standardisation~\cite{danecek2021twelve}, we plan to
-include support in later versions of VCF Zarr.
+non-signalling NaN for floats). This is problematic in situations
+where some sites have a large number of alleles (e.g., 95 in the All
+of Us case study above), and 
+in particular for fields that have a quadratic dependency on the number of alleles
+(Number=G) such as PL. We use the local alleles
+encoding~\cite{poterba2025scalable} to reduce the storage 
+requirements for such fields (see the next section for 
+implementation details).
 
@@
 
-\subsection{Local alleles}
-[TODO Implementation of local alleles in vcf2zarr]
+The implementation of the local alleles data reduction technique 
+in \texttt{vcf2zarr} follows the approach of Poterba 
+et al~\cite{poterba2025scalable}.
+Specifically, we identify the alleles in use for each call by
+examining the genotypes and then store the LA field,
+listing the (maximum of ploidy) unique alleles observed as indexes into the 
+global list for that variant.
+Using the LA field we then compute derived fields such as 
+AD and PL in terms of the local alleles during Zarr encoding
+(see e.g.\ the \texttt{call\_LAD} field in Table~\ref{tab-aou-data}
+and
+\texttt{call\_LPL} field in Table~\ref{tab-spruce-data}).
+Note that our approach (along with Poterba et al)
+differs from the details in the VCF 4.5 specification~\citep{vcf4_5},
+which defines the LAA field containing indexes of ALT 
+alleles only. Explicitly listing the alleles observed leads
+to simpler code and means we do not need to refer back to the 
+genotypes to distinguish missing data from homozygous reference 
+calls. Note also that the local alleles encoding may be lossy,
+as only information relevant to the chosen alleles for 
+each call is retained.
+
+To facilitate efficient region search queries, we compute an
+additional array \texttt{region\_index} which stores information
+about the range of genomic positions in each chunk
+and provides sufficient information to implement interval
+search queries. We can then load this index into memory and 
+use it to determine which variant chunks need to be retrieved.
+In the Norway Spruce example (Table~\ref{tab-spruce-data}),
+which has 3.75 billion variants,
+this index requires 3.14 MiB of storage over 12 chunk files
+to store information about all 
+374,518 variant chunks 
+(the \texttt{variant\_position} array, for comparison 
+is 13.95 GiB uncompressed).
 
@@
 the first 100,000 lines of the VCF
-for chromosome 22 (giving a 1.1GiB compressed VCF)
+for chromosome 22 (giving a 1.1 GiB compressed VCF)
 and converted to Zarr using
@@
 examined the effects of varying chunk sizes and compressor settings
-on the compression ratio for call-level fields. We excluded call\_PL
-from this analysis as it requires conversion to a ``local alleles''
-encoding~\citep{poterba2024scalable} to be efficient,
-which is planned for implementation in a future
-version of \texttt{vcf2zarr}.
+on the compression ratio for call-level fields. (
+We excluded \texttt{call\_PL} from this analysis as this analysis predated 
+our implementation of local alleles.)
 
@@
 
-\subsection{vcztools}
-Vcztools is available for download on the Python package index, and
-supports all major platforms [TODO do we support Windows??],
-and is written in a combination of Python and C.
-We use
-the Zarr-Python package~\citep{zarrpython} for all Zarr operations.
-The mini-language in the \texttt{query} command is parsed using
-PyParsing [TODO cite].
-[Add a note here on why it was easier to to write a Python application
-from scratch than to update bcftools.]
-
-As a general principle, data is only retrieved from the Zarr store
-when it is required to fulfil a particular query. For example,
-when a subset of samples are specified, we only retrieve chunks
-in arrays such as \texttt{call\_genotype} for those
-samples. Chunks in the variants dimension are retrieved as needed,
-and we use a multithreaded strategy to fetch and decode variant chunks
-while another thread encodes VCF. Generating VCF text at speed
-is a challenging problem, and we use a C extension module to convert
-a given set of Zarr chunks represented as numpy arrays to VCF.
-Based on the \texttt{region\_index} array computed by \texttt{vcf2zarr},
-we use PyRanges~\citep{stovner2020pyranges} to perform efficient interval
-search queries.
-
 \subsection{Benchmarks}
@@
 ARG simplification~\cite{kelleher2018efficient,wong2024general},
-storing the subsets in \texttt{tskit} format~\cite{tskit2024}.
+storing the subsets in \texttt{tskit} format~\cite{ralph2020efficiently,tskit2024}.
 Note that this procedure captures the growth
@@
 dual CPU (Intel Xeon E5-2680 v2)
-server with 256GiB of RAM running Debian GNU/Linux 11.
+server with 256 GiB of RAM running Debian GNU/Linux 11.
 To ensure that the true effects of having data distributed over a large
@@
 \subsection{AWS parallel Zarr prototype}
-[TODO Implementation of prototype Zarr implementaion]
+This section describes the methodology used for the AWS parallel Zarr
+throughput measurements in Fig~\ref{fig-s3-throughput}
+and Fig~\ref{fig-s3-network-throughput}, and timing measurements
+detailed in the cloud computing results.
+These measurements were run on EC2 instances using Jupyter notebooks
+hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
+were hosted in a standard profile S3 bucket. Both instance and bucket were
+in the same availability zone in the eu-west-2 region. Each
+of the parallel Python processes used the aioboto3 library (version 13.2.0) to
+fetch chunks concurrently from S3.
+% The underlying aiohttp library
+% limits the concurrent fetching to 100 requests per process. 
+For decompressing
+and computing af-dist the same architecture was used but as fetched chunks
+became available in memory in each process, they were decompressed and
+processed synchronously. Decompression was performed by using the 
+numcodecs library~\citep{numcodecs}.
+
+This prototype did not use any existing Zarr implementation as 
+the levels of concurrency involved require specialised 
+asynchronous programming techniques that are not widely supported.
+The fact that such a high-performance prototype can be implemented
+in a notebook with no Zarr-specific library support is a testament to the 
+transparent simplicity of the Zarr format.
 
@@
 \url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
-All source code for running benchmarks, analyses and creating
+The SAIGE prototype is available at \url{https://github.com/Will-Tyler/SAIGE}
+and all other code for running benchmarks, analyses and creating
 plots in this article is available at
@@
 is freely available under the terms of the Apache 2.0
-and can be installed from the Python Package Index
+license and can be installed from the Python Package Index
 (\url{https://pypi.org/project/vcztools/}).
@@
 \section{Acknowledgements}
-We would like to thank Gil McVean and Ben Neale for helpful discussions and
-comments on the manuscript.
+We are grateful to Augusto Rendon and the Genomics 
+England support team for facilitating our cloud computing experiments,
+and to Gil McVean and Ben Neale for helpful discussions and comments on the 
+manuscript.
+
 This research was made possible through access to data in the National Genomic
@@
 Elapsed time is also reported (dotted line).
-This corresponds to a maximum rate of 1.2GiB/s for Zarr (Zstd + BitShuffle),
+This corresponds to a maximum rate of 1.2 GiB/s for Zarr (Zstd + BitShuffle),
 2.7 GiB/s Zarr (Zstd), 2.9 GiB/s Zarr (LZ4), and 6.6 GiB/s for Savvy.
@@
 Elapsed time is also reported (dotted line).
-These benchmarks were run on an 8-core CPU (Intel i7-9700) with 32GiB RAM running
+These benchmarks were run on an 8-core CPU (Intel i7-9700) with 32 GiB RAM running
 Linux Mint 21.3 with data on an NVMe SSD.
@@
 \caption{Scalability of genotype data processing on AWS.
+Using the prototype parallel Zarr implementation (see Methods) we 
+measured the rate genotype data can be decoded to memory (decompression
+only) and the rate at which we can perform the full af-dist
+calculation as we vary the number of parallel processes.
 \label{fig-s3-throughput}}
@@
 \caption{Chunk data download rate on AWS.
+Using the prototype parallel Zarr implementation (see Methods) we 
+measured the rate at which chunks of compressed genotype data 
+can be downloaded on two different instance types.
 \label{fig-s3-network-throughput}}
@@
 \begin{figure}[h]
-\begin{tabular}{ccc}
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_LRR.png}&
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_BAF.png}&
-\includegraphics[width=2.5cm]{ofh_example/plots/default_call_GS.png}
-\end{tabular}
-\caption{Distribution of floating point fields in the Our Future Health data.
+\includegraphics{figures/ofh-histograms}
+\caption{Distribution of values in the GS, LRR and BAF fields in the Our Future
+Health data with no truncation (Default), and truncation using the BitRound
+and Quantize filters.
 \label{fig-ofh-field-distributions}}

### 2025-01-28 paper.tex
Shuffle Jeff and Eric's affils

--- 2025-01-28 paper.tex
@@
 % First author
-\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
+\author[1,\authfn{1},\authfn{3}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
 \author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
@@
 % Senior
-\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
-
-\affil[1]{Related Sciences}
+\author[1,\authfn{2},\authfn{3}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
+\author[10,\authfn{2},\authfn{4}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+
+\affil[1]{Open Athena AI Foundation}
 \affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
@@
 \authnote{\authfn{2}Joint senior author.}
-\authnote{\authfn{3}jerome.kelleher@bdi.ox.ac.uk}
+\authnote{\authfn{3}Work partially performed at Related Sciences.}
+\authnote{\authfn{4}jerome.kelleher@bdi.ox.ac.uk}
 
@@
 To explore the use of cloud computing services with VCF Zarr,
-we uploaded the Zarr directory hierarchy for the Genomics England
-data chromosome 2 described in Table~\ref{tab-genomics-england-data}
-to an AWS S3 standard bucket. 
-We repeated the benchmarks above on the same subset
+Genomics England provided access to data within their secure
+cloud environment. The VCF Zarr files for Chromosome 2
+described in Table~\ref{tab-genomics-england-data}
+were hosted in a standard AWS S3 bucket.
+
+Using this data we repeated the benchmarks above on the same subset
 of 562,640 variants and 78,195 samples using the same
@@
 \subsection{AWS parallel Zarr prototype}
-[TODO Implementation of prototype Zarr implementaion]
+This section describes the methodology used for the AWS parallel Zarr
+throughput measurements in Fig~\ref{fig-s3-throughput}
+and Fig~\ref{fig-s3-network-throughput}, and timing measurements
+detailed in the cloud computing results.
+These measurements were run on EC2 instances using jupyter notebooks
+hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
+were hosted in a standard profile S3 bucket. Both instance and bucket were
+in the same availability zone in the eu-west-2 region. Each
+of the parallel Python processes used the aioboto3 library (version 13.2.0) to
+fetch chunks concurrently from S3. The underlying aiohttp library
+limits the concurrent fetching to 100 requests per process. For decompressing
+and computing af-dist the same architecture was used but as fetched chunks
+became available in memory in each process, they were decompressed and
+computed sequentially. Decompression was performed by the blosc library
+without the use of zarr.
 

### 2025-01-28 paper.tex
Decent discussion of local alleles
Closes #186
Closes #187
￼
￼

--- 2025-01-28 paper.tex
@@
 
-\subsection{Local alleles}
-[TODO Implementation of local alleles in vcf2zarr]
+The implementation of the local alleles data reduction technique 
+in \texttt{vcf2zarr} follows the approach of Poterba 
+et al~\cite{poterba2025scalable}.
+Specifically, we identify the alleles in use for each call by
+examining the genotypes and then store the LA field,
+listing the (maximum of ploidy) unique alleles observed as indexes into the 
+global list for that variant.
+Using the LA field we then compute derived fields such as 
+AD and PL in terms of the local alleles during Zarr encoding
+(see e.g.\ the \texttt{call\_LAD} field in Table~\ref{tab-aou-data}
+and
+\texttt{call\_LPL} field in Table~\ref{tab-spruce-data}).
+Note that our approach (along with Poterba et al)
+differs from the details in the VCF 4.5 specification~\citep{vcf4_5},
+which defines the LAA field containing indexes of ALT 
+alleles only. Explicitly listing the alleles observed leads
+to simpler code and means we do not need to refer back to the 
+genotypes to distinguish missing data from homozygous reference 
+calls. Note also that the local alleles encoding may be lossy,
+as only information relevant to the chosen alleles for 
+each call is retained.
 

### 2025-01-28 paper.tex
Pass through AoU data

@@
 
-[TODO discuss local alleles and mention Table~\ref{tab-aou-data}].
-
-[Discuss why we're bigger than VCF here, people will want to understand.
+Local alleles (see Methods) is essential in this example
+as the maximum number of alleles observed at a site is 95.
+Thus, as defined, the AD field field has an inner dimension of 95
+each $1000\times10000$ chunk 
+% >>> 1000 * 10_000 * 95 * 2 / 1024**3
+% 1.7695128917694092
+has an uncompressed size of 1.77GiB (in this 16 bit integers are required
+to store the values). Given that the vast majority
+of this encoding consists of the dimension padding sentinel (-2, see 
+Methods), it is an inefficient means of representing and processing 
+allele depth data. Translating to the local alleles representation
+here reduces the inner dimension to 2, and the uncompressed 
+% >>> 1000 * 10_000 * 2 * 2 / 1024**2
+% 38.14697265625
+size of each chunk to 38.15MiB. 
+Note that there is also no loss of information in this case 
+by encoding as local alleles, as the per-sample gVCFs used 
+in earlier stages in the All of Us pipeline were truncated
+to a maximum of two ALT alleles per site.
+
+[Discussion of the sparsity of the QC field data, and how this 
+explains the difference to the VCF compression.]
 

### 2025-01-28 paper.tex
Fixup affils

@@
 % First author
-\author[1,\authfn{1},\authfn{3}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
-\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
-\author[4,\authfn{1}]{Will Tyler}
-\author[5,\authfn{1}]{Tom White}
+\author[1,2\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
+\author[3,4\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
+\author[5,\authfn{1}]{Will Tyler}
+\author[6,\authfn{1}]{Tom White}
 
 % Middle, alphabetical order
-\author[6]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
-\author[7,8]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
-\author[9]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
-\author[10]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[7,8,11]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
-\author[12]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
-\author[13]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
-\author[14]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
+\author[7]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
+\author[8,9]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
+\author[10]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
+\author[11]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[8,9,12]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
+\author[13]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
+\author[14]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
+\author[15]{Per Unneberg} % https://orcid.org/0000-0001-5735-3315
 \author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
-\author[15]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
+\author[16]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284
 
 % Senior
-\author[1,\authfn{2},\authfn{3}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[10,\authfn{2},\authfn{4}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[1,2,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
+\author[11,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
 \affil[1]{Open Athena AI Foundation}
-\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
+\affil[2]{Related Sciences}
+\affil[3]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
 New Zealand}
-\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
-\affil[4]{Independent researcher}
-\affil[5]{Tom White Consulting Ltd.}
-\affil[6]{Our Future Health, Manchester, UK.}
-\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
+\affil[4]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
+\affil[5]{Independent researcher}
+\affil[6]{Tom White Consulting Ltd.}
+\affil[7]{Our Future Health, Manchester, UK.}
+\affil[8]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[9]{Analytic and Translational Genetics Unit, Massachusetts General
 Hospital, Boston, Massachusetts 02114, USA}
-\affil[9]{NVIDIA Ltd, Reading, UK}
-\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
+\affil[10]{NVIDIA Ltd, Reading, UK}
+\affil[11]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
 University of Oxford, UK}
-\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[12]{Wellcome Sanger Institute}
-\affil[13]{Genomics England}
-\affil[14]{Department of Cell and Molecular Biology, National
+\affil[12]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[13]{Wellcome Sanger Institute}
+\affil[14]{Genomics England}
+\affil[15]{Department of Cell and Molecular Biology, National
   Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
   Uppsala University, Uppsala, Sweden}
-\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}
+\affil[16]{School of Computer Science, McGill University, Montreal, QC, Canada}
 
@@
 \authnote{\authfn{2}Joint senior author.}
-\authnote{\authfn{3}Work partially performed at Related Sciences.}
-\authnote{\authfn{4}jerome.kelleher@bdi.ox.ac.uk}
+\authnote{\authfn{3}jerome.kelleher@bdi.ox.ac.uk}
 

### 2025-01-28 paper.tex

--- 2025-01-28 paper.tex
@@
 % First author
-\author[1,\authfn{1},\authfn{3}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
+\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
 \author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
@@
 % Senior
-\author[1,\authfn{2},\authfn{3}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[10,\authfn{2},\authfn{4}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
-
-\affil[1]{Open Athena AI Foundation}
+\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
+\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+
+\affil[1]{Related Sciences}
 \affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
@@
 quadratically on the number of alleles.
-While the local alleles representation~\cite{poterba2025scalable}
+While the local alleles representation~\cite{poterba2024scalable}
 is a substantial improvement (see All of Us and Norway Spruce
@@
 is padded with a sentinel value (-2 for integers and a specific
-non-signalling NaN for floats). This is problematic in situations
-where some sites have a large number of alleles (e.g., 95 in the All
-of Us case study above), and 
-in particular for fields that have a quadratic dependency on the number of alleles
-(Number=G) such as PL. We use the local alleles
-encoding~\cite{poterba2025scalable} to reduce the storage 
-requirements for such fields (see the next section for 
-implementation details).
+non-signalling NaN for floats). While this is not a problem
+in practice for datasets in which all four bases are observed,
+it is a substantial issue for
+fields that have a quadratic dependency on the number of alleles
+(Number=G) such as PL. Such fields are already known to cause
+significant problems, and the ``local alleles'' proposal
+provides an elegant solution~\cite{poterba2024scalable}.
+As this approach is on a likely path to
+standardisation~\cite{danecek2021twelve}, we plan to
+include support in later versions of VCF Zarr.
 
@@
 
-The implementation of the local alleles data reduction technique 
-in \texttt{vcf2zarr} follows the approach of Poterba 
-et al~\cite{poterba2025scalable}.
-Specifically, we identify the alleles in use for each call by
-examining the genotypes and then store the LA field,
-listing the (maximum of ploidy) unique alleles observed as indexes into the 
-global list for that variant.
-Using the LA field we then compute derived fields such as 
-AD and PL in terms of the local alleles during Zarr encoding
-(see e.g.\ the \texttt{call\_LAD} field in Table~\ref{tab-aou-data}
-and
-\texttt{call\_LPL} field in Table~\ref{tab-spruce-data}).
-Note that our approach (along with Poterba et al)
-differs from the details in the VCF 4.5 specification~\citep{vcf4_5},
-which defines the LAA field containing indexes of ALT 
-alleles only. Explicitly listing the alleles observed leads
-to simpler code and means we do not need to refer back to the 
-genotypes to distinguish missing data from homozygous reference 
-calls. Note also that the local alleles encoding may be lossy,
-as only information relevant to the chosen alleles for 
-each call is retained.
+\subsection{Local alleles}
+[TODO Implementation of local alleles in vcf2zarr]
 
@@
 examined the effects of varying chunk sizes and compressor settings
-on the compression ratio for call-level fields. (
-We excluded \texttt{call\_PL} from this analysis as this analysis predated 
-our implementation of local alleles.)
+on the compression ratio for call-level fields. We excluded call\_PL
+from this analysis as it requires conversion to a ``local alleles''
+encoding~\citep{poterba2024scalable} to be efficient,
+which is planned for implementation in a future
+version of \texttt{vcf2zarr}.
 
@@
 \subsection{AWS parallel Zarr prototype}
-This section describes the methodology used for the AWS parallel Zarr
-throughput measurements in Fig~\ref{fig-s3-throughput}
-and Fig~\ref{fig-s3-network-throughput}, and timing measurements
-detailed in the cloud computing results.
-These measurements were run on EC2 instances using jupyter notebooks
-hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
-were hosted in a standard profile S3 bucket. Both instance and bucket were
-in the same availability zone in the eu-west-2 region. Each
-of the parallel Python processes used the aioboto3 library (version 13.2.0) to
-fetch chunks concurrently from S3. The underlying aiohttp library
-limits the concurrent fetching to 100 requests per process. For decompressing
-and computing af-dist the same architecture was used but as fetched chunks
-became available in memory in each process, they were decompressed and
-computed sequentially. Decompression was performed by the blosc library
-without the use of zarr.
+[TODO Implementation of prototype Zarr implementaion]
 

### 2025-01-28 paper.tex
Updates

--- 2025-01-28 paper.tex
@@
 To explore the use of cloud computing services with VCF Zarr,
-Genomics England provided access to data within their secure
+Genomics England provided access to Aggv2 data 
+in VCF Zarr format within their secure
 cloud environment. The VCF Zarr files for Chromosome 2
 described in Table~\ref{tab-genomics-england-data}
-were hosted in a standard AWS S3 bucket.
-
+were copied to a standard AWS S3 bucket.
 Using this data we repeated the benchmarks above on the same subset
@@
 The maximum decompression rate of 47.4GiB/s
-occurs at 64 processes, with a corresponding
-network data rate of 121 MiB/s
+occurs at 64 processes (each with 100 concurrent requests), 
+with a corresponding network data rate of 121 MiB/s
 Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
@@
 detailed in the cloud computing results.
-These measurements were run on EC2 instances using jupyter notebooks
+These measurements were run on EC2 instances using Jupyter notebooks
 hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
@@
 of the parallel Python processes used the aioboto3 library (version 13.2.0) to
-fetch chunks concurrently from S3. The underlying aiohttp library
-limits the concurrent fetching to 100 requests per process. For decompressing
+fetch chunks concurrently from S3.
+% The underlying aiohttp library
+% limits the concurrent fetching to 100 requests per process. 
+For decompressing
 and computing af-dist the same architecture was used but as fetched chunks
 became available in memory in each process, they were decompressed and
-computed sequentially. Decompression was performed by the blosc library
-without the use of zarr.
+processed syncronously. Decompression was performed by using the 
+numcodecs library~\citep{numcodecs}.
+
+This prototype did not use any existing Zarr implementation as 
+the levels of concurrency involved require specialised 
+asynchronous programming techniques that are not widely supported.
+The fact that such a high-performance prototype can be implemented
+in a notebook with no Zarr-specific library support is a testament to the 
+transparent simplicity of the Zarr format.
 

### 2025-01-28 paper.tex

@@
 totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
-distributed over 165 VCF files. Since the dataset originally was
-generated for downstream analyses that require genotype likelihoods,
-the genotypes include the PL field.
-
-Following the conversion process detailed in the previous section, we
-first converted the 165 VCF files (7.33TiB) to ICF, followed by
+distributed over 165 VCF files. 
+% Since the dataset originally was
+% generated for downstream analyses that require genotype likelihoods,
+% the genotypes include the PL field.
+
+We first converted the 165 VCF files (7.33TiB) to ICF, followed by
 conversion to Zarr. The tasks were run as single jobs on a compute
@@
 required 13h45min runtime. The ICF representation used a total of 6.77
-TiB over 1,340,804 data storage files. Conversion to Zarr required
-18h36min runtime, generating a dataset with 32 arrays, consuming a
-total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
-total storage (1.1X) is negligible, mainly due to the inherent
-difficulty in compressing the PL field. The output dataset storage
-size is dominated by a major component, call\_PL, which accounts for
-nearly 93\% of the total. call\_PL is yet another field with low
-compression potential due to inherent noisiness.
-
-Therefore, to reduce storage requirements we generated a Zarr schema
-for encoding that uses local allele fields with \texttt{vcf2zarr
+TiB over 1,340,804 data storage files. Conversion to Zarr 
+using the default schema and a chunk size of 10,000 variants
+$\times$ 1,063 samples
+required 18h36min runtime, generating a dataset with 34 arrays, 
+and required a total of 6.6 TiB storage over 10,861,038 chunk files. 
+Over 90\% of this was used by the call\_PL field, mostly due to 
+the maximum number of alleles at a site being 4,
+and consequently the inner dimension of call\_PL 10 because 
+of its quadratic dependency on the number of alleles (see Methods).
+
+We therefore generated a vcf2zarr schema
+using local allele fields with \texttt{vcf2zarr
 mkschema --local-alleles}, and used the generated schema as input to
@@
 generating a dataset with 35 arrays, consuming a total of 2.3TiB
-storage over 11,235,556 chunk files, with a reduction in storage of
-3.2X. Therefore, by using local allele fields we achieve approximately
-3X improvement of compression. The top fields in terms of storage are
-detailed inTable~\ref{tab-spruce-data}.
+storage over 11,235,556 chunk files. 
+Using local allele fields in this example we therefore 
+achieve approximately 3X improvement of compression. 
+The top fields in terms of storage are
+detailed in Table~\ref{tab-spruce-data}.
 

### 2025-01-29 paper.tex

@@
 datasets for Norway Spruce ($n$=1,063) and 
-SARS-Cov-2 ($n$=3,960,704), % TODO update for tranche 2 data 
+SARS-CoV-2 ($n$=3,960,704), % TODO update for tranche 2 data 
 demonstrating the scalability and flexibility of VCF Zarr.
@@
 and in a large collection of short whole genome alignments
-using SARS-Cov-2 data 
+using SARS-CoV-2 data 
 (100X % FIXME! this is roughly correct, but need to get the numbers straight

### 2025-01-29 paper.tex
Spelling, clarity and consistency issues

--- 2025-01-28 paper.tex
@@
 % ==42.3 GiB - assuming the G is power-of-two here
-10,221 variants and 42.3GiB of storage. We first converted the VCF
+10,221 variants and 42.3 GiB of storage. We first converted the VCF
 data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
-The ICF representation required a total of 41GiB over XX files.
+%TODO There is an XX HERE
+The ICF representation required a total of 41 GiB over XX files.
 We then converted to VCF Zarr format using the default schema, which resulted
-in a 41GiB VCZ. On inspection, this was dominated by three floating point
+in a 41 GiB VCZ. On inspection, this was dominated by three floating point
 fields LRR, BAF and GS stored
@@
 $1.88\times10^{-6}$ (GS).
-Using the Bitround filter with a parameter of 5 bits reduced the overall storage for the Zarr to
+Using the BitRound filter with a parameter of 5 bits reduced the overall storage for the Zarr to
 XX with an MAE of
@@
 \caption{Summary for VCF Zarr conversion of 
-All of Us exome-like data genotype data for chromosome 22 (245,394 samples, 
+%TODO Text says chr 22, this caption says chr 20, I'm not sure which is right
+All of Us exome-like data genotype data for chromosome 20 (245,394 samples, 
 715,256 variants) 
-consisting of 25 fields and 8.5GiB of storage ($\sim1.1$X larger
+consisting of 25 fields and 8.5 GiB of storage ($\sim1.1$X larger
 % 8.5 / 7.44 = 1.142
@@
 % 1.7695128917694092
-has an uncompressed size of 1.77GiB (in this 16 bit integers are required
+has an uncompressed size of 1.77 GiB (in this 16 bit integers are required
 to store the values). Given that the vast majority
@@
 % 38.14697265625
-size of each chunk to 38.15MiB. 
+size of each chunk to 38.15 MiB. 
 Note that there is also no loss of information in this case 
@@
 
-We first converted the 165 VCF files (7.33TiB) to ICF, followed by
+We first converted the 165 VCF files (7.33 TiB) to ICF, followed by
 conversion to Zarr. The tasks were run as single jobs on a compute
@@
 \texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
-generating a dataset with 35 arrays, consuming a total of 2.3TiB
+generating a dataset with 35 arrays, consuming a total of 2.3 TiB
 storage over 11,235,556 chunk files. 
@@
 \begin{table}
-\caption{Summary for VCF Zarr conversion of the Norway spruce
+\caption{Summary for VCF Zarr conversion of the Norway Spruce
+%TODO The variants number here does not match the text "3.8 million"
 WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
-consisting of 34 fields and 2.3TiB of storage ($\sim 3.2$X smaller 
+consisting of 34 fields and 2.3 TiB of storage ($\sim 3.2$X smaller 
 than source gzipped VCFs). The chunk size is 10,000 variants $\times$
-1,063 samples (i..e, a single sample chunk). This uses local 
+1,063 samples (i.e., a single sample chunk). This uses local 
 alleles to reduce the 
@@
 encoding information about those samples. This full dataset is stored in a
-single 345MiB Zip file, 241MiB of which is used by the alignment data. This
+single 345 MiB Zip file, 241 MiB of which is used by the alignment data. This
 gives an average compression ratio of 468X.
@@
 average number of deletions in samples with a Scorpio designation of Alpha
+%TODO "x" on next line
 required x seconds and 5 lines of code using Dask.
@@
 and Zip archives, and support for major cloud stores will be added in the near
+%TODO Resolve this note
 future. [JK: note this is a pretty simple feature, maybe we should just try
@@
 API~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
-above (3.73 GiB) this took 31s, giving a parse rate of 123.1 MiB/s
+above (3.73 GiB) this took 31s, giving a parsing rate of 123.1 MiB/s
 (almost identical to the write-rate of vcztools above). Given this limiting
@@
 
-To test the performance of this implemenentation we compared the time required
+To test the performance of this implementation we compared the time required
 to perform an association test using the simulated dataset of 10,000 samples
@@
 
-We compared the same analyis using the existing VCF, BCF and Savvy backends
+We compared the same analysis using the existing VCF, BCF and Savvy backends
 to our prototype Zarr backend.
@@
 an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
-The maximum decompression rate of 47.4GiB/s
+The maximum decompression rate of 47.4 GiB/s
 occurs at 64 processes (each with 100 concurrent requests), 
@@
 Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
-calculation is performed at a maximum rate of 25.5GiB/s using 128
+calculation is performed at a maximum rate of 25.5 GiB/s using 128
 processes. At this rate, the af-dist calculation for the target region
@@
 became available in memory in each process, they were decompressed and
-processed syncronously. Decompression was performed by using the 
+processed synchronously. Decompression was performed by using the 
 numcodecs library~\citep{numcodecs}.

### 2025-01-29 paper.tex

@@
 \subsection{Interoperability: SAIGE proof-of-concept}
-To demonstrate the potential for integrating VCF Zarr into existing utilities,
+To demonstrate the potential for integrating VCF Zarr into existing
+high-performance analysis tools,
 we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
 zhou2020scalable}, a popular package for performing genome wide
-association tests. Using the Tensorstore C++ library~\cite{tensorstore}
-we added support for VCF Zarr via an additional backend class [some details
-about the implementation].
-
-To test the performance of this implementation we compared the time required
+association tests. 
+Using the TensorStore C++ library~\cite{tensorstore}
+we added support for VCF Zarr via an additional backend 
+(SAIGE reading data from PLINK, BGEN, VCF, BCF and SAV formats) 
+for which the core functionality is defined in a 264 line C++ file.
+
+To assess the performance of this implementation we compared the time required
 to perform an association test using the simulated dataset of 10,000 samples
-from Fig~\ref{fig-data-storage}. We generated a phenotype using
-\texttt{tstrait}~\citep{tagami2024tstrait}.
+from Fig~\ref{fig-data-storage}. We performed a single-variant association
+test based on a phenotype simulated using
+\texttt{tstrait}~\citep{tagami2024tstrait},
+and measure the overall time required when using the 
+equivalent VCF, BCF, SAV and Zarr files.
 

### 2025-01-29 paper.tex
Tweaks to OFH section

@@
 
-Release 9~\citep{ofhdatarelease9}
-was made available in December 2024. The genetic component includes genotype data
-from 651,050 individuals on 707,522 variants. Here we use the genotype data from
-chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
-three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling
-% 20G     ofh.chr22-b0.vcf.gz
-% 19G     ofh.chr22-b1.vcf.gz
-% 3.3G    ofh.chr22-b2.vcf.gz
-% ==42.3 GiB - assuming the G is power-of-two here
-10,221 variants and 42.3 GiB of storage. We first converted the VCF
+Conversion was performed on a Standard\_D32s\_v5 Azure VM instance, with 
+32 CPUs and 128 GB RAM. We first converted the VCF
 data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
-%TODO There is an XX HERE
-The ICF representation required a total of 41 GiB over XX files.
+The ICF representation required a total of 41 GiB of storage.
 We then converted to VCF Zarr format using the default schema, which resulted
-in a 41 GiB VCZ. On inspection, this was dominated by three floating point
-fields LRR, BAF and GS stored
+in a 41 GiB VCZ over 4514 chunk files. 
+On inspection, this was dominated by three floating point
+fields LRR (Log R Ratio), BAF (B allele frequency) and GS (GenCall Score)
+stored
 to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
@@
 Using the BitRound filter with a parameter of 5 bits reduced the overall storage for the Zarr to
-XX with an MAE of
+17 GiB with an MAE of
 % 0   call_LRR    bitround    6.93e-04
@@
 \caption{Summary for VCF Zarr conversion of 
-%TODO Text says chr 22, this caption says chr 20, I'm not sure which is right
 All of Us exome-like data genotype data for chromosome 20 (245,394 samples, 

### 2025-01-29 paper.tex
Finalise summary in OFH section
Closes  #250

@@
 apply the BitRound(5) filter to the large floating point fields (see text for 
-details).
-For column details, see Table~\ref{tab-genomics-england-data}.
+details). Shown are the top 7 fields consuming at 0.001\% or more of the 
+overall storage (see Table~\ref{tab-genomics-england-data} for column details).
 \label{tab-ofh-data}}
@@
 \midrule
-/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
-/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
-/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
-/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
-/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
-/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
-/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
-/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
-/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
-/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
-/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
-/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
-/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
-/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
-/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
-/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
-/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
+call\_LRR & float32 & 8.35 GiB & 3.0 & 52.08\% \\
+call\_BAF & float32 & 6.02 GiB & 4.1 & 37.55\% \\
+call\_GS & float32 & 961.37 MiB & 26.0 & 5.86\% \\
+call\_genotype & int8 & 657.21 MiB & 19.0 & 4.00\% \\
+call\_genotype\_mask & bool & 79.48 MiB & 160.0 & 0.48\% \\
+sample\_id & str & 2.93 MiB & 1.7 & 0.02\% \\
+call\_genotype\_phased & bool & 644.72 KiB & 10000.0 & 0.00\% \\
 \bottomrule
@@
 from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
-Table~\ref{tab-ofh-data} shows a summary of the Zarr using the BitRound filter.
+Table~\ref{tab-ofh-data} shows a summary of the Zarr dataset using the 
+BitRound filter.
 

### 2025-01-29 paper.tex
Make data tables systematic

@@
 apply the BitRound(5) filter to the large floating point fields (see text for 
-details). Shown are the top 7 fields consuming at 0.001\% or more of the 
-overall storage (see Table~\ref{tab-genomics-england-data} for column details).
+details). 
+Shown are the top 6 fields consuming at least 0.01\% of the overall storage 
+(see Table~\ref{tab-genomics-england-data} for column details).
 \label{tab-ofh-data}}
@@
 sample\_id & str & 2.93 MiB & 1.7 & 0.02\% \\
-call\_genotype\_phased & bool & 644.72 KiB & 10000.0 & 0.00\% \\
 \bottomrule
@@
 space required by the PL field; see the text for discussion.
-For column details see the caption of
-Table~\ref{tab-genomics-england-data}.
+Shown are the top 21 fields consuming at least 0.01\% of the 
+overall storage (see Table~\ref{tab-genomics-england-data} for column details).
 \label{tab-spruce-data}}
-
 \begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
 \toprule
-{Field} & {type} & {storage} & {compress} & {\%total} \\
+{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
 \midrule
-/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
-/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
-/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
-/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
-/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
-/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
-/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
-/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
-/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
-/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
-/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
-/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
-/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
-/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
-/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
-/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
-/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
-/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
-/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
-/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
-/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
-/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
-% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
-% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
-% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
-% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
-% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
-% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
-% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
-% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
-% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
-% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
-% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
+call\_LPL & int16 & 1.26 TiB & 17.0 & 54.76\% \\
+call\_LA & int8 & 415.39 GiB & 18.0 & 17.63\% \\
+call\_genotype & int8 & 381.28 GiB & 19.0 & 16.18\% \\
+call\_genotype\_mask & bool & 103.84 GiB & 71.0 & 4.41\% \\
+variant\_DP4 & int32 & 26.24 GiB & 2.1 & 1.11\% \\
+variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56\% \\
+variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55\% \\
+variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55\% \\
+variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55\% \\
+variant\_quality & float32 & 12.84 GiB & 1.1 & 0.55\% \\
+variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.55\% \\
+variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54\% \\
+variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.51\% \\
+variant\_position & int32 & 9.65 GiB & 1.4 & 0.41\% \\
+variant\_AC & int16 & 7.42 GiB & 2.8 & 0.31\% \\
+variant\_DP & int32 & 6.71 GiB & 2.1 & 0.28\% \\
+variant\_allele & str & 5.37 GiB & 21.0 & 0.23\% \\
+variant\_AN & int16 & 3.05 GiB & 2.3 & 0.13\% \\
+call\_genotype\_phased & bool & 1.75 GiB & 2100.0 & 0.07\% \\
+variant\_filter & bool & 1.46 GiB & 2.4 & 0.06\% \\
+variant\_MQ & int8 & 489.41 MiB & 7.3 & 0.02\% \\
 \bottomrule

### 2025-01-29 paper.tex
Closes #255 ￼

@@
 $3.38\times10^{-3}$ (GS).
-The histograms of the truncated fields in all cases were visually indistinguishable
-from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
+The histograms of the truncated fields in all cases were very similar to the
+to the original (Fig~\ref{fig-ofh-field-distributions}).
 Table~\ref{tab-ofh-data} shows a summary of the Zarr dataset using the 

### 2025-01-29 paper.tex

@@
 datasets for Norway Spruce ($n$=1,063) and 
-SARS-CoV-2 ($n$=3,960,704), % TODO update for tranche 2 data 
+SARS-CoV-2 ($n$=4,384,310), 
 demonstrating the scalability and flexibility of VCF Zarr.
@@
 using SARS-CoV-2 data 
-(100X % FIXME! this is roughly correct, but need to get the numbers straight
-smaller than gzipped FASTA). 
+(81X smaller than gzipped FASTA). 
 Interoperability with the vast ecosystem of tools
@@
 is that the \texttt{variant\_alleles} field contains the REF and ALT
-alleles as observed (such that the zeroth allele is always the reference).
+alleles as observed (such that the zero'th allele is always the reference).
 In this case it is more convenient to encode the alleles as the original
-bases and IUPAC uncertainty codes, such that the allele 0 is always A, 1 is C etc.
-Even in this case with very high sequence diversity, uncertainty codes and
-storing data for the entire genome, compression rates are very high.
-
-Based on the batch 1 release of Viridian, the Zarr dataset contains
-3,960,704 sequences of length 29903 base pairs along with 31 metadata columns
-encoding information about those samples. This full dataset is stored in a
-single 345 MiB Zip file, 241 MiB of which is used by the alignment data. This
-gives an average compression ratio of 468X.
-
-Computational accessibility is excellent. For example, computing the number of
-samples with missing or ambiguous data at each site requires around 3 minutes,
-iterating over sample chunks sequentially (5 lines of Python).
-Retrieving the full alignment for a given sample requires 163ms and
-getting the calls for all samples at a given site 209ms. [TODO confirm hardware
-details]. The storage of metadata and genetic data in a single file also
+bases and IUPAC uncertainty codes~\citep{cornish1985nomenclature}, 
+such that the allele 0 is always A, 1 is C etc.
+Other than these minor deviations in interpretation, the Zarr follows
+the VCF Zarr specification and can be processed using the same tools.
+
+\begin{table}
+\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
+whole genome alignments and metadata 
+(4,384,310 samples, 29903 variants)
+consisting of 37 fields and 391 MiB of storage, 
+of which 280MiB is used by aligments ($\sim 81$X smaller 
+than gzipped FASTA). The chunk size is 100 variants $\times$
+10,000 samples.
+Shown are the top 9 fields consuming at least 1\% of the 
+overall storage (see Table~\ref{tab-genomics-england-data} for column details).
+Field names (derived from source CSV file) are shortened for 
+display purposes by replacing ``Genbank'' with ``G''
+and ``Viridian'' with ``V''.
+\label{tab-sc2-data}}
+\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
+\toprule
+{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
+\midrule
+call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
+sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
+sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
+sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
+sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
+sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
+sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
+sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
+sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Decompressed, the Viridian dataset consists of 48
+FASTA files totaling 125 GiB of storage and a 1.4 GiB
+TSV file for sample metadata.
+When aligned to the reference and converted to Zarr, the
+entire dataset is stored in a single 391 MiB Zip file.
+When compressed with bgzip (so they can be indexed; see next 
+paragraph) the alignments require 22 GiB of storage.
+Table~\ref{tab-sc2-data} summarises the storage used by individual
+arrays, and shows that the alignment data takes up 298 MiB. 
+This is around 81X smaller than gzipped FASTA.
+The original dataset is distributed as xz compressed FASTA files, which
+achieves even higher compression levels at a total of 218 MiB.
+However, these files cannot be accessed in-situ and must be decompressed
+in order to access the data.
+
+Computational accessibility for the Zarr is excellent. 
+Retrieving the full alignment for the 1 millionth sample required
+[x ms]. Performing the same operation on the corresponding indexed FASTA
+file using pyfaidx~\citep{shirley2015efficient} required [x ms].
+Similarly, accessing calls for all samples 
+at the 10,000th site requires xms
+The storage of metadata and genetic data in a single file also
 makes many queries efficient and convenient. For example, computing the

### 2025-01-30 paper.tex
Update saige to 10_5

@@
 we added support for VCF Zarr via an additional backend 
-(SAIGE reading data from PLINK, BGEN, VCF, BCF and SAV formats) 
+(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats) 
 for which the core functionality is defined in a 264 line C++ file.
@@
 To assess the performance of this implementation we compared the time required
-to perform an association test using the simulated dataset of 10,000 samples
+to perform an association test using the simulated dataset of 100,000 samples
 from Fig~\ref{fig-data-storage}. We performed a single-variant association
@@
 to our prototype Zarr backend.
-[Times: BCF 3m, VCF6.5m, Savvy 39s, Zarr 55s]
-
-% BCF:
-% real 3m1.068s user 2m56.944s sys 0m2.941s
-
-% VCF:
-% real 6m30.335s user 6m25.193s sys 0m2.842s
-
-% Savvy:
-% real 0m39.660s user 0m37.498s sys 0m2.183s
-
-% Zarr:
-% real 0m55.957s user 1m10.375s sys 0m12.111s
+[Times: BCF 70m, VCF 162m, Savvy 7m, Zarr 20m]
+
+%BCF:
+%user   system  elapsed 
+%4195.218    4.671 4200.508
+%VCF:
+%user  system elapsed 
+%9718.428    4.735 9727.995 
+%Savvy:
+%user  system elapsed 
+%415.371   2.420 418.098 
+%Zarr:
+%user   system  elapsed 
+%1758.976  259.924 1223.767
 

### 2025-01-30 paper.tex
Final pass through SC2

@@
 bases and IUPAC uncertainty codes~\citep{cornish1985nomenclature}, 
-such that the allele 0 is always A, 1 is C etc.
+such that allele 0 is always A, 1 is C etc.
 Other than these minor deviations in interpretation, the Zarr follows
@@
 overall storage (see Table~\ref{tab-genomics-england-data} for column details).
-Field names (derived from source CSV file) are shortened for 
+Field names (derived from source TSV file) are shortened for 
 display purposes by replacing ``Genbank'' with ``G''
@@
 Table~\ref{tab-sc2-data} summarises the storage used by individual
-arrays, and shows that the alignment data takes up 298 MiB. 
-This is around 81X smaller than gzipped FASTA.
+arrays, and shows that the alignment data requires 298 MiB,
+around 81X smaller than gzipped FASTA.
 The original dataset is distributed as xz compressed FASTA files, which
 achieves even higher compression levels at a total of 218 MiB.
-However, these files cannot be accessed in-situ and must be decompressed
-in order to access the data.
-
-Computational accessibility for the Zarr is excellent. 
-Retrieving the full alignment for the 1 millionth sample required
-[x ms]. Performing the same operation on the corresponding indexed FASTA
-file using pyfaidx~\citep{shirley2015efficient} required [x ms].
-Similarly, accessing calls for all samples 
-at the 10,000th site requires xms
-The storage of metadata and genetic data in a single file also
-makes many queries efficient and convenient. For example, computing the
-average number of deletions in samples with a Scorpio designation of Alpha
-%TODO "x" on next line
-required x seconds and 5 lines of code using Dask.
+However, these files must be decompressed in order to access the data
+using standard FASTA processing techniques.
+
+Computational accessibility for the Zarr is excellent. For example,
+computing the total number of missing or ambiguous bases for 
+all 4.4 million samples required 55.5 seconds elapsed (6min 45s CPU)
+using 8 threads. This required 11 lines of code, using only the 
+Python standard library and NumPy. Similarly, computing allele
+counts at each site required 4min 53s (25min 35s CPU) using a slightly
+more complicated 16 line Python function.
+Existing methods for working with FASTA data provided efficient 
+access to individual alignments using the fai index format, which
+works with uncompressed or bgzip compressed files.
+For example, retrieving the alignment for the 1 millionth sample 
+from the corresponding chunk file requires only a few milliseconds
+using pyfaidx~\citep{shirley2015efficient} or 
+pysam~\citep{pysam}. Retrieving the same alignment using Zarr 
+required 365ms. 
+Accessing calls for all samples at at a given 
+site requires a similar time (here, accessing site 14,951
+required 276ms). 
+FASTA cannot support efficient retrieval
+of data for a particular site in a set of alignments.
+Storing data chunked in two dimensions means that we can avoid
+the classic choice between sample-major or variant-major storage,
+and access it efficiently in both dimensions.
 

### 2025-01-30 paper.tex
Update on SAIGE implementation
Closes #173

@@
 Vcztools currently supports Zarr stored in the local file system
-and Zip archives, and support for major cloud stores will be added in the near
-%TODO Resolve this note
-future. [JK: note this is a pretty simple feature, maybe we should just try
-to add it?]
+and Zip archives, and support for major cloud stores will be added in the near future. 
 Thus, users can convert their VCF data to Zarr using \texttt{bio2zarr},
@@
 it. To get an estimate of the rate at which VCF can be parsed, we wrote a simple
-program that reads a VCF file and parses it line-by-line using the \texttt{htslib}
-API~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
+C program that reads a VCF file and parses it line-by-line using 
+\texttt{htslib}~\citep{bonfield2021htslib}. 
+For the 10,000 variant subset discussed
 above (3.73 GiB) this took 31s, giving a parsing rate of 123.1 MiB/s
@@
 we added support for VCF Zarr via an additional backend 
-(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats) 
+(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats),
 for which the core functionality is defined in a 264 line C++ file.
+TensorStore provides efficient read and write access to array
+data in several formats, used, for example, in
+checkpointing~\cite{orbax-tensorstore} 
+TensorFlow model training~\citep{tensorflow2015-whitepaper}.
+It supports several cloud platforms, and is based on an
+asyncronous, parallel processing model (see Discussion).
 
@@
 to perform an association test using the simulated dataset of 100,000 samples
-from Fig~\ref{fig-data-storage}. We performed a single-variant association
+from Fig~\ref{fig-data-storage}. 
+We performed a single-variant association
 test based on a phenotype simulated using
 \texttt{tstrait}~\citep{tagami2024tstrait},
-and measure the overall time required when using the 
-equivalent VCF, BCF, SAV and Zarr files.
-
-We compared the same analysis using the existing VCF, BCF and Savvy backends
-to our prototype Zarr backend.
-[Times: BCF 70m, VCF 162m, Savvy 7m, Zarr 20m]
+and measured the overall time required when using the 
+equivalent VCF (162m), BCF (70m), SAV (7m) and Zarr (20m) files.
+Thus, in this case, substantially improved performance over the 
+most popular formats, along with native cloud-store support are 
+possible with relatively minor coding effort.
+We validated the implementation by comparing the association
+test results with the output of the other backends,
+confirming that they were numerically identical. 
+As TensorStore does not currently support string data for Zarr, 
+however,
+we were not able to fully replicate the output of fields 
+such as MarkerID, Allele1 etc.
 

### 2025-01-30 paper.tex
Drop separate vcztools methods section
Closes #256

@@
 
+Vcztools is implemented in Python and C, and is available for download 
+from the Python package index. 
+Based on the \texttt{region\_index} array computed by \texttt{vcf2zarr},
+we use PyRanges~\citep{stovner2020pyranges} to perform efficient interval
+search queries. 
+% While support direct support for VCF Zarr in \texttt{bcftools} would be the 
+% ideal outcome, implementation in C within \texttt{htslib} would be significant 
+% undertaking.
+
 \subsection{Interoperability: SAIGE proof-of-concept}

### 2025-01-30 paper.tex
Add region index text

@@
 
+To facilitate efficient region search queries, we compute an
+additional array \texttt{region\_index} which stores information
+about the range of genomic positions in each chunk
+and provides sufficient information to implement interval
+search queries. We can then load this index into memory and 
+use it to determine which variant chunks need to be retrieved.
+In the Norway Spruce example (Table~\ref{tab-spruce-data}),
+which has 3.75 billion variants,
+this index requires 3.14 MiB of storage over 12 chunk files
+to store information about all 
+374,518 variant chunks 
+(the \texttt{variant\_position} array, for comparison 
+is 13.95 GiB uncompressed).
+
 \subsection{Choosing default compressor settings}

### 2025-01-30 paper.tex
Fix spruce m number. 3.75 billion!

@@
 where resequencing data has been mapped to a chromosome-scale
-reference genome. The dataset consists of approximately 3.8 million
+reference genome. The dataset consists of approximately 3.75 billion
 single-nucleotide variants and small indels from the 12 autosomes
@@
 \caption{Summary for VCF Zarr conversion of the Norway Spruce
-%TODO The variants number here does not match the text "3.8 million"
-WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
+WGS data on all 12 autosomes (1,063 samples, 
+%3745170452
+3,745,170,452 variants)
 consisting of 34 fields and 2.3 TiB of storage ($\sim 3.2$X smaller 

### 2025-01-30 paper.tex

@@
 calculation is performed at a maximum rate of 25.5 GiB/s using 128
-processes. At this rate, the af-dist calculation for the target region
-was completed in 4.15s (compare to 5.23min for the naive single-threaded
-method above).
+processes. At this rate, the af-dist calculation for the target 
+% 180.80 MB in 4503 files
+% Total data: 90060.00 MB in 4503 chnnks
+region---172 MiB retrieved in 4503 chunks, 83.87 GiB of genotype data
+decompressed and processed---
+% NOTE: this doesn't quite up to the 25.5G above:  83.87 / 4.15 = 20.2
+was completed in 4.15s.
 This level of throughput would complete af-dist for all chromosomes for all samples

### 2025-01-30 paper.tex
Document X numbers for SAIGE

@@
 datasets for Norway Spruce ($n$=1,063) and 
-SARS-CoV-2 ($n$=4,384,310), 
-demonstrating the scalability and flexibility of VCF Zarr.
+SARS-CoV-2 ($n$=4,384,310).
 We demonstrate the potential for VCF Zarr to enable a new generation
 of high-performance and cost-effective applications
-via illustrative highly parallel benchmarks on AWS and experimentation with GPUs.
+via illustrative examples using cloud computing and GPUs.
 
@@
 
+% BCF: 4200.508 / 1223.767 = 3.43
+% VCF 9727.995 / 1223.767 = 7.95
+
 \subsection{Future applications: Cloud computing}

### 2025-01-30 response-to-reviewers.tex
Add reviewer 1s comments

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule}
                  % \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication.  We have addressed
the points you and Reviewer 1 raised below in turn.

\reviewersection
\section*{Reviewer 1 comments}

\begin{point}
The authors present VCF Zarr, a specification that translates the variant call
format (VCF) data model into an array-based representation for the Zarr storage
format. They also present the `vcf2zarr` utility to convert large VCFs to Zarr.
They provide data compression and analysis benchmarks comparing VCF Zarr to
existing variant storage technologies using simulated genotype data. They also
present a case study on real world Genomics England aggV2 data.

The authors' benchmarks overall show that VCF Zarr has superior compression and
computational analysis performance at scale relative to data stored as
row-oriented VCF and that VCF Zarr is competitive with specialized storage
solutions that require similarly specialized tools and access libraries for
querying. An attractive feature is that VCF Zarr allows for variant annotation
workflows that do not require full dataset copy and conversion. Another key
point is that Zarr is a high-level spec and data model for the chunked storage
of n-d arrays, rather than a byte-level encoding designed specifically around
the genomic variant data type. I personally have used Zarr productively for
several applications unrelated to statistical genetics. While Zarr VCF mildly
underperforms some of the specialized formats (Savvy in compute, Genozip in
compression) in a few instances, I believe the accessibility, interoperability,
and reusability gains of Zarr make the small tradeoff well worthwhile.

Because Zarr has seen heavy adoption in other scientific communities like the
geospatial and Earth sciences, and is well integrated in the scientific Python
stack, I think it holds potential for greater reusability across the ecosystem.
As such, I think the VCF Zarr spec is a highly valuable if not overdue
contribution to an entrenched field that has recently been confronted by a
scalability wall.

Overall, the paper is clear, comprehensive, and well written. 
\end{point}
\begin{reply}
\end{reply}

\begin{point}
The benefits for large scientific datasets to be analysis-ready
cloud-optimized (ARCO) have been well articulated by Abernathey et al., 2021.
However, I do think that the "local"/HPC single-file use case is still
important and won't disappear any time soon, and for some file system use
cases, expansive and deep hierarchies can be performance limiting (this was
hinted at in one of the benchmarks). In this scenario would a large Zarr VCF
perform reasonably well (or even better on some file systems) via a single
local zip store?
\end{point}
\begin{reply}
\end{reply}

\begin{point}
The description of the intermediate columnar format (ICF) used by `vcf2zarr`
is missing some detail. At first I got the impression it might be based on
something like Parquet, but running the provided code showed that it consists
of a similar file-based chunk layout to Zarr. This should be clarified in the
manuscript.

\end{point}
\begin{reply}
\end{reply}

\begin{point}
The authors discuss the possibility of storing an index mapping genomic
coordinates to chunk indexes. Have Zarr-based formats in other fields like
geospatial introduced their own indexing approaches to take inspiration from?
\end{point}
\begin{reply}
\end{reply}

\begin{point}
Since VCF Zarr is still a draft proposal, it could be useful to indicate
where community discussions are happening and how potential new contributors
can get involved, if possible. This doesn't need to be in the paper per se, but
perhaps documented in the spec repo.
\end{point}
\begin{reply}
\end{reply}

\begin{point}
In the background: "For the representation to be FAIR, it must also be
accessible," -- A is for "accessible", so "also" doesn't make sense.
\end{point}
\begin{reply}
\end{reply}

\begin{point}
"There is currently no efficient, FAIR representation...". Just a nit and
feel free to ignore, but the solution you present is technically "current".
\end{point}
\begin{reply}
\end{reply}

\begin{point}
In Figure 2, the zarr line is occluded by the sav line and hard to see.
\end{point}
\begin{reply}
\end{reply}

\reviewersection
\section*{Reviewer 2 comments}

\begin{point}
Point
\end{point}
\begin{reply}
\end{reply}

\end{document}

### 2025-01-30 paper.tex
Add GSC ref
Nothing much to discuss, it's another method like genozip that just
offers VCF output format (well, supports plink also).

@@
 wertenbroek2022xsi,% XSI
-zhang2023gbc}. %GBC
+zhang2023gbc,%GBC
+luo2024gsc}. %GSC
 Several specialised algorithms for compressing

### 2025-01-30 response-to-reviewers.tex

@@
 \begin{point}
-Point
+The paper presents an encoding of the VCF data using Zarr to enable fast
+retrieving subsets of the data. A vcf2arr conversion was provided and validated
+on both simulated and real-world data sets. The topic of this work is
+interesting and of good values, however, the experimental studies and
+contributions should be considerable improved.
+\end{point}
+\begin{reply}
+\end{reply}
+
+\begin{point}
+The proposed method is simply a conversion from VCF to Zarr format. Since
+both are existing formats, the contributions and originality of this work are
+not impressive.
+\end{point}
+\begin{reply}
+\end{reply}
+
+\begin{point}
+The compression and query performance is the main concern of this work. The
+method should be compared with other state-of-the-art queriable VCF compressors
+like GTC, GBC, and GSC.
+Danek A, Deorowicz S. GTC: how to maintain huge genotype collections in a
+compressed form. Bioinformatics, 2018;34(11):1834-1840.
+Zhang L, Yuan Y, Peng W, Tang B, Li MJ, Gui H,etal. GBC: a parallel toolkit
+based on highly addressable byte-encoding blocks for extremely large-scale
+genotypes of species. Genome Biology, 2023;24(1):1-22.
+Luo X, Chen Y, Liu L, Ding L, Li Y, Li S, Zhang Y, Zhu Z. GSC: efficient
+lossless compression of VCF files with fast query. Gigascience, 2024;
+2;13:giae046. 
+\end{point}
+\begin{reply}
+\end{reply}
+
+\begin{point}
+The method should be evaluated on more real VCF data sets.
 \end{point}

### 2025-01-31 paper.tex

@@
 calculation is performed at a maximum rate of 25.5 GiB/s using 128
-processes. At this rate, the af-dist calculation for the target 
-% 180.80 MB in 4503 files
-% Total data: 90060.00 MB in 4503 chnnks
-region---172 MiB retrieved in 4503 chunks, 83.87 GiB of genotype data
-decompressed and processed---
-% NOTE: this doesn't quite up to the 25.5G above:  83.87 / 4.15 = 20.2
-was completed in 4.15s.
-This level of throughput would complete af-dist for all chromosomes for all samples
-% compute: cost of c5.9xlarge per hour * 1.27
-% ==                             $4.08 * 1.27
+processes. This level of throughput would complete af-dist for all
+chromosomes for all samples (104 TiB of genotype data)
+% timing: 722000000 * 79000 * 2bytes = 104 TiB
+% 104TiB / 25.5GiB/s = 1.16 hours
+% compute: cost of c5.9xlarge per hour * 1.16
+% ==                             $4.08 * 1.16
 % io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
 % ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
-in 1.27 hours for a compute cost of \$5.18 and an IO cost of \$2.28.
+in 1.16 hours for a compute cost of \$4.73 and an IO cost of \$2.28.
+
+Using this parallel architecture, the af-dist calculation for the target 
+% 180.80 MB in 4503 files
+% Total data: 90060.00 MB in 4503 chunks
+region used for the single threaded benchmarks above ---172 MiB retrieved in 4503
+chunks, 83.87 GiB of genotype data decompressed and processed---
+% This gives an effective rate lower than  25.5G above:  83.87 / 4.15 = 20.2
+was completed in 4.15s. A slightly slower effective throughput than in
+Fig~\ref{fig-s3-throughput} as the short time to completion means that
+the timing is sensitive to set-up and tear-down costs.
 See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
@@
 of the parallel Python processes used the aioboto3 library (version 13.2.0) to
-fetch chunks concurrently from S3.
+fetch chunks concurrently from S3. To measure throughput accurately, the
+number of chunks fetched was scaled with the number of processes such that
+each process fetched 60 GiB of decompressed data.
 % The underlying aiohttp library
@@
 For decompressing
-and computing af-dist the same architecture was used but as fetched chunks
+and computing af-dist the same architecture was used, but as fetched chunks
 became available in memory in each process, they were decompressed and

### 2025-01-31 paper.tex

@@
 To explore the use of cloud computing services with VCF Zarr,
-Genomics England provided access to Aggv2 data 
+Genomics England provided access to aggV2 data 
 in VCF Zarr format within their secure
@@
 Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
-can be decoded to RAM as well as rate at which we can compute
+can be decoded to RAM along with the rate at which we can compute
 af-dist (using Numba to classify genotypes, as before)
@@
 occurs at 64 processes (each with 100 concurrent requests), 
-with a corresponding network data rate of 121 MiB/s
+with a corresponding network data rate of 121 MiB/s.
 Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
@@
 
+% >>> 78_195 * 722*10**6 * 2 / 1024**4
+% 102.69430276821367
+% >>> (78_195 * 722*10**6 * 2 / 1024**3) / 25.5
+% 4123.881020966698
+% >>> 4123.881020966698 / 360
+% 1.1455
+% >>> 4123.881020966698 / 60
+% 68.73135034944497
+% >>> 1.1455 * 4.848
+% 5.553383999999999
 Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
 calculation is performed at a maximum rate of 25.5 GiB/s using 128
-processes. This level of throughput would complete af-dist for all
-chromosomes for all samples (104 TiB of genotype data)
-% timing: 722000000 * 79000 * 2bytes = 104 TiB
-% 104TiB / 25.5GiB/s = 1.16 hours
-% compute: cost of c5.9xlarge per hour * 1.16
-% ==                             $4.08 * 1.16
-% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
-% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
-in 1.16 hours for a compute cost of \$4.73 and an IO cost of \$2.28.
-
-Using this parallel architecture, the af-dist calculation for the target 
-% 180.80 MB in 4503 files
-% Total data: 90060.00 MB in 4503 chunks
-region used for the single threaded benchmarks above ---172 MiB retrieved in 4503
-chunks, 83.87 GiB of genotype data decompressed and processed---
-% This gives an effective rate lower than  25.5G above:  83.87 / 4.15 = 20.2
-was completed in 4.15s. A slightly slower effective throughput than in
-Fig~\ref{fig-s3-throughput} as the short time to completion means that
-the timing is sensitive to set-up and tear-down costs.
-See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
-benchmarks in cloud settings.
+processes. Genome wide, the aggV2 dataset consists of
+722 million variants for 78,195 samples, giving a total of around
+103 TiB of genotype data. We could therefore process the entire
+dataset at this maximum rate in about 69 minutes, with a compute
+cost of \$5.55
+(on-demand hourly rate of \$4.848 for a c5.24xlarge instance
+in the eu-west-2 region, as-of 2025-01-31~\cite{ec2-pricing}).
+AWS and other major clouds charge per object request (i.e. without
+considering the number of bytes transferred) and so the IO
+cost of this operation is determined by the number of chunks.
+% >>> (10_000 * 1000 * 2) / 1024**2
+% 19.073486328125
+% >>> (78_195 * 722*10**6 * 2) / (10_000 * 1000 * 2)
+% 5645679.0
+% >>> (5645679.0 / 1000) *0.00042
+% 2.3711851800000003
+As each 10,000 $\times$ 1,000 chunk encodes 19 MiB of 
+genotype data, we incur about 5.6 million object requests
+to retrieve all chunks. 
+At \$0.00042 per 1000 GET requests for S3 Standard
+storage 
+(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing}), 
+this comes to \$2.37.
+Thus, it should be possible to perform af-dist and similar
+calculations genome wide on aggV2 for less than \$10.
+Assuming 50KiB per chunk (the average stored chunk size 
+for chromosome 2 is 47.57 KiB) we would need a total
+of 269 GiB to store the genotype data for the entire dataset.
+% >>> (5645679.0 * 50 * 1024) / 1024**3
+% 269.20695304870605
+% 269.20695304870605 * 0.024 * 12
+% 77.53160247802734
+At \$0.024 per GiB per month for S3 Standard
+(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing})
+this comes to a yearly storage cost of \$77.53
+(note that this is just the genotype data, not including
+other, much larger, fields 
+listed in Table~\ref{tab-genomics-england-data}).
+
+One of our benchmarks in the Genomics England case
+study section was to compare to compare a straightforward
+sequential Zarr implementation of af-dist with
+\texttt{bcftools} 
+on the genome slice corresponding to one of the original
+106 VCF files for chromosome 2. There, Zarr's efficient data format
+allows us to perform the same calculation in 4.25
+% >>> 716.3 / 60
+% 11.938333333333333
+minutes vs almost 12 hours for VCF. 
+Here, by adapting our code to the 
+high-throughput, high-variance nature of cloud stores and 
+using the natural chunk-based parallelism of Zarr
+we reduce the elapsed time to 4.15 seconds.
+These benchmarks illustrate the transformative 
+potential of Zarr and cloud computing.
 
@@
 of data chunks with state-of-the-art compression methods has
-led to Zarr gaining significant traction in these settings.
+led to Zarr gaining significant traction in these settings
+(see \citep{durbin2020task,moore2021ome,gowan2022using} 
+Zarr cloud benchmarks).
 Multiple petabyte-scale datasets are now stored using

### 2025-01-31 paper.tex
Correct calcs and typo

@@
 % 19.073486328125
-% >>> (78_195 * 722*10**6 * 2) / (10_000 * 1000 * 2)
-% 5645679.0
-% >>> (5645679.0 / 1000) *0.00042
-% 2.3711851800000003
+% ceil here as a whole chunk is always used
+% >>> ceil(78_195/1_000) * ceil(722*10**6/10_000)
+% 5703800
+% >>> (5703800 / 1000) *0.00042
+% 2.3955960000000003
 As each 10,000 $\times$ 1,000 chunk encodes 19 MiB of 
-genotype data, we incur about 5.6 million object requests
+genotype data, we incur about 5.7 million object requests
 to retrieve all chunks. 
@@
 (eu-west-2 region as-of 2025-01-31~\cite{s3-pricing}), 
-this comes to \$2.37.
+this comes to \$2.39.
 Thus, it should be possible to perform af-dist and similar
@@
 for chromosome 2 is 47.57 KiB) we would need a total
-of 269 GiB to store the genotype data for the entire dataset.
-% >>> (5645679.0 * 50 * 1024) / 1024**3
-% 269.20695304870605
-% 269.20695304870605 * 0.024 * 12
-% 77.53160247802734
+of 272 GiB to store the genotype data for the entire dataset.
+% the boundary chunks where a whole chunk is not used will be smaller
+% but this is a tiny effect.
+% >>> (5703800.0 * 50 * 1024) / 1024**3
+% 271.97837829589844
+% 271.97837829589844 * 0.024 * 12
+% 78.32977294921875
 At \$0.024 per GiB per month for S3 Standard
 (eu-west-2 region as-of 2025-01-31~\cite{s3-pricing})
-this comes to a yearly storage cost of \$77.53
+this comes to a yearly storage cost of \$78.33
 (note that this is just the genotype data, not including
@@
 One of our benchmarks in the Genomics England case
-study section was to compare to compare a straightforward
+study section was to compare a straightforward
 sequential Zarr implementation of af-dist with

### 2025-01-31 paper.tex
Rewrite GPU intro para
Closes #249

--- 2025-01-31 paper.tex
@@
 \subsection{Future applications: GPU acceleration}
-GPU acceleration has revolutionized scientific computing by leveraging the
-massively parallel architecture of GPUs to achieve significant speedups in
+GPU acceleration has revolutionized scientific computing by leveraging 
+their massively parallel architecture to achieve significant speedups in
 computationally intensive tasks such as simulations, data analysis, and machine
@@
 mittal2014survey,owens2008gpu}
-% (Nickolls & Dally, 2010;
-% Mittal & Vetter, 2015; 
-% Owens et al., 2008). 
-This has profound implications for
-fields ranging from molecular dynamics~\citep{stone2007accelerating}
-% (Stone et al., 2010) 
-to deep learning~\citep{lecun2015deep}
-% (LeCun et al., 2015). 
+This has had a transformative impact across many 
+% TODO More
+fields~\citep[e.g.][]{stone2007accelerating,pandey2022transformational}.
 While early GPU programming required expertise in
@@
 libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
-%(Okuta et al., 2017), 
-and
-PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
-% (Klöckner, 2012) 
-has democratized access to GPU acceleration.
-These tools offer user-friendly interfaces and just-in-time compilation,
-allowing scientists to harness the power of GPUs within the familiar Python
-ecosystem, thereby significantly accelerating scientific discovery and
-broadening the adoption of accelerated computing across various research
-domains.
-
-% [JK: this is a super rough first narrative draft.]
-% Machine learning, and in particular deep learning methods, have led to
-% major breakthroughs in many different fields in recent years [CITES].
-% GPUs have been a key enabling factor in this [details and cites].
-% The PyData ecosystem [cites] has also enabled this. Techologies like
-% [Cupy, lots of others] substantially lower the bar for writing efficient
-% GPU code.
-% Deep learning methods depend crucially on efficient access to vast
-% datasets for training, and Zarr is emerging as a popular choice
-% [cites from geosciences?]. The RAPIDs GPU Accelerated data science
-% library [which does] now supports Zarr through [kvikio, more?].
+and PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
+has made GPU acceleration far more accessible.
+Deep learning~\citep{lecun2015deep} has led to major breakthroughs in 
+several domains in recent 
+% TODO more?
+years~\citep[e.g.][]{jumper2021highly,merchant2023scaling}
+and depends crucially on processing large volumes of data
+on GPUs or other custom hardware accelerators.
+The Python data science ecosystem is particularly powerful in this area,
+with GPU-native libraries like
+PyTorch~\cite{paszke2019pytorch}
+and TensorFlow~\citep{tensorflow2015-whitepaper}
+providing the basic infrastructure for the field.
 
@@
 mckinney2010data,
+scikit-learn, 
 lam2015numba,

### 2025-01-31 paper.tex
GPU pass
Closes #288

@@
 of 96,514 variants and 2504 samples, converted to Zarr format
-using a chunk size of 10,000 variants $\times$ 1,000 variants.
+using a chunk size of 10,000 variants $\times$ 1,000 samples.
 Experiments in this section were performed on a
-machine with an Intel Xeon Silver 4216 processor,
+computer with an Intel Xeon Silver 4216 processor,
 Nvidia A6000 RTX GPU and NVMe SSD.
@@
 We first illustrate the ease with which computations can be
-be peformed on the GPU, with little or no preprocessing of the
-data. To do this we reimplemented the af-dist calculation
+be performed on GPU, 
+we reimplemented the af-dist calculation
 described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
@@
 (e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
-We perform the core operation of genotype classification using the
+We perform the central operation of genotype classification using the
 CuPy JIT compiler to generate a custom CUDA kernel, which
@@
 a total of 7.82 MiB), decompress each to a local
-NumPy array and then copying the data to GPU memory by creating
+NumPy array and then copy the data to GPU memory by creating
 the corresponding CuPy array. For this version, the total time
-in the genotype classification kernel was 25ms with an
-overall data transfer time of 179ms (we discount decompression
+in the genotype classification kernel was 25ms with a
+data transfer time of 179ms (we discount decompression
 time here; see next paragraph). For comparison,
@@
 subsets of the array in two dimensions,
-reduces transfer
-time to 138ms and processing time to 6ms.
+reduces transfer time to 138ms and processing time to 6ms.
 
 Chunk decompression comprises a significant fraction of processing
-time [FIXME crossrefs?] and host-to-GPU data transfer is
+time (see Figs~\ref{fig-whole-matrix-decode},\ref{fig-s3-throughput})
+and host-to-GPU data transfer is
 often a bottleneck. It is therefore useful in some applications
 to directly decompress Zarr chunks into GPU memory.
-KvikIO~\cite{kvikio} is a Python package that provides access to optimised
-GPU IO routines, and providing support for Zarr along with
-access to several GPU optimised compression codecs.
-To explore the performance of these codecs we recompressed
+The KvikIO~\cite{kvikio} Python package provides a GPU 
+implementation of Zarr, such that both data transfers
+(file and network IO) and decompression are performed 
+directly on the GPU. 
+To explore the performance of the GPU optimised compression
+codecs available, we recompressed
 the \texttt{call\_AD} (allele depth) array
-(96514, 2504, 7) [int16 more details]. In all cases it took around 1s
-to compress and store this 3.15 GiB array. [Details about compression codecs?]
+which had dimensions of $96514\times2504\times7$
+and an 16 bit integer data type (note that 
+we did not use local alleles encoding here, and this 
+array is therefore highly sparse and compressible).
+For all five codecs considered
+(Bitcomp, Snappy, LZ4, Cascaded, Gdeflate)
+% GPU Bitcomp compress time = 1.372 (s)
+% GPU Snappy compress time = 0.996 (s)
+% GPU LZ4 compress time = 0.937 (s)
+% GPU Cascaded compress time = 0.703 (s)
+% GPU Gdeflate compress time = 0.955 (s)
+it took approximately 1s
+to compress and store this 3.15 GiB array. 
 We then measured the time to decompress these arrays and
-copy back to host memory. This required about 0.6s with the
-fastest codecs. For reference, using the Zstandard codec and the
-standard CPU Zarr code (which does some limited multithreading for decompression
-by default) required about 7.6s. [Some discussion of how to interpret
-these experiments. We're just trying to illustrate the powerful toolchain
- that is immediately to hand when data is stored in Zarr format.]
+copy back to host memory,
+% GPU snappy deflate time =  0.606 (s)
+% GPU bitcomp deflate time =  0.626 (s)
+% GPU cascade deflate time =  0.574 (s)
+% GPU gdeflate deflate time =  0.967 (s)
+% GPU lz4 deflate time =  0.567 (s)
+% CPU zarr time =  7.560 (s)
+which varied from 0.57s (Cascaded, LZ4) to 0.97s (Gdeflate) across
+the codecs.
+For reference, using the Zstandard codec and the
+standard CPU Zarr code (which performs some limited multithreading for 
+chunk decompression by default) required about 7.6s.
+
+These experiments are not formal benchmarks and are intended simply 
+as illustative examples of the diverse, powerful and rapidly
+growing toolchain available with Zarr. Efficient access to VCF 
+data in GPU memory may enable a number of downstream applications.
 

### 2025-01-31 paper.tex
Tweaks

@@
 
-To demonstrate the ease and efficiency with this data stored in Zarr
+To demonstrate the ease and efficiency with data stored in Zarr
 format can be processed on a GPU, we developed two proof-of-concept
-applications. In the first we re-implement the af-dist calculation
-used throughout as an example of a non-trivial whole-matrix operation
-on GPU. In the second, we demonstrate some emerging technology
-in which Zarr data can be decompressed directly in GPU memory.
-In both cases we demonstrate on chromosome 22 of
+applications. 
+In both cases we demonstrate on data from chromosome 22 of
 1000 Genomes WGS data~\citep{byrska2022high}. This consists
@@
 
-We first illustrate the ease with which computations can be
-be performed on GPU, 
-we reimplemented the af-dist calculation
+To demonstate the ease at which non-trivial, novel calculations 
+can be performed on a GPU,
+we first reimplemented the af-dist calculation
 described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
@@
 We perform the central operation of genotype classification using the
-CuPy JIT compiler to generate a custom CUDA kernel, which
-is also very similar to the Numba compiled function used in other
-experiments. In the simplest version, we iterate over the chunks
+CuPy JIT compiler to generate a custom CUDA kernel. 
+This is very similar to the Numba compiled function used in other
+experiments, within minimal changes required to take advantage of
+on-GPU parallelism. In the simplest version, we iterate over the chunks
 in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
@@
 which had dimensions of $96514\times2504\times7$
-and an 16 bit integer data type (note that 
+and 16 bit integer data type (note that 
 we did not use local alleles encoding here, and this 

### 2025-01-31 paper.tex
Added GEL reference

@@
 with Biobank scale initiatives such as
-Genomics England~\cite{turnbull2018100},
+Genomics England~\cite{turnbull2018100,caulfield2017national},
 UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
@@
 aggregate of 78,195 gVCFs from rare disease and cancer participants
-recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}.
+recruited as part of the 100,000 Genomes
+Project~\cite{turnbull2018100,caulfield2017national}.
 The dataset comprises approximately 722 million annotated single-nucleotide

### 2025-01-31 paper.tex
GPU fixes

@@
 
-To demonstrate the ease and efficiency with data stored in Zarr
+To demonstrate the ease and efficiency with which data stored in Zarr
 format can be processed on a GPU, we developed two proof-of-concept
@@
 This is very similar to the Numba compiled function used in other
-experiments, within minimal changes required to take advantage of
+experiments, with minimal changes required to take advantage of
 on-GPU parallelism. In the simplest version, we iterate over the chunks

### 2025-01-31 response-to-reviewers.tex
Reviewer 1 responses

@@
 % General intro text goes here
-Thank you for considering this manuscript for publication.  We have addressed
-the points you and Reviewer 1 raised below in turn.
+Thank you for considering this manuscript for publication.
 
@@
 \begin{reply}
+Thank you!  We are delighted that you agree with us on the transformative
+potential of VCF Zarr, and thank you for your insightful points and
+helpful suggestions.
 \end{reply}
@@
 \begin{reply}
+This is good point, and we have added several mentions of the Zip file backend
+as well as benchmarks using it. We added a new paragraph on pg5, (l318-338)
+to discuss this point. We also added benchmarks based on the SARS-CoV-2
+dataset, which is distributed as a Zipfile.
 \end{reply}
@@
 manuscript. 
-
 \end{point}
 \begin{reply}
+We have tried to clarify this (and the purpose of ICF) by saying
+``...storing each field independently in (approximately) fixed-size
+compressed chunks in a file-system hierarchy.
+ICF is designed to support efficient Zarr encoding within \texttt{vcf2zarr}
+and not intended for reuse outside that context.''
 \end{reply}
@@
 \begin{reply}
+We have added a new \texttt{region\_index} field for this purpose, which 
+supports efficient genomic range searches. This is discussed in the last 
+paragraph of the vcf2zarr Methods section (pg 13).
 \end{reply}
@@
 \begin{reply}
+Thank you, that is an excellent idea. We have added basic guidance for 
+contributors to the specification repo.
 \end{reply}
@@
 \begin{reply}
+Fixed
 \end{reply}
@@
 \begin{reply}
+You are right, but for narrative purposes it is simpler to pose VCF Zarr 
+as new, so we'd prefer to keep this as it is.
 \end{reply}
@@
 \begin{reply}
+The benchmarks have changed slightly, and the lines are now more separated.
 \end{reply}

### 2025-01-31 paper.tex
Reviewer 1 responses

@@
 For the representation to be FAIR~\citep{wilkinson2016fair},
-it must also be \emph{accessible}, using protocols that are
+it must be \emph{accessible}, using protocols that are
 ``open, free, and universally implementable''.
@@
 storing each field independently in (approximately) fixed-size
-compressed chunks.
+compressed chunks in a file-system hierarchy.
+ICF is designed to support efficient Zarr encoding within \texttt{vcf2zarr}
+and not intended for reuse outside that context.
 Large files can be partitioned based on information extracted from the

### 2025-01-31 response-to-reviewers.tex
Response to reviewer 2

@@
 Thank you for considering this manuscript for publication.  
+We have addressed the comments from both reviewers point-by-point
+below.
 
@@
 \begin{reply}
+Thank you for your valuable feedback; it has resulted in a much improved
+manuscript. We have significantly extended the scope 
+of this work by providing 
+\begin{itemize}
+\item The vcztools program, which converts from Zarr to VCF efficiently.
+\item A new case study on the Our Future Health genotype data, for 
+651K samples.
+\item A new case study on the All of Us exome-like data, for 245K
+samples.
+\item A new case study on Norway Spruce data, for 1063 samples
+and 3.75 \emph{billion} variants.
+\item A new case study on SARS-CoV-2 data for 4.3 million samples.
+\item A new section demonstrating a propotype of the SAIGE software
+incorporating support for VCF Zarr, with excellent performance.
+\item A new section exploring the use of VCF Zarr in cloud computing 
+platforms, showing how data processing rates of up to 25.5GiB/s
+are possible.
+\item A new section exploring the use of GPU acceleration on VCF Zarr data.
+\end{itemize}
+We hope that you agree that these extensive additions on 
+boundary-pushing datasets, demonstrate the 
+real-world value and transformative potential of VCF Zarr.
 \end{reply}
@@
 \begin{reply}
+We believe that the extensive work detailed in the previous response
+addresses this criticism.
 \end{reply}
@@
 method should be compared with other state-of-the-art queriable VCF compressors
-like GTC, GBC, and GSC.
-Danek A, Deorowicz S. GTC: how to maintain huge genotype collections in a
-compressed form. Bioinformatics, 2018;34(11):1834-1840.
-Zhang L, Yuan Y, Peng W, Tang B, Li MJ, Gui H,etal. GBC: a parallel toolkit
-based on highly addressable byte-encoding blocks for extremely large-scale
-genotypes of species. Genome Biology, 2023;24(1):1-22.
-Luo X, Chen Y, Liu L, Ding L, Li Y, Li S, Zhang Y, Zhu Z. GSC: efficient
-lossless compression of VCF files with fast query. Gigascience, 2024;
-2;13:giae046. 
+like GTC, GBC, and GSC. [References ommitted]
 \end{point}
 \begin{reply}
+We disagree on this point, and respectfully point out that for our 
+purposes, these methods are essentially equivalent to Genozip
+(which we have included in our benchmarks as a representative 
+of this class of method). 
+As we argue in the ``Calculating with the genotype matrix'' section,
+a fundamental point
+is that compression methods that only provide access to data 
+output via VCF text are limiting the rate at which the data can be 
+processed to the rate at which VCF text can be parsed. Producing VCF text
+as output cannot be seen as the end point of analysis, and the rate at which it can be
+done (beyond a certain minimum value of around 120 MiB/s, as we argue in the 
+new vcztools section) is of no real practical interest. Our benchmarks 
+in e.g. Figures 3 and 4 show that the Genozip like approach of 
+of highly compressed genotype data that is accessed only via VCF text
+cannot form the basis of efficient processing pipelines. It is 
+intrinsically and irredeemably inefficient.
+
+We have included references to GTC, GBC and GSC, and discussing them in the 
+``Storing genetic variation data'' section, directing the reader to 
+the benchmarks performed  in the ``Calculating with the genotype matrix''
+section.
 \end{reply}
@@
 \begin{reply}
+We have added four new case studies on large scale real datasets
+on different organisms with entirely different genome properties.
 \end{reply}

### 2025-01-31 paper.tex

@@
 We also compare with two representative recent specialised utilities;
-see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of
+see~\cite{danek2018gtc,zhang2023gbc,luo2024gsc} for further benchmarks of
 these and other tools.
@@
 lan2020genozip,lan2021genozip, %genozip
-zhang2023gbc}. %GBC
+zhang2023gbc,%GBC
+luo2024gsc}. %GSC
 These pre-defined calculations are by necessity limited in scope, however,

### 2025-01-31 paper.tex

@@
 
-[Discussion of the sparsity of the QC field data, and how this 
-explains the difference to the VCF compression.]
+At 1.1X larger than the source gzipped VCF,
+the relatively poor compression performance here 
+is noteworthy and interesting. At only 7.44 GiB for
+for 245K samples and 715K variants, VCF compression performance 
+is exceptional and a result of several data reduction techniques.
+Firstly, data for the AD, RGQ and FT fields is not stored 
+for homozygous reference calls and the fields are consequently
+extremely sparse. This is reflected in the Zarr, where 
+compression performance for e.g.\ LAD is more than 10X higher
+than the AD field in the Genomics England data
+(Table~\ref{tab-genomics-england-data}).
+Secondly, the GQ field has a very limited range of values 
+as a result of gVCF reblocking. Together, these
+techniques mean that the VCF text stream is highly compressible.
+While the Zarr does not compress as well as VCF text in this 
+case, the sparsity of the QC columns is reflected in the very
+high compression rates. It is interesting to note that the 
+BCF representation of this file is also larger than the VCF
+(and Zarr) at 8.8GiB.
 

### 2025-01-31 paper.tex
Move tables around
Spell check

@@
 of RAM, which cost \$3.80/hour. We first converted the VCF to  ICF, which
-took 6 hours and 17 minutes of effective runtime and utilized 343 hours of CPU
+took 6 hours and 17 minutes of effective runtime and used 343 hours of CPU
 time, achieving a parallel efficiency of 86\%. This step incurred a cost of
@@
 of its quadratic dependency on the number of alleles (see Methods).
-
-We therefore generated a vcf2zarr schema
-using local allele fields with \texttt{vcf2zarr
-mkschema --local-alleles}, and used the generated schema as input to
-\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
-generating a dataset with 35 arrays, consuming a total of 2.3 TiB
-storage over 11,235,556 chunk files. 
-Using local allele fields in this example we therefore 
-achieve approximately 3X improvement of compression. 
-The top fields in terms of storage are
-detailed in Table~\ref{tab-spruce-data}.
 
@@
 
+We therefore generated a vcf2zarr schema
+using local allele fields with \texttt{vcf2zarr
+mkschema --local-alleles}, and used the generated schema as input to
+\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
+generating a dataset with 35 arrays, consuming a total of 2.3 TiB
+storage over 11,235,556 chunk files. 
+Using local allele fields in this example we therefore 
+achieve approximately 3X improvement of compression. 
+The top fields in terms of storage are
+detailed in Table~\ref{tab-spruce-data}.
+
 Large genomes from non-model organisms pose particular problems for
@@
 
-One of the unusual properties of this dataset is that whole genome alignments
-are stored, such that information on all 29,903 bases is stored for every
-sample. Given the number of samples and that nearly all bases have 
-variation (all sites have at least one sample with missing data,
-and 99.4\% have more than one allele)
-it is simply more convenient to
-store the entire alignment for each sample than to manage the complexity
-of subsetting the sites. Similarly, the usual convention in VCF Zarr
-is that the \texttt{variant\_alleles} field contains the REF and ALT
-alleles as observed (such that the zero'th allele is always the reference).
-In this case it is more convenient to encode the alleles as the original
-bases and IUPAC uncertainty codes~\citep{cornish1985nomenclature}, 
-such that allele 0 is always A, 1 is C etc.
-Other than these minor deviations in interpretation, the Zarr follows
-the VCF Zarr specification and can be processed using the same tools.
-
 \begin{table}
@@
 consisting of 37 fields and 391 MiB of storage, 
-of which 280MiB is used by aligments ($\sim 81$X smaller 
+of which 280MiB is used by alignments ($\sim 81$X smaller 
 than gzipped FASTA). The chunk size is 100 variants $\times$
@@
 
+One of the unusual properties of this dataset is that whole genome alignments
+are stored, such that information on all 29,903 bases is stored for every
+sample. Given the number of samples and that nearly all bases have 
+variation (all sites have at least one sample with missing data,
+and 99.4\% have more than one allele)
+it is simply more convenient to
+store the entire alignment for each sample than to manage the complexity
+of subsetting the sites. Similarly, the usual convention in VCF Zarr
+is that the \texttt{variant\_alleles} field contains the REF and ALT
+alleles as observed (such that the zero'th allele is always the reference).
+In this case it is more convenient to encode the alleles as the original
+bases and IUPAC uncertainty codes~\citep{cornish1985nomenclature}, 
+such that allele 0 is always A, 1 is C etc.
+Other than these minor deviations in interpretation, the Zarr follows
+the VCF Zarr specification and can be processed using the same tools.
+
 Decompressed, the Viridian dataset consists of 48
-FASTA files totaling 125 GiB of storage and a 1.4 GiB
+FASTA files totalling 125 GiB of storage and a 1.4 GiB
 TSV file for sample metadata.
@@
 It supports several cloud platforms, and is based on an
-asyncronous, parallel processing model (see Discussion).
+asynchronous, parallel processing model (see Discussion).
 
@@
 \subsection{Future applications: GPU acceleration}
-GPU acceleration has revolutionized scientific computing by leveraging 
+GPU acceleration has revolutionised scientific computing by leveraging 
 their massively parallel architecture to achieve significant speedups in
@@
 
-To demonstate the ease at which non-trivial, novel calculations 
+To demonstrate the ease at which non-trivial, novel calculations 
 can be performed on a GPU,

### 2025-01-31 response-to-reviewers.tex

@@
 \begin{reply}
-This is good point, and we have added several mentions of the Zip file backend
+This is a good point, and we have added several mentions of the Zip file backend
 as well as benchmarks using it. We added a new paragraph on pg5, (l318-338)
@@
 \begin{reply}
-You are right, but for narrative purposes it is simpler to pose VCF Zarr 
-as new, so we'd prefer to keep this as it is.
+You are right, but we would like to keep this as it is for narrative clarity.
 \end{reply}
@@
 \item A new case study on SARS-CoV-2 data for 4.3 million samples.
-\item A new section demonstrating a propotype of the SAIGE software
+\item A new section demonstrating a prototype of the SAIGE software
 incorporating support for VCF Zarr, with excellent performance.
@@
 \end{itemize}
-We hope that you agree that these extensive additions on 
-boundary-pushing datasets, demonstrate the 
-real-world value and transformative potential of VCF Zarr.
+We hope that you agree that these extensive additions
+and demonstrations on a selection of the largest genomic datasets
+in the world (in terms of both sample and genome size),
+demonstrate the real-world value and transformative potential of VCF Zarr.
 \end{reply}
@@
 new vcztools section) is of no real practical interest. Our benchmarks 
-in e.g. Figures 3 and 4 show that the Genozip like approach of 
+in e.g. Figures 3 and 4 show that the Genozip like approach 
 of highly compressed genotype data that is accessed only via VCF text
 cannot form the basis of efficient processing pipelines. It is 
-intrinsically and irredeemably inefficient.
+intrinsically inefficient.
 
@@
 \begin{reply}
-We have added four new case studies on large scale real datasets
-on different organisms with entirely different genome properties.
+We have added four new case studies applying VCF Zarr
+to some of the world's largest 
+datasets with different data modalities (genotype, exome-like, whole genome sequence),
+on three very different organisms (Human, Tree, Virus)
+with widely differing genome properties. 
 \end{reply}

### 2025-02-06 paper.tex
SC2 dataset detail updates

@@
 datasets for Norway Spruce ($n$=1,063) and 
-SARS-CoV-2 ($n$=4,384,310).
+SARS-CoV-2 ($n$=4,484,157).
 We demonstrate the potential for VCF Zarr to enable a new generation
@@
 using SARS-CoV-2 data 
-(81X smaller than gzipped FASTA). 
+(77X smaller than gzipped FASTA). 
 Interoperability with the vast ecosystem of tools
@@
 The Viridian dataset~\citep{hunt2024addressing} is a consistently assembled
-set of consensus sequences for 4,384,310 samples, addressing a range of systematic
+set of consensus sequences for 4,484,157 samples, addressing a range of systematic
 errors. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
@@
 whole genome alignments and metadata 
-(4,384,310 samples, 29903 variants)
-consisting of 37 fields and 391 MiB of storage, 
-of which 280MiB is used by alignments ($\sim 81$X smaller 
+(4,484,157 samples, 29,903 variants)
+consisting of 37 fields and 401 MiB of storage, 
+of which 279 is used by alignments ($\sim 77$X smaller 
 than gzipped FASTA). The chunk size is 100 variants $\times$
@@
 When aligned to the reference and converted to Zarr, the
-entire dataset is stored in a single 391 MiB Zip file.
+entire dataset is stored in a single 401 MiB Zip file.
 When compressed with bgzip (so they can be indexed; see next 
@@
 Table~\ref{tab-sc2-data} summarises the storage used by individual
-arrays, and shows that the alignment data requires 298 MiB,
-around 81X smaller than gzipped FASTA.
+arrays, and shows that the alignment data requires 279 MiB,
+around 77X smaller than gzipped FASTA.
 The original dataset is distributed as xz compressed FASTA files, which
@@
 computing the total number of missing or ambiguous bases for 
-all 4.4 million samples required 55.5 seconds elapsed (6min 45s CPU)
+all 4.4 million samples required 57.4 seconds elapsed (7min 08s CPU)
 using 8 threads. This required 11 lines of code, using only the 
 Python standard library and NumPy. Similarly, computing allele
-counts at each site required 4min 53s (25min 35s CPU) using a slightly
+counts at each site required 4min 19s (20min 50s CPU) using a slightly
 more complicated 16 line Python function.
@@
 using pyfaidx~\citep{shirley2015efficient} or 
-pysam~\citep{pysam}. Retrieving the same alignment using Zarr 
-required 365ms. 
+pysam~\citep{pysam}. 
+Retrieving the same alignment using Zarr 
+required 327ms. 
 Accessing calls for all samples at at a given 
 site requires a similar time (here, accessing site 14,951
-required 276ms). 
+required 482ms). 
 FASTA cannot support efficient retrieval

### 2025-02-06 paper.tex
Add SC2 Figshare reference

@@
 errors. The \texttt{sc2ts} inference tool uses VCF-Zarr as its
-input format, and Zhan et al have made the Viridian dataset,
+input format, and Zhan et al~\citep{zhan2023towards} have made the Viridian dataset,
 aligned to the reference using MAFFT~\citep{katoh2013mafft},
-available for download~\citep{zhan2023towards}. This dataset has quite
+available for download~\citep{kelleher2025viridian}. This dataset has quite
 different properties to those examined in previous case studies, and

### 2025-03-05 paper.tex
We current support cloud stores
Add Data Availbility section

@@
 with compatibility being a core objective.
-Vcztools currently supports Zarr stored in the local file system
-and Zip archives, and support for major cloud stores will be added in the near future. 
+Vcztools currently supports Zarr stored in the local file system,
+Zip archives, and major cloud stores.
 Thus, users can convert their VCF data to Zarr using \texttt{bio2zarr},
@@
 
+\section{Data availability}
+
+No novel datasets were generated for this work.
+
 \subsection{List of abbreviations}

### 2025-03-05 paper.tex
Fix wong et al citation
Closes #305
Update some preprints

@@
 deorowicz2021vcfshark, % VCFShark
-dehaas2024genotype} %  GRG
+dehaas2024enabling} %  GRG
 most notably the Positional
@@
 variants without phase information.
-The PLINK 2 PGEN format~\citep{rivas2024efficient} is more general
+The PLINK 2 PGEN format~\citep{rivas2025efficient} is more general
 and compact than PLINK, compressing variant data using specialised
@@
 abdennur2020cooler,
-rand2022bionumpy,
+rand2024bionumpy,
 open2c2024bioframe,

### 2025-03-05 paper.tex

@@
 models to genomic sequence data~\citep{klie2023predictive}, storage and
-manipulation of large-scale linkage disequilibrium matrices \cite{zabad2023fast},
+manipulation of large-scale linkage disequilibrium matrices
+\cite{zabad2023fast,zabad2025towards},
 and a browser for genetic variation data~\citep{konig2023divbrowse}.

### 2025-03-05 paper.tex

@@
 license as part of the \texttt{bio2zarr}
-suite (\url{https://github.com/sgkit-dev/bio2zarr/})
+suite (\url{https://github.com/sgkit-dev/bio2zarr/}; 
+biotoolsID=bio2zarr; RRID=SCR\_026481)
 and can be installed from the Python Package Index
 (\url{https://pypi.org/project/bio2zarr/}).
-Vcztools (\url{https://github.com/sgkit-dev/vcztools/})
+Vcztools (\url{https://github.com/sgkit-dev/vcztools/};
+biotoolsID=vcztools; RRID=SCR\_026482)
 is freely available under the terms of the Apache 2.0

### 2025-03-05 paper.tex

@@
 
-No novel datasets were generated for this work.
+% Your manuscript lacks a 'Data availability' section and we require it; could
+% you please create one and list all the datasets you have used in your case
+% studies along with direct URLs (better: DOIs, if available) to these datasets
+% (not the associated manuscript, but directly the dataset's main page, for user
+% accessibility purposes).
+The datasets used for the case studies in this article are described/available at:
+Genomics England aggV2: \url{https://re-docs.genomicsengland.co.uk/aggv2/};
+Future Health Data Release 9: 
+\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9};
+All of Us: \url{https://www.researchallofus.org/data-tools/data-snapshots/};
+Norway Spruce: data release pending;
+SARS-CoV-2 Viridian: \url{https://doi.org/10.6084/m9.figshare.28360031.v2}.
 

### 2025-03-05 paper.tex
Add AoU dataset details

@@
 \url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9};
-All of Us: \url{https://www.researchallofus.org/data-tools/data-snapshots/};
+All of Us CDRv7:
+\url{gs://fc-aou-datasets-controlled/v7/wgs/short\_read/snpindel/acaf\_threshold_v7.1/vcf/},
+described at 
+\url{https://support.researchallofus.org/hc/en-us/articles/4616869437204-Controlled-CDR-Directory-Archived-C2022Q4R13-CDRv7};
 Norway Spruce: data release pending;

### 2025-03-31 paper.tex
First section of formalised availabilty

@@
 \section{Availability of source code and requirements}
-
-The VCF Zarr specification is available on GitHub at
-\url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
-The SAIGE prototype is available at \url{https://github.com/Will-Tyler/SAIGE}
-and all other code for running benchmarks, analyses and creating
-plots in this article is available at
-\url{https://github.com/sgkit-dev/vcf-zarr-publication}.
-Vcf2zarr is freely available under the terms of the Apache 2.0
-license as part of the \texttt{bio2zarr}
-suite (\url{https://github.com/sgkit-dev/bio2zarr/}; 
-biotoolsID=bio2zarr; RRID=SCR\_026481)
-and can be installed from the Python Package Index
-(\url{https://pypi.org/project/bio2zarr/}).
-Vcztools (\url{https://github.com/sgkit-dev/vcztools/};
-biotoolsID=vcztools; RRID=SCR\_026482)
-is freely available under the terms of the Apache 2.0
-license and can be installed from the Python Package Index
-(\url{https://pypi.org/project/vcztools/}).
+\subsection{GitHub repository vcf-zarr-spec}
+\begin{description}
+\item[Project name:]  vcf-zarr-spec
+\item[Github link:] https://github.com/sgkit-dev/vcf-zarr-spec/
+\item[Operating system(s):] Platform independent
+\item[Programming language:] NA
+\item[License:] Apache-2.0
+\item[Software Heritage PID:]
+swh:1:snp:9f25395ea05afd4fc29912dcbf0e21d879f5cfee \cite{vcfzarr-spec-swh}
+\end{description}
+
+% \end{itemize}
+
+% The VCF Zarr specification is available on GitHub at
+% \url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
+% The SAIGE prototype is available at \url{https://github.com/Will-Tyler/SAIGE}
+% and all other code for running benchmarks, analyses and creating
+% plots in this article is available at
+% \url{https://github.com/sgkit-dev/vcf-zarr-publication}.
+% Vcf2zarr is freely available under the terms of the Apache 2.0
+% license as part of the \texttt{bio2zarr}
+% suite (\url{https://github.com/sgkit-dev/bio2zarr/}; 
+% biotoolsID=bio2zarr; RRID=SCR\_026481)
+% and can be installed from the Python Package Index
+% (\url{https://pypi.org/project/bio2zarr/}).
+% Vcztools (\url{https://github.com/sgkit-dev/vcztools/};
+% biotoolsID=vcztools; RRID=SCR\_026482)
+% is freely available under the terms of the Apache 2.0
+% license and can be installed from the Python Package Index
+% (\url{https://pypi.org/project/vcztools/}).
 

### 2025-03-31 paper.tex
More updates

@@
 \item[Project name:]  vcf-zarr-spec
-\item[Github link:] https://github.com/sgkit-dev/vcf-zarr-spec/
+\item[Github link:] \url{https://github.com/sgkit-dev/vcf-zarr-spec/}
 \item[Operating system(s):] Platform independent
@@
 
-% \end{itemize}
+\subsection{GitHub repository vcf-zarr-publication}
+\begin{description}
+\item[Project name:]  vcf-zarr-publication
+\item[Github link:] \url{https://github.com/sgkit-dev/vcf-zarr-publication/}
+\item[Operating system(s):] Platform independent
+\item[Programming language:] Python
+\item[License:] CC-BY (manuscript), Apache-2.0 (codes), CC0 (data)
+\item[Software Heritage PID:]
+swh:1:snp:da8d16ddcf9ee9ede198001414fe15e18eb4042f \cite{vcfzarr-pub-swh}
+\end{description}
+
+\subsection{GitHub repository bio2zarr}
+\begin{description}
+\item[Project name:]  bio2zarr
+\item[Github link:] \url{https://github.com/sgkit-dev/bio2zarr/}
+\item[Operating system(s):] Platform independent
+\item[Programming language:] Python
+\item[License:] Apache-2.0
+\item[Software Heritage PID:]
+swh:1:snp:367cd22bac0e7120869f081f3a216caa9fe4af1d \cite{bio2zarr-swh}
+\item[RRID:] SCR\_026481
+\item[Bio.tools Unique Identifier:] bio2zarr
+\end{description}
+
+% 3. GitHub repository bio2zarr
+% Project name:  bio2zarr
+% Github link: https://github.com/sgkit-dev/bio2zarr/
+% Python module: https://pypi.org/project/bio2zarr/
+% Operating system(s): Platform independent
+% Programming language: Python
+% License: Apache-2.0
+% Software Heritage PID: [207]
 

### 2025-04-01 paper.tex
Added software heritage references and metadata

@@
 
-% 3. GitHub repository bio2zarr
-% Project name:  bio2zarr
-% Github link: https://github.com/sgkit-dev/bio2zarr/
-% Python module: https://pypi.org/project/bio2zarr/
-% Operating system(s): Platform independent
-% Programming language: Python
-% License: Apache-2.0
-% Software Heritage PID: [207]
+\subsection{GitHub repository SAIGE-VCZ}
+\begin{description}
+\item[Project name:]  SAIGE-VCZ
+\item[Github link:] \url{https://github.com/sgkit-dev/SAIGE-VCZ/}
+\item[Operating system(s):] Platform independent
+\item[Programming language:] C++, R
+\item[License:] GPL-3.0
+\item[Software Heritage PID:]
+swh:1:snp:b868873e7e5dec417459cdf74de95647deb15167 \cite{saige-vcz-swh}
+\end{description}
+
+\subsection{GitHub repository vcztools}
+\begin{description}
+\item[Project name:]  vcztools
+\item[Github link:] \url{https://github.com/sgkit-dev/vcztools/}
+\item[Operating system(s):] Platform independent
+\item[Programming language:] Python, C
+\item[License:] Apache-2.0
+\item[Software Heritage PID:] 
+swh:1:snp:bdf1771ddb19144626df61a74cbe1e6fcdc634bc \cite{vcztools-swh}
+\item[RRID:] SCR\_026482
+\item[Bio.tools Unique Identifier:] vcztools
+\end{description}
 

### 2025-04-01 paper.tex
Closes #313
Finish up data access stuff

@@
 
-% The VCF Zarr specification is available on GitHub at
-% \url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
-% The SAIGE prototype is available at \url{https://github.com/Will-Tyler/SAIGE}
-% and all other code for running benchmarks, analyses and creating
-% plots in this article is available at
-% \url{https://github.com/sgkit-dev/vcf-zarr-publication}.
-% Vcf2zarr is freely available under the terms of the Apache 2.0
-% license as part of the \texttt{bio2zarr}
-% suite (\url{https://github.com/sgkit-dev/bio2zarr/}; 
-% biotoolsID=bio2zarr; RRID=SCR\_026481)
-% and can be installed from the Python Package Index
-% (\url{https://pypi.org/project/bio2zarr/}).
-% Vcztools (\url{https://github.com/sgkit-dev/vcztools/};
-% biotoolsID=vcztools; RRID=SCR\_026482)
-% is freely available under the terms of the Apache 2.0
-% license and can be installed from the Python Package Index
-% (\url{https://pypi.org/project/vcztools/}).
-
 \section{Data availability}
-
-% Your manuscript lacks a 'Data availability' section and we require it; could
-% you please create one and list all the datasets you have used in your case
-% studies along with direct URLs (better: DOIs, if available) to these datasets
-% (not the associated manuscript, but directly the dataset's main page, for user
-% accessibility purposes).
-The datasets used for the case studies in this article are described/available at:
-Genomics England aggV2: \url{https://re-docs.genomicsengland.co.uk/aggv2/};
-Future Health Data Release 9: 
-\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9};
-All of Us CDRv7:
-\url{gs://fc-aou-datasets-controlled/v7/wgs/short\_read/snpindel/acaf\_threshold_v7.1/vcf/},
-described at 
-\url{https://support.researchallofus.org/hc/en-us/articles/4616869437204-Controlled-CDR-Directory-Archived-C2022Q4R13-CDRv7};
-Norway Spruce: data release pending;
-SARS-CoV-2 Viridian: \url{https://doi.org/10.6084/m9.figshare.28360031.v2}.
+The public datasets used in this paper can be found by the following
+accessions:
+1) Genomics England aggV2~\cite{gel-aggv2} 
+which is under controlled access and researchers can apply for access
+here~\citep{gel-join};
+2) Our Future Health data release 9~\cite{ofhdatarelease9},
+which is under controlled access and researchers can apply for access here~\citep{ofh-apply-for-access};
+3) All of Us CDRv7~\citep{aou-cdrv7}
+which is under controlled access and researchers can apply for access
+here~\citep{aou-apply};
+4) The Norway Spruce dataset can be found here~\cite{plantgenie};
+5) the SARS-CoV-2 Viridian data \cite{kelleher2025viridian}.
 

### 2025-04-01 paper.tex
Make second author bucket
Fix up affils

@@
 \author[1,2\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
-\author[3,4\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
-\author[5,\authfn{1}]{Will Tyler}
-\author[6,\authfn{1}]{Tom White}
+\author[3,\authfn{1}]{Will Tyler}
+\author[4,\authfn{1}]{Tom White}
+
+\author[5\authfn{2}]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
+\author[6,7\authfn{2}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
 
 % Middle, alphabetical order
-\author[7]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
-\author[8,9]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
-\author[10]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
-\author[11]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
-\author[8,9,12]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
+\author[8]{Benjamin Elsworth} % https://orcid.org/0000-0001-7328-4233
+\author[9,10]{Jérémy Guez} % https://orcid.org/0009-0007-6406-5187
+\author[11]{Jonny Hancox} % https://orcid.org/0009-0003-5799-5299
+\author[9,10,12]{ Konrad J. Karczewski} % https://orcid.org/0000-0003-2878-4671
 \author[13]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
@@
 % Senior
-\author[1,2,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
-\author[11,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
+\author[1,2,\authfn{3}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
+\author[5,\authfn{3},\authfn{4}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253
 
@@
 \affil[2]{Related Sciences}
-\affil[3]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
+\affil[3]{Independent researcher}
+\affil[4]{Tom White Consulting Ltd.}
+\affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, UK}
+\affil[6]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
 New Zealand}
-\affil[4]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
-\affil[5]{Independent researcher}
-\affil[6]{Tom White Consulting Ltd.}
-\affil[7]{Our Future Health, Manchester, UK.}
-\affil[8]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
-\affil[9]{Analytic and Translational Genetics Unit, Massachusetts General
+\affil[7]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
+\affil[8]{Our Future Health, Manchester, UK.}
+\affil[9]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
+\affil[10]{Analytic and Translational Genetics Unit, Massachusetts General
 Hospital, Boston, Massachusetts 02114, USA}
-\affil[10]{NVIDIA Ltd, Reading, UK}
-\affil[11]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
-University of Oxford, UK}
+\affil[11]{NVIDIA Ltd, Reading, UK}
+
 \affil[12]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
@@
 \authnote{\authfn{1}Joint first author.}
-\authnote{\authfn{2}Joint senior author.}
-\authnote{\authfn{3}jerome.kelleher@bdi.ox.ac.uk}
+\authnote{\authfn{2}Joint second author.}
+\authnote{\authfn{3}Joint senior author.}
+\authnote{\authfn{4}jerome.kelleher@bdi.ox.ac.uk}
 
